### 2017-01-19 levin_etal/reanalyses.rmd
added levin files from https://github.com/caseywdunn/levin2016

---
output:
  github_document: default
---
# Levin *et al.* reanalyses

```{r setup, echo=FALSE, warning=FALSE, message=FALSE}

```

by [Casey W. Dunn](http://dunnlab.org/)

Please see the [git repository](https://github.com/caseywdunn/levin2016) for additional files related to this document, including data and [source code for all analyses](https://github.com/caseywdunn/levin2016/blob/master/reanalyses.rmd).

This notebook describes reanalyses of Levin *et al*. 2016:

> Levin M, Anavy L, Cole AG, Winter E, Mostov N, Khair S, Senderovich N, Kovalev E, Silver DH, Feder M, et al. 2016. The mid-developmental transition and the evolution of animal body plans. Nature 531: 637-641. [doi:10.1038/nature16994](http://dx.doi.org/10.1038/nature16994)

We addressed concerns regarding the study design of Levin *et al*. in a previous manuscript:

> Hejnol A, Dunn CW, 2016. Animal Evolution: Are Phyla Real? Current Biology 26:R424-R426. [doi:10.1016/j.cub.2016.03.058](http://dx.doi.org/10.1016/j.cub.2016.03.058)

Here we examine support for their central finding - the inverse hourglass model.

## Figure 4d data

This box plot corresponds to that of Levin *et al.* [Figure 4d](http://www.nature.com/nature/journal/v531/n7596/fig_tab/nature16994_F4.html):
```{r preprocess, echo=FALSE}
D = read.table("data_processed/Figure_4d_data.txt", header=T)

# Replace taxon indeces with taxon names

taxa = c("Porifera", "Nematodes", "Arthropods", "Platyhelminthes", "Tardigrades", "Cnidaria", "Annelides", "Echinodermata", "Chordata", "Ctenophores") # from Figure 4d data.xlsx
D$Species_A = as.factor( sapply(D$Species_A, function(x) taxa[x]) )
D$Species_B = as.factor( sapply(D$Species_B, function(x) taxa[x]) )

boxplot(as.matrix(D[,3:5]))
```

The KS p-value for the early-mid comparison is `r format(ks.test(D$Early, D$Mid)$p.value, digits=3)`, while the p-value for the late-mid comparison is `r format(ks.test(D$Late, D$Mid)$p.value, digits=3, scientific=TRUE)`.
The KS tests as reproduced here are still significant, but the p values are much less significant than those published by Levin *et al.*:

> we found that the mid-developmental transition profiles are significantly less conserved than the early and late phase expression (Fig. 4d, P<10-6 compared with the early phase and P<10-12 with the late phase, Kolmogorov–Smirnov test)

Question: How was the published P value calculated? Was it two sided? Was it on a different data set?

Levin *et al.* answer: They used a a two-sided KS test in matlab:

    [i,j]=kstest2(S(:,2),S(:,1)); ['P-value: early to switch: ' mat2str(j)]
    [i,j]=kstest2(S(:,2),S(:,3)); ['P-value: late to switch: ' mat2str(j)]

To compare my new results as closely as possible with their published results, I ran the following in matlab (version R2016a):

    # With duplicated values
    B = tdfread('Figure_4d_data_full.txt')
    [i,p]=kstest2(B.Early,B.Mid)
    # p = 3.1161e-06
    [i,p]=kstest2(B.Late,B.Mid)
    # p = 5.3257e-12

The matlab ks test p values are the same orders of magnitude as the published results (10<sup>-6</sup> for early-mid and 10<sup>-12</sup> for early-late) when analyzing the `Figure_4d_data_full.txt` file obtained from the authors. This shows I can reproduce their result. The problem, though, is that every pairwise comparison is represented twice in this file. Each species was compared to every other species, but both reciprocal comparisons (which have the same values) were retained. For example, there is an entry for both a `Nematodes` to `Arthropods` comparison and a `Arthropods` to `Nematodes` comparison. Both these comparisons have the same values. As a consequence, there are 90 entries for the 45 pairwise comparisons. These duplicate values inflates the p values by several orders of magnitude as it seems that twice the actual data are available.

I derived the `Figure_4d_data.txt` file from `Figure_4d_data_full.txt` by removing duplicate pairwise comparisons. The KS test on this non-redundant file returns p values that are orders of magnitude greater (i.e., less significant) than those published (0.002 for early-mid and 10<sup>-6</sup> for early-late):

    # Without duplicated values
    A = tdfread('Figure_4d_data.txt')
    [i,p]=kstest2(A.Early,A.Mid)
    # p = 0.0022
    [i,p]=kstest2(A.Late,A.Mid)
    # p = 2.5199e-06

The R and matlab results are consistent with each other when analyzing `Figure_4d_data.txt`, indicating that my results don't differ from those of Levin *et al.* due to software differences and that the discrepancy is due to the duplicated values.

### Evaluating the test method

The primary result of Levin *et al.*, *i.e.* support for the inverse hourglass model, is based on the box plot in their [Figure 4d](http://www.nature.com/nature/journal/v531/n7596/fig_tab/nature16994_F4.html) and the corresponding KS tests. In this figure, the box plots for each temporal phase largely overlap and there is little differentiation between the distributions. The published KS tests did reject the hypothesis that the early and late distributions are drawn from the same distribution as the mid phase, though the result was much less significant when correcting for duplicate values as described above. There are a couple fundamental problems using the KS test in this context, though:

- The KS test does not just test whether one distribution is greater than the other, it tests if the distributions are the same. This takes into consideration other features, such as distribution shape. This is a much more specific null hypothesis than that discussed in the paper (that they have the same mean). The null hypothesis would be rejected if the two distributions had the same mean but different variances, for example.

- The samples in this dataset are matched - for each pairwise comparison there is a early, mid, and late expression value. The KS test is for unmatched samples. In this study, the question is not whether the distributions are the same, as the KS test can evaluate, but whether each value in one phase (early, late) tends to be higher than its matched value in the mid phase. The Wilcoxon test is appropriate for this question.

Here I rerun the analyses taking the matched samples into account.

```{r matched_test, echo=FALSE}

# hist(D$Early - D$Mid)
w_early = wilcox.test(D$Early, D$Mid, alternative="greater", paired=T)$p.value

# hist(D$Late - D$Mid)
w_late = wilcox.test(D$Late, D$Mid, alternative="greater", paired=T)$p.value
```

The p-values for the Wilcoxon test are `r format(w_early, digits=3)` for the early-mid comparison, which is barely significant, and `r format(w_late, digits=3, scientific=TRUE)` for the late-mid comparison.

Levin *et al.* comment: "I think you make a good point regarding the KS test, that significance could result even when the means are the same. We used it in the paper as two-sided non-parametric test to quantify a P-value for the distributions which we found to have different means (Figure 4d)."

### Evaluation of support across taxa

Their [Figure 4d](http://www.nature.com/nature/journal/v531/n7596/fig_tab/nature16994_F4.html) shows that the the medians are close in value for the three phases, but the mid phase includes three low outliers. These outliers are much lower than any values in the other phases.

Here I order the matrix by Mid to identify these values:

```{r sorted_mid}
D[order(D[,4]),]

```

The five lowest mid values, including the three outliers, are pairwise comparisons that include the ctenophore. Here I rerun the analyses without ctenophores:

```{r no_cteno}
D_no_ct = D[ ((D$Species_A != "Ctenophores") &   (D$Species_B != "Ctenophores")), ]
boxplot(as.matrix(D_no_ct[,3:5]))
wilcox.test(D_no_ct$Early, D_no_ct$Mid, alternative="greater", paired=T)
wilcox.test(D_no_ct$Late, D_no_ct$Mid, alternative="greater", paired=T)

```

There is no significant difference between the early and mid phase.

### Summary figures

```{r echo=FALSE}

# Multiple plot function
#
# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
#
# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
# then plot 1 will go in the upper left, 2 will go in the upper right, and
# 3 will go all the way across the bottom.
#
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}

```

```{r summary, echo=FALSE, message=FALSE}

library(ggplot2)
library(reshape2)
library(grid)

limits=c(-0.4, 0.75)

Dm = melt(D, id=c("Species_A", "Species_B"), variable.name = "phase", value.name = "similarity")
Dm_no_ct = melt(D_no_ct, id=c("Species_A", "Species_B"), variable.name = "phase", value.name = "similarity")

has_cteno = (Dm$Species_A == "Ctenophores") | (Dm$Species_B == "Ctenophores")
alpha = rep(0.4, length(has_cteno))
alpha[has_cteno] = 0.9
color = rep("black", length(has_cteno))
color[has_cteno] = "red"

pa = ggplot(Dm, aes(phase, similarity)) +
    geom_boxplot(outlier.shape=NA)+ # Outliers are drawn in bellow with jitter, so supress them here so they aren't duplicated
    scale_y_continuous(limits=limits) +
    geom_jitter(shape=16, position=position_jitter(0.2), alpha=alpha, color=color)

pa = pa + coord_flip() + ggtitle("(a) Original, with ctenophore in red")

pb = ggplot(Dm_no_ct, aes(phase, similarity)) +
    geom_boxplot(outlier.shape=NA)+ # Outliers are drawn in bellow with jitter, so supress them here so they aren't duplicated
    scale_y_continuous(limits=limits) +
    geom_jitter(shape=16, position=position_jitter(0.2), alpha=0.4)

pb = pb + coord_flip() + ggtitle("(b) New, without ctenophore")

#pa
#ggplotly()

multiplot(pa, pb, cols=1)

pdf("Figure_levin_raw.pdf", width=4, height=4)
multiplot(pa, pb, cols=1)
dev.off()

# Plots without points
pa = ggplot(Dm, aes(phase, similarity)) + geom_boxplot()  + ggtitle("(a) Original, with ctenophore") + scale_y_continuous(limits=limits) + coord_flip()
pb = ggplot(Dm_no_ct, aes(phase, similarity)) + geom_boxplot() + ggtitle("(b) New, without ctenophore") +  scale_y_continuous(limits=limits) + coord_flip() 
#multiplot(pa, pb, cols=1)

```

### Overview of comparisons

<a name="overview"></a>

Based in the results shown in their [Figure 4d](http://www.nature.com/nature/journal/v531/n7596/fig_tab/nature16994_F4.html), Levin *et al.* concluded that the "mid" distribution was significantly less than the "early" and "late" distributions, and that a low "mid" value is a general evolutionary pattern for animals.

When all the annotated points are shown in the interactive plot below, however, it becomes apparent that minimum similarity at the "mid" point is not a general pattern. Many pairwise comparisons show other patterns, and the low values in the "mid" distribution are dominated by comparisons that include the ctenophore (shown in red).

```{r matched, echo=FALSE}

ggplot(data=Dm, aes(x=phase,y=similarity, group=interaction(Species_A, Species_B), text=paste(Species_A, Species_B, sep="-")))  + geom_line(size=0.5) + geom_point(color=color)

inverse_hourglass = D[ (D$Mid<D$Early) & (D$Mid < D$Late),]

```

Only `r nrow(inverse_hourglass)` of the `r nrow(D)` comparisons show an inverse hourglass pattern. Of these, `r nrow(inverse_hourglass[inverse_hourglass$Species_B == "Ctenophores",])` are for pairs that include the ctenophore.

The results for the subset of comparisons that show the inverse hourglass pattern is below.

```{r matched_table, echo=FALSE}

inverse_hourglass

table(c(as.character(inverse_hourglass$Species_A), as.character(inverse_hourglass$Species_B)))

```

## Figure 4c data

These data are an upstream intermediate that the 4d data are derived from. I explored these data a bit to better understand the broader analysis workflow.

```{r preprocess_4c}
Phase = as.matrix(read.table("data_processed/Figure_4c_mode.txt", header=T))
Ortho = as.matrix(read.table("data_processed/Figure_4c_orthology.txt", header=T))

dim(Phase)
dim(Ortho)
```

First, check to see if there are any elements that are 0 in one table but not the other.

```{r comare_4c}
sum((Phase>0) != (Ortho>0))

table(Phase)

nrow(Phase[rowSums(Phase)>0,]) # number of nonzero rows

```

There aren't. This verifies that every annotated ortholog present in each species is categorized as belonging to one of the three phases.

From Levin et al:

> To compare gene expression across these species, we delineated 11,139 orthologous protein families, with each orthologous family having representatives from an average of six species.

Question: What is the source of the discrepancy between the reported number 11,139 in the manuscript and the number of genes (ie, rows) in this Figure_4c matrix, ie `r nrow(Phase[rowSums(Phase)>0,])`? Were there additional filters applied?

Levin *et al.* answer: "The 11,139 refers to the number ortholog families, not the number of genes (rows)."

### C. elegans and D. melanogaster comparison

Here I try to replicate the example described by Levin et al.:

> Finally, we measured the extent of evolutionary change within the two conserved phases and the mid-developmental transition by determining whether orthologues annotated for a particular temporal category in one species are also annotated to the same temporal category in another species. Figure 4c shows an example of this analysis for D. melanogaster and C. elegans. For 4,395 orthologues delineated between these two species, the early phase, mid-developmental transition, and the late phase expression account for 51%, 14%, and 35% of the C. elegans orthologues, respectively. A total of 28% of the orthologues are annotated to the early phase in both C. elegans and D. melanogaster, while by chance only 22% are expected given the fraction of genes in each category across the species (Fig. 4c). In contrast, 3% were expected to be conserved at the mid-developmental transition at random, and 3% were observed. The log-odds ratios between observed and expected for the early phase and the mid-developmental transition between C. elegans and D. melanogaster are thus 0.35 and 0, respectively.

```{r DC_pre}
# Sample the columns for these two species
DC = Phase[,c(2,3)]
colnames(DC)

# sample only rows that have phase annotations for both

DC = DC[(DC[,1]>0) & (DC[,2]>0),]
nrow(DC)
```

So the number of genes selected is consistent with the reported number.

Now calculate the fraction of genes in each phase for each species.

```{r frac_4c_ce}
colnames(DC)[1]
ce = table(DC[,1])/sum(table(DC[,1]))
ce
```

These values do correspond to the 51%, 14%, and 35% reported in the manuscript for C. elegans.

```{r frac_4c_dm}
colnames(DC)[2]
dm = table(DC[,2])/sum(table(DC[,2]))
dm
```

```{r scores}
expected = dm * ce

observed = c( sum(DC[,1] ==1 & DC[,2] ==1), sum(DC[,1] ==2 & DC[,2] ==2), sum(DC[,1] ==3 & DC[,2] ==3) )
observed = observed / nrow(DC)
observed

observed/expected

log(observed/expected, 2) # have not yet verified the base

```

## Software versions

This document can be executed to an html file in RStudio (or with the `knitr` package at the command line).

This document was computed on `r format(Sys.time(), "%a %b %d %X %Y")` with the following R package versions.

```{r session_summary, echo=FALSE, comment=NA}
	sessionInfo()
```

### 2017-01-19 manuscript.rmd
added Levin material

---
bibliography: manuscript.bib
csl: nature.csl
output:
  github_document: default
---

```{r setup, cache=TRUE, message=F, echo=FALSE, warning=F}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Here we reconsider two recent comparative analyses of gene expression data [@KryuchkovaMostacci:2016iw; @Levin:2016bp].

## KMRR Reanalysis

```{r preliminaries, message=F, warning=F}
	library( treeio )
	library( tidyverse )
	library( magrittr )
	library( stringr )
	library( parallel )
	library( ggtree )
	library( ape )
	library( digest )
	library( hutan )

	# Set system computational parameters
	cores = detectCores() - 1
	if ( cores < 1 ) {
		cores = 1
	}
	set.seed( 23456 )

	# Set biology starting parameters, including files

	gene_tree_name = "Compara.75.protein.nhx.emf"

	kmrr_directory = "kmrr/"

	expression_file_names = c(
		"ChickenBrawandTScomparisonTable_9_6chPC.txt", 
		"ChimpBrawandTScomparisonTable_9_6cmPC.txt", 
		"GorillaBrawandTScomparisonTable_9_6gPC.txt", 
		"HumBrawandTScomparisonTable_9_6hPC.txt", 
		"MacacaBrawandTScomparisonTable_9_6mcPC.txt", 
		"MusBrawandTScomparisonTable_9_6mPC.txt", 
		"OpossumBrawandTScomparisonTable_9_6oPC.txt", 
		"PlatypusBrawandTScomparisonTable_9_6pPC.txt"
	)

	expression_file_names %<>% paste(kmrr_directory, ., sep="")

	gene_tree_name %<>% paste(kmrr_directory, ., sep="")

	# These are from the KRR code
	calibration_times = data.frame(
		age.min =
			c(20, 92, 167, 8, 42, 74, 296, 535, 104, 937, 29, 722, 441, 65, 15, 1215, 
				414, 371, 162, 25, 74, 77, 86), 
		clade = c("Hominoidea", "Euarchontoglires", "Mammalia", "Homininae", 
							"Simiiformes", "Primates", "Amniota", "Vertebrata", "Eutheria", 
							"Bilateria", "Catarrhini", "Chordata", "Euteleostomi", 
							"Haplorrhini", "Hominidae", "Opisthokonta", "Sarcopterygii", 
							"Tetrapoda", "Theria", "Murinae", "Sciurognathi", "Rodentia", "Glires"),
		stringsAsFactors=FALSE
	)

	# Functions

	# Returns a logical vector indicating which rows in nhx_tags are for tips
	is.tip.nhx = function( nhx ) {
	  is.tip = rep( FALSE, nrow( nhx@data ) )
	  is.tip[ 1:length( nhx@phylo$tip.label ) ] = TRUE
	  is.tip
	}

	# put node ages in @data
	store_node_age = function( nhx ) {

	  if ( class( nhx ) != "treedata" ) {
	    return( nhx )
	  }

	  node_age = hutan::distance_from_tip( nhx@phylo )

	  # make sure the dataframe is ordered by consecutive nodes
	  stopifnot( all( nhx@data$node == 1:length( nhx@data$node ) ) )

	  nhx@data$node_age = node_age

	  return( nhx )
	}

	# Make tree ultrametric, calibrating speciation nodes to consistent times
	calibrate_tree = function ( nhx, calibration_times ) {

			# Create calibration matrix for speciation nodes
			calibration = 
				nhx@data[ !is.tip.nhx( nhx ) ,  ] %>%
				filter( D=="N" ) %>%
				left_join( calibration_times, c( "label" = "clade" ) ) %>%
				select( node, age.min ) %>%
				mutate( age.max = age.min ) %>% 
				mutate( soft.bounds = NA )

			tree = try( ape::chronos( nhx@phylo, calibration=calibration, model="correlated" ) )
			if( "phylo" %in% class(tree) ){
				class( tree ) = "phylo"
				nhx@phylo = tree
				return( nhx )
			}
			else{
				return( NA )
			}
		}

```

```{r load_expression, message=F, warning=F}
	expression = lapply(
		expression_file_names, read.table, stringsAsFactors=FALSE, header=TRUE
	) %>%
	bind_rows()
```

```{r load_trees, cache=TRUE, message=F, warning=F}
	lines = readLines( gene_tree_name )

	# Only consider the lines with trees
	lines = lines[ grepl( "^\\(", lines, perl = TRUE ) ]

	gene_trees = mclapply(
		lines,
		function( x ){
			tree_tc = textConnection( x )
			tree <- treeio::read.nhx( tree_tc )
			tree@data$label = c( tree@phylo$tip.label, tree@phylo$node.label )
			return( tree )
		}, mc.cores=cores
	)
```

There were `r length(gene_trees)` parsed.

```{r add_expression_to_trees, cache=TRUE, message=F, warning=F}
	gene_trees_annotated = mclapply(
		gene_trees,
		function( tree ){
			tree@data %<>% left_join( expression, by = c("G" = "Ensembl.Gene.ID") )
			return( tree )
		}, mc.cores=cores
	)
```

```{r prep_trees, cache=TRUE, message=F, warning=F}
	drop_empty_tips = function( nhx ) {
	  to_drop = which( is.na( nhx@data[ 1:length(nhx@phylo$tip.label), ]$Tau ) )
	  remaining = length(nhx@phylo$tip.label) - length(to_drop)

	  if( remaining > 3 ){
	  	pruned = treeio::drop.tip( nhx, to_drop )
	  	return( pruned )
	  }
	  else{
	  	return( NA )
	  }
	}

	gene_trees_sub = mclapply( gene_trees_annotated, drop_empty_tips, mc.cores=cores )
	gene_trees_sub = gene_trees_sub[ ! is.na( gene_trees_sub ) ]
#	gene_trees_sub = gene_trees_sub[ lapply(gene_trees_sub, class) == "treedata" ]

	gene_trees_calibrated = mclapply( 
		gene_trees_sub,
		calibrate_tree, 
		calibration_times=calibration_times,
		mc.cores=cores
	)

	gene_trees_calibrated = gene_trees_calibrated[ ! is.na( gene_trees_calibrated ) ]

	gene_trees_calibrated = mclapply( gene_trees_calibrated, store_node_age, mc.cores=cores )

```

`r length(gene_trees_sub)` trees passed taxon sampling criteria.

`r length(gene_trees_calibrated)` were time calibrated.

```{r contrasts, cache=TRUE, message=F, warning=F}

	pic.nhx = function( nhx ) {

	  p = pic( nhx@data$Tau[ is.tip.nhx( nhx ) ], nhx@phylo, var.contrasts=TRUE )

	  nhx@data$pic = c( rep( NA, length( nhx@phylo$tip.label ) ), p[ ,1 ] )
	  nhx@data$var_exp = c( rep( NA, length( nhx@phylo$tip.label ) ), p[ ,2 ] )

	  return( nhx )
	}

	gene_trees_pic = mclapply( gene_trees_calibrated,  pic.nhx, mc.cores=cores )

	all_nodes = lapply( 
	      gene_trees_pic, 
	  		function( nhx ){
	  			tags = nhx@data
	  			tags$gene = digest( nhx )
	  			tags %<>% select( -(B) ) # B has inconsistent types, remove it
	  			return(tags)
	  		}
  		) %>%
		bind_rows()

```

```{r show_result, cache=TRUE, message=F, warning=F}

	nodes_contrast = 
		all_nodes %>%
		filter( ! is.na(pic) ) %>%
		mutate( pic=abs(pic) ) %>%
		filter( ! is.na(D) )

	nodes_contrast %>%
		ggplot( aes( x=pic, y=..density.., col=D ) ) + 
        geom_freqpoly( binwidth=.001, position="identity" ) +
        xlab("Phylogenetic independent \ncontrasts (PIC)") +
				xlim(0,0.075)

	nodes_contrast %>%
		filter( pic < 0.2 ) %>%
		ggplot( aes( x=node_age, y=pic, col=D ) ) +
		geom_point( alpha=0.2) +
		xlim(0,1000)

	wilcox.test( nodes_contrast %>% filter(D=="Y") %>% .$pic, nodes_contrast %>% filter(D=="N") %>% .$pic, alternative="greater" )

```

There were `r nodes_contrast %>% filter(D=="Y") %>% nrow()` duplication nodes and `r nodes_contrast %>% filter(D=="N") %>% nrow()` speciation nodes.

```{r pairwise, cache=TRUE, message=F, warning=F}
	get_pairwise_summary = function ( nhx ) {

		# Create a tibble with one row for each pairwise combination of tip nodes
		ntips = length( nhx@phylo$tip.label )
		D = 1:ntips %>% combn( 2 ) %>% t() %>% as_tibble()
		names( D ) = c( "tip_a", "tip_b" )

		# Add a colmn with the gene tree hash 
		D %<>% mutate( gene_tree = digest( nhx ) )

		# Calculate the phenotypic difference between the tips
		D$difference = 
			apply(
				D,
				1,
				function(y) {
					tau_values = nhx@data$Tau[ c( as.integer(y['tip_a']), as.integer( y['tip_b'] ) ) ]
					max( tau_values ) - min( tau_values )
				}
			)

		# For each row, bet the most recent common ancestor of the pairwise tip combination
		D$mrca = 
			apply(
				D, 
				1, 
				function(y) ape::getMRCA( 
					nhx@phylo, 
					c( as.integer(y['tip_a']), as.integer(y['tip_b'] ))
				)
			) %>%
			as.integer()

		# Merge data on this mrca to the node
		D %<>% left_join(
			nhx@data %>%
				select( D, node, node_age, pic ),
			by = c( "mrca" = "node")

		)

		# Calculate the distance between the tips, which for an ultrametric tree 
		# is twice the age of their most recent common ancestor
		D %<>% mutate( distance = 2 * node_age )

		return( D )

	}

	# Build a tibble of all pairwise comparisons between tips
	pairwise_summary = 
		mclapply( gene_trees_pic, get_pairwise_summary, mc.cores=cores ) %>% 
		bind_rows()

	pairwise_summary %>% 
		filter( ! is.na(D) ) %>%
  	ggplot(aes(x=distance, y=difference, color=D)) + 
  		geom_point(alpha=0.1) +
  		geom_smooth(method=lm, se=FALSE) +
			xlim( 0, 2000 )

```

## Levin *et al.* reanalysis

How changes in animal development relate to the evolution of animal diversity is a major question in evolutionary developmental biology (EvoDevo). To address this topic, Levin *et al.* <sup>1</sup> analyzed gene expression through the course of embryonic development for ten animal species, each from a different phylum. As their title ("The mid-developmental transition and the evolution of animal body plans") indicates, they arrived at two major conclusions. First, animal development is characterized by a well-defined mid-developmental transition. Second, this transition helps explain the evolution of features observed among distantly related animals. Specifically, they concluded that animals from different phyla exhibit an "inverse hourglass" model for the evolution of gene expression, where there is more evolutionary variance in gene expression at a mid phase of development than there is at early and late phases. Closely related animals have previously been described as having an hourglass model of gene expression, where evolutionary variance in expression is greater early and late in development than at the midpoint of development <sup>2,3</sup>. Levin *et al.* conclude that this contrast between distantly and closely related animals provides biological justification for the concept of phyla and may provide a long-sought operational definition of phyla. We previously described some concerns with their interpretations of these results <sup>4</sup>. Here we directly address problems with the analyses behind their evolutionary conclusions. Rather than support the inverse hourglass as a general pattern among distantly related species, our reanalyses of Levin *et al.* <sup>1</sup> indicate their existing data support the opposite conclusion - that the inverse hourglass is due to large differences specific to a single lineage, the ctenophore (comb jelly). Better understanding these differences will be of interest to future work, especially in light of current uncertainty regarding the phylogenetic placement of ctenophores and their unique biology <sup>5,6</sup>.

To understand our concerns with their analyses, it is helpful to first outline Levin *et al.*'s published methods and results. Although there are mature methods and tools for statistical analysis of character evolution <sup>7</sup> and gene expression <sup>8</sup>, Levin *et al.* <sup>1</sup> drew on neither of these very active areas of research and instead applied an *ad hoc* pairwise comparison of orthologous gene expression between species. They characterized each gene in each species as having expression that peaks in early, mid, or late temporal phase of development while the goodness of fit to these patterns was not considered, and alternative patterns were not evaluated. For each species pair, they then identified the orthologs shared by these species (shared orthologs vary from pair to pair). They then calculated a similarity score for each temporal phase for each species pair based on the fraction of genes that exhibited the same patterns in each species. The distributions of similarity scores are plotted in their [Figure 4d](http://www.nature.com/nature/journal/v531/n7596/fig_tab/nature16994_F4.html), and their Kolmogorov–Smirnov (KS) tests indicated that the early distribution and late distribution were each significantly different from mid distribution (P < 10<sup>-6</sup> and P < 10<sup>-12</sup>, respectively). This is the support they presented for the inverse hourglass model.

We began by examining the matrix of pairwise comparisons that their KS tests and Figure 4d are based on. As this figure shows (regenerated here as our Figure 1a), there is not a strong distinction between these temporal phases - the distributions for the early, mid, and late distributions overlap considerably. The medians are close in value, and the range of mid phase includes the entire range of the early phase. Further inspection (Figure 1a) reveals that the mid phase distribution has several outliers with very low similarity scores. We found that all five of the lowest values in the mid phase distribution are for pairwise comparisons that include the ctenophore (Figure 1a). When the nine pairwise comparisons that include the ctenophore are removed, the differences between the early and mid phase distributions are greatly reduced (Figure 1b).

We also found several problems with the statistical tests that were used to evaluate the inverse hourglass hypothesis. First, in the published analyses every data point was included twice because both reciprocal comparisons (which have the same values) were retained. For example, there is both a nematode to arthropod comparison and an arthropod to nematode comparison. As a consequence, there are 90 entries for the 45 pairwise comparisons, and by doubling the data the significance of the result appears much stronger than it actually is. After removing the duplicate values, the p values are far less significant, 0.002 for the early-mid comparison and on the order of 10<sup>-6</sup> for early-late. Second, the test they used (KS test) is not appropriate for the hypothesis they seek to evaluate. The KS test does not just evaluate whether one distribution is greater than the other, it also tests whether the shape of the distributions are the same. In addition, the samples in this dataset are matched (*i.e.*, for each pairwise comparison there is a early, mid, and late expression value), which the KS test does not take into account. The Wilcoxon test is instead appropriate in this case. After removing the duplicate scores, removing the ctenophore, and applying the Wilcoxon test, there is no significant difference between the early phase and mid phase distributions (P  = 0.1428 for the early-mid comparison and P < 10<sup>-5</sup> for the late-mid comparison), and the inverse hourglass turns into a wine bottle (Figure 1b).

Our results highlight the importance of explicitly accounting for phylogenetic relationships when studying character evolution, including developmental <sup>9</sup> and functional genomic traits <sup>10</sup>. This is particularly true for evolutionary analyses of quantitative gene expression <sup>11</sup>. Pairwise comparisons result in the same evolutionary changes being counted multiple times in each species pair, as for the ctenophore comparisons that were mistaken for a global pattern here. This is a well understood property of pairwise comparisons that has been specifically addressed by phylogenetic comparative methods <sup>12</sup>. This particular case illustrates the shortcomings of *ad hoc* rank-based methods <sup>3</sup> and pairwise comparisons <sup>1</sup> for the study of the evolution of gene expression. Phylogenetic analyses of character evolution provide not only a way to avoid these problems, but a richer context for interpreting the biological implications of the results.

![Figure 1](levin_etal/Figure_levin.png?raw=true)

> Figure 1 | Distributions of pairwise similarity scores for each phase of development. Pairwise scores for the ctenophore are red. Wilcoxon test p-values for the significance of the differences between early-mid distributions and late-mid distributions are on the right. Model of variance, which is inversely related to similarity, is on the left. (a) The distributions as published. Low similarity (*i.e.*, high variance) in the mid phase of development was interpreted as support for an inverse hourglass model for the evolution of gene expression. The five least-similar mid phase scores were all from the ctenophore. Published KS p-values, based on duplicated data, are in parentheses. The inset ctenophore image is by S. Haddock from phylopic.org. (b) The distributions after the exclusion of the ctenophore. The early and mid phase distributions are not statistically distinct. This suggest a wine bottle model, with similar evolutionary variance at the early and mid phase and less at the late phase.

## Methods

### KMRR reanalysis

Based on their scripts from: https://figshare.com/articles/Tissue-specificity_of_gene_expression_diverges_slowly_between_orthologs_and_rapidly_between_paralogs/3493010/2

To replciate the published analysis, `cd` to `Data_used/Expression_data`, launch `R`, and run:

    source( "../Rscript.R" )

After running the analysis above, also obtain the `Compara.75.protein.nh.emf` gene trees from ftp://ftp.ensembl.org/pub/release-75/emf/ensembl-compara/homologies/ and place them in the same directory as this file.

### Levin et al. reanalysis

Levin *et al.* helpfully provided data and clarification on methods. We obtained the matrix of pairwise scores that underlies their [Figure 4d](http://www.nature.com/nature/journal/v531/n7596/fig_tab/nature16994_F4.html) and confirmed we could reproduce their published results. We then removed duplicate rows, identified ctenophores as overrepresented among the low outliers in the mid-developmental transition column, and applied the Wilcoxon test in place of the Kolmogorov-Smirnov test. These data, our analysis code, and additional information are available in a git repository at [https://github.com/caseywdunn/levin2016](https://github.com/caseywdunn/levin2016). This repository includes our analysis notebook at [https://rawgit.com/caseywdunn/levin2016/master/reanalyses.html](https://rawgit.com/caseywdunn/levin2016/master/reanalyses.html).

## Ackowledgements

## References

### 2017-01-19 manuscript.rmd
updated hard citations to bib

@@
 ```
+
+# Comparing gene expression across species
+
+Dunn, Zapata, Munro, Siebert, Hejnol
 
@@
 How changes in animal development relate to the evolution of animal diversity is a major question in evolutionary developmental biology (EvoDevo).
-To address this topic, Levin *et al.* <sup>1</sup> analyzed gene expression through the course of embryonic development for ten animal species, each from a different phylum.
+To address this topic, Levin *et al.* [@Levin:2016bp] analyzed gene expression through the course of embryonic development for ten animal species, each from a different phylum.
 As their title ("The mid-developmental transition and the evolution of animal body plans") indicates, they arrived at two major conclusions.
@@
 Specifically, they concluded that animals from different phyla exhibit an "inverse hourglass" model for the evolution of gene expression, where there is more evolutionary variance in gene expression at a mid phase of development than there is at early and late phases.
-Closely related animals have previously been described as having an hourglass model of gene expression, where evolutionary variance in expression is greater early and late in development than at the midpoint of development <sup>2,3</sup>.
+Closely related animals have previously been described as having an hourglass model of gene expression, where evolutionary variance in expression is greater early and late in development than at the midpoint of development [@Kalinka:2010dw; DomazetLoso:2010cn].
 Levin *et al.* conclude that this contrast between distantly and closely related animals provides biological justification for the concept of phyla and may provide a long-sought operational definition of phyla.
-We previously described some concerns with their interpretations of these results <sup>4</sup>.
+We previously described some concerns with their interpretations of these results [@Hejnol:2016hw].
 Here we directly address problems with the analyses behind their evolutionary conclusions.
-Rather than support the inverse hourglass as a general pattern among distantly related species, our reanalyses of Levin *et al.* <sup>1</sup> indicate their existing data support the opposite conclusion - that the inverse hourglass is due to large differences specific to a single lineage, the ctenophore (comb jelly).
-Better understanding these differences will be of interest to future work, especially in light of current uncertainty regarding the phylogenetic placement of ctenophores and their unique biology <sup>5,6</sup>.
+Rather than support the inverse hourglass as a general pattern among distantly related species, our reanalyses of Levin *et al.* [@Levin:2016bp] indicate their existing data support the opposite conclusion - that the inverse hourglass is due to large differences specific to a single lineage, the ctenophore (comb jelly).
+Better understanding these differences will be of interest to future work, especially in light of current uncertainty regarding the phylogenetic placement of ctenophores and their unique biology [@Martindale:2015ve, Dunn:2015ij].
 
 To understand our concerns with their analyses, it is helpful to first outline Levin *et al.*'s published methods and results.
-Although there are mature methods and tools for statistical analysis of character evolution <sup>7</sup> and gene expression <sup>8</sup>, Levin *et al.* <sup>1</sup> drew on neither of these very active areas of research and instead applied an *ad hoc* pairwise comparison of orthologous gene expression between species.
+Although there are mature methods and tools for statistical analysis of character evolution [@Garamszegi:2014ew] and gene expression [@Robinson:2009cw], Levin *et al.* [@Levin:2016bp] drew on neither of these very active areas of research and instead applied an *ad hoc* pairwise comparison of orthologous gene expression between species.
 They characterized each gene in each species as having expression that peaks in early, mid, or late temporal phase of development while the goodness of fit to these patterns was not considered, and alternative patterns were not evaluated.
@@
 
-Our results highlight the importance of explicitly accounting for phylogenetic relationships when studying character evolution, including developmental <sup>9</sup> and functional genomic traits <sup>10</sup>.
-This is particularly true for evolutionary analyses of quantitative gene expression <sup>11</sup>.
+Our results highlight the importance of explicitly accounting for phylogenetic relationships when studying character evolution, including developmental [@Telford:2003tsa] and functional genomic traits [@Hejnol:2015gr].
+This is particularly true for evolutionary analyses of quantitative gene expression [@Dunn:2013cpa].
 Pairwise comparisons result in the same evolutionary changes being counted multiple times in each species pair, as for the ctenophore comparisons that were mistaken for a global pattern here.
-This is a well understood property of pairwise comparisons that has been specifically addressed by phylogenetic comparative methods <sup>12</sup>.
-This particular case illustrates the shortcomings of *ad hoc* rank-based methods <sup>3</sup> and pairwise comparisons <sup>1</sup> for the study of the evolution of gene expression.
+This is a well understood property of pairwise comparisons that has been specifically addressed by phylogenetic comparative methods [@Felsenstein:1985ua].
+This particular case illustrates the shortcomings of *ad hoc* rank-based methods [@DomazetLoso:2010cn] and pairwise comparisons [@Levin:2016bp] for the study of the evolution of gene expression.
 Phylogenetic analyses of character evolution provide not only a way to avoid these problems, but a richer context for interpreting the biological implications of the results.
@@
 
+Thanks to
+
 ## References

### 2017-01-21 manuscript.rmd
added some text

@@
 
-# Comparing gene expression across species
+# The importance of phylogenetic methods when comparing functional genomic data across species
 
@@
 
+Casey W. Dunn^1^*, Felipe Zapata^1,2^, Cat Munro^1^, Stefan Siebert^1,3^, Andreas Hejnol^4^
+
+^1^ Department of Ecology and Evolutionary Biology, Brown University, Providence, RI, USA
+
+^2^ Current address: Department of Ecology and Evolutionary Biology, University of California Los Angeles, Los Angeles, CA, USA
+
+^3^ Current address: Department of Molecular & Cellular Biology, University of California at Davis, Davis, CA, USA
+
+^4^ Sars Institute, Bergen, Norway
+
+\* Corresponding author, casey_dunn@brown.edu
+
+## Abstract
+
+Functional genomics, the study of how genomes work, is one of the most rapidly advancing fields in Biology.
+There is now considerable interest in comparing genome function across species to understand what is conserved, what is variable, and how the diversity of organism traits (including anatomy, physiology, and development) relates to genome diversity.
+Most published comparisons of genome function across species have, however, neglected to use methods that consider the evolutionary relationships among species.
+This has two potential negative impacts.
+First, it can lead to the wrong conclusions about the evolution of genome function.
+Second, it is a missed opportunity to learn about biology that can only be understood in an explicit historical context that considers these relationships.
+Here we examine two recently published comparative expression studies that used multiple pairwise comparison rather than phylogenetic approaches to investigate evolutionary patterns of gene expression.
+We find problems with the pairwise comparisons used in both studies that undermine their published conclusions.
+This is a concrete demonstration of the inadequacy of pairwise comparisons in comparative genomics studies, and indicates that it will be critical to adopt phylogenetic comparative methods in future analyses.
+
 ## Introduction
 
+The focus of genomics research has quickly shifted from describing genome sequences to Functional Genomics (FG), the study of how genomes "work" using tools that measure functional attributes such as expression, chromatin state, and transcription initiation.
+FG, in turn, is now becoming more comparative- there is great interest in understanding how FG variation across species gives rise to a diversity of development, morphology, physiology, and other phenotypes [@Wray:2013gh].
+Comparative Functional Genomics (CFG) analyses are also critical to transferring functional insight across species, and will grow in importance in coming years.
+
+A rich theoretical and statistical methodology of phylogenetic comparative methods have been developed over the last three decades to address the statistical challenges and opportunities of comparisons across species [@Felsenstein:1985ua; Grafen:1989um; Chamberlain:2012vx; Garamszegi:2014ew; Pagel:1999gc].
+These have largely been applied to the evolution of morphological and ecological traits, but are just as relevant to CFG.
+Most CFG studies have abstained from phylogenetic approaches and instead heavily rely on pairwise comparisons (Fig.
+1A) that do not account for the fact that evolutionary relationships explain much of the structure of variation across species.
+This leaves CFG studies susceptible to serious statistical errors and is a missed opportunity to ask questions that are only accessible in an explicit phylogenetic context.
+
+One reason that CFG has not yet embraced phylogenetic approaches is that it has not yet been concretely demonstrated that pairwise and phylogenetic approaches can lead to different results in CFG.
+Although the value of phylogenetic approaches has been repeatedly shown in comparative analyses of other character data [@Chamberlain:2012vx; Ricklefs:1996wb], this value is apparently not well known to the FG community since the field has until recently largely focused on within-species analyses.
+Many initial CFG studies have also been focused on only two species at a time, where the differences between phylogenetic and pairwise approaches are not as pronounced.
+As the number of species considered in each CFG study grows, the shortcomings of pairwise approaches will become far more acute.
+
+Most CFG studies to date compare gene expression across species, often between distinct tissues or developmental time points.
+Gene expression has been a popular target of CFG studies because, with the advent of RNA-Seq, it is the most readily collected FG trait across diverse species.
 Here we reconsider two recent comparative analyses of gene expression data [@KryuchkovaMostacci:2016iw; @Levin:2016bp].
 
-## KMRR Reanalysis
+Kryuchkova-Mostacci and Robinson-Rechavi (KMRR) 2016 [@KryuchkovaMostacci:2016iw] integrated vertebrate expression datasets to test the ortholog conjecture - the hypothesis that orthologs tend of have more conserved attributes (expression in this case) than do paralogs [@Nehrt:2011bm].
+Using pairwise comparisons (Fig.
+1A), they found lower expression covariance between paralogs than between orthologs and interpreted this as strong support for the ortholog conjecture.
+Orthology and paralogy is annotated based on gene phylogenies, but these phylogenies are not used when making the expression comparisons.
 
-```{r preliminaries, message=F, warning=F}
+The second CFG study we evaluate here is Levin *et al.* [@Levin:2016bp].
+This study analyzed gene expression through the course of embryonic development for ten animal species, each from a different phylum.
+They concluded that animals from different phyla exhibit an "inverse hourglass" model for the evolution of gene expression, where there is more evolutionary variance in gene expression at a mid phase of development than there is at early and late phases (Fig.
+2).
+This contrasts with the "hourglass" model previously proposed for closely related species [@Kalinka:2010dw].
+Levin *et al.* conclude that this provides biological justification for the concept of phyla and may provide a definition of phyla.
+We previously described concerns with their interpretations of these results [@Hejnol:2016hw].
+Here we directly address the analyses themselves by examining the structure of the pairwise comparisons.
+
+![Figure Overview](Figure_overview.svg?raw=true)
+
+> Figure 1 | Pairwise and phylogenetic comparative approaches.
+(a) Many comparative functional genomic studies across species rely on pairwise comparisons, where traits of each gene are compared to traits of every other gene.
+This leads to many more comparisons than unique observations, making each comparison non-independent.
+(b) Comparative phylogenetic methods, including phylogenetic independent contrasts [@Felsenstein:1985ua], make a smaller number of independent comparisons, where each contrast measures changes along different branches.
+Phylogenetic approaches are rarely used for functional genomics studies.
+
+## Results
+
+### KMRR Reanalysis
+
+When we reanalyze the data with phylogenetic independent contrasts (Fig. 1B), we did not find increased evolutionary change in expression following duplication events (Fig. XX).
+
+```{r preliminaries, echo=FALSE, message=F, warning=F}
 	library( treeio )
@@
 
-```{r load_expression, message=F, warning=F}
+```{r load_expression, echo=FALSE, message=F, warning=F}
 	expression = lapply(
@@
 
-```{r load_trees, cache=TRUE, message=F, warning=F}
+```{r load_trees, echo=FALSE, cache=TRUE, message=F, warning=F}
 	lines = readLines( gene_tree_name )
@@
 			tree <- treeio::read.nhx( tree_tc )
+			close( tree_tc )
 			tree@data$label = c( tree@phylo$tip.label, tree@phylo$node.label )
@@
 
-There were `r length(gene_trees)` parsed.
-
-```{r add_expression_to_trees, cache=TRUE, message=F, warning=F}
+```{r add_expression_to_trees, echo=FALSE, cache=TRUE, message=F, warning=F}
 	gene_trees_annotated = mclapply(
@@
 
-```{r prep_trees, cache=TRUE, message=F, warning=F}
+```{r prep_trees, echo=FALSE, cache=TRUE, message=F, warning=F}
 	drop_empty_tips = function( nhx ) {
@@
 
-`r length(gene_trees_sub)` trees passed taxon sampling criteria.
+Of the `r length(gene_trees)` trees that were parsed, `r length(gene_trees_sub)` passed taxon sampling criteria and `r length(gene_trees_calibrated)` were successfully time calibrated.
 
-`r length(gene_trees_calibrated)` were time calibrated.
-
-```{r contrasts, cache=TRUE, message=F, warning=F}
+```{r contrasts, cache=TRUE, echo=FALSE, message=F, warning=F}
 
@@
 
-```{r show_result, cache=TRUE, message=F, warning=F}
+```{r show_result, echo=FALSE, cache=TRUE, message=F, warning=F}
 
@@
 
-```{r pairwise, cache=TRUE, message=F, warning=F}
+```{r pairwise, echo=FALSE, cache=TRUE, message=F, warning=F}
 	get_pairwise_summary = function ( nhx ) {
@@
 
-## Levin *et al.* reanalysis
+### Levin *et al.* reanalysis
+
+![Figure Levin](levin_etal/Figure_levin.png?raw=true)
+
+> Figure XXLevin | Distributions of pairwise similarity scores for each phase of development.
+Pairwise scores for the ctenophore are red.
+Wilcoxon test p-values for the significance of the differences between early-mid distributions and late-mid distributions are on the right.
+Model of variance, which is inversely related to similarity, is on the left.
+(a) The distributions as published.
+Low similarity (*i.e.*, high variance) in the mid phase of development was interpreted as support for an inverse hourglass model for the evolution of gene expression.
+The five least-similar mid phase scores were all from the ctenophore.
+Published KS p-values, based on duplicated data, are in parentheses.
+The inset ctenophore image is by S.
+Haddock from phylopic.org.
+(b) The distributions after the exclusion of the ctenophore.
+The early and mid phase distributions are not statistically distinct.
+This suggest a wine bottle model, with similar evolutionary variance at the early and mid phase and less at the late phase.
 
@@
 We previously described some concerns with their interpretations of these results [@Hejnol:2016hw].
-Here we directly address problems with the analyses behind their evolutionary conclusions.
+
 Rather than support the inverse hourglass as a general pattern among distantly related species, our reanalyses of Levin *et al.* [@Levin:2016bp] indicate their existing data support the opposite conclusion - that the inverse hourglass is due to large differences specific to a single lineage, the ctenophore (comb jelly).
@@
 We began by examining the matrix of pairwise comparisons that their KS tests and Figure 4d are based on.
-As this figure shows (regenerated here as our Figure 1a), there is not a strong distinction between these temporal phases - the distributions for the early, mid, and late distributions overlap considerably.
+As this figure shows (regenerated here as our Figure XXLevin a), there is not a strong distinction between these temporal phases - the distributions for the early, mid, and late distributions overlap considerably.
 The medians are close in value, and the range of mid phase includes the entire range of the early phase.
-Further inspection (Figure 1a) reveals that the mid phase distribution has several outliers with very low similarity scores.
-We found that all five of the lowest values in the mid phase distribution are for pairwise comparisons that include the ctenophore (Figure 1a).
-When the nine pairwise comparisons that include the ctenophore are removed, the differences between the early and mid phase distributions are greatly reduced (Figure 1b).
+Further inspection (Figure XXLevin a) reveals that the mid phase distribution has several outliers with very low similarity scores.
+We found that all five of the lowest values in the mid phase distribution are for pairwise comparisons that include the ctenophore (Figure XXLevin a).
+When the nine pairwise comparisons that include the ctenophore are removed, the differences between the early and mid phase distributions are greatly reduced (Figure XXLevin b).
 
@@
 The Wilcoxon test is instead appropriate in this case.
-After removing the duplicate scores, removing the ctenophore, and applying the Wilcoxon test, there is no significant difference between the early phase and mid phase distributions (P  = 0.1428 for the early-mid comparison and P < 10<sup>-5</sup> for the late-mid comparison), and the inverse hourglass turns into a wine bottle (Figure 1b).
+After removing the duplicate scores, removing the ctenophore, and applying the Wilcoxon test, there is no significant difference between the early phase and mid phase distributions (P  = 0.1428 for the early-mid comparison and P < 10<sup>-5</sup> for the late-mid comparison), and the inverse hourglass turns into a wine bottle (Figure XXLevin b).
 
@@
 
-![Figure 1](levin_etal/Figure_levin.png?raw=true)
+While demonstrate a common problem with pairwise comparisons in Levin *et al.* dataset, we did not perform a phylogenetic reanalysis of this study, as we did for the KMRR study.
+This is because the similarity metric computed in the pairwise comparisons of Levin *et al.* are based on different genes for different species pairs.
+This means that there is not a trait that can be mapped onto the tree, as KMRR's tau can be.
+A full phylogenetic reanalysis would of course be possible be using upstream analysis products to rederive a new expression summary statistic that was suitable for mapping onto a phylogenetic tree.
 
-> Figure 1 | Distributions of pairwise similarity scores for each phase of development.
-Pairwise scores for the ctenophore are red.
-Wilcoxon test p-values for the significance of the differences between early-mid distributions and late-mid distributions are on the right.
-Model of variance, which is inversely related to similarity, is on the left.
-(a) The distributions as published.
-Low similarity (*i.e.*, high variance) in the mid phase of development was interpreted as support for an inverse hourglass model for the evolution of gene expression.
-The five least-similar mid phase scores were all from the ctenophore.
-Published KS p-values, based on duplicated data, are in parentheses.
-The inset ctenophore image is by S.
-Haddock from phylopic.org.
-(b) The distributions after the exclusion of the ctenophore.
-The early and mid phase distributions are not statistically distinct.
-This suggest a wine bottle model, with similar evolutionary variance at the early and mid phase and less at the late phase.
+## Discussion
+
+Some of the most widely used phylogenetic comparative methods [1, 5] are directly applicable to CFG data.
+There are also unique challenges that will need to be addressed at this new intersection, though.
+One of the greatest challenges is that most phylogenetic comparative methods have been developed to address problems with many more species (e.g., dozens or more) relative to the number of traits being examined (e.g., 2-5).
+In CFG analyses, there are often far fewer species (e.g., 3-10) because adding taxa is still expensive, but tens of thousands of traits.
+This creates challenges as the resulting covariance matrices are singular and, if not treated appropriately, imply many false correlations that are artefacts of project design.
+We outlined these challenges and potential solutions in the context of gene expression [31].
+
+It is not enough to demonstrate the impact of phylogenetics on CFG analyses, these methods need to be made accessible to the FG community so they can be widely.
+
+Most early phylogenetic comparative methods attempted to remove evolutionary signal to correct statistical tests for correlations between traits, while more recent methods tend to focus on testing hypotheses of evolutionary processes [34].
+The application of this newer focus to pCFG provides an exciting opportunity to address long standing questions of broad interest, including the order of changes in FG traits and shifts in rates of evolution of one FG trait following changes in another trait.
+
+A small number of CFG studies have employed some phylogenetic comparative approaches [18–22].
+For instance, a phylogenetic ANOVA [21] of the evolution of gene expression improves statistical power and drastically reduces the rate of false positives relative to pairwise approaches.
+
+## Conclusions
+
+The fact that the first two CFG studies we reanalyzed show serious problems with pairwise comparisons indicates that there likely to be similar problems in other CFG studies, and that future CFG studies will be compromised if they continue to use pairwise methods.
 
@@
 
+## Supplementary Material
+
 ## References

### 2017-01-22 manuscript.rmd
added ascertainment bias explorations and ortholog conjecture text

@@
 
-A rich theoretical and statistical methodology of phylogenetic comparative methods have been developed over the last three decades to address the statistical challenges and opportunities of comparisons across species [@Felsenstein:1985ua; Grafen:1989um; Chamberlain:2012vx; Garamszegi:2014ew; Pagel:1999gc].
+A rich theoretical and statistical methodology of phylogenetic comparative methods have been developed over the last three decades to address the challenges and opportunities of comparisons across species [@Felsenstein:1985ua; Grafen:1989um; Chamberlain:2012vx; Garamszegi:2014ew; Pagel:1999gc].
 These have largely been applied to the evolution of morphological and ecological traits, but are just as relevant to CFG.
@@
 
+The KMRR considers several publicly available datasets.
+Here we consider their conclusions in the context of Brawand *et al.* 2011 [Brawand:2011du], one of the studies they considered.
+This analysis of XX different organs across XX mammal species.
+
+For each internal node in each gene tree (Figure 1b), we then measured the phylogenetically independent contrast (PIC) in the expression across this node [@Felsenstein:1985ua].
+This is the difference in values of descendent nodes (i.e., the evulutionary change along the branches that descend from the node) scaled by the expected variance (determined largely by the lengths of these branches).
+These contrasts were then annotated by whether they are made across branches that descend from a speciation or duplication event.
+The original description of independent contrasts [@Felsenstein:1985ua] focused on assessing covariance between changes in two traits on a tree.
+Our use of it to look for differences in evolutionary changes of one trait (differential expression) between two categories of nodes (speciation and duplication) on a tree is a bit different than the usual application of the method.
+
+![Figure Expectations](figure_expectations.svg?raw=true)
+
+> Figure XXExpectations | (a) Null hypothesis. (b) Under the ortholog conjecture.
+
+To test the ortholog conjecture it is essental to have a null hypothesis that results in a distinct prediction from the ortholog conjecture.
+A suitable null hypothesis is that there is no difference in evolutionary change in differential expression along branches that descend from speciation or duplication events (Figure 1b).
+Under this hypothesis, we would predict that contrasts across speciation nodes and duplication nodes are drawn from the same distribution (Figure XXEXPECTATIONS a).
+Under the alternative hypothesis specified by the ortholog conjecture, that there is a higher rate of change following duplication events than speciation events, we would expect to see the distribution of duplication contrasts shifted to higher values relative to the speciation contrasts (Figure XXEXPECTATIONS b).
+
 When we reanalyze the data with phylogenetic independent contrasts (Fig. 1B), we did not find increased evolutionary change in expression following duplication events (Fig. XX).
@@
 	library( hutan )
+	library( gridExtra )
 
@@
 
-	# These are from the KRR code
+	# These are from the KMRR code
 	calibration_times = data.frame(
@@
 	gene_trees_sub = gene_trees_sub[ ! is.na( gene_trees_sub ) ]
-#	gene_trees_sub = gene_trees_sub[ lapply(gene_trees_sub, class) == "treedata" ]
 
@@
 
+	all_nodes$Event = NA
+
+	all_nodes$Event[ all_nodes$D == "Y" ] = "Duplication"
+	all_nodes$Event[ all_nodes$D == "N" ] = "Speciation"
+
 ```
@@
 	nodes_contrast %>%
-		ggplot( aes( x=pic, y=..density.., col=D ) ) + 
+		ggplot( aes( x=pic, y=..density.., col=Event ) ) + 
         geom_freqpoly( binwidth=.001, position="identity" ) +
         xlab("Phylogenetic independent \ncontrasts (PIC)") +
-				xlim(0,0.075)
+				xlim(0,0.075) + 
+				theme_classic()
 
-	nodes_contrast %>%
-		filter( pic < 0.2 ) %>%
-		ggplot( aes( x=node_age, y=pic, col=D ) ) +
-		geom_point( alpha=0.2) +
-		xlim(0,1000)
-
-	wilcox.test( nodes_contrast %>% filter(D=="Y") %>% .$pic, nodes_contrast %>% filter(D=="N") %>% .$pic, alternative="greater" )
+	wilcox_test_result = wilcox.test( nodes_contrast %>% filter(D=="Y") %>% .$pic, nodes_contrast %>% filter(D=="N") %>% .$pic, alternative="greater" )
 
@@
 
+> Figure XXKMRR | Density plot of the magnitude of phylogenetic independent contrast values following duplication and speciation events.
+These contrasts are not larger following duplication events, as predicted by the ortholog conjecture.
+
 There were `r nodes_contrast %>% filter(D=="Y") %>% nrow()` duplication nodes and `r nodes_contrast %>% filter(D=="N") %>% nrow()` speciation nodes.
 
-```{r pairwise, echo=FALSE, cache=TRUE, message=F, warning=F}
-	get_pairwise_summary = function ( nhx ) {
+Given the ortholog conjecture, we would expect to see larger expression contrasts following duplication events than speciation events.
+Instead,we find no such distribution (Figure XXKMRR).
+The Wilcox rank test does not reject the null hypothesis that the rate of evolution following duplications is the same as or less than the rate following speciation (p value = `r wilcox_test_result$p.value`).
+Our results therefore find no evidence for the ortholog conjecture in this system.
 
-		# Create a tibble with one row for each pairwise combination of tip nodes
-		ntips = length( nhx@phylo$tip.label )
-		D = 1:ntips %>% combn( 2 ) %>% t() %>% as_tibble()
-		names( D ) = c( "tip_a", "tip_b" )
+We next examined the possibility that ascertainment biases were differentially impacting the inference of evolutionary change in differential expression following duplication and speciation events.
+We focused on two possible sources of bias - node depth and branch length.
 
-		# Add a colmn with the gene tree hash 
-		D %<>% mutate( gene_tree = digest( nhx ) )
+While the age of speciation nodes is constrained, duplication nodes can be much older and therefore have a wider range of ages (Supplementary Figure XXAscertainment a).
+Biological reasons for this include the fact that duplication events can be older than the root of the species tree, but speciation events cannot.
+There are also technical factors that can lead to an excess of duplication events deeper in the tree, including the impact of assembly errors that lead to the overestimation of deep duplications [@Hahn:2007cj].
+If independent contrast values also tended to to be lower at greater node depth, it could interact with the preponderance of duplications at greater depth to create a pattern of lower contrasts following duplication events.
+When we remove duplication nodes that are older than the oldest speciation node, our results are unchanged and we cannot reject the null hypothesis that the rate of evolution following duplications is the same as or less than the rate following speciation (Supplementary Figure XXAscertainment b).
 
-		# Calculate the phenotypic difference between the tips
-		D$difference = 
-			apply(
-				D,
-				1,
-				function(y) {
-					tau_values = nhx@data$Tau[ c( as.integer(y['tip_a']), as.integer( y['tip_b'] ) ) ]
-					max( tau_values ) - min( tau_values )
-				}
-			)
+The independent contrast across a node is the amount of change observed between the daughter notes, scaled by the expected variance [Felsenstein:1985ua].
+The expected variance is principaly determined by the lengths of the branches leading from the node to these daughters.
+The same observed difference between daughter nodes will lead to a larger contrast the shorter the branches are to these nodes, as this indicates a greater rate of evolutionary change.
+The expected variance of contrasts for speciation nodes is constrained by the branch lengths on the species tree, but the expected variance of contrasts for duplications can range much wider (Figure XXAscertainment c).
 
-		# For each row, bet the most recent common ancestor of the pairwise tip combination
-		D$mrca = 
-			apply(
-				D, 
-				1, 
-				function(y) ape::getMRCA( 
-					nhx@phylo, 
-					c( as.integer(y['tip_a']), as.integer(y['tip_b'] ))
-				)
-			) %>%
-			as.integer()
+The fact that the expected variance of speciation contrasts is a subset of the range of the expected variance of duplication contrasts could lead to ascertainment biases.
+We therefore also examine only the contrasts that fell within in the range of expected variance seen for speciation contrasts, excluding duplication contrasts that fall outside of this range.
+This reanalysys does not reject the ortholog conjecture either (Figure XXAscertainment d).
 
-		# Merge data on this mrca to the node
-		D %<>% left_join(
-			nhx@data %>%
-				select( D, node, node_age, pic ),
-			by = c( "mrca" = "node")
-
-		)
-
-		# Calculate the distance between the tips, which for an ultrametric tree 
-		# is twice the age of their most recent common ancestor
-		D %<>% mutate( distance = 2 * node_age )
-
-		return( D )
-
-	}
-
-	# Build a tibble of all pairwise comparisons between tips
-	pairwise_summary = 
-		mclapply( gene_trees_pic, get_pairwise_summary, mc.cores=cores ) %>% 
-		bind_rows()
-
-	pairwise_summary %>% 
-		filter( ! is.na(D) ) %>%
-  	ggplot(aes(x=distance, y=difference, color=D)) + 
-  		geom_point(alpha=0.1) +
-  		geom_smooth(method=lm, se=FALSE) +
-			xlim( 0, 2000 )
-
-```
+The widespread invokation of the ortholog conjecture follows from the expectation that it is strong, widely distributed phenomenon.
+Our failure to detect a signature consistent with the ortholog conjecture suggests that it may not be, and that it should not be assumed in systems where it has not been explicitly tested.
 
@@
 
+### Comparative methods in functional genomics
+
 Some of the most widely used phylogenetic comparative methods [1, 5] are directly applicable to CFG data.
@@
 
+### Ortholog conjecture
+
+The ortholog conjecture [@Nehrt:2011bm] is the proposition that orthologs (genes that diverged from each other due to a speciation event) have more similar attributes than do paralogs (genes that diverged from each other due to a gene duplication event).
+It has important biological and technical implications.
+It shapes our understanding of the functional diversity of gene families.
+It is used to relate findings from well-studied genes to related genes that haven't been investigated in detail.
+It can be applied to any trait of genes, from gene seqeunce to biochemical properties to expression.
+While the ortholog conjecture describes a specific pattern of functional diversity across genes, it can also be considered a hypothesis about the process of evolution.
+Specifically, it postulates that there is greater evolutionary change in gene attributes following a duplication event than a speciation event.
+
+Despite its importance, there have been relatively few tests of the ortolog conjecture.
+Previous work has shown that ontology annotations are not sufficient to test the ortholog conjecture [@Chen:2012jz; @Thomas:2012hz].
+Analyses of domain structure were consistent with ortholog conjecture [@Forslund:2011cw].
+There have been few tests of the ortholog conjecture with regards to gene expression [@Chen:2012jz].
+
+Our reanalysis of KMRR using phylogenetic methods suggests that the ortholog conjectures is not a dominant pattern that is central to explaining the evolution of phenotypic diversity in gene families.
+This has impotant biological implications.
+It suggests that the mechanism of gene divergence (speciation versus duplication) may not have as strong an impact on phenotypic divergence as thought.
+It also has technical implications.
+Rather than focus on whether genes are orthologs or paralogs when attempting to predict function, it may be more effective to simply focus on how closely related or distantly related they are.
+Closely related paralogs, for example, may tend to have more similar phenotypes than more distantly related orthologs.
+
 ## Conclusions
@@
 
+All files needed to reexecute the anlyses presented in this document are available at https://github.com/caseywdunn/comparative_expression_2017.
+The most recent commit at the time of the analysis presented here was `r system("git log | head -n 1", intern=TRUE) %>% str_replace("commit ", "")`.
+
 ### KMRR reanalysis
 
-Based on their scripts from: https://figshare.com/articles/Tissue-specificity_of_gene_expression_diverges_slowly_between_orthologs_and_rapidly_between_paralogs/3493010/2
+The KMRR study [@KryuchkovaMostacci:2016iw] followed excellent practices in reproducibility.
+They posted all data and code needed to re-execute their analyses at figshare:  https://figshare.com/articles/Tissue-specificity_of_gene_expression_diverges_slowly_between_orthologs_and_rapidly_between_paralogs/3493010/2 .
+We slightly altered their `Rscript.R` to simplify file paths.
+This midified script and their data files are available in the gihub respository for this paper, as are the intermediate files that were generated by their analysis script that we used in our own analyses.
+We also obtained the `Compara.75.protein.nh.emf` gene trees [@Herrero:2016jj] from ftp://ftp.ensembl.org/pub/release-75/emf/ensembl-compara/homologies/ and place them in the same directory as this file.
 
-To replciate the published analysis, `cd` to `Data_used/Expression_data`, launch `R`, and run:
-
-    source( "../Rscript.R" )
-
-After running the analysis above, also obtain the `Compara.75.protein.nh.emf` gene trees from ftp://ftp.ensembl.org/pub/release-75/emf/ensembl-compara/homologies/ and place them in the same directory as this file.
+Each of these gene trees includes includes homologous sequences derived from a common ancestral sequence.
+The tip nodes of a gene tree can include both orthogs and paralogs, and each internal node is an infered speciation or duplication event.
+Prior to phylogenetic analyses of gene expression, each tree was first pruned to include only tips for which expression data were available.
+The speciation nodes in the tree were then time calibrated so that branch lengths were consistent across trees.
+We then calculated the magnitude of the independent contrast for differential expression at each internal node in each of these trees.
+The independent contrast is the amount of change in expression observed along the branches that arise from the node, scaled by the expected variance [Felsenstein:1985ua].
+The expected variance is determined mostly by the length of the branches, but can also be influenced by uncertainty regarding the reconstruction of differential expression at internal nodes.
+We took the absolute value of these contrasts, since the sign depends only on the arbitrary ordering of the nodes in the trees.
 
@@
 
-Thanks to
+Thanks to XX. This work was supported by the National Science Foundation (DEB-1256695 and the Waterman Award).
 
@@
 
+### KMRR Analyses
+
+```{r filtered, cache=TRUE, echo=FALSE, fig.cap="XX_PIC_VAR", warning=FALSE, message=FALSE}
+
+	# Examine by node age
+
+  g_age =
+  nodes_contrast %>%
+    ggplot( aes( x=node_age, y=pic, col=Event ) ) + 
+		  geom_point( alpha=0.2 ) + 
+		  xlab( "Node age (millions of years)" ) + 
+		  ylab("Phylogenetic independent contrasts (PIC)\n of differential expression") +
+		  ylim(0,0.3) +
+			xlim(0,1000) +
+		  ggtitle( "(a)" ) +
+		  theme_classic()	
+		  #theme( legend.justification=c( 1, 1 ), legend.position=c( 1, 1 ) )
+
+	max_speciation_age = 
+    nodes_contrast %>% 
+    filter( D=="N" ) %>%
+    select( node_age ) %>%
+    max( na.rm=TRUE )
+
+	nodes_contrast_filtered_age = 
+		nodes_contrast %>% 
+		filter( node_age <= max_speciation_age )
+
+	wilcox_filtered_age = 
+		wilcox.test( 
+			nodes_contrast_filtered_age %>% filter(D=="Y") %>% .$pic, 
+			nodes_contrast_filtered_age %>% filter(D=="N") %>% .$pic, 
+			alternative="greater" 
+		)$p.value
+
+	label_age = str_c( "Wilcoxon p=", wilcox_filtered_age )
+
+	g_density_filtered_age =   
+		nodes_contrast_filtered_age %>%
+		ggplot( aes( x=pic, y=..density.., col=Event ) ) + 
+      geom_freqpoly( binwidth=.001, position="identity" ) +
+      xlab("Phylogenetic independent \ncontrasts (PIC)") +
+			ylab("Density") +
+			xlim(0,0.075) + 
+			ggtitle( "(b)" ) +
+			annotate("text", x = Inf, y = Inf, hjust="right", vjust="top", label=label_age ) +
+			theme_classic()
+
+	# Examine by expected variance (branch length)
+  g_var_exp =
+    nodes_contrast %>%
+      ggplot( aes( x=var_exp, y=pic, col=Event ) ) + 
+      geom_point( alpha=0.2 ) + 
+      xlab( "Expected variance, \n depends mostly on branch length" ) + 
+      ylab("Phylogenetic independent contrasts (PIC)\n of differential expression") +
+      ylim(0,0.3) +
+			xlim(0,2000) +
+      ggtitle( "(c)" ) +
+      theme_classic()	
+      # theme( legend.justification=c( 1, 1 ), legend.position=c( 1, 1 ) )
+
+  var_exp_max = 
+    nodes_contrast %>% 
+    filter( D=="N" ) %>%
+    select( var_exp ) %>%
+    max( na.rm=TRUE )
+
+  var_exp_min = 
+    nodes_contrast %>% 
+    filter( D=="N" ) %>%
+    select( var_exp ) %>%
+    min( na.rm=TRUE )
+
+  nodes_contrast_filtered_var = 
+    nodes_contrast %>% 
+      filter ( var_exp >=var_exp_min & var_exp <=var_exp_max )
+
+	wilcox_filtered_var = 
+	wilcox.test( 
+		nodes_contrast_filtered_var %>% filter(D=="Y") %>% .$pic, 
+		nodes_contrast_filtered_var %>% filter(D=="N") %>% .$pic, 
+		alternative="greater" 
+	)$p.value
+
+	label_var = str_c( "Wilcoxon p=", wilcox_filtered_var )
+
+  g_density_filtered_var =   
+    nodes_contrast_filtered_var  %>% 
+    ggplot( aes( x=pic, y=..density.., col=Event ) ) + 
+      geom_freqpoly( binwidth=.001, position="identity" ) + 
+      ggtitle( "(d)" ) +
+      theme( legend.justification=c( 1, 1 ), legend.position=c( 1, 1 ) ) +
+      xlim(0,0.075) + 
+			theme_classic() +
+  		annotate("text", x = Inf, y = Inf, hjust="right", vjust="top", label=label_var ) +
+      xlab("Phylogenetic independent contrasts (PIC)\n of differential expression") +
+  		ylab("Density")
+
+  grid.arrange(
+  	g_age,
+  	g_density_filtered_age,
+		g_var_exp,
+		g_density_filtered_var,
+		ncol = 2
+	)
+```
+
+> Supplementary Figure XXAscertainment | Investigation of possible assertainment biases.
+(a) Magnitude of independent contrasts ploted against node age.
+Speciation nodes are calibrated to particular times, whereas duplication nodes have a wider range.
+(b) Density plot of contrasts for only the nodes that have an age less than or equal to the maximum age of speciation nodes.
+(c) Magnitude of independent contrasts ploted against expected variance, which is largely determined by branch lengths.
+Contrasts for speciation nodes have a narrower range of expected variance than do contrasts for duplication nodes.
+(d) Density plot of contrasts for only the nodes that have expected variance within the range of contrasts across speciation nodes.
+
+```{r pairwise, echo=FALSE, cache=TRUE, message=F, warning=F}
+	get_pairwise_summary = function ( nhx ) {
+
+		# Create a tibble with one row for each pairwise combination of tip nodes
+		ntips = length( nhx@phylo$tip.label )
+		D = 1:ntips %>% combn( 2 ) %>% t() %>% as_tibble()
+		names( D ) = c( "tip_a", "tip_b" )
+
+		# Add a colmn with the gene tree hash 
+		D %<>% mutate( gene_tree = digest( nhx ) )
+
+		# Calculate the phenotypic difference between the tips
+		D$difference = 
+			apply(
+				D,
+				1,
+				function(y) {
+					tau_values = nhx@data$Tau[ c( as.integer(y['tip_a']), as.integer( y['tip_b'] ) ) ]
+					max( tau_values ) - min( tau_values )
+				}
+			)
+
+		# For each row, bet the most recent common ancestor of the pairwise tip combination
+		D$mrca = 
+			apply(
+				D, 
+				1, 
+				function(y) ape::getMRCA( 
+					nhx@phylo, 
+					c( as.integer(y['tip_a']), as.integer(y['tip_b'] ))
+				)
+			) %>%
+			as.integer()
+
+		# Merge data on this mrca to the node
+		D %<>% left_join(
+			nhx@data %>%
+				select( D, node, node_age, pic ),
+			by = c( "mrca" = "node")
+
+		)
+
+		# Calculate the distance between the tips, which for an ultrametric tree 
+		# is twice the age of their most recent common ancestor
+		D %<>% mutate( distance = 2 * node_age )
+
+		return( D )
+
+	}
+
+	# Build a tibble of all pairwise comparisons between tips
+	pairwise_summary = 
+		mclapply( gene_trees_pic, get_pairwise_summary, mc.cores=cores ) %>% 
+		bind_rows()
+
+	pairwise_summary %>% 
+		filter( ! is.na(D) ) %>%
+  	ggplot(aes(x=distance, y=difference, color=D)) + 
+  		geom_point(alpha=0.1) +
+  		geom_smooth(method=lm, se=FALSE) +
+			xlim( 0, 2000 )
+
+```
+
+> Supplementary Figure XXPairwise | Pairwise comparisons
+
 ## References

### 2017-01-22 manuscript.rmd
minor formatting

@@
   github_document: default
+  pdf_document:
+    dev: png
 ---
@@
 # The importance of phylogenetic methods when comparing functional genomic data across species
-
-Dunn, Zapata, Munro, Siebert, Hejnol
 
@@
 
-A rich theoretical and statistical methodology of phylogenetic comparative methods have been developed over the last three decades to address the challenges and opportunities of comparisons across species [@Felsenstein:1985ua; Grafen:1989um; Chamberlain:2012vx; Garamszegi:2014ew; Pagel:1999gc].
+A rich theoretical and statistical methodology of phylogenetic comparative methods have been developed over the last three decades to address the challenges and opportunities of comparisons across species [@Felsenstein:1985ua; @Grafen:1989um; @Chamberlain:2012vx; @Garamszegi:2014ew; @Pagel:1999gc].
 These have largely been applied to the evolution of morphological and ecological traits, but are just as relevant to CFG.
@@
 One reason that CFG has not yet embraced phylogenetic approaches is that it has not yet been concretely demonstrated that pairwise and phylogenetic approaches can lead to different results in CFG.
-Although the value of phylogenetic approaches has been repeatedly shown in comparative analyses of other character data [@Chamberlain:2012vx; Ricklefs:1996wb], this value is apparently not well known to the FG community since the field has until recently largely focused on within-species analyses.
+Although the value of phylogenetic approaches has been repeatedly shown in comparative analyses of other character data [@Chamberlain:2012vx; @Ricklefs:1996wb], this value is apparently not well known to the FG community since the field has until recently largely focused on within-species analyses.
 Many initial CFG studies have also been focused on only two species at a time, where the differences between phylogenetic and pairwise approaches are not as pronounced.
@@
 
-![Figure Overview](Figure_overview.svg?raw=true)
+![Figure Overview](Figure_overview.png?raw=true)
 
@@
 
-![Figure Expectations](figure_expectations.svg?raw=true)
+![Figure Expectations](figure_expectations.png?raw=true)
 
@@
 Specifically, they concluded that animals from different phyla exhibit an "inverse hourglass" model for the evolution of gene expression, where there is more evolutionary variance in gene expression at a mid phase of development than there is at early and late phases.
-Closely related animals have previously been described as having an hourglass model of gene expression, where evolutionary variance in expression is greater early and late in development than at the midpoint of development [@Kalinka:2010dw; DomazetLoso:2010cn].
+Closely related animals have previously been described as having an hourglass model of gene expression, where evolutionary variance in expression is greater early and late in development than at the midpoint of development [@Kalinka:2010dw; @DomazetLoso:2010cn].
 Levin *et al.* conclude that this contrast between distantly and closely related animals provides biological justification for the concept of phyla and may provide a long-sought operational definition of phyla.
@@
 Rather than support the inverse hourglass as a general pattern among distantly related species, our reanalyses of Levin *et al.* [@Levin:2016bp] indicate their existing data support the opposite conclusion - that the inverse hourglass is due to large differences specific to a single lineage, the ctenophore (comb jelly).
-Better understanding these differences will be of interest to future work, especially in light of current uncertainty regarding the phylogenetic placement of ctenophores and their unique biology [@Martindale:2015ve, Dunn:2015ij].
+Better understanding these differences will be of interest to future work, especially in light of current uncertainty regarding the phylogenetic placement of ctenophores and their unique biology [@Martindale:2015ve, @Dunn:2015ij].
 

### 2017-01-24 manuscript.rmd
reorganized a bit, added to pairwise analyses

@@
 We find problems with the pairwise comparisons used in both studies that undermine their published conclusions.
-This is a concrete demonstration of the inadequacy of pairwise comparisons in comparative genomics studies, and indicates that it will be critical to adopt phylogenetic comparative methods in future analyses.
+The first study we examine found that expression tends to be more different between paralogs than orthologs, and interpretted this as evidence of the ortholog conjecture- the hypothesis that there are greater rates of evolutionary change following duplication.
+Our phylogenetic analysis finds no difference in rates following duplication and speciation.
+Instead, the greater difference between paralogs is because they tend to be more distantly related to each other than orthologs.
+The second study claimed to find more variance in gene expression at the midpoint of development than earlier and later when comparing animals from different phyla.
+We instead find that changes along the branch of a single taxon, the ctenophore, were overcounted becuase they were included in each pairwise comparison.
+These reanalyses are concrete demonstrations of the inadequacy of pairwise comparisons in comparative genomics studies, and indicates that it will be critical to adopt phylogenetic comparative methods in future analyses.
 
@@
 
-## Results
+## Results and Discussion
 
@@
 
-The KMRR considers several publicly available datasets.
-Here we consider their conclusions in the context of Brawand *et al.* 2011 [Brawand:2011du], one of the studies they considered.
-This analysis of XX different organs across XX mammal species.
+#### Original analyses
 
-For each internal node in each gene tree (Figure 1b), we then measured the phylogenetically independent contrast (PIC) in the expression across this node [@Felsenstein:1985ua].
-This is the difference in values of descendent nodes (i.e., the evulutionary change along the branches that descend from the node) scaled by the expected variance (determined largely by the lengths of these branches).
+The KMRR study [@KryuchkovaMostacci:2016iw] sought to test the ortholog conjecture.
+The ortholog conjecture [@Nehrt:2011bm] is the proposition that orthologs (genes that diverged from each other due to a speciation event) have more similar attributes than do paralogs (genes that diverged from each other due to a gene duplication event).
+It has important biological and technical implications.
+It shapes our understanding of the functional diversity of gene families.
+It is used to relate findings from well-studied genes to related genes that haven't been investigated in detail.
+It can be applied to any trait of genes, from gene seqeunce to biochemical properties to expression.
+While the ortholog conjecture describes a specific pattern of functional diversity across genes, it can also be considered a hypothesis about the process of evolution.
+Specifically, it postulates that there is greater evolutionary change in gene attributes following a duplication event than a speciation event.
+
+Despite its importance, there have been relatively few tests of the ortolog conjecture.
+Previous work has shown that ontology annotations are not sufficient to test the ortholog conjecture [@Chen:2012jz; @Thomas:2012hz].
+Analyses of domain structure were consistent with ortholog conjecture [@Forslund:2011cw].
+There have been few tests of the ortholog conjecture with regards to gene expression [@Chen:2012jz], and KMRR is the most thorough such study to date.
+
+The KMRR study considered several publicly available datasets.
+Here we consider their conclusions in the context of Brawand *et al.* 2011 [@Brawand:2011du], one of the studies they considered.
+This dataset of gene expression for XX different organs across XX mammal species is the best sampled in their reanalyses.
+Their expression summary statistic is Tau [@Yanai:2005dh, @KryuchkovaMostacci:2016in], an indicator of tissue specificity of gene expression.
+It can range from a value of 0, which indiciates no specifificity (*i.e.*, uniform expression across tissues), to a value of 1, which indicates his specificity.
+It is convenient in that it is a single number of defined range for each gene in each species, though of course since the original expression is multidemensional this means much information is discarded.
+For example, it does not convey which tissue expression is specific to.
+Homologous genes could each have a tau value of 1, but one gene could have expression that is specific to the kidney and the other to the brain, for example.
+Their published analyses are based on pairwise comparisons (Figure 1a) betwen tau for each gene family.
+They found the correlation for orthologs to be significantly greater than the correlation of tau for paralogs, *i.e.* that orthologs tend to be more similar to each other than paralogs.
+From this concluded that their analyses support the ortholog conjecture.
+They also went one step further - they concluded that this pattern provides support for a particular evolutionary process, that "tissue-specificity evolves very slowly in the absence of duplication, while immediately after duplication the new gene copy differs."
+
+#### Phylogenetic reanalyses
+
+We reanalyzed the their study using phylogenetic comparative methods, focusing on the dataset of Brawand *et al.* 2011 [Brawand:2011du].
+For each internal node in each gene tree (Figure 1b), we measured the phylogenetically independent contrast (PIC) in the expression across this node [@Felsenstein:1985ua].
+This is the difference in values of descendent nodes (*i.e.*, the evulutionary change along the branches that descend from the node) scaled by the expected variance (determined largely by the lengths of these branches).
 These contrasts were then annotated by whether they are made across branches that descend from a speciation or duplication event.
@@
 
+```{r ascertainment, cache=TRUE, echo=FALSE, fig.cap="XX_PIC_VAR", warning=FALSE, message=FALSE}
+
+	# Examine by node age
+
+  g_age =
+  nodes_contrast %>%
+    ggplot( aes( x=node_age, y=pic, col=Event ) ) + 
+		  geom_point( alpha=0.2 ) + 
+		  xlab( "Node age (millions of years)" ) + 
+		  ylab("Phylogenetic independent contrasts (PIC)\n of differential expression") +
+		  ylim(0,0.3) +
+			xlim(0,1000) +
+		  ggtitle( "(a)" ) +
+		  theme_classic()
+
+	max_speciation_age = 
+    nodes_contrast %>% 
+    filter( D=="N" ) %>%
+    select( node_age ) %>%
+    max( na.rm=TRUE )
+
+	nodes_contrast_filtered_age = 
+		nodes_contrast %>% 
+		filter( node_age <= max_speciation_age )
+
+	wilcox_filtered_age = 
+		wilcox.test( 
+			nodes_contrast_filtered_age %>% filter(D=="Y") %>% .$pic, 
+			nodes_contrast_filtered_age %>% filter(D=="N") %>% .$pic, 
+			alternative="greater" 
+		)$p.value
+
+	# Examine by expected variance, ie branch length
+  var_exp_max = 
+    nodes_contrast %>% 
+    filter( D=="N" ) %>%
+    select( var_exp ) %>%
+    max( na.rm=TRUE )
+
+  var_exp_min = 
+    nodes_contrast %>% 
+    filter( D=="N" ) %>%
+    select( var_exp ) %>%
+    min( na.rm=TRUE )
+
+  nodes_contrast_filtered_var = 
+    nodes_contrast %>% 
+      filter ( var_exp >=var_exp_min & var_exp <=var_exp_max )
+
+	wilcox_filtered_var = 
+	wilcox.test( 
+		nodes_contrast_filtered_var %>% filter(D=="Y") %>% .$pic, 
+		nodes_contrast_filtered_var %>% filter(D=="N") %>% .$pic, 
+		alternative="greater" 
+	)$p.value
+
+```
+
 We next examined the possibility that ascertainment biases were differentially impacting the inference of evolutionary change in differential expression following duplication and speciation events.
@@
 
-The widespread invokation of the ortholog conjecture follows from the expectation that it is strong, widely distributed phenomenon.
-Our failure to detect a signature consistent with the ortholog conjecture suggests that it may not be, and that it should not be assumed in systems where it has not been explicitly tested.
+#### Understanding the incongruence between pairwise and phylogenetic methods
+
+```{r pairwise, echo=FALSE, cache=TRUE, message=F, warning=F}
+	get_pairwise_summary = function ( nhx ) {
+
+		# Create a tibble with one row for each pairwise combination of tip nodes
+		ntips = length( nhx@phylo$tip.label )
+		D = 1:ntips %>% combn( 2 ) %>% t() %>% as_tibble()
+		names( D ) = c( "tip_a", "tip_b" )
+
+		# Add a colmn with the gene tree hash 
+		D %<>% mutate( gene_tree = digest( nhx ) )
+
+		# Add the tip names
+		D %<>% mutate ( name_a = nhx@data$G[ D$tip_a ] )
+		D %<>% mutate ( name_b = nhx@data$G[ D$tip_b ] )
+
+		# Add the values at the tips
+		D %<>% mutate ( tau_a = nhx@data$Tau[ D$tip_a ] )
+		D %<>% mutate ( tau_b = nhx@data$Tau[ D$tip_b ] )
+		D %<>% mutate ( difference = abs( tau_a - tau_b ) )
+
+		# Calculate the phenotypic difference between the tips with a different method as 
+		# check on parsing
+		D$difference_old = 
+			apply(
+				D,
+				1,
+				function(y) {
+					tau_values = nhx@data$Tau[ c( as.integer(y['tip_a']), as.integer( y['tip_b'] ) ) ]
+					max( tau_values ) - min( tau_values )
+				}
+			)
+
+		if ( ! all ( near( D$difference, D$difference_old ) ) ){
+			stop( "Inconsistent difference calculations" )
+		}
+
+		# For each row, bet the most recent common ancestor of the pairwise tip combination
+		D$mrca = 
+			apply(
+				D, 
+				1, 
+				function(y) ape::getMRCA( 
+					nhx@phylo, 
+					c( as.integer(y['tip_a']), as.integer(y['tip_b'] ))
+				)
+			) %>%
+			as.integer()
+
+		# Merge data on this mrca to the node
+		D %<>% left_join(
+			nhx@data %>%
+				select( D, node, node_age, pic ),
+			by = c( "mrca" = "node")
+
+		)
+
+		# Calculate the distance between the tips, which for an ultrametric tree 
+		# is twice the age of their most recent common ancestor
+		D %<>% mutate( distance = 2 * node_age )
+
+		return( D )
+
+	}
+
+	# Build a tibble of all pairwise comparisons between tips
+	pairwise_summary = 
+		mclapply( gene_trees_pic, get_pairwise_summary, mc.cores=cores ) %>% 
+		bind_rows()
+
+	# Calcualte some summary statistics 
+	ortholog_r = 
+		pairwise_summary %>% 
+		filter( D=="N" ) %>% 
+		select( tau_a, tau_b ) %>% 
+		cor( ) %>% 
+		.[ 1, 2 ]
+
+	paralog_r = 
+		pairwise_summary %>% 
+		filter( D=="Y" ) %>% 
+		select( tau_a, tau_b ) %>% 
+		cor( ) %>% 
+		.[ 1, 2 ]
+
+	paralog_subset_r = 
+		pairwise_summary %>% 
+		filter( D=="Y" ) %>% 
+		filter( node_age <= max_speciation_age  ) %>% 
+		select( tau_a, tau_b ) %>% 
+		cor( ) %>% 
+		.[ 1, 2 ]
+
+	ortholog_distance_mean = 
+		pairwise_summary %>% 
+		filter(D=="N") %>% 
+		.[["distance"]] %>% 
+		mean()
+
+	paralog_distance_mean = 
+		pairwise_summary %>% 
+		filter(D=="Y") %>% 
+		.[["distance"]] %>% 
+		mean()
+
+```
+
+In order to better understand why our phylogenetic analysis seem to support a different conclusion (*i.e.*, no support for the ortholog conjecture) than the published analysis of KMRR [@KryuchkovaMostacci:2016iw] (*i.e.*, strong support for the ortholog conjecture), we first checked to make sure we could reproduce their specific result.
+This is imporant since we are only looking at at a subset of the data they considered, the Brawand *et al.* 2011 [@Brawand:2011du] dataset.
+We can replicate their result - we find a higher tau correlation between orthologs (R=`r round( ortholog_r, 3) `) then between paralogs (R=`r round( paralog_r, 3)`).
+
+Why is it that pairwise methods and phylogenetic methods lead to the opposite conclusion?
+The reason is that they describe different things.
+Pairwise comparisons describe contemporary patterns, while phylogenetic methods infer historical processes.
+There need not be a different process of evolution following speciation and duplication for there to be different patterns for orthologs and paralogs.
+That is because there is one other important link between evolutionary process and contemporary patterns - the structure of the tree.
+
+Paralogs could be more different from each other than ortholgs if they tend to be more distantly related to each other than orthologs are.
+That would provide more time for differences to accumulate between them, regardless of the history of duplication and speciation.
+This is, in fact, what we find in this dataset.
+While the mean distance between orthologs is `r round( ortholog_distance_mean, 1 )` years, the mean distance between paralogs is `r round( paralog_distance_mean, 1 )` years.
+This is at least in part because, while the maximum distance between orthogs is two times the depth of the oldest speciation event in the tree, many duplications precede this oldest sampled speciation event.
+The distance between paralogs resulting from these deeper duplications can be much larger.
+
+If we consider only the paralogs that arise from the duplication events the same age or younger than the oldest speciation event, the observed pattern is very different.
+The correlation of tau between these shallower paralog pairs, R=`r round( paralog_subset_r, 3)`, is much stronger than for the correlation between all paralog pairs, R=`r round( paralog_r, 3)`.
+This is much closer to the ortholog comparison, which has R=`r round( ortholog_r, 3) `.
+
+The other issue with the pairwise comparisons is that many are summarizing changes along many branches in the phylogeny, not just the branches that follow the divergence event.
+Two paralogs that diverged from a duplication event deep in the tree may have many subsequent duplication and speciation events, and cnages along all these branches will impact the final pattern.
+The phylogenetic methods isolate the changes under consideration.
+There may still be missing speciation events, due to extinction and incomplete taxon sampling, and missing duplication events, due to gene loss.
+But these affect both methods.
+The phylogenetic methods avoid diluting the change that occurs along the branches that follow the node in question with changes along all subsequent branches.
+
+The reason that paralogs pairs have tau values that are more different than orthog pairs is that they tend to be more distantly related, not because there are differences in evolutionary rates of change in tau following duplication and speciation events.
+This means that having information on whether two genes are orthologs or paralogs provides little added information about expression beyond knowing how distintly related the two genes are.
+This has several implications.
+One is that it is an example of the limitations of the terms orthology and paralogy [@Dunn:2016er].
 
@@
 
-While demonstrate a common problem with pairwise comparisons in Levin *et al.* dataset, we did not perform a phylogenetic reanalysis of this study, as we did for the KMRR study.
+While we do demonstrate a common problem with pairwise comparisons in the Levin *et al.* analysis, we did not perform a phylogenetic reanalysis of this study, as we did for the KMRR study.
 This is because the similarity metric computed in the pairwise comparisons of Levin *et al.* are based on different genes for different species pairs.
@@
 A full phylogenetic reanalysis would of course be possible be using upstream analysis products to rederive a new expression summary statistic that was suitable for mapping onto a phylogenetic tree.
-
-## Discussion
 
@@
 
-### Ortholog conjecture
-
-The ortholog conjecture [@Nehrt:2011bm] is the proposition that orthologs (genes that diverged from each other due to a speciation event) have more similar attributes than do paralogs (genes that diverged from each other due to a gene duplication event).
-It has important biological and technical implications.
-It shapes our understanding of the functional diversity of gene families.
-It is used to relate findings from well-studied genes to related genes that haven't been investigated in detail.
-It can be applied to any trait of genes, from gene seqeunce to biochemical properties to expression.
-While the ortholog conjecture describes a specific pattern of functional diversity across genes, it can also be considered a hypothesis about the process of evolution.
-Specifically, it postulates that there is greater evolutionary change in gene attributes following a duplication event than a speciation event.
-
-Despite its importance, there have been relatively few tests of the ortolog conjecture.
-Previous work has shown that ontology annotations are not sufficient to test the ortholog conjecture [@Chen:2012jz; @Thomas:2012hz].
-Analyses of domain structure were consistent with ortholog conjecture [@Forslund:2011cw].
-There have been few tests of the ortholog conjecture with regards to gene expression [@Chen:2012jz].
-
 Our reanalysis of KMRR using phylogenetic methods suggests that the ortholog conjectures is not a dominant pattern that is central to explaining the evolution of phenotypic diversity in gene families.
@@
 
+The widespread invokation of the ortholog conjecture follows from the expectation that it is strong, widely distributed phenomenon.
+Our failure to detect a signature consistent with the ortholog conjecture suggests that it may not be, and that it should not be assumed in systems where it has not been explicitly tested.
+
 ## Conclusions
@@
 The fact that the first two CFG studies we reanalyzed show serious problems with pairwise comparisons indicates that there likely to be similar problems in other CFG studies, and that future CFG studies will be compromised if they continue to use pairwise methods.
+
+Studies of evolutionary functional genomics should not be focused on the tips of the tree, they should explicitly delve into the tree itself.
 
@@
 We also obtained the `Compara.75.protein.nh.emf` gene trees [@Herrero:2016jj] from ftp://ftp.ensembl.org/pub/release-75/emf/ensembl-compara/homologies/ and place them in the same directory as this file.
+
+We considered only the data from Brawand *et al.* 2011 [@Brawand:2011du].
+KMRR examined multiple subsets of the data.
+We left in sex chromosome genes and testes expression data, as this corresponded to the analyses that provided the strongest support for the ortholog conjecture and therefore the most conservative reconsideration of it.
 
@@
 
-```{r filtered, cache=TRUE, echo=FALSE, fig.cap="XX_PIC_VAR", warning=FALSE, message=FALSE}
-
-	# Examine by node age
-
-  g_age =
-  nodes_contrast %>%
-    ggplot( aes( x=node_age, y=pic, col=Event ) ) + 
-		  geom_point( alpha=0.2 ) + 
-		  xlab( "Node age (millions of years)" ) + 
-		  ylab("Phylogenetic independent contrasts (PIC)\n of differential expression") +
-		  ylim(0,0.3) +
-			xlim(0,1000) +
-		  ggtitle( "(a)" ) +
-		  theme_classic()	
-		  #theme( legend.justification=c( 1, 1 ), legend.position=c( 1, 1 ) )
-
-	max_speciation_age = 
-    nodes_contrast %>% 
-    filter( D=="N" ) %>%
-    select( node_age ) %>%
-    max( na.rm=TRUE )
-
-	nodes_contrast_filtered_age = 
-		nodes_contrast %>% 
-		filter( node_age <= max_speciation_age )
-
-	wilcox_filtered_age = 
-		wilcox.test( 
-			nodes_contrast_filtered_age %>% filter(D=="Y") %>% .$pic, 
-			nodes_contrast_filtered_age %>% filter(D=="N") %>% .$pic, 
-			alternative="greater" 
-		)$p.value
+```{r ascertainment_plot, cache=TRUE, echo=FALSE, fig.cap="XX_PIC_VAR", warning=FALSE, message=FALSE}
 
@@
       # theme( legend.justification=c( 1, 1 ), legend.position=c( 1, 1 ) )
-
-  var_exp_max = 
-    nodes_contrast %>% 
-    filter( D=="N" ) %>%
-    select( var_exp ) %>%
-    max( na.rm=TRUE )
-
-  var_exp_min = 
-    nodes_contrast %>% 
-    filter( D=="N" ) %>%
-    select( var_exp ) %>%
-    min( na.rm=TRUE )
-
-  nodes_contrast_filtered_var = 
-    nodes_contrast %>% 
-      filter ( var_exp >=var_exp_min & var_exp <=var_exp_max )
-
-	wilcox_filtered_var = 
-	wilcox.test( 
-		nodes_contrast_filtered_var %>% filter(D=="Y") %>% .$pic, 
-		nodes_contrast_filtered_var %>% filter(D=="N") %>% .$pic, 
-		alternative="greater" 
-	)$p.value
 
@@
 
-```{r pairwise, echo=FALSE, cache=TRUE, message=F, warning=F}
-	get_pairwise_summary = function ( nhx ) {
-
-		# Create a tibble with one row for each pairwise combination of tip nodes
-		ntips = length( nhx@phylo$tip.label )
-		D = 1:ntips %>% combn( 2 ) %>% t() %>% as_tibble()
-		names( D ) = c( "tip_a", "tip_b" )
-
-		# Add a colmn with the gene tree hash 
-		D %<>% mutate( gene_tree = digest( nhx ) )
-
-		# Calculate the phenotypic difference between the tips
-		D$difference = 
-			apply(
-				D,
-				1,
-				function(y) {
-					tau_values = nhx@data$Tau[ c( as.integer(y['tip_a']), as.integer( y['tip_b'] ) ) ]
-					max( tau_values ) - min( tau_values )
-				}
-			)
-
-		# For each row, bet the most recent common ancestor of the pairwise tip combination
-		D$mrca = 
-			apply(
-				D, 
-				1, 
-				function(y) ape::getMRCA( 
-					nhx@phylo, 
-					c( as.integer(y['tip_a']), as.integer(y['tip_b'] ))
-				)
-			) %>%
-			as.integer()
-
-		# Merge data on this mrca to the node
-		D %<>% left_join(
-			nhx@data %>%
-				select( D, node, node_age, pic ),
-			by = c( "mrca" = "node")
-
-		)
-
-		# Calculate the distance between the tips, which for an ultrametric tree 
-		# is twice the age of their most recent common ancestor
-		D %<>% mutate( distance = 2 * node_age )
-
-		return( D )
-
-	}
-
-	# Build a tibble of all pairwise comparisons between tips
-	pairwise_summary = 
-		mclapply( gene_trees_pic, get_pairwise_summary, mc.cores=cores ) %>% 
-		bind_rows()
-
+```{r pairwise_plot, echo=FALSE, cache=TRUE, message=F, warning=F}	
 	pairwise_summary %>% 

### 2017-01-24 manuscript.rmd
typos, polishing

@@
 
-Functional genomics, the study of how genomes work, is one of the most rapidly advancing fields in Biology.
+Functional genomics, the study of how genomes work, is one of the most rapidly advancing fields in biology.
 There is now considerable interest in comparing genome function across species to understand what is conserved, what is variable, and how the diversity of organism traits (including anatomy, physiology, and development) relates to genome diversity.
@@
 Second, it is a missed opportunity to learn about biology that can only be understood in an explicit historical context that considers these relationships.
-Here we examine two recently published comparative expression studies that used multiple pairwise comparison rather than phylogenetic approaches to investigate evolutionary patterns of gene expression.
+Here we examine two recently published comparative gene expression studies that used multiple pairwise comparison rather than phylogenetic approaches to investigate evolutionary patterns of gene expression.
 We find problems with the pairwise comparisons used in both studies that undermine their published conclusions.
-The first study we examine found that expression tends to be more different between paralogs than orthologs, and interpretted this as evidence of the ortholog conjecture- the hypothesis that there are greater rates of evolutionary change following duplication.
+The first study we examine found that expression tends to be more different between paralogs than orthologs, and interpreted this as evidence of the ortholog conjecture- the hypothesis that there are greater rates of evolutionary change following duplication.
 Our phylogenetic analysis finds no difference in rates following duplication and speciation.
-Instead, the greater difference between paralogs is because they tend to be more distantly related to each other than orthologs.
-The second study claimed to find more variance in gene expression at the midpoint of development than earlier and later when comparing animals from different phyla.
-We instead find that changes along the branch of a single taxon, the ctenophore, were overcounted becuase they were included in each pairwise comparison.
-These reanalyses are concrete demonstrations of the inadequacy of pairwise comparisons in comparative genomics studies, and indicates that it will be critical to adopt phylogenetic comparative methods in future analyses.
+Instead, the greater difference between paralogs is because they tend to be more distantly related to each other than orthologs are to each other.
+The second study claimed to find more evolutionary variance in gene expression at a midpoint of embryonic development than earlier and later when comparing distantly related animals.
+We instead find that changes along the branch of a single taxon, the ctenophore, were over counted because they were included in each pairwise comparison with this species.
+These reanalyses are concrete demonstrations of the inadequacy of pairwise comparisons in comparative genomics studies, including gene expression analyses, and indicates that it will be critical to adopt phylogenetic comparative methods in future work.
 
@@
 
-The focus of genomics research has quickly shifted from describing genome sequences to Functional Genomics (FG), the study of how genomes "work" using tools that measure functional attributes such as expression, chromatin state, and transcription initiation.
-FG, in turn, is now becoming more comparative- there is great interest in understanding how FG variation across species gives rise to a diversity of development, morphology, physiology, and other phenotypes [@Wray:2013gh].
+The focus of genomics research has quickly shifted from describing genome sequences to functional genomics, the study of how genomes "work" using tools that measure functional attributes such as expression, chromatin state, and transcription initiation.
+functional genomics, in turn, is now becoming more comparative- there is great interest in understanding how functional genomic variation across species gives rise to a diversity of development, morphology, physiology, and other phenotypes [@Wray:2013gh].
 Comparative Functional Genomics (CFG) analyses are also critical to transferring functional insight across species, and will grow in importance in coming years.
@@
 A rich theoretical and statistical methodology of phylogenetic comparative methods have been developed over the last three decades to address the challenges and opportunities of comparisons across species [@Felsenstein:1985ua; @Grafen:1989um; @Chamberlain:2012vx; @Garamszegi:2014ew; @Pagel:1999gc].
-These have largely been applied to the evolution of morphological and ecological traits, but are just as relevant to CFG.
-Most CFG studies have abstained from phylogenetic approaches and instead heavily rely on pairwise comparisons (Fig.
-1A) that do not account for the fact that evolutionary relationships explain much of the structure of variation across species.
+These challenges include non-independence of obserbations across species -- more closely related species will tend to be more similar to each other than distantly related species.
+Phylogenetic comparative methods have largely been applied to the evolution of morphological and ecological traits, but are just as relevant to CFG.
+Most CFG studies have abstained from phylogenetic approaches and instead heavily rely on multiple pairwise comparisons (Fig.
+1A) between contemporary traits that do not account for the fact that evolutionary relationships explain much of the structure of variation across species.
 This leaves CFG studies susceptible to serious statistical errors and is a missed opportunity to ask questions that are only accessible in an explicit phylogenetic context.
 
-One reason that CFG has not yet embraced phylogenetic approaches is that it has not yet been concretely demonstrated that pairwise and phylogenetic approaches can lead to different results in CFG.
-Although the value of phylogenetic approaches has been repeatedly shown in comparative analyses of other character data [@Chamberlain:2012vx; @Ricklefs:1996wb], this value is apparently not well known to the FG community since the field has until recently largely focused on within-species analyses.
+One reason that CFG has not yet embraced phylogenetic approaches is that it has not yet been concretely demonstrated that pairwise and phylogenetic approaches can lead to different results in CFG studies.
+In this manuscript we examine this issue.
+Although the value of phylogenetic approaches has been repeatedly shown in comparative analyses of other types of character data [@Chamberlain:2012vx; @Ricklefs:1996wb], this value is apparently not well known to the functional genomics community since the field has until recently largely focused on within-species analyses.
 Many initial CFG studies have also been focused on only two species at a time, where the differences between phylogenetic and pairwise approaches are not as pronounced.
@@
 Most CFG studies to date compare gene expression across species, often between distinct tissues or developmental time points.
-Gene expression has been a popular target of CFG studies because, with the advent of RNA-Seq, it is the most readily collected FG trait across diverse species.
+Gene expression has been a popular target of CFG studies because, with the advent of RNA-Seq and related transcriptome sequencing methods, it is the most readily collected functional genomics trait across diverse species.
 Here we reconsider two recent comparative analyses of gene expression data [@KryuchkovaMostacci:2016iw; @Levin:2016bp].
 
-Kryuchkova-Mostacci and Robinson-Rechavi (KMRR) 2016 [@KryuchkovaMostacci:2016iw] integrated vertebrate expression datasets to test the ortholog conjecture - the hypothesis that orthologs tend of have more conserved attributes (expression in this case) than do paralogs [@Nehrt:2011bm].
+Kryuchkova-Mostacci and Robinson-Rechavi (KMRR) 2016 [@KryuchkovaMostacci:2016iw] analyzed multiple vertebrate expression datasets to test the ortholog conjecture - the hypothesis that orthologs tend of have more conserved attributes (specificity of expression across organs in this case) than do paralogs [@Nehrt:2011bm].
 Using pairwise comparisons (Fig.
-1A), they found lower expression covariance between paralogs than between orthologs and interpreted this as strong support for the ortholog conjecture.
-Orthology and paralogy is annotated based on gene phylogenies, but these phylogenies are not used when making the expression comparisons.
+1A), they found lower expression correlation between paralogs than between orthologs and interpreted this as strong support for the ortholog conjecture.
 
@@
 This study analyzed gene expression through the course of embryonic development for ten animal species, each from a different phylum.
-They concluded that animals from different phyla exhibit an "inverse hourglass" model for the evolution of gene expression, where there is more evolutionary variance in gene expression at a mid phase of development than there is at early and late phases (Fig.
+They concluded that these animals exhibit an "inverse hourglass" model for the evolution of gene expression, where there is more evolutionary variance in gene expression at a mid phase of development than there is at early and late phases (Fig.
 2).
@@
 
-#### Original analyses
+#### Original pairwise analyses
 
@@
 The ortholog conjecture [@Nehrt:2011bm] is the proposition that orthologs (genes that diverged from each other due to a speciation event) have more similar attributes than do paralogs (genes that diverged from each other due to a gene duplication event).
-It has important biological and technical implications.
+The ortholog conjecture has important biological and technical implications.
 It shapes our understanding of the functional diversity of gene families.
@@
 
-Despite its importance, there have been relatively few tests of the ortolog conjecture.
+Despite its importance, there have been relatively few tests of the ortholog conjecture.
 Previous work has shown that ontology annotations are not sufficient to test the ortholog conjecture [@Chen:2012jz; @Thomas:2012hz].
 Analyses of domain structure were consistent with ortholog conjecture [@Forslund:2011cw].
-There have been few tests of the ortholog conjecture with regards to gene expression [@Chen:2012jz], and KMRR is the most thorough such study to date.
+There have been few tests of the ortholog conjecture with regards to gene expression [@Chen:2012jz], and KMRR is the most thorough expression study to date.
 
@@
 Here we consider their conclusions in the context of Brawand *et al.* 2011 [@Brawand:2011du], one of the studies they considered.
-This dataset of gene expression for XX different organs across XX mammal species is the best sampled in their reanalyses.
+This dataset of gene expression for six organs across 10 species (nine mammals and a bird) is the best sampled in their reanalyses.
 Their expression summary statistic is Tau [@Yanai:2005dh, @KryuchkovaMostacci:2016in], an indicator of tissue specificity of gene expression.
-It can range from a value of 0, which indiciates no specifificity (*i.e.*, uniform expression across tissues), to a value of 1, which indicates his specificity.
-It is convenient in that it is a single number of defined range for each gene in each species, though of course since the original expression is multidemensional this means much information is discarded.
+It can range from a value of 0, which indicates no specificity (*i.e.*, uniform expression across tissues), to a value of 1, which indicates high specificity (*i.e.*, expression in only one tissue).
+It is convenient in that it is a single number of defined range for each gene, though of course since the original expression is multidimensional this means much information is discarded.
 For example, it does not convey which tissue expression is specific to.
 Homologous genes could each have a tau value of 1, but one gene could have expression that is specific to the kidney and the other to the brain, for example.
-Their published analyses are based on pairwise comparisons (Figure 1a) betwen tau for each gene family.
-They found the correlation for orthologs to be significantly greater than the correlation of tau for paralogs, *i.e.* that orthologs tend to be more similar to each other than paralogs.
-From this concluded that their analyses support the ortholog conjecture.
-They also went one step further - they concluded that this pattern provides support for a particular evolutionary process, that "tissue-specificity evolves very slowly in the absence of duplication, while immediately after duplication the new gene copy differs."
+
+KMRR's published analyses are based on pairwise comparisons (Figure 1a) between tau within each gene family.
+They found the correlation of tau for orthologs to be significantly greater than the correlation of tau for paralogs, *i.e.* that orthologs tend to be more similar to each other than paralogs.
+From this they concluded that their analyses support the ortholog conjecture.
+They also went one step further.
+They concluded that this pattern provides support for a particular evolutionary process, that "tissue-specificity evolves very slowly in the absence of duplication, while immediately after duplication the new gene copy differs".
 
@@
 
-We reanalyzed the their study using phylogenetic comparative methods, focusing on the dataset of Brawand *et al.* 2011 [Brawand:2011du].
+We reanalyzed the the KMRR study using phylogenetic comparative methods, focusing on the dataset of Brawand *et al.* 2011 [Brawand:2011du].
 For each internal node in each gene tree (Figure 1b), we measured the phylogenetically independent contrast (PIC) in the expression across this node [@Felsenstein:1985ua].
-This is the difference in values of descendent nodes (*i.e.*, the evulutionary change along the branches that descend from the node) scaled by the expected variance (determined largely by the lengths of these branches).
+This is the difference in values of descendant nodes (*i.e.*, the evulutionary change along the branches that descend from the node) scaled by the expected variance (determined largely by the lengths of these branches).
 These contrasts were then annotated by whether they are made across branches that descend from a speciation or duplication event.
@@
 
-To test the ortholog conjecture it is essental to have a null hypothesis that results in a distinct prediction from the ortholog conjecture.
+To test the ortholog conjecture it is essential to have a null hypothesis that results in a distinct prediction from the ortholog conjecture.
 A suitable null hypothesis is that there is no difference in evolutionary change in differential expression along branches that descend from speciation or duplication events (Figure 1b).
@@
         geom_freqpoly( binwidth=.001, position="identity" ) +
-        xlab("Phylogenetic independent \ncontrasts (PIC)") +
-				xlim(0,0.075) + 
+        xlab( "Phylogenetic independent \ncontrasts (PIC)" ) +
+				ylab( "Density" ) +
+				xlim( 0, 0.075 ) + 
 				theme_classic()
 
-	wilcox_test_result = wilcox.test( nodes_contrast %>% filter(D=="Y") %>% .$pic, nodes_contrast %>% filter(D=="N") %>% .$pic, alternative="greater" )
+	wilcox_test_result = 
+		wilcox.test( 
+			nodes_contrast %>% filter(D=="Y") %>% .$pic, 
+			nodes_contrast %>% filter(D=="N") %>% .$pic, 
+			alternative="greater" 
+		)
 
@@
 The independent contrast across a node is the amount of change observed between the daughter notes, scaled by the expected variance [Felsenstein:1985ua].
-The expected variance is principaly determined by the lengths of the branches leading from the node to these daughters.
+The expected variance is principally determined by the lengths of the branches leading from the node to these daughters.
 The same observed difference between daughter nodes will lead to a larger contrast the shorter the branches are to these nodes, as this indicates a greater rate of evolutionary change.
@@
 We therefore also examine only the contrasts that fell within in the range of expected variance seen for speciation contrasts, excluding duplication contrasts that fall outside of this range.
-This reanalysys does not reject the ortholog conjecture either (Figure XXAscertainment d).
+This reanalysis does not reject the ortholog conjecture either (Figure XXAscertainment d).
 
@@
 In order to better understand why our phylogenetic analysis seem to support a different conclusion (*i.e.*, no support for the ortholog conjecture) than the published analysis of KMRR [@KryuchkovaMostacci:2016iw] (*i.e.*, strong support for the ortholog conjecture), we first checked to make sure we could reproduce their specific result.
-This is imporant since we are only looking at at a subset of the data they considered, the Brawand *et al.* 2011 [@Brawand:2011du] dataset.
+This is important since we are only looking at at a subset of the data they considered, the Brawand *et al.* 2011 [@Brawand:2011du] dataset.
 We can replicate their result - we find a higher tau correlation between orthologs (R=`r round( ortholog_r, 3) `) then between paralogs (R=`r round( paralog_r, 3)`).
@@
 
-Paralogs could be more different from each other than ortholgs if they tend to be more distantly related to each other than orthologs are.
+Paralogs could be more different from each other than orthologs if they tend to be more distantly related to each other than orthologs are.
 That would provide more time for differences to accumulate between them, regardless of the history of duplication and speciation.
@@
 While the mean distance between orthologs is `r round( ortholog_distance_mean, 1 )` years, the mean distance between paralogs is `r round( paralog_distance_mean, 1 )` years.
-This is at least in part because, while the maximum distance between orthogs is two times the depth of the oldest speciation event in the tree, many duplications precede this oldest sampled speciation event.
+This is at least in part because, while the maximum distance between orthologs is two times the depth of the oldest speciation event in the tree, many duplications precede this oldest sampled speciation event.
 The distance between paralogs resulting from these deeper duplications can be much larger.
@@
 The other issue with the pairwise comparisons is that many are summarizing changes along many branches in the phylogeny, not just the branches that follow the divergence event.
-Two paralogs that diverged from a duplication event deep in the tree may have many subsequent duplication and speciation events, and cnages along all these branches will impact the final pattern.
+Two paralogs that diverged from a duplication event deep in the tree may have many subsequent duplication and speciation events, and changes along all these branches will impact the final pattern.
 The phylogenetic methods isolate the changes under consideration.
@@
 
-The reason that paralogs pairs have tau values that are more different than orthog pairs is that they tend to be more distantly related, not because there are differences in evolutionary rates of change in tau following duplication and speciation events.
-This means that having information on whether two genes are orthologs or paralogs provides little added information about expression beyond knowing how distintly related the two genes are.
+The reason that paralogs pairs have tau values that are more different than ortholog pairs is that they tend to be more distantly related, not because there are differences in evolutionary rates of change in tau following duplication and speciation events.
+This means that having information on whether two genes are orthologs or paralogs provides little added information about expression beyond knowing how distantly related the two genes are.
 This has several implications.
 One is that it is an example of the limitations of the terms orthology and paralogy [@Dunn:2016er].
+These terms can have straightforward meaning in small gene trees with simple duplication/speciation histories, but the utility of the terms breaks down on larger more complex gene trees.
+Orthology and paralogy are annotations on the tips of the tree that are derived from the infered structure of the tree and annotations of duplication and speciation at internal tree nodes.
+In this sense, orthology and paralogy are statements about the internals of the tree that are distilled into statements at the tips of the tree.
+Much is lost in the process, though, and for most questions it is much more direct to focus on the structure of the tree and the inferred processes of internal nodes.
+In this case, KMRR [@KryuchkovaMostacci:2016iw] relied on orthology and paralogy annotations that are derived from the infered duplication and speciation history on gene phylogenies [@Herrero:2016jj].
+These are the same pre-computed trees we use in our phylogenetic re-analyses of their study.
 
@@
 This means that there is not a trait that can be mapped onto the tree, as KMRR's tau can be.
-A full phylogenetic reanalysis would of course be possible be using upstream analysis products to rederive a new expression summary statistic that was suitable for mapping onto a phylogenetic tree.
+A full phylogenetic reanalysis would of course be possible be using upstream analysis products to re-derive a new expression summary statistic that was suitable for mapping onto a phylogenetic tree.
 
@@
 In CFG analyses, there are often far fewer species (e.g., 3-10) because adding taxa is still expensive, but tens of thousands of traits.
-This creates challenges as the resulting covariance matrices are singular and, if not treated appropriately, imply many false correlations that are artefacts of project design.
+This creates challenges as the resulting covariance matrices are singular and, if not treated appropriately, imply many false correlations that are artifacts of project design.
 We outlined these challenges and potential solutions in the context of gene expression [31].
 
-It is not enough to demonstrate the impact of phylogenetics on CFG analyses, these methods need to be made accessible to the FG community so they can be widely.
+It is not enough to demonstrate the impact of phylogenetic methods on CFG analyses, these methods need to be made accessible to the functional genomics community so they can be widely applied.
 
 Most early phylogenetic comparative methods attempted to remove evolutionary signal to correct statistical tests for correlations between traits, while more recent methods tend to focus on testing hypotheses of evolutionary processes [34].
-The application of this newer focus to pCFG provides an exciting opportunity to address long standing questions of broad interest, including the order of changes in FG traits and shifts in rates of evolution of one FG trait following changes in another trait.
+The application of this newer focus to pCFG provides an exciting opportunity to address long standing questions of broad interest, including the order of changes in functional genomics traits and shifts in rates of evolution of one functional genomics trait following changes in another trait.
 
@@
 Our reanalysis of KMRR using phylogenetic methods suggests that the ortholog conjectures is not a dominant pattern that is central to explaining the evolution of phenotypic diversity in gene families.
-This has impotant biological implications.
+This has important biological implications.
 It suggests that the mechanism of gene divergence (speciation versus duplication) may not have as strong an impact on phenotypic divergence as thought.
@@
 
-The widespread invokation of the ortholog conjecture follows from the expectation that it is strong, widely distributed phenomenon.
+The widespread invocation of the ortholog conjecture follows from the expectation that it is strong, widely distributed phenomenon.
 Our failure to detect a signature consistent with the ortholog conjecture suggests that it may not be, and that it should not be assumed in systems where it has not been explicitly tested.
@@
 
-All files needed to reexecute the anlyses presented in this document are available at https://github.com/caseywdunn/comparative_expression_2017.
+All files needed to re-execute the analyses presented in this document are available at https://github.com/caseywdunn/comparative_expression_2017.
 The most recent commit at the time of the analysis presented here was `r system("git log | head -n 1", intern=TRUE) %>% str_replace("commit ", "")`.
@@
 We slightly altered their `Rscript.R` to simplify file paths.
-This midified script and their data files are available in the gihub respository for this paper, as are the intermediate files that were generated by their analysis script that we used in our own analyses.
+This modified script and their data files are available in the github repository for this paper, as are the intermediate files that were generated by their analysis script that we used in our own analyses.
 We also obtained the `Compara.75.protein.nh.emf` gene trees [@Herrero:2016jj] from ftp://ftp.ensembl.org/pub/release-75/emf/ensembl-compara/homologies/ and place them in the same directory as this file.
@@
 Each of these gene trees includes includes homologous sequences derived from a common ancestral sequence.
-The tip nodes of a gene tree can include both orthogs and paralogs, and each internal node is an infered speciation or duplication event.
+The tip nodes of a gene tree can include both orthologs and paralogs, and each internal node is an inferred speciation or duplication event.
 Prior to phylogenetic analyses of gene expression, each tree was first pruned to include only tips for which expression data were available.
@@
 
-## Ackowledgements
+## Acknowledgments
 
@@
 
-> Supplementary Figure XXAscertainment | Investigation of possible assertainment biases.
-(a) Magnitude of independent contrasts ploted against node age.
+> Supplementary Figure XXAscertainment | Investigation of possible ascertainment biases.
+(a) Magnitude of independent contrasts plotted against node age.
 Speciation nodes are calibrated to particular times, whereas duplication nodes have a wider range.
 (b) Density plot of contrasts for only the nodes that have an age less than or equal to the maximum age of speciation nodes.
-(c) Magnitude of independent contrasts ploted against expected variance, which is largely determined by branch lengths.
+(c) Magnitude of independent contrasts plotted against expected variance, which is largely determined by branch lengths.
 Contrasts for speciation nodes have a narrower range of expected variance than do contrasts for duplication nodes.

### 2017-01-25 manuscript.rmd
Update manuscript.rmd
code formatting

@@
 ```{r setup, cache=TRUE, message=F, echo=FALSE, warning=F}
-knitr::opts_chunk$set(echo = TRUE)
+
+	knitr::opts_chunk$set( echo = TRUE )
+
 ```
@@
 
-^4^ Sars Institute, Bergen, Norway
+^4^ Sars International Centre for Marine Molecular Biology, University of Bergen, Bergen, Norway
 
@@
 
-Kryuchkova-Mostacci and Robinson-Rechavi (KMRR) 2016 [@KryuchkovaMostacci:2016iw] analyzed multiple vertebrate expression datasets to test the ortholog conjecture - the hypothesis that orthologs tend of have more conserved attributes (specificity of expression across organs in this case) than do paralogs [@Nehrt:2011bm].
+The first study, Kryuchkova-Mostacci and Robinson-Rechavi (KMRR) 2016 [@KryuchkovaMostacci:2016iw], analyzed multiple vertebrate expression datasets to test the ortholog conjecture - the hypothesis that orthologs tend of have more conserved attributes (specificity of expression across organs in this case) than do paralogs [@Nehrt:2011bm].
 Using pairwise comparisons (Fig.
@@
 > Figure 1 | Pairwise and phylogenetic comparative approaches.
-(a) Many comparative functional genomic studies across species rely on pairwise comparisons, where traits of each gene are compared to traits of every other gene.
+(a) Many comparative functional genomic studies rely on pairwise comparisons, where traits of each gene are compared to traits of other genes (such as orthologs and paralogs).
 This leads to many more comparisons than unique observations, making each comparison non-independent.
-(b) Comparative phylogenetic methods, including phylogenetic independent contrasts [@Felsenstein:1985ua], make a smaller number of independent comparisons, where each contrast measures changes along different branches.
+(b) Comparative phylogenetic methods, including phylogenetic independent contrasts [@Felsenstein:1985ua], make a smaller number of independent comparisons, where each contrast measures independent changes along different branches.
 Phylogenetic approaches are rarely used for functional genomics studies.
@@
 
-#### Original pairwise analyses
+#### Original pairwise test of the ortholog conjecture
 
@@
 It shapes our understanding of the functional diversity of gene families.
-It is used to relate findings from well-studied genes to related genes that haven't been investigated in detail.
+It is used to relate findings from well-studied genes to related genes that have not been investigated in detail.
 It can be applied to any trait of genes, from gene seqeunce to biochemical properties to expression.
-While the ortholog conjecture describes a specific pattern of functional diversity across genes, it can also be considered a hypothesis about the process of evolution.
-Specifically, it postulates that there is greater evolutionary change in gene attributes following a duplication event than a speciation event.
+While the ortholog conjecture describes a specific pattern of functional diversity across genes, it is also articulated as a hypothesis about the process of evolution-- that there is greater evolutionary change in gene attributes following a duplication event than a speciation event.
 
@@
 Analyses of domain structure were consistent with ortholog conjecture [@Forslund:2011cw].
-There have been few tests of the ortholog conjecture with regards to gene expression [@Chen:2012jz], and KMRR is the most thorough expression study to date.
+There have been few tests of the ortholog conjecture with regards to gene expression [@Chen:2012jz], and KMRR is the most thorough such expression study to date.
 
 The KMRR study considered several publicly available datasets.
-Here we consider their conclusions in the context of Brawand *et al.* 2011 [@Brawand:2011du], one of the studies they considered.
-This dataset of gene expression for six organs across 10 species (nine mammals and a bird) is the best sampled in their reanalyses.
+Here we consider their conclusions in the context of one of these studies, Brawand *et al.* 2011 [@Brawand:2011du].
+This dataset of gene expression for six organs across 10 species (nine mammals and a bird) is the best sampled in their analyses.
 Their expression summary statistic is Tau [@Yanai:2005dh, @KryuchkovaMostacci:2016in], an indicator of tissue specificity of gene expression.
@@
 
-	expression_file_names %<>% paste(kmrr_directory, ., sep="")
+	expression_file_names %<>% paste( kmrr_directory, ., sep="" )
 
-	gene_tree_name %<>% paste(kmrr_directory, ., sep="")
+	gene_tree_name %<>% paste( kmrr_directory, ., sep="" )
 
@@
 		age.min =
-			c(20, 92, 167, 8, 42, 74, 296, 535, 104, 937, 29, 722, 441, 65, 15, 1215, 
-				414, 371, 162, 25, 74, 77, 86), 
-		clade = c("Hominoidea", "Euarchontoglires", "Mammalia", "Homininae", 
+			c( 20, 92, 167, 8, 42, 74, 296, 535, 104, 937, 29, 722, 441, 65, 15, 1215, 
+				414, 371, 162, 25, 74, 77, 86 ), 
+		clade = 
+			c("Hominoidea", "Euarchontoglires", "Mammalia", "Homininae", 
 					"Simiiformes", "Primates", "Amniota", "Vertebrata", "Eutheria", 
@@
 					"Haplorrhini", "Hominidae", "Opisthokonta", "Sarcopterygii", 
-							"Tetrapoda", "Theria", "Murinae", "Sciurognathi", "Rodentia", "Glires"),
+					"Tetrapoda", "Theria", "Murinae", "Sciurognathi", "Rodentia", "Glires"
+			),
 		stringsAsFactors=FALSE
@@
 			calibration = 
-				nhx@data[ !is.tip.nhx( nhx ) ,  ] %>%
-				filter( D=="N" ) %>%
+				nhx@data[ !is.tip.nhx( nhx ), ] %>%
+				filter( D == "N" ) %>%
 				left_join( calibration_times, c( "label" = "clade" ) ) %>%
@@
 
-			tree = try( ape::chronos( nhx@phylo, calibration=calibration, model="correlated" ) )
-			if( "phylo" %in% class(tree) ){
+			tree = try( 
+				ape::chronos( nhx@phylo, calibration=calibration, model="correlated" ) 
+			)
+
+			if( "phylo" %in% class( tree ) ){
 				class( tree ) = "phylo"
@@
 ```{r load_expression, echo=FALSE, message=F, warning=F}
-	expression = lapply(
-		expression_file_names, read.table, stringsAsFactors=FALSE, header=TRUE
+
+	expression = 
+	lapply(
+		expression_file_names, 
+		read.table, 
+		stringsAsFactors=FALSE, 
+		header=TRUE
 	) %>%
 	bind_rows()
+
 ```
@@
 			tree_tc = textConnection( x )
-			tree <- treeio::read.nhx( tree_tc )
+			tree = treeio::read.nhx( tree_tc )
 			close( tree_tc )
@@
 ```{r add_expression_to_trees, echo=FALSE, cache=TRUE, message=F, warning=F}
+
 	gene_trees_annotated = mclapply(
@@
 		function( tree ){
-			tree@data %<>% left_join( expression, by = c("G" = "Ensembl.Gene.ID") )
+			tree@data %<>% left_join( expression, by = c( "G" = "Ensembl.Gene.ID" ) )
 			return( tree )
-		}, mc.cores=cores
+		}, 
+		mc.cores=cores
 	)
+
 ```
@@
 ```{r prep_trees, echo=FALSE, cache=TRUE, message=F, warning=F}
+
 	drop_empty_tips = function( nhx ) {
-	  to_drop = which( is.na( nhx@data[ 1:length(nhx@phylo$tip.label), ]$Tau ) )
-	  remaining = length(nhx@phylo$tip.label) - length(to_drop)
+		to_drop = which( is.na( nhx@data[ 1:length( nhx@phylo$tip.label ), ]$Tau ) )
+		remaining = length( nhx@phylo$tip.label ) - length( to_drop )
 
@@
 
-	gene_trees_calibrated = mclapply( 
+	gene_trees_calibrated = 
+		mclapply( 
 			gene_trees_sub,
@@
 
-	gene_trees_calibrated = mclapply( gene_trees_calibrated, store_node_age, mc.cores=cores )
+	gene_trees_calibrated = 
+		mclapply( 
+			gene_trees_calibrated, 
+			store_node_age, 
+			mc.cores=cores 
+		)
 
@@
 
-	all_nodes = lapply( 
+	all_nodes = 
+		lapply( 
 			gene_trees_pic, 
@@
 				tags$gene = digest( nhx )
-	  			tags %<>% select( -(B) ) # B has inconsistent types, remove it
-	  			return(tags)
+				tags %<>% select( -( B ) ) # B has inconsistent types, remove it
+				return( tags )
 			}
@@
 		all_nodes %>%
-		filter( ! is.na(pic) ) %>%
-		mutate( pic=abs(pic) ) %>%
-		filter( ! is.na(D) )
+		filter( ! is.na( pic ) ) %>%
+		mutate( pic = abs( pic ) ) %>%
+		filter( ! is.na( D ) )
 
@@
 		wilcox.test( 
-			nodes_contrast %>% filter(D=="Y") %>% .$pic, 
-			nodes_contrast %>% filter(D=="N") %>% .$pic, 
+			nodes_contrast %>% filter( D=="Y" ) %>% .$pic, 
+			nodes_contrast %>% filter( D=="N" ) %>% .$pic, 
 			alternative="greater" 
@@
 			xlab( "Node age (millions of years)" ) + 
-		  ylab("Phylogenetic independent contrasts (PIC)\n of differential expression") +
-		  ylim(0,0.3) +
-			xlim(0,1000) +
+			ylab( "Phylogenetic independent contrasts (PIC)\n of differential expression" ) +
+			ylim( 0, 0.3 ) +
+			xlim( 0, 1000 ) +
 			ggtitle( "(a)" ) +
@@
 		wilcox.test( 
-			nodes_contrast_filtered_age %>% filter(D=="Y") %>% .$pic, 
-			nodes_contrast_filtered_age %>% filter(D=="N") %>% .$pic, 
+			nodes_contrast_filtered_age %>% filter( D=="Y" ) %>% .$pic, 
+			nodes_contrast_filtered_age %>% filter( D=="N" ) %>% .$pic, 
 			alternative="greater" 
@@
 		nodes_contrast %>% 
-      filter ( var_exp >=var_exp_min & var_exp <=var_exp_max )
+		filter ( ( var_exp >= var_exp_min ) & ( var_exp <= var_exp_max ) )
 
@@
 	wilcox.test( 
-		nodes_contrast_filtered_var %>% filter(D=="Y") %>% .$pic, 
-		nodes_contrast_filtered_var %>% filter(D=="N") %>% .$pic, 
+		nodes_contrast_filtered_var %>% filter( D=="Y" ) %>% .$pic, 
+		nodes_contrast_filtered_var %>% filter( D=="N" ) %>% .$pic, 
 		alternative="greater" 
@@
 
-		# Calculate the phenotypic difference between the tips with a different method as 
-		# check on parsing
-		D$difference_old = 
-			apply(
-				D,
-				1,
-				function(y) {
-					tau_values = nhx@data$Tau[ c( as.integer(y['tip_a']), as.integer( y['tip_b'] ) ) ]
-					max( tau_values ) - min( tau_values )
-				}
-			)
-
-		if ( ! all ( near( D$difference, D$difference_old ) ) ){
-			stop( "Inconsistent difference calculations" )
-		}
-
 		# For each row, bet the most recent common ancestor of the pairwise tip combination
@@
 				1, 
-				function(y) ape::getMRCA( 
+				function( y ) ape::getMRCA( 
 					nhx@phylo, 
-					c( as.integer(y['tip_a']), as.integer(y['tip_b'] ))
+					c( as.integer( y['tip_a'] ), as.integer( y['tip_b'] ) )
 				)
@@
 		pairwise_summary %>% 
-		filter(D=="N") %>% 
+		filter( D=="N" ) %>% 
 		.[["distance"]] %>% 
@@
 		pairwise_summary %>% 
-		filter(D=="Y") %>% 
+		filter( D=="Y" ) %>% 
 		.[["distance"]] %>% 
@@
 		geom_freqpoly( binwidth=.001, position="identity" ) +
-      xlab("Phylogenetic independent \ncontrasts (PIC)") +
-			ylab("Density") +
-			xlim(0,0.075) + 
+		xlab( "Phylogenetic independent \ncontrasts (PIC)" ) +
+		ylab( "Density" ) +
+		xlim( 0, 0.075 ) + 
 		ggtitle( "(b)" ) +
-			annotate("text", x = Inf, y = Inf, hjust="right", vjust="top", label=label_age ) +
+		annotate( "text", x = Inf, y = Inf, hjust="right", vjust="top", label=label_age ) +
 		theme_classic()
@@
 		xlab( "Expected variance, \n depends mostly on branch length" ) + 
-      ylab("Phylogenetic independent contrasts (PIC)\n of differential expression") +
-      ylim(0,0.3) +
-			xlim(0,2000) +
+		ylab( "Phylogenetic independent contrasts (PIC)\n of differential expression" ) +
+		ylim( 0, 0.3 ) +
+		xlim( 0, 2000 ) +
 		ggtitle( "(c)" ) +
 		theme_classic()
-      # theme( legend.justification=c( 1, 1 ), legend.position=c( 1, 1 ) )
 
@@
 		theme( legend.justification=c( 1, 1 ), legend.position=c( 1, 1 ) ) +
-      xlim(0,0.075) + 
+		xlim( 0, 0.075 ) + 
 		theme_classic() +
-  		annotate("text", x = Inf, y = Inf, hjust="right", vjust="top", label=label_var ) +
-      xlab("Phylogenetic independent contrasts (PIC)\n of differential expression") +
-  		ylab("Density")
+		annotate( "text", x = Inf, y = Inf, hjust="right", vjust="top", label=label_var ) +
+		xlab( "Phylogenetic independent contrasts (PIC)\n of differential expression" ) +
+		ylab( "Density" )
 
@@
 	pairwise_summary %>% 
-		filter( ! is.na(D) ) %>%
-  	ggplot(aes(x=distance, y=difference, color=D)) + 
-  		geom_point(alpha=0.1) +
-  		geom_smooth(method=lm, se=FALSE) +
+	filter( ! is.na( D ) ) %>%
+	ggplot( aes( x=distance, y=difference, color=D ) ) + 
+		geom_point( alpha=0.1 ) +
+		geom_smooth( method=lm, se=FALSE ) +
 		xlim( 0, 2000 )

### 2017-01-25 manuscript.rmd
more polishing

@@
 Here we consider their conclusions in the context of one of these studies, Brawand *et al.* 2011 [@Brawand:2011du].
-This dataset of gene expression for six organs across 10 species (nine mammals and a bird) is the best sampled in their analyses.
+This dataset of gene expression for six organs across 10 species (nine mammals and one bird) is the best sampled in their analyses.
 Their expression summary statistic is Tau [@Yanai:2005dh, @KryuchkovaMostacci:2016in], an indicator of tissue specificity of gene expression.
@@
 It is convenient in that it is a single number of defined range for each gene, though of course since the original expression is multidimensional this means much information is discarded.
-For example, it does not convey which tissue expression is specific to.
-Homologous genes could each have a tau value of 1, but one gene could have expression that is specific to the kidney and the other to the brain, for example.
+This includes information about which tissue expression is specific to.
+For example, if one gene has expression specific to the brain and another expression specific to the kidney, both would have a Tau of 1.
 
-KMRR's published analyses are based on pairwise comparisons (Figure 1a) between tau within each gene family.
-They found the correlation of tau for orthologs to be significantly greater than the correlation of tau for paralogs, *i.e.* that orthologs tend to be more similar to each other than paralogs.
+KMRR's published analyses are based on pairwise comparisons (Figure 1a) between Tau within each gene family.
+They found the correlation of Tau for orthologs to be significantly greater than the correlation of Tau for paralogs, *i.e.* that orthologs tend to be more similar to each other than paralogs.
 From this they concluded that their analyses support the ortholog conjecture.
-They also went one step further.
-They concluded that this pattern provides support for a particular evolutionary process, that "tissue-specificity evolves very slowly in the absence of duplication, while immediately after duplication the new gene copy differs".
+They also concluded that this pattern provides support for a particular evolutionary process, that "tissue-specificity evolves very slowly in the absence of duplication, while immediately after duplication the new gene copy differs".
 
@@
 
-We reanalyzed the the KMRR study using phylogenetic comparative methods, focusing on the dataset of Brawand *et al.* 2011 [Brawand:2011du].
-For each internal node in each gene tree (Figure 1b), we measured the phylogenetically independent contrast (PIC) in the expression across this node [@Felsenstein:1985ua].
-This is the difference in values of descendant nodes (*i.e.*, the evulutionary change along the branches that descend from the node) scaled by the expected variance (determined largely by the lengths of these branches).
-These contrasts were then annotated by whether they are made across branches that descend from a speciation or duplication event.
+We reanalyzed the the KMRR study using phylogenetic comparative methods, focusing on the dataset of Brawand *et al.* 2011 [@Brawand:2011du].
+For each internal node in each gene tree (Figure 1b), we measured the phylogenetic independent contrast[@Felsenstein:1985ua] (PIC) of the expression statistic Tau.
+This is the difference in values of Tau for descendant nodes scaled by the expected variance (determined largely by the lengths of these branches).
+These contrasts were then annotated by whether they are made across a speciation or duplication event.
 The original description of independent contrasts [@Felsenstein:1985ua] focused on assessing covariance between changes in two traits on a tree.
-Our use of it to look for differences in evolutionary changes of one trait (differential expression) between two categories of nodes (speciation and duplication) on a tree is a bit different than the usual application of the method.
+Our use of contrasts is a bit different-- we look for differences in evolutionary changes of one trait (differential expression) between two categories of nodes (speciation and duplication) on a tree.
 
@@
 
-> Figure XXExpectations | (a) Null hypothesis. (b) Under the ortholog conjecture.
+> Figure XXExpectations | (a) Under the null hypothesis that there is no difference in the rate of evolution after duplication or speciation events, the distribution of PICs would be the same across both types of nodes.
+(b) Under the ortholog conjecture, contrasts across duplication nodes would tend to be larger than contrasts across speciation nodes.
 
-To test the ortholog conjecture it is essential to have a null hypothesis that results in a distinct prediction from the ortholog conjecture.
+To test the ortholog conjecture it is essential to have a null hypothesis that makes predictions distinct from the ortholog conjecture.
 A suitable null hypothesis is that there is no difference in evolutionary change in differential expression along branches that descend from speciation or duplication events (Figure 1b).
@@
 Under the alternative hypothesis specified by the ortholog conjecture, that there is a higher rate of change following duplication events than speciation events, we would expect to see the distribution of duplication contrasts shifted to higher values relative to the speciation contrasts (Figure XXEXPECTATIONS b).
-
-When we reanalyze the data with phylogenetic independent contrasts (Fig. 1B), we did not find increased evolutionary change in expression following duplication events (Fig. XX).
 
@@
 	gene_tree_name %<>% paste( kmrr_directory, ., sep="" )
+
+	# The minimum number of genes with expression data that a gene needs to have to be considered
+	min_genes_with_expression = 4
 
@@
 
-		if( remaining > 3 ){
+		if( remaining >= min_genes_with_expression ){
 			pruned = treeio::drop.tip( nhx, to_drop )
@@
 ```
-
-Of the `r length(gene_trees)` trees that were parsed, `r length(gene_trees_sub)` passed taxon sampling criteria and `r length(gene_trees_calibrated)` were successfully time calibrated.
 
@@
 
-There were `r nodes_contrast %>% filter(D=="Y") %>% nrow()` duplication nodes and `r nodes_contrast %>% filter(D=="N") %>% nrow()` speciation nodes.
+When we reanalyze the data with phylogenetic independent contrasts (Fig.
+1B), we did not find increased evolutionary change in expression following duplication events (Fig.
+XXKMRR).
+Of the `r length(gene_trees)` trees that were parsed from ENSMBL Compara [@Herrero:2016jj], `r length(gene_trees_sub)` passed taxon sampling criteria (`r min_genes_with_expression` genes) after removing tips without Tau values.
+Of these, `r length(gene_trees_calibrated)` were successfully time calibrated.
+These calibrated pruned trees were used to calculate phylogenetic independent contrasts for Tau.
+They contain `r nodes_contrast %>% filter(D=="Y") %>% nrow()` duplication nodes and `r nodes_contrast %>% filter(D=="N") %>% nrow()` speciation nodes.
 
-Given the ortholog conjecture, we would expect to see larger expression contrasts following duplication events than speciation events.
-Instead,we find no such distribution (Figure XXKMRR).
+Given the ortholog conjecture, we would expect to see larger contrasts following duplication events than speciation events.
+Instead, we find no such distribution (Figure XXKMRR).
 The Wilcox rank test does not reject the null hypothesis that the rate of evolution following duplications is the same as or less than the rate following speciation (p value = `r wilcox_test_result$p.value`).
@@
 
-We next examined the possibility that ascertainment biases were differentially impacting the inference of evolutionary change in differential expression following duplication and speciation events.
+We next examined the possibility that ascertainment biases were differentially impacting the inference of evolutionary following duplication and speciation events.
 We focused on two possible sources of bias - node depth and branch length.
@@
 While the age of speciation nodes is constrained, duplication nodes can be much older and therefore have a wider range of ages (Supplementary Figure XXAscertainment a).
-Biological reasons for this include the fact that duplication events can be older than the root of the species tree, but speciation events cannot.
-There are also technical factors that can lead to an excess of duplication events deeper in the tree, including the impact of assembly errors that lead to the overestimation of deep duplications [@Hahn:2007cj].
-If independent contrast values also tended to to be lower at greater node depth, it could interact with the preponderance of duplications at greater depth to create a pattern of lower contrasts following duplication events.
-When we remove duplication nodes that are older than the oldest speciation node, our results are unchanged and we cannot reject the null hypothesis that the rate of evolution following duplications is the same as or less than the rate following speciation (Supplementary Figure XXAscertainment b).
+This is because many gene duplication events are older than the most recent common ancestor of the species in the study.
+There are also technical factors that can lead to an excess of duplication events deeper in the tree.
+Assembly errors, for example, tend to lead to the overestimation of deep duplications [@Hahn:2007cj].
+If independent contrast values also tended to to be lower at greater node depth, it could interact with the preponderance of duplications at greater depth to create a pattern of lower contrasts associated with duplication events.
+To test for such an effect, we remove duplication nodes that are older than the oldest speciation node.
+Our results are unchanged and this reduced dataset does not reject the null hypothesis that the rate of evolution following duplications is the same as or less than the rate following speciation (Supplementary Figure XXAscertainment b).
 
@@
 The expected variance is principally determined by the lengths of the branches leading from the node to these daughters.
-The same observed difference between daughter nodes will lead to a larger contrast the shorter the branches are to these nodes, as this indicates a greater rate of evolutionary change.
-The expected variance of contrasts for speciation nodes is constrained by the branch lengths on the species tree, but the expected variance of contrasts for duplications can range much wider (Figure XXAscertainment c).
-
-The fact that the expected variance of speciation contrasts is a subset of the range of the expected variance of duplication contrasts could lead to ascertainment biases.
-We therefore also examine only the contrasts that fell within in the range of expected variance seen for speciation contrasts, excluding duplication contrasts that fall outside of this range.
-This reanalysis does not reject the ortholog conjecture either (Figure XXAscertainment d).
+The shorter the total length of the two branches leading to dauter nodes, the larger the contrast for a given observed difference.
+This is because the same difference across a shorter total branch length indicates a greater rate of evolutionary change.
+The expected variance of contrasts for speciation nodes is constrained by the branch lengths on the species tree, but the expected variance of contrasts for duplications has a much wider range (Figure XXAscertainment c).
+This could lead to biases if the lengths of branches that descend from duplication nodes tend to be overestimated.
+We thereforeexamined only the contrasts that fell within in the range of expected variance seen for speciation contrasts, excluding duplication contrasts that fall outside of this range.
+This reanalysis does not reject the ortholog conjecture either (Figure XXAscertainment d), indicating that branch length bias is not responsible for the result.
 
@@
 
-In order to better understand why our phylogenetic analysis seem to support a different conclusion (*i.e.*, no support for the ortholog conjecture) than the published analysis of KMRR [@KryuchkovaMostacci:2016iw] (*i.e.*, strong support for the ortholog conjecture), we first checked to make sure we could reproduce their specific result.
-This is important since we are only looking at at a subset of the data they considered, the Brawand *et al.* 2011 [@Brawand:2011du] dataset.
-We can replicate their result - we find a higher tau correlation between orthologs (R=`r round( ortholog_r, 3) `) then between paralogs (R=`r round( paralog_r, 3)`).
+In order to better understand why our phylogenetic analysis seems to support a different conclusion (*i.e.*, no support for the ortholog conjecture) than the published analysis of KMRR [@KryuchkovaMostacci:2016iw] (*i.e.*, strong support for the ortholog conjecture), we first checked to make sure we could reproduce their result based on pairwise analyses.
+This is important since we are only looking at at a subset of the data they considered, the Brawand *et al.* 2011 [@Brawand:2011du] dataset for gene trees that could be successfully time calibrated.
+We can replicate their result - we find a higher Tau correlation between orthologs (R=`r round( ortholog_r, 3) `) then between paralogs (R=`r round( paralog_r, 3)`).
 
-Why is it that pairwise methods and phylogenetic methods lead to the opposite conclusion?
+Why is it that pairwise methods and phylogenetic methods lead to opposite conclusions?
 The reason is that they describe different things.
 Pairwise comparisons describe contemporary patterns, while phylogenetic methods infer historical processes.
-There need not be a different process of evolution following speciation and duplication for there to be different patterns for orthologs and paralogs.
-That is because there is one other important link between evolutionary process and contemporary patterns - the structure of the tree.
+There need not be a different process of evolution following speciation and duplication for paralogs to be more different than orthologs.
+This could be due to the structure of the gene phylogenies.
+If paralogs tend to be more distantly related to each other than orthologs, then there would be more time for differences to accumulate even if the rate of change is the same between the two.
+This is, in fact, the case for these data.
+While the mean distance between orthologs is `r round( ortholog_distance_mean, 1 )` years, the mean distance between paralogs is `r round( paralog_distance_mean, 1 )` years.
+As described above, this is because the oldest speciation event is be definition the most recent common ancestor of the the species included in the study, but many gene families underwent duplication before this time.
 
-Paralogs could be more different from each other than orthologs if they tend to be more distantly related to each other than orthologs are.
-That would provide more time for differences to accumulate between them, regardless of the history of duplication and speciation.
-This is, in fact, what we find in this dataset.
-While the mean distance between orthologs is `r round( ortholog_distance_mean, 1 )` years, the mean distance between paralogs is `r round( paralog_distance_mean, 1 )` years.
-This is at least in part because, while the maximum distance between orthologs is two times the depth of the oldest speciation event in the tree, many duplications precede this oldest sampled speciation event.
-The distance between paralogs resulting from these deeper duplications can be much larger.
+To test the hypothesis that distantly related paralogs arising from these older duplication events are the reason for the lower correlation of Tau between paralogs than between orthologs, we removed them.
+When we consider only the paralogs that arise from duplication events the same age or younger than the oldest speciation event, the observed pattern is very different.
+The correlation of Tau between these shallower paralog pairs, R=`r round( paralog_subset_r, 3)`, is much stronger than for the correlation between all paralog pairs, R=`r round( paralog_r, 3)`.
+This is much closer to the ortholog correlation, which has R=`r round( ortholog_r, 3) `.
 
-If we consider only the paralogs that arise from the duplication events the same age or younger than the oldest speciation event, the observed pattern is very different.
-The correlation of tau between these shallower paralog pairs, R=`r round( paralog_subset_r, 3)`, is much stronger than for the correlation between all paralog pairs, R=`r round( paralog_r, 3)`.
-This is much closer to the ortholog comparison, which has R=`r round( ortholog_r, 3) `.
-
-The other issue with the pairwise comparisons is that many are summarizing changes along many branches in the phylogeny, not just the branches that follow the divergence event.
-Two paralogs that diverged from a duplication event deep in the tree may have many subsequent duplication and speciation events, and changes along all these branches will impact the final pattern.
-The phylogenetic methods isolate the changes under consideration.
-There may still be missing speciation events, due to extinction and incomplete taxon sampling, and missing duplication events, due to gene loss.
-But these affect both methods.
-The phylogenetic methods avoid diluting the change that occurs along the branches that follow the node in question with changes along all subsequent branches.
-
-The reason that paralogs pairs have tau values that are more different than ortholog pairs is that they tend to be more distantly related, not because there are differences in evolutionary rates of change in tau following duplication and speciation events.
+The reason that paralog pairs have Tau values that are more different than ortholog pairs is that they tend to be more distantly related, not because there are different rates of Tau evolution following duplication and speciation events.
 This means that having information on whether two genes are orthologs or paralogs provides little added information about expression beyond knowing how distantly related the two genes are.
 This has several implications.
-One is that it is an example of the limitations of the terms orthology and paralogy [@Dunn:2016er].
+It is an example of the limitations of the concepts of orthology and paralogy [@Dunn:2016er].
 These terms can have straightforward meaning in small gene trees with simple duplication/speciation histories, but the utility of the terms breaks down on larger more complex gene trees.
@@
 In this sense, orthology and paralogy are statements about the internals of the tree that are distilled into statements at the tips of the tree.
-Much is lost in the process, though, and for most questions it is much more direct to focus on the structure of the tree and the inferred processes of internal nodes.
+Much is lost in the process, though.
 In this case, KMRR [@KryuchkovaMostacci:2016iw] relied on orthology and paralogy annotations that are derived from the infered duplication and speciation history on gene phylogenies [@Herrero:2016jj].
 These are the same pre-computed trees we use in our phylogenetic re-analyses of their study.
+For most questions it is much more direct to focus on the structure of the tree and the inferred processes within the tree, such as which internal nodes are duplication or speciation events and how much change occurs along the branches.
 
@@
 This is because the similarity metric computed in the pairwise comparisons of Levin *et al.* are based on different genes for different species pairs.
-This means that there is not a trait that can be mapped onto the tree, as KMRR's tau can be.
+This means that there is not a trait that can be mapped onto the tree, as KMRR's Tau can be.
 A full phylogenetic reanalysis would of course be possible be using upstream analysis products to re-derive a new expression summary statistic that was suitable for mapping onto a phylogenetic tree.
@@
 Our failure to detect a signature consistent with the ortholog conjecture suggests that it may not be, and that it should not be assumed in systems where it has not been explicitly tested.
+
+The other issue with the pairwise comparisons is that, except at the tips, they summarize changes along many branches in the phylogeny.
+Two paralogs that diverged from a duplication event deep in the tree may have many subsequent duplication and speciation events, and changes along all these branches will impact the final pairwise comparison.
+Phylogenetic methods have the advantage of isolating the changes under consideration (Fig.
+1b).
+The phylogenetic methods avoid diluting the change that occurs along the branches that follow the node in question with changes along all subsequent branches.
+There may still be missing speciation events, due to extinction and incomplete taxon sampling, and missing duplication events, due to gene loss, but these omissions affect both methods.
 

### 2017-01-25 manuscript.rmd
more polishing, added some references

@@
 A rich theoretical and statistical methodology of phylogenetic comparative methods have been developed over the last three decades to address the challenges and opportunities of comparisons across species [@Felsenstein:1985ua; @Grafen:1989um; @Chamberlain:2012vx; @Garamszegi:2014ew; @Pagel:1999gc].
-These challenges include non-independence of obserbations across species -- more closely related species will tend to be more similar to each other than distantly related species.
+These challenges include non-independence of observations across species -- more closely related species will tend to be more similar to each other than distantly related species.
 Phylogenetic comparative methods have largely been applied to the evolution of morphological and ecological traits, but are just as relevant to CFG.
@@
 It is used to relate findings from well-studied genes to related genes that have not been investigated in detail.
-It can be applied to any trait of genes, from gene seqeunce to biochemical properties to expression.
+It can be applied to any trait of genes, from gene sequence to biochemical properties to expression.
 While the ortholog conjecture describes a specific pattern of functional diversity across genes, it is also articulated as a hypothesis about the process of evolution-- that there is greater evolutionary change in gene attributes following a duplication event than a speciation event.
@@
 
-The independent contrast across a node is the amount of change observed between the daughter notes, scaled by the expected variance [Felsenstein:1985ua].
+The independent contrast across a node is the amount of change observed between the daughter notes, scaled by the expected variance [@Felsenstein:1985ua].
 The expected variance is principally determined by the lengths of the branches leading from the node to these daughters.
-The shorter the total length of the two branches leading to dauter nodes, the larger the contrast for a given observed difference.
+The shorter the total length of the two branches leading to daughter nodes, the larger the contrast for a given observed difference.
 This is because the same difference across a shorter total branch length indicates a greater rate of evolutionary change.
@@
 This could lead to biases if the lengths of branches that descend from duplication nodes tend to be overestimated.
-We thereforeexamined only the contrasts that fell within in the range of expected variance seen for speciation contrasts, excluding duplication contrasts that fall outside of this range.
+We therefore examined only the contrasts that fell within in the range of expected variance seen for speciation contrasts, excluding duplication contrasts that fall outside of this range.
 This reanalysis does not reject the ortholog conjecture either (Figure XXAscertainment d), indicating that branch length bias is not responsible for the result.
@@
 These terms can have straightforward meaning in small gene trees with simple duplication/speciation histories, but the utility of the terms breaks down on larger more complex gene trees.
-Orthology and paralogy are annotations on the tips of the tree that are derived from the infered structure of the tree and annotations of duplication and speciation at internal tree nodes.
+Orthology and paralogy are annotations on the tips of the tree that are derived from the inferred structure of the tree and annotations of duplication and speciation at internal tree nodes.
 In this sense, orthology and paralogy are statements about the internals of the tree that are distilled into statements at the tips of the tree.
 Much is lost in the process, though.
-In this case, KMRR [@KryuchkovaMostacci:2016iw] relied on orthology and paralogy annotations that are derived from the infered duplication and speciation history on gene phylogenies [@Herrero:2016jj].
+In this case, KMRR [@KryuchkovaMostacci:2016iw] relied on orthology and paralogy annotations that are derived from the inferred duplication and speciation history on gene phylogenies [@Herrero:2016jj].
 These are the same pre-computed trees we use in our phylogenetic re-analyses of their study.
@@
 
+#### Implications for the ortholog conjecture
+
+Our reanalysis of KMRR using phylogenetic methods suggests that the ortholog conjectures is not a dominant pattern that is central to explaining the evolution of phenotypic diversity in gene families.
+This has important biological implications.
+It suggests that the mechanism of gene divergence (speciation versus duplication) may not have as strong an impact on phenotypic divergence as thought.
+It also has technical implications.
+Rather than focus on whether genes are orthologs or paralogs when attempting to predict function, it may be more effective to simply focus on how closely related or distantly related they are.
+Closely related paralogs, for example, may tend to have more similar phenotypes than more distantly related orthologs.
+
+The widespread invocation of the ortholog conjecture follows from the expectation that it is strong, widely distributed phenomenon.
+Our failure to detect a signature consistent with the ortholog conjecture suggests that it may not be, and that it should not be assumed in systems where it has not been explicitly tested.
+
 ### Levin *et al.* reanalysis
+
+#### Original pairwise analyses of developmental gene expression
 
@@
 As their title ("The mid-developmental transition and the evolution of animal body plans") indicates, they arrived at two major conclusions.
-First, animal development is characterized by a well-defined mid-developmental transition.
+First, animal development is characterized by a well-defined mid-developmental transition that marks the transition from an early phase of gene expression to a late stage of gene expression.
 Second, this transition helps explain the evolution of features observed among distantly related animals.
@@
 Closely related animals have previously been described as having an hourglass model of gene expression, where evolutionary variance in expression is greater early and late in development than at the midpoint of development [@Kalinka:2010dw; @DomazetLoso:2010cn].
-Levin *et al.* conclude that this contrast between distantly and closely related animals provides biological justification for the concept of phyla and may provide a long-sought operational definition of phyla.
-We previously described some concerns with their interpretations of these results [@Hejnol:2016hw].
+Levin *et al.* conclude that this contrast between distantly and closely related animals provides biological justification for the concept of phyla and may provide a definition of phyla.
 
-Rather than support the inverse hourglass as a general pattern among distantly related species, our reanalyses of Levin *et al.* [@Levin:2016bp] indicate their existing data support the opposite conclusion - that the inverse hourglass is due to large differences specific to a single lineage, the ctenophore (comb jelly).
-Better understanding these differences will be of interest to future work, especially in light of current uncertainty regarding the phylogenetic placement of ctenophores and their unique biology [@Martindale:2015ve, @Dunn:2015ij].
-
-To understand our concerns with their analyses, it is helpful to first outline Levin *et al.*'s published methods and results.
 Although there are mature methods and tools for statistical analysis of character evolution [@Garamszegi:2014ew] and gene expression [@Robinson:2009cw], Levin *et al.* [@Levin:2016bp] drew on neither of these very active areas of research and instead applied an *ad hoc* pairwise comparison of orthologous gene expression between species.
-They characterized each gene in each species as having expression that peaks in early, mid, or late temporal phase of development while the goodness of fit to these patterns was not considered, and alternative patterns were not evaluated.
-For each species pair, they then identified the orthologs shared by these species (shared orthologs vary from pair to pair).
+They characterized each gene in each species as having expression that peaks in early, mid, or late temporal phase of development.
+The goodness of fit to these patterns was not considered, and alternative patterns, such as the possibility that there are more phases of expression or no distinct phases of expression, were not evaluated.
+For each species pair, they identified the orthologs shared by these species (shared orthologs vary from pair to pair).
 They then calculated a similarity score for each temporal phase for each species pair based on the fraction of genes that exhibited the same patterns in each species.
@@
 
+We previously described some concerns with their interpretations of these results [@Hejnol:2016hw].
+The main interpretation issues we identified with their interpretation is that because they only sampled one species per phylum, there is no way to tell if observed differences are unique to the branches that gave rise to phyla or occurred along the many other branches in the tree.
+By the stated logic this study could just as easily be a test of unique expression differences between classes, orders, or families, since each of the sampled animals come from a different one of these groups.
+It is far more likely that changes between such distantly related taxa do not correspond to taxonomic ranks in any way.
+Phyla were invented as taxonomic convenience to refer to animal groups of unknown relationship to each other, so they are a statement of historical phylogenetic uncertainty by humans rather than an intrinsic biological phenomenon to be explained.
+
+#### Reexamination of pairwise comparisons
+
+Here we examine the support for the inverse hourglass model itself.
 We began by examining the matrix of pairwise comparisons that their KS tests and Figure 4d are based on.
@@
 Further inspection (Figure XXLevin a) reveals that the mid phase distribution has several outliers with very low similarity scores.
-We found that all five of the lowest values in the mid phase distribution are for pairwise comparisons that include the ctenophore (Figure XXLevin a).
+We found that all five of the lowest values in the mid phase distribution (Figure XXLevin a) are for pairwise comparisons that include the ctenophore (comb jelly).
 When the nine pairwise comparisons that include the ctenophore are removed, the differences between the early and mid phase distributions are greatly reduced (Figure XXLevin b).
@@
 After removing the duplicate scores, removing the ctenophore, and applying the Wilcoxon test, there is no significant difference between the early phase and mid phase distributions (P = 0.1428 for the early-mid comparison and P < 10<sup>-5</sup> for the late-mid comparison), and the inverse hourglass turns into a wine bottle (Figure XXLevin b).
+Rather than support the inverse hourglass as a general pattern among distantly related species, our reanalyses of Levin *et al.* [@Levin:2016bp] comes to a very conclusion - the greater variance at the midpoint of development is specific to a single lineage, the ctenophore.
 
@@
 This particular case illustrates the shortcomings of *ad hoc* rank-based methods [@DomazetLoso:2010cn] and pairwise comparisons [@Levin:2016bp] for the study of the evolution of gene expression.
-Phylogenetic analyses of character evolution provide not only a way to avoid these problems, but a richer context for interpreting the biological implications of the results.
 
@@
 This is because the similarity metric computed in the pairwise comparisons of Levin *et al.* are based on different genes for different species pairs.
-This means that there is not a trait that can be mapped onto the tree, as KMRR's Tau can be.
-A full phylogenetic reanalysis would of course be possible be using upstream analysis products to re-derive a new expression summary statistic that was suitable for mapping onto a phylogenetic tree.
+This means that they do not generate a trait that can be evaluated across a phylogeny, as KMRR's Tau can be.
+A full phylogenetic reanalysis would be possible by using upstream analysis products to re-derive new expression summary statistics suitable for phylogenetic comparative methods.
 
-### Comparative methods in functional genomics
+### Phylogenetic comparative methods in functional genomics
 
-Some of the most widely used phylogenetic comparative methods [1, 5] are directly applicable to CFG data.
-There are also unique challenges that will need to be addressed at this new intersection, though.
-One of the greatest challenges is that most phylogenetic comparative methods have been developed to address problems with many more species (e.g., dozens or more) relative to the number of traits being examined (e.g., 2-5).
-In CFG analyses, there are often far fewer species (e.g., 3-10) because adding taxa is still expensive, but tens of thousands of traits.
-This creates challenges as the resulting covariance matrices are singular and, if not treated appropriately, imply many false correlations that are artifacts of project design.
-We outlined these challenges and potential solutions in the context of gene expression [31].
+Some of the most widely used phylogenetic comparative methods [@Felsenstein:1985ua; @Grafen:1989um] are directly applicable to CFG data.
+There are also unique challenges that will need to be addressed at this new interdisciplinary interface, though.
+One of the greatest challenges is that most phylogenetic comparative methods have been developed to address problems with many more species (e.g., dozens or more) relative to the number of traits being examined.
+In CFG analyses, there are often far fewer species because adding taxa is still expensive, but tens of thousands of traits.
+This creates statistical challenges as the resulting covariance matrices are singular and, if not treated appropriately, imply many false correlations that are artifacts of project design.
+We outlined these challenges and potential solutions in the context of gene expression [@Dunn:2013cpa].
+In addition to methods development at this interface, there is a critical need for software tools that make it simpler to integrate existing functional genomic and phylogenetic comparative packages.
 
-It is not enough to demonstrate the impact of phylogenetic methods on CFG analyses, these methods need to be made accessible to the functional genomics community so they can be widely applied.
+Recent advances in phylogenetic comparative methods are particularly well suited to addressing questions about the evolution of functional genomic traits.
+Most early phylogenetic comparative methods attempted to remove evolutionary signal to correct statistical tests for correlations between traits, while more recent methods tend to focus on testing hypotheses of evolutionary processes [@Pennell:2013cw].
+The application of this newer focus to functional genomics provides an exciting opportunity to address long standing questions of broad interest, including the order of changes in functional genomics traits and shifts in rates of evolution of one functional genomics trait following changes in another trait.
 
-Most early phylogenetic comparative methods attempted to remove evolutionary signal to correct statistical tests for correlations between traits, while more recent methods tend to focus on testing hypotheses of evolutionary processes [34].
-The application of this newer focus to pCFG provides an exciting opportunity to address long standing questions of broad interest, including the order of changes in functional genomics traits and shifts in rates of evolution of one functional genomics trait following changes in another trait.
+We are not the first to apply phylogenetic comparative methods to functional genomic data.
+While the vast majority of comparative functional genomics studies have used pairwise similarity methods, a small number of CFG studies have employed phylogenetic comparative approaches [@Oakley:2005ky; @Eng:2009gp; @Chang:2014hq].
+For instance, a phylogenetic ANOVA [@Rohlfs:2015bd] of the evolution of gene expression improves statistical power and drastically reduces the rate of false positives relative to pairwise approaches.
 
-A small number of CFG studies have employed some phylogenetic comparative approaches [18–22].
-For instance, a phylogenetic ANOVA [21] of the evolution of gene expression improves statistical power and drastically reduces the rate of false positives relative to pairwise approaches.
-
-Our reanalysis of KMRR using phylogenetic methods suggests that the ortholog conjectures is not a dominant pattern that is central to explaining the evolution of phenotypic diversity in gene families.
-This has important biological implications.
-It suggests that the mechanism of gene divergence (speciation versus duplication) may not have as strong an impact on phenotypic divergence as thought.
-It also has technical implications.
-Rather than focus on whether genes are orthologs or paralogs when attempting to predict function, it may be more effective to simply focus on how closely related or distantly related they are.
-Closely related paralogs, for example, may tend to have more similar phenotypes than more distantly related orthologs.
-
-The widespread invocation of the ortholog conjecture follows from the expectation that it is strong, widely distributed phenomenon.
-Our failure to detect a signature consistent with the ortholog conjecture suggests that it may not be, and that it should not be assumed in systems where it has not been explicitly tested.
-
-The other issue with the pairwise comparisons is that, except at the tips, they summarize changes along many branches in the phylogeny.
+Addressing the statistical non-independence of pairwise comparisons is not the only advantage of using phylogenetic comparative methods for functional genomics analyses.
+Another problem with the pairwise comparisons is that, except at the tips, they summarize changes along many branches in the phylogeny.
 Two paralogs that diverged from a duplication event deep in the tree may have many subsequent duplication and speciation events, and changes along all these branches will impact the final pairwise comparison.
@@
 The fact that the first two CFG studies we reanalyzed show serious problems with pairwise comparisons indicates that there likely to be similar problems in other CFG studies, and that future CFG studies will be compromised if they continue to use pairwise methods.
-
 Studies of evolutionary functional genomics should not be focused on the tips of the tree, they should explicitly delve into the tree itself.
@@
 We then calculated the magnitude of the independent contrast for differential expression at each internal node in each of these trees.
-The independent contrast is the amount of change in expression observed along the branches that arise from the node, scaled by the expected variance [Felsenstein:1985ua].
-The expected variance is determined mostly by the length of the branches, but can also be influenced by uncertainty regarding the reconstruction of differential expression at internal nodes.
-We took the absolute value of these contrasts, since the sign depends only on the arbitrary ordering of the nodes in the trees.
 

### 2017-01-25 manuscript.rmd
author name changed
typos

@@
 
-Casey W. Dunn^1^*, Felipe Zapata^1,2^, Cat Munro^1^, Stefan Siebert^1,3^, Andreas Hejnol^4^
+Casey W. Dunn^1^*, Felipe Zapata^1,2^, Catriona Munro^1^, Stefan Siebert^1,3^, Andreas Hejnol^4^
 
@@
 The focus of genomics research has quickly shifted from describing genome sequences to functional genomics, the study of how genomes "work" using tools that measure functional attributes such as expression, chromatin state, and transcription initiation.
-functional genomics, in turn, is now becoming more comparative- there is great interest in understanding how functional genomic variation across species gives rise to a diversity of development, morphology, physiology, and other phenotypes [@Wray:2013gh].
+Functional genomics, in turn, is now becoming more comparative- there is great interest in understanding how functional genomic variation across species gives rise to a diversity of development, morphology, physiology, and other phenotypes [@Wray:2013gh].
 Comparative Functional Genomics (CFG) analyses are also critical to transferring functional insight across species, and will grow in importance in coming years.
@@
 Previous work has shown that ontology annotations are not sufficient to test the ortholog conjecture [@Chen:2012jz; @Thomas:2012hz].
-Analyses of domain structure were consistent with ortholog conjecture [@Forslund:2011cw].
+Analyses of domain structure were consistent with the ortholog conjecture [@Forslund:2011cw].
 There have been few tests of the ortholog conjecture with regards to gene expression [@Chen:2012jz], and KMRR is the most thorough such expression study to date.

### 2017-01-26 manuscript.rmd
code review - added comments, consolidated and documented functions, added to readme

@@
 Here we consider their conclusions in the context of one of these studies, Brawand *et al.* 2011 [@Brawand:2011du].
-This dataset of gene expression for six organs across 10 species (nine mammals and one bird) is the best sampled in their analyses.
+This dataset of gene expression for six organs across ten species (nine mammals and one bird), eight of which were analyzed by KMRR, is the best sampled in their analyses.
 Their expression summary statistic is Tau [@Yanai:2005dh, @KryuchkovaMostacci:2016in], an indicator of tissue specificity of gene expression.
@@
 
-	# Set biology starting parameters, including files
-
+	# Name of the compara gene trees file
 	gene_tree_name = "Compara.75.protein.nhx.emf"
 
-	kmrr_directory = "kmrr/"
-
+	# Files priduced by KMRR Rscript.R with Tau values and other expression statistics
 	expression_file_names = c(
@@
 
+	kmrr_directory = "kmrr/"
 	expression_file_names %<>% paste( kmrr_directory, ., sep="" )
-
 	gene_tree_name %<>% paste( kmrr_directory, ., sep="" )
@@
 
-	# These are from the KMRR code
+	# These clade dates are from the KMRR code, they got them from http://timetree.org/
 	calibration_times = data.frame(
@@
 
-	# Returns a logical vector indicating which rows in nhx_tags are for tips
+	#' Get a boolean vector corresponding to all tips and internal nodes of 
+	#' a tree, with value TRUE for tips and FALSE for internal nodes
+	#' 
+	#' @param nhx A phylogenetic tree as a treeio::treedata object
+	#' @return A boolean vector
+	#' @export
 	is.tip.nhx = function( nhx ) {
@@
 
-	# put node ages in @data
+	#' Record the distance from a tip for each node in a tree in 
+	#' the data slot
+	#' 
+	#' @param nhx A phylogenetic tree as a  treeio::treedata
+	#' @return A treeio::treedata object, with an additional node_age column
+	#' in the data slot, or the original object if it was not of class treedata
+	#' @export
 	store_node_age = function( nhx ) {
@@
 
-	# Make tree ultrametric, calibrating speciation nodes to consistent times
+	#' Adjust branch lengths of a phylogenetic tree to make it ultrametric and 
+	#' to time calibrate the speciation nodes to fixed values
+	#' 
+	#' @param nhx A phylogenetic tree as a treeio::treedata object, with speciation 
+	#' nodes annotated with a value of "N" in @data$D
+	#' @param calibration_times A dataframe with two columns: age.min is the age 
+	#' of a clade, and clade is the name of the clade
+	#' @return A treeio::treedata object if successfully calibrated
+	#' @export
 	calibrate_tree = function ( nhx, calibration_times ) {
@@
 
+	#' Drop tips without expression data from a tree 
+	#' 
+	#' @param nhx A phylogenetic tree as a treeio::treedata object, with a 
+	#' column @data$Tau containing expression data
+	#' @param min_genes_with_expression The minimum number of tips with 
+	#' expression data for a tree to be retained
+	#' @return A treeio::treedata object, or NA if less than 
+	#' min_genes_with_expression tips would have been retained
+	#' @export
+	drop_empty_tips = function( nhx, min_genes_with_expression ) {
+
+		# Identify the tips without expression data
+		to_drop = which( is.na( nhx@data[ 1:length( nhx@phylo$tip.label ), ]$Tau ) )
+
+		# Return the pruned tree if it meets sampling criteria, otherwise return NA
+		remaining = length( nhx@phylo$tip.label ) - length( to_drop )
+		if( remaining >= min_genes_with_expression ){
+			pruned = treeio::drop.tip( nhx, to_drop )
+			return( pruned )
+		}
+		else{
+			return( NA )
+		}
+	}
+
+	#' Calculate phylogenetic independent contrasts (PIC) with the tree and 
+	#' character data in a treeio::treedata object
+	#' 
+	#' @param nhx A phylogenetic tree as a treeio::treedata object, with a 
+	#' column @data$Tau containing expression data
+	#' @return A treeio::treedata object, with new @data columns pic and var_exp
+	#' @export
+	pic.nhx = function( nhx ) {
+
+		# Calculate the contrasts
+		p = pic( nhx@data$Tau[ is.tip.nhx( nhx ) ], nhx@phylo, var.contrasts=TRUE )
+
+		# Add the results back to the @data slot, padding the rows that correspond to 
+		# tips with NA
+		nhx@data$pic = c( rep( NA, length( nhx@phylo$tip.label ) ), p[ ,1 ] )
+		nhx@data$var_exp = c( rep( NA, length( nhx@phylo$tip.label ) ), p[ ,2 ] )
+
+		return( nhx )
+	}
+
 ```
@@
 
+	# Read the expression data for each species and combine into a single tibble
 	expression = 
@@
 ```{r load_trees, echo=FALSE, cache=TRUE, message=F, warning=F}
+
+	# Read in the tree file as a vector of strings, one per line
 	lines = readLines( gene_tree_name )
 
-	# Only consider the lines with trees
+	# Only retain strings that are trees
 	lines = lines[ grepl( "^\\(", lines, perl = TRUE ) ]
 
+	# Parse each string to a treeio::treedata object, adding a label column 
+	# for node names
 	gene_trees = mclapply(
@@
 
+	# Annotate each tree by joining the corresponding expression data to the 
+	# @data object. This maps the expression data to the tree tips
 	gene_trees_annotated = mclapply(
@@
 
-	drop_empty_tips = function( nhx ) {
-		to_drop = which( is.na( nhx@data[ 1:length( nhx@phylo$tip.label ), ]$Tau ) )
-		remaining = length( nhx@phylo$tip.label ) - length( to_drop )
+	# Remove tips from the trees that do not have expression data
+	gene_trees_sub = mclapply( 
+		gene_trees_annotated, 
+		drop_empty_tips, 
+		min_genes_with_expression=min_genes_with_expression,
+		mc.cores=cores 
+	)
 
-		if( remaining >= min_genes_with_expression ){
-			pruned = treeio::drop.tip( nhx, to_drop )
-			return( pruned )
-		}
-		else{
-			return( NA )
-		}
-	}
-
-	gene_trees_sub = mclapply( gene_trees_annotated, drop_empty_tips, mc.cores=cores )
+	# Remove trees with less than min_genes_with_expression tips with expression data
 	gene_trees_sub = gene_trees_sub[ ! is.na( gene_trees_sub ) ]
 
+	# Make the trees ultrametric and calibrate the speciation nodes to 
+	# specified times
 	gene_trees_calibrated = 
@@
 
+	# Remove trees that could not be successfully calibrated
 	gene_trees_calibrated = gene_trees_calibrated[ ! is.na( gene_trees_calibrated ) ]
 
+	# Parse the node ages from the internal @phylo object and store them with the 
+	# corresponding rows in the @data object
 	gene_trees_calibrated = 
@@
 
-	pic.nhx = function( nhx ) {
-
-		p = pic( nhx@data$Tau[ is.tip.nhx( nhx ) ], nhx@phylo, var.contrasts=TRUE )
-
-		nhx@data$pic = c( rep( NA, length( nhx@phylo$tip.label ) ), p[ ,1 ] )
-		nhx@data$var_exp = c( rep( NA, length( nhx@phylo$tip.label ) ), p[ ,2 ] )
-
-		return( nhx )
-	}
-
+	# Calculate independent contrasts for Tau on each tree, storing the results 
+	# back into the @data slot of the tree objects
 	gene_trees_pic = mclapply( gene_trees_calibrated, pic.nhx, mc.cores=cores )
 
+	# Combine the @data slots across all trees. These have labels, node annotations,
+	# expression data, contrast results, etc...
 	all_nodes = 
@@
 				tags = nhx@data
-				tags$gene = digest( nhx )
+				tags$gene = digest( nhx ) # Creates a  hash that is unique to each gene tree
 				tags %<>% select( -( B ) ) # B has inconsistent types, remove it
@@
 
+	# Make a new human readable column for plots that records speciation/ duplication
 	all_nodes$Event = NA
-
 	all_nodes$Event[ all_nodes$D == "Y" ] = "Duplication"
@@
 
+	# Create a subset that corresponds to nodes with contrasts. This excludes tips.
+	# Also take the absolute value of the contrast, since we only consider magnitude.
 	nodes_contrast = 
@@
 
+	# Plot the contrasts for duplication events and speciation events
 	nodes_contrast %>%
@@
 
+	# Check if the mean rank of duplication pics is greater than speciation pics
 	wilcox_test_result = 
@@
 
-We considered only the data from Brawand *et al.* 2011 [@Brawand:2011du].
-KMRR examined multiple subsets of the data.
+We considered only the data from Brawand *et al.* 2011 [@Brawand:2011du] for the eight taxa included in KMRR.
 We left in sex chromosome genes and testes expression data, as this corresponded to the analyses that provided the strongest support for the ortholog conjecture and therefore the most conservative reconsideration of it.
@@
 ```{r ascertainment_plot, cache=TRUE, echo=FALSE, fig.cap="XX_PIC_VAR", warning=FALSE, message=FALSE}
+
+	# Plot results of the ascertainment bias analyses
 

### 2017-01-26 manuscript.rmd
added url for compara gene trees

@@
 
-	# Name of the compara gene trees file
+	# Name of the compara gene trees file. This is from 
+	# ftp://ftp.ensembl.org/pub/release-75/emf/ensembl-compara/homologies/
+	# but not included here due to its size. Put it in the kmrr/ directory
 	gene_tree_name = "Compara.75.protein.nhx.emf"

### 2017-01-26 manuscript.rmd
Adds some initial suggestions to text

--- 2017-01-25 manuscript.rmd
@@
 
-We reanalyzed the the KMRR study using phylogenetic comparative methods, focusing on the dataset of Brawand *et al.* 2011 [@Brawand:2011du].
+We reanalyzed the KMRR study using phylogenetic comparative methods, focusing on the dataset of Brawand *et al.* 2011 [@Brawand:2011du].
 For each internal node in each gene tree (Figure 1b), we measured the phylogenetic independent contrast[@Felsenstein:1985ua] (PIC) of the expression statistic Tau.
@@
 There are also technical factors that can lead to an excess of duplication events deeper in the tree.
-Assembly errors, for example, tend to lead to the overestimation of deep duplications [@Hahn:2007cj].
-If independent contrast values also tended to to be lower at greater node depth, it could interact with the preponderance of duplications at greater depth to create a pattern of lower contrasts associated with duplication events.
+Gene tree estimation errors, for example, tend to lead to the overestimation of deep duplications [@Hahn:2007cj].
+If independent contrast values also tended to be lower at greater node depth, it could interact with the preponderance of duplications at greater depth to create a pattern of lower contrasts associated with duplication events.
 To test for such an effect, we remove duplication nodes that are older than the oldest speciation node.
@@
 This could lead to biases if the lengths of branches that descend from duplication nodes tend to be overestimated.
-We therefore examined only the contrasts that fell within in the range of expected variance seen for speciation contrasts, excluding duplication contrasts that fall outside of this range.
-This reanalysis does not reject the ortholog conjecture either (Figure XXAscertainment d), indicating that branch length bias is not responsible for the result.
+We therefore examined only the contrasts that fell within the range of expected variance seen for speciation contrasts, excluding duplication contrasts that fall outside of this range.
+This reanalysis does not reject the null hypothesis either (Figure XXAscertainment d), indicating that branch length bias is not responsible for rejecting the ortholog conjecture.
 
@@
 This is important since we are only looking at at a subset of the data they considered, the Brawand *et al.* 2011 [@Brawand:2011du] dataset for gene trees that could be successfully time calibrated.
-We can replicate their result - we find a higher Tau correlation between orthologs (R=`r round( ortholog_r, 3) `) then between paralogs (R=`r round( paralog_r, 3)`).
+We can replicate their result - we find a higher Tau correlation between orthologs (R=`r round( ortholog_r, 3) `) than between paralogs (R=`r round( paralog_r, 3)`).
 
@@
 It also has technical implications.
-Rather than focus on whether genes are orthologs or paralogs when attempting to predict function, it may be more effective to simply focus on how closely related or distantly related they are.
+Rather than focus on whether genes are orthologs or paralogs when attempting to predict function, it may be more effective to simply focus on their evolutionary history and how closely related or distantly related they are.
 Closely related paralogs, for example, may tend to have more similar phenotypes than more distantly related orthologs.
@@
 To address this topic, Levin *et al.* [@Levin:2016bp] analyzed gene expression through the course of embryonic development for ten animal species, each from a different phylum.
-As their title ("The mid-developmental transition and the evolution of animal body plans") indicates, they arrived at two major conclusions.
+They arrived at two major conclusions.
 First, animal development is characterized by a well-defined mid-developmental transition that marks the transition from an early phase of gene expression to a late stage of gene expression.
@@
 Closely related animals have previously been described as having an hourglass model of gene expression, where evolutionary variance in expression is greater early and late in development than at the midpoint of development [@Kalinka:2010dw; @DomazetLoso:2010cn].
-Levin *et al.* conclude that this contrast between distantly and closely related animals provides biological justification for the concept of phyla and may provide a definition of phyla.
+Levin *et al.* conclude that the inverse hourglass model provides biological justification for the concept of phyla and may provide an operational definition of phyla.
 
@@
 After removing the duplicate scores, removing the ctenophore, and applying the Wilcoxon test, there is no significant difference between the early phase and mid phase distributions (P = 0.1428 for the early-mid comparison and P < 10<sup>-5</sup> for the late-mid comparison), and the inverse hourglass turns into a wine bottle (Figure XXLevin b).
-Rather than support the inverse hourglass as a general pattern among distantly related species, our reanalyses of Levin *et al.* [@Levin:2016bp] comes to a very conclusion - the greater variance at the midpoint of development is specific to a single lineage, the ctenophore.
+Rather than support the inverse hourglass as a general pattern among distantly related species, our reanalyses of Levin *et al.* [@Levin:2016bp] comes to a different conclusion - the greater variance at the midpoint of development is specific to a single lineage, the ctenophore.
 

### 2017-01-26 manuscript.rmd
reorganized code in chunks and functions to improve clarity and make additional subanalyses easier

--- 2017-01-26 manuscript.rmd
@@
 	)
+```
 
-	# Functions
+```{r define_functions, echo=FALSE, message=F, warning=F}
 
@@
 
-	#' Adjust branch lengths of a phylogenetic tree to make it ultrametric and 
-	#' to time calibrate the speciation nodes to fixed values
-	#' 
-	#' @param nhx A phylogenetic tree as a treeio::treedata object, with speciation 
-	#' nodes annotated with a value of "N" in @data$D
-	#' @param calibration_times A dataframe with two columns: age.min is the age 
-	#' of a clade, and clade is the name of the clade
-	#' @return A treeio::treedata object if successfully calibrated
-	#' @export
-	calibrate_tree = function ( nhx, calibration_times ) {
-
-		# Create calibration matrix for speciation nodes
-		calibration = 
-			nhx@data[ !is.tip.nhx( nhx ), ] %>%
-			filter( D == "N" ) %>%
-			left_join( calibration_times, c( "label" = "clade" ) ) %>%
-			select( node, age.min ) %>%
-			mutate( age.max = age.min ) %>% 
-			mutate( soft.bounds = NA )
-
-		tree = try( 
-			ape::chronos( nhx@phylo, calibration=calibration, model="correlated" ) 
-		)
-
-		if( "phylo" %in% class( tree ) ){
-			class( tree ) = "phylo"
-			nhx@phylo = tree
-			return( nhx )
-		}
-		else{
-			return( NA )
-		}
-	}
-
 	#' Drop tips without expression data from a tree 
@@
 
+	#' Adjust branch lengths of a phylogenetic tree to make it ultrametric and 
+	#' to time calibrate the speciation nodes to fixed values
+	#' 
+	#' @param nhx A phylogenetic tree as a treeio::treedata object, with speciation 
+	#' nodes annotated with a value of "N" in @data$D
+	#' @param calibration_times A dataframe with two columns: age.min is the age 
+	#' of a clade, and clade is the name of the clade
+	#' @param ... Any additional arguments to pass to ape::chronos()
+	#' @return A treeio::treedata object if successfully calibrated
+	#' @export
+	calibrate_tree = function ( nhx, calibration_times, ... ) {
+
+		# Create calibration matrix for speciation nodes
+		calibration = 
+			nhx@data[ !is.tip.nhx( nhx ), ] %>%
+			filter( D == "N" ) %>%
+			left_join( calibration_times, c( "label" = "clade" ) ) %>%
+			select( node, age.min ) %>%
+			mutate( age.max = age.min ) %>% 
+			mutate( soft.bounds = NA )
+
+		tree = try( 
+			ape::chronos( nhx@phylo, calibration=calibration ) 
+		)
+
+		if( "phylo" %in% class( tree ) ){
+			class( tree ) = "phylo"
+			nhx@phylo = tree
+			return( nhx )
+		}
+		else{
+			return( NA )
+		}
+	}
+
 	#' Calculate phylogenetic independent contrasts (PIC) with the tree and 
@@
 		return( nhx )
+	}
+
+	#' Add phylogenetic independent contrasts (PIC) with the tree and 
+	#' character data in a treeio::treedata object
+	#' 
+	#' @param gene_trees_calibrated A list of time calibrated phylogenetic trees 
+	#' as treeio::treedata objects 
+	#' @return A list of time phylogenetic trees with pics 
+	#' as treeio::treedata objects
+	#' @export
+	add_pics_to_trees = function( gene_trees_calibrated ) {
+
+		# Calculate independent contrasts for Tau on each tree, storing the results 
+		# back into the @data slot of the tree objects
+		gene_trees_pic = mclapply( gene_trees_calibrated, pic.nhx, mc.cores=cores )
+
+		return( gene_trees_pic )
+	}
+
+	#' Add phylogenetic independent contrasts (PIC) with the tree and 
+	#' character data in a treeio::treedata object
+	#' 
+	#' @param gene_trees_pic A list of phylogenetic trees with pics
+	#' as treeio::treedata objects 
+	#' @return A tibble with combined data, including phylogenetic independent 
+	#' contrasts, for internal nodes of all trees
+	#' @export
+	summarize_contrasts = function( gene_trees_pic ) {
+		# Combine the @data slots across all trees. These have labels, node annotations,
+		# expression data, contrast results, etc...
+		all_nodes = 
+			lapply( 
+				gene_trees_pic, 
+				function( nhx ){
+					tags = nhx@data
+					tags$gene = digest( nhx ) # Creates a  hash that is unique to each gene tree
+					tags %<>% select( -( B ) ) # B has inconsistent types, remove it
+					return( tags )
+				}
+			) %>%
+			bind_rows()
+
+		# Make a new human readable column for plots that records speciation/ duplication
+		all_nodes$Event = NA
+		all_nodes$Event[ all_nodes$D == "Y" ] = "Duplication"
+		all_nodes$Event[ all_nodes$D == "N" ] = "Speciation"
+
+		# Create a subset that corresponds to nodes with contrasts. This excludes tips.
+		# Also take the absolute value of the contrast, since we only consider magnitude.
+		nodes_contrast = 
+			all_nodes %>%
+			filter( ! is.na( pic ) ) %>%
+			mutate( pic = abs( pic ) ) %>%
+			filter( ! is.na( D ) )
+
+		return( nodes_contrast )
+	}
+
+	#' Calibrate a list of gene trees given a set of speciation node dates
+	#' 
+	#' @param gene_trees_sub A list of phylogenetic trees as treeio::treedata objects 
+	#' with associated character data
+	#' @param calibration_times A dataframe with two columns: age.min is the age 
+	#' of a clade, and clade is the name of the clade
+	#' @param ... Any additional arguments to pass to ape::chronos()
+	#' @return A list of calibrated phylogenetic trees as treeio::treedata objects. May be 
+	#' shorter than gene_trees_sub if some calibrations fail.
+	#' @export
+	calibrate_trees = function( gene_trees_sub, calibration_times, ... ) {
+		# Make the trees ultrametric and calibrate the speciation nodes to 
+		# specified times
+		gene_trees_calibrated = 
+			mclapply( 
+				gene_trees_sub,
+				calibrate_tree, 
+				calibration_times=calibration_times,
+				...,
+				mc.cores=cores
+			)
+
+		# Remove trees that could not be successfully calibrated
+		gene_trees_calibrated = gene_trees_calibrated[ ! is.na( gene_trees_calibrated ) ]
+
+		# Parse the calibrated node ages from the internal @phylo object and store them with the 
+		# corresponding rows in the @data object
+		gene_trees_calibrated = 
+			mclapply( 
+				gene_trees_calibrated, 
+				store_node_age, 
+				mc.cores=cores 
+			)
+
+		return( gene_trees_calibrated )
+	}
+
+	#' Wilcoxon test of the ortholog conjecture, specifically that contrasts
+	#' associated with duplication nodes are greater than those for 
+	#' speciation nodes. A significant result indicates rejection of the null
+	#' hypothesis that they are not greater.
+	#' 
+	#' @param nodes_contrast A tibble of independent contrasts
+	#' @return Wilcoxon test p value
+	#' @export
+	wilcox_oc = function( nodes_contrast ) {
+
+		p = wilcox.test( 
+			nodes_contrast %>% filter( D=="Y" ) %>% .$pic, 
+			nodes_contrast %>% filter( D=="N" ) %>% .$pic, 
+			alternative="greater" 
+		)$p.value
+
+		return( p )
 	}
@@
 
-```
-
-```{r prep_trees, echo=FALSE, cache=TRUE, message=F, warning=F}
-
 	# Remove tips from the trees that do not have expression data
@@
 
-	# Make the trees ultrametric and calibrate the speciation nodes to 
-	# specified times
-	gene_trees_calibrated = 
-		mclapply( 
-			gene_trees_sub,
-			calibrate_tree, 
-			calibration_times=calibration_times,
-			mc.cores=cores
-		)
+```
 
-	# Remove trees that could not be successfully calibrated
-	gene_trees_calibrated = gene_trees_calibrated[ ! is.na( gene_trees_calibrated ) ]
+```{r calibrate_trees, echo=FALSE, cache=TRUE, message=F, warning=F}
 
-	# Parse the node ages from the internal @phylo object and store them with the 
-	# corresponding rows in the @data object
-	gene_trees_calibrated = 
-		mclapply( 
-			gene_trees_calibrated, 
-			store_node_age, 
-			mc.cores=cores 
-		)
+	# Time calibrate all speciation nodes to the same years
+	gene_trees_calibrated = calibrate_trees( gene_trees_sub, calibration_times, model="correlated" )
 
@@
 
-```{r contrasts, cache=TRUE, echo=FALSE, message=F, warning=F}
+```{r trees_to_contrasts, cache=TRUE, echo=FALSE, message=F, warning=F}
 
-	# Calculate independent contrasts for Tau on each tree, storing the results 
-	# back into the @data slot of the tree objects
-	gene_trees_pic = mclapply( gene_trees_calibrated, pic.nhx, mc.cores=cores )
+	# Calculate the contrasts from the trees and associated expression data
+	gene_trees_pic = add_pics_to_trees( gene_trees_calibrated )
 
-	# Combine the @data slots across all trees. These have labels, node annotations,
-	# expression data, contrast results, etc...
-	all_nodes = 
-		lapply( 
-			gene_trees_pic, 
-			function( nhx ){
-				tags = nhx@data
-				tags$gene = digest( nhx ) # Creates a  hash that is unique to each gene tree
-				tags %<>% select( -( B ) ) # B has inconsistent types, remove it
-				return( tags )
-			}
-		) %>%
-		bind_rows()
+	# Collect all the contrasts in a single tibble
+	nodes_contrast = summarize_contrasts( gene_trees_pic )
 
-	# Make a new human readable column for plots that records speciation/ duplication
-	all_nodes$Event = NA
-	all_nodes$Event[ all_nodes$D == "Y" ] = "Duplication"
-	all_nodes$Event[ all_nodes$D == "N" ] = "Speciation"
+	# Test if the mean rank of duplication pics is greater than speciation pics
+	wilcox_test_result = wilcox_oc( nodes_contrast )
 
@@
 
-```{r show_result, echo=FALSE, cache=TRUE, message=F, warning=F}
-
-	# Create a subset that corresponds to nodes with contrasts. This excludes tips.
-	# Also take the absolute value of the contrast, since we only consider magnitude.
-	nodes_contrast = 
-		all_nodes %>%
-		filter( ! is.na( pic ) ) %>%
-		mutate( pic = abs( pic ) ) %>%
-		filter( ! is.na( D ) )
+```{r show_contrast_result, echo=FALSE, cache=TRUE, message=F, warning=F}
 
@@
 		theme_classic()
-
-	# Check if the mean rank of duplication pics is greater than speciation pics
-	wilcox_test_result = 
-		wilcox.test( 
-			nodes_contrast %>% filter( D=="Y" ) %>% .$pic, 
-			nodes_contrast %>% filter( D=="N" ) %>% .$pic, 
-			alternative="greater" 
-		)
 
@@
 Instead, we find no such distribution (Figure XXKMRR).
-The Wilcox rank test does not reject the null hypothesis that the rate of evolution following duplications is the same as or less than the rate following speciation (p value = `r wilcox_test_result$p.value`).
+The Wilcox rank test does not reject the null hypothesis that the rate of evolution following duplications is the same as or less than the rate following speciation (p value = `r wilcox_test_result`).
 Our results therefore find no evidence for the ortholog conjecture in this system.
@@
 
-	wilcox_filtered_age = 
-		wilcox.test( 
-			nodes_contrast_filtered_age %>% filter( D=="Y" ) %>% .$pic, 
-			nodes_contrast_filtered_age %>% filter( D=="N" ) %>% .$pic, 
-			alternative="greater" 
-		)$p.value
+	wilcox_filtered_age = wilcox_oc( nodes_contrast_filtered_age )
 
@@
 
-	wilcox_filtered_var = 
-	wilcox.test( 
-		nodes_contrast_filtered_var %>% filter( D=="Y" ) %>% .$pic, 
-		nodes_contrast_filtered_var %>% filter( D=="N" ) %>% .$pic, 
-		alternative="greater" 
-	)$p.value
+	wilcox_filtered_var = wilcox_oc( nodes_contrast_filtered_var )
 

### 2017-01-26 manuscript.rmd
changed title, removed unused calibrations, moved around and deleted some text

@@
 
-# The importance of phylogenetic methods when comparing functional genomic data across species
+# Multiple pairwise comparisons between species strongly mislead functional genomics studies
 
@@
 There is now considerable interest in comparing genome function across species to understand what is conserved, what is variable, and how the diversity of organism traits (including anatomy, physiology, and development) relates to genome diversity.
-Most published comparisons of genome function across species have, however, neglected to use methods that consider the evolutionary relationships among species.
+Most published comparisons of genome function across species have relied on multiple pairwise comparisons between species, an approach that neglects to account for the evolutionary relationships among species.
 This has two potential negative impacts.
@@
 		age.min =
-			c( 20, 92, 167, 8, 42, 74, 296, 535, 104, 937, 29, 722, 441, 65, 15, 1215, 
-				414, 371, 162, 25, 74, 77, 86 ), 
+			c( 92, 167, 8, 296, 29, 162 ), 
 		clade = 
-			c("Hominoidea", "Euarchontoglires", "Mammalia", "Homininae", 
-					"Simiiformes", "Primates", "Amniota", "Vertebrata", "Eutheria", 
-					"Bilateria", "Catarrhini", "Chordata", "Euteleostomi", 
-					"Haplorrhini", "Hominidae", "Opisthokonta", "Sarcopterygii", 
-					"Tetrapoda", "Theria", "Murinae", "Sciurognathi", "Rodentia", "Glires"
+			c( "Euarchontoglires", "Mammalia", "Homininae", 
+					"Amniota", "Catarrhini", "Theria"
 			),
@@
 
-We next examined the possibility that ascertainment biases were differentially impacting the inference of evolutionary following duplication and speciation events.
+We next examined the possibility that ascertainment biases were differentially impacting the inference of evolution following duplication and speciation events.
 We focused on two possible sources of bias - node depth and branch length.
-
-While the age of speciation nodes is constrained, duplication nodes can be much older and therefore have a wider range of ages (Supplementary Figure XXAscertainment a).
-This is because many gene duplication events are older than the most recent common ancestor of the species in the study.
-There are also technical factors that can lead to an excess of duplication events deeper in the tree.
-Assembly errors, for example, tend to lead to the overestimation of deep duplications [@Hahn:2007cj].
-If independent contrast values also tended to to be lower at greater node depth, it could interact with the preponderance of duplications at greater depth to create a pattern of lower contrasts associated with duplication events.
-To test for such an effect, we remove duplication nodes that are older than the oldest speciation node.
-Our results are unchanged and this reduced dataset does not reject the null hypothesis that the rate of evolution following duplications is the same as or less than the rate following speciation (Supplementary Figure XXAscertainment b).
-
-The independent contrast across a node is the amount of change observed between the daughter notes, scaled by the expected variance [@Felsenstein:1985ua].
-The expected variance is principally determined by the lengths of the branches leading from the node to these daughters.
-The shorter the total length of the two branches leading to daughter nodes, the larger the contrast for a given observed difference.
-This is because the same difference across a shorter total branch length indicates a greater rate of evolutionary change.
-The expected variance of contrasts for speciation nodes is constrained by the branch lengths on the species tree, but the expected variance of contrasts for duplications has a much wider range (Figure XXAscertainment c).
-This could lead to biases if the lengths of branches that descend from duplication nodes tend to be overestimated.
-We therefore examined only the contrasts that fell within in the range of expected variance seen for speciation contrasts, excluding duplication contrasts that fall outside of this range.
-This reanalysis does not reject the ortholog conjecture either (Figure XXAscertainment d), indicating that branch length bias is not responsible for the result.
+We found no evidence that either affacted our results (Supplementary materials).
 
@@
 
-Although there are mature methods and tools for statistical analysis of character evolution [@Garamszegi:2014ew] and gene expression [@Robinson:2009cw], Levin *et al.* [@Levin:2016bp] drew on neither of these very active areas of research and instead applied an *ad hoc* pairwise comparison of orthologous gene expression between species.
+Although there are mature methods and tools for statistical analysis of character evolution [@Garamszegi:2014ew] and gene expression [@Robinson:2009cw], Levin *et al.* [@Levin:2016bp] drew on neither of these active areas of research and instead applied an *ad hoc* pairwise comparison of ortholog gene expression between species.
 They characterized each gene in each species as having expression that peaks in early, mid, or late temporal phase of development.
 The goodness of fit to these patterns was not considered, and alternative patterns, such as the possibility that there are more phases of expression or no distinct phases of expression, were not evaluated.
-For each species pair, they identified the orthologs shared by these species (shared orthologs vary from pair to pair).
+For each species pair, they identified the orthologs shared by these species.
+This list of genes was different from species pair to species pair).
 They then calculated a similarity score for each temporal phase for each species pair based on the fraction of genes that exhibited the same patterns in each species.
@@
 
-We previously described some concerns with their interpretations of these results [@Hejnol:2016hw].
-The main interpretation issues we identified with their interpretation is that because they only sampled one species per phylum, there is no way to tell if observed differences are unique to the branches that gave rise to phyla or occurred along the many other branches in the tree.
-By the stated logic this study could just as easily be a test of unique expression differences between classes, orders, or families, since each of the sampled animals come from a different one of these groups.
-It is far more likely that changes between such distantly related taxa do not correspond to taxonomic ranks in any way.
-Phyla were invented as taxonomic convenience to refer to animal groups of unknown relationship to each other, so they are a statement of historical phylogenetic uncertainty by humans rather than an intrinsic biological phenomenon to be explained.
-
 #### Reexamination of pairwise comparisons
 
+We previously described some concerns with their interpretations of these results [@Hejnol:2016hw].
+The main issue is that because they only sampled one species per phylum, there is no way to tell if observed differences are unique to the branches that gave rise to clades designated as phyla or occurred along the many other branches in the tree.
 Here we examine the support for the inverse hourglass model itself.
-We began by examining the matrix of pairwise comparisons that their KS tests and Figure 4d are based on.
+We focus on the matrix of pairwise comparisons that their KS tests and their Figure 4d are based on.
 As this figure shows (regenerated here as our Figure XXLevin a), there is not a strong distinction between these temporal phases - the distributions for the early, mid, and late distributions overlap considerably.
@@
 Further inspection (Figure XXLevin a) reveals that the mid phase distribution has several outliers with very low similarity scores.
-We found that all five of the lowest values in the mid phase distribution (Figure XXLevin a) are for pairwise comparisons that include the ctenophore (comb jelly).
-When the nine pairwise comparisons that include the ctenophore are removed, the differences between the early and mid phase distributions are greatly reduced (Figure XXLevin b).
 
-We also found several problems with the statistical tests that were used to evaluate the inverse hourglass hypothesis.
+The first problems we identified were general stastistical issues that are not problems with pairwise comparisons *per se*.
 First, in the published analyses every data point was included twice because both reciprocal comparisons (which have the same values) were retained.
 For example, there is both a nematode to arthropod comparison and an arthropod to nematode comparison.
-As a consequence, there are 90 entries for the 45 pairwise comparisons, and by doubling the data the significance of the result appears much stronger than it actually is.
+As a consequence, there are 90 entries for the 45 pairwise comparisons, and by doubling the data the significance of the result appears stronger than it actually is.
 After removing the duplicate values, the p values are far less significant, 0.002 for the early-mid comparison and on the order of 10<sup>-6</sup> for early-late.
@@
 The Wilcoxon test is instead appropriate in this case.
-After removing the duplicate scores, removing the ctenophore, and applying the Wilcoxon test, there is no significant difference between the early phase and mid phase distributions (P = 0.1428 for the early-mid comparison and P < 10<sup>-5</sup> for the late-mid comparison), and the inverse hourglass turns into a wine bottle (Figure XXLevin b).
-Rather than support the inverse hourglass as a general pattern among distantly related species, our reanalyses of Levin *et al.* [@Levin:2016bp] comes to a very conclusion - the greater variance at the midpoint of development is specific to a single lineage, the ctenophore.
 
-Our results highlight the importance of explicitly accounting for phylogenetic relationships when studying character evolution, including developmental [@Telford:2003tsa] and functional genomic traits [@Hejnol:2015gr].
-This is particularly true for evolutionary analyses of quantitative gene expression [@Dunn:2013cpa].
-Pairwise comparisons result in the same evolutionary changes being counted multiple times in each species pair, as for the ctenophore comparisons that were mistaken for a global pattern here.
-This is a well understood property of pairwise comparisons that has been specifically addressed by phylogenetic comparative methods [@Felsenstein:1985ua].
-This particular case illustrates the shortcomings of *ad hoc* rank-based methods [@DomazetLoso:2010cn] and pairwise comparisons [@Levin:2016bp] for the study of the evolution of gene expression.
+The next issue we found is a general problem with pairwise comparisons.
+Inspecting all five of the lowest values in the mid phase distribution (Figure XXLevin a) are for pairwise comparisons that include the ctenophore (comb jelly).
+When the nine pairwise comparisons that include the ctenophore are removed and the general statistical issues noted above are addressed, there is no significant difference between the early phase and mid phase distributions (P = 0.1428 for the early-mid comparison and P < 10<sup>-5</sup> for the late-mid comparison).
+The inverse hourglass turns into a bottle (Figure XXLevin b).
 
-While we do demonstrate a common problem with pairwise comparisons in the Levin *et al.* analysis, we did not perform a phylogenetic reanalysis of this study, as we did for the KMRR study.
+This highlights a well understood property of pairwise comparisons across species [@Felsenstein:1985ua]: evolutionary changes along a given branch, like those along the ctenophore branch, impact each of the multiple pairwise comparisons that includes that branch.
+The pairwise comparisons are therefore not independent - one evolutionary change impacts multiple comparisons.
+This can give the impression of a general pattern across the tree that is instead specific to changes along one part of the tree.
+The number of comparisons impacted by each change depends on the structure of the phylogenetic tree, *i.e.* how the species are related to each other.
+
+Phylogenetic comparative methods were developed specifically to address this problem [@Felsenstein:1985ua].
+While we do demonstrate it impacts the Levin *et al.* analysis, we did not perform a phylogenetic reanalysis of this study, as we did for the KMRR study.
 This is because the similarity metric computed in the pairwise comparisons of Levin *et al.* are based on different genes for different species pairs.
@@
 
-Some of the most widely used phylogenetic comparative methods [@Felsenstein:1985ua; @Grafen:1989um] are directly applicable to CFG data.
-There are also unique challenges that will need to be addressed at this new interdisciplinary interface, though.
-One of the greatest challenges is that most phylogenetic comparative methods have been developed to address problems with many more species (e.g., dozens or more) relative to the number of traits being examined.
+Our results highlight the importance of explicitly accounting for phylogenetic relationships when studying character evolution, including developmental [@Telford:2003tsa] and functional genomic traits [@Hejnol:2015gr].
+This is particularly true for evolutionary analyses of quantitative gene expression [@Dunn:2013cpa].
+
+Some of the most widely used phylogenetic comparative methods [@Felsenstein:1985ua; @Grafen:1989um] are already directly applicable to CFG data.
+There are interesting new challenges at this interdisciplinary interface that will need to be addressed to fully realize the potential of phylogenetic comparative functional genomics stidues.
+One such challenge is that most phylogenetic comparative analyses of covariance between traits have been developed to address problems with many more species (e.g., dozens or more) relative to the number of traits being examined.
 In CFG analyses, there are often far fewer species because adding taxa is still expensive, but tens of thousands of traits.
@@
 We outlined these challenges and potential solutions in the context of gene expression [@Dunn:2013cpa].
-In addition to methods development at this interface, there is a critical need for software tools that make it simpler to integrate existing functional genomic and phylogenetic comparative packages.
+In addition to methods development, there is a critical need for software tools that make it simpler to integrate existing functional genomic and phylogenetic comparative packages.
 
@@
 ### KMRR Analyses
+
+#### Investigation of potential ascertainment bias
+
+While the age of speciation nodes is constrained, duplication nodes can be much older and therefore have a wider range of ages (Supplementary Figure XXAscertainment a).
+This is because many gene duplication events are older than the most recent common ancestor of the species in the study.
+There are also technical factors that can lead to an excess of duplication events deeper in the tree.
+Assembly errors, for example, tend to lead to the overestimation of deep duplications [@Hahn:2007cj].
+If independent contrast values also tended to to be lower at greater node depth, it could interact with the preponderance of duplications at greater depth to create a pattern of lower contrasts associated with duplication events.
+To test for such an effect, we remove duplication nodes that are older than the oldest speciation node.
+Our results are unchanged and this reduced dataset does not reject the null hypothesis that the rate of evolution following duplications is the same as or less than the rate following speciation (Supplementary Figure XXAscertainment b).
+
+The independent contrast across a node is the amount of change observed between the daughter notes, scaled by the expected variance [@Felsenstein:1985ua].
+The expected variance is principally determined by the lengths of the branches leading from the node to these daughters.
+The shorter the total length of the two branches leading to daughter nodes, the larger the contrast for a given observed difference.
+This is because the same difference across a shorter total branch length indicates a greater rate of evolutionary change.
+The expected variance of contrasts for speciation nodes is constrained by the branch lengths on the species tree, but the expected variance of contrasts for duplications has a much wider range (Figure XXAscertainment c).
+This could lead to biases if the lengths of branches that descend from duplication nodes tend to be overestimated.
+We therefore examined only the contrasts that fell within in the range of expected variance seen for speciation contrasts, excluding duplication contrasts that fall outside of this range.
+This reanalysis does not reject the ortholog conjecture either (Figure XXAscertainment d), indicating that branch length bias is not responsible for the result.
 
@@
 
-```{r pairwise_plot, echo=FALSE, cache=TRUE, message=F, warning=F}	
-	pairwise_summary %>% 
-	filter( ! is.na( D ) ) %>%
-	ggplot( aes( x=distance, y=difference, color=D ) ) + 
-		geom_point( alpha=0.1 ) +
-		geom_smooth( method=lm, se=FALSE ) +
-		xlim( 0, 2000 )
-
-```
-
-> Supplementary Figure XXPairwise | Pairwise comparisons
-
 ## References

### 2017-01-26 manuscript.rmd
small tweaks to text

@@
 
-Although there are mature methods and tools for statistical analysis of character evolution [@Garamszegi:2014ew] and gene expression [@Robinson:2009cw], Levin *et al.* [@Levin:2016bp] drew on neither of these active areas of research and instead applied an *ad hoc* pairwise comparison of ortholog gene expression between species.
+Levin *et al.* [@Levin:2016bp] arrived at this conclusion my making pairwise comparisons of ortholog expression through development between species.
 They characterized each gene in each species as having expression that peaks in early, mid, or late temporal phase of development.
-The goodness of fit to these patterns was not considered, and alternative patterns, such as the possibility that there are more phases of expression or no distinct phases of expression, were not evaluated.
+The goodness of fit to these patterns was not considered, and alternative patterns, such as the possibility that there are more phases of expression, were not evaluated.
 For each species pair, they identified the orthologs shared by these species.
-This list of genes was different from species pair to species pair).
+This list of shared genes was different from species pair to species pair.
 They then calculated a similarity score for each temporal phase for each species pair based on the fraction of genes that exhibited the same patterns in each species.
@@
 We previously described some concerns with their interpretations of these results [@Hejnol:2016hw].
-The main issue is that because they only sampled one species per phylum, there is no way to tell if observed differences are unique to the branches that gave rise to clades designated as phyla or occurred along the many other branches in the tree.
+The main concern was that because they only sampled one species per phylum, there is no way to tell if observed differences are unique to the branches that gave rise to clades designated as phyla or occurred along the many other branches in the tree.
 Here we examine the support for the inverse hourglass model itself.
 We focus on the matrix of pairwise comparisons that their KS tests and their Figure 4d are based on.
-As this figure shows (regenerated here as our Figure XXLevin a), there is not a strong distinction between these temporal phases - the distributions for the early, mid, and late distributions overlap considerably.
-The medians are close in value, and the range of mid phase includes the entire range of the early phase.
-Further inspection (Figure XXLevin a) reveals that the mid phase distribution has several outliers with very low similarity scores.
 
@@
 The Wilcoxon test is instead appropriate in this case.
+When applied to the de-duplicated data, the significance of this test is 0.02 for the early-mid comparison and on the order of 10<sup>-7</sup> for early-late.
 
 The next issue we found is a general problem with pairwise comparisons.
-Inspecting all five of the lowest values in the mid phase distribution (Figure XXLevin a) are for pairwise comparisons that include the ctenophore (comb jelly).
+We found that all five of the lowest values in the mid phase distribution (Figure XXLevin a) are for pairwise comparisons that include the ctenophore (comb jelly).
 When the nine pairwise comparisons that include the ctenophore are removed and the general statistical issues noted above are addressed, there is no significant difference between the early phase and mid phase distributions (P = 0.1428 for the early-mid comparison and P < 10<sup>-5</sup> for the late-mid comparison).
@@
 This highlights a well understood property of pairwise comparisons across species [@Felsenstein:1985ua]: evolutionary changes along a given branch, like those along the ctenophore branch, impact each of the multiple pairwise comparisons that includes that branch.
-The pairwise comparisons are therefore not independent - one evolutionary change impacts multiple comparisons.
+The pairwise comparisons are therefore not independent - different pairwise comparisons are impacted by canges along some of the same branches.
 This can give the impression of a general pattern across the tree that is instead specific to changes along one part of the tree.
@@
 Phylogenetic comparative methods were developed specifically to address this problem [@Felsenstein:1985ua].
-While we do demonstrate it impacts the Levin *et al.* analysis, we did not perform a phylogenetic reanalysis of this study, as we did for the KMRR study.
+While we do demonstrate this problem impacts the Levin *et al.* analysis, we did not perform a phylogenetic reanalysis of this study, as we did for the KMRR study.
 This is because the similarity metric computed in the pairwise comparisons of Levin *et al.* are based on different genes for different species pairs.

### 2017-01-26 manuscript.rmd
typos

@@
 While the mean distance between orthologs is `r round( ortholog_distance_mean, 1 )` years, the mean distance between paralogs is `r round( paralog_distance_mean, 1 )` years.
-As described above, this is because the oldest speciation event is be definition the most recent common ancestor of the the species included in the study, but many gene families underwent duplication before this time.
+As described above, this is because the oldest speciation event is by definition the most recent common ancestor of the the species included in the study, but many gene families underwent duplication before this time.
 
@@
 
-Levin *et al.* [@Levin:2016bp] arrived at this conclusion my making pairwise comparisons of ortholog expression through development between species.
+Levin *et al.* [@Levin:2016bp] arrived at this conclusion by making pairwise comparisons of ortholog expression through development between species.
 They characterized each gene in each species as having expression that peaks in early, mid, or late temporal phase of development.

### 2017-01-26 manuscript.rmd
typo

@@
 Some of the most widely used phylogenetic comparative methods [@Felsenstein:1985ua; @Grafen:1989um] are already directly applicable to CFG data.
-There are interesting new challenges at this interdisciplinary interface that will need to be addressed to fully realize the potential of phylogenetic comparative functional genomics stidues.
+There are interesting new challenges at this interdisciplinary interface that will need to be addressed to fully realize the potential of phylogenetic comparative functional genomics studies.
 One such challenge is that most phylogenetic comparative analyses of covariance between traits have been developed to address problems with many more species (e.g., dozens or more) relative to the number of traits being examined.

### 2017-01-26 manuscript.rmd
adds only 2 minor fixes, the other ones don't matter for now

@@
 There are also technical factors that can lead to an excess of duplication events deeper in the tree.
-Assembly errors, for example, tend to lead to the overestimation of deep duplications [@Hahn:2007cj].
+Gene tree estimation errors, for example, tend to lead to the overestimation of deep duplications [@Hahn:2007cj].
 If independent contrast values also tended to to be lower at greater node depth, it could interact with the preponderance of duplications at greater depth to create a pattern of lower contrasts associated with duplication events.
@@
 We therefore examined only the contrasts that fell within in the range of expected variance seen for speciation contrasts, excluding duplication contrasts that fall outside of this range.
-This reanalysis does not reject the ortholog conjecture either (Figure XXAscertainment d), indicating that branch length bias is not responsible for the result.
+This reanalysis does not reject the null hypothesis either (Figure XXAscertainment d), indicating that branch length bias is not responsible for rejecting the ortholog conjecture.
 

### 2017-01-26 manuscript.rmd
added function for adding noise to calibration dates

--- 2017-01-26 manuscript.rmd
@@
 		age.min =
-			c( 92, 167, 8, 296, 29, 162 ), 
+			c(  8, 29, 92, 162, 167, 296 ), 
 		clade = 
-			c( "Euarchontoglires", "Mammalia", "Homininae", 
-					"Amniota", "Catarrhini", "Theria"
-			),
+			c( "Homininae", "Catarrhini", "Euarchontoglires", "Theria", "Mammalia", "Amniota" ),
 		stringsAsFactors=FALSE
@@
 ```{r define_functions, echo=FALSE, message=F, warning=F}
+
+	#' Adds some random noise from a normal distribution to calibration
+	#' times
+	#' 
+	#' @param calibration_times A dataframe with two columns: age.min is the age 
+	#' of a clade, and clade is the name of the clade
+	#' @param sd_fraction this is multiplied by each value to set sd of that value 
+	#' @return A dataframe as above, but with noise
+	#' @export
+	add_noise_calibration_times = function( calibration_times, sd_fraction ) {
+		age = calibration_times$age.min
+		new_age = NA
+
+		while( NA %in% new_age ) {
+			x = sapply( 
+				age,
+				function( i ){
+					rnorm( 1, mean = i, sd = ( sd_fraction * i ) )
+				}
+			)
+
+			if( ! is.unsorted( x ) ){
+				new_age = x
+			}
+
+		}
+
+		calibration_times$age.min = new_age
+		return( calibration_times )
+	}
 
@@
 		tree = try( 
-			ape::chronos( nhx@phylo, calibration=calibration ) 
+			ape::chronos( nhx@phylo, calibration=calibration, ... ) 
 		)
@@
 	#' 
-	#' @param gene_trees_sub A list of phylogenetic trees as treeio::treedata objects 
+	#' @param gene_trees_pruned A list of phylogenetic trees as treeio::treedata objects 
 	#' with associated character data
@@
 	#' @return A list of calibrated phylogenetic trees as treeio::treedata objects. May be 
-	#' shorter than gene_trees_sub if some calibrations fail.
+	#' shorter than gene_trees_pruned if some calibrations fail.
 	#' @export
-	calibrate_trees = function( gene_trees_sub, calibration_times, ... ) {
+	calibrate_trees = function( gene_trees_pruned, calibration_times, ... ) {
 		# Make the trees ultrametric and calibrate the speciation nodes to 
@@
 			mclapply( 
-				gene_trees_sub,
+				gene_trees_pruned,
 				calibrate_tree, 
@@
 	# Remove tips from the trees that do not have expression data
-	gene_trees_sub = mclapply( 
+	gene_trees_pruned = mclapply( 
 		gene_trees_annotated, 
@@
 	# Remove trees with less than min_genes_with_expression tips with expression data
-	gene_trees_sub = gene_trees_sub[ ! is.na( gene_trees_sub ) ]
+	gene_trees_pruned = gene_trees_pruned[ ! is.na( gene_trees_pruned ) ]
 
@@
 	# Time calibrate all speciation nodes to the same years
-	gene_trees_calibrated = calibrate_trees( gene_trees_sub, calibration_times, model="correlated" )
+	gene_trees_calibrated = calibrate_trees( gene_trees_pruned, calibration_times, model="correlated" )
 
@@
 XXKMRR).
-Of the `r length(gene_trees)` trees that were parsed from ENSMBL Compara [@Herrero:2016jj], `r length(gene_trees_sub)` passed taxon sampling criteria (`r min_genes_with_expression` genes) after removing tips without Tau values.
+Of the `r length(gene_trees)` trees that were parsed from ENSMBL Compara [@Herrero:2016jj], `r length(gene_trees_pruned)` passed taxon sampling criteria (`r min_genes_with_expression` genes) after removing tips without Tau values.
 Of these, `r length(gene_trees_calibrated)` were successfully time calibrated.
@@
 While the mean distance between orthologs is `r round( ortholog_distance_mean, 1 )` years, the mean distance between paralogs is `r round( paralog_distance_mean, 1 )` years.
-As described above, this is because the oldest speciation event is by definition the most recent common ancestor of the the species included in the study, but many gene families underwent duplication before this time.
+As described above, this is because the oldest speciation event is be definition the most recent common ancestor of the the species included in the study, but many gene families underwent duplication before this time.
 
@@
 
-Levin *et al.* [@Levin:2016bp] arrived at this conclusion by making pairwise comparisons of ortholog expression through development between species.
+Levin *et al.* [@Levin:2016bp] arrived at this conclusion my making pairwise comparisons of ortholog expression through development between species.
 They characterized each gene in each species as having expression that peaks in early, mid, or late temporal phase of development.

### 2017-01-27 manuscript.rmd
adds minor fix

--- 2017-01-26 manuscript.rmd
@@
 
-	# Files priduced by KMRR Rscript.R with Tau values and other expression statistics
+	# Files produced by KMRR Rscript.R with Tau values and other expression statistics
 	expression_file_names = c(

### 2017-01-27 manuscript.rmd
typo

@@
 
-The first study, Kryuchkova-Mostacci and Robinson-Rechavi (KMRR) 2016 [@KryuchkovaMostacci:2016iw], analyzed multiple vertebrate expression datasets to test the ortholog conjecture - the hypothesis that orthologs tend of have more conserved attributes (specificity of expression across organs in this case) than do paralogs [@Nehrt:2011bm].
+The first study, Kryuchkova-Mostacci and Robinson-Rechavi (KMRR) 2016 [@KryuchkovaMostacci:2016iw], analyzed multiple vertebrate expression datasets to test the ortholog conjecture - the hypothesis that orthologs tend to have more conserved attributes (specificity of expression across organs in this case) than do paralogs [@Nehrt:2011bm].
 Using pairwise comparisons (Fig.

### 2017-01-27 manuscript.rmd
typo

@@
 We focused on two possible sources of bias - node depth and branch length.
-We found no evidence that either affacted our results (Supplementary materials).
+We found no evidence that either affected our results (Supplementary materials).
 
@@
 
-	# Calcualte some summary statistics 
+	# Calculate some summary statistics 
 	ortholog_r = 

### 2017-01-27 manuscript.rmd
typo

@@
 
-		# Add a colmn with the gene tree hash 
+		# Add a column with the gene tree hash 
 		D %<>% mutate( gene_tree = digest( nhx ) )
@@
 
-		# For each row, bet the most recent common ancestor of the pairwise tip combination
+		# For each row, get the most recent common ancestor of the pairwise tip combination
 		D$mrca = 
@@
 
-		# Merge data on this mrca to the node
+		# Merge data of this mrca to the node
 		D %<>% left_join(

### 2017-01-27 manuscript.rmd
Changes wording

--- 2017-01-27 manuscript.rmd
@@
 
-The first study, Kryuchkova-Mostacci and Robinson-Rechavi (KMRR) 2016 [@KryuchkovaMostacci:2016iw], analyzed multiple vertebrate expression datasets to test the ortholog conjecture - the hypothesis that orthologs tend to have more conserved attributes (specificity of expression across organs in this case) than do paralogs [@Nehrt:2011bm].
+The first study, Kryuchkova-Mostacci and Robinson-Rechavi (KMRR) 2016 [@KryuchkovaMostacci:2016iw], analyzed multiple vertebrate expression datasets to test the ortholog conjecture - the hypothesis that orthologs tend of have more conserved attributes (specificity of expression across organs in this case) than do paralogs [@Nehrt:2011bm].
 Using pairwise comparisons (Fig.
@@
 		age.min =
-			c(  8, 29, 92, 162, 167, 296 ), 
+			c( 92, 167, 8, 296, 29, 162 ), 
 		clade = 
-			c( "Homininae", "Catarrhini", "Euarchontoglires", "Theria", "Mammalia", "Amniota" ),
+			c( "Euarchontoglires", "Mammalia", "Homininae", 
+					"Amniota", "Catarrhini", "Theria"
+			),
 		stringsAsFactors=FALSE
@@
 ```{r define_functions, echo=FALSE, message=F, warning=F}
-
-	#' Adds some random noise from a normal distribution to calibration
-	#' times
-	#' 
-	#' @param calibration_times A dataframe with two columns: age.min is the age 
-	#' of a clade, and clade is the name of the clade
-	#' @param sd_fraction this is multiplied by each value to set sd of that value 
-	#' @return A dataframe as above, but with noise
-	#' @export
-	add_noise_calibration_times = function( calibration_times, sd_fraction ) {
-		age = calibration_times$age.min
-		new_age = NA
-
-		while( NA %in% new_age ) {
-			x = sapply( 
-				age,
-				function( i ){
-					rnorm( 1, mean = i, sd = ( sd_fraction * i ) )
-				}
-			)
-
-			if( ! is.unsorted( x ) ){
-				new_age = x
-			}
-
-		}
-
-		calibration_times$age.min = new_age
-		return( calibration_times )
-	}
 
@@
 		tree = try( 
-			ape::chronos( nhx@phylo, calibration=calibration, ... ) 
+			ape::chronos( nhx@phylo, calibration=calibration ) 
 		)
@@
 	#' 
-	#' @param gene_trees_pruned A list of phylogenetic trees as treeio::treedata objects 
+	#' @param gene_trees_sub A list of phylogenetic trees as treeio::treedata objects 
 	#' with associated character data
@@
 	#' @return A list of calibrated phylogenetic trees as treeio::treedata objects. May be 
-	#' shorter than gene_trees_pruned if some calibrations fail.
+	#' shorter than gene_trees_sub if some calibrations fail.
 	#' @export
-	calibrate_trees = function( gene_trees_pruned, calibration_times, ... ) {
+	calibrate_trees = function( gene_trees_sub, calibration_times, ... ) {
 		# Make the trees ultrametric and calibrate the speciation nodes to 
@@
 			mclapply( 
-				gene_trees_pruned,
+				gene_trees_sub,
 				calibrate_tree, 
@@
 	# Remove tips from the trees that do not have expression data
-	gene_trees_pruned = mclapply( 
+	gene_trees_sub = mclapply( 
 		gene_trees_annotated, 
@@
 	# Remove trees with less than min_genes_with_expression tips with expression data
-	gene_trees_pruned = gene_trees_pruned[ ! is.na( gene_trees_pruned ) ]
+	gene_trees_sub = gene_trees_sub[ ! is.na( gene_trees_sub ) ]
 
@@
 	# Time calibrate all speciation nodes to the same years
-	gene_trees_calibrated = calibrate_trees( gene_trees_pruned, calibration_times, model="correlated" )
+	gene_trees_calibrated = calibrate_trees( gene_trees_sub, calibration_times, model="correlated" )
 
@@
 XXKMRR).
-Of the `r length(gene_trees)` trees that were parsed from ENSMBL Compara [@Herrero:2016jj], `r length(gene_trees_pruned)` passed taxon sampling criteria (`r min_genes_with_expression` genes) after removing tips without Tau values.
+Of the `r length(gene_trees)` trees that were parsed from ENSMBL Compara [@Herrero:2016jj], `r length(gene_trees_sub)` passed taxon sampling criteria (`r min_genes_with_expression` genes) after removing tips without Tau values.
 Of these, `r length(gene_trees_calibrated)` were successfully time calibrated.
@@
 Recent advances in phylogenetic comparative methods are particularly well suited to addressing questions about the evolution of functional genomic traits.
-Most early phylogenetic comparative methods attempted to remove evolutionary signal to correct statistical tests for correlations between traits, while more recent methods tend to focus on testing hypotheses of evolutionary processes [@Pennell:2013cw].
+Most early phylogenetic comparative methods attempted to account for evolutionary signal to correct statistical tests for correlations between traits, while more recent methods tend to focus on testing hypotheses of evolutionary processes [@Pennell:2013cw].
 The application of this newer focus to functional genomics provides an exciting opportunity to address long standing questions of broad interest, including the order of changes in functional genomics traits and shifts in rates of evolution of one functional genomics trait following changes in another trait.

### 2017-01-27 manuscript.rmd
Update manuscript.rmd

--- 2017-01-27 manuscript.rmd
@@
 
-Most CFG studies to date compare gene expression across species, often between distinct tissues or developmental time points.
+Most CFG studies to date compare gene expression across species, often between distinct cells, tissues or developmental time points.
 Gene expression has been a popular target of CFG studies because, with the advent of RNA-Seq and related transcriptome sequencing methods, it is the most readily collected functional genomics trait across diverse species.
@@
 We focused on two possible sources of bias - node depth and branch length.
-We found no evidence that either affected our results (Supplementary materials).
+We found no evidence that either affected our results (Supplementary Materials).
+We also examined the sensitivity of the results to the calibration times used for speciation events on the gene tree.
+This is important because it is expected that the most recent common ancestor of two genes that diverge due to speciation is older than the speciation event itself due to incomplete lineage sorting.
+There is also uncertainty associated with the timing of these speciation events.
+In replicate analyses we added random noise to the calibration times and found that it does not impact our inability to reject the null hypothesis (Supplementary Materials).
 
@@
 
+#### Investigation of sensitivity to calibration times
+
+```{r calibration_sensitivity, cache=TRUE, echo=FALSE, fig.cap="XX_PIC_VAR", warning=FALSE, message=FALSE}
+
+	noised_replicate_n = 10
+	sd_fraction = 0.2
+
+	noised_dates = 
+		replicate( 
+			noised_replicate_n, 
+			add_noise_calibration_times( calibration_times, sd_fraction ),
+			simplify = FALSE
+		)
+
+	nodes_contrast_noised_list = lapply(
+		noised_dates,
+		function( x ){
+			gene_trees_pruned %>% 
+				calibrate_trees( x, model="correlated" ) %>% 
+				add_pics_to_trees() %>% 
+				summarize_contrasts()
+		}
+	)
+
+	p_noised = lapply( nodes_contrast_noised_list, wilcox_oc )
+
+```
+
+We examined the ensitivity of our results to the specification of calibration dates for the speciation nodes.
+In `r noised_replicate_n` reanalyses, we drew a new date for each calibration from a normal distribution with the mean of the original date and a standard deviation `r sd_fraction` times the original date.
+If any daugter nodes became older than their parent, we repeated the sampling until the dates were congruent with the topology.
+The minimum Wilcoxon p in these reanalyses was `r min( as.numeric( unlist( p_noised ) ) )`, *i.e.* none of them reject the null hypothesis that the rate of evolution of Tau is greater following duplication events than speciation events.
+This is consistent with the analysis that uses the calibration dates as specified, indicating that our results are robust to the selection of calibration times for speciation nodes.
+
 ## References

### 2017-01-27 manuscript.rmd
updated md after merge
added final paragraph to the conclusion

@@
 
-Most CFG studies to date compare gene expression across species, often between distinct cells, tissues or developmental time points.
+Most CFG studies to date compare gene expression across species, often between distinct tissues or developmental time points.
 Gene expression has been a popular target of CFG studies because, with the advent of RNA-Seq and related transcriptome sequencing methods, it is the most readily collected functional genomics trait across diverse species.
@@
 
+These analyses also illustrate the importance of not conflating evolutionary processes with the patterns they generate.
+Stating that there is a pattern wherein paralogs tend to be more different than orthologs does not mean that there is a difference in the process by which orthologs and paralogs evolve.
+When paralogs tend to be more distantly related than orthologs, they can be more different from wach other when both evolve under the same processes.
+Noting that there are multiple pairwise comparisons between distantly related animals that show lower similarity in comparison at the midpoint of development does not mean that there is a general process by which this phase of development evolved deep in the animal tree.
+Changes restricted to a single branch will be included in multiple pairwise comparisons, and can give the false impression of a general pattern.
+In both cases, there is no way to interpret the impications of patterns between living organisms for understanding evolutionary processes that happened in the past without taking the relationships between the organisms - the phylogenetic tree - into account.
+
 ## Methods

### 2017-01-27 manuscript.rmd
Changes wording

@@
 We reanalyzed the the KMRR study using phylogenetic comparative methods, focusing on the dataset of Brawand *et al.* 2011 [@Brawand:2011du].
-For each internal node in each gene tree (Figure 1b), we measured the phylogenetic independent contrast[@Felsenstein:1985ua] (PIC) of the expression statistic Tau.
+For each internal node in each gene tree (Figure 1b), we measured the phylogenetic independent contrast [@Felsenstein:1985ua] (PIC) of the expression statistic Tau.
 This is the difference in values of Tau for descendant nodes scaled by the expected variance (determined largely by the lengths of these branches).

### 2017-01-27 manuscript.rmd

@@
 
-The first study, Kryuchkova-Mostacci and Robinson-Rechavi (KMRR) 2016 [@KryuchkovaMostacci:2016iw], analyzed multiple vertebrate expression datasets to test the ortholog conjecture - the hypothesis that orthologs tend to have more conserved attributes (specificity of expression across organs in this case) than do paralogs [@Nehrt:2011bm].
+The first study, Kryuchkova-Mostacci and Robinson-Rechavi (KMRR) 2016 [@KryuchkovaMostacci:2016iw], analyzed multiple vertebrate expression datasets to test the ortholog conjecture - the hypothesis that orthologs tend to have more conserved attributes (specificity of expression across organs in this case) than do paralogs [@Nehrt:2011bm, @Koonin:2005fp].
 Using pairwise comparisons (Fig.
@@
 To test the ortholog conjecture it is essential to have a null hypothesis that makes predictions distinct from the ortholog conjecture.
-A suitable null hypothesis is that there is no difference in evolutionary change in differential expression along branches that descend from speciation or duplication events (Figure 1b).
+A suitable null hypothesis is that there is no difference in evolutionary change in differential expression along branches that descend from speciation or duplication events [@Studer:2009bm] (Figure 1b).
 Under this hypothesis, we would predict that contrasts across speciation nodes and duplication nodes are drawn from the same distribution (Figure XXEXPECTATIONS a).
@@
 	library( digest )
-	library( hutan )
+	library( hutan )  # From https://github.com/caseywdunn/hutan
 	library( gridExtra )
@@
 
-			if( ! is.unsorted( x ) ){
+			if( ( ! is.unsorted( x ) ) & all( x > 0 ) ){
 				new_age = x
@@
 
-	#' Drop tips without expression data from a tree 
+	#' Get the number of speciation events in an nhx tree
+	#' 
+	#' @param nhx A phylogenetic tree as a treeio::treedata object, with a 
+	#' column @data$D that has value N for speciation events
+	#' @return An integer indicating the number of speciation events
+	#' @export
+	get_n_speciation = function( nhx ){
+		sum( nhx@data$D[ ! is.tip.nhx( nhx ) ] == "N", na.rm = TRUE )
+	}
+
+	#' Drop tips without expression data from a tree. 
 	#' 
@@
 	#' @return A treeio::treedata object, or NA if less than 
-	#' min_genes_with_expression tips would have been retained
+	#' min_genes_with_expression
 	#' @export
@@
 
+	n_removed_for_no_speciation = length( gene_trees_pruned )
+
+	# Remove trees with no speciation events
+	gene_trees_pruned = gene_trees_pruned[  unlist( lapply( gene_trees_pruned, get_n_speciation ) ) > 0 ]
+
+	n_removed_for_no_speciation = n_removed_for_no_speciation - length( gene_trees_pruned )
+
 ```
@@
 XXKMRR).
-Of the `r length(gene_trees)` trees that were parsed from ENSMBL Compara [@Herrero:2016jj], `r length(gene_trees_pruned)` passed taxon sampling criteria (`r min_genes_with_expression` genes) after removing tips without Tau values.
+Of the `r length(gene_trees)` trees that were parsed from ENSMBL Compara [@Herrero:2016jj], `r length(gene_trees_pruned)` passed taxon sampling criteria (`r min_genes_with_expression` genes) after removing tips without Tau values and had at least one speciation event.
 Of these, `r length(gene_trees_calibrated)` were successfully time calibrated.
@@
 This means that having information on whether two genes are orthologs or paralogs provides little added information about expression beyond knowing how distantly related the two genes are.
+Closely related paralogs, for example, are expected to be more similar than distantly related orthologs [@Studer:2009bm].
 This has several implications.
@@
 Closely related paralogs, for example, may tend to have more similar phenotypes than more distantly related orthologs.
-
-The widespread invocation of the ortholog conjecture follows from the expectation that it is strong, widely distributed phenomenon.
-Our failure to detect a signature consistent with the ortholog conjecture suggests that it may not be, and that it should not be assumed in systems where it has not been explicitly tested.
+There has been considerable recent intrest in, and controversy about, the ortholog conjecture [@Koonin:2005fp; @Studer:2009bm; @Nehrt:2011bm; @Gabaldon:2013fi].
+Our reanalysys is consistent with multiple studies that have not found support for the ortholog conjecture in comparative analyses of gene ontology [@Studer:2009bm; @Nehrt:2011bm] or gene expression [@Yanai:2004hk].
+A previous study of expression concluded that there was strong support for the ortholog conjecture , though like KMRR it used pairwise comparisons.
+
+; @Altenhoff:2012ea
 
@@
 
-The fact that the first two CFG studies we reanalyzed show serious problems with pairwise comparisons indicates that there likely to be similar problems in other CFG studies, and that future CFG studies will be compromised if they continue to use pairwise methods.
+The fact that the first two comparative functional genomics studies we reanalyzed show serious problems with pairwise comparisons indicates that there likely to be similar problems in other studies that use these methods, and that future comparative studies will be compromised if they continue to use pairwise methods.
 Studies of evolutionary functional genomics should not be focused on the tips of the tree, they should explicitly delve into the tree itself.
 
-These analyses also illustrate the importance of not conflating evolutionary processes with the patterns they generate.
-Stating that there is a pattern wherein paralogs tend to be more different than orthologs does not mean that there is a difference in the process by which orthologs and paralogs evolve.
-When paralogs tend to be more distantly related than orthologs, they can be more different from wach other when both evolve under the same processes.
-Noting that there are multiple pairwise comparisons between distantly related animals that show lower similarity in comparison at the midpoint of development does not mean that there is a general process by which this phase of development evolved deep in the animal tree.
-Changes restricted to a single branch will be included in multiple pairwise comparisons, and can give the false impression of a general pattern.
-In both cases, there is no way to interpret the impications of patterns between living organisms for understanding evolutionary processes that happened in the past without taking the relationships between the organisms - the phylogenetic tree - into account.
+These analyses illustrate how important it is to not conflate evolutionary patterns with the processes that generated them.
+Finding a pattern wherein paralogs tend to be more different than orthologs is not evidence that there are different processes by which orthologs and paralogs evolve.
+This is also the expected pattern when they evolve under the same process but paralogs tend to be more distantly related to each other than orthologs are.
+The fact that multiple pairwise comparisons of developmental gene expression across diverse species share a particular pattern is not evidence of a general process that explains the differences between all species in the analysis.
+It is also the expected pattern when there are unique changes specific to a single species, as these changes will impact all pairwise comparisons to that species and therefore be counted multiple times.
+To use patterns across living species to test hypotheses about evolutionary processes, it is also necessary to incorporate information about evolutionary relationships, *i.e.* phylogenies.
+There have been decades of work on building comparative phylogenetic methods that do exactly that, and they are just as relevant to comparing functional genomics traits across species as they are to comparing morphology or any of the other traits they are already routinely applied to.
 
@@
 The most recent commit at the time of the analysis presented here was `r system("git log | head -n 1", intern=TRUE) %>% str_replace("commit ", "")`.
+See the `readme.md` file in this repository for more information on the contents of the source file and how to re-execute them.
 
@@
 They posted all data and code needed to re-execute their analyses at figshare: https://figshare.com/articles/Tissue-specificity_of_gene_expression_diverges_slowly_between_orthologs_and_rapidly_between_paralogs/3493010/2 .
-We slightly altered their `Rscript.R` to simplify file paths.
+We slightly altered their `Rscript.R` to simplify file paths and specify one missing variable.
 This modified script and their data files are available in the github repository for this paper, as are the intermediate files that were generated by their analysis script that we used in our own analyses.
 We also obtained the `Compara.75.protein.nh.emf` gene trees [@Herrero:2016jj] from ftp://ftp.ensembl.org/pub/release-75/emf/ensembl-compara/homologies/ and place them in the same directory as this file.
+These gene trees include branch lengths and annotate each internal node as being a duplication or speciation event.
 
 We considered only the data from Brawand *et al.* 2011 [@Brawand:2011du] for the eight taxa included in KMRR.
-We left in sex chromosome genes and testes expression data, as this corresponded to the analyses that provided the strongest support for the ortholog conjecture and therefore the most conservative reconsideration of it.
-
-Each of these gene trees includes includes homologous sequences derived from a common ancestral sequence.
-The tip nodes of a gene tree can include both orthologs and paralogs, and each internal node is an inferred speciation or duplication event.
-Prior to phylogenetic analyses of gene expression, each tree was first pruned to include only tips for which expression data were available.
-The speciation nodes in the tree were then time calibrated so that branch lengths were consistent across trees.
-We then calculated the magnitude of the independent contrast for differential expression at each internal node in each of these trees.
+We left in sex chromosome genes and testes expression data, which KMRR removed in some of their sensitivity analyses.
+This corresponded to the analyses that provided the strongest support for the ortholog conjecture and therefore the most conservative reconsideration of it.
+
+After parsing the trees from the Compara file with treeio, which was recently split from ggtree [@Yu:2016fo], we added Tau estimates generated by the KMRR `Rscript.R` to the tree data objects.
+We then pruned away tips without expression data, retaining only the trees with `r min_genes_with_expression` or more tips.
+We also only retained trees with one or more speciation events, as speciation events are required for calibration steps.
+This removes trees that have multiple genes from only one species after pruning away tips without expression data.
+
+The speciation nodes in the gene tree were then calibrated to the dates specified in the KMRR code.
+This was necessary to scale the branch lengths so that they are equivalent across the trees, and in turn to scale the phylogenetic independent contrasts (which take branch length into account) so they are equivalent across trees.
+We used the `chronos()` function from the R package ape [@Paradis:2004dv] for this scaling, with the `correlated` model.
+See supplementary Materials for additional sensitivity analyses to time calibration.
+Some trees could not be calibrated with these hard node constraints, and were discarded.
+
+For each node in the remaining calibrated trees, we calculated the phylogenetic independent contrast for Tau across its daughter branches with the `pic()` function in ape [@Paradis:2004dv].
+We then collected the contrasts from all trees into a single table, along with otherr annotations including whether the node is a speciation or duplication event.
+This table, `nodes_contrast`, was then analyzed as described in the main text for the presented plots and tests.
 
@@
 We obtained the matrix of pairwise scores that underlies their [Figure 4d](http://www.nature.com/nature/journal/v531/n7596/fig_tab/nature16994_F4.html) and confirmed we could reproduce their published results.
-We then removed duplicate rows, identified ctenophores as overrepresented among the low outliers in the mid-developmental transition column, and applied the Wilcoxon test in place of the Kolmogorov-Smirnov test.
-These data, our analysis code, and additional information are available in a git repository at [https://github.com/caseywdunn/levin2016](https://github.com/caseywdunn/levin2016).
-This repository includes our analysis notebook at [https://rawgit.com/caseywdunn/levin2016/master/reanalyses.html](https://rawgit.com/caseywdunn/levin2016/master/reanalyses.html).
+We then removed duplicate rows, applied the Wilcoxon test in place of the Kolmogorov-Smirnov test, and identified ctenophores as overrepresented among the low outliers in the mid-developmental transition column.
+An annotated explanation of these analyses is included in the git repository at
 

### 2017-01-27 manuscript.rmd
merged...

@@
 We reanalyzed the the KMRR study using phylogenetic comparative methods, focusing on the dataset of Brawand *et al.* 2011 [@Brawand:2011du].
-For each internal node in each gene tree (Figure 1b), we measured the phylogenetic independent contrast [@Felsenstein:1985ua] (PIC) of the expression statistic Tau.
+For each internal node in each gene tree (Figure 1b), we measured the phylogenetic independent contrast[@Felsenstein:1985ua] (PIC) of the expression statistic Tau.
 This is the difference in values of Tau for descendant nodes scaled by the expected variance (determined largely by the lengths of these branches).
@@
 Recent advances in phylogenetic comparative methods are particularly well suited to addressing questions about the evolution of functional genomic traits.
-Most early phylogenetic comparative methods attempted to account for evolutionary signal to correct statistical tests for correlations between traits, while more recent methods tend to focus on testing hypotheses of evolutionary processes [@Pennell:2013cw].
+Most early phylogenetic comparative methods attempted to remove evolutionary signal to correct statistical tests for correlations between traits, while more recent methods tend to focus on testing hypotheses of evolutionary processes [@Pennell:2013cw].
 The application of this newer focus to functional genomics provides an exciting opportunity to address long standing questions of broad interest, including the order of changes in functional genomics traits and shifts in rates of evolution of one functional genomics trait following changes in another trait.
@@
 There are also technical factors that can lead to an excess of duplication events deeper in the tree.
-Gene tree estimation errors, for example, tend to lead to the overestimation of deep duplications [@Hahn:2007cj].
+Assembly errors, for example, tend to lead to the overestimation of deep duplications [@Hahn:2007cj].
 If independent contrast values also tended to to be lower at greater node depth, it could interact with the preponderance of duplications at greater depth to create a pattern of lower contrasts associated with duplication events.
@@
 We therefore examined only the contrasts that fell within in the range of expected variance seen for speciation contrasts, excluding duplication contrasts that fall outside of this range.
-This reanalysis does not reject the null hypothesis either (Figure XXAscertainment d), indicating that branch length bias is not responsible for rejecting the ortholog conjecture.
+This reanalysis does not reject the ortholog conjecture either (Figure XXAscertainment d), indicating that branch length bias is not responsible for the result.
 

### 2017-01-28 manuscript.rmd
adds my edits again

@@
 There are also technical factors that can lead to an excess of duplication events deeper in the tree.
-Assembly errors, for example, tend to lead to the overestimation of deep duplications [@Hahn:2007cj].
+Gene tree estimation errors errors, for example, tend to lead to the overestimation of deep duplications [@Hahn:2007cj].
 If independent contrast values also tended to to be lower at greater node depth, it could interact with the preponderance of duplications at greater depth to create a pattern of lower contrasts associated with duplication events.

### 2017-01-28 manuscript.rmd
Fixes text on ILS and adds citations

@@
 We found no evidence that either affected our results (Supplementary Materials).
-We also examined the sensitivity of the results to the calibration times used for speciation events on the gene tree.
-This is important because it is expected that the most recent common ancestor of two genes that diverge due to speciation is older than the speciation event itself due to incomplete lineage sorting.
+We also examined the sensitivity of the results to the calibration times used for speciation events on the gene trees.
+This is important because it is expected that genes from separate species trace backward in time to a common ancestor at older times than the splitting times of the species to which they belong [Takahata:1989uh, Degnan:2009hr].
 There is also uncertainty associated with the timing of these speciation events.

### 2017-01-28 manuscript.rmd
now parse species names

--- 2017-01-27 manuscript.rmd
@@
 
+		# Keep trying until dates are coherent with tree structure
+		# This is a ladder tree, so that means checking that they 
+		# are still in ascending order
 		while( NA %in% new_age ) {
@@
 		# expression data, contrast results, etc...
+		# Also take the absolute value of the contrast, since we only consider magnitude.
 		all_nodes = 
@@
 			) %>%
-			bind_rows()
+			bind_rows() %>%
+			mutate( pic = abs( pic ) )
 
@@
 		# Create a subset that corresponds to nodes with contrasts. This excludes tips.
-		# Also take the absolute value of the contrast, since we only consider magnitude.
 		nodes_contrast = 
@@
 			filter( ! is.na( pic ) ) %>%
-			mutate( pic = abs( pic ) ) %>%
 			filter( ! is.na( D ) )
@@
 
+```{r load_trees, echo=FALSE, cache=TRUE, message=F, warning=F}
+
+	# Read in the tree file as a vector of strings, one per line
+	lines = readLines( gene_tree_name )
+
+	# Isolate strings that are trees
+	# These start with (
+	tree_lines = lines[ grepl( "^\\(", lines, perl = TRUE ) ]
+
+	# Parse each string to a treeio::treedata object, adding a label column 
+	# for node names
+	gene_trees = mclapply(
+		tree_lines,
+		function( x ){
+			tree_tc = textConnection( x )
+			tree = treeio::read.nhx( tree_tc )
+			close( tree_tc )
+			tree@data$label = c( tree@phylo$tip.label, tree@phylo$node.label )
+			return( tree )
+		}, mc.cores=cores
+	)
+
+	# Parse the tree annotations, these are lines that start with SEQ
+	# These have the species names, which are not in the trees themselves
+	tip_annotations = 
+		lines[ grepl( "^SEQ", lines, perl = TRUE ) ] %>%	# Get lines that start with SEQ
+		str_c( collapse="\n" ) %>%												# Combine them into one string
+		read_delim( delim=" ", col_names = FALSE ) %>% 		# Read string as tibble
+		select( 2, 8 ) %>%																# Select the columns with species and gene ID
+		rename(  Ensembl.Gene.ID = X8 ) %>%
+		rename(  species = X2 )
+
+```
+
 ```{r load_expression, echo=FALSE, message=F, warning=F}
 
-	# Read the expression data for each species and combine into a single tibble
+	# Read the expression data for each species and combine into a single tibble,
+	# adding the species names from tip_annotations
 	expression = 
@@
 		) %>%
-	bind_rows()
-
-```
-
-```{r load_trees, echo=FALSE, cache=TRUE, message=F, warning=F}
-
-	# Read in the tree file as a vector of strings, one per line
-	lines = readLines( gene_tree_name )
-
-	# Only retain strings that are trees
-	lines = lines[ grepl( "^\\(", lines, perl = TRUE ) ]
-
-	# Parse each string to a treeio::treedata object, adding a label column 
-	# for node names
-	gene_trees = mclapply(
-		lines,
-		function( x ){
-			tree_tc = textConnection( x )
-			tree = treeio::read.nhx( tree_tc )
-			close( tree_tc )
-			tree@data$label = c( tree@phylo$tip.label, tree@phylo$node.label )
-			return( tree )
-		}, mc.cores=cores
-	)
+		bind_rows() %>%
+		left_join( tip_annotations, by="Ensembl.Gene.ID" )
+
 ```

### 2017-01-28 manuscript.rmd
fixed Hominini node names

@@
 		age.min =
-			c(  8, 29, 92, 162, 167, 296 ), 
+			c(  7, 9, 29, 92, 162, 167, 296 ), 
 		clade = 
-			c( "Homininae", "Catarrhini", "Euarchontoglires", "Theria", "Mammalia", "Amniota" ),
+			c( "Hominini", "Homininae", "Catarrhini", "Euarchontoglires", "Theria", "Mammalia", "Amniota" ),
 		stringsAsFactors=FALSE
@@
 		is.tip
+	}
+
+	#' Fix ensembl clade names.
+	#' Homininae is pan + homo + gorilla
+	#' Hominini is pan + homo
+	#' ENSEMBL Compara trees lislabel Hominini as Homininae
+	#' 
+	#' @param nhx A phylogenetic tree as a treeio::treedata object
+	#' @return A phylogenetic tree as a treeio::treedata object
+	#' @export
+	fix_hominini = function( nhx ) {
+		# Get the numbers for all nodes labeled as homininae
+		homininae_nodes = nhx@data$node[ nhx@data$label == "Homininae" ]
+
+		# Check which have no gorilla as a descendent
+		no_gorilla = sapply(
+			homininae_nodes,
+			function( x ){
+				tips = hutan::tip_descendants( nhx@phylo, x )
+				tip_species = nhx@data$species[ tips ]
+				return( ! "gorilla_gorilla" %in% tip_species )
+			}
+		)
+
+		if( any( no_gorilla ) ){
+			hominini_nodes = homininae_nodes[ no_gorilla ]
+			nhx@data$label[ hominini_nodes ] = "Hominini"
+		}
+
+		return( nhx )
 	}
@@
 
+	# fix clade names
+	gene_trees_pruned = lapply( gene_trees_pruned, fix_hominini )
+
 ```

### 2017-01-28 manuscript.rmd
added summary statistics and example trees

@@
 		# Also take the absolute value of the contrast, since we only consider magnitude.
-		all_nodes = 
+		nodes_all = 
 			lapply( 
@@
 		# Make a new human readable column for plots that records speciation/ duplication
-		all_nodes$Event = NA
-		all_nodes$Event[ all_nodes$D == "Y" ] = "Duplication"
-		all_nodes$Event[ all_nodes$D == "N" ] = "Speciation"
+		nodes_all$Event = NA
+		nodes_all$Event[ nodes_all$D == "Y" ] = "Duplication"
+		nodes_all$Event[ nodes_all$D == "N" ] = "Speciation"
 
@@
 		nodes_contrast = 
-			all_nodes %>%
+			nodes_all %>%
 			filter( ! is.na( pic ) ) %>%
@@
 
-The speciation nodes in the gene tree were then calibrated to the dates specified in the KMRR code.
-This was necessary to scale the branch lengths so that they are equivalent across the trees, and in turn to scale the phylogenetic independent contrasts (which take branch length into account) so they are equivalent across trees.
-We used the `chronos()` function from the R package ape [@Paradis:2004dv] for this scaling, with the `correlated` model.
-See supplementary Materials for additional sensitivity analyses to time calibration.
+The gene tree were then time calibrated.
+The goal is not necesarily to have precise dates for each node, but to scale branch lengths so that they are equivalent across the trees.
+This in turn scales the phylogenetic independent contrasts (which take branch length into account) so they can be compared appropriately.
+We found a problem with the clade names on the ENSEMBL Compara trees, which others have also identified [@Daub:2016gd], that first had to be fixed.
+Hominini is the name for the clade that includes humans and chimps, while Homininae is the clade that includes humans, chimps, and gorillas.
+Both clades are labeled as Homininae in the Compara trees, though, which would interfere with calibration of these nodes.
+To fix the problem, we identified all clades labeled Homininae that have no gorilla sequence and renamed them Hominini.
+We then calibrated the trees by fixing the speciation nodes to the dates specified in the KMRR code, with the exception of Hominini and Homininae.
+These we set to 7 million years and 9 million years, drawing on the same TimeTree source [@Hedges:2006jv] that KMRR used.
+We used the `chronos()` function from the R package ape [@Paradis:2004dv] for this calibration, with the `correlated` model.
+See Supplementary Materials for additional sensitivity analyses to time calibration.
 Some trees could not be calibrated with these hard node constraints, and were discarded.
@@
 ### KMRR Analyses
+
+#### Summary statistics
+
+```{r contrast_table, cache=TRUE, echo=FALSE, fig.cap="XX_PIC_VAR", warning=FALSE, message=FALSE}
+
+clades_table = 
+	nodes_contrast %>% 
+	filter(D=="N") %>% 
+	.$label %>% 
+	table() %>% 
+	as.data.frame()
+
+names( clades_table ) = c( "Clade", "N" )
+
+clades_table %<>%  slice( match( calibration_times$clade, Clade) )
+
+knitr::kable(
+  clades_table,
+  caption = "Supplementary Table 1. Number of speciation nodes with contrasts for each clade on the tree."
+)
+```
+
+```{r species_table, cache=TRUE, echo=FALSE, fig.cap="XX_PIC_VAR", warning=FALSE, message=FALSE}
+
+	nodes_all = 
+			lapply( 
+				gene_trees_pic, 
+				function( nhx ){
+					tags = nhx@data
+					tags$gene = digest( nhx ) # Creates a  hash that is unique to each gene tree
+					tags %<>% select( -( B ) ) # B has inconsistent types, remove it
+					return( tags )
+				}
+			) %>%
+			bind_rows() %>%
+			mutate( pic = abs( pic ) )
+
+	species_table = 
+		nodes_all %>% 
+		filter( ! is.na( species ) ) %>% 
+		.$species %>% 
+		table() %>% 
+		as.data.frame()
+
+	names( species_table ) = c( "Species", "N" )
+
+	species_table %<>% 
+		arrange( Species ) %>%
+		#mutate( Species = str_to_title( Species ) ) %>%
+		mutate( Species = str_replace( Species, "_", " ") )
+
+	knitr::kable(
+	  species_table,
+	  caption = "Supplementary Table 2. Number of tips for each species in trees that were used to calculate the independent contrasts."
+	)
+
+```
+
+```{r example_trees, cache=TRUE, echo=FALSE, fig.cap="XX_PIC_VAR", warning=FALSE, message=FALSE}
+
+	ntips = lapply( gene_trees_pic, function(x) length( x@phylo$tip.label ))
+	to_show = which( ntips < 20 & ntips >15 )[1:8]
+
+	example_trees = lapply( 
+		to_show,
+		function( x ){
+				nhx = gene_trees_pic[[x]]
+				nhx@data %<>% mutate( clade_name=label )
+				nhx@data$clade_name[ is.tip.nhx(nhx) | nhx@data$D !="N" ] = NA
+
+				nhx@data$Event = NA
+				nhx@data$Event[ nhx@data$D == "Y" ] = "Duplication"
+				nhx@data$Event[ nhx@data$D == "N" ] = "Speciation"
+
+				nhx %>% 
+					ggtree() + 
+					geom_tiplab( aes( label=species ), hjust=-0.1 ) + 
+					xlim( NA, 1400 ) + 
+					geom_tippoint( aes( size=Tau ) ) +
+					geom_nodepoint( aes( color=Event ), shape=15, size = 2 ) +
+					geom_tippoint( aes( size=Tau ) ) +
+					geom_text( aes( label=round( abs( pic ), 3 ) ), vjust=1.5, hjust=1, size=2.5, col="blue" ) + 
+					geom_text( aes( label=clade_name ), vjust=-.5, hjust=1, size=2.5, col="black" ) + 
+					theme( legend.position="right" )
+			}
+		)
+
+	print( example_trees )
+```
+
+> Supplementary Figure XX_EXAMPLE_TREES | A set of example trees to demonstrate the data and analyses that underlie the results.
+The value of Tau (expression specificity) is indicated by the sizes of the circles at the tips of the tree.
+Whether an internal node is a speciation or duplication is indicated by color.
+Speciation nodes are labeled by clade name.
+Branch lengths are scaled to time.
+The blue number is the independent contrast for each node.
 

### 2017-01-29 manuscript.rmd
typo

@@
 Second, it is a missed opportunity to learn about biology that can only be understood in an explicit historical context that considers these relationships.
-Here we examine two recently published comparative gene expression studies that used multiple pairwise comparison rather than phylogenetic approaches to investigate evolutionary patterns of gene expression.
+Here we examine two recently published comparative gene expression studies that used multiple pairwise comparisons rather than phylogenetic approaches to investigate evolutionary patterns of gene expression.
 We find problems with the pairwise comparisons used in both studies that undermine their published conclusions.

### 2017-01-29 manuscript.rmd
changes to text

--- 2017-01-28 manuscript.rmd
@@
 This means that having information on whether two genes are orthologs or paralogs provides little added information about expression beyond knowing how distantly related the two genes are.
-Closely related paralogs, for example, are expected to be more similar than distantly related orthologs [@Studer:2009bm].
 This has several implications.
@@
 These terms can have straightforward meaning in small gene trees with simple duplication/speciation histories, but the utility of the terms breaks down on larger more complex gene trees.
-Orthology and paralogy are annotations on the tips of the tree that are derived from the inferred structure of the tree and annotations of duplication and speciation at internal tree nodes.
+Orthology and paralogy are annotations on the tips of the tree that are derived from the inferred structure of the tree and history of duplication and speciation at internal tree nodes.
 In this sense, orthology and paralogy are statements about the internals of the tree that are distilled into statements at the tips of the tree.
 Much is lost in the process, though.
-In this case, KMRR [@KryuchkovaMostacci:2016iw] relied on orthology and paralogy annotations that are derived from the inferred duplication and speciation history on gene phylogenies [@Herrero:2016jj].
 These are the same pre-computed trees we use in our phylogenetic re-analyses of their study.
 For most questions it is much more direct to focus on the structure of the tree and the inferred processes within the tree, such as which internal nodes are duplication or speciation events and how much change occurs along the branches.
+In this case, KMRR [@KryuchkovaMostacci:2016iw] relied on orthology and paralogy annotations that are derived from the inferred duplication and speciation history on gene phylogenies [@Herrero:2016jj].
 
 #### Implications for the ortholog conjecture
+
+There has been considerable recent intrest in, and controversy about, the ortholog conjecture [@Koonin:2005fp; @Studer:2009bm; @Nehrt:2011bm; @Gabaldon:2013fi].
+Our results are consistent with multiple studies that have not found support for the ortholog conjecture [@Studer:2009bm; @Nehrt:2011bm; @Yanai:2004hk].
+Another sudy found only slightly greater similarity between orthologs than paralogs [@Altenhoff:2012ea].
 
@@
 This has important biological implications.
-It suggests that the mechanism of gene divergence (speciation versus duplication) may not have as strong an impact on phenotypic divergence as thought.
+It suggests that the mechanism of gene divergence (speciation versus duplication) may not have as strong an impact on phenotypic divergence as sometimes proposed.
 It also has technical implications.
 Rather than focus on whether genes are orthologs or paralogs when attempting to predict function, it may be more effective to simply focus on how closely related or distantly related they are.
-Closely related paralogs, for example, may tend to have more similar phenotypes than more distantly related orthologs.
-There has been considerable recent intrest in, and controversy about, the ortholog conjecture [@Koonin:2005fp; @Studer:2009bm; @Nehrt:2011bm; @Gabaldon:2013fi].
-Our reanalysys is consistent with multiple studies that have not found support for the ortholog conjecture in comparative analyses of gene ontology [@Studer:2009bm; @Nehrt:2011bm] or gene expression [@Yanai:2004hk].
-A previous study of expression concluded that there was strong support for the ortholog conjecture , though like KMRR it used pairwise comparisons.
-
-; @Altenhoff:2012ea
+Closely related paralogs, for example, may tend to have more similar phenotypes than more distantly related orthologs [@Studer:2009bm].
+
+The ortholog conjecture does not have to be an all or nothing question.
+It may be the case that the rates of phenotypic evolution following duplication may be greater than that following duplication in some organisms, gene families, and evolutionary processes.
+We just don't find evidence for it when summarizing gene expression acriss tissues with Tau in these mammals.
+This calls into question the general predictive power of the ortholog conjecture, and until these processes are better understood it will be necessary to test for it in each situation.
+These tests should be articulated in terms of clear alternative hypotheses [@Studer:2009bm] that make distinct phylogenetic comparative predictions.
 

### 2017-01-29 manuscript.rmd
added kmrr fib 2a equivalent

@@
 
+		# Add the species
+		D %<>% mutate ( species_a = nhx@data$species[ D$tip_a ] )
+		D %<>% mutate ( species_b = nhx@data$species[ D$tip_b ] )
+
 		# Add the values at the tips
@@
 			nhx@data %>%
-				select( D, node, node_age, pic ),
+				select( D, node, node_age, label, pic ),
 			by = c( "mrca" = "node")
@@
 		bind_rows()
+
+	pairwise_summary$Event = NA
+	pairwise_summary$Event[ pairwise_summary$D == "N" ] = "Speciation"
+	pairwise_summary$Event[ pairwise_summary$D == "Y" ] = "Duplication"
 
@@
 
-```
+	calibration_times_all = data.frame(
+        age =
+            c( 20, 92, 167, 9, 42, 74, 296, 535, 104, 937, 29, 722, 441, 65, 15, 1215, 
+                414, 371, 162, 25, 74, 77, 86, 7 ), 
+        clade = 
+            c("Hominoidea", "Euarchontoglires", "Mammalia", "Homininae", 
+                    "Simiiformes", "Primates", "Amniota", "Vertebrata", "Eutheria", 
+                    "Bilateria", "Catarrhini", "Chordata", "Euteleostomi", 
+                    "Haplorrhini", "Hominidae", "Opisthokonta", "Sarcopterygii", 
+                    "Tetrapoda", "Theria", "Murinae", "Sciurognathi", "Rodentia", "Glires", "Hominini"
+            ),
+        stringsAsFactors=FALSE
+    )
+
+	pairwise_summary %>% 
+		filter( species_a == "homo_sapiens" | species_b == "homo_sapiens" ) %>% 
+		filter( ! is.na( D ) ) %>%
+		group_by( Event, label ) %>% 
+		summarise( cor.test(  tau_a, tau_b )$estimate, n=n() ) %>% 
+		rename( rho=`cor.test(tau_a, tau_b)$estimate` ) %>%
+		rename( clade = label ) %>% 
+		left_join( calibration_times_all, clade=clade ) %>%
+		ggplot() + 
+			geom_point( aes( x=age, y=rho, col=Event, size=n ) ) +
+			theme_classic()
+
+```
+
+> Figure XX_KMRR_Pairwise | This figures demonstrates that the pairwise analyses presented in Figure 2a of KMRR can be reproduced with the subset of data we consider here and our own independent code base.
+Each circle indicates the Tau Pearson correlation (rho) for a set of pairwise comparisons annotated with a specific node name of a given age and event type (*i.e.*, whether the divergence at that nodes was due to speciation or duplication).
+Following KMRR Figure 2a, this figure shows only the comparisons that include a human sequence.
 

### 2017-01-29 manuscript.rmd
typo, changed unit years to My

@@
 This is, in fact, the case for these data.
-While the mean distance between orthologs is `r round( ortholog_distance_mean, 1 )` years, the mean distance between paralogs is `r round( paralog_distance_mean, 1 )` years.
+While the mean distance between orthologs is `r round( ortholog_distance_mean, 1 )` My, the mean distance between paralogs is `r round( paralog_distance_mean, 1 )` My.
 As described above, this is because the oldest speciation event is by definition the most recent common ancestor of the the species included in the study, but many gene families underwent duplication before this time.

### 2017-01-30 manuscript.rmd
typos

@@
 
-There has been considerable recent intrest in, and controversy about, the ortholog conjecture [@Koonin:2005fp; @Studer:2009bm; @Nehrt:2011bm; @Gabaldon:2013fi].
+There has been considerable recent interest in, and controversy about, the ortholog conjecture [@Koonin:2005fp; @Studer:2009bm; @Nehrt:2011bm; @Gabaldon:2013fi].
 Our results are consistent with multiple studies that have not found support for the ortholog conjecture [@Studer:2009bm; @Nehrt:2011bm; @Yanai:2004hk].
-Another sudy found only slightly greater similarity between orthologs than paralogs [@Altenhoff:2012ea].
-
-Our reanalysis of KMRR using phylogenetic methods suggests that the ortholog conjectures is not a dominant pattern that is central to explaining the evolution of phenotypic diversity in gene families.
+Another study found only slightly greater similarity between orthologs than paralogs [@Altenhoff:2012ea].
+
+Our reanalysis of KMRR using phylogenetic methods suggests that the ortholog conjecture is not a dominant pattern that is central to explaining the evolution of phenotypic diversity in gene families.
 This has important biological implications.
@@
 It may be the case that the rates of phenotypic evolution following duplication may be greater than that following duplication in some organisms, gene families, and evolutionary processes.
-We just don't find evidence for it when summarizing gene expression acriss tissues with Tau in these mammals.
+We just don't find evidence for it when summarizing gene expression across tissues with Tau in these mammals.
 This calls into question the general predictive power of the ortholog conjecture, and until these processes are better understood it will be necessary to test for it in each situation.

### 2017-01-30 manuscript.rmd
typos

@@
 
-The gene tree were then time calibrated.
-The goal is not necesarily to have precise dates for each node, but to scale branch lengths so that they are equivalent across the trees.
+The gene trees were then time calibrated.
+The goal is not necessarily to have precise dates for each node, but to scale branch lengths so that they are equivalent across the trees.
 This in turn scales the phylogenetic independent contrasts (which take branch length into account) so they can be compared appropriately.

### 2017-01-30 manuscript.rmd
Update manuscript.rmd
Adds bibliography PCMs

@@
 
-A rich theoretical and statistical methodology of phylogenetic comparative methods have been developed over the last three decades to address the challenges and opportunities of comparisons across species [@Felsenstein:1985ua; @Grafen:1989um; @Chamberlain:2012vx; @Garamszegi:2014ew; @Pagel:1999gc].
+A rich theoretical and statistical methodology of phylogenetic comparative methods have been developed over the last three decades to address the challenges and opportunities of comparisons across species [@Felsenstein:1985ua; @Grafen:1989um; @Pagel:1999gc; Revell:2012jd; @FitzJohn:2012ey; @Uyeda:2014jl; @Garamszegi:2014ew].
 These challenges include non-independence of observations across species -- more closely related species will tend to be more similar to each other than distantly related species.
@@
 In this manuscript we examine this issue.
-Although the value of phylogenetic approaches has been repeatedly shown in comparative analyses of other types of character data [@Chamberlain:2012vx; @Ricklefs:1996wb], this value is apparently not well known to the functional genomics community since the field has until recently largely focused on within-species analyses.
+Although the value of phylogenetic approaches has been repeatedly shown in comparative analyses of other types of character data [@Ricklefs:1996wb; @Chamberlain:2012vx; @OMeara:2012gq], this value is apparently not well known to the functional genomics community since the field has until recently largely focused on within-species analyses.
 Many initial CFG studies have also been focused on only two species at a time, where the differences between phylogenetic and pairwise approaches are not as pronounced.
@@
 We also examined the sensitivity of the results to the calibration times used for speciation events on the gene trees.
-This is important because it is expected that genes from separate species trace backward in time to a common ancestor at older times than the splitting times of the species to which they belong [Takahata:1989uh, Degnan:2009hr].
+This is important because it is expected that genes from separate species trace backward in time to a common ancestor at older times than the splitting times of the species to which they belong [@Takahata:1989uh; @Degnan:2009hr].
 There is also uncertainty associated with the timing of these speciation events.

### 2017-01-30 manuscript.rmd
Update manuscript.rmd
Minor edits to describe results in KMRR

@@
 
-KMRR's published analyses are based on pairwise comparisons (Figure 1a) between Tau within each gene family.
+Results in KMRR are based on pairwise comparisons (Figure 1A) between Tau within each gene family.
 They found the correlation of Tau for orthologs to be significantly greater than the correlation of Tau for paralogs, *i.e.* that orthologs tend to be more similar to each other than paralogs.

### 2017-01-30 manuscript.rmd
Update manuscript.rmd
Minor fix to citation of PIC

@@
 We reanalyzed the the KMRR study using phylogenetic comparative methods, focusing on the dataset of Brawand *et al.* 2011 [@Brawand:2011du].
-For each internal node in each gene tree (Figure 1b), we measured the phylogenetic independent contrast [@Felsenstein:1985ua] (PIC) of the expression statistic Tau.
+For each internal node in each gene tree (Figure 1b), we measured the phylogenetic independent contrast (PIC) [@Felsenstein:1985ua] of the expression statistic Tau.
 This is the difference in values of Tau for descendant nodes scaled by the expected variance (determined largely by the lengths of these branches).

### 2017-01-30 manuscript.rmd
Update manuscript.rmd
Minor edits

@@
 Instead, we find no such distribution (Figure XXKMRR).
-The Wilcox rank test does not reject the null hypothesis that the rate of evolution following duplications is the same as or less than the rate following speciation (p value = `r wilcox_test_result`).
+The Wilcoxon rank test does not reject the null hypothesis that the rate of evolution following duplications is the same as or less than the rate following speciation (p value = `r wilcox_test_result`).
 Our results therefore find no evidence for the ortholog conjecture in this system.
@@
 
-In order to better understand why our phylogenetic analysis seems to support a different conclusion (*i.e.*, no support for the ortholog conjecture) than the published analysis of KMRR [@KryuchkovaMostacci:2016iw] (*i.e.*, strong support for the ortholog conjecture), we first checked to make sure we could reproduce their result based on pairwise analyses.
+In order to better understand why our phylogenetic analysis supports a different conclusion (*i.e.*, no support for the ortholog conjecture) than the published analysis of KMRR [@KryuchkovaMostacci:2016iw] (*i.e.*, strong support for the ortholog conjecture), we first checked to make sure we could reproduce their result based on pairwise analyses.
 This is important since we are only looking at at a subset of the data they considered, the Brawand *et al.* 2011 [@Brawand:2011du] dataset for gene trees that could be successfully time calibrated.
@@
 The reason is that they describe different things.
-Pairwise comparisons describe contemporary patterns, while phylogenetic methods infer historical processes.
+Pairwise comparisons describe contemporary patterns, while phylogenetic methods infer historical processes [@OMeara:2012gq].
 There need not be a different process of evolution following speciation and duplication for paralogs to be more different than orthologs.
@@
 While the mean distance between orthologs is `r round( ortholog_distance_mean, 1 )` My, the mean distance between paralogs is `r round( paralog_distance_mean, 1 )` My.
-As described above, this is because the oldest speciation event is by definition the most recent common ancestor of the the species included in the study, but many gene families underwent duplication before this time.
+As described above, this is because the oldest speciation event is by definition the most recent common ancestor of the species included in the study, but many gene families underwent duplication before this time.
 
@@
 The ortholog conjecture does not have to be an all or nothing question.
-It may be the case that the rates of phenotypic evolution following duplication may be greater than that following duplication in some organisms, gene families, and evolutionary processes.
+It may be the case that the rates of phenotypic evolution following duplication may be greater than that following duplication in some organisms, gene families, and evolutionary processes [@Gabaldon:2013fi].
 We just don't find evidence for it when summarizing gene expression across tissues with Tau in these mammals.
@@
 We are not the first to apply phylogenetic comparative methods to functional genomic data.
-While the vast majority of comparative functional genomics studies have used pairwise similarity methods, a small number of CFG studies have employed phylogenetic comparative approaches [@Oakley:2005ky; @Eng:2009gp; @Chang:2014hq].
+While the vast majority of comparative functional genomics studies have used standard pairwise similarity methods, a small number of CFG studies have employed phylogenetic comparative approaches [@Oakley:2005ky; @Eng:2009gp; @Chang:2014hq].
 For instance, a phylogenetic ANOVA [@Rohlfs:2015bd] of the evolution of gene expression improves statistical power and drastically reduces the rate of false positives relative to pairwise approaches.

### 2017-01-31 manuscript.rmd
incorporated multiple changes from co-authors

@@
 A rich theoretical and statistical methodology of phylogenetic comparative methods have been developed over the last three decades to address the challenges and opportunities of comparisons across species [@Felsenstein:1985ua; @Grafen:1989um; @Pagel:1999gc; Revell:2012jd; @FitzJohn:2012ey; @Uyeda:2014jl; @Garamszegi:2014ew].
-These challenges include non-independence of observations across species -- more closely related species will tend to be more similar to each other than distantly related species.
+A central challenge is the dependence of observations across species due to the evolutionary history of species -- more closely related species share many traits that evolve once in a common ancestor.
+This violates the fundamental assumption of independence of observations in standard statistical methods.
 Phylogenetic comparative methods have largely been applied to the evolution of morphological and ecological traits, but are just as relevant to CFG.
@@
 This study analyzed gene expression through the course of embryonic development for ten animal species, each from a different phylum.
-They concluded that these animals exhibit an "inverse hourglass" model for the evolution of gene expression, where there is more evolutionary variance in gene expression at a mid phase of development than there is at early and late phases (Fig.
-2).
-This contrasts with the "hourglass" model previously proposed for closely related species [@Kalinka:2010dw].
-Levin *et al.* conclude that this provides biological justification for the concept of phyla and may provide a definition of phyla.
-We previously described concerns with their interpretations of these results [@Hejnol:2016hw].
+Using pairwise comparisons (Figure 1A), they found there is more evolutionary variance in gene expression at a mid phase of development than there is at early and late phases for these ten animals (Fig.
+XX).
+They suggest that this supports an "inverse hourglass" model for the evolution of gene expression, in contrast with the "hourglass" model previously proposed for closely related species [@Kalinka:2010dw], and provides biological justification for the concept of phyla and may provide an operational definition of phyla.
+We previously described concerns with these interpretations [@Hejnol:2016hw].
 Here we directly address the analyses themselves by examining the structure of the pairwise comparisons.
@@
 (a) Many comparative functional genomic studies rely on pairwise comparisons, where traits of each gene are compared to traits of other genes (such as orthologs and paralogs).
-This leads to many more comparisons than unique observations, making each comparison non-independent.
+This leads to many more comparisons than unique observations, making each comparison dependent on others.
 (b) Comparative phylogenetic methods, including phylogenetic independent contrasts [@Felsenstein:1985ua], make a smaller number of independent comparisons, where each contrast measures independent changes along different branches.
@@
 Results in KMRR are based on pairwise comparisons (Figure 1A) between Tau within each gene family.
-They found the correlation of Tau for orthologs to be significantly greater than the correlation of Tau for paralogs, *i.e.* that orthologs tend to be more similar to each other than paralogs.
+They found the correlation coefficient of Tau for orthologs to be significantly greater than the correlation coefficient of Tau for paralogs, *i.e.* that orthologs tend to be more similar to each other than paralogs.
 From this they concluded that their analyses support the ortholog conjecture.
@@
 
-When we reanalyze the data with phylogenetic independent contrasts (Fig.
-1B), we did not find increased evolutionary change in expression following duplication events (Fig.
+When we reanalyze the data with phylogenetic independent contrasts, we did not find increased evolutionary change in expression following duplication events (Fig.
 XXKMRR).
@@
 > Figure XX_KMRR_Pairwise | This figures demonstrates that the pairwise analyses presented in Figure 2a of KMRR can be reproduced with the subset of data we consider here and our own independent code base.
-Each circle indicates the Tau Pearson correlation (rho) for a set of pairwise comparisons annotated with a specific node name of a given age and event type (*i.e.*, whether the divergence at that nodes was due to speciation or duplication).
+Each circle indicates the Tau Pearson correlation (rho) for a set of pairwise comparisons annotated with a specific node name of a given age and event type, *i.e.*, whether the divergence at that nodes was due to speciation (giving rise to orthologs) or duplication (giving rise to paralogs).
 Following KMRR Figure 2a, this figure shows only the comparisons that include a human sequence.
@@
 
-To test the hypothesis that distantly related paralogs arising from these older duplication events are the reason for the lower correlation of Tau between paralogs than between orthologs, we removed them.
-When we consider only the paralogs that arise from duplication events the same age or younger than the oldest speciation event, the observed pattern is very different.
+To test the hypothesis that distantly related paralogs arising from these older duplication events (Supplementary Figure XXAscertainment a) are the reason for the lower correlation of Tau between paralogs than between orthologs, we removed them.
+When we consider only the paralogs that arise from duplication events the same age or younger than the oldest speciation event (Supplementary Figure XXAscertainment b), the observed pattern is very different.
 The correlation of Tau between these shallower paralog pairs, R=`r round( paralog_subset_r, 3)`, is much stronger than for the correlation between all paralog pairs, R=`r round( paralog_r, 3)`.
@@
 This highlights a well understood property of pairwise comparisons across species [@Felsenstein:1985ua]: evolutionary changes along a given branch, like those along the ctenophore branch, impact each of the multiple pairwise comparisons that includes that branch.
-The pairwise comparisons are therefore not independent - different pairwise comparisons are impacted by canges along some of the same branches.
+The pairwise comparisons are therefore not independent - different pairwise comparisons are impacted by canges along some of the same branches (Figure 1a).
 This can give the impression of a general pattern across the tree that is instead specific to changes along one part of the tree.
@@
 
-Addressing the statistical non-independence of pairwise comparisons is not the only advantage of using phylogenetic comparative methods for functional genomics analyses.
+Addressing the statistical dependence of pairwise comparisons is not the only advantage of using phylogenetic comparative methods for functional genomics analyses.
 Another problem with the pairwise comparisons is that, except at the tips, they summarize changes along many branches in the phylogeny.

### 2017-02-01 manuscript.rmd
Update manuscript.rmd
Adds changes suggested today on Levin.

@@
 
-Levin *et al.* [@Levin:2016bp] arrived at this conclusion by making pairwise comparisons of ortholog expression through development between species.
-They characterized each gene in each species as having expression that peaks in early, mid, or late temporal phase of development.
-The goodness of fit to these patterns was not considered, and alternative patterns, such as the possibility that there are more phases of expression, were not evaluated.
+Levin *et al.* [@Levin:2016bp] arrived at this conclusion by making multiple pairwise comparisons of ortholog expression through development between each pair of species.
 For each species pair, they identified the orthologs shared by these species.
 This list of shared genes was different from species pair to species pair.
+They characterized each gene in each species as having expression that peaks in early, mid, or late temporal phase of development.
 They then calculated a similarity score for each temporal phase for each species pair based on the fraction of genes that exhibited the same patterns in each species.
@@
 
-We previously described some concerns with their interpretations of these results [@Hejnol:2016hw].
-The main concern was that because they only sampled one species per phylum, there is no way to tell if observed differences are unique to the branches that gave rise to clades designated as phyla or occurred along the many other branches in the tree.
-Here we examine the support for the inverse hourglass model itself.
-We focus on the matrix of pairwise comparisons that their KS tests and their Figure 4d are based on.
-
-The first problems we identified were general stastistical issues that are not problems with pairwise comparisons *per se*.
-First, in the published analyses every data point was included twice because both reciprocal comparisons (which have the same values) were retained.
+We examined the matrix of pairwise comparisons used as the base for the KS tests and Figure 4d in *Levin et al.* [@Levin:2016bp], and thus as evidence to support the "inverse hourglass model".
+We found several problems resulting from the use of multiple pairwise comparisons.
+First, we found that every data point was included twice because both reciprocal pairwise comparisons (which have the same values) were retained.
 For example, there is both a nematode to arthropod comparison and an arthropod to nematode comparison.
@@
 After removing the duplicate values, the p values are far less significant, 0.002 for the early-mid comparison and on the order of 10<sup>-6</sup> for early-late.
-Second, the test they used (KS test) is not appropriate for the hypothesis they seek to evaluate.
+In addition, the test they used (KS test) is not appropriate for the hypothesis they seek to evaluate.
 The KS test does not just evaluate whether one distribution is greater than the other, it also tests whether the shape of the distributions are the same.
@@
 
-The next issue we found is a general problem with pairwise comparisons.
-We found that all five of the lowest values in the mid phase distribution (Figure XXLevin a) are for pairwise comparisons that include the ctenophore (comb jelly).
+The next issue we found was that all five of the lowest values in the mid phase distribution (Figure XXLevin a) are for pairwise comparisons that include the ctenophore (comb jelly).
 When the nine pairwise comparisons that include the ctenophore are removed and the general statistical issues noted above are addressed, there is no significant difference between the early phase and mid phase distributions (P = 0.1428 for the early-mid comparison and P < 10<sup>-5</sup> for the late-mid comparison).
 The inverse hourglass turns into a bottle (Figure XXLevin b).
-
 This highlights a well understood property of pairwise comparisons across species [@Felsenstein:1985ua]: evolutionary changes along a given branch, like those along the ctenophore branch, impact each of the multiple pairwise comparisons that includes that branch.
@@
 This can give the impression of a general pattern across the tree that is instead specific to changes along one part of the tree.
-The number of comparisons impacted by each change depends on the structure of the phylogenetic tree, *i.e.* how the species are related to each other.
+The number of comparisons impacted by each change depends on the structure of the phylogenetic tree, i.e.
+how the species are related to each other.
 
 Phylogenetic comparative methods were developed specifically to address this problem [@Felsenstein:1985ua].
-While we do demonstrate this problem impacts the Levin *et al.* analysis, we did not perform a phylogenetic reanalysis of this study, as we did for the KMRR study.
-This is because the similarity metric computed in the pairwise comparisons of Levin *et al.* are based on different genes for different species pairs.
+While we do demonstrate this problem impacts the Levin et al.
+analysis, we did not perform a phylogenetic reanalysis of this study, as we did for the KMRR study.
+This is because the similarity metric computed in the pairwise comparisons of Levin et al.
+are based on different genes for different species pairs.
 This means that they do not generate a trait that can be evaluated across a phylogeny, as KMRR's Tau can be.
 A full phylogenetic reanalysis would be possible by using upstream analysis products to re-derive new expression summary statistics suitable for phylogenetic comparative methods.
+
+In addition to the problems with the pairwise comparisons discussed above, we previously described several other concerns with the experimental design, analysis, and interpretation of the results of this study [@Hejnol:2016hw].
+For instance, because Levin *et al.* [@Levin:2016bp] only sampled one species per phylum, there is no way to tell if observed differences are unique to the branches that gave rise to clades designated as phyla or occurred along the many other branches in the tree.
+Moreover, the goodness of fit of each gene having expression peaking at three phases was not considered, and alternative patterns, such as the possibility that there are more phases of expression, were not evaluated.
 

### 2017-02-01 manuscript.rmd
bib fix
added null simulation

--- 2017-01-31 manuscript.rmd
@@
 	library( gridExtra )
+	library( geiger )
+	library( phytools )
 
@@
 
+		tau = nhx@data$Tau[ is.tip.nhx( nhx ) ]
+
+		if( any( is.na(tau) ) ){
+			stop( "A trait value is NA." )
+		}
+
 		# Calculate the contrasts
-		p = pic( nhx@data$Tau[ is.tip.nhx( nhx ) ], nhx@phylo, var.contrasts=TRUE )
+		p = pic( tau, nhx@phylo, var.contrasts=TRUE )
 
@@
 
-		# Make a new human readable column for plots that records speciation/ duplication
-		nodes_all$Event = NA
-		nodes_all$Event[ nodes_all$D == "Y" ] = "Duplication"
-		nodes_all$Event[ nodes_all$D == "N" ] = "Speciation"
-
 		# Create a subset that corresponds to nodes with contrasts. This excludes tips.
@@
 			tree@data %<>% left_join( expression, by = c( "G" = "Ensembl.Gene.ID" ) )
+
+			tree@data$Event = NA
+			tree@data$Event[ tree@data$D == "N" ] = "Speciation"
+			tree@data$Event[ tree@data$D == "Y" ] = "Duplication"
+
 			return( tree )
@@
 			nhx@data %>%
-				select( D, node, node_age, label, pic ),
+				select( D, Event, node, node_age, label, pic ),
 			by = c( "mrca" = "node")
@@
 		bind_rows()
-
-	pairwise_summary$Event = NA
-	pairwise_summary$Event[ pairwise_summary$D == "N" ] = "Speciation"
-	pairwise_summary$Event[ pairwise_summary$D == "Y" ] = "Duplication"
 
@@
 
-				nhx@data$Event = NA
-				nhx@data$Event[ nhx@data$D == "Y" ] = "Duplication"
-				nhx@data$Event[ nhx@data$D == "N" ] = "Speciation"
-
 				nhx %>% 
@@
 
+#### Simulations
+
+```{r simulations_null, cache=TRUE, echo=FALSE, fig.cap="XX_PIC_VAR", warning=FALSE, message=FALSE}
+
+	sim_tau = function( nhx ) {
+		tau_original = nhx@data$Tau [ 1:length(nhx@phylo$tip.label) ]
+		names( tau_original ) = nhx@phylo$tip.label
+		brownian_model = fitContinuous( nhx@phylo, tau_original )
+		x = fastBM( 
+				nhx@phylo, 
+				a=brownian_model$opt$z0, 
+				sig2=brownian_model$opt$sigsq, 
+				bounds=c(-Inf,Inf) 
+			) %>% abs()
+		x[ x > 1 ] = 1
+		x[ is.nan( x ) ] = 1
+		names( x ) = NULL
+		nhx@data$Tau = c( x, rep( NA, nrow( nhx@data ) - length (x ) ) )
+		return( nhx )
+	}
+
+	# Replace the actual Tau values with simulated Tau values, using parameters 
+	# fitted from each tree
+	gene_trees_sim_null = lapply( gene_trees_calibrated[1:1000], sim_tau )
+
+	# Perform contrasts 
+	gene_trees_sim_null %<>% add_pics_to_trees()
+
+	# Summarize results
+	nodes_sim_null_contrasts = gene_trees_sim_null %>% summarize_contrasts()
+
+	# Plot
+	g_phylo_null = 
+		nodes_sim_null_contrasts %>% 
+		ggplot( aes( x=pic, y=..density.., col=Event ) ) + 
+			geom_freqpoly( binwidth=.001, position="identity" ) +
+			xlab( "Phylogenetic independent \ncontrasts (PIC)" ) +
+			ylab( "Density" ) +
+			xlim( 0, 0.075 ) + 
+			theme_classic() + 
+			ggtitle( "(b) Phylogenetic result under null simulation" )
+
+	pairwise_sim_null_summary = 
+		mclapply( gene_trees_sim_null, get_pairwise_summary, mc.cores=cores ) %>% 
+		bind_rows()
+
+	# Calculate some summary statistics 
+	ortholog_r_sim_null = 
+		pairwise_sim_null_summary %>% 
+		filter( D=="N" ) %>% 
+		select( tau_a, tau_b ) %>% 
+		cor( ) %>% 
+		.[ 1, 2 ]
+
+	paralog_r_sim_null = 
+		pairwise_sim_null_summary %>% 
+		filter( D=="Y" ) %>% 
+		select( tau_a, tau_b ) %>% 
+		cor( ) %>% 
+		.[ 1, 2 ]
+
+	g_pair_null = 
+		pairwise_sim_null_summary %>% 
+		filter( species_a == "homo_sapiens" | species_b == "homo_sapiens" ) %>% 
+		filter( ! is.na( D ) ) %>%
+		group_by( Event, label ) %>% 
+		summarise( cor.test(  tau_a, tau_b )$estimate, n=n() ) %>% 
+		rename( rho=`cor.test(tau_a, tau_b)$estimate` ) %>%
+		rename( clade = label ) %>% 
+		left_join( calibration_times_all, clade=clade ) %>%
+		ggplot() + 
+			geom_point( aes( x=age, y=rho, col=Event, size=n ) ) +
+			theme_classic() +
+			ggtitle( "(a) Pairwise result under null simulation" )
+
+```
+
+```{r simulations_plot, cache=TRUE, echo=FALSE, fig.cap="XX_PIC_VAR", warning=FALSE, message=FALSE}
+
+	grid.arrange(
+		g_pair_null,
+		g_phylo_null,
+		ncol = 2
+	)
+
+```
+
+> Figure XXSim
+
+We first simulated an analysis under the null model that tau evolves at the same rate following duplication and speciation events.
+Under the null model, pairwise approaches are very similar as for the observed data (Figure XXSim a), with higher rho for orthologs (`r ortholog_r_sim_null`) than paralogs (`r paralog_r_sim_null`).
+Phylogenetic approaches do show similar divergence following speciation and duplication (Figure XXSim b) and does not reject the null hypothesis that the rates differ (Wilcoxon p=`r wilcox_oc( nodes_sim_null_contrasts )`).
+
 ## References

### 2017-02-02 manuscript.rmd
added oc simulation

@@
 
-	sim_tau = function( nhx ) {
-		tau_original = nhx@data$Tau [ 1:length(nhx@phylo$tip.label) ]
-		names( tau_original ) = nhx@phylo$tip.label
-		brownian_model = fitContinuous( nhx@phylo, tau_original )
+	sim_tau = function( nhx, dup_adjust=1 ) {
+
+		phy = nhx@phylo
+
+		# Estimate parameters
+		tau_original = nhx@data$Tau [ 1:length( phy$tip.label ) ]
+		names( tau_original ) = phy$tip.label
+		brownian_model = fitContinuous( phy, tau_original )
+
+		# Adjust length of branches that descend from duplication events. 
+		# This is how we implement heterogeneous rates after duplication 
+		# rather than speciation. Extending the branch lengths is equivalent 
+		# to using a higher rate along them.
+		dup_nodes = nhx@data$node[ nhx@data$D == "Y" ]
+		dup_edges = phy$edge[,1] %in% dup_nodes
+		phy$edge.length[ dup_edges ] = phy$edge.length[ dup_edges ] * dup_adjust
+
+		# Simulate trait given the tree and parameter estimates
 		x = fastBM( 
-				nhx@phylo, 
+				phy, 
 				a=brownian_model$opt$z0, 
@@
 			) %>% abs()
+
+		# Adjust the results so they fall between 0 and 1
 		x[ x > 1 ] = 1
 		x[ is.nan( x ) ] = 1
+
 		names( x ) = NULL
@@
 	gene_trees_sim_null %<>% add_pics_to_trees()
-
-	# Summarize results
 	nodes_sim_null_contrasts = gene_trees_sim_null %>% summarize_contrasts()
 
-	# Plot
+	# Calculate pairwise comparisons
+	pairwise_sim_null_summary = 
+		mclapply( gene_trees_sim_null, get_pairwise_summary, mc.cores=cores ) %>% 
+		bind_rows()
+
+	# Calculate some summary statistics 
+	ortholog_r_sim_null = 
+		pairwise_sim_null_summary %>% 
+		filter( D=="N" ) %>% 
+		select( tau_a, tau_b ) %>% 
+		cor( ) %>% 
+		.[ 1, 2 ]
+
+	paralog_r_sim_null = 
+		pairwise_sim_null_summary %>% 
+		filter( D=="Y" ) %>% 
+		select( tau_a, tau_b ) %>% 
+		cor( ) %>% 
+		.[ 1, 2 ]
+
+```
+
+```{r simulations_oc, cache=TRUE, echo=FALSE, fig.cap="XX_PIC_VAR", warning=FALSE, message=FALSE}
+
+	dup_adjust = 2
+
+	# Replace the actual Tau values with simulated Tau values, using parameters 
+	# fitted from each tree. 
+	gene_trees_sim_oc = lapply( gene_trees_calibrated[1:1000], sim_tau, dup_adjust=dup_adjust )
+
+	# Calculate contrasts 
+	gene_trees_sim_oc %<>% add_pics_to_trees()
+	nodes_sim_oc_contrasts = gene_trees_sim_oc %>% summarize_contrasts()
+
+	# Calculate pairwise comparisons 
+	pairwise_sim_oc_summary = 
+		mclapply( gene_trees_sim_oc, get_pairwise_summary, mc.cores=cores ) %>% 
+		bind_rows()
+
+	ortholog_r_sim_oc = 
+		pairwise_sim_oc_summary %>% 
+		filter( D=="N" ) %>% 
+		select( tau_a, tau_b ) %>% 
+		cor( ) %>% 
+		.[ 1, 2 ]
+
+	paralog_r_sim_oc = 
+		pairwise_sim_oc_summary %>% 
+		filter( D=="Y" ) %>% 
+		select( tau_a, tau_b ) %>% 
+		cor( ) %>% 
+		.[ 1, 2 ]
+
+```
+
+```{r simulations_plot, cache=TRUE, echo=FALSE, fig.cap="XX_PIC_VAR", warning=FALSE, message=FALSE}
+
 	g_phylo_null = 
@@
 			xlim( 0, 0.075 ) + 
+			ylim( 0, 100 ) +
 			theme_classic() + 
 			ggtitle( "(b) Phylogenetic result under null simulation" )
-
-	pairwise_sim_null_summary = 
-		mclapply( gene_trees_sim_null, get_pairwise_summary, mc.cores=cores ) %>% 
-		bind_rows()
-
-	# Calculate some summary statistics 
-	ortholog_r_sim_null = 
-		pairwise_sim_null_summary %>% 
-		filter( D=="N" ) %>% 
-		select( tau_a, tau_b ) %>% 
-		cor( ) %>% 
-		.[ 1, 2 ]
-
-	paralog_r_sim_null = 
-		pairwise_sim_null_summary %>% 
-		filter( D=="Y" ) %>% 
-		select( tau_a, tau_b ) %>% 
-		cor( ) %>% 
-		.[ 1, 2 ]
 
@@
 
-```
-
-```{r simulations_plot, cache=TRUE, echo=FALSE, fig.cap="XX_PIC_VAR", warning=FALSE, message=FALSE}
+	g_phylo_oc = 
+	nodes_sim_oc_contrasts %>% 
+	ggplot( aes( x=pic, y=..density.., col=Event ) ) + 
+		geom_freqpoly( binwidth=.001, position="identity" ) +
+		xlab( "Phylogenetic independent \ncontrasts (PIC)" ) +
+		ylab( "Density" ) +
+		xlim( 0, 0.075 ) + 
+		ylim( 0, 100 ) +
+		theme_classic() + 
+		ggtitle( "(d) Phylogenetic result under OC simulation" )
+
+	g_pair_oc = 
+		pairwise_sim_oc_summary %>% 
+		filter( species_a == "homo_sapiens" | species_b == "homo_sapiens" ) %>% 
+		filter( ! is.na( D ) ) %>%
+		group_by( Event, label ) %>% 
+		summarise( cor.test(  tau_a, tau_b )$estimate, n=n() ) %>% 
+		rename( rho=`cor.test(tau_a, tau_b)$estimate` ) %>%
+		rename( clade = label ) %>% 
+		left_join( calibration_times_all, clade=clade ) %>%
+		ggplot() + 
+			geom_point( aes( x=age, y=rho, col=Event, size=n ) ) +
+			theme_classic() +
+			ggtitle( "(c) Pairwise result under OC simulation" )
 
@@
 		g_phylo_null,
+		g_pair_oc,
+		g_phylo_oc,
 		ncol = 2
@@
 
-We first simulated an analysis under the null model that tau evolves at the same rate following duplication and speciation events.
-Under the null model, pairwise approaches are very similar as for the observed data (Figure XXSim a), with higher rho for orthologs (`r ortholog_r_sim_null`) than paralogs (`r paralog_r_sim_null`).
-Phylogenetic approaches do show similar divergence following speciation and duplication (Figure XXSim b) and does not reject the null hypothesis that the rates differ (Wilcoxon p=`r wilcox_oc( nodes_sim_null_contrasts )`).
+We first simulated a data set under the null model that tau evolves at the same rate following duplication and speciation events.
+Under the null model, pairwise analysis results (Figure XXSim a) are very similar as for the observed data, with higher rho for orthologs (`r ortholog_r_sim_null`) than paralogs (`r paralog_r_sim_null`).
+In the phylogenetic analysis of the null simulation, contrasts show similar divergence following speciation and duplication (Figure XXSim b) and does not reject the null hypothesis (Wilcoxon p=`r wilcox_oc( nodes_sim_null_contrasts )`).
+
+We next simulated a data set under the ortholog conjecture, where the rate of evolution of Tau following duplication was `r dup_adjust` fold the rate following speciation.
+The results of this heterogeneous model are nearly indistinguishable from the null model  (Figure XXSim c), and also have a higher rho for orthologs (`r ortholog_r_sim_oc`) than paralogs (`r paralog_r_sim_oc`).
+In the phylogenetic analysis of the ortholog conjecture simulation, contrasts XX (Figure XXSim c) and XX (Wilcoxon p=`r wilcox_oc( nodes_sim_oc_contrasts )`).
 

### 2017-02-02 manuscript.rmd
moved simulations to man text, re knit from scratch

@@
 
+	#' Simulate Tau on a tree::treedata. Existing observed values are
+	#' fit to a brownian model to estimate parameters. These parameters
+	#' are then used to replace the original values with similated 
+	#' values.
+	#' 
+	#' @param nhx A phylogenetic tree and associated Tau values 
+	#' as a treeio::treedata object
+	#' @param dup_adjust A multiplier for adjusting the branch lengths 
+	#' following duplication to effectively change the rate following duplication
+	#' @return A phylogenetic trees and simulated Tau values as a 
+	#' treeio::treedata object
+	#' @export
+	sim_tau = function( nhx, dup_adjust=1 ) {
+
+		phy = nhx@phylo
+
+		# Estimate parameters
+		tau_original = nhx@data$Tau [ 1:length( phy$tip.label ) ]
+		names( tau_original ) = phy$tip.label
+		brownian_model = fitContinuous( phy, tau_original )
+
+		# Adjust length of branches that descend from duplication events. 
+		# This is how we implement heterogeneous rates after duplication 
+		# rather than speciation. Extending the branch lengths is equivalent 
+		# to using a higher rate along them.
+		dup_nodes = nhx@data$node[ nhx@data$D == "Y" ]
+		dup_edges = phy$edge[,1] %in% dup_nodes
+		phy$edge.length[ dup_edges ] = phy$edge.length[ dup_edges ] * dup_adjust
+
+		# Simulate trait given the tree and parameter estimates
+		x = fastBM( 
+				phy, 
+				a=brownian_model$opt$z0, 
+				sig2=brownian_model$opt$sigsq, 
+				bounds=c(-Inf,Inf) 
+			) %>% abs()
+
+		# Adjust the results so they fall between 0 and 1
+		x[ x > 1 ] = 1
+		x[ is.nan( x ) ] = 1
+
+		names( x ) = NULL
+		nhx@data$Tau = c( x, rep( NA, nrow( nhx@data ) - length (x ) ) )
+		return( nhx )
+	}
+
 ```
@@
 			tree@data$Event[ tree@data$D == "Y" ] = "Duplication"
+			tree@data$Event = factor( tree@data$Event, levels=c( "Speciation", "Duplication" ) )
 
@@
 In this case, KMRR [@KryuchkovaMostacci:2016iw] relied on orthology and paralogy annotations that are derived from the inferred duplication and speciation history on gene phylogenies [@Herrero:2016jj].
+
+```{r simulations_null, cache=TRUE, echo=FALSE, fig.cap="XX_PIC_VAR", warning=FALSE, message=FALSE}
+
+	# Replace the actual Tau values with simulated Tau values, using parameters 
+	# fitted from each tree
+	gene_trees_sim_null = lapply( gene_trees_calibrated[1:1000], sim_tau )
+
+	# Perform contrasts 
+	gene_trees_sim_null %<>% add_pics_to_trees()
+	nodes_sim_null_contrasts = gene_trees_sim_null %>% summarize_contrasts()
+
+	# Calculate pairwise comparisons
+	pairwise_sim_null_summary = 
+		mclapply( gene_trees_sim_null, get_pairwise_summary, mc.cores=cores ) %>% 
+		bind_rows()
+
+	# Calculate some summary statistics 
+	ortholog_r_sim_null = 
+		pairwise_sim_null_summary %>% 
+		filter( D=="N" ) %>% 
+		select( tau_a, tau_b ) %>% 
+		cor( ) %>% 
+		.[ 1, 2 ]
+
+	paralog_r_sim_null = 
+		pairwise_sim_null_summary %>% 
+		filter( D=="Y" ) %>% 
+		select( tau_a, tau_b ) %>% 
+		cor( ) %>% 
+		.[ 1, 2 ]
+
+```
+
+```{r simulations_oc, cache=TRUE, echo=FALSE, fig.cap="XX_PIC_VAR", warning=FALSE, message=FALSE}
+
+	dup_adjust = 2
+
+	# Replace the actual Tau values with simulated Tau values, using parameters 
+	# fitted from each tree. 
+	gene_trees_sim_oc = lapply( gene_trees_calibrated[1:1000], sim_tau, dup_adjust=dup_adjust )
+
+	# Calculate contrasts 
+	gene_trees_sim_oc %<>% add_pics_to_trees()
+	nodes_sim_oc_contrasts = gene_trees_sim_oc %>% summarize_contrasts()
+
+	# Calculate pairwise comparisons 
+	pairwise_sim_oc_summary = 
+		mclapply( gene_trees_sim_oc, get_pairwise_summary, mc.cores=cores ) %>% 
+		bind_rows()
+
+	ortholog_r_sim_oc = 
+		pairwise_sim_oc_summary %>% 
+		filter( D=="N" ) %>% 
+		select( tau_a, tau_b ) %>% 
+		cor( ) %>% 
+		.[ 1, 2 ]
+
+	paralog_r_sim_oc = 
+		pairwise_sim_oc_summary %>% 
+		filter( D=="Y" ) %>% 
+		select( tau_a, tau_b ) %>% 
+		cor( ) %>% 
+		.[ 1, 2 ]
+
+```
+
+```{r simulations_plot, cache=TRUE, echo=FALSE, fig.cap="XX_PIC_VAR", warning=FALSE, message=FALSE}
+
+	g_phylo_null = 
+		nodes_sim_null_contrasts %>% 
+		ggplot( aes( x=pic, y=..density.., col=Event ) ) + 
+			geom_freqpoly( binwidth=.001, position="identity" ) +
+			xlab( "Phylogenetic independent \ncontrasts (PIC)" ) +
+			ylab( "Density" ) +
+			xlim( 0, 0.075 ) + 
+			ylim( 0, 100 ) +
+			theme_classic() + 
+			ggtitle( "(b) Phylogenetic result under null simulation" )
+
+	g_pair_null = 
+		pairwise_sim_null_summary %>% 
+		filter( species_a == "homo_sapiens" | species_b == "homo_sapiens" ) %>% 
+		filter( ! is.na( D ) ) %>%
+		group_by( Event, label ) %>% 
+		summarise( cor.test(  tau_a, tau_b )$estimate, n=n() ) %>% 
+		rename( rho=`cor.test(tau_a, tau_b)$estimate` ) %>%
+		rename( clade = label ) %>% 
+		left_join( calibration_times_all, clade=clade ) %>%
+		ggplot() + 
+			geom_point( aes( x=age, y=rho, col=Event, size=n ) ) +
+			theme_classic() +
+			ggtitle( "(a) Pairwise result under null simulation" )
+
+	g_phylo_oc = 
+	nodes_sim_oc_contrasts %>% 
+	ggplot( aes( x=pic, y=..density.., col=Event ) ) + 
+		geom_freqpoly( binwidth=.001, position="identity" ) +
+		xlab( "Phylogenetic independent \ncontrasts (PIC)" ) +
+		ylab( "Density" ) +
+		xlim( 0, 0.075 ) + 
+		ylim( 0, 100 ) +
+		theme_classic() + 
+		ggtitle( "(d) Phylogenetic result under OC simulation" )
+
+	g_pair_oc = 
+		pairwise_sim_oc_summary %>% 
+		filter( species_a == "homo_sapiens" | species_b == "homo_sapiens" ) %>% 
+		filter( ! is.na( D ) ) %>%
+		group_by( Event, label ) %>% 
+		summarise( cor.test(  tau_a, tau_b )$estimate, n=n() ) %>% 
+		rename( rho=`cor.test(tau_a, tau_b)$estimate` ) %>%
+		rename( clade = label ) %>% 
+		left_join( calibration_times_all, clade=clade ) %>%
+		ggplot() + 
+			geom_point( aes( x=age, y=rho, col=Event, size=n ) ) +
+			theme_classic() +
+			ggtitle( "(c) Pairwise result under OC simulation" )
+
+	grid.arrange(
+		g_pair_null,
+		g_phylo_null,
+		g_pair_oc,
+		g_phylo_oc,
+		ncol = 2
+	)
+
+```
+
+> Figure XXSim
+
+We simulated a data set under the null model that tau evolves at the same rate following duplication and speciation events.
+Under the null model, pairwise analysis results (Figure XXSim a) are very similar as for the observed data, with higher rho for orthologs (`r ortholog_r_sim_null`) than paralogs (`r paralog_r_sim_null`).
+In the phylogenetic analysis of the null simulation, contrasts show similar divergence following speciation and duplication (Figure XXSim b) and does not reject the null hypothesis (Wilcoxon p=`r wilcox_oc( nodes_sim_null_contrasts )`).
+
+We next simulated a data set under the ortholog conjecture, where the rate of evolution of Tau following duplication was `r dup_adjust` fold the rate following speciation.
+The results of this heterogeneous model are nearly indistinguishable from the null model  (Figure XXSim c), and also have a higher rho for orthologs (`r ortholog_r_sim_oc`) than paralogs (`r paralog_r_sim_oc`).
+In the phylogenetic analysis of the ortholog conjecture simulation, contrasts XX (Figure XXSim c) and XX (Wilcoxon p=`r wilcox_oc( nodes_sim_oc_contrasts )`).
 
@@
 
-We examined the ensitivity of our results to the specification of calibration dates for the speciation nodes.
+We examined the sensitivity of our results to the specification of calibration dates for the speciation nodes.
 In `r noised_replicate_n` reanalyses, we drew a new date for each calibration from a normal distribution with the mean of the original date and a standard deviation `r sd_fraction` times the original date.
@@
 
-#### Simulations
-
-```{r simulations_null, cache=TRUE, echo=FALSE, fig.cap="XX_PIC_VAR", warning=FALSE, message=FALSE}
-
-	sim_tau = function( nhx, dup_adjust=1 ) {
-
-		phy = nhx@phylo
-
-		# Estimate parameters
-		tau_original = nhx@data$Tau [ 1:length( phy$tip.label ) ]
-		names( tau_original ) = phy$tip.label
-		brownian_model = fitContinuous( phy, tau_original )
-
-		# Adjust length of branches that descend from duplication events. 
-		# This is how we implement heterogeneous rates after duplication 
-		# rather than speciation. Extending the branch lengths is equivalent 
-		# to using a higher rate along them.
-		dup_nodes = nhx@data$node[ nhx@data$D == "Y" ]
-		dup_edges = phy$edge[,1] %in% dup_nodes
-		phy$edge.length[ dup_edges ] = phy$edge.length[ dup_edges ] * dup_adjust
-
-		# Simulate trait given the tree and parameter estimates
-		x = fastBM( 
-				phy, 
-				a=brownian_model$opt$z0, 
-				sig2=brownian_model$opt$sigsq, 
-				bounds=c(-Inf,Inf) 
-			) %>% abs()
-
-		# Adjust the results so they fall between 0 and 1
-		x[ x > 1 ] = 1
-		x[ is.nan( x ) ] = 1
-
-		names( x ) = NULL
-		nhx@data$Tau = c( x, rep( NA, nrow( nhx@data ) - length (x ) ) )
-		return( nhx )
-	}
-
-	# Replace the actual Tau values with simulated Tau values, using parameters 
-	# fitted from each tree
-	gene_trees_sim_null = lapply( gene_trees_calibrated[1:1000], sim_tau )
-
-	# Perform contrasts 
-	gene_trees_sim_null %<>% add_pics_to_trees()
-	nodes_sim_null_contrasts = gene_trees_sim_null %>% summarize_contrasts()
-
-	# Calculate pairwise comparisons
-	pairwise_sim_null_summary = 
-		mclapply( gene_trees_sim_null, get_pairwise_summary, mc.cores=cores ) %>% 
-		bind_rows()
-
-	# Calculate some summary statistics 
-	ortholog_r_sim_null = 
-		pairwise_sim_null_summary %>% 
-		filter( D=="N" ) %>% 
-		select( tau_a, tau_b ) %>% 
-		cor( ) %>% 
-		.[ 1, 2 ]
-
-	paralog_r_sim_null = 
-		pairwise_sim_null_summary %>% 
-		filter( D=="Y" ) %>% 
-		select( tau_a, tau_b ) %>% 
-		cor( ) %>% 
-		.[ 1, 2 ]
-
-```
-
-```{r simulations_oc, cache=TRUE, echo=FALSE, fig.cap="XX_PIC_VAR", warning=FALSE, message=FALSE}
-
-	dup_adjust = 2
-
-	# Replace the actual Tau values with simulated Tau values, using parameters 
-	# fitted from each tree. 
-	gene_trees_sim_oc = lapply( gene_trees_calibrated[1:1000], sim_tau, dup_adjust=dup_adjust )
-
-	# Calculate contrasts 
-	gene_trees_sim_oc %<>% add_pics_to_trees()
-	nodes_sim_oc_contrasts = gene_trees_sim_oc %>% summarize_contrasts()
-
-	# Calculate pairwise comparisons 
-	pairwise_sim_oc_summary = 
-		mclapply( gene_trees_sim_oc, get_pairwise_summary, mc.cores=cores ) %>% 
-		bind_rows()
-
-	ortholog_r_sim_oc = 
-		pairwise_sim_oc_summary %>% 
-		filter( D=="N" ) %>% 
-		select( tau_a, tau_b ) %>% 
-		cor( ) %>% 
-		.[ 1, 2 ]
-
-	paralog_r_sim_oc = 
-		pairwise_sim_oc_summary %>% 
-		filter( D=="Y" ) %>% 
-		select( tau_a, tau_b ) %>% 
-		cor( ) %>% 
-		.[ 1, 2 ]
-
-```
-
-```{r simulations_plot, cache=TRUE, echo=FALSE, fig.cap="XX_PIC_VAR", warning=FALSE, message=FALSE}
-
-	g_phylo_null = 
-		nodes_sim_null_contrasts %>% 
-		ggplot( aes( x=pic, y=..density.., col=Event ) ) + 
-			geom_freqpoly( binwidth=.001, position="identity" ) +
-			xlab( "Phylogenetic independent \ncontrasts (PIC)" ) +
-			ylab( "Density" ) +
-			xlim( 0, 0.075 ) + 
-			ylim( 0, 100 ) +
-			theme_classic() + 
-			ggtitle( "(b) Phylogenetic result under null simulation" )
-
-	g_pair_null = 
-		pairwise_sim_null_summary %>% 
-		filter( species_a == "homo_sapiens" | species_b == "homo_sapiens" ) %>% 
-		filter( ! is.na( D ) ) %>%
-		group_by( Event, label ) %>% 
-		summarise( cor.test(  tau_a, tau_b )$estimate, n=n() ) %>% 
-		rename( rho=`cor.test(tau_a, tau_b)$estimate` ) %>%
-		rename( clade = label ) %>% 
-		left_join( calibration_times_all, clade=clade ) %>%
-		ggplot() + 
-			geom_point( aes( x=age, y=rho, col=Event, size=n ) ) +
-			theme_classic() +
-			ggtitle( "(a) Pairwise result under null simulation" )
-
-	g_phylo_oc = 
-	nodes_sim_oc_contrasts %>% 
-	ggplot( aes( x=pic, y=..density.., col=Event ) ) + 
-		geom_freqpoly( binwidth=.001, position="identity" ) +
-		xlab( "Phylogenetic independent \ncontrasts (PIC)" ) +
-		ylab( "Density" ) +
-		xlim( 0, 0.075 ) + 
-		ylim( 0, 100 ) +
-		theme_classic() + 
-		ggtitle( "(d) Phylogenetic result under OC simulation" )
-
-	g_pair_oc = 
-		pairwise_sim_oc_summary %>% 
-		filter( species_a == "homo_sapiens" | species_b == "homo_sapiens" ) %>% 
-		filter( ! is.na( D ) ) %>%
-		group_by( Event, label ) %>% 
-		summarise( cor.test(  tau_a, tau_b )$estimate, n=n() ) %>% 
-		rename( rho=`cor.test(tau_a, tau_b)$estimate` ) %>%
-		rename( clade = label ) %>% 
-		left_join( calibration_times_all, clade=clade ) %>%
-		ggplot() + 
-			geom_point( aes( x=age, y=rho, col=Event, size=n ) ) +
-			theme_classic() +
-			ggtitle( "(c) Pairwise result under OC simulation" )
-
-	grid.arrange(
-		g_pair_null,
-		g_phylo_null,
-		g_pair_oc,
-		g_phylo_oc,
-		ncol = 2
-	)
-
-```
-
-> Figure XXSim
-
-We first simulated a data set under the null model that tau evolves at the same rate following duplication and speciation events.
-Under the null model, pairwise analysis results (Figure XXSim a) are very similar as for the observed data, with higher rho for orthologs (`r ortholog_r_sim_null`) than paralogs (`r paralog_r_sim_null`).
-In the phylogenetic analysis of the null simulation, contrasts show similar divergence following speciation and duplication (Figure XXSim b) and does not reject the null hypothesis (Wilcoxon p=`r wilcox_oc( nodes_sim_null_contrasts )`).
-
-We next simulated a data set under the ortholog conjecture, where the rate of evolution of Tau following duplication was `r dup_adjust` fold the rate following speciation.
-The results of this heterogeneous model are nearly indistinguishable from the null model  (Figure XXSim c), and also have a higher rho for orthologs (`r ortholog_r_sim_oc`) than paralogs (`r paralog_r_sim_oc`).
-In the phylogenetic analysis of the ortholog conjecture simulation, contrasts XX (Figure XXSim c) and XX (Wilcoxon p=`r wilcox_oc( nodes_sim_oc_contrasts )`).
-
 ## References

### 2017-02-02 manuscript.rmd
plot formatting, increased dup_adjust to 5

@@
 	# These clade dates are from the KMRR code, they got them from http://timetree.org/
+	# only the following are used in phylogenetic analyses - 
+	# c( "Hominini", "Homininae", "Catarrhini", "Euarchontoglires", "Theria", "Mammalia", "Amniota" )
 	calibration_times = data.frame(
-		age.min =
-			c(  7, 9, 29, 92, 162, 167, 296 ), 
+        age =
+            c( 20, 92, 167, 9, 42, 74, 296, 535, 104, 937, 29, 722, 441, 65, 15, 1215, 
+                414, 371, 162, 25, 74, 77, 86, 7 ), 
         clade = 
-			c( "Hominini", "Homininae", "Catarrhini", "Euarchontoglires", "Theria", "Mammalia", "Amniota" ),
+            c("Hominoidea", "Euarchontoglires", "Mammalia", "Homininae", 
+                    "Simiiformes", "Primates", "Amniota", "Vertebrata", "Eutheria", 
+                    "Bilateria", "Catarrhini", "Chordata", "Euteleostomi", 
+                    "Haplorrhini", "Hominidae", "Opisthokonta", "Sarcopterygii", 
+                    "Tetrapoda", "Theria", "Murinae", "Sciurognathi", "Rodentia", "Glires", "Hominini"
+            ),
         stringsAsFactors=FALSE
     )
+
+	# In simulation, the fold change in rate following duplication relative to speciation
+	dup_adjust = 5
+
 ```
@@
 
-	calibration_times_all = data.frame(
-        age =
-            c( 20, 92, 167, 9, 42, 74, 296, 535, 104, 937, 29, 722, 441, 65, 15, 1215, 
-                414, 371, 162, 25, 74, 77, 86, 7 ), 
-        clade = 
-            c("Hominoidea", "Euarchontoglires", "Mammalia", "Homininae", 
-                    "Simiiformes", "Primates", "Amniota", "Vertebrata", "Eutheria", 
-                    "Bilateria", "Catarrhini", "Chordata", "Euteleostomi", 
-                    "Haplorrhini", "Hominidae", "Opisthokonta", "Sarcopterygii", 
-                    "Tetrapoda", "Theria", "Murinae", "Sciurognathi", "Rodentia", "Glires", "Hominini"
-            ),
-        stringsAsFactors=FALSE
-    )
-
 	pairwise_summary %>% 
@@
 		rename( clade = label ) %>% 
-		left_join( calibration_times_all, clade=clade ) %>%
+		left_join( calibration_times, clade=clade ) %>%
 		ggplot() + 
@@
 This is important since we are only looking at at a subset of the data they considered, the Brawand *et al.* 2011 [@Brawand:2011du] dataset for gene trees that could be successfully time calibrated.
-We can replicate their result - we find a higher Tau correlation between orthologs (R=`r round( ortholog_r, 3) `) than between paralogs (R=`r round( paralog_r, 3)`).
+In their Figure 1, they present a higher Tau correlation coefficient between ortholog pairs than paralog ortholog pairs.
+We find the same here, with correlation coefficients of R=`r round( ortholog_r, 3)` between ortholog pairs and `r round( paralog_r, 3)` for paralog pairs.
 
@@
 
-The reason that paralog pairs have Tau values that are more different than ortholog pairs is that they tend to be more distantly related, not because there are different rates of Tau evolution following duplication and speciation events.
-This means that having information on whether two genes are orthologs or paralogs provides little added information about expression beyond knowing how distantly related the two genes are.
+```{r simulations_null, cache=TRUE, echo=FALSE, fig.cap="XX_PIC_VAR", warning=FALSE, message=FALSE}
+
+	# Replace the actual Tau values with simulated Tau values, using parameters 
+	# fitted from each tree
+	gene_trees_sim_null = lapply( gene_trees_calibrated[1:1000], sim_tau )
+
+	# Perform contrasts 
+	gene_trees_sim_null %<>% add_pics_to_trees()
+	nodes_sim_null_contrast = gene_trees_sim_null %>% summarize_contrasts()
+
+	# Calculate pairwise comparisons
+	pairwise_sim_null_summary = 
+		mclapply( gene_trees_sim_null, get_pairwise_summary, mc.cores=cores ) %>% 
+		bind_rows()
+
+	# Calculate some summary statistics 
+	ortholog_r_sim_null = 
+		pairwise_sim_null_summary %>% 
+		filter( D=="N" ) %>% 
+		select( tau_a, tau_b ) %>% 
+		cor( ) %>% 
+		.[ 1, 2 ]
+
+	paralog_r_sim_null = 
+		pairwise_sim_null_summary %>% 
+		filter( D=="Y" ) %>% 
+		select( tau_a, tau_b ) %>% 
+		cor( ) %>% 
+		.[ 1, 2 ]
+
+```
+
+```{r simulations_oc, cache=TRUE, echo=FALSE, fig.cap="XX_PIC_VAR", warning=FALSE, message=FALSE}
+
+	# Replace the actual Tau values with simulated Tau values, using parameters 
+	# fitted from each tree. 
+	gene_trees_sim_oc = lapply( gene_trees_calibrated[1:1000], sim_tau, dup_adjust=dup_adjust )
+
+	# Calculate contrasts 
+	gene_trees_sim_oc %<>% add_pics_to_trees()
+	nodes_sim_oc_contrast = gene_trees_sim_oc %>% summarize_contrasts()
+
+	# Calculate pairwise comparisons 
+	pairwise_sim_oc_summary = 
+		mclapply( gene_trees_sim_oc, get_pairwise_summary, mc.cores=cores ) %>% 
+		bind_rows()
+
+	ortholog_r_sim_oc = 
+		pairwise_sim_oc_summary %>% 
+		filter( D=="N" ) %>% 
+		select( tau_a, tau_b ) %>% 
+		cor( ) %>% 
+		.[ 1, 2 ]
+
+	paralog_r_sim_oc = 
+		pairwise_sim_oc_summary %>% 
+		filter( D=="Y" ) %>% 
+		select( tau_a, tau_b ) %>% 
+		cor( ) %>% 
+		.[ 1, 2 ]
+
+```
+
+```{r simulations_plot, cache=TRUE, echo=FALSE, fig.cap="XX_PIC_VAR", warning=FALSE, message=FALSE}
+
+	g_phylo_null = 
+		nodes_sim_null_contrast %>% 
+		ggplot( aes( x=pic, y=..density.., col=Event ) ) + 
+			geom_freqpoly( binwidth=.001, position="identity" ) +
+			xlab( "Phylogenetic Independent Contrast" ) +
+			ylab( "Density" ) +
+			xlim( 0, 0.05 ) + 
+			ylim( 0, 75 ) +
+			theme_classic() + 
+			ggtitle( "(b) Null simulation, phylogenetic" ) +
+			theme(legend.justification=c(1,1), legend.position=c(1,1))
+
+	g_pair_null = 
+		pairwise_sim_null_summary %>% 
+		filter( species_a == "homo_sapiens" | species_b == "homo_sapiens" ) %>% 
+		filter( ! is.na( D ) ) %>%
+		group_by( Event, label ) %>% 
+		summarise( cor.test(  tau_a, tau_b )$estimate, n=n() ) %>% 
+		rename( rho=`cor.test(tau_a, tau_b)$estimate` ) %>%
+		rename( clade = label ) %>% 
+		left_join( calibration_times, clade=clade ) %>%
+		ggplot() + 
+			geom_point( aes( x=age, y=rho, col=Event, size=n ) ) +
+			theme_classic() +
+			xlab( "Age" ) +
+			ylab( "Rho" ) +
+			ggtitle( "(a) Null simulation, pairwise" ) +
+			theme(legend.position="none")
+
+	g_phylo_oc = 
+	nodes_sim_oc_contrast %>% 
+	ggplot( aes( x=pic, y=..density.., col=Event ) ) + 
+		geom_freqpoly( binwidth=.001, position="identity" ) +
+		xlab( "Phylogenetic Independent Contrast" ) +
+		ylab( "Density" ) +
+		xlim( 0, 0.05 ) + 
+		ylim( 0, 75 ) +
+		theme_classic() + 
+		ggtitle( "(d) OC simulation, phylogenetic" ) +
+		theme(legend.position="none")
+
+	g_pair_oc = 
+		pairwise_sim_oc_summary %>% 
+		filter( species_a == "homo_sapiens" | species_b == "homo_sapiens" ) %>% 
+		filter( ! is.na( D ) ) %>%
+		group_by( Event, label ) %>% 
+		summarise( cor.test(  tau_a, tau_b )$estimate, n=n() ) %>% 
+		rename( rho=`cor.test(tau_a, tau_b)$estimate` ) %>%
+		rename( clade = label ) %>% 
+		left_join( calibration_times, clade=clade ) %>%
+		ggplot() + 
+			geom_point( aes( x=age, y=rho, col=Event, size=n ) ) +
+			theme_classic() +
+			xlab( "Age" ) +
+			ylab( "Rho" ) +
+			ggtitle( "(c) OC simulation, pairwise" ) +
+			theme(legend.position="none")
+
+	grid.arrange(
+		g_pair_null,
+		g_phylo_null,
+		g_pair_oc,
+		g_phylo_oc,
+		ncol = 2
+	)
+
+```
+
+> Figure XXSim
+
+We simulated a data set under the null model that tau evolves at the same rate following duplication and speciation events.
+Under the null model, pairwise analysis results (Figure XXSim a) are very similar as for the observed data, with higher rho for orthologs (`r ortholog_r_sim_null`) than paralogs (`r paralog_r_sim_null`).
+In the phylogenetic analysis of the null simulation, contrasts show similar divergence following speciation and duplication (Figure XXSim b) and does not reject the null hypothesis (Wilcoxon p=`r wilcox_oc( nodes_sim_null_contrast )`).
+
+We next simulated a data set under the ortholog conjecture, where the rate of evolution of Tau following duplication was `r dup_adjust` fold the rate following speciation.
+The results of this heterogeneous model are nearly indistinguishable from the null model  (Figure XXSim c), and also have a higher rho for orthologs (`r ortholog_r_sim_oc`) than paralogs (`r paralog_r_sim_oc`).
+In the phylogenetic analysis of the ortholog conjecture simulation, contrasts XX (Figure XXSim c) and XX (Wilcoxon p=`r wilcox_oc( nodes_sim_oc_contrast )`).
+
+#### Implications for the ortholog conjecture
+
+We find that having information on whether two genes are orthologs or paralogs provides little added information about expression beyond knowing how distantly related the two genes are.
 This has several implications.
@@
 In this case, KMRR [@KryuchkovaMostacci:2016iw] relied on orthology and paralogy annotations that are derived from the inferred duplication and speciation history on gene phylogenies [@Herrero:2016jj].
-
-```{r simulations_null, cache=TRUE, echo=FALSE, fig.cap="XX_PIC_VAR", warning=FALSE, message=FALSE}
-
-	# Replace the actual Tau values with simulated Tau values, using parameters 
-	# fitted from each tree
-	gene_trees_sim_null = lapply( gene_trees_calibrated[1:1000], sim_tau )
-
-	# Perform contrasts 
-	gene_trees_sim_null %<>% add_pics_to_trees()
-	nodes_sim_null_contrasts = gene_trees_sim_null %>% summarize_contrasts()
-
-	# Calculate pairwise comparisons
-	pairwise_sim_null_summary = 
-		mclapply( gene_trees_sim_null, get_pairwise_summary, mc.cores=cores ) %>% 
-		bind_rows()
-
-	# Calculate some summary statistics 
-	ortholog_r_sim_null = 
-		pairwise_sim_null_summary %>% 
-		filter( D=="N" ) %>% 
-		select( tau_a, tau_b ) %>% 
-		cor( ) %>% 
-		.[ 1, 2 ]
-
-	paralog_r_sim_null = 
-		pairwise_sim_null_summary %>% 
-		filter( D=="Y" ) %>% 
-		select( tau_a, tau_b ) %>% 
-		cor( ) %>% 
-		.[ 1, 2 ]
-
-```
-
-```{r simulations_oc, cache=TRUE, echo=FALSE, fig.cap="XX_PIC_VAR", warning=FALSE, message=FALSE}
-
-	dup_adjust = 2
-
-	# Replace the actual Tau values with simulated Tau values, using parameters 
-	# fitted from each tree. 
-	gene_trees_sim_oc = lapply( gene_trees_calibrated[1:1000], sim_tau, dup_adjust=dup_adjust )
-
-	# Calculate contrasts 
-	gene_trees_sim_oc %<>% add_pics_to_trees()
-	nodes_sim_oc_contrasts = gene_trees_sim_oc %>% summarize_contrasts()
-
-	# Calculate pairwise comparisons 
-	pairwise_sim_oc_summary = 
-		mclapply( gene_trees_sim_oc, get_pairwise_summary, mc.cores=cores ) %>% 
-		bind_rows()
-
-	ortholog_r_sim_oc = 
-		pairwise_sim_oc_summary %>% 
-		filter( D=="N" ) %>% 
-		select( tau_a, tau_b ) %>% 
-		cor( ) %>% 
-		.[ 1, 2 ]
-
-	paralog_r_sim_oc = 
-		pairwise_sim_oc_summary %>% 
-		filter( D=="Y" ) %>% 
-		select( tau_a, tau_b ) %>% 
-		cor( ) %>% 
-		.[ 1, 2 ]
-
-```
-
-```{r simulations_plot, cache=TRUE, echo=FALSE, fig.cap="XX_PIC_VAR", warning=FALSE, message=FALSE}
-
-	g_phylo_null = 
-		nodes_sim_null_contrasts %>% 
-		ggplot( aes( x=pic, y=..density.., col=Event ) ) + 
-			geom_freqpoly( binwidth=.001, position="identity" ) +
-			xlab( "Phylogenetic independent \ncontrasts (PIC)" ) +
-			ylab( "Density" ) +
-			xlim( 0, 0.075 ) + 
-			ylim( 0, 100 ) +
-			theme_classic() + 
-			ggtitle( "(b) Phylogenetic result under null simulation" )
-
-	g_pair_null = 
-		pairwise_sim_null_summary %>% 
-		filter( species_a == "homo_sapiens" | species_b == "homo_sapiens" ) %>% 
-		filter( ! is.na( D ) ) %>%
-		group_by( Event, label ) %>% 
-		summarise( cor.test(  tau_a, tau_b )$estimate, n=n() ) %>% 
-		rename( rho=`cor.test(tau_a, tau_b)$estimate` ) %>%
-		rename( clade = label ) %>% 
-		left_join( calibration_times_all, clade=clade ) %>%
-		ggplot() + 
-			geom_point( aes( x=age, y=rho, col=Event, size=n ) ) +
-			theme_classic() +
-			ggtitle( "(a) Pairwise result under null simulation" )
-
-	g_phylo_oc = 
-	nodes_sim_oc_contrasts %>% 
-	ggplot( aes( x=pic, y=..density.., col=Event ) ) + 
-		geom_freqpoly( binwidth=.001, position="identity" ) +
-		xlab( "Phylogenetic independent \ncontrasts (PIC)" ) +
-		ylab( "Density" ) +
-		xlim( 0, 0.075 ) + 
-		ylim( 0, 100 ) +
-		theme_classic() + 
-		ggtitle( "(d) Phylogenetic result under OC simulation" )
-
-	g_pair_oc = 
-		pairwise_sim_oc_summary %>% 
-		filter( species_a == "homo_sapiens" | species_b == "homo_sapiens" ) %>% 
-		filter( ! is.na( D ) ) %>%
-		group_by( Event, label ) %>% 
-		summarise( cor.test(  tau_a, tau_b )$estimate, n=n() ) %>% 
-		rename( rho=`cor.test(tau_a, tau_b)$estimate` ) %>%
-		rename( clade = label ) %>% 
-		left_join( calibration_times_all, clade=clade ) %>%
-		ggplot() + 
-			geom_point( aes( x=age, y=rho, col=Event, size=n ) ) +
-			theme_classic() +
-			ggtitle( "(c) Pairwise result under OC simulation" )
-
-	grid.arrange(
-		g_pair_null,
-		g_phylo_null,
-		g_pair_oc,
-		g_phylo_oc,
-		ncol = 2
-	)
-
-```
-
-> Figure XXSim
-
-We simulated a data set under the null model that tau evolves at the same rate following duplication and speciation events.
-Under the null model, pairwise analysis results (Figure XXSim a) are very similar as for the observed data, with higher rho for orthologs (`r ortholog_r_sim_null`) than paralogs (`r paralog_r_sim_null`).
-In the phylogenetic analysis of the null simulation, contrasts show similar divergence following speciation and duplication (Figure XXSim b) and does not reject the null hypothesis (Wilcoxon p=`r wilcox_oc( nodes_sim_null_contrasts )`).
-
-We next simulated a data set under the ortholog conjecture, where the rate of evolution of Tau following duplication was `r dup_adjust` fold the rate following speciation.
-The results of this heterogeneous model are nearly indistinguishable from the null model  (Figure XXSim c), and also have a higher rho for orthologs (`r ortholog_r_sim_oc`) than paralogs (`r paralog_r_sim_oc`).
-In the phylogenetic analysis of the ortholog conjecture simulation, contrasts XX (Figure XXSim c) and XX (Wilcoxon p=`r wilcox_oc( nodes_sim_oc_contrasts )`).
-
-#### Implications for the ortholog conjecture
 

### 2017-02-03 manuscript.rmd
touched up overview figure, generate expectation figure with code

@@
 ```
-
-# Multiple pairwise comparisons between species strongly mislead functional genomics studies
-
-Casey W. Dunn^1^*, Felipe Zapata^1,2^, Catriona Munro^1^, Stefan Siebert^1,3^, Andreas Hejnol^4^
-
-^1^ Department of Ecology and Evolutionary Biology, Brown University, Providence, RI, USA
-
-^2^ Current address: Department of Ecology and Evolutionary Biology, University of California Los Angeles, Los Angeles, CA, USA
-
-^3^ Current address: Department of Molecular & Cellular Biology, University of California at Davis, Davis, CA, USA
-
-^4^ Sars International Centre for Marine Molecular Biology, University of Bergen, Bergen, Norway
-
-\* Corresponding author, casey_dunn@brown.edu
-
-## Abstract
-
-Functional genomics, the study of how genomes work, is one of the most rapidly advancing fields in biology.
-There is now considerable interest in comparing genome function across species to understand what is conserved, what is variable, and how the diversity of organism traits (including anatomy, physiology, and development) relates to genome diversity.
-Most published comparisons of genome function across species have relied on multiple pairwise comparisons between species, an approach that neglects to account for the evolutionary relationships among species.
-This has two potential negative impacts.
-First, it can lead to the wrong conclusions about the evolution of genome function.
-Second, it is a missed opportunity to learn about biology that can only be understood in an explicit historical context that considers these relationships.
-Here we examine two recently published comparative gene expression studies that used multiple pairwise comparisons rather than phylogenetic approaches to investigate evolutionary patterns of gene expression.
-We find problems with the pairwise comparisons used in both studies that undermine their published conclusions.
-The first study we examine found that expression tends to be more different between paralogs than orthologs, and interpreted this as evidence of the ortholog conjecture- the hypothesis that there are greater rates of evolutionary change following duplication.
-Our phylogenetic analysis finds no difference in rates following duplication and speciation.
-Instead, the greater difference between paralogs is because they tend to be more distantly related to each other than orthologs are to each other.
-The second study claimed to find more evolutionary variance in gene expression at a midpoint of embryonic development than earlier and later when comparing distantly related animals.
-We instead find that changes along the branch of a single taxon, the ctenophore, were over counted because they were included in each pairwise comparison with this species.
-These reanalyses are concrete demonstrations of the inadequacy of pairwise comparisons in comparative genomics studies, including gene expression analyses, and indicates that it will be critical to adopt phylogenetic comparative methods in future work.
-
-## Introduction
-
-The focus of genomics research has quickly shifted from describing genome sequences to functional genomics, the study of how genomes "work" using tools that measure functional attributes such as expression, chromatin state, and transcription initiation.
-Functional genomics, in turn, is now becoming more comparative- there is great interest in understanding how functional genomic variation across species gives rise to a diversity of development, morphology, physiology, and other phenotypes [@Wray:2013gh].
-Comparative Functional Genomics (CFG) analyses are also critical to transferring functional insight across species, and will grow in importance in coming years.
-
-A rich theoretical and statistical methodology of phylogenetic comparative methods have been developed over the last three decades to address the challenges and opportunities of comparisons across species [@Felsenstein:1985ua; @Grafen:1989um; @Pagel:1999gc; Revell:2012jd; @FitzJohn:2012ey; @Uyeda:2014jl; @Garamszegi:2014ew].
-A central challenge is the dependence of observations across species due to the evolutionary history of species -- more closely related species share many traits that evolve once in a common ancestor.
-This violates the fundamental assumption of independence of observations in standard statistical methods.
-Phylogenetic comparative methods have largely been applied to the evolution of morphological and ecological traits, but are just as relevant to CFG.
-Most CFG studies have abstained from phylogenetic approaches and instead heavily rely on multiple pairwise comparisons (Fig.
-1A) between contemporary traits that do not account for the fact that evolutionary relationships explain much of the structure of variation across species.
-This leaves CFG studies susceptible to serious statistical errors and is a missed opportunity to ask questions that are only accessible in an explicit phylogenetic context.
-
-One reason that CFG has not yet embraced phylogenetic approaches is that it has not yet been concretely demonstrated that pairwise and phylogenetic approaches can lead to different results in CFG studies.
-In this manuscript we examine this issue.
-Although the value of phylogenetic approaches has been repeatedly shown in comparative analyses of other types of character data [@Ricklefs:1996wb; @Chamberlain:2012vx; @OMeara:2012gq], this value is apparently not well known to the functional genomics community since the field has until recently largely focused on within-species analyses.
-Many initial CFG studies have also been focused on only two species at a time, where the differences between phylogenetic and pairwise approaches are not as pronounced.
-As the number of species considered in each CFG study grows, the shortcomings of pairwise approaches will become far more acute.
-
-Most CFG studies to date compare gene expression across species, often between distinct tissues or developmental time points.
-Gene expression has been a popular target of CFG studies because, with the advent of RNA-Seq and related transcriptome sequencing methods, it is the most readily collected functional genomics trait across diverse species.
-Here we reconsider two recent comparative analyses of gene expression data [@KryuchkovaMostacci:2016iw; @Levin:2016bp].
-
-The first study, Kryuchkova-Mostacci and Robinson-Rechavi (KMRR) 2016 [@KryuchkovaMostacci:2016iw], analyzed multiple vertebrate expression datasets to test the ortholog conjecture - the hypothesis that orthologs tend to have more conserved attributes (specificity of expression across organs in this case) than do paralogs [@Nehrt:2011bm, @Koonin:2005fp].
-Using pairwise comparisons (Fig.
-1A), they found lower expression correlation between paralogs than between orthologs and interpreted this as strong support for the ortholog conjecture.
-
-The second CFG study we evaluate here is Levin *et al.* [@Levin:2016bp].
-This study analyzed gene expression through the course of embryonic development for ten animal species, each from a different phylum.
-Using pairwise comparisons (Figure 1A), they found there is more evolutionary variance in gene expression at a mid phase of development than there is at early and late phases for these ten animals (Fig.
-XX).
-They suggest that this supports an "inverse hourglass" model for the evolution of gene expression, in contrast with the "hourglass" model previously proposed for closely related species [@Kalinka:2010dw], and provides biological justification for the concept of phyla and may provide an operational definition of phyla.
-We previously described concerns with these interpretations [@Hejnol:2016hw].
-Here we directly address the analyses themselves by examining the structure of the pairwise comparisons.
-
-![Figure Overview](Figure_overview.png?raw=true)
-
-> Figure 1 | Pairwise and phylogenetic comparative approaches.
-(a) Many comparative functional genomic studies rely on pairwise comparisons, where traits of each gene are compared to traits of other genes (such as orthologs and paralogs).
-This leads to many more comparisons than unique observations, making each comparison dependent on others.
-(b) Comparative phylogenetic methods, including phylogenetic independent contrasts [@Felsenstein:1985ua], make a smaller number of independent comparisons, where each contrast measures independent changes along different branches.
-Phylogenetic approaches are rarely used for functional genomics studies.
-
-## Results and Discussion
-
-### KMRR Reanalysis
-
-#### Original pairwise test of the ortholog conjecture
-
-The KMRR study [@KryuchkovaMostacci:2016iw] sought to test the ortholog conjecture.
-The ortholog conjecture [@Nehrt:2011bm] is the proposition that orthologs (genes that diverged from each other due to a speciation event) have more similar attributes than do paralogs (genes that diverged from each other due to a gene duplication event).
-The ortholog conjecture has important biological and technical implications.
-It shapes our understanding of the functional diversity of gene families.
-It is used to relate findings from well-studied genes to related genes that have not been investigated in detail.
-It can be applied to any trait of genes, from gene sequence to biochemical properties to expression.
-While the ortholog conjecture describes a specific pattern of functional diversity across genes, it is also articulated as a hypothesis about the process of evolution-- that there is greater evolutionary change in gene attributes following a duplication event than a speciation event.
-
-Despite its importance, there have been relatively few tests of the ortholog conjecture.
-Previous work has shown that ontology annotations are not sufficient to test the ortholog conjecture [@Chen:2012jz; @Thomas:2012hz].
-Analyses of domain structure were consistent with the ortholog conjecture [@Forslund:2011cw].
-There have been few tests of the ortholog conjecture with regards to gene expression [@Chen:2012jz], and KMRR is the most thorough such expression study to date.
-
-The KMRR study considered several publicly available datasets.
-Here we consider their conclusions in the context of one of these studies, Brawand *et al.* 2011 [@Brawand:2011du].
-This dataset of gene expression for six organs across ten species (nine mammals and one bird), eight of which were analyzed by KMRR, is the best sampled in their analyses.
-Their expression summary statistic is Tau [@Yanai:2005dh, @KryuchkovaMostacci:2016in], an indicator of tissue specificity of gene expression.
-It can range from a value of 0, which indicates no specificity (*i.e.*, uniform expression across tissues), to a value of 1, which indicates high specificity (*i.e.*, expression in only one tissue).
-It is convenient in that it is a single number of defined range for each gene, though of course since the original expression is multidimensional this means much information is discarded.
-This includes information about which tissue expression is specific to.
-For example, if one gene has expression specific to the brain and another expression specific to the kidney, both would have a Tau of 1.
-
-Results in KMRR are based on pairwise comparisons (Figure 1A) between Tau within each gene family.
-They found the correlation coefficient of Tau for orthologs to be significantly greater than the correlation coefficient of Tau for paralogs, *i.e.* that orthologs tend to be more similar to each other than paralogs.
-From this they concluded that their analyses support the ortholog conjecture.
-They also concluded that this pattern provides support for a particular evolutionary process, that "tissue-specificity evolves very slowly in the absence of duplication, while immediately after duplication the new gene copy differs".
-
-#### Phylogenetic reanalyses
-
-We reanalyzed the the KMRR study using phylogenetic comparative methods, focusing on the dataset of Brawand *et al.* 2011 [@Brawand:2011du].
-For each internal node in each gene tree (Figure 1b), we measured the phylogenetic independent contrast (PIC) [@Felsenstein:1985ua] of the expression statistic Tau.
-This is the difference in values of Tau for descendant nodes scaled by the expected variance (determined largely by the lengths of these branches).
-These contrasts were then annotated by whether they are made across a speciation or duplication event.
-The original description of independent contrasts [@Felsenstein:1985ua] focused on assessing covariance between changes in two traits on a tree.
-Our use of contrasts is a bit different-- we look for differences in evolutionary changes of one trait (differential expression) between two categories of nodes (speciation and duplication) on a tree.
-
-![Figure Expectations](figure_expectations.png?raw=true)
-
-> Figure XXExpectations | (a) Under the null hypothesis that there is no difference in the rate of evolution after duplication or speciation events, the distribution of PICs would be the same across both types of nodes.
-(b) Under the ortholog conjecture, contrasts across duplication nodes would tend to be larger than contrasts across speciation nodes.
-
-To test the ortholog conjecture it is essential to have a null hypothesis that makes predictions distinct from the ortholog conjecture.
-A suitable null hypothesis is that there is no difference in evolutionary change in differential expression along branches that descend from speciation or duplication events [@Studer:2009bm] (Figure 1b).
-Under this hypothesis, we would predict that contrasts across speciation nodes and duplication nodes are drawn from the same distribution (Figure XXEXPECTATIONS a).
-Under the alternative hypothesis specified by the ortholog conjecture, that there is a higher rate of change following duplication events than speciation events, we would expect to see the distribution of duplication contrasts shifted to higher values relative to the speciation contrasts (Figure XXEXPECTATIONS b).
 
@@
 
+# Multiple pairwise comparisons between species strongly mislead functional genomics studies
+
+Casey W. Dunn^1^*, Felipe Zapata^1,2^, Catriona Munro^1^, Stefan Siebert^1,3^, Andreas Hejnol^4^
+
+^1^ Department of Ecology and Evolutionary Biology, Brown University, Providence, RI, USA
+
+^2^ Current address: Department of Ecology and Evolutionary Biology, University of California Los Angeles, Los Angeles, CA, USA
+
+^3^ Current address: Department of Molecular & Cellular Biology, University of California at Davis, Davis, CA, USA
+
+^4^ Sars International Centre for Marine Molecular Biology, University of Bergen, Bergen, Norway
+
+\* Corresponding author, casey_dunn@brown.edu
+
+## Abstract
+
+Functional genomics, the study of how genomes work, is one of the most rapidly advancing fields in biology.
+There is now considerable interest in comparing genome function across species to understand what is conserved, what is variable, and how the diversity of organism traits (including anatomy, physiology, and development) relates to genome diversity.
+Most published comparisons of genome function across species have relied on multiple pairwise comparisons between species, an approach that neglects to account for the evolutionary relationships among species.
+This has two potential negative impacts.
+First, it can lead to the wrong conclusions about the evolution of genome function.
+Second, it is a missed opportunity to learn about biology that can only be understood in an explicit historical context that considers these relationships.
+Here we examine two recently published comparative gene expression studies that used multiple pairwise comparisons rather than phylogenetic approaches to investigate evolutionary patterns of gene expression.
+We find problems with the pairwise comparisons used in both studies that undermine their published conclusions.
+The first study we examine found that expression tends to be more different between paralogs than orthologs, and interpreted this as evidence of the ortholog conjecture- the hypothesis that there are greater rates of evolutionary change following duplication.
+Our phylogenetic analysis finds no difference in rates following duplication and speciation.
+Instead, the greater difference between paralogs is because they tend to be more distantly related to each other than orthologs are to each other.
+The second study claimed to find more evolutionary variance in gene expression at a midpoint of embryonic development than earlier and later when comparing distantly related animals.
+We instead find that changes along the branch of a single taxon, the ctenophore, were over counted because they were included in each pairwise comparison with this species.
+These reanalyses are concrete demonstrations of the inadequacy of pairwise comparisons in comparative genomics studies, including gene expression analyses, and indicates that it will be critical to adopt phylogenetic comparative methods in future work.
+
+## Introduction
+
+The focus of genomics research has quickly shifted from describing genome sequences to functional genomics, the study of how genomes "work" using tools that measure functional attributes such as expression, chromatin state, and transcription initiation.
+Functional genomics, in turn, is now becoming more comparative- there is great interest in understanding how functional genomic variation across species gives rise to a diversity of development, morphology, physiology, and other phenotypes [@Wray:2013gh].
+Comparative Functional Genomics (CFG) analyses are also critical to transferring functional insight across species, and will grow in importance in coming years.
+
+A rich theoretical and statistical methodology of phylogenetic comparative methods have been developed over the last three decades to address the challenges and opportunities of comparisons across species [@Felsenstein:1985ua; @Grafen:1989um; @Pagel:1999gc; Revell:2012jd; @FitzJohn:2012ey; @Uyeda:2014jl; @Garamszegi:2014ew].
+A central challenge is the dependence of observations across species due to the evolutionary history of species -- more closely related species share many traits that evolve once in a common ancestor.
+This violates the fundamental assumption of independence of observations in standard statistical methods.
+Phylogenetic comparative methods have largely been applied to the evolution of morphological and ecological traits, but are just as relevant to CFG.
+Most CFG studies have abstained from phylogenetic approaches and instead heavily rely on multiple pairwise comparisons (Fig.
+1A) between contemporary traits that do not account for the fact that evolutionary relationships explain much of the structure of variation across species.
+This leaves CFG studies susceptible to serious statistical errors and is a missed opportunity to ask questions that are only accessible in an explicit phylogenetic context.
+
+One reason that CFG has not yet embraced phylogenetic approaches is that it has not yet been concretely demonstrated that pairwise and phylogenetic approaches can lead to different results in CFG studies.
+In this manuscript we examine this issue.
+Although the value of phylogenetic approaches has been repeatedly shown in comparative analyses of other types of character data [@Ricklefs:1996wb; @Chamberlain:2012vx; @OMeara:2012gq], this value is apparently not well known to the functional genomics community since the field has until recently largely focused on within-species analyses.
+Many initial CFG studies have also been focused on only two species at a time, where the differences between phylogenetic and pairwise approaches are not as pronounced.
+As the number of species considered in each CFG study grows, the shortcomings of pairwise approaches will become far more acute.
+
+Most CFG studies to date compare gene expression across species, often between distinct tissues or developmental time points.
+Gene expression has been a popular target of CFG studies because, with the advent of RNA-Seq and related transcriptome sequencing methods, it is the most readily collected functional genomics trait across diverse species.
+Here we reconsider two recent comparative analyses of gene expression data [@KryuchkovaMostacci:2016iw; @Levin:2016bp].
+
+The first study, Kryuchkova-Mostacci and Robinson-Rechavi (KMRR) 2016 [@KryuchkovaMostacci:2016iw], analyzed multiple vertebrate expression datasets to test the ortholog conjecture - the hypothesis that orthologs tend to have more conserved attributes (specificity of expression across organs in this case) than do paralogs [@Nehrt:2011bm, @Koonin:2005fp].
+Using pairwise comparisons (Fig.
+1A), they found lower expression correlation between paralogs than between orthologs and interpreted this as strong support for the ortholog conjecture.
+
+The second CFG study we evaluate here is Levin *et al.* [@Levin:2016bp].
+This study analyzed gene expression through the course of embryonic development for ten animal species, each from a different phylum.
+Using pairwise comparisons (Figure 1A), they found there is more evolutionary variance in gene expression at a mid phase of development than there is at early and late phases for these ten animals (Fig.
+XX).
+They suggest that this supports an "inverse hourglass" model for the evolution of gene expression, in contrast with the "hourglass" model previously proposed for closely related species [@Kalinka:2010dw], and provides biological justification for the concept of phyla and may provide an operational definition of phyla.
+We previously described concerns with these interpretations [@Hejnol:2016hw].
+Here we directly address the analyses themselves by examining the structure of the pairwise comparisons.
+
+![Figure Overview](Figure_overview.png?raw=true)
+
+> Figure 1 | Pairwise and phylogenetic comparative approaches.
+(a) Many comparative functional genomic studies rely on pairwise comparisons, where traits of each gene are compared to traits of other genes (such as orthologs and paralogs).
+This leads to many more comparisons than unique observations, making each comparison dependent on others.
+(b) Comparative phylogenetic methods, including phylogenetic independent contrasts [@Felsenstein:1985ua], make a smaller number of independent comparisons, where each contrast measures independent changes along different branches.
+Phylogenetic approaches are rarely used for functional genomics studies.
+
+## Results and Discussion
+
+### KMRR Reanalysis
+
+#### Original pairwise test of the ortholog conjecture
+
+The KMRR study [@KryuchkovaMostacci:2016iw] sought to test the ortholog conjecture.
+The ortholog conjecture [@Nehrt:2011bm] is the proposition that orthologs (genes that diverged from each other due to a speciation event) have more similar attributes than do paralogs (genes that diverged from each other due to a gene duplication event).
+The ortholog conjecture has important biological and technical implications.
+It shapes our understanding of the functional diversity of gene families.
+It is used to relate findings from well-studied genes to related genes that have not been investigated in detail.
+It can be applied to any trait of genes, from gene sequence to biochemical properties to expression.
+While the ortholog conjecture describes a specific pattern of functional diversity across genes, it is also articulated as a hypothesis about the process of evolution-- that there is greater evolutionary change in gene attributes following a duplication event than a speciation event.
+
+Despite its importance, there have been relatively few tests of the ortholog conjecture.
+Previous work has shown that ontology annotations are not sufficient to test the ortholog conjecture [@Chen:2012jz; @Thomas:2012hz].
+Analyses of domain structure were consistent with the ortholog conjecture [@Forslund:2011cw].
+There have been few tests of the ortholog conjecture with regards to gene expression [@Chen:2012jz], and KMRR is the most thorough such expression study to date.
+
+The KMRR study considered several publicly available datasets.
+Here we consider their conclusions in the context of one of these studies, Brawand *et al.* 2011 [@Brawand:2011du].
+This dataset of gene expression for six organs across ten species (nine mammals and one bird), eight of which were analyzed by KMRR, is the best sampled in their analyses.
+Their expression summary statistic is Tau [@Yanai:2005dh, @KryuchkovaMostacci:2016in], an indicator of tissue specificity of gene expression.
+It can range from a value of 0, which indicates no specificity (*i.e.*, uniform expression across tissues), to a value of 1, which indicates high specificity (*i.e.*, expression in only one tissue).
+It is convenient in that it is a single number of defined range for each gene, though of course since the original expression is multidimensional this means much information is discarded.
+This includes information about which tissue expression is specific to.
+For example, if one gene has expression specific to the brain and another expression specific to the kidney, both would have a Tau of 1.
+
+Results in KMRR are based on pairwise comparisons (Figure 1A) between Tau within each gene family.
+They found the correlation coefficient of Tau for orthologs to be significantly greater than the correlation coefficient of Tau for paralogs, *i.e.* that orthologs tend to be more similar to each other than paralogs.
+From this they concluded that their analyses support the ortholog conjecture.
+They also concluded that this pattern provides support for a particular evolutionary process, that "tissue-specificity evolves very slowly in the absence of duplication, while immediately after duplication the new gene copy differs".
+
+#### Phylogenetic reanalyses
+
+We reanalyzed the the KMRR study using phylogenetic comparative methods, focusing on the dataset of Brawand *et al.* 2011 [@Brawand:2011du].
+For each internal node in each gene tree (Figure 1b), we measured the phylogenetic independent contrast (PIC) [@Felsenstein:1985ua] of the expression statistic Tau.
+This is the difference in values of Tau for descendant nodes scaled by the expected variance (determined largely by the lengths of these branches).
+These contrasts were then annotated by whether they are made across a speciation or duplication event.
+The original description of independent contrasts [@Felsenstein:1985ua] focused on assessing covariance between changes in two traits on a tree.
+Our use of contrasts is a bit different-- we look for differences in evolutionary changes of one trait (differential expression) between two categories of nodes (speciation and duplication) on a tree.
+
+```{r expectation_plots, echo=FALSE, cache=TRUE, message=F, warning=F}
+
+	sim_expectation = function( scale = 1, n = 10000 ){
+		expectation = 
+			bind_rows(
+				data.frame( 
+					pic = rgamma( n, 1, scale = scale ),
+					Event = "Duplication",
+					stringsAsFactors = FALSE
+				),
+				data.frame( 
+					pic = rgamma( n, 1, scale = 1 ),
+					Event = "Speciation",
+					stringsAsFactors = FALSE
+				)
+			)
+
+		event_levels = c( "Speciation", "Duplication" )
+		expectation$Event = 
+			factor( expectation$Event, levels=event_levels )
+
+		return( expectation )
+	}
+
+	g_expectation_null =
+		sim_expectation() %>%
+		ggplot( aes( x=pic, y=..density.., col=Event ) ) + 
+			geom_freqpoly(  binwidth=0.2, position="identity" ) +
+			xlab( "Phylogenetic independent contrasts (PIC)" ) +
+			ylab( "Density" ) +
+			xlim( 0, 6 ) + 
+			theme_classic() +
+			theme( 
+				axis.text.x=element_blank(), 
+				axis.ticks.x=element_blank(), 
+				axis.text.y=element_blank(), 
+				axis.ticks.y=element_blank(),
+				legend.justification=c( 1, 1 ), 
+				legend.position=c( 1, 1 )
+			) +
+			ggtitle( "(a) Null expectation" )
+
+	g_expectation_oc =
+		sim_expectation( scale=2 ) %>%
+		ggplot( aes( x=pic, y=..density.., col=Event ) ) + 
+			geom_freqpoly(  binwidth=0.2, position="identity" ) +
+			xlab( "Phylogenetic independent contrasts (PIC)" ) +
+			ylab( "Density" ) +
+			xlim( 0, 6 ) + 
+			theme_classic() +
+			theme( 
+				axis.text.x=element_blank(), 
+				axis.ticks.x=element_blank(), 
+				axis.text.y=element_blank(), 
+				axis.ticks.y=element_blank(),
+				legend.position="none"
+			) +
+			ggtitle( "(b) Ortholog Conjecture (OC) expectation" )
+
+	grid.arrange(
+			g_expectation_null,
+			g_expectation_oc,
+			ncol = 2
+		)
+
+```
+
+> Figure XXExpectations | (a) Under the null hypothesis that there is no difference in the rate of evolution after duplication or speciation events, the distribution of PICs would be the same across both types of nodes.
+(b) Under the ortholog conjecture, contrasts across duplication nodes would tend to be larger than contrasts across speciation nodes.
+
+To test the ortholog conjecture it is essential to have a null hypothesis that makes predictions distinct from the ortholog conjecture.
+A suitable null hypothesis is that there is no difference in evolutionary change in differential expression along branches that descend from speciation or duplication events [@Studer:2009bm] (Figure 1b).
+Under this hypothesis, we would predict that contrasts across speciation nodes and duplication nodes are drawn from the same distribution (Figure XXEXPECTATIONS a).
+Under the alternative hypothesis specified by the ortholog conjecture, that there is a higher rate of change following duplication events than speciation events, we would expect to see the distribution of duplication contrasts shifted to higher values relative to the speciation contrasts (Figure XXEXPECTATIONS b).
+
 ```{r load_trees, echo=FALSE, cache=TRUE, message=F, warning=F}
@@
 	# fitted from each tree
-	gene_trees_sim_null = lapply( gene_trees_calibrated[1:1000], sim_tau )
+	gene_trees_sim_null = lapply( gene_trees_calibrated, sim_tau )
 
@@
 	# fitted from each tree. 
-	gene_trees_sim_oc = lapply( gene_trees_calibrated[1:1000], sim_tau, dup_adjust=dup_adjust )
+	gene_trees_sim_oc = lapply( gene_trees_calibrated, sim_tau, dup_adjust=dup_adjust )
 
@@
 		ggplot( aes( x=pic, y=..density.., col=Event ) ) + 
-			geom_freqpoly( binwidth=.001, position="identity" ) +
+			geom_freqpoly( binwidth=.0025, position="identity" ) +
 			xlab( "Phylogenetic Independent Contrast" ) +
@@
 			ggtitle( "(b) Null simulation, phylogenetic" ) +
-			theme(legend.justification=c(1,1), legend.position=c(1,1))
+			theme( legend.justification=c( 1, 1 ), legend.position=c( 1, 1 ) )
 
@@
 			ylab( "Rho" ) +
+			ylim( 0, 1 ) +
 			ggtitle( "(a) Null simulation, pairwise" ) +
@@
 	ggplot( aes( x=pic, y=..density.., col=Event ) ) + 
-		geom_freqpoly( binwidth=.001, position="identity" ) +
+		geom_freqpoly( binwidth=.0025, position="identity" ) +
 		xlab( "Phylogenetic Independent Contrast" ) +
@@
 			ylab( "Rho" ) +
+			ylim( 0, 1 ) +
 			ggtitle( "(c) OC simulation, pairwise" ) +
@@
 
+### Software versions
+
+This manuscript was computed on `r format( Sys.time(), "%a %b %d %X %Y" )` with the following R package versions.
+
+```{r session_summary, echo=FALSE, comment=NA}
+	sessionInfo()
+```
+
 ## References

### 2017-02-04 manuscript.rmd
fixed error that led to crash, reworked abstract and intro

@@
 	# These clade dates are from the KMRR code, they got them from http://timetree.org/
-	# only the following are used in phylogenetic analyses - 
-	# c( "Hominini", "Homininae", "Catarrhini", "Euarchontoglires", "Theria", "Mammalia", "Amniota" )
 	calibration_times = data.frame(
@@
 
+	# The calibration times used for speciation nodes
+	focal_calibrations_clades = 
+		c( "Hominini", "Homininae", "Catarrhini", "Euarchontoglires", "Theria", "Mammalia", "Amniota" )
+
 	# In simulation, the fold change in rate following duplication relative to speciation
@@
 	add_noise_calibration_times = function( calibration_times, sd_fraction ) {
-		age = calibration_times$age.min
+		age = calibration_times$age
 		new_age = NA
@@
 
-		calibration_times$age.min = new_age
+		calibration_times$age = new_age
 		return( calibration_times )
@@
 	#' nodes annotated with a value of "N" in @data$D
-	#' @param calibration_times A dataframe with two columns: age.min is the age 
+	#' @param calibration_times A dataframe with two columns: age is the age 
 	#' of a clade, and clade is the name of the clade
@@
 			left_join( calibration_times, c( "label" = "clade" ) ) %>%
-			select( node, age.min ) %>%
-			mutate( age.max = age.min ) %>% 
+			mutate( age.min = age ) %>%
+			mutate( age.max = age ) %>% 
+			select( node, age.min, age.max ) %>%
 			mutate( soft.bounds = NA )
@@
 
-```
-
-# Multiple pairwise comparisons between species strongly mislead functional genomics studies
-
-Casey W. Dunn^1^*, Felipe Zapata^1,2^, Catriona Munro^1^, Stefan Siebert^1,3^, Andreas Hejnol^4^
-
-^1^ Department of Ecology and Evolutionary Biology, Brown University, Providence, RI, USA
-
-^2^ Current address: Department of Ecology and Evolutionary Biology, University of California Los Angeles, Los Angeles, CA, USA
-
-^3^ Current address: Department of Molecular & Cellular Biology, University of California at Davis, Davis, CA, USA
-
-^4^ Sars International Centre for Marine Molecular Biology, University of Bergen, Bergen, Norway
-
-\* Corresponding author, casey_dunn@brown.edu
-
-## Abstract
-
-Functional genomics, the study of how genomes work, is one of the most rapidly advancing fields in biology.
-There is now considerable interest in comparing genome function across species to understand what is conserved, what is variable, and how the diversity of organism traits (including anatomy, physiology, and development) relates to genome diversity.
-Most published comparisons of genome function across species have relied on multiple pairwise comparisons between species, an approach that neglects to account for the evolutionary relationships among species.
-This has two potential negative impacts.
-First, it can lead to the wrong conclusions about the evolution of genome function.
-Second, it is a missed opportunity to learn about biology that can only be understood in an explicit historical context that considers these relationships.
-Here we examine two recently published comparative gene expression studies that used multiple pairwise comparisons rather than phylogenetic approaches to investigate evolutionary patterns of gene expression.
-We find problems with the pairwise comparisons used in both studies that undermine their published conclusions.
-The first study we examine found that expression tends to be more different between paralogs than orthologs, and interpreted this as evidence of the ortholog conjecture- the hypothesis that there are greater rates of evolutionary change following duplication.
-Our phylogenetic analysis finds no difference in rates following duplication and speciation.
-Instead, the greater difference between paralogs is because they tend to be more distantly related to each other than orthologs are to each other.
-The second study claimed to find more evolutionary variance in gene expression at a midpoint of embryonic development than earlier and later when comparing distantly related animals.
-We instead find that changes along the branch of a single taxon, the ctenophore, were over counted because they were included in each pairwise comparison with this species.
-These reanalyses are concrete demonstrations of the inadequacy of pairwise comparisons in comparative genomics studies, including gene expression analyses, and indicates that it will be critical to adopt phylogenetic comparative methods in future work.
-
-## Introduction
-
-The focus of genomics research has quickly shifted from describing genome sequences to functional genomics, the study of how genomes "work" using tools that measure functional attributes such as expression, chromatin state, and transcription initiation.
-Functional genomics, in turn, is now becoming more comparative- there is great interest in understanding how functional genomic variation across species gives rise to a diversity of development, morphology, physiology, and other phenotypes [@Wray:2013gh].
-Comparative Functional Genomics (CFG) analyses are also critical to transferring functional insight across species, and will grow in importance in coming years.
-
-A rich theoretical and statistical methodology of phylogenetic comparative methods have been developed over the last three decades to address the challenges and opportunities of comparisons across species [@Felsenstein:1985ua; @Grafen:1989um; @Pagel:1999gc; Revell:2012jd; @FitzJohn:2012ey; @Uyeda:2014jl; @Garamszegi:2014ew].
-A central challenge is the dependence of observations across species due to the evolutionary history of species -- more closely related species share many traits that evolve once in a common ancestor.
-This violates the fundamental assumption of independence of observations in standard statistical methods.
-Phylogenetic comparative methods have largely been applied to the evolution of morphological and ecological traits, but are just as relevant to CFG.
-Most CFG studies have abstained from phylogenetic approaches and instead heavily rely on multiple pairwise comparisons (Fig.
-1A) between contemporary traits that do not account for the fact that evolutionary relationships explain much of the structure of variation across species.
-This leaves CFG studies susceptible to serious statistical errors and is a missed opportunity to ask questions that are only accessible in an explicit phylogenetic context.
-
-One reason that CFG has not yet embraced phylogenetic approaches is that it has not yet been concretely demonstrated that pairwise and phylogenetic approaches can lead to different results in CFG studies.
-In this manuscript we examine this issue.
-Although the value of phylogenetic approaches has been repeatedly shown in comparative analyses of other types of character data [@Ricklefs:1996wb; @Chamberlain:2012vx; @OMeara:2012gq], this value is apparently not well known to the functional genomics community since the field has until recently largely focused on within-species analyses.
-Many initial CFG studies have also been focused on only two species at a time, where the differences between phylogenetic and pairwise approaches are not as pronounced.
-As the number of species considered in each CFG study grows, the shortcomings of pairwise approaches will become far more acute.
-
-Most CFG studies to date compare gene expression across species, often between distinct tissues or developmental time points.
-Gene expression has been a popular target of CFG studies because, with the advent of RNA-Seq and related transcriptome sequencing methods, it is the most readily collected functional genomics trait across diverse species.
-Here we reconsider two recent comparative analyses of gene expression data [@KryuchkovaMostacci:2016iw; @Levin:2016bp].
-
-The first study, Kryuchkova-Mostacci and Robinson-Rechavi (KMRR) 2016 [@KryuchkovaMostacci:2016iw], analyzed multiple vertebrate expression datasets to test the ortholog conjecture - the hypothesis that orthologs tend to have more conserved attributes (specificity of expression across organs in this case) than do paralogs [@Nehrt:2011bm, @Koonin:2005fp].
-Using pairwise comparisons (Fig.
-1A), they found lower expression correlation between paralogs than between orthologs and interpreted this as strong support for the ortholog conjecture.
-
-The second CFG study we evaluate here is Levin *et al.* [@Levin:2016bp].
-This study analyzed gene expression through the course of embryonic development for ten animal species, each from a different phylum.
-Using pairwise comparisons (Figure 1A), they found there is more evolutionary variance in gene expression at a mid phase of development than there is at early and late phases for these ten animals (Fig.
-XX).
-They suggest that this supports an "inverse hourglass" model for the evolution of gene expression, in contrast with the "hourglass" model previously proposed for closely related species [@Kalinka:2010dw], and provides biological justification for the concept of phyla and may provide an operational definition of phyla.
-We previously described concerns with these interpretations [@Hejnol:2016hw].
-Here we directly address the analyses themselves by examining the structure of the pairwise comparisons.
-
-![Figure Overview](Figure_overview.png?raw=true)
-
-> Figure 1 | Pairwise and phylogenetic comparative approaches.
-(a) Many comparative functional genomic studies rely on pairwise comparisons, where traits of each gene are compared to traits of other genes (such as orthologs and paralogs).
-This leads to many more comparisons than unique observations, making each comparison dependent on others.
-(b) Comparative phylogenetic methods, including phylogenetic independent contrasts [@Felsenstein:1985ua], make a smaller number of independent comparisons, where each contrast measures independent changes along different branches.
-Phylogenetic approaches are rarely used for functional genomics studies.
-
-## Results and Discussion
-
-### KMRR Reanalysis
-
-#### Original pairwise test of the ortholog conjecture
-
-The KMRR study [@KryuchkovaMostacci:2016iw] sought to test the ortholog conjecture.
-The ortholog conjecture [@Nehrt:2011bm] is the proposition that orthologs (genes that diverged from each other due to a speciation event) have more similar attributes than do paralogs (genes that diverged from each other due to a gene duplication event).
-The ortholog conjecture has important biological and technical implications.
-It shapes our understanding of the functional diversity of gene families.
-It is used to relate findings from well-studied genes to related genes that have not been investigated in detail.
-It can be applied to any trait of genes, from gene sequence to biochemical properties to expression.
-While the ortholog conjecture describes a specific pattern of functional diversity across genes, it is also articulated as a hypothesis about the process of evolution-- that there is greater evolutionary change in gene attributes following a duplication event than a speciation event.
-
-Despite its importance, there have been relatively few tests of the ortholog conjecture.
-Previous work has shown that ontology annotations are not sufficient to test the ortholog conjecture [@Chen:2012jz; @Thomas:2012hz].
-Analyses of domain structure were consistent with the ortholog conjecture [@Forslund:2011cw].
-There have been few tests of the ortholog conjecture with regards to gene expression [@Chen:2012jz], and KMRR is the most thorough such expression study to date.
-
-The KMRR study considered several publicly available datasets.
-Here we consider their conclusions in the context of one of these studies, Brawand *et al.* 2011 [@Brawand:2011du].
-This dataset of gene expression for six organs across ten species (nine mammals and one bird), eight of which were analyzed by KMRR, is the best sampled in their analyses.
-Their expression summary statistic is Tau [@Yanai:2005dh, @KryuchkovaMostacci:2016in], an indicator of tissue specificity of gene expression.
-It can range from a value of 0, which indicates no specificity (*i.e.*, uniform expression across tissues), to a value of 1, which indicates high specificity (*i.e.*, expression in only one tissue).
-It is convenient in that it is a single number of defined range for each gene, though of course since the original expression is multidimensional this means much information is discarded.
-This includes information about which tissue expression is specific to.
-For example, if one gene has expression specific to the brain and another expression specific to the kidney, both would have a Tau of 1.
-
-Results in KMRR are based on pairwise comparisons (Figure 1A) between Tau within each gene family.
-They found the correlation coefficient of Tau for orthologs to be significantly greater than the correlation coefficient of Tau for paralogs, *i.e.* that orthologs tend to be more similar to each other than paralogs.
-From this they concluded that their analyses support the ortholog conjecture.
-They also concluded that this pattern provides support for a particular evolutionary process, that "tissue-specificity evolves very slowly in the absence of duplication, while immediately after duplication the new gene copy differs".
-
-#### Phylogenetic reanalyses
-
-We reanalyzed the the KMRR study using phylogenetic comparative methods, focusing on the dataset of Brawand *et al.* 2011 [@Brawand:2011du].
-For each internal node in each gene tree (Figure 1b), we measured the phylogenetic independent contrast (PIC) [@Felsenstein:1985ua] of the expression statistic Tau.
-This is the difference in values of Tau for descendant nodes scaled by the expected variance (determined largely by the lengths of these branches).
-These contrasts were then annotated by whether they are made across a speciation or duplication event.
-The original description of independent contrasts [@Felsenstein:1985ua] focused on assessing covariance between changes in two traits on a tree.
-Our use of contrasts is a bit different-- we look for differences in evolutionary changes of one trait (differential expression) between two categories of nodes (speciation and duplication) on a tree.
-
-```{r expectation_plots, echo=FALSE, cache=TRUE, message=F, warning=F}
-
+	#' Simulate a set of pics from a gamma distribution
+	#' 
+	#' @param scale Gamma scale parameter
+	#' @param n Number of observations per event type	
+	#' @return A dataframe
+	#' @export
 	sim_expectation = function( scale = 1, n = 10000 ){
@@
 	}
+
+```
+
+# Multiple pairwise comparisons between species strongly mislead functional genomics studies
+
+Casey W. Dunn^1^*, Felipe Zapata^1,2^, Catriona Munro^1^, Stefan Siebert^1,3^, Andreas Hejnol^4^
+
+^1^ Department of Ecology and Evolutionary Biology, Brown University, Providence, RI, USA
+
+^2^ Current address: Department of Ecology and Evolutionary Biology, University of California Los Angeles, Los Angeles, CA, USA
+
+^3^ Current address: Department of Molecular & Cellular Biology, University of California at Davis, Davis, CA, USA
+
+^4^ Sars International Centre for Marine Molecular Biology, University of Bergen, Bergen, Norway
+
+\* Corresponding author, casey_dunn@brown.edu
+
+## Abstract
+
+There is now considerable interest in comparing functional genomics data across species, with the hope of providing an integrated understanding of genome and phenotype evolution.
+Most such studies have relied on multiple pairwise comparisons between species, an approach that does not incoporate information about the evolutionary relationships among species.
+This has the potential to mislead conclusions about the evolution of genome function, and is a missed opportunity to learn about biology that can only be understood in an explicit phylogenetic context.
+Here we examine two recently published studies that compare expression across species with pairwise methods.
+We find problems with both that call their conclusions into question.
+One study interpreted higher expression correlation between orthologs than paralogs as evidence of the ortholog conjecture, *i.e.* the hypothesis that gene function evolution is more rapid after duplication than speciation.
+Instead, we find that this pattern is due to the structure of the gene phylogenies, not different rates of expression evolution, and is the expected pattern when rates are the same following duplication and speciation.
+The second study interpreted pairwise comparisons of embryonic gene expression across distantly related animals as evidence for a distinct evolutionary process that gave rise to animal phyla.
+Instead, we find that the pattern they identified is due to unique feature of a single species that impact multiple pairwise comparisons that include that species.
+In each study, distinct patterns of pairwise similarity among species were interpreted as evidence of particular evolutionary processes, but instead reflect the structure of the phylogenetic trees themselves.
+These reanalyses concretely demonstrate the inadequacy of pairwise comparisons for analyzing functional genomic data across species, and indicate that it will be critical to adopt phylogenetic comparative methods in future work.
+Fortunately, phylogenetic comparative biology is also a rapidly advancing field with many methods that can be directly applied to functional genomic data.
+
+## Introduction
+
+The focus of genomics research has quickly shifted from describing genome sequences to functional genomics, the study of how genomes "work" using tools that measure functional attributes such as expression, chromatin state, and transcription initiation.
+Functional genomics, in turn, is now becoming more comparative-- there is great interest in understanding how functional genomic variation across species gives rise to a diversity of development, morphology, physiology, and other phenotypes [@Wray:2013gh].
+These analyses are also critical to transferring functional insight across species, and will grow in importance in coming years.
+
+A rich theoretical and statistical methodology of phylogenetic comparative methods have been developed over the last three decades to address the challenges and opportunities of comparisons across species [@Felsenstein:1985ua; @Grafen:1989um; @Pagel:1999gc; Revell:2012jd; @FitzJohn:2012ey; @Uyeda:2014jl; @Garamszegi:2014ew].
+A central challenge is the dependence of observations across species due to the evolutionary history of species -- more closely related species share many traits that evolve once in a common ancestor.
+This violates the fundamental assumption of independence of observations in standard statistical methods.
+Phylogenetic comparative methods have largely been applied to the evolution of morphological and ecological traits, but are just as relevant to functional genomics.
+Most Comparative Functional Genomics (CFG) studies have abstained from phylogenetic approaches and instead rely on multiple pairwise comparisons (Fig.
+1A).
+These do not account for the fact that evolutionary relationships are expected to explain much of the structure of variation across species.
+This leaves CFG studies susceptible to statistical problems and is a missed opportunity to ask questions that are only accessible in an explicit phylogenetic context.
+
+Phylogenetic comparative methods explicitly model trait change along the branches of evolutionary trees.
+The value of these methods relative to to pairwise comparisons has been repeatedly shown in analyses of other types of character data [@Ricklefs:1996wb; @Chamberlain:2012vx; @OMeara:2012gq].
+One reason that CFG has not yet embraced phylogenetic approaches is that it has not yet been concretely demonstrated that pairwise and phylogenetic comparative methods can lead to different results when considering functional genomics data.
+Here we examine this issue be re-evaluating the pairwise comparisons used in two recent studies that compared gene expression across species.
+
+The first study, Kryuchkova-Mostacci and Robinson-Rechavi (KMRR) 2016 [@KryuchkovaMostacci:2016iw], analyzed multiple vertebrate expression datasets to test the ortholog conjecture - the hypothesis that orthologs tend to have more conserved attributes (specificity of expression across organs in this case) than do paralogs [@Nehrt:2011bm, @Koonin:2005fp].
+Using pairwise comparisons (Fig.
+1A), they found lower expression correlation between paralogs than between orthologs and interpreted this as strong support for the ortholog conjecture.
+
+The second CFG study we evaluate here is Levin *et al.* [@Levin:2016bp].
+This study analyzed gene expression through the course of embryonic development for ten animal species, each from a different phylum.
+Using pairwise comparisons (Figure 1A), they found there is more evolutionary variance in gene expression at a mid phase of development than there is at early and late phases for these ten animals (Fig.
+XX).
+They suggest that this supports an "inverse hourglass" model for the evolution of gene expression, in contrast with the "hourglass" model previously proposed for closely related species [@Kalinka:2010dw], and provides biological justification for the concept of phyla.
+We previously described concerns with the interpretations of this result [@Hejnol:2016hw].
+Here we directly address the analyses themselves by examining the structure of the pairwise comparisons.
+
+![Figure Overview](Figure_overview.png?raw=true)
+
+> Figure 1 | Pairwise and phylogenetic comparative approaches.
+(a) Many comparative functional genomic studies rely on pairwise comparisons, where traits of each gene are compared to traits of other genes (such as orthologs and paralogs).
+This leads to many more comparisons than unique observations, making each comparison dependent on others.
+(b) Comparative phylogenetic methods, including phylogenetic independent contrasts [@Felsenstein:1985ua], make a smaller number of independent comparisons, where each contrast measures independent changes along different branches.
+Phylogenetic approaches are rarely used for functional genomics studies.
+
+## Results and Discussion
+
+### KMRR Reanalysis
+
+#### Original pairwise test of the ortholog conjecture
+
+The KMRR study [@KryuchkovaMostacci:2016iw] sought to test the ortholog conjecture.
+The ortholog conjecture [@Nehrt:2011bm] is the proposition that orthologs (genes that diverged from each other due to a speciation event) have more similar attributes than do paralogs (genes that diverged from each other due to a gene duplication event).
+The ortholog conjecture has important biological and technical implications.
+It shapes our understanding of the functional diversity of gene families.
+It is used to relate findings from well-studied genes to related genes that have not been investigated in detail.
+It can be applied to any trait of genes, from gene sequence to biochemical properties to expression.
+While the ortholog conjecture describes a specific pattern of functional diversity across genes, it is also articulated as a hypothesis about the process of evolution-- that there is greater evolutionary change in gene attributes following a duplication event than a speciation event.
+
+Despite its importance, there have been relatively few tests of the ortholog conjecture.
+Previous work has shown that ontology annotations are not sufficient to test the ortholog conjecture [@Chen:2012jz; @Thomas:2012hz].
+Analyses of domain structure were consistent with the ortholog conjecture [@Forslund:2011cw].
+There have been few tests of the ortholog conjecture with regards to gene expression [@Chen:2012jz], and KMRR is the most thorough such expression study to date.
+
+The KMRR study considered several publicly available datasets.
+Here we consider their conclusions in the context of one of these studies, Brawand *et al.* 2011 [@Brawand:2011du].
+This dataset of gene expression for six organs across ten species (nine mammals and one bird), eight of which were analyzed by KMRR, is the best sampled in their analyses.
+Their expression summary statistic is Tau [@Yanai:2005dh, @KryuchkovaMostacci:2016in], an indicator of tissue specificity of gene expression.
+It can range from a value of 0, which indicates no specificity (*i.e.*, uniform expression across tissues), to a value of 1, which indicates high specificity (*i.e.*, expression in only one tissue).
+It is convenient in that it is a single number of defined range for each gene, though of course since the original expression is multidimensional this means much information is discarded.
+This includes information about which tissue expression is specific to.
+For example, if one gene has expression specific to the brain and another expression specific to the kidney, both would have a Tau of 1.
+
+Results in KMRR are based on pairwise comparisons (Figure 1A) between Tau within each gene family.
+They found the correlation coefficient of Tau for orthologs to be significantly greater than the correlation coefficient of Tau for paralogs, *i.e.* that orthologs tend to be more similar to each other than paralogs.
+From this they concluded that their analyses support the ortholog conjecture.
+They also concluded that this pattern provides support for a particular evolutionary process, that "tissue-specificity evolves very slowly in the absence of duplication, while immediately after duplication the new gene copy differs".
+
+#### Phylogenetic reanalyses
+
+We reanalyzed the the KMRR study using phylogenetic comparative methods, focusing on the dataset of Brawand *et al.* 2011 [@Brawand:2011du].
+For each internal node in each gene tree (Figure 1b), we measured the phylogenetic independent contrast (PIC) [@Felsenstein:1985ua] of the expression statistic Tau.
+This is the difference in values of Tau for descendant nodes scaled by the expected variance (determined largely by the lengths of these branches).
+These contrasts were then annotated by whether they are made across a speciation or duplication event.
+The original description of independent contrasts [@Felsenstein:1985ua] focused on assessing covariance between changes in two traits on a tree.
+Our use of contrasts is a bit different-- we look for differences in evolutionary changes of one trait (differential expression) between two categories of nodes (speciation and duplication) on a tree.
+
+```{r expectation_plots, echo=FALSE, cache=TRUE, message=F, warning=F}
 
@@
 
-In addition to the problems with the pairwise comparisons discussed above, we previously described several other concerns with the experimental design, analysis, and interpretation of the results of this study [@Hejnol:2016hw].
-For instance, because Levin *et al.* [@Levin:2016bp] only sampled one species per phylum, there is no way to tell if observed differences are unique to the branches that gave rise to clades designated as phyla or occurred along the many other branches in the tree.
-Moreover, the goodness of fit of each gene having expression peaking at three phases was not considered, and alternative patterns, such as the possibility that there are more phases of expression, were not evaluated.
-
 ### Phylogenetic comparative methods in functional genomics
@@
 
+	calibration_times_focal = 
+		calibration_times %>%
+		filter( clade %in% focal_calibrations_clades ) %>%
+		arrange( age )
+
 	noised_dates = 
@@
 			noised_replicate_n, 
-			add_noise_calibration_times( calibration_times, sd_fraction ),
+			add_noise_calibration_times( calibration_times_focal, sd_fraction ),
 			simplify = FALSE

### 2017-02-04 manuscript.rmd
Update manuscript.rmd
Fixes a couple of typos

@@
 
-We reanalyzed the the KMRR study using phylogenetic comparative methods, focusing on the dataset of Brawand *et al.* 2011 [@Brawand:2011du].
+We reanalyzed the KMRR study using phylogenetic comparative methods, focusing on the dataset of Brawand *et al.* 2011 [@Brawand:2011du].
 For each internal node in each gene tree (Figure 1b), we measured the phylogenetic independent contrast (PIC) [@Felsenstein:1985ua] of the expression statistic Tau.
@@
 We therefore examined only the contrasts that fell within in the range of expected variance seen for speciation contrasts, excluding duplication contrasts that fall outside of this range.
-This reanalysis does not reject the ortholog conjecture either (Figure XXAscertainment d), indicating that branch length bias is not responsible for the result.
+This reanalysis does not reject the null hypothesis either (Figure XXAscertainment d), indicating that branch length bias is not responsible for the result.
 

### 2017-02-05 levin_etal/reanalyses.rmd
minor tidying of levin et al reanalysis

@@
 ```{r setup, echo=FALSE, warning=FALSE, message=FALSE}
-
+	library(ggplot2)
+	library(tidyverse)
+	library(grid)
 ```
@@
 	w_late = wilcox.test(D$Late, D$Mid, alternative="greater", paired=T)$p.value
+
 ```
@@
 ```{r sorted_mid}
+
 	D[order(D[,4]),]
@@
 ```{r no_cteno}
+
 	D_no_ct = D[ ((D$Species_A != "Ctenophores") &   (D$Species_B != "Ctenophores")), ]
@@
 	multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
-  library(grid)
 
@@
 
-library(ggplot2)
-library(reshape2)
-library(grid)
-
 	limits=c(-0.4, 0.75)
 
-Dm = melt(D, id=c("Species_A", "Species_B"), variable.name = "phase", value.name = "similarity")
-Dm_no_ct = melt(D_no_ct, id=c("Species_A", "Species_B"), variable.name = "phase", value.name = "similarity")
+	phase_levels = c( "Early", "Mid", "Late" )
+
+	Dm = D %>% 
+		gather( Early, Mid, Late, key="phase", value="similarity" ) %>%
+		mutate( phase=factor( phase, levels=phase_levels ) )
+
+	Dm_no_ct =  D_no_ct %>% 
+		gather( Early, Mid, Late, key="phase", value="similarity" ) %>%
+		mutate( phase=factor( phase, levels=phase_levels ) )
 
@@
 
-pa = ggplot(Dm, aes(phase, similarity)) +
+	pa = Dm %>% 
+		ggplot( aes(x=phase, y=similarity) ) +
+	    geom_boxplot(outlier.shape=NA) + # Outliers are drawn in bellow with jitter, so supress them here so they aren't duplicated
+	    scale_y_continuous(limits=limits) +
+	    geom_jitter(shape=16, position=position_jitter(0.2), alpha=alpha, color=color) + 
+			coord_flip() + 
+			ggtitle("(a) Original, with ctenophore in red")
+
+	pb = Dm_no_ct %>% 
+		ggplot( aes(x=phase, y=similarity)) +
 	    geom_boxplot(outlier.shape=NA)+ # Outliers are drawn in bellow with jitter, so supress them here so they aren't duplicated
 	    scale_y_continuous(limits=limits) +
-    geom_jitter(shape=16, position=position_jitter(0.2), alpha=alpha, color=color)
-
-pa = pa + coord_flip() + ggtitle("(a) Original, with ctenophore in red")
-
-pb = ggplot(Dm_no_ct, aes(phase, similarity)) +
-    geom_boxplot(outlier.shape=NA)+ # Outliers are drawn in bellow with jitter, so supress them here so they aren't duplicated
-    scale_y_continuous(limits=limits) +
-    geom_jitter(shape=16, position=position_jitter(0.2), alpha=0.4)
-
-pb = pb + coord_flip() + ggtitle("(b) New, without ctenophore")
-
-#pa
-#ggplotly()
+	    geom_jitter(shape=16, position=position_jitter(0.2), alpha=0.4) + 
+			coord_flip() + 
+			ggtitle("(b) New, without ctenophore")
 
@@
 ```{r DC_pre}
+
 	# Sample the columns for these two species
@@
 	nrow(DC)
+
 ```
@@
 ```{r frac_4c_ce}
+
 	colnames(DC)[1]
@@
 	ce
+
 ```
@@
 ```{r frac_4c_dm}
+
 	colnames(DC)[2]
@@
 	dm
+
 ```
@@
 
+### Matrix occupancy
+
+```{r gene_phase_matrix}
+
+	n_species_sampled = apply( Phase, 1, function(x) sum( x > 0 ) )
+
+	phase_sub = Phase[ n_species_sampled >= 2, ]
+
+	hist( n_species_sampled )
+
+```
+
+There were `r nrow(Phase)` gene in the original data file.
+`r sum(n_species_sampled == 0)` of these had no phase annotations for any species.
+`r nrow(phase_sub)` had phase annotations for 2 or more species.
+
 ## Software versions

### 2017-02-06 manuscript.rmd
made a complete pass at text, now compute BM model parameters once

@@
 
+	#' Estimate Tau evolution parameters from a tree::treedata tree
+	#' and add them back into the tree object.
+	#' 
+	#' @param nhx A phylogenetic tree and associated Tau values 
+	#' as a treeio::treedata object
+	#' @param ... Additional arguments to pass to geiger::fitContinuous()
+	#' @return A phylogenetic trees as a treeio::treedata object, 
+	#' with estimated brownian motion model parameters stored 
+	#' in @phylo$model
+	#' @export
+	add_model_parameters = function( nhx, ... ) {
+		phy = nhx@phylo
+		tau_original = nhx@data$Tau [ 1:length( phy$tip.label ) ]
+		names( tau_original ) = phy$tip.label
+		nhx@phylo$model = fitContinuous( phy, tau_original, ... )
+
+		return( nhx )
+	}
+
 	#' Simulate Tau on a tree::treedata. Existing observed values are
@@
 
-		# Estimate parameters
-		tau_original = nhx@data$Tau [ 1:length( phy$tip.label ) ]
-		names( tau_original ) = phy$tip.label
-		brownian_model = fitContinuous( phy, tau_original )
+		# Get parameter estimates
+		brownian_model = nhx@phylo$model
 
@@
 
-There is now considerable interest in comparing functional genomics data across species, with the hope of providing an integrated understanding of genome and phenotype evolution.
+With the goal of providing an integrated understanding of genome and phenotype evolution, there is now considerable interest in comparing functional genomics data across species.
 Most such studies have relied on multiple pairwise comparisons between species, an approach that does not incoporate information about the evolutionary relationships among species.
-This has the potential to mislead conclusions about the evolution of genome function, and is a missed opportunity to learn about biology that can only be understood in an explicit phylogenetic context.
+These pairwise approaches have the potential to mislead conclusions about the evolution of genome function, and are a missed opportunity to learn about biology that can only be understood in an explicit phylogenetic context.
 Here we examine two recently published studies that compare expression across species with pairwise methods.
@@
 Instead, we find that the pattern they identified is due to unique feature of a single species that impact multiple pairwise comparisons that include that species.
-In each study, distinct patterns of pairwise similarity among species were interpreted as evidence of particular evolutionary processes, but instead reflect the structure of the phylogenetic trees themselves.
-These reanalyses concretely demonstrate the inadequacy of pairwise comparisons for analyzing functional genomic data across species, and indicate that it will be critical to adopt phylogenetic comparative methods in future work.
+In each study, distinct patterns of pairwise similarity among species were interpreted as evidence of particular evolutionary processes, but instead reflect the structure of phylogenies.
+These reanalyses concretely demonstrate the inadequacy of pairwise comparisons for analyzing functional genomic data across species, and indicate that it will be critical to adopt phylogenetic comparative methods in future funcitonal genomics work.
 Fortunately, phylogenetic comparative biology is also a rapidly advancing field with many methods that can be directly applied to functional genomic data.
@@
 
-A rich theoretical and statistical methodology of phylogenetic comparative methods have been developed over the last three decades to address the challenges and opportunities of comparisons across species [@Felsenstein:1985ua; @Grafen:1989um; @Pagel:1999gc; Revell:2012jd; @FitzJohn:2012ey; @Uyeda:2014jl; @Garamszegi:2014ew].
+A rich theoretical and statistical methodology of phylogenetic comparative methods have been developed over the last three decades to address the challenges and opportunities of trait comparisons across species [@Felsenstein:1985ua; @Grafen:1989um; @Pagel:1999gc; Revell:2012jd; @FitzJohn:2012ey; @Uyeda:2014jl; @Garamszegi:2014ew].
 A central challenge is the dependence of observations across species due to the evolutionary history of species -- more closely related species share many traits that evolve once in a common ancestor.
-This violates the fundamental assumption of independence of observations in standard statistical methods.
-Phylogenetic comparative methods have largely been applied to the evolution of morphological and ecological traits, but are just as relevant to functional genomics.
-Most Comparative Functional Genomics (CFG) studies have abstained from phylogenetic approaches and instead rely on multiple pairwise comparisons (Fig.
+This violates the fundamental assumption of observation independence in standard statistical methods.
+Phylogenetic comparative methods address this dependence.
+They have largely been applied to morphological and ecological traits, but are just as relevant to functional genomics.
+Even so, most comparative functional genomics studies have abstained from phylogenetic approaches and instead rely on multiple pairwise comparisons (Fig.
 1A).
-These do not account for the fact that evolutionary relationships are expected to explain much of the structure of variation across species.
-This leaves CFG studies susceptible to statistical problems and is a missed opportunity to ask questions that are only accessible in an explicit phylogenetic context.
-
-Phylogenetic comparative methods explicitly model trait change along the branches of evolutionary trees.
+This leaves comparative functional genomics studies susceptible to statistical problems and is a missed opportunity to ask questions that are only accessible in an explicit phylogenetic context.
+
+Phylogenetic comparative methods account for evolutionary history and explicitly model trait change along the branches of evolutionary trees.
 The value of these methods relative to to pairwise comparisons has been repeatedly shown in analyses of other types of character data [@Ricklefs:1996wb; @Chamberlain:2012vx; @OMeara:2012gq].
-One reason that CFG has not yet embraced phylogenetic approaches is that it has not yet been concretely demonstrated that pairwise and phylogenetic comparative methods can lead to different results when considering functional genomics data.
+The application of these methods is illustrated by a study of leaf functional attributes for about 100 plant species [@Ackerly:1999iu].
+If phylogenetic relationships are not considered, analyses indicate a negative correlation of leaf lifespan with leaf size.
+With the application of phylogenetic comparative methods, however, this correlation disappears.
+This is because most of the observed variance is due to differences between conifers and flowering plants.
+Phylogenetic comparative methods capture this shift as change along a single branch deep in the tree, revealing that there is not a tendancy of correlated change between these traits across the phylogeny.
+One reason that comparative functional genomics has not yet embraced phylogenetic approaches is that it has not yet been concretely demonstrated that pairwise and phylogenetic comparative methods can lead to different results when considering functional genomics data.
 Here we examine this issue be re-evaluating the pairwise comparisons used in two recent studies that compared gene expression across species.
@@
 
-The second CFG study we evaluate here is Levin *et al.* [@Levin:2016bp].
+The second comparative functional genomics study we evaluate here is Levin *et al.* [@Levin:2016bp].
 This study analyzed gene expression through the course of embryonic development for ten animal species, each from a different phylum.
-Using pairwise comparisons (Figure 1A), they found there is more evolutionary variance in gene expression at a mid phase of development than there is at early and late phases for these ten animals (Fig.
+Using pairwise comparisons (Figure 1A), they found there is more evolutionary variance in gene expression at a mid phase of development than there is at early and late phases (Fig.
 XX).
@@
 We previously described concerns with the interpretations of this result [@Hejnol:2016hw].
-Here we directly address the analyses themselves by examining the structure of the pairwise comparisons.
+Here we address the analyses themselves by examining the structure of the pairwise comparisons.
 
@@
 
-The KMRR study considered several publicly available datasets.
-Here we consider their conclusions in the context of one of these studies, Brawand *et al.* 2011 [@Brawand:2011du].
-This dataset of gene expression for six organs across ten species (nine mammals and one bird), eight of which were analyzed by KMRR, is the best sampled in their analyses.
+The KMRR study considered several publicly available datasets of gene expression across tissues and species.
 Their expression summary statistic is Tau [@Yanai:2005dh, @KryuchkovaMostacci:2016in], an indicator of tissue specificity of gene expression.
@@
 
-Results in KMRR are based on pairwise comparisons (Figure 1A) between Tau within each gene family.
+The KMRR analyses are based on pairwise comparisons (Figure 1A) between Tau within each gene family.
 They found the correlation coefficient of Tau for orthologs to be significantly greater than the correlation coefficient of Tau for paralogs, *i.e.* that orthologs tend to be more similar to each other than paralogs.
 From this they concluded that their analyses support the ortholog conjecture.
-They also concluded that this pattern provides support for a particular evolutionary process, that "tissue-specificity evolves very slowly in the absence of duplication, while immediately after duplication the new gene copy differs".
+They also concluded that this pattern provides support for a particular evolutionary process, that "tissue-specificity evolves very slowly in the absence of duplication, while immediately after duplication the new gene copy differs" [@KryuchkovaMostacci:2016iw].
 
@@
 
-We reanalyzed the KMRR study using phylogenetic comparative methods, focusing on the dataset of Brawand *et al.* 2011 [@Brawand:2011du].
-For each internal node in each gene tree (Figure 1b), we measured the phylogenetic independent contrast (PIC) [@Felsenstein:1985ua] of the expression statistic Tau.
-This is the difference in values of Tau for descendant nodes scaled by the expected variance (determined largely by the lengths of these branches).
-These contrasts were then annotated by whether they are made across a speciation or duplication event.
-The original description of independent contrasts [@Felsenstein:1985ua] focused on assessing covariance between changes in two traits on a tree.
-Our use of contrasts is a bit different-- we look for differences in evolutionary changes of one trait (differential expression) between two categories of nodes (speciation and duplication) on a tree.
+We reanalyzed the KMRR study using phylogenetic comparative methods.
+We focused on one of the datasets included in their analyses, that of Brawand *et al.* 2011 [@Brawand:2011du].
+This dataset is the best sampled in their analyses.
+It has gene expression data for six organs across ten species (nine mammals and one bird), eight of which were analyzed by KMRR and further considered here.
 
@@
 
-To test the ortholog conjecture it is essential to have a null hypothesis that makes predictions distinct from the ortholog conjecture.
-A suitable null hypothesis is that there is no difference in evolutionary change in differential expression along branches that descend from speciation or duplication events [@Studer:2009bm] (Figure 1b).
-Under this hypothesis, we would predict that contrasts across speciation nodes and duplication nodes are drawn from the same distribution (Figure XXEXPECTATIONS a).
-Under the alternative hypothesis specified by the ortholog conjecture, that there is a higher rate of change following duplication events than speciation events, we would expect to see the distribution of duplication contrasts shifted to higher values relative to the speciation contrasts (Figure XXEXPECTATIONS b).
+For each internal node in each gene tree, we measured the phylogenetic independent contrast (PIC) [@Felsenstein:1985ua] of the expression statistic Tau.
+This is the difference in values of Tau for descendant nodes scaled by the expected variance, which is largely determined by the lengths of the two branches that connect the node to its two descendants (Figure 1b).
+These contrasts were then annotated by whether each is made across a speciation or duplication event.
+The original description of independent contrasts [@Felsenstein:1985ua] focused on assessing covariance between changes in two traits on a tree.
+Our use of contrasts is a bit different-- we look for differences in evolutionary changes of one trait (differential expression) between two categories of nodes (speciation and duplication) on a tree.
 
@@
 
-When we reanalyze the data with phylogenetic independent contrasts, we did not find increased evolutionary change in expression following duplication events (Fig.
+We mapped the Tau values calculated by KMRR [@KryuchkovaMostacci:2016iw] for the Brawand *et al.* 2011 [@Brawand:2011du] datset onto `r length(gene_trees)` gene trees parsed from ENSMBL Compara [@Herrero:2016jj].
+These are the same pre-computed trees that the orthology/ paralogy annotations KMRR used are based on.
+`r length(gene_trees_pruned)` passed taxon sampling criteria (`r min_genes_with_expression` genes) after removing tips without Tau values and had at least one speciation event.
+Of these, `r length(gene_trees_calibrated)` were successfully time calibrated.
+These calibrated trees were used to calculate phylogenetic independent contrasts for  `r nodes_contrast %>% filter(D=="Y") %>% nrow()` duplication nodes and `r nodes_contrast %>% filter(D=="N") %>% nrow()` speciation nodes.
+
+It is essential to have a null hypothesis that makes a distinct prediction from the prediction of the hypothesis under consideration.
+A suitable null hypothesis in this case  is that there is no difference in the evolution of expression following speciation or duplication events [@Studer:2009bm].
+Under this hypothesis, we would predict that contrasts across speciation nodes and duplication nodes are drawn from the same distribution (Figure XXEXPECTATIONS a).
+Under the alternative hypothesis specified by the ortholog conjecture, that there is a higher rate of change following duplication events than speciation events, we would expect to see the distribution of duplication contrasts shifted to higher values relative to the speciation contrasts (Figure XXEXPECTATIONS b).
+
+We did not find increased evolutionary change in expression following duplication events (Fig.
 XXKMRR).
-Of the `r length(gene_trees)` trees that were parsed from ENSMBL Compara [@Herrero:2016jj], `r length(gene_trees_pruned)` passed taxon sampling criteria (`r min_genes_with_expression` genes) after removing tips without Tau values and had at least one speciation event.
-Of these, `r length(gene_trees_calibrated)` were successfully time calibrated.
-These calibrated pruned trees were used to calculate phylogenetic independent contrasts for Tau.
-They contain `r nodes_contrast %>% filter(D=="Y") %>% nrow()` duplication nodes and `r nodes_contrast %>% filter(D=="N") %>% nrow()` speciation nodes.
-
-Given the ortholog conjecture, we would expect to see larger contrasts following duplication events than speciation events.
-Instead, we find no such distribution (Figure XXKMRR).
 The Wilcoxon rank test does not reject the null hypothesis that the rate of evolution following duplications is the same as or less than the rate following speciation (p value = `r wilcox_test_result`).
-Our results therefore find no evidence for the ortholog conjecture in this system.
+Our phylogenetic comparative analysis, unlike the previously published pairwise comparative analysis [@KryuchkovaMostacci:2016iw], therefore find no evidence for the ortholog conjecture in this system.
 
@@
 
-We next examined the possibility that ascertainment biases were differentially impacting the inference of evolution following duplication and speciation events.
-We focused on two possible sources of bias - node depth and branch length.
+We next examined the possibility that ascertainment biases were differentially impacting the inference of expression evolution following duplication and speciation events.
+Such a bias might obscure support for the ortholog conjecture.
+We focused on two possible sources of bias, node depth and branch length.
 We found no evidence that either affected our results (Supplementary Materials).
-We also examined the sensitivity of the results to the calibration times used for speciation events on the gene trees.
-This is important because it is expected that genes from separate species trace backward in time to a common ancestor at older times than the splitting times of the species to which they belong [@Takahata:1989uh; @Degnan:2009hr].
+We also examined the sensitivity of the results to the calibration times applied to speciation events on the gene trees.
+This is important because it is expected that genes from separate species have a common ancestor older than the time at which the species diverged from each other [@Takahata:1989uh; @Degnan:2009hr].
 There is also uncertainty associated with the timing of these speciation events.
-In replicate analyses we added random noise to the calibration times and found that it does not impact our inability to reject the null hypothesis (Supplementary Materials).
+We added random noise to the calibration times in replicate analyses, and all still failed to reject the null hypothesis (Supplementary Materials).
 
@@
 			geom_point( aes( x=age, y=rho, col=Event, size=n ) ) +
-			theme_classic()
+			theme_classic() +
+			xlab( "Node Age (million years)" ) +
+			ylab( "Correlation Coefficient" )
 
@@
 > Figure XX_KMRR_Pairwise | This figures demonstrates that the pairwise analyses presented in Figure 2a of KMRR can be reproduced with the subset of data we consider here and our own independent code base.
-Each circle indicates the Tau Pearson correlation (rho) for a set of pairwise comparisons annotated with a specific node name of a given age and event type, *i.e.*, whether the divergence at that nodes was due to speciation (giving rise to orthologs) or duplication (giving rise to paralogs).
+Each circle indicates the Tau Pearson correlation coefficient for a set of pairwise comparisons annotated with a specific node name of a given age and event type, *i.e.*, whether the divergence at that nodes was due to speciation (giving rise to orthologs) or duplication (giving rise to paralogs).
 Following KMRR Figure 2a, this figure shows only the comparisons that include a human sequence.
@@
 In their Figure 1, they present a higher Tau correlation coefficient between ortholog pairs than paralog ortholog pairs.
-We find the same here, with correlation coefficients of R=`r round( ortholog_r, 3)` between ortholog pairs and `r round( paralog_r, 3)` for paralog pairs.
+We find the same here, with correlation coefficients of `r round( ortholog_r, 3)` for orthologs and `r round( paralog_r, 3)` for paralogs.
 
@@
 This is, in fact, the case for these data.
-While the mean distance between orthologs is `r round( ortholog_distance_mean, 1 )` My, the mean distance between paralogs is `r round( paralog_distance_mean, 1 )` My.
-As described above, this is because the oldest speciation event is by definition the most recent common ancestor of the species included in the study, but many gene families underwent duplication before this time.
-
-To test the hypothesis that distantly related paralogs arising from these older duplication events (Supplementary Figure XXAscertainment a) are the reason for the lower correlation of Tau between paralogs than between orthologs, we removed them.
-When we consider only the paralogs that arise from duplication events the same age or younger than the oldest speciation event (Supplementary Figure XXAscertainment b), the observed pattern is very different.
-The correlation of Tau between these shallower paralog pairs, R=`r round( paralog_subset_r, 3)`, is much stronger than for the correlation between all paralog pairs, R=`r round( paralog_r, 3)`.
-This is much closer to the ortholog correlation, which has R=`r round( ortholog_r, 3) `.
+While the mean distance between orthologs is `r round( ortholog_distance_mean, 1 )` million years, the mean distance between paralogs is `r round( paralog_distance_mean, 1 )` million years.
+This is because the oldest speciation event is by definition the most recent common ancestor of the species included in the study, but many gene families underwent duplication before this time.
+
+To test the hypothesis that distantly paralogs that arose from duplications that precede the oldest speciation event (Supplementary Figure XXAscertainment a) impact the lower correlation of Tau between paralogs than between orthologs, we removed them.
+When we consider only the paralogs that arise from duplication events the same age or younger than the oldest speciation event (Supplementary Figure XXAscertainment b), the paralog correlation coefficient increases from `r round( paralog_r, 3)` to  `r round( paralog_subset_r, 3)`.
+This is much closer to the ortholog correlation of `r round( ortholog_r, 3) `.
+
+```{r estimate_tau_model, cache=TRUE, echo=FALSE, fig.cap="XX_PIC_VAR", warning=FALSE, message=FALSE}
+	# Estimate model parameters and add them to the tree objects
+	# Subcalls to fitContinuous() are parallel, so don't use mclapply here
+	# This takes a while, so do it once here and reuse the values as needed
+	gene_trees_calibrated = lapply( gene_trees_calibrated, add_model_parameters )
+```
 
@@
 
-	# Replace the actual Tau values with simulated Tau values, using parameters 
-	# fitted from each tree
-	gene_trees_sim_null = lapply( gene_trees_calibrated, sim_tau )
+	# Replace the actual Tau values with simulated Tau values, using the parameters 
+	# estimated for each tree
+	gene_trees_sim_null = mclapply( gene_trees_calibrated, sim_tau, mc.cores=cores )
 
@@
 
-	# Replace the actual Tau values with simulated Tau values, using parameters 
-	# fitted from each tree. 
+	# Replace the actual Tau values with simulated Tau values
 	gene_trees_sim_oc = lapply( gene_trees_calibrated, sim_tau, dup_adjust=dup_adjust )
@@
 			theme_classic() +
-			xlab( "Age" ) +
-			ylab( "Rho" ) +
+			xlab( "Node Age (million years)" ) +
+			ylab( "Correlation Coefficient" ) +
 			ylim( 0, 1 ) +
@@
 
-We simulated a data set under the null model that tau evolves at the same rate following duplication and speciation events.
-Under the null model, pairwise analysis results (Figure XXSim a) are very similar as for the observed data, with higher rho for orthologs (`r ortholog_r_sim_null`) than paralogs (`r paralog_r_sim_null`).
-In the phylogenetic analysis of the null simulation, contrasts show similar divergence following speciation and duplication (Figure XXSim b) and does not reject the null hypothesis (Wilcoxon p=`r wilcox_oc( nodes_sim_null_contrast )`).
-
-We next simulated a data set under the ortholog conjecture, where the rate of evolution of Tau following duplication was `r dup_adjust` fold the rate following speciation.
-The results of this heterogeneous model are nearly indistinguishable from the null model  (Figure XXSim c), and also have a higher rho for orthologs (`r ortholog_r_sim_oc`) than paralogs (`r paralog_r_sim_oc`).
-In the phylogenetic analysis of the ortholog conjecture simulation, contrasts XX (Figure XXSim c) and XX (Wilcoxon p=`r wilcox_oc( nodes_sim_oc_contrast )`).
+The KMRR study did investigate the impact of node age on correlation, but in a different way.
+In their Figure 2, they grouped orthologs and paralogs according to the ENSEMBL node name of their most recent common ancestor, and plotted the correlation of Tau for each of these groups by the node age.
+They found that across the investigated range of node ages, ortholog pairs have higher tau correlation than paralogs.
+We confirmed that we can replicate this result (Figure XX_KMRR_Pairwise).
+There are several difficulty with interpretting this plot, though.
+First, it does not just reflect the evolutionary porcesses that generated the data, it is also impacted by the phylogenies along which these processes acted.
+The expected covariance of traits that evolve under neutral processes is in fact defined by the phylogeny [@Grafen:1989um].
+Second, the correlation for each group is based on multiple non-independent pairwise comparisons.
+
+To better understand this plot (Figure XX_KMRR_Pairwise), we performed a couple simulations of Tau on the calibrated gene trees and then regenerated the figures.
+First, we simulated the evolution of tau under the null model that it evolves at the same rate following duplication and speciation events.
+Under the null model, the plot of mean correlation to node age is very similar as for the observed data (Figure XXSim a).
+As in the original study, there is higher correlation coefficient across orthologs (`r ortholog_r_sim_null`) than paralogs (`r paralog_r_sim_null`) when not considering node age.
+Phylogenetic analysis of the data simulated under the null hypothesis (Figure XXSim b) do not reject the null hypothesis (Wilcoxon p=`r wilcox_oc( nodes_sim_null_contrast )`), as expected.
+
+We next simulated the evolution of Tau under the ortholog conjecture, where the rate of evolution of Tau following duplication was `r dup_adjust` fold the rate following speciation.
+The pairwise results of this heterogeneous model are nearly indistinguishable from the null model (Figure XXSim c), and also have a higher correlation coefficient for orthologs (`r ortholog_r_sim_oc`) than paralogs (`r paralog_r_sim_oc`).
+The phylogenetic analysis of the ortholog conjecture simulation (Figure XXSim c) does reject the null hypothesis (Wilcoxon p=`r wilcox_oc( nodes_sim_oc_contrast )`).
+
+These simulations have several implcations.
+The pairwise comparisons used by KMRR cannot distinguish betweeh the null hypothesis and ortholog conjecture, as the expected results are the same under both hypotheses.
+They also serve as a validation of phylogenetic methods as applied to this problem.
+Our analysis of independent contrasts does not reject the null hypothesis when data are simulated under the null model, and does reject the null hypothesis when the data are simulated under the ortholog conjecture.
+In contrast to pairwise methods, the phylogenetic analyses make testable predictions based on explicit hyptheses about evolutionary process.
 
@@
 
-We find that having information on whether two genes are orthologs or paralogs provides little added information about expression beyond knowing how distantly related the two genes are.
-This has several implications.
-It is an example of the limitations of the concepts of orthology and paralogy [@Dunn:2016er].
+There has been considerable recent interest in, and controversy about, the ortholog conjecture [@Koonin:2005fp; @Studer:2009bm; @Nehrt:2011bm; @Gabaldon:2013fi; @Altenhoff:2012ea].
+While some studies have presented support for the ortholog conjecture, our results are consistent with multiple studies that have not [@Studer:2009bm; @Nehrt:2011bm; @Yanai:2004hk].
+
+Our results suggest, at a minimum, that the ortholog conjecture is not a dominant pattern that is central to explaining the evolution of phenotypic diversity in gene families.
+The ortholog conjecture does not have to be an all or nothing question, though.
+It may be the case that the rates of phenotypic evolution following duplication may be greater than that following duplication in some organisms, gene families, and evolutionary processes [@Gabaldon:2013fi].
+We just don't find evidence for it when summarizing gene expression across tissues with Tau in these organisms.
+This calls into question the general predictive power of the ortholog conjecture, and until these processes are better understood it will be necessary to test for it in each situation.
+These tests should be articulated in terms of clear alternative hypotheses [@Studer:2009bm] that make distinct phylogenetic comparative predictions.
+
+Lack of support for the ortholog conjecture has important biological implications.
+It indicates that the mechanism of gene divergence (speciation versus duplication) may not have as strong an impact on phenotypic divergence as sometimes proposed.
+It also has technical implications.
+Having information on whether two genes are orthologs or paralogs provides little added information about expression beyond knowing how distantly related the two genes are.
+Rather than focus on whether genes are orthologs or paralogs when attempting to predict function, it may be more effective to simply focus on how closely related or distantly related they are.
+Closely related paralogs, for example, may tend to have more similar phenotypes than more distantly related orthologs [@Studer:2009bm].
+
+This is also an example of the limitations of the concepts of orthology and paralogy [@Dunn:2016er].
 These terms can have straightforward meaning in small gene trees with simple duplication/speciation histories, but the utility of the terms breaks down on larger more complex gene trees.
-Orthology and paralogy are annotations on the tips of the tree that are derived from the inferred structure of the tree and history of duplication and speciation at internal tree nodes.
+Orthology and paralogy are annotations on the tips of the phylogeny that are derived from the structure of the tree and history of duplication and speciation at internal tree nodes.
 In this sense, orthology and paralogy are statements about the internals of the tree that are distilled into statements at the tips of the tree.
 Much is lost in the process, though.
-These are the same pre-computed trees we use in our phylogenetic re-analyses of their study.
 For most questions it is much more direct to focus on the structure of the tree and the inferred processes within the tree, such as which internal nodes are duplication or speciation events and how much change occurs along the branches.
-In this case, KMRR [@KryuchkovaMostacci:2016iw] relied on orthology and paralogy annotations that are derived from the inferred duplication and speciation history on gene phylogenies [@Herrero:2016jj].
-
-There has been considerable recent interest in, and controversy about, the ortholog conjecture [@Koonin:2005fp; @Studer:2009bm; @Nehrt:2011bm; @Gabaldon:2013fi].
-Our results are consistent with multiple studies that have not found support for the ortholog conjecture [@Studer:2009bm; @Nehrt:2011bm; @Yanai:2004hk].
-Another study found only slightly greater similarity between orthologs than paralogs [@Altenhoff:2012ea].
-
-Our reanalysis of KMRR using phylogenetic methods suggests that the ortholog conjecture is not a dominant pattern that is central to explaining the evolution of phenotypic diversity in gene families.
-This has important biological implications.
-It suggests that the mechanism of gene divergence (speciation versus duplication) may not have as strong an impact on phenotypic divergence as sometimes proposed.
-It also has technical implications.
-Rather than focus on whether genes are orthologs or paralogs when attempting to predict function, it may be more effective to simply focus on how closely related or distantly related they are.
-Closely related paralogs, for example, may tend to have more similar phenotypes than more distantly related orthologs [@Studer:2009bm].
-
-The ortholog conjecture does not have to be an all or nothing question.
-It may be the case that the rates of phenotypic evolution following duplication may be greater than that following duplication in some organisms, gene families, and evolutionary processes [@Gabaldon:2013fi].
-We just don't find evidence for it when summarizing gene expression across tissues with Tau in these mammals.
-This calls into question the general predictive power of the ortholog conjecture, and until these processes are better understood it will be necessary to test for it in each situation.
-These tests should be articulated in terms of clear alternative hypotheses [@Studer:2009bm] that make distinct phylogenetic comparative predictions.
 
@@
 
-How changes in animal development relate to the evolution of animal diversity is a major question in evolutionary developmental biology (EvoDevo).
-To address this topic, Levin *et al.* [@Levin:2016bp] analyzed gene expression through the course of embryonic development for ten animal species, each from a different phylum.
+Levin *et al.* [@Levin:2016bp] analyzed gene expression through the course of embryonic development for ten animal species, each from a different clade that has been designated as having the rank of phylum.
 They arrived at two major conclusions.
@@
 
-Levin *et al.* [@Levin:2016bp] arrived at this conclusion by making multiple pairwise comparisons of ortholog expression through development between each pair of species.
+Levin *et al.* [@Levin:2016bp] arrived at this conclusion by making multiple pairwise comparisons of ortholog expression data sampled throughout the course of embryonic development.
 For each species pair, they identified the orthologs shared by these species.
 This list of shared genes was different from species pair to species pair.
-They characterized each gene in each species as having expression that peaks in early, mid, or late temporal phase of development.
+They characterized each of these orthologs in each species as having expression that peaks in early, mid, or late temporal phase of development.
 They then calculated a similarity score for each temporal phase for each species pair based on the fraction of genes that exhibited the same patterns in each species.
@@
 We found several problems resulting from the use of multiple pairwise comparisons.
-First, we found that every data point was included twice because both reciprocal pairwise comparisons (which have the same values) were retained.
+The first problems are specific to this particular implementation of pairwise comparisons.
+We found that every data point was included twice because both reciprocal pairwise comparisons (which have the same values) were retained.
 For example, there is both a nematode to arthropod comparison and an arthropod to nematode comparison.
@@
 
-The next issue we found was that all five of the lowest values in the mid phase distribution (Figure XXLevin a) are for pairwise comparisons that include the ctenophore (comb jelly).
-When the nine pairwise comparisons that include the ctenophore are removed and the general statistical issues noted above are addressed, there is no significant difference between the early phase and mid phase distributions (P = 0.1428 for the early-mid comparison and P < 10<sup>-5</sup> for the late-mid comparison).
-The inverse hourglass turns into a bottle (Figure XXLevin b).
+Once we addressed the issues above with the implementation of pairwise comparisons in Levin *et al.* [@Levin:2016bp], we were able to explore more general issues that can be a problem when making multiple pairwise comparisons between the same species.
+We found that all five of the lowest values in the mid phase distribution (Figure XXLevin a) are for pairwise comparisons that include the ctenophore (comb jelly).
+When the nine pairwise comparisons that include the ctenophore are removed, there is no significant difference between the early phase and mid phase distributions (P = 0.1428 for the early-mid comparison and P < 10<sup>-5</sup> for the late-mid comparison) and no support for the inverse hourglass (Figure XXLevin b).
 This highlights a well understood property of pairwise comparisons across species [@Felsenstein:1985ua]: evolutionary changes along a given branch, like those along the ctenophore branch, impact each of the multiple pairwise comparisons that includes that branch.
@@
 This can give the impression of a general pattern across the tree that is instead specific to changes along one part of the tree.
-The number of comparisons impacted by each change depends on the structure of the phylogenetic tree, i.e.
-how the species are related to each other.
+The number of comparisons impacted by each change depends on the structure of the phylogenetic tree, *i.e.* how the species are related to each other.
 
 Phylogenetic comparative methods were developed specifically to address this problem [@Felsenstein:1985ua].
-While we do demonstrate this problem impacts the Levin et al.
-analysis, we did not perform a phylogenetic reanalysis of this study, as we did for the KMRR study.
-This is because the similarity metric computed in the pairwise comparisons of Levin et al.
-are based on different genes for different species pairs.
+While we do demonstrate this problem impacts the Levin *et al.* analysis, we did not perform a phylogenetic reanalysis of this study, as we did for the KMRR study.
+This is because the similarity metric computed in the pairwise comparisons of Levin *et al.* are based on different genes for different species pairs.
 This means that they do not generate a trait that can be evaluated across a phylogeny, as KMRR's Tau can be.
-A full phylogenetic reanalysis would be possible by using upstream analysis products to re-derive new expression summary statistics suitable for phylogenetic comparative methods.
+A full phylogenetic reanalysis would be possible using upstream analysis products to re-derive new expression summary statistics suitable for phylogenetic comparative methods.
 
@@
 
-Our results highlight the importance of explicitly accounting for phylogenetic relationships when studying character evolution, including developmental [@Telford:2003tsa] and functional genomic traits [@Hejnol:2015gr].
-This is particularly true for evolutionary analyses of quantitative gene expression [@Dunn:2013cpa].
-
-Some of the most widely used phylogenetic comparative methods [@Felsenstein:1985ua; @Grafen:1989um] are already directly applicable to CFG data.
-There are interesting new challenges at this interdisciplinary interface that will need to be addressed to fully realize the potential of phylogenetic comparative functional genomics studies.
+Our results highlight the importance of explicitly incorporating information about phylogenetic relationships when comparing functional genomic traits across species.
+Some of the most widely used phylogenetic comparative methods [@Felsenstein:1985ua; @Grafen:1989um] are already directly applicable to comparative functional genomics stidues.
+There are also interesting new challenges at this interdisciplinary interface that will need to be addressed to fully realize the potential of phylogenetic comparative functional genomics studies.
 One such challenge is that most phylogenetic comparative analyses of covariance between traits have been developed to address problems with many more species (e.g., dozens or more) relative to the number of traits being examined.
-In CFG analyses, there are often far fewer species because adding taxa is still expensive, but tens of thousands of traits.
+In comparative functional genomics analyses, there are often far fewer species because adding taxa is still expensive, but tens of thousands of traits.
 This creates statistical challenges as the resulting covariance matrices are singular and, if not treated appropriately, imply many false correlations that are artifacts of project design.
@@
 We are not the first to apply phylogenetic comparative methods to functional genomic data.
-While the vast majority of comparative functional genomics studies have used standard pairwise similarity methods, a small number of CFG studies have employed phylogenetic comparative approaches [@Oakley:2005ky; @Eng:2009gp; @Chang:2014hq].
+While the vast majority of comparative functional genomics studies have used standard pairwise similarity methods, a small number of comparative functional genomics studies have employed phylogenetic comparative approaches [@Oakley:2005ky; @Eng:2009gp; @Chang:2014hq].
 For instance, a phylogenetic ANOVA [@Rohlfs:2015bd] of the evolution of gene expression improves statistical power and drastically reduces the rate of false positives relative to pairwise approaches.
@@
 There may still be missing speciation events, due to extinction and incomplete taxon sampling, and missing duplication events, due to gene loss, but these omissions affect both methods.
+
+This is particularly true for evolutionary analyses of quantitative gene expression [@Dunn:2013cpa].
 

### 2017-02-07 manuscript.rmd
wor smithing, added tau v max expression plot

@@
 With the goal of providing an integrated understanding of genome and phenotype evolution, there is now considerable interest in comparing functional genomics data across species.
-Most such studies have relied on multiple pairwise comparisons between species, an approach that does not incoporate information about the evolutionary relationships among species.
+Most such studies have relied on multiple pairwise comparisons between species, an approach that does not incorporate information about the evolutionary relationships among species.
 These pairwise approaches have the potential to mislead conclusions about the evolution of genome function, and are a missed opportunity to learn about biology that can only be understood in an explicit phylogenetic context.
@@
 In each study, distinct patterns of pairwise similarity among species were interpreted as evidence of particular evolutionary processes, but instead reflect the structure of phylogenies.
-These reanalyses concretely demonstrate the inadequacy of pairwise comparisons for analyzing functional genomic data across species, and indicate that it will be critical to adopt phylogenetic comparative methods in future funcitonal genomics work.
+These reanalyses concretely demonstrate the inadequacy of pairwise comparisons for analyzing functional genomic data across species, and indicate that it will be critical to adopt phylogenetic comparative methods in future functional genomics work.
 Fortunately, phylogenetic comparative biology is also a rapidly advancing field with many methods that can be directly applied to functional genomic data.
@@
 This is because most of the observed variance is due to differences between conifers and flowering plants.
-Phylogenetic comparative methods capture this shift as change along a single branch deep in the tree, revealing that there is not a tendancy of correlated change between these traits across the phylogeny.
+Phylogenetic comparative methods capture this shift as change along a single branch deep in the tree, revealing that there is not a tendency of correlated change between these traits across the phylogeny.
 One reason that comparative functional genomics has not yet embraced phylogenetic approaches is that it has not yet been concretely demonstrated that pairwise and phylogenetic comparative methods can lead to different results when considering functional genomics data.
@@
 Our use of contrasts is a bit different-- we look for differences in evolutionary changes of one trait (differential expression) between two categories of nodes (speciation and duplication) on a tree.
+We present eight of the actual gene trees with Tau values and independent contrasts in Figure XX_EXAMPLE_TREES as an example of how these methods are applied.
 
@@
 We confirmed that we can replicate this result (Figure XX_KMRR_Pairwise).
-There are several difficulty with interpretting this plot, though.
-First, it does not just reflect the evolutionary porcesses that generated the data, it is also impacted by the phylogenies along which these processes acted.
+There are several difficulty with interpreting this plot, though.
+First, it does not just reflect the evolutionary processes that generated the data, it is also impacted by the phylogenies along which these processes acted.
 The expected covariance of traits that evolve under neutral processes is in fact defined by the phylogeny [@Grafen:1989um].
@@
 
-These simulations have several implcations.
-The pairwise comparisons used by KMRR cannot distinguish betweeh the null hypothesis and ortholog conjecture, as the expected results are the same under both hypotheses.
+These simulations have several implications.
+The pairwise comparisons used by KMRR cannot distinguish between the null hypothesis and ortholog conjecture even when the evolution of expression is `r dup_adjust` fold higher following duplication than speciation.
+The pairwise results are the same under both hypotheses.
 They also serve as a validation of phylogenetic methods as applied to this problem.
 Our analysis of independent contrasts does not reject the null hypothesis when data are simulated under the null model, and does reject the null hypothesis when the data are simulated under the ortholog conjecture.
-In contrast to pairwise methods, the phylogenetic analyses make testable predictions based on explicit hyptheses about evolutionary process.
+In contrast to pairwise methods, the phylogenetic analyses make testable predictions based on explicit hypotheses about evolutionary process.
+
+Their Figure 3 differs from their analyses in that it presents independent changes in expression between triplets of genes.
+Each triplet has two paralogs that arose from a duplication event and one unduplicated homolog.
+They find that there is a tendency for the paralog with the lowest expression to have the highest Tau.
+From this they conclude that as paralogs diverge in expression, there is a tendency for the one that evolves to have lower expression also evolve to become more tissue specific.
+There is, however, a negative relationship between Tau and maximum expression across all genes (Supplementary Figure XX_TAU_V_EXP).
+A simpler explanation for this plot is that it reflects the global tendency for Tau to decrease with maximum expression.
+This global pattern could be due to both biological factors and the technical details of how Tau is calculated.
 
@@
 This highlights a well understood property of pairwise comparisons across species [@Felsenstein:1985ua]: evolutionary changes along a given branch, like those along the ctenophore branch, impact each of the multiple pairwise comparisons that includes that branch.
-The pairwise comparisons are therefore not independent - different pairwise comparisons are impacted by canges along some of the same branches (Figure 1a).
+The pairwise comparisons are therefore not independent - different pairwise comparisons are impacted by changes along some of the same branches (Figure 1a).
 This can give the impression of a general pattern across the tree that is instead specific to changes along one part of the tree.
@@
 Our results highlight the importance of explicitly incorporating information about phylogenetic relationships when comparing functional genomic traits across species.
-Some of the most widely used phylogenetic comparative methods [@Felsenstein:1985ua; @Grafen:1989um] are already directly applicable to comparative functional genomics stidues.
+Some of the most widely used phylogenetic comparative methods [@Felsenstein:1985ua; @Grafen:1989um] are already directly applicable to comparative functional genomics studies.
 There are also interesting new challenges at this interdisciplinary interface that will need to be addressed to fully realize the potential of phylogenetic comparative functional genomics studies.
@@
 We outlined these challenges and potential solutions in the context of gene expression [@Dunn:2013cpa].
-In addition to methods development, there is a critical need for software tools that make it simpler to integrate existing functional genomic and phylogenetic comparative packages.
+In the same manuscript we also considered another issue of relevance here -- the read counts generated by RNA-seq expression studies cannot be directly compared across species.
+This is because there are various species-specific technical factors that can be mistaken for differences in expression across species.
+These can be canceled out within species before making comparisons across species.
 
@@
 
-This is particularly true for evolutionary analyses of quantitative gene expression [@Dunn:2013cpa].
-
 ## Conclusions
@@
 The fact that the first two comparative functional genomics studies we reanalyzed show serious problems with pairwise comparisons indicates that there likely to be similar problems in other studies that use these methods, and that future comparative studies will be compromised if they continue to use pairwise methods.
-Studies of evolutionary functional genomics should not be focused on the tips of the tree, they should explicitly delve into the tree itself.
+Studies of evolutionary functional genomics should not be focused on the tips of the tree using pairwise comparisons.
+They should explicitly delve into the tree with phylogenetic comparative methods.
+Here we show a concrete case where the predictions of a null hypotheses and an alternatives hypothesis do not differ when considering pairwise comparisons, but make distinct predictions with phylogenetic comparative methods that can be used to test them.
 
@@
 The fact that multiple pairwise comparisons of developmental gene expression across diverse species share a particular pattern is not evidence of a general process that explains the differences between all species in the analysis.
-It is also the expected pattern when there are unique changes specific to a single species, as these changes will impact all pairwise comparisons to that species and therefore be counted multiple times.
-To use patterns across living species to test hypotheses about evolutionary processes, it is also necessary to incorporate information about evolutionary relationships, *i.e.* phylogenies.
+It is also the expected pattern when there are unique differences specific to a single species, and the evolutionary changes responsible for these differences are sampled multiple times in pairwise comparisons that span the same phylogenetic branches along which they occurred.
+To use patterns across living species to test hypotheses about evolutionary processes it is also necessary to incorporate information about evolutionary relationships, *i.e.* phylogenies.
 There have been decades of work on building comparative phylogenetic methods that do exactly that, and they are just as relevant to comparing functional genomics traits across species as they are to comparing morphology or any of the other traits they are already routinely applied to.
@@
 
+#### Relationship between Tau and maximum expression
+
+```{r Tau_v_max, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
+
+	nodes_tips = nodes_all %>% filter( ! is.na( species ) )
+	max_expression = nodes_tips %>% select( contains("Averaged") ) %>% apply( 1, FUN=max, na.rm=TRUE )
+	nodes_tips$max_expression = max_expression
+
+	ggplot( nodes_tips, aes( x=max_expression, y=Tau  ) ) + 
+		geom_point(alpha=0.05) + 
+		geom_smooth(method="lm") +
+		xlab("Maximum Expression") +
+		theme_classic()
+
+```
+
+> Supplementary Figure XX_TAU_V_EXP | There is a negaative relationship between Tau and the maximum observed expression for each gene.
+
 #### Investigation of potential ascertainment bias
@@
 In `r noised_replicate_n` reanalyses, we drew a new date for each calibration from a normal distribution with the mean of the original date and a standard deviation `r sd_fraction` times the original date.
-If any daugter nodes became older than their parent, we repeated the sampling until the dates were congruent with the topology.
+If any daughter nodes became older than their parent, we repeated the sampling until the dates were congruent with the topology.
 The minimum Wilcoxon p in these reanalyses was `r min( as.numeric( unlist( p_noised ) ) )`, *i.e.* none of them reject the null hypothesis that the rate of evolution of Tau is greater following duplication events than speciation events.

### 2017-02-07 manuscript.rmd
typos

@@
 We find problems with both that call their conclusions into question.
-One study interpreted higher expression correlation between orthologs than paralogs as evidence of the ortholog conjecture, *i.e.* the hypothesis that gene function evolution is more rapid after duplication than speciation.
+One study interpreted higher expression correlation between orthologs than paralogs as evidence for the ortholog conjecture, *i.e.* the hypothesis that gene function evolution is more rapid after duplication than speciation.
 Instead, we find that this pattern is due to the structure of the gene phylogenies, not different rates of expression evolution, and is the expected pattern when rates are the same following duplication and speciation.
@@
 A rich theoretical and statistical methodology of phylogenetic comparative methods have been developed over the last three decades to address the challenges and opportunities of trait comparisons across species [@Felsenstein:1985ua; @Grafen:1989um; @Pagel:1999gc; Revell:2012jd; @FitzJohn:2012ey; @Uyeda:2014jl; @Garamszegi:2014ew].
-A central challenge is the dependence of observations across species due to the evolutionary history of species -- more closely related species share many traits that evolve once in a common ancestor.
+A central challenge is the dependence of observations across species due to the evolutionary history of species -- more closely related species share many traits that evolved once in a common ancestor.
 This violates the fundamental assumption of observation independence in standard statistical methods.
@@
 
-> Supplementary Figure XX_TAU_V_EXP | There is a negaative relationship between Tau and the maximum observed expression for each gene.
+> Supplementary Figure XX_TAU_V_EXP | There is a negative relationship between Tau and the maximum observed expression for each gene.
 

### 2017-02-07 manuscript.rmd
added pdf

--- 2017-02-07 manuscript.rmd
@@
 output:
-  github_document: default
   pdf_document:
     dev: png
+    fig_caption: yes
+    fig_width: 6
+    fig_height: 4
+  github_document:
+    fig_caption: yes
+header-includes:
+  - \setlength{\parindent}{4em}
 ---
@@
 They have largely been applied to morphological and ecological traits, but are just as relevant to functional genomics.
-Even so, most comparative functional genomics studies have abstained from phylogenetic approaches and instead rely on multiple pairwise comparisons (Fig.
-1A).
+Even so, most comparative functional genomics studies have abstained from phylogenetic approaches and instead rely on multiple pairwise comparisons (Figure 1a).
 This leaves comparative functional genomics studies susceptible to statistical problems and is a missed opportunity to ask questions that are only accessible in an explicit phylogenetic context.
 
-Phylogenetic comparative methods account for evolutionary history and explicitly model trait change along the branches of evolutionary trees.
+Phylogenetic comparative methods account for evolutionary history and explicitly model trait change along the branches of evolutionary trees (*e.g.*, Figure 1b).
 The value of these methods relative to to pairwise comparisons has been repeatedly shown in analyses of other types of character data [@Ricklefs:1996wb; @Chamberlain:2012vx; @OMeara:2012gq].
@@
 The first study, Kryuchkova-Mostacci and Robinson-Rechavi (KMRR) 2016 [@KryuchkovaMostacci:2016iw], analyzed multiple vertebrate expression datasets to test the ortholog conjecture - the hypothesis that orthologs tend to have more conserved attributes (specificity of expression across organs in this case) than do paralogs [@Nehrt:2011bm, @Koonin:2005fp].
-Using pairwise comparisons (Fig.
-1A), they found lower expression correlation between paralogs than between orthologs and interpreted this as strong support for the ortholog conjecture.
+Using pairwise comparisons (Figure 1a), they found lower expression correlation between paralogs than between orthologs and interpreted this as strong support for the ortholog conjecture.
 
@@
 This study analyzed gene expression through the course of embryonic development for ten animal species, each from a different phylum.
-Using pairwise comparisons (Figure 1A), they found there is more evolutionary variance in gene expression at a mid phase of development than there is at early and late phases (Fig.
-XX).
+Using pairwise comparisons, they found there is more evolutionary variance in gene expression at a mid phase of development than there is at early and late phases (Figure 1a).
 They suggest that this supports an "inverse hourglass" model for the evolution of gene expression, in contrast with the "hourglass" model previously proposed for closely related species [@Kalinka:2010dw], and provides biological justification for the concept of phyla.
@@
 
-![Figure Overview](Figure_overview.png?raw=true)
-
-> Figure 1 | Pairwise and phylogenetic comparative approaches.
+![Pairwise and phylogenetic comparative approaches.
 (a) Many comparative functional genomic studies rely on pairwise comparisons, where traits of each gene are compared to traits of other genes (such as orthologs and paralogs).
@@
 (b) Comparative phylogenetic methods, including phylogenetic independent contrasts [@Felsenstein:1985ua], make a smaller number of independent comparisons, where each contrast measures independent changes along different branches.
-Phylogenetic approaches are rarely used for functional genomics studies.
+Phylogenetic approaches are rarely used for functional genomics studies.](Figure_overview.png?raw=true)
 
@@
 
-The KMRR analyses are based on pairwise comparisons (Figure 1A) between Tau within each gene family.
+The KMRR analyses are based on pairwise comparisons (Figure 1a) between Tau within each gene family.
 They found the correlation coefficient of Tau for orthologs to be significantly greater than the correlation coefficient of Tau for paralogs, *i.e.* that orthologs tend to be more similar to each other than paralogs.
@@
 
-```{r expectation_plots, echo=FALSE, cache=TRUE, message=F, warning=F}
+```{r expectation_plots, echo=FALSE, cache=TRUE, message=F, warning=F, fig.cap="(a) Under the null hypothesis that there is no difference in the rate of evolution after duplication or speciation events, the distribution of phylogenetic independent contrasts would be drawn from the same distribution across both types of nodes.
+(b) Under the ortholog conjecture, contrasts across duplication nodes would tend to be larger than contrasts across speciation nodes."}
 
@@
 
-> Figure XXExpectations | (a) Under the null hypothesis that there is no difference in the rate of evolution after duplication or speciation events, the distribution of PICs would be the same across both types of nodes.
-(b) Under the ortholog conjecture, contrasts across duplication nodes would tend to be larger than contrasts across speciation nodes.
-
-For each internal node in each gene tree, we measured the phylogenetic independent contrast (PIC) [@Felsenstein:1985ua] of the expression statistic Tau.
+For each internal node in each gene tree, we measured the phylogenetic independent contrast (PIC) of the expression statistic Tau.
 This is the difference in values of Tau for descendant nodes scaled by the expected variance, which is largely determined by the lengths of the two branches that connect the node to its two descendants (Figure 1b).
@@
 Our use of contrasts is a bit different-- we look for differences in evolutionary changes of one trait (differential expression) between two categories of nodes (speciation and duplication) on a tree.
-We present eight of the actual gene trees with Tau values and independent contrasts in Figure XX_EXAMPLE_TREES as an example of how these methods are applied.
 
@@
 
-```{r show_contrast_result, echo=FALSE, cache=TRUE, message=F, warning=F}
+```{r show_contrast_result, echo=FALSE, cache=TRUE, message=F, warning=F, fig.cap="Density plot of the magnitude of phylogenetic independent contrast values following duplication and speciation events.
+As the contrasts for duplication events are not larger than those of speciation events, the results are not consistent with the predictions of the ortholog conjecture."}
 
@@
 
-> Figure XXKMRR | Density plot of the magnitude of phylogenetic independent contrast values following duplication and speciation events.
-These contrasts are not larger following duplication events, as predicted by the ortholog conjecture.
-
 We mapped the Tau values calculated by KMRR [@KryuchkovaMostacci:2016iw] for the Brawand *et al.* 2011 [@Brawand:2011du] datset onto `r length(gene_trees)` gene trees parsed from ENSMBL Compara [@Herrero:2016jj].
@@
 These calibrated trees were used to calculate phylogenetic independent contrasts for  `r nodes_contrast %>% filter(D=="Y") %>% nrow()` duplication nodes and `r nodes_contrast %>% filter(D=="N") %>% nrow()` speciation nodes.
+One of these trees is shown in Supplementary Figure 7 to demonstrate the analysis.
 
@@
 A suitable null hypothesis in this case  is that there is no difference in the evolution of expression following speciation or duplication events [@Studer:2009bm].
-Under this hypothesis, we would predict that contrasts across speciation nodes and duplication nodes are drawn from the same distribution (Figure XXEXPECTATIONS a).
-Under the alternative hypothesis specified by the ortholog conjecture, that there is a higher rate of change following duplication events than speciation events, we would expect to see the distribution of duplication contrasts shifted to higher values relative to the speciation contrasts (Figure XXEXPECTATIONS b).
-
-We did not find increased evolutionary change in expression following duplication events (Fig.
-XXKMRR).
+Under this hypothesis, we would predict that contrasts across speciation nodes and duplication nodes are drawn from the same distribution (Figure 2a).
+Under the alternative hypothesis specified by the ortholog conjecture, that there is a higher rate of change following duplication events than speciation events, we would expect to see the distribution of duplication contrasts shifted to higher values relative to the speciation contrasts (Figure 2b).
+
+We did not find increased evolutionary change in expression following duplication events (Figure 3).
 The Wilcoxon rank test does not reject the null hypothesis that the rate of evolution following duplications is the same as or less than the rate following speciation (p value = `r wilcox_test_result`).
@@
 We focused on two possible sources of bias, node depth and branch length.
-We found no evidence that either affected our results (Supplementary Materials).
+We found no evidence that either affected our results (Supplementary Information, Supplementary Figure 8).
 We also examined the sensitivity of the results to the calibration times applied to speciation events on the gene trees.
@@
 There is also uncertainty associated with the timing of these speciation events.
-We added random noise to the calibration times in replicate analyses, and all still failed to reject the null hypothesis (Supplementary Materials).
+We added random noise to the calibration times in replicate analyses, and all still failed to reject the null hypothesis (Supplementary Information).
 
@@
 
-```{r pairwise, echo=FALSE, cache=TRUE, message=F, warning=F}
+```{r pairwise, echo=FALSE, cache=TRUE, message=F, warning=F, fig.cap="The pairwise analyses presented in Figure 2a of KMRR [@KryuchkovaMostacci:2016iw] can be reproduced with the subset of data we consider here.
+Each circle indicates the Tau Pearson correlation coefficient for a set of pairwise comparisons annotated with a specific node name of a given age and event type, *i.e.*, whether the divergence at that nodes was due to speciation (giving rise to orthologs) or duplication (giving rise to paralogs).
+Following KMRR Figure 2a, this figure shows only the comparisons that include a human sequence."}
 	get_pairwise_summary = function ( nhx ) {
@@
 		ggplot() + 
-			geom_point( aes( x=age, y=rho, col=Event, size=n ) ) +
+			geom_point( aes( x=age, y=rho, col=Event ), size=3 ) +
 			theme_classic() +
@@
 ```
-
-> Figure XX_KMRR_Pairwise | This figures demonstrates that the pairwise analyses presented in Figure 2a of KMRR can be reproduced with the subset of data we consider here and our own independent code base.
-Each circle indicates the Tau Pearson correlation coefficient for a set of pairwise comparisons annotated with a specific node name of a given age and event type, *i.e.*, whether the divergence at that nodes was due to speciation (giving rise to orthologs) or duplication (giving rise to paralogs).
-Following KMRR Figure 2a, this figure shows only the comparisons that include a human sequence.
 
@@
 Why is it that pairwise methods and phylogenetic methods lead to opposite conclusions?
-The reason is that they describe different things.
+One reason is that multiple pairwise comparisons repeatedly sample the same evolutionary changes and in so doing violate statistical assumptions of independence, whereas phylogenetic comparative methods make multiple independent comparisons across non-overlapping branches of the tree.
+The other reason for the different conclusions is that pairwise comparisons between the tips and phylogenetic comparative methods describe different things.
 Pairwise comparisons describe contemporary patterns, while phylogenetic methods infer historical processes [@OMeara:2012gq].
@@
 
-To test the hypothesis that distantly paralogs that arose from duplications that precede the oldest speciation event (Supplementary Figure XXAscertainment a) impact the lower correlation of Tau between paralogs than between orthologs, we removed them.
-When we consider only the paralogs that arise from duplication events the same age or younger than the oldest speciation event (Supplementary Figure XXAscertainment b), the paralog correlation coefficient increases from `r round( paralog_r, 3)` to  `r round( paralog_subset_r, 3)`.
+To test the hypothesis that distantly paralogs that arose from duplications that precede the oldest speciation event (Supplementary Figure 8a) impact the lower correlation of Tau between paralogs than between orthologs, we removed them.
+When we consider only the duplication events the same age or younger than the oldest speciation event (Supplementary Figure 8b), the paralog correlation coefficient increases from `r round( paralog_r, 3)` to  `r round( paralog_subset_r, 3)`.
 This is much closer to the ortholog correlation of `r round( ortholog_r, 3) `.
@@
 
-```{r simulations_plot, cache=TRUE, echo=FALSE, fig.cap="XX_PIC_VAR", warning=FALSE, message=FALSE}
+```{r simulations_plot, cache=TRUE, echo=FALSE, fig.cap="XX_PIC_VAR", warning=FALSE, message=FALSE, fig.cap="Analyses of simulated data reveal that pairwise methods produce similar results under both the null model and the ortholog conjecture, while the phylogenetic method distinguishes between the two.
+(a) Pairwise analysis of dataset simulated under the null model that there is no difference in the rate of expresison evolution following duplication and speciation.
+(b) Phylogenetic analysis of data simulated under the null model.
+(c) Pairwise analysis of dataset simulated under the ortholog conjecture, in which there is a higher rate of expresison evolution following duplication than speciation.
+This plot is very similar to the corresponding plot for the null model.
+(d) Phylogenetic analysis of data simulated under the ortholog conjecture.
+This plot is distinct from that for the null model in that phylogenetic independent contrasts for duplication events are significantly larger than after speciation events."}
 
@@
 
-> Figure XXSim
-
 The KMRR study did investigate the impact of node age on correlation, but in a different way.
@@
 They found that across the investigated range of node ages, ortholog pairs have higher tau correlation than paralogs.
-We confirmed that we can replicate this result (Figure XX_KMRR_Pairwise).
+We confirmed that we can replicate this result (Figure 4).
 There are several difficulty with interpreting this plot, though.
@@
 
-To better understand this plot (Figure XX_KMRR_Pairwise), we performed a couple simulations of Tau on the calibrated gene trees and then regenerated the figures.
+To better understand this plot (Figure 4), we performed a couple simulations of Tau on the calibrated gene trees and then regenerated the figures.
 First, we simulated the evolution of tau under the null model that it evolves at the same rate following duplication and speciation events.
-Under the null model, the plot of mean correlation to node age is very similar as for the observed data (Figure XXSim a).
+Under the null model, the plot of correlation coefficient to node age is very similar as for the observed data (Figure 5a).
 As in the original study, there is higher correlation coefficient across orthologs (`r ortholog_r_sim_null`) than paralogs (`r paralog_r_sim_null`) when not considering node age.
-Phylogenetic analysis of the data simulated under the null hypothesis (Figure XXSim b) do not reject the null hypothesis (Wilcoxon p=`r wilcox_oc( nodes_sim_null_contrast )`), as expected.
+Phylogenetic analysis of the data simulated under the null hypothesis (Figure 5b) do not reject the null hypothesis (Wilcoxon p = $`r wilcox_oc( nodes_sim_null_contrast )`$), as expected.
 
 We next simulated the evolution of Tau under the ortholog conjecture, where the rate of evolution of Tau following duplication was `r dup_adjust` fold the rate following speciation.
-The pairwise results of this heterogeneous model are nearly indistinguishable from the null model (Figure XXSim c), and also have a higher correlation coefficient for orthologs (`r ortholog_r_sim_oc`) than paralogs (`r paralog_r_sim_oc`).
-The phylogenetic analysis of the ortholog conjecture simulation (Figure XXSim c) does reject the null hypothesis (Wilcoxon p=`r wilcox_oc( nodes_sim_oc_contrast )`).
+The pairwise results of this heterogeneous model are nearly indistinguishable from the null model (Figure 5c), and also have a higher correlation coefficient for orthologs (`r ortholog_r_sim_oc`) than paralogs (`r paralog_r_sim_oc`).
+The phylogenetic analysis of the ortholog conjecture simulation (Figure 5d) does reject the null hypothesis (Wilcoxon p = $`r wilcox_oc( nodes_sim_oc_contrast )`$).
 
@@
 
-Their Figure 3 differs from their analyses in that it presents independent changes in expression between triplets of genes.
+Their Figure 3 differs from their other analyses in that it presents independent changes in expression between triplets of genes.
 Each triplet has two paralogs that arose from a duplication event and one unduplicated homolog.
 They find that there is a tendency for the paralog with the lowest expression to have the highest Tau.
-From this they conclude that as paralogs diverge in expression, there is a tendency for the one that evolves to have lower expression also evolve to become more tissue specific.
-There is, however, a negative relationship between Tau and maximum expression across all genes (Supplementary Figure XX_TAU_V_EXP).
+From this they conclude that following duplication there is a common trajectory of evolutionary change, where one paralog evolves to have lower expression and also become more tissue specific.
+There is, however, a negative relationship between Tau and maximum expression across all genes (Supplementary Figure 9).
 A simpler explanation for this plot is that it reflects the global tendency for Tau to decrease with maximum expression.
-This global pattern could be due to both biological factors and the technical details of how Tau is calculated.
+The null expectation for any two genes sampled at random is for one to have hhigher maximum expression and lower  This global pattern could be due to both biological factors and the technical details of how Tau is calculated.
 
@@
 
-![Figure Levin](levin_etal/Figure_levin.png?raw=true)
-
-> Figure XXLevin | Distributions of pairwise similarity scores for each phase of development.
+![Figure XXLevin | Distributions of pairwise similarity scores for each phase of development.
 Pairwise scores for the ctenophore are red.
@@
 (b) The distributions after the exclusion of the ctenophore.
-The early and mid phase distributions are not statistically distinct.
-This suggest a wine bottle model, with similar evolutionary variance at the early and mid phase and less at the late phase.
+The early and mid phase distributions are not statistically distinct.](levin_etal/Figure_levin.png?raw=true)
 
@@
 As a consequence, there are 90 entries for the 45 pairwise comparisons, and by doubling the data the significance of the result appears stronger than it actually is.
-After removing the duplicate values, the p values are far less significant, 0.002 for the early-mid comparison and on the order of 10<sup>-6</sup> for early-late.
+After removing the duplicate values, the p values are far less significant, 0.002 for the early-mid comparison and on the order of $10^-6$ for early-late.
 In addition, the test they used (KS test) is not appropriate for the hypothesis they seek to evaluate.
@@
 The Wilcoxon test is instead appropriate in this case.
-When applied to the de-duplicated data, the significance of this test is 0.02 for the early-mid comparison and on the order of 10<sup>-7</sup> for early-late.
+When applied to the de-duplicated data, the significance of this test is 0.02 for the early-mid comparison and on the order of $10^-7$ for early-late.
 
 Once we addressed the issues above with the implementation of pairwise comparisons in Levin *et al.* [@Levin:2016bp], we were able to explore more general issues that can be a problem when making multiple pairwise comparisons between the same species.
-We found that all five of the lowest values in the mid phase distribution (Figure XXLevin a) are for pairwise comparisons that include the ctenophore (comb jelly).
-When the nine pairwise comparisons that include the ctenophore are removed, there is no significant difference between the early phase and mid phase distributions (P = 0.1428 for the early-mid comparison and P < 10<sup>-5</sup> for the late-mid comparison) and no support for the inverse hourglass (Figure XXLevin b).
+We found that all five of the lowest values in the mid phase distribution (Figure 6a) are for pairwise comparisons that include the ctenophore (comb jelly).
+When the nine pairwise comparisons that include the ctenophore are removed, there is no significant difference between the early phase and mid phase distributions (P = 0.1428 for the early-mid comparison and P < 10<sup>-5</sup> for the late-mid comparison) and no support for the inverse hourglass (Figure 6b).
 This highlights a well understood property of pairwise comparisons across species [@Felsenstein:1985ua]: evolutionary changes along a given branch, like those along the ctenophore branch, impact each of the multiple pairwise comparisons that includes that branch.
@@
 Two paralogs that diverged from a duplication event deep in the tree may have many subsequent duplication and speciation events, and changes along all these branches will impact the final pairwise comparison.
-Phylogenetic methods have the advantage of isolating the changes under consideration (Fig.
-1b).
+Phylogenetic methods have the advantage of isolating the changes under consideration (Figure 1b).
 The phylogenetic methods avoid diluting the change that occurs along the branches that follow the node in question with changes along all subsequent branches.
@@
 We used the `chronos()` function from the R package ape [@Paradis:2004dv] for this calibration, with the `correlated` model.
-See Supplementary Materials for additional sensitivity analyses to time calibration.
+See Supplementary Information for additional sensitivity analyses to time calibration.
 Some trees could not be calibrated with these hard node constraints, and were discarded.
@@
 
-## Supplementary Material
+\pagebreak
+
+## Supplementary Information
 
@@
   clades_table,
-  caption = "Supplementary Table 1. Number of speciation nodes with contrasts for each clade on the tree."
+  caption = "Number of speciation nodes N with contrasts by clade name."
 )
@@
 		arrange( Species ) %>%
-		#mutate( Species = str_to_title( Species ) ) %>%
-		mutate( Species = str_replace( Species, "_", " ") )
+		mutate( Species = str_replace( Species, "_", "Q") ) %>%
+		mutate( Species = str_to_title( Species ) ) %>%
+		mutate( Species = str_replace( Species, "q", " ") )
 
@@
 	  species_table,
-	  caption = "Supplementary Table 2. Number of tips for each species in trees that were used to calculate the independent contrasts."
-	)
-
-```
-
-```{r example_trees, cache=TRUE, echo=FALSE, fig.cap="XX_PIC_VAR", warning=FALSE, message=FALSE}
+	  caption = "Number of tips for each species in trees that were used to calculate the independent contrasts."
+	)
+
+```
+
+```{r example_trees, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, eval=TRUE, fig.cap="One of the gene trees from the phylogenetic analysis.
+The value of Tau (expression specificity) is indicated by the sizes of the circles at the tips of the tree.
+Whether an internal node is a speciation or duplication is indicated by color.
+Speciation nodes are labeled by clade name.
+Branch lengths are scaled to time.
+The blue number is the independent contrast for each node."}
 
@@
 
-	print( example_trees )
-```
-
-> Supplementary Figure XX_EXAMPLE_TREES | A set of example trees to demonstrate the data and analyses that underlie the results.
-The value of Tau (expression specificity) is indicated by the sizes of the circles at the tips of the tree.
-Whether an internal node is a speciation or duplication is indicated by color.
-Speciation nodes are labeled by clade name.
-Branch lengths are scaled to time.
-The blue number is the independent contrast for each node.
-
-#### Relationship between Tau and maximum expression
-
-```{r Tau_v_max, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
-
-	nodes_tips = nodes_all %>% filter( ! is.na( species ) )
-	max_expression = nodes_tips %>% select( contains("Averaged") ) %>% apply( 1, FUN=max, na.rm=TRUE )
-	nodes_tips$max_expression = max_expression
-
-	ggplot( nodes_tips, aes( x=max_expression, y=Tau  ) ) + 
-		geom_point(alpha=0.05) + 
-		geom_smooth(method="lm") +
-		xlab("Maximum Expression") +
-		theme_classic()
-
-```
-
-> Supplementary Figure XX_TAU_V_EXP | There is a negaative relationship between Tau and the maximum observed expression for each gene.
+	print( example_trees[[6]] )
+```
 
@@
 
-While the age of speciation nodes is constrained, duplication nodes can be much older and therefore have a wider range of ages (Supplementary Figure XXAscertainment a).
+While the age of speciation nodes is constrained, duplication nodes can be much older and therefore have a wider range of ages (Supplementary Figure 8a).
 This is because many gene duplication events are older than the most recent common ancestor of the species in the study.
@@
 To test for such an effect, we remove duplication nodes that are older than the oldest speciation node.
-Our results are unchanged and this reduced dataset does not reject the null hypothesis that the rate of evolution following duplications is the same as or less than the rate following speciation (Supplementary Figure XXAscertainment b).
+Our results are unchanged and this reduced dataset does not reject the null hypothesis that the rate of evolution following duplications is the same as or less than the rate following speciation (Supplementary Figure 8b).
 
@@
 This is because the same difference across a shorter total branch length indicates a greater rate of evolutionary change.
-The expected variance of contrasts for speciation nodes is constrained by the branch lengths on the species tree, but the expected variance of contrasts for duplications has a much wider range (Figure XXAscertainment c).
+The expected variance of contrasts for speciation nodes is constrained by the branch lengths on the species tree, but the expected variance of contrasts for duplications has a much wider range (Figure 8c).
 This could lead to biases if the lengths of branches that descend from duplication nodes tend to be overestimated.
 We therefore examined only the contrasts that fell within in the range of expected variance seen for speciation contrasts, excluding duplication contrasts that fall outside of this range.
-This reanalysis does not reject the null hypothesis either (Figure XXAscertainment d), indicating that branch length bias is not responsible for the result.
-
-```{r ascertainment_plot, cache=TRUE, echo=FALSE, fig.cap="XX_PIC_VAR", warning=FALSE, message=FALSE}
+This reanalysis does not reject the null hypothesis either (Figure 8d), indicating that branch length bias is not responsible for the result.
+
+```{r ascertainment_plot, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Investigation of possible ascertainment biases.
+(a) Magnitude of independent contrasts plotted against node age.
+Speciation nodes are calibrated to particular times, whereas duplication nodes have a wider range.
+(b) Density plot of contrasts for only the nodes that have an age less than or equal to the maximum age of speciation nodes.
+(c) Magnitude of independent contrasts plotted against expected variance, which is largely determined by branch lengths.
+Contrasts for speciation nodes have a narrower range of expected variance than do contrasts for duplication nodes.
+(d) Density plot of contrasts for only the nodes that have expected variance within the range of contrasts across speciation nodes."}
 
@@
 
-> Supplementary Figure XXAscertainment | Investigation of possible ascertainment biases.
-(a) Magnitude of independent contrasts plotted against node age.
-Speciation nodes are calibrated to particular times, whereas duplication nodes have a wider range.
-(b) Density plot of contrasts for only the nodes that have an age less than or equal to the maximum age of speciation nodes.
-(c) Magnitude of independent contrasts plotted against expected variance, which is largely determined by branch lengths.
-Contrasts for speciation nodes have a narrower range of expected variance than do contrasts for duplication nodes.
-(d) Density plot of contrasts for only the nodes that have expected variance within the range of contrasts across speciation nodes.
-
 #### Investigation of sensitivity to calibration times
@@
 If any daughter nodes became older than their parent, we repeated the sampling until the dates were congruent with the topology.
-The minimum Wilcoxon p in these reanalyses was `r min( as.numeric( unlist( p_noised ) ) )`, *i.e.* none of them reject the null hypothesis that the rate of evolution of Tau is greater following duplication events than speciation events.
+The minimum Wilcoxon p in these reanalyses was $`r min( as.numeric( unlist( p_noised ) ) )`$, *i.e.* none of them reject the null hypothesis that the rate of evolution of Tau is greater following duplication events than speciation events.
 This is consistent with the analysis that uses the calibration dates as specified, indicating that our results are robust to the selection of calibration times for speciation nodes.
+
+#### Relationship between Tau and maximum expression
+
+```{r Tau_v_max, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="There is a negaative relationship between Tau and the maximum observed expression for each gene."}
+
+	nodes_tips = nodes_all %>% filter( ! is.na( species ) )
+	max_expression = nodes_tips %>% select( contains("Averaged") ) %>% apply( 1, FUN=max, na.rm=TRUE )
+	nodes_tips$max_expression = max_expression
+
+	ggplot( nodes_tips, aes( x=max_expression, y=Tau  ) ) + 
+		geom_point(alpha=0.05) + 
+		geom_smooth(method="lm") +
+		xlab("Maximum Expression") +
+		theme_classic()
+
+```
 
@@
 
+\pagebreak
+
 ## References

### 2017-02-07 manuscript.rmd
minor pdf formatting

@@
 
-```{r expectation_plots, echo=FALSE, cache=TRUE, message=F, warning=F, fig.cap="(a) Under the null hypothesis that there is no difference in the rate of evolution after duplication or speciation events, the distribution of phylogenetic independent contrasts would be drawn from the same distribution across both types of nodes.
+```{r expectation_plots, echo=FALSE, cache=TRUE, message=F, warning=F, fig.height=2, fig.cap="(a) Under the null hypothesis that there is no difference in the rate of evolution after duplication or speciation events, the distribution of phylogenetic independent contrasts would be drawn from the same distribution across both types of nodes.
 (b) Under the ortholog conjecture, contrasts across duplication nodes would tend to be larger than contrasts across speciation nodes."}
@@
 
-```{r show_contrast_result, echo=FALSE, cache=TRUE, message=F, warning=F, fig.cap="Density plot of the magnitude of phylogenetic independent contrast values following duplication and speciation events.
+```{r show_contrast_result, echo=FALSE, cache=TRUE, message=F, warning=F, fig.height=2, fig.width=5, fig.cap="Density plot of the magnitude of phylogenetic independent contrast values following duplication and speciation events.
 As the contrasts for duplication events are not larger than those of speciation events, the results are not consistent with the predictions of the ortholog conjecture."}
@@
 
-```{r pairwise, echo=FALSE, cache=TRUE, message=F, warning=F, fig.cap="The pairwise analyses presented in Figure 2a of KMRR [@KryuchkovaMostacci:2016iw] can be reproduced with the subset of data we consider here.
+```{r pairwise, echo=FALSE, cache=TRUE, message=F, warning=F, fig.height=2, fig.width=5, fig.cap="The pairwise analyses presented in Figure 2a of KMRR [@KryuchkovaMostacci:2016iw] can be reproduced with the subset of data we consider here.
 Each circle indicates the Tau Pearson correlation coefficient for a set of pairwise comparisons annotated with a specific node name of a given age and event type, *i.e.*, whether the divergence at that nodes was due to speciation (giving rise to orthologs) or duplication (giving rise to paralogs).
@@
 
-![Figure XXLevin | Distributions of pairwise similarity scores for each phase of development.
+![Distributions of pairwise similarity scores for each phase of development.
 Pairwise scores for the ctenophore are red.
@@
 They then calculated a similarity score for each temporal phase for each species pair based on the fraction of genes that exhibited the same patterns in each species.
-The distributions of similarity scores are plotted in their [Figure 4d](http://www.nature.com/nature/journal/v531/n7596/fig_tab/nature16994_F4.html), and their Kolmogorov–Smirnov (KS) tests indicated that the early distribution and late distribution were each significantly different from mid distribution (P < 10<sup>-6</sup> and P < 10<sup>-12</sup>, respectively).
+The distributions of similarity scores are plotted in their [Figure 4d](http://www.nature.com/nature/journal/v531/n7596/fig_tab/nature16994_F4.html), and their Kolmogorov–Smirnov (KS) tests indicated that the early distribution and late distribution were each significantly different from mid distribution (P < $10^{-6}$ and P < $10^{-12}$, respectively).
 This is the support they presented for the inverse hourglass model.
@@
 As a consequence, there are 90 entries for the 45 pairwise comparisons, and by doubling the data the significance of the result appears stronger than it actually is.
-After removing the duplicate values, the p values are far less significant, 0.002 for the early-mid comparison and on the order of $10^-6$ for early-late.
+After removing the duplicate values, the p values are far less significant, 0.002 for the early-mid comparison and on the order of $10^{-6}$ for early-late.
 In addition, the test they used (KS test) is not appropriate for the hypothesis they seek to evaluate.
@@
 The Wilcoxon test is instead appropriate in this case.
-When applied to the de-duplicated data, the significance of this test is 0.02 for the early-mid comparison and on the order of $10^-7$ for early-late.
+When applied to the de-duplicated data, the significance of this test is 0.02 for the early-mid comparison and on the order of $10^{-7}$ for early-late.
 
@@
 We found that all five of the lowest values in the mid phase distribution (Figure 6a) are for pairwise comparisons that include the ctenophore (comb jelly).
-When the nine pairwise comparisons that include the ctenophore are removed, there is no significant difference between the early phase and mid phase distributions (P = 0.1428 for the early-mid comparison and P < 10<sup>-5</sup> for the late-mid comparison) and no support for the inverse hourglass (Figure 6b).
+When the nine pairwise comparisons that include the ctenophore are removed, there is no significant difference between the early phase and mid phase distributions (P = 0.1428 for the early-mid comparison and P < $10^{-5}$ for the late-mid comparison) and no support for the inverse hourglass (Figure 6b).
 This highlights a well understood property of pairwise comparisons across species [@Felsenstein:1985ua]: evolutionary changes along a given branch, like those along the ctenophore branch, impact each of the multiple pairwise comparisons that includes that branch.
@@
 
-```{r contrast_table, cache=TRUE, echo=FALSE, fig.cap="XX_PIC_VAR", warning=FALSE, message=FALSE}
+```{r contrast_table, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
 
@@
 
-```{r species_table, cache=TRUE, echo=FALSE, fig.cap="XX_PIC_VAR", warning=FALSE, message=FALSE}
+```{r species_table, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
 
@@
 
-```{r ascertainment_plot, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Investigation of possible ascertainment biases.
+```{r ascertainment_plot, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, dev="png", fig.cap="Investigation of possible ascertainment biases.
 (a) Magnitude of independent contrasts plotted against node age.
@@
 
-```{r Tau_v_max, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="There is a negaative relationship between Tau and the maximum observed expression for each gene."}
+```{r Tau_v_max, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, dev="png", fig.cap="There is a negaative relationship between Tau and the maximum observed expression for each gene."}
 

### 2017-02-07 manuscript.rmd
removed pdf indenting

@@
     fig_caption: yes
-header-includes:
-  - \setlength{\parindent}{4em}
+
 ---

### 2017-02-08 manuscript.rmd
figure formatting, typos

@@
   pdf_document:
-    dev: png
     fig_caption: yes
@@
     fig_height: 4
-  github_document:
-    fig_caption: yes
 
@@
 
-# Multiple pairwise comparisons between species strongly mislead functional genomics studies
+# Multiple pairwise comparisons between species can strongly mislead functional genomics studies
 
@@
 
-With the goal of providing an integrated understanding of genome and phenotype evolution, there is now considerable interest in comparing functional genomics data across species.
-Most such studies have relied on multiple pairwise comparisons between species, an approach that does not incorporate information about the evolutionary relationships among species.
+There is now considerable interest in comparing functional genomics data across species.
+One goal of this work is to provide an integrated understanding of genome and phenotype evolution.
+Most studies of this type have relied on multiple pairwise comparisons between species, an approach that does not incorporate information about the evolutionary relationships among species.
 These pairwise approaches have the potential to mislead conclusions about the evolution of genome function, and are a missed opportunity to learn about biology that can only be understood in an explicit phylogenetic context.
@@
 (b) Comparative phylogenetic methods, including phylogenetic independent contrasts [@Felsenstein:1985ua], make a smaller number of independent comparisons, where each contrast measures independent changes along different branches.
-Phylogenetic approaches are rarely used for functional genomics studies.](Figure_overview.png?raw=true)
+Phylogenetic approaches are rarely used for functional genomics studies.](Figure_overview.pdf)
 
@@
 			geom_freqpoly(  binwidth=0.2, position="identity" ) +
-			xlab( "Phylogenetic independent contrasts (PIC)" ) +
+			xlab( "Phylogenetic independent contrasts" ) +
 			ylab( "Density" ) +
@@
 			) +
-			ggtitle( "(b) Ortholog Conjecture (OC) expectation" )
+			ggtitle( "(b) Ortholog Conjecture \nexpectation" )
 
@@
 		geom_freqpoly( binwidth=.001, position="identity" ) +
-		xlab( "Phylogenetic independent \ncontrasts (PIC)" ) +
+		xlab( "Phylogenetic independent contrasts" ) +
 		ylab( "Density" ) +
 		xlim( 0, 0.075 ) + 
+		theme( legend.justification=c( 1, 1 ), legend.position=c( 1, 1 ) ) +
 		theme_classic()
@@
 
-```{r ascertainment, cache=TRUE, echo=FALSE, fig.cap="XX_PIC_VAR", warning=FALSE, message=FALSE}
+```{r ascertainment, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
 
@@
 			xlab( "Node age (millions of years)" ) + 
-			ylab( "Phylogenetic independent contrasts (PIC)\n of differential expression" ) +
+			ylab( "Phylogenetic independent \ncontrast" ) +
 			ylim( 0, 0.3 ) +
@@
 			ggtitle( "(a)" ) +
-			theme_classic()
+			theme_classic()	+
+			theme( legend.justification=c( 1, 1 ), legend.position=c( 1, 1 ) )
 
@@
 
-```{r estimate_tau_model, cache=TRUE, echo=FALSE, fig.cap="XX_PIC_VAR", warning=FALSE, message=FALSE}
+```{r estimate_tau_model, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
 	# Estimate model parameters and add them to the tree objects
@@
 
-```{r simulations_null, cache=TRUE, echo=FALSE, fig.cap="XX_PIC_VAR", warning=FALSE, message=FALSE}
+```{r simulations_null, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
 
@@
 
-```{r simulations_oc, cache=TRUE, echo=FALSE, fig.cap="XX_PIC_VAR", warning=FALSE, message=FALSE}
+```{r simulations_oc, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
 
@@
 
-```{r simulations_plot, cache=TRUE, echo=FALSE, fig.cap="XX_PIC_VAR", warning=FALSE, message=FALSE, fig.cap="Analyses of simulated data reveal that pairwise methods produce similar results under both the null model and the ortholog conjecture, while the phylogenetic method distinguishes between the two.
+```{r simulations_plot, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Analyses of simulated data reveal that pairwise methods produce similar results under both the null model and the ortholog conjecture, while the phylogenetic method distinguishes between the two.
 (a) Pairwise analysis of dataset simulated under the null model that there is no difference in the rate of expresison evolution following duplication and speciation.
@@
 We confirmed that we can replicate this result (Figure 4).
-There are several difficulty with interpreting this plot, though.
+There are several difficulties with interpreting this plot, though.
 First, it does not just reflect the evolutionary processes that generated the data, it is also impacted by the phylogenies along which these processes acted.
@@
 A simpler explanation for this plot is that it reflects the global tendency for Tau to decrease with maximum expression.
-The null expectation for any two genes sampled at random is for one to have hhigher maximum expression and lower  This global pattern could be due to both biological factors and the technical details of how Tau is calculated.
+The null expectation for any two genes sampled at random is for one to have higher maximum expression and lower Tau, and the other to have lower maximum expression and higher Tau.
+This global pattern could be due to both biological factors and the technical details of how Tau is calculated.
 
@@
 
-```{r ascertainment_plot, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, dev="png", fig.cap="Investigation of possible ascertainment biases.
+```{r ascertainment_plot, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, dev="png", dpi=600, fig.cap="Investigation of possible ascertainment biases.
 (a) Magnitude of independent contrasts plotted against node age.
@@
 		geom_freqpoly( binwidth=.001, position="identity" ) +
-		xlab( "Phylogenetic independent \ncontrasts (PIC)" ) +
+		xlab( "Phylogenetic independent contrast" ) +
 		ylab( "Density" ) +
@@
 		annotate( "text", x = Inf, y = Inf, hjust="right", vjust="top", label=label_age ) +
-		theme_classic()
+		theme_classic() +
+		theme(legend.position="none")
 
@@
 		xlab( "Expected variance, \n depends mostly on branch length" ) + 
-		ylab( "Phylogenetic independent contrasts (PIC)\n of differential expression" ) +
+		ylab( "Phylogenetic independent \ncontrast" ) +
 		ylim( 0, 0.3 ) +
@@
 		ggtitle( "(c)" ) +
-		theme_classic()
+		theme_classic()	+
+		theme(legend.position="none")
 
@@
 		annotate( "text", x = Inf, y = Inf, hjust="right", vjust="top", label=label_var ) +
-		xlab( "Phylogenetic independent contrasts (PIC)\n of differential expression" ) +
-		ylab( "Density" )
+		xlab( "Phylogenetic independent contrast" ) +
+		ylab( "Density" ) +
+		theme(legend.position="none")
 
@@
 
-```{r calibration_sensitivity, cache=TRUE, echo=FALSE, fig.cap="XX_PIC_VAR", warning=FALSE, message=FALSE}
+```{r calibration_sensitivity, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
 
@@
 
-```{r Tau_v_max, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, dev="png", fig.cap="There is a negaative relationship between Tau and the maximum observed expression for each gene."}
+```{r Tau_v_max, cache=TRUE, echo=FALSE, warning=FALSE, dev="png", dpi=300, message=FALSE, fig.cap="There is a negative relationship between Tau and the maximum observed expression for each gene."}
 
@@
 	ggplot( nodes_tips, aes( x=max_expression, y=Tau  ) ) + 
-		geom_point(alpha=0.05) + 
-		geom_smooth(method="lm") +
-		xlab("Maximum Expression") +
+		geom_point( size=0.5, alpha = 0.05 ) + 
+		geom_smooth( method="lm" ) +
+		xlab( "Maximum Expression" ) +
 		theme_classic()

### 2017-02-08 manuscript.rmd
pdf formatting, small changes to text throughout

@@
 
-# Multiple pairwise comparisons between species can strongly mislead functional genomics studies
+# Pairwise comparisons are problematic when analyzing functional genomic data across species
 
@@
 Most studies of this type have relied on multiple pairwise comparisons between species, an approach that does not incorporate information about the evolutionary relationships among species.
-These pairwise approaches have the potential to mislead conclusions about the evolution of genome function, and are a missed opportunity to learn about biology that can only be understood in an explicit phylogenetic context.
-Here we examine two recently published studies that compare expression across species with pairwise methods.
+The statistical problems that arise from not considering these relationships can lead pairwise approaches to the wrong conclusions, and are a missed opportunity to learn about biology that can only be understood in an explicit phylogenetic context.
+Here we examine two recently published studies that compare gene expression across species with pairwise methods.
 We find problems with both that call their conclusions into question.
@@
 The second study interpreted pairwise comparisons of embryonic gene expression across distantly related animals as evidence for a distinct evolutionary process that gave rise to animal phyla.
-Instead, we find that the pattern they identified is due to unique feature of a single species that impact multiple pairwise comparisons that include that species.
+Instead, we find that the pattern they identified is due to unique features of a single species that impact multiple pairwise comparisons that include that species.
 In each study, distinct patterns of pairwise similarity among species were interpreted as evidence of particular evolutionary processes, but instead reflect the structure of phylogenies.
@@
 ### KMRR Reanalysis
-
-#### Original pairwise test of the ortholog conjecture
-
-The KMRR study [@KryuchkovaMostacci:2016iw] sought to test the ortholog conjecture.
-The ortholog conjecture [@Nehrt:2011bm] is the proposition that orthologs (genes that diverged from each other due to a speciation event) have more similar attributes than do paralogs (genes that diverged from each other due to a gene duplication event).
-The ortholog conjecture has important biological and technical implications.
-It shapes our understanding of the functional diversity of gene families.
-It is used to relate findings from well-studied genes to related genes that have not been investigated in detail.
-It can be applied to any trait of genes, from gene sequence to biochemical properties to expression.
-While the ortholog conjecture describes a specific pattern of functional diversity across genes, it is also articulated as a hypothesis about the process of evolution-- that there is greater evolutionary change in gene attributes following a duplication event than a speciation event.
-
-Despite its importance, there have been relatively few tests of the ortholog conjecture.
-Previous work has shown that ontology annotations are not sufficient to test the ortholog conjecture [@Chen:2012jz; @Thomas:2012hz].
-Analyses of domain structure were consistent with the ortholog conjecture [@Forslund:2011cw].
-There have been few tests of the ortholog conjecture with regards to gene expression [@Chen:2012jz], and KMRR is the most thorough such expression study to date.
-
-The KMRR study considered several publicly available datasets of gene expression across tissues and species.
-Their expression summary statistic is Tau [@Yanai:2005dh, @KryuchkovaMostacci:2016in], an indicator of tissue specificity of gene expression.
-It can range from a value of 0, which indicates no specificity (*i.e.*, uniform expression across tissues), to a value of 1, which indicates high specificity (*i.e.*, expression in only one tissue).
-It is convenient in that it is a single number of defined range for each gene, though of course since the original expression is multidimensional this means much information is discarded.
-This includes information about which tissue expression is specific to.
-For example, if one gene has expression specific to the brain and another expression specific to the kidney, both would have a Tau of 1.
-
-The KMRR analyses are based on pairwise comparisons (Figure 1a) between Tau within each gene family.
-They found the correlation coefficient of Tau for orthologs to be significantly greater than the correlation coefficient of Tau for paralogs, *i.e.* that orthologs tend to be more similar to each other than paralogs.
-From this they concluded that their analyses support the ortholog conjecture.
-They also concluded that this pattern provides support for a particular evolutionary process, that "tissue-specificity evolves very slowly in the absence of duplication, while immediately after duplication the new gene copy differs" [@KryuchkovaMostacci:2016iw].
-
-#### Phylogenetic reanalyses
-
-We reanalyzed the KMRR study using phylogenetic comparative methods.
-We focused on one of the datasets included in their analyses, that of Brawand *et al.* 2011 [@Brawand:2011du].
-This dataset is the best sampled in their analyses.
-It has gene expression data for six organs across ten species (nine mammals and one bird), eight of which were analyzed by KMRR and further considered here.
 
@@
 
+#### Original pairwise test of the ortholog conjecture
+
+The KMRR study [@KryuchkovaMostacci:2016iw] sought to test the ortholog conjecture.
+The ortholog conjecture [@Nehrt:2011bm] is the proposition that orthologs (genes that diverged from each other due to a speciation event) have more similar attributes than do paralogs (genes that diverged from each other due to a gene duplication event).
+The ortholog conjecture has important biological and technical implications.
+It shapes our understanding of the functional diversity of gene families.
+It is used to relate findings from well-studied genes to related genes that have not been investigated in detail.
+It can be applied to any trait of genes, from gene sequence to biochemical properties to expression.
+While the ortholog conjecture describes a specific pattern of functional diversity across genes, it is also articulated as a hypothesis about the process of evolution-- that there is greater evolutionary change in gene attributes following a duplication event than a speciation event.
+
+Despite its importance, there have been relatively few tests of the ortholog conjecture.
+Previous work has shown that ontology annotations are not sufficient to test the ortholog conjecture [@Chen:2012jz; @Thomas:2012hz].
+Analyses of domain structure were consistent with the ortholog conjecture [@Forslund:2011cw].
+There have been few tests of the ortholog conjecture with regards to gene expression [@Chen:2012jz], and KMRR is the most thorough such expression study to date.
+
+The KMRR study considered several publicly available datasets of gene expression across tissues and species.
+Their expression summary statistic is Tau [@Yanai:2005dh, @KryuchkovaMostacci:2016in], an indicator of tissue specificity of gene expression.
+It can range from a value of 0, which indicates no specificity (*i.e.*, uniform expression across tissues), to a value of 1, which indicates high specificity (*i.e.*, expression in only one tissue).
+It is convenient in that it is a single number of defined range for each gene, though of course since the original expression is multidimensional this means much information is discarded.
+This includes information about which tissue expression is specific to.
+For example, if one gene has expression specific to the brain and another expression specific to the kidney, both would have a Tau of 1.
+
+The KMRR analyses are based on pairwise comparisons (Figure 1a) between Tau within each gene family.
+They found the correlation coefficient of Tau for orthologs to be significantly greater than the correlation coefficient of Tau for paralogs, *i.e.* that orthologs tend to be more similar to each other than paralogs.
+From this they concluded that their analyses support the ortholog conjecture.
+They also concluded that this pattern provides support for a particular evolutionary process, that "tissue-specificity evolves very slowly in the absence of duplication, while immediately after duplication the new gene copy differs" [@KryuchkovaMostacci:2016iw].
+
+#### Phylogenetic reanalyses
+
+We reanalyzed the KMRR study using phylogenetic comparative methods.
+We focused on one of the datasets included in their analyses, that of Brawand *et al.* 2011 [@Brawand:2011du].
+This dataset is the best sampled in their analyses.
+It has gene expression data for six organs across ten species (nine mammals and one bird), eight of which were analyzed by KMRR and further considered here.
+
 For each internal node in each gene tree, we measured the phylogenetic independent contrast (PIC) of the expression statistic Tau.
@@
 
-```{r show_contrast_result, echo=FALSE, cache=TRUE, message=F, warning=F, fig.height=2, fig.width=5, fig.cap="Density plot of the magnitude of phylogenetic independent contrast values following duplication and speciation events.
+```{r show_contrast_result, echo=FALSE, cache=TRUE, message=F, warning=F, fig.height=2, fig.width=4, fig.cap="Phylogneetic reanalysis of the Tau values calculated by KMRR [@KryuchkovaMostacci:2016iw].
+Density plot of the magnitude of phylogenetic independent contrasts for Tau following duplication and speciation events.
 As the contrasts for duplication events are not larger than those of speciation events, the results are not consistent with the predictions of the ortholog conjecture."}
@@
 		xlim( 0, 0.075 ) + 
-		theme( legend.justification=c( 1, 1 ), legend.position=c( 1, 1 ) ) +
-		theme_classic()
+		theme_classic() + 
+		theme( legend.justification=c( 1, 1 ), legend.position=c( 1, 1 ) )
 
@@
 
-```{r simulations_plot, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Analyses of simulated data reveal that pairwise methods produce similar results under both the null model and the ortholog conjecture, while the phylogenetic method distinguishes between the two.
+```{r simulations_plot, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Analyses of simulated data reveal that pairwise methods produce similar results under both the null model and the ortholog conjecture (OC), while the phylogenetic method provides different results under each model.
 (a) Pairwise analysis of dataset simulated under the null model that there is no difference in the rate of expresison evolution following duplication and speciation.
@@
 			theme_classic() + 
-			ggtitle( "(b) Null simulation, phylogenetic" ) +
+			ggtitle( "(b) Null, phylogenetic" ) +
 			theme( legend.justification=c( 1, 1 ), legend.position=c( 1, 1 ) )
@@
 			ylim( 0, 1 ) +
-			ggtitle( "(a) Null simulation, pairwise" ) +
+			ggtitle( "(a) Null, pairwise" ) +
 			theme(legend.position="none")
@@
 		theme_classic() + 
-		ggtitle( "(d) OC simulation, phylogenetic" ) +
+		ggtitle( "(d) OC, phylogenetic" ) +
 		theme(legend.position="none")
@@
 			ylim( 0, 1 ) +
-			ggtitle( "(c) OC simulation, pairwise" ) +
+			ggtitle( "(c) OC, pairwise" ) +
 			theme(legend.position="none")
@@
 To better understand this plot (Figure 4), we performed a couple simulations of Tau on the calibrated gene trees and then regenerated the figures.
-First, we simulated the evolution of tau under the null model that it evolves at the same rate following duplication and speciation events.
+First, we simulated the evolution of Tau under the null model that it evolves at the same rate following duplication and speciation events.
 Under the null model, the plot of correlation coefficient to node age is very similar as for the observed data (Figure 5a).
@@
 (b) The distributions after the exclusion of the ctenophore.
-The early and mid phase distributions are not statistically distinct.](levin_etal/Figure_levin.png?raw=true)
+The early and mid phase distributions are not statistically distinct.](levin_etal/Figure_levin.pdf)
 
@@
 
-Thanks to XX. This work was supported by the National Science Foundation (DEB-1256695 and the Waterman Award).
+Thanks to Steve Haddock, Alex Damian Serrano, and August Guang for feedback on the manuscript. This work was supported by the National Science Foundation (DEB-1256695 and the Waterman Award).
 
@@
 
-```{r Tau_v_max, cache=TRUE, echo=FALSE, warning=FALSE, dev="png", dpi=300, message=FALSE, fig.cap="There is a negative relationship between Tau and the maximum observed expression for each gene."}
+There is a negative relationship between miximum expression and Tau for each gene (Supplementary Figure 9).
+
+```{r Tau_v_max, cache=TRUE, echo=FALSE, warning=FALSE, dev="png", dpi=300, message=FALSE, fig.cap="Genes with higher maximum observed expression tend to have lower Tau.
+The blue line is the linear model for the relationship between the two."}
 
@@
 		geom_point( size=0.5, alpha = 0.05 ) + 
-		geom_smooth( method="lm" ) +
+		geom_smooth( method="lm", se=FALSE ) +
 		xlab( "Maximum Expression" ) +
+		ylim( 0, 1 ) +
 		theme_classic()
@@
 ```
+
+\pagebreak
 

### 2017-02-08 manuscript.rmd
reformatted with new bib, cherry picked bm model parameters to reduce memory footprint

@@
 		names( tau_original ) = phy$tip.label
-		nhx@phylo$model = fitContinuous( phy, tau_original, ... )
+		brownian_model = fitContinuous( phy, tau_original, ... )
+
+		nhx@phylo$model$opt$z0 = brownian_model$opt$z0
+		nhx@phylo$model$opt$sigsq = brownian_model$opt$sigsq
 

### 2017-02-08 manuscript.rmd
re knit from scratch

@@
 
+	#' Summarize pairwise comparisons between tips of a tree
+	#' 
+	#' @param nhx A phylogenetic tree as a treeio::treedata object 
+	#' 	
+	#' @return A dataframe
+	#' @export
+	get_pairwise_summary = function ( nhx ) {
+
+		# Create a tibble with one row for each pairwise combination of tip nodes
+		ntips = length( nhx@phylo$tip.label )
+		D = 1:ntips %>% combn( 2 ) %>% t() %>% as_tibble()
+		names( D ) = c( "tip_a", "tip_b" )
+
+		# Add a column with the gene tree hash 
+		D %<>% mutate( gene_tree = digest( nhx ) )
+
+		# Add the tip names
+		D %<>% mutate ( name_a = nhx@data$G[ D$tip_a ] )
+		D %<>% mutate ( name_b = nhx@data$G[ D$tip_b ] )
+
+		# Add the species
+		D %<>% mutate ( species_a = nhx@data$species[ D$tip_a ] )
+		D %<>% mutate ( species_b = nhx@data$species[ D$tip_b ] )
+
+		# Add the values at the tips
+		D %<>% mutate ( tau_a = nhx@data$Tau[ D$tip_a ] )
+		D %<>% mutate ( tau_b = nhx@data$Tau[ D$tip_b ] )
+		D %<>% mutate ( difference = abs( tau_a - tau_b ) )
+
+		# For each row, get the most recent common ancestor of the pairwise tip combination
+		D$mrca = 
+			apply(
+				D, 
+				1, 
+				function( y ) ape::getMRCA( 
+					nhx@phylo, 
+					c( as.integer( y['tip_a'] ), as.integer( y['tip_b'] ) )
+				)
+			) %>%
+			as.integer()
+
+		# Merge data of this mrca to the node
+		D %<>% left_join(
+			nhx@data %>%
+				select( D, Event, node, node_age, label, pic ),
+			by = c( "mrca" = "node")
+
+		)
+
+		# Calculate the distance between the tips, which for an ultrametric tree 
+		# is twice the age of their most recent common ancestor
+		D %<>% mutate( distance = 2 * node_age )
+
+		return( D )
+
+	}
+
 ```
@@
 One reason that comparative functional genomics has not yet embraced phylogenetic approaches is that it has not yet been concretely demonstrated that pairwise and phylogenetic comparative methods can lead to different results when considering functional genomics data.
-Here we examine this issue be re-evaluating the pairwise comparisons used in two recent studies that compared gene expression across species.
+Here we examine this issue by re-evaluating the pairwise comparisons used in two recent studies that compared gene expression across species.
 
@@
 
+	# Remove the large string objects to free up memory
+	rm( lines )
+	rm( tree_lines )
+
 ```
@@
 
-```{r show_contrast_result, echo=FALSE, cache=TRUE, message=F, warning=F, fig.height=2, fig.width=4, fig.cap="Phylogneetic reanalysis of the Tau values calculated by KMRR [@KryuchkovaMostacci:2016iw].
+```{r show_contrast_result, echo=FALSE, cache=TRUE, message=F, warning=F, fig.height=2, fig.width=4, fig.cap="Phylogenetic reanalysis of the Tau values calculated by KMRR [@KryuchkovaMostacci:2016iw].
 Density plot of the magnitude of phylogenetic independent contrasts for Tau following duplication and speciation events.
@@
 The Wilcoxon rank test does not reject the null hypothesis that the rate of evolution following duplications is the same as or less than the rate following speciation (p value = `r wilcox_test_result`).
-Our phylogenetic comparative analysis, unlike the previously published pairwise comparative analysis [@KryuchkovaMostacci:2016iw], therefore find no evidence for the ortholog conjecture in this system.
+Our phylogenetic comparative analysis, unlike the previously published pairwise comparative analysis [@KryuchkovaMostacci:2016iw], therefore finds no evidence for the ortholog conjecture in this system.
 
@@
 Following KMRR Figure 2a, this figure shows only the comparisons that include a human sequence."}
-	get_pairwise_summary = function ( nhx ) {
-
-		# Create a tibble with one row for each pairwise combination of tip nodes
-		ntips = length( nhx@phylo$tip.label )
-		D = 1:ntips %>% combn( 2 ) %>% t() %>% as_tibble()
-		names( D ) = c( "tip_a", "tip_b" )
-
-		# Add a column with the gene tree hash 
-		D %<>% mutate( gene_tree = digest( nhx ) )
-
-		# Add the tip names
-		D %<>% mutate ( name_a = nhx@data$G[ D$tip_a ] )
-		D %<>% mutate ( name_b = nhx@data$G[ D$tip_b ] )
-
-		# Add the species
-		D %<>% mutate ( species_a = nhx@data$species[ D$tip_a ] )
-		D %<>% mutate ( species_b = nhx@data$species[ D$tip_b ] )
-
-		# Add the values at the tips
-		D %<>% mutate ( tau_a = nhx@data$Tau[ D$tip_a ] )
-		D %<>% mutate ( tau_b = nhx@data$Tau[ D$tip_b ] )
-		D %<>% mutate ( difference = abs( tau_a - tau_b ) )
-
-		# For each row, get the most recent common ancestor of the pairwise tip combination
-		D$mrca = 
-			apply(
-				D, 
-				1, 
-				function( y ) ape::getMRCA( 
-					nhx@phylo, 
-					c( as.integer( y['tip_a'] ), as.integer( y['tip_b'] ) )
-				)
-			) %>%
-			as.integer()
-
-		# Merge data of this mrca to the node
-		D %<>% left_join(
-			nhx@data %>%
-				select( D, Event, node, node_age, label, pic ),
-			by = c( "mrca" = "node")
-
-		)
-
-		# Calculate the distance between the tips, which for an ultrametric tree 
-		# is twice the age of their most recent common ancestor
-		D %<>% mutate( distance = 2 * node_age )
-
-		return( D )
-
-	}
 
@@
 In their Figure 1, they present a higher Tau correlation coefficient between ortholog pairs than paralog ortholog pairs.
-We find the same here, with correlation coefficients of `r round( ortholog_r, 3)` for orthologs and `r round( paralog_r, 3)` for paralogs.
+We find the same here, with correlation coefficients of `r round( ortholog_r, 2)` for orthologs and `r round( paralog_r, 2)` for paralogs.
 
@@
 This is, in fact, the case for these data.
-While the mean distance between orthologs is `r round( ortholog_distance_mean, 1 )` million years, the mean distance between paralogs is `r round( paralog_distance_mean, 1 )` million years.
+While the mean distance (*i.e.*, total branch length) between orthologs is `r round( ortholog_distance_mean, 1 )` million years, the mean distance between paralogs is `r round( paralog_distance_mean, 1 )` million years.
 This is because the oldest speciation event is by definition the most recent common ancestor of the species included in the study, but many gene families underwent duplication before this time.
 
-To test the hypothesis that distantly paralogs that arose from duplications that precede the oldest speciation event (Supplementary Figure 8a) impact the lower correlation of Tau between paralogs than between orthologs, we removed them.
-When we consider only the duplication events the same age or younger than the oldest speciation event (Supplementary Figure 8b), the paralog correlation coefficient increases from `r round( paralog_r, 3)` to  `r round( paralog_subset_r, 3)`.
-This is much closer to the ortholog correlation of `r round( ortholog_r, 3) `.
+To test the hypothesis that ancient duplications that precede the oldest speciation event (Supplementary Figure 8a) impact the lower correlation of Tau between paralogs than between orthologs, we removed them.
+When we consider only the duplication events the same age or younger than the oldest speciation event (Supplementary Figure 8b), the paralog correlation coefficient increases from `r round( paralog_r, 2)` to `r round( paralog_subset_r, 2)`.
+This is much closer to the ortholog correlation of `r round( ortholog_r, 2) `.
 
@@
 ```{r simulations_plot, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Analyses of simulated data reveal that pairwise methods produce similar results under both the null model and the ortholog conjecture (OC), while the phylogenetic method provides different results under each model.
-(a) Pairwise analysis of dataset simulated under the null model that there is no difference in the rate of expresison evolution following duplication and speciation.
+(a) Pairwise analysis of dataset simulated under the null model that there is no difference in the rate of expression evolution following duplication and speciation.
 (b) Phylogenetic analysis of data simulated under the null model.
-(c) Pairwise analysis of dataset simulated under the ortholog conjecture, in which there is a higher rate of expresison evolution following duplication than speciation.
+(c) Pairwise analysis of dataset simulated under the ortholog conjecture, in which there is a higher rate of expression evolution following duplication than speciation.
 This plot is very similar to the corresponding plot for the null model.
@@
 We just don't find evidence for it when summarizing gene expression across tissues with Tau in these organisms.
-This calls into question the general predictive power of the ortholog conjecture, and until these processes are better understood it will be necessary to test for it in each situation.
+This calls into question the general predictive power of the ortholog conjecture with respect to gene expression, and until these processes are better understood it will be necessary to test for it in each situation.
 These tests should be articulated in terms of clear alternative hypotheses [@Studer:2009bm] that make distinct phylogenetic comparative predictions.
@@
 For each node in the remaining calibrated trees, we calculated the phylogenetic independent contrast for Tau across its daughter branches with the `pic()` function in ape [@Paradis:2004dv].
-We then collected the contrasts from all trees into a single table, along with otherr annotations including whether the node is a speciation or duplication event.
+We then collected the contrasts from all trees into a single table, along with other annotations including whether the node is a speciation or duplication event.
 This table, `nodes_contrast`, was then analyzed as described in the main text for the presented plots and tests.
@@
 
-Thanks to Steve Haddock, Alex Damian Serrano, and August Guang for feedback on the manuscript. This work was supported by the National Science Foundation (DEB-1256695 and the Waterman Award).
+Thanks to Steve Haddock, Alex Damian Serrano, August Guang, and Zack Lewis for feedback on the manuscript.
+This work was supported by the National Science Foundation (DEB-1256695 and the Waterman Award).
 

### 2017-02-08 manuscript.rmd
added p values to figures

@@
 
-![Pairwise and phylogenetic comparative approaches.
+![Pairwise and phylogenetic comparative approaches, illustrated on an example gene tree with multiple genes per species.
+The internal nodes of the tree are speciation and gene duplication events.
 (a) Many comparative functional genomic studies rely on pairwise comparisons, where traits of each gene are compared to traits of other genes (such as orthologs and paralogs).
@@
 
+	label_p = str_c( "Wilcoxon p=", signif( wilcox_test_result, 2 ) )
+
 	# Plot the contrasts for duplication events and speciation events
@@
 		theme_classic() + 
+		annotate( "text", x = Inf, y = 50, hjust="right", vjust="top", label=label_p ) +
 		theme( legend.justification=c( 1, 1 ), legend.position=c( 1, 1 ) )
@@
 In their Figure 1, they present a higher Tau correlation coefficient between ortholog pairs than paralog ortholog pairs.
-We find the same here, with correlation coefficients of `r round( ortholog_r, 2)` for orthologs and `r round( paralog_r, 2)` for paralogs.
+We find the same here, with correlation coefficients of `r signif( ortholog_r, 2)` for orthologs and `r signif( paralog_r, 2)` for paralogs.
 
@@
 To test the hypothesis that ancient duplications that precede the oldest speciation event (Supplementary Figure 8a) impact the lower correlation of Tau between paralogs than between orthologs, we removed them.
-When we consider only the duplication events the same age or younger than the oldest speciation event (Supplementary Figure 8b), the paralog correlation coefficient increases from `r round( paralog_r, 2)` to `r round( paralog_subset_r, 2)`.
-This is much closer to the ortholog correlation of `r round( ortholog_r, 2) `.
+When we consider only the duplication events the same age or younger than the oldest speciation event (Supplementary Figure 8b), the paralog correlation coefficient increases from `r signif( paralog_r, 2)` to `r signif( paralog_subset_r, 2)`.
+This is much closer to the ortholog correlation of `r signif( ortholog_r, 2) `.
 
@@
 This plot is distinct from that for the null model in that phylogenetic independent contrasts for duplication events are significantly larger than after speciation events."}
+
+	label_p_sim_null = str_c( "Wilcoxon p=", signif( wilcox_oc( nodes_sim_null_contrast ), 2 ) )
 
@@
 			ylim( 0, 75 ) +
+			annotate( "text", x = Inf, y = 20, hjust="right", vjust="top", label=label_p_sim_null ) +
 			theme_classic() + 
@@
 
+	label_p_sim_oc = str_c( "Wilcoxon p=", signif( wilcox_oc( nodes_sim_oc_contrast ), 2 ) )
+
 	g_phylo_oc = 
@@
 		ylim( 0, 75 ) +
+		annotate( "text", x = Inf, y = 30, hjust="right", vjust="top", label=label_p_sim_oc ) +
 		theme_classic() + 
@@
 To better understand this plot (Figure 4), we performed a couple simulations of Tau on the calibrated gene trees and then regenerated the figures.
+We did not modify the gene tree topologies or their inferred histories of duplication and loss.
 First, we simulated the evolution of Tau under the null model that it evolves at the same rate following duplication and speciation events.
 Under the null model, the plot of correlation coefficient to node age is very similar as for the observed data (Figure 5a).
-As in the original study, there is higher correlation coefficient across orthologs (`r ortholog_r_sim_null`) than paralogs (`r paralog_r_sim_null`) when not considering node age.
+As in the original study, there is higher correlation coefficient across orthologs (`r signif( ortholog_r_sim_null, 2 )`) than paralogs (`r signif( paralog_r_sim_null, 2 )`) when not considering node age.
 Phylogenetic analysis of the data simulated under the null hypothesis (Figure 5b) do not reject the null hypothesis (Wilcoxon p = $`r wilcox_oc( nodes_sim_null_contrast )`$), as expected.
@@
 We next simulated the evolution of Tau under the ortholog conjecture, where the rate of evolution of Tau following duplication was `r dup_adjust` fold the rate following speciation.
-The pairwise results of this heterogeneous model are nearly indistinguishable from the null model (Figure 5c), and also have a higher correlation coefficient for orthologs (`r ortholog_r_sim_oc`) than paralogs (`r paralog_r_sim_oc`).
-The phylogenetic analysis of the ortholog conjecture simulation (Figure 5d) does reject the null hypothesis (Wilcoxon p = $`r wilcox_oc( nodes_sim_oc_contrast )`$).
+The pairwise results of this heterogeneous model are nearly indistinguishable from the null model (Figure 5c), and also have a higher correlation coefficient for orthologs (`r signif( ortholog_r_sim_oc, 2 )`) than paralogs (`r signif( paralog_r_sim_oc, 2 )`).
+The phylogenetic analysis of the ortholog conjecture simulation (Figure 5d) does reject the null hypothesis (Wilcoxon p = $`r signif( wilcox_oc( nodes_sim_oc_contrast ), 2 )`$).
 
@@
 
-We examined the matrix of pairwise comparisons used as the base for the KS tests and Figure 4d in *Levin et al.* [@Levin:2016bp], and thus as evidence to support the "inverse hourglass model".
+We examined the matrix of pairwise comparisons used as the base for the KS tests and Figure 4d in *Levin et al.* [@Levin:2016bp], and thus as evidence to support the "inverse hourglass" model.
 We found several problems resulting from the use of multiple pairwise comparisons.
@@
 
-Once we addressed the issues above with the implementation of pairwise comparisons in Levin *et al.* [@Levin:2016bp], we were able to explore more general issues that can be a problem when making multiple pairwise comparisons between the same species.
+Once we addressed the issues above with the implementation of pairwise comparisons, we were able to explore more general issues that can be a problem when making multiple pairwise comparisons between the same species.
 We found that all five of the lowest values in the mid phase distribution (Figure 6a) are for pairwise comparisons that include the ctenophore (comb jelly).
-When the nine pairwise comparisons that include the ctenophore are removed, there is no significant difference between the early phase and mid phase distributions (P = 0.1428 for the early-mid comparison and P < $10^{-5}$ for the late-mid comparison) and no support for the inverse hourglass (Figure 6b).
-This highlights a well understood property of pairwise comparisons across species [@Felsenstein:1985ua]: evolutionary changes along a given branch, like those along the ctenophore branch, impact each of the multiple pairwise comparisons that includes that branch.
+When the nine pairwise comparisons that include the ctenophore are removed, there is no significant difference between the early phase and mid phase distributions (p = 0.14 for the early-mid comparison and p < $10^{-5}$ for the late-mid comparison) and no support for the inverse hourglass (Figure 6b).
+This highlights a well understood property of pairwise comparisons across species [@Felsenstein:1985ua; @Ackerly:1999iu]: evolutionary changes along a given branch, like those along the ctenophore branch, impact each of the multiple pairwise comparisons that includes that branch.
 The pairwise comparisons are therefore not independent - different pairwise comparisons are impacted by changes along some of the same branches (Figure 1a).
@@
 The number of comparisons impacted by each change depends on the structure of the phylogenetic tree, *i.e.* how the species are related to each other.
-
 Phylogenetic comparative methods were developed specifically to address this problem [@Felsenstein:1985ua].
-While we do demonstrate this problem impacts the Levin *et al.* analysis, we did not perform a phylogenetic reanalysis of this study, as we did for the KMRR study.
-This is because the similarity metric computed in the pairwise comparisons of Levin *et al.* are based on different genes for different species pairs.
-This means that they do not generate a trait that can be evaluated across a phylogeny, as KMRR's Tau can be.
-A full phylogenetic reanalysis would be possible using upstream analysis products to re-derive new expression summary statistics suitable for phylogenetic comparative methods.
+
+While we demonstrate problems with pairwise comparisons that impacts the Levin *et al.* analysis, we did not perform a phylogenetic reanalysis of this study, as we did for the KMRR study.
+This is because the similarity metric computed in the pairwise comparisons of Levin *et al.* is not suitable for phylogenetic analysis.
+One issue with the similarity metric is that it is based on different genes for different species pairs, and therefore is not a trait whose evolution can be modeled across the phylogeny.
+A full phylogenetic reanalysis would be possible using upstream analysis products to re-derive new expression summary statistics.
 
@@
 This is because there are various species-specific technical factors that can be mistaken for differences in expression across species.
-These can be canceled out within species before making comparisons across species.
+These can be canceled out within species before making comparisons across species [@Dunn:2013cpa].
 
@@
 The fact that multiple pairwise comparisons of developmental gene expression across diverse species share a particular pattern is not evidence of a general process that explains the differences between all species in the analysis.
-It is also the expected pattern when there are unique differences specific to a single species, and the evolutionary changes responsible for these differences are sampled multiple times in pairwise comparisons that span the same phylogenetic branches along which they occurred.
+It is also the expected pattern when a single species has unique differences, and the evolutionary changes responsible for these differences are sampled multiple times in pairwise comparisons that span the same phylogenetic branches along which these differences arose.
 To use patterns across living species to test hypotheses about evolutionary processes it is also necessary to incorporate information about evolutionary relationships, *i.e.* phylogenies.
@@
 We then removed duplicate rows, applied the Wilcoxon test in place of the Kolmogorov-Smirnov test, and identified ctenophores as overrepresented among the low outliers in the mid-developmental transition column.
-An annotated explanation of these analyses is included in the git repository at
+An annotated explanation of these analyses is included in the git repository at https://github.com/caseywdunn/comparative_expression_2017/blob/master/levin_etal/reanalyses.md .
 
@@
 
-Thanks to Steve Haddock, Alex Damian Serrano, August Guang, and Zack Lewis for feedback on the manuscript.
+Thanks to Steve Haddock, Alex Damian Serrano, August Guang, Bruno Vellutini, and Zack Lewis for feedback on the manuscript.
 This work was supported by the National Science Foundation (DEB-1256695 and the Waterman Award).

### 2017-02-08 manuscript.rmd
word smithing

@@
 One goal of this work is to provide an integrated understanding of genome and phenotype evolution.
-Most studies of this type have relied on multiple pairwise comparisons between species, an approach that does not incorporate information about the evolutionary relationships among species.
+Most studies of this type have relied on multiple pairwise comparisons of functional genomic data between species, an approach that does not incorporate information about the evolutionary relationships among species.
 The statistical problems that arise from not considering these relationships can lead pairwise approaches to the wrong conclusions, and are a missed opportunity to learn about biology that can only be understood in an explicit phylogenetic context.
@@
 
-A rich theoretical and statistical methodology of phylogenetic comparative methods have been developed over the last three decades to address the challenges and opportunities of trait comparisons across species [@Felsenstein:1985ua; @Grafen:1989um; @Pagel:1999gc; Revell:2012jd; @FitzJohn:2012ey; @Uyeda:2014jl; @Garamszegi:2014ew].
+A rich theoretical and statistical methodology of phylogenetic comparative methods has been developed over the last three decades to address the challenges and opportunities of trait comparisons across species [@Felsenstein:1985ua; @Grafen:1989um; @Pagel:1999gc; Revell:2012jd; @FitzJohn:2012ey; @Uyeda:2014jl; @Garamszegi:2014ew].
 A central challenge is the dependence of observations across species due to the evolutionary history of species -- more closely related species share many traits that evolved once in a common ancestor.
@@
 They have largely been applied to morphological and ecological traits, but are just as relevant to functional genomics.
-Even so, most comparative functional genomics studies have abstained from phylogenetic approaches and instead rely on multiple pairwise comparisons (Figure 1a).
+Even so, most comparative functional genomics studies have abstained from phylogenetic approaches and instead rely on multiple pairwise comparisons across species (Figure 1a).
 This leaves comparative functional genomics studies susceptible to statistical problems and is a missed opportunity to ask questions that are only accessible in an explicit phylogenetic context.
@@
 If phylogenetic relationships are not considered, analyses indicate a negative correlation of leaf lifespan with leaf size.
-With the application of phylogenetic comparative methods, however, this correlation disappears.
-This is because most of the observed variance is due to differences between conifers and flowering plants.
+With the application of phylogenetic comparative methods, however, this correlation disappears because most of the observed variance is due to differences between conifers and flowering plants.
 Phylogenetic comparative methods capture this shift as change along a single branch deep in the tree, revealing that there is not a tendency of correlated change between these traits across the phylogeny.
-One reason that comparative functional genomics has not yet embraced phylogenetic approaches is that it has not yet been concretely demonstrated that pairwise and phylogenetic comparative methods can lead to different results when considering functional genomics data.
-Here we examine this issue by re-evaluating the pairwise comparisons used in two recent studies that compared gene expression across species.
+One reason that comparative functional genomic sudies have not embraced phylogenetic approaches is that it has not yet been concretely demonstrated that pairwise and phylogenetic comparative methods can lead to different results when considering functional genomics data.
+Here we examine this issue by re-evaluating the pairwise comparisons in two recent studies that compared gene expression across species.
 
@@
 This study analyzed gene expression through the course of embryonic development for ten animal species, each from a different phylum.
-Using pairwise comparisons, they found there is more evolutionary variance in gene expression at a mid phase of development than there is at early and late phases (Figure 1a).
-They suggest that this supports an "inverse hourglass" model for the evolution of gene expression, in contrast with the "hourglass" model previously proposed for closely related species [@Kalinka:2010dw], and provides biological justification for the concept of phyla.
+Using pairwise comparisons, they found there is more evolutionary variance in gene expression at a mid phase of development than there is at early and late phases.
+They suggest that this supports an "inverse hourglass" model for the evolution of gene expression, in contrast with the "hourglass" model previously proposed for closely related species [@Kalinka:2010dw].
+Furthermore, they suggested that this provides biological justification for the concept of phyla.
 We previously described concerns with the interpretations of this result [@Hejnol:2016hw].
@@
 The internal nodes of the tree are speciation and gene duplication events.
-(a) Many comparative functional genomic studies rely on pairwise comparisons, where traits of each gene are compared to traits of other genes (such as orthologs and paralogs).
+(a) Many comparative functional genomic studies rely on pairwise comparisons, where traits of each gene are compared to traits of other genes across species.
 This leads to many more comparisons than unique observations, making each comparison dependent on others.
 (b) Comparative phylogenetic methods, including phylogenetic independent contrasts [@Felsenstein:1985ua], make a smaller number of independent comparisons, where each contrast measures independent changes along different branches.
-Phylogenetic approaches are rarely used for functional genomics studies.](Figure_overview.pdf)
+Phylogenetic approaches are rarely used for functional genomic studies.](Figure_overview.pdf)
 
@@
 It shapes our understanding of the functional diversity of gene families.
-It is used to relate findings from well-studied genes to related genes that have not been investigated in detail.
+It is also used to relate findings from well-studied genes to related genes that have not been investigated in detail.
 It can be applied to any trait of genes, from gene sequence to biochemical properties to expression.
@@
 It can range from a value of 0, which indicates no specificity (*i.e.*, uniform expression across tissues), to a value of 1, which indicates high specificity (*i.e.*, expression in only one tissue).
-It is convenient in that it is a single number of defined range for each gene, though of course since the original expression is multidimensional this means much information is discarded.
+Tau is convenient in that it is a single number of defined range for each gene, though of course since the original expression is multidimensional this means much information is discarded.
 This includes information about which tissue expression is specific to.
@@
 
-For each internal node in each gene tree, we measured the phylogenetic independent contrast (PIC) of the expression statistic Tau.
+For each internal node in each gene tree, we calculated the phylogenetic independent contrast (PIC) of Tau.
 This is the difference in values of Tau for descendant nodes scaled by the expected variance, which is largely determined by the lengths of the two branches that connect the node to its two descendants (Figure 1b).
 These contrasts were then annotated by whether each is made across a speciation or duplication event.
-The original description of independent contrasts [@Felsenstein:1985ua] focused on assessing covariance between changes in two traits on a tree.
-Our use of contrasts is a bit different-- we look for differences in evolutionary changes of one trait (differential expression) between two categories of nodes (speciation and duplication) on a tree.
+The original description of independent contrasts [@Felsenstein:1985ua] focused on assessing covariance between changes in two traits.
+Our use of contrasts is a bit different-- we look for differences in evolutionary changes of one trait (differential expression) between two categories of nodes (speciation and duplication).
 
@@
 These are the same pre-computed trees that the orthology/ paralogy annotations KMRR used are based on.
-`r length(gene_trees_pruned)` passed taxon sampling criteria (`r min_genes_with_expression` genes) after removing tips without Tau values and had at least one speciation event.
+`r length(gene_trees_pruned)` gene trees passed taxon sampling criteria (`r min_genes_with_expression` genes) after removing tips without Tau values and had at least one speciation event.
 Of these, `r length(gene_trees_calibrated)` were successfully time calibrated.
@@
 One reason is that multiple pairwise comparisons repeatedly sample the same evolutionary changes and in so doing violate statistical assumptions of independence, whereas phylogenetic comparative methods make multiple independent comparisons across non-overlapping branches of the tree.
-The other reason for the different conclusions is that pairwise comparisons between the tips and phylogenetic comparative methods describe different things.
+The other reason for the different conclusions is that pairwise comparisons and phylogenetic comparative methods describe different things.
 Pairwise comparisons describe contemporary patterns, while phylogenetic methods infer historical processes [@OMeara:2012gq].
 There need not be a different process of evolution following speciation and duplication for paralogs to be more different than orthologs.
-This could be due to the structure of the gene phylogenies.
+Any difference could be due to the structure of the gene phylogenies alone.
 If paralogs tend to be more distantly related to each other than orthologs, then there would be more time for differences to accumulate even if the rate of change is the same between the two.
@@
 In their Figure 2, they grouped orthologs and paralogs according to the ENSEMBL node name of their most recent common ancestor, and plotted the correlation of Tau for each of these groups by the node age.
-They found that across the investigated range of node ages, ortholog pairs have higher tau correlation than paralogs.
+They found that across the investigated range of node ages, ortholog pairs have higher Tau correlation than paralogs.
 We confirmed that we can replicate this result (Figure 4).
@@
 The pairwise comparisons used by KMRR cannot distinguish between the null hypothesis and ortholog conjecture even when the evolution of expression is `r dup_adjust` fold higher following duplication than speciation.
-The pairwise results are the same under both hypotheses.
-They also serve as a validation of phylogenetic methods as applied to this problem.
-Our analysis of independent contrasts does not reject the null hypothesis when data are simulated under the null model, and does reject the null hypothesis when the data are simulated under the ortholog conjecture.
+The pairwise results are strikingly similar under both hypotheses.
+These simulations also serve as a validation of phylogenetic methods as applied to this problem.
+Our phylogenetic analysis of independent contrasts does not reject the null hypothesis when data are simulated under the null model, and does reject the null hypothesis when the data are simulated under the ortholog conjecture.
 In contrast to pairwise methods, the phylogenetic analyses make testable predictions based on explicit hypotheses about evolutionary process.
@@
 
-Once we addressed the issues above with the implementation of pairwise comparisons, we were able to explore more general issues that can be a problem when making multiple pairwise comparisons between the same species.
+Once we addressed the issues above with the implementation of pairwise comparisons, we were able to explore more general issues that can be a problem when making multiple pairwise comparisons between species.
 We found that all five of the lowest values in the mid phase distribution (Figure 6a) are for pairwise comparisons that include the ctenophore (comb jelly).

### 2017-02-08 manuscript.rmd
added funding

@@
 Thanks to Steve Haddock, Alex Damian Serrano, August Guang, Bruno Vellutini, and Zack Lewis for feedback on the manuscript.
-This work was supported by the National Science Foundation (DEB-1256695 and the Waterman Award).
-
+CWD's contribution was supported by the National Science Foundation (DEB-1256695 and the Waterman Award).
+AH was supported by the European Research Council Community’s Framework Program Horizon 2020 (2014–2020) ERC grant agreement 648861.
 \pagebreak

### 2017-02-11 manuscript.rmd
reduced memory footprint

@@
 
+	n_gene_trees = length( gene_trees )
+
 	# Parse the tree annotations, these are lines that start with SEQ
@@
 
+	# Free up memory
+	rm( tip_annotations )
+
 ```
@@
 
+	# Free up memory
+	rm( gene_trees )
+
 	# Remove tips from the trees that do not have expression data
@@
 	)
+
+	# Free up memory
+	rm( gene_trees_annotated )
 
@@
 
-We mapped the Tau values calculated by KMRR [@KryuchkovaMostacci:2016iw] for the Brawand *et al.* 2011 [@Brawand:2011du] datset onto `r length(gene_trees)` gene trees parsed from ENSMBL Compara [@Herrero:2016jj].
+We mapped the Tau values calculated by KMRR [@KryuchkovaMostacci:2016iw] for the Brawand *et al.* 2011 [@Brawand:2011du] datset onto `r n_gene_trees` gene trees parsed from ENSMBL Compara [@Herrero:2016jj].
 These are the same pre-computed trees that the orthology/ paralogy annotations KMRR used are based on.
@@
 Model of variance, which is inversely related to similarity, is on the left.
-(a) The distributions as published.
+(a) The distributions as published by Levin *et al.* [@Levin:2016bp].
 Low similarity (*i.e.*, high variance) in the mid phase of development was interpreted as support for an inverse hourglass model for the evolution of gene expression.

### 2017-02-12 manuscript.rmd
modified tree parsing so nearly all are now time callibrated. Moved functions to separate file.

@@
 
-	#' Adds some random noise from a normal distribution to calibration
-	#' times
-	#' 
-	#' @param calibration_times A dataframe with two columns: age.min is the age 
-	#' of a clade, and clade is the name of the clade
-	#' @param sd_fraction this is multiplied by each value to set sd of that value 
-	#' @return A dataframe as above, but with noise
-	#' @export
-	add_noise_calibration_times = function( calibration_times, sd_fraction ) {
-		age = calibration_times$age
-		new_age = NA
-
-		# Keep trying until dates are coherent with tree structure
-		# This is a ladder tree, so that means checking that they 
-		# are still in ascending order
-		while( NA %in% new_age ) {
-			x = sapply( 
-				age,
-				function( i ){
-					rnorm( 1, mean = i, sd = ( sd_fraction * i ) )
-				}
-			)
-
-			if( ( ! is.unsorted( x ) ) & all( x > 0 ) ){
-				new_age = x
-			}
-
-		}
-
-		calibration_times$age = new_age
-		return( calibration_times )
-	}
-
-	#' Get a boolean vector corresponding to all tips and internal nodes of 
-	#' a tree, with value TRUE for tips and FALSE for internal nodes
-	#' 
-	#' @param nhx A phylogenetic tree as a treeio::treedata object
-	#' @return A boolean vector
-	#' @export
-	is.tip.nhx = function( nhx ) {
-		is.tip = rep( FALSE, nrow( nhx@data ) )
-		is.tip[ 1:length( nhx@phylo$tip.label ) ] = TRUE
-		is.tip
-	}
-
-	#' Fix ensembl clade names.
-	#' Homininae is pan + homo + gorilla
-	#' Hominini is pan + homo
-	#' ENSEMBL Compara trees lislabel Hominini as Homininae
-	#' 
-	#' @param nhx A phylogenetic tree as a treeio::treedata object
-	#' @return A phylogenetic tree as a treeio::treedata object
-	#' @export
-	fix_hominini = function( nhx ) {
-		# Get the numbers for all nodes labeled as homininae
-		homininae_nodes = nhx@data$node[ nhx@data$label == "Homininae" ]
-
-		# Check which have no gorilla as a descendent
-		no_gorilla = sapply(
-			homininae_nodes,
-			function( x ){
-				tips = hutan::tip_descendants( nhx@phylo, x )
-				tip_species = nhx@data$species[ tips ]
-				return( ! "gorilla_gorilla" %in% tip_species )
-			}
-		)
-
-		if( any( no_gorilla ) ){
-			hominini_nodes = homininae_nodes[ no_gorilla ]
-			nhx@data$label[ hominini_nodes ] = "Hominini"
-		}
-
-		return( nhx )
-	}
-
-	#' Record the distance from a tip for each node in a tree in 
-	#' the data slot
-	#' 
-	#' @param nhx A phylogenetic tree as a treeio::treedata
-	#' @return A treeio::treedata object, with an additional node_age column
-	#' in the data slot, or the original object if it was not of class treedata
-	#' @export
-	store_node_age = function( nhx ) {
-
-		if ( class( nhx ) != "treedata" ) {
-			return( nhx )
-		}
-
-		node_age = hutan::distance_from_tip( nhx@phylo )
-
-		# make sure the dataframe is ordered by consecutive nodes
-		stopifnot( all( nhx@data$node == 1:length( nhx@data$node ) ) )
-
-		nhx@data$node_age = node_age
-
-		return( nhx )
-	}
-
-	#' Get the number of speciation events in an nhx tree
-	#' 
-	#' @param nhx A phylogenetic tree as a treeio::treedata object, with a 
-	#' column @data$D that has value N for speciation events
-	#' @return An integer indicating the number of speciation events
-	#' @export
-	get_n_speciation = function( nhx ){
-		sum( nhx@data$D[ ! is.tip.nhx( nhx ) ] == "N", na.rm = TRUE )
-	}
-
-	#' Drop tips without expression data from a tree. 
-	#' 
-	#' @param nhx A phylogenetic tree as a treeio::treedata object, with a 
-	#' column @data$Tau containing expression data
-	#' @param min_genes_with_expression The minimum number of tips with 
-	#' expression data for a tree to be retained
-	#' @return A treeio::treedata object, or NA if less than 
-	#' min_genes_with_expression
-	#' @export
-	drop_empty_tips = function( nhx, min_genes_with_expression ) {
-
-		# Identify the tips without expression data
-		to_drop = which( is.na( nhx@data[ 1:length( nhx@phylo$tip.label ), ]$Tau ) )
-
-		# Return the pruned tree if it meets sampling criteria, otherwise return NA
-		remaining = length( nhx@phylo$tip.label ) - length( to_drop )
-		if( remaining >= min_genes_with_expression ){
-			pruned = treeio::drop.tip( nhx, to_drop )
-			return( pruned )
-		}
-		else{
-			return( NA )
-		}
-	}
-
-	#' Adjust branch lengths of a phylogenetic tree to make it ultrametric and 
-	#' to time calibrate the speciation nodes to fixed values
-	#' 
-	#' @param nhx A phylogenetic tree as a treeio::treedata object, with speciation 
-	#' nodes annotated with a value of "N" in @data$D
-	#' @param calibration_times A dataframe with two columns: age is the age 
-	#' of a clade, and clade is the name of the clade
-	#' @param ... Any additional arguments to pass to ape::chronos()
-	#' @return A treeio::treedata object if successfully calibrated
-	#' @export
-	calibrate_tree = function ( nhx, calibration_times, ... ) {
-
-		# Create calibration matrix for speciation nodes
-		calibration = 
-			nhx@data[ !is.tip.nhx( nhx ), ] %>%
-			filter( D == "N" ) %>%
-			left_join( calibration_times, c( "label" = "clade" ) ) %>%
-			mutate( age.min = age ) %>%
-			mutate( age.max = age ) %>% 
-			select( node, age.min, age.max ) %>%
-			mutate( soft.bounds = NA )
-
-		tree = try( 
-			ape::chronos( nhx@phylo, calibration=calibration, ... ) 
-		)
-
-		if( "phylo" %in% class( tree ) ){
-			class( tree ) = "phylo"
-			nhx@phylo = tree
-			return( nhx )
-		}
-		else{
-			return( NA )
-		}
-	}
-
-	#' Calculate phylogenetic independent contrasts (PIC) with the tree and 
-	#' character data in a treeio::treedata object
-	#' 
-	#' @param nhx A phylogenetic tree as a treeio::treedata object, with a 
-	#' column @data$Tau containing expression data
-	#' @return A treeio::treedata object, with new @data columns pic and var_exp
-	#' @export
-	pic.nhx = function( nhx ) {
-
-		tau = nhx@data$Tau[ is.tip.nhx( nhx ) ]
-
-		if( any( is.na(tau) ) ){
-			stop( "A trait value is NA." )
-		}
-
-		# Calculate the contrasts
-		p = pic( tau, nhx@phylo, var.contrasts=TRUE )
-
-		# Add the results back to the @data slot, padding the rows that correspond to 
-		# tips with NA
-		nhx@data$pic = c( rep( NA, length( nhx@phylo$tip.label ) ), p[ ,1 ] )
-		nhx@data$var_exp = c( rep( NA, length( nhx@phylo$tip.label ) ), p[ ,2 ] )
-
-		return( nhx )
-	}
-
-	#' Add phylogenetic independent contrasts (PIC) with the tree and 
-	#' character data in a treeio::treedata object
-	#' 
-	#' @param gene_trees_calibrated A list of time calibrated phylogenetic trees 
-	#' as treeio::treedata objects 
-	#' @return A list of time phylogenetic trees with pics 
-	#' as treeio::treedata objects
-	#' @export
-	add_pics_to_trees = function( gene_trees_calibrated ) {
-
-		# Calculate independent contrasts for Tau on each tree, storing the results 
-		# back into the @data slot of the tree objects
-		gene_trees_pic = mclapply( gene_trees_calibrated, pic.nhx, mc.cores=cores )
-
-		return( gene_trees_pic )
-	}
-
-	#' Add phylogenetic independent contrasts (PIC) with the tree and 
-	#' character data in a treeio::treedata object
-	#' 
-	#' @param gene_trees_pic A list of phylogenetic trees with pics
-	#' as treeio::treedata objects 
-	#' @return A tibble with combined data, including phylogenetic independent 
-	#' contrasts, for internal nodes of all trees
-	#' @export
-	summarize_contrasts = function( gene_trees_pic ) {
-		# Combine the @data slots across all trees. These have labels, node annotations,
-		# expression data, contrast results, etc...
-		# Also take the absolute value of the contrast, since we only consider magnitude.
-		nodes_all = 
-			lapply( 
-				gene_trees_pic, 
-				function( nhx ){
-					tags = nhx@data
-					tags$gene = digest( nhx ) # Creates a hash that is unique to each gene tree
-					tags %<>% select( -( B ) ) # B has inconsistent types, remove it
-					return( tags )
-				}
-			) %>%
-			bind_rows() %>%
-			mutate( pic = abs( pic ) )
-
-		# Create a subset that corresponds to nodes with contrasts. This excludes tips.
-		nodes_contrast = 
-			nodes_all %>%
-			filter( ! is.na( pic ) ) %>%
-			filter( ! is.na( D ) )
-
-		return( nodes_contrast )
-	}
-
-	#' Calibrate a list of gene trees given a set of speciation node dates
-	#' 
-	#' @param gene_trees_pruned A list of phylogenetic trees as treeio::treedata objects 
-	#' with associated character data
-	#' @param calibration_times A dataframe with two columns: age.min is the age 
-	#' of a clade, and clade is the name of the clade
-	#' @param ... Any additional arguments to pass to ape::chronos()
-	#' @return A list of calibrated phylogenetic trees as treeio::treedata objects. May be 
-	#' shorter than gene_trees_pruned if some calibrations fail.
-	#' @export
-	calibrate_trees = function( gene_trees_pruned, calibration_times, ... ) {
-		# Make the trees ultrametric and calibrate the speciation nodes to 
-		# specified times
-		gene_trees_calibrated = 
-			mclapply( 
-				gene_trees_pruned,
-				calibrate_tree, 
-				calibration_times=calibration_times,
-				...,
-				mc.cores=cores
-			)
-
-		# Remove trees that could not be successfully calibrated
-		gene_trees_calibrated = gene_trees_calibrated[ ! is.na( gene_trees_calibrated ) ]
-
-		# Parse the calibrated node ages from the internal @phylo object and store them with the 
-		# corresponding rows in the @data object
-		gene_trees_calibrated = 
-			mclapply( 
-				gene_trees_calibrated, 
-				store_node_age, 
-				mc.cores=cores 
-			)
-
-		return( gene_trees_calibrated )
-	}
-
-	#' Wilcoxon test of the ortholog conjecture, specifically that contrasts
-	#' associated with duplication nodes are greater than those for 
-	#' speciation nodes. A significant result indicates rejection of the null
-	#' hypothesis that they are not greater.
-	#' 
-	#' @param nodes_contrast A tibble of independent contrasts
-	#' @return Wilcoxon test p value
-	#' @export
-	wilcox_oc = function( nodes_contrast ) {
-
-		p = wilcox.test( 
-			nodes_contrast %>% filter( D=="Y" ) %>% .$pic, 
-			nodes_contrast %>% filter( D=="N" ) %>% .$pic, 
-			alternative="greater" 
-		)$p.value
-
-		return( p )
-	}
-
-	#' Estimate Tau evolution parameters from a tree::treedata tree
-	#' and add them back into the tree object.
-	#' 
-	#' @param nhx A phylogenetic tree and associated Tau values 
-	#' as a treeio::treedata object
-	#' @param ... Additional arguments to pass to geiger::fitContinuous()
-	#' @return A phylogenetic trees as a treeio::treedata object, 
-	#' with estimated brownian motion model parameters stored 
-	#' in @phylo$model
-	#' @export
-	add_model_parameters = function( nhx, ... ) {
-		phy = nhx@phylo
-		tau_original = nhx@data$Tau [ 1:length( phy$tip.label ) ]
-		names( tau_original ) = phy$tip.label
-		brownian_model = fitContinuous( phy, tau_original, ... )
-
-		nhx@phylo$model$opt$z0 = brownian_model$opt$z0
-		nhx@phylo$model$opt$sigsq = brownian_model$opt$sigsq
-
-		return( nhx )
-	}
-
-	#' Simulate Tau on a tree::treedata. Existing observed values are
-	#' fit to a brownian model to estimate parameters. These parameters
-	#' are then used to replace the original values with similated 
-	#' values.
-	#' 
-	#' @param nhx A phylogenetic tree and associated Tau values 
-	#' as a treeio::treedata object
-	#' @param dup_adjust A multiplier for adjusting the branch lengths 
-	#' following duplication to effectively change the rate following duplication
-	#' @return A phylogenetic trees and simulated Tau values as a 
-	#' treeio::treedata object
-	#' @export
-	sim_tau = function( nhx, dup_adjust=1 ) {
-
-		phy = nhx@phylo
-
-		# Get parameter estimates
-		brownian_model = nhx@phylo$model
-
-		# Adjust length of branches that descend from duplication events. 
-		# This is how we implement heterogeneous rates after duplication 
-		# rather than speciation. Extending the branch lengths is equivalent 
-		# to using a higher rate along them.
-		dup_nodes = nhx@data$node[ nhx@data$D == "Y" ]
-		dup_edges = phy$edge[,1] %in% dup_nodes
-		phy$edge.length[ dup_edges ] = phy$edge.length[ dup_edges ] * dup_adjust
-
-		# Simulate trait given the tree and parameter estimates
-		x = fastBM( 
-				phy, 
-				a=brownian_model$opt$z0, 
-				sig2=brownian_model$opt$sigsq, 
-				bounds=c(-Inf,Inf) 
-			) %>% abs()
-
-		# Adjust the results so they fall between 0 and 1
-		x[ x > 1 ] = 1
-		x[ is.nan( x ) ] = 1
-
-		names( x ) = NULL
-		nhx@data$Tau = c( x, rep( NA, nrow( nhx@data ) - length (x ) ) )
-		return( nhx )
-	}
-
-	#' Simulate a set of pics from a gamma distribution
-	#' 
-	#' @param scale Gamma scale parameter
-	#' @param n Number of observations per event type	
-	#' @return A dataframe
-	#' @export
-	sim_expectation = function( scale = 1, n = 10000 ){
-		expectation = 
-			bind_rows(
-				data.frame( 
-					pic = rgamma( n, 1, scale = scale ),
-					Event = "Duplication",
-					stringsAsFactors = FALSE
-				),
-				data.frame( 
-					pic = rgamma( n, 1, scale = 1 ),
-					Event = "Speciation",
-					stringsAsFactors = FALSE
-				)
-			)
-
-		event_levels = c( "Speciation", "Duplication" )
-		expectation$Event = 
-			factor( expectation$Event, levels=event_levels )
-
-		return( expectation )
-	}
-
-	#' Summarize pairwise comparisons between tips of a tree
-	#' 
-	#' @param nhx A phylogenetic tree as a treeio::treedata object 
-	#' 	
-	#' @return A dataframe
-	#' @export
-	get_pairwise_summary = function ( nhx ) {
-
-		# Create a tibble with one row for each pairwise combination of tip nodes
-		ntips = length( nhx@phylo$tip.label )
-		D = 1:ntips %>% combn( 2 ) %>% t() %>% as_tibble()
-		names( D ) = c( "tip_a", "tip_b" )
-
-		# Add a column with the gene tree hash 
-		D %<>% mutate( gene_tree = digest( nhx ) )
-
-		# Add the tip names
-		D %<>% mutate ( name_a = nhx@data$G[ D$tip_a ] )
-		D %<>% mutate ( name_b = nhx@data$G[ D$tip_b ] )
-
-		# Add the species
-		D %<>% mutate ( species_a = nhx@data$species[ D$tip_a ] )
-		D %<>% mutate ( species_b = nhx@data$species[ D$tip_b ] )
-
-		# Add the values at the tips
-		D %<>% mutate ( tau_a = nhx@data$Tau[ D$tip_a ] )
-		D %<>% mutate ( tau_b = nhx@data$Tau[ D$tip_b ] )
-		D %<>% mutate ( difference = abs( tau_a - tau_b ) )
-
-		# For each row, get the most recent common ancestor of the pairwise tip combination
-		D$mrca = 
-			apply(
-				D, 
-				1, 
-				function( y ) ape::getMRCA( 
-					nhx@phylo, 
-					c( as.integer( y['tip_a'] ), as.integer( y['tip_b'] ) )
-				)
-			) %>%
-			as.integer()
-
-		# Merge data of this mrca to the node
-		D %<>% left_join(
-			nhx@data %>%
-				select( D, Event, node, node_age, label, pic ),
-			by = c( "mrca" = "node")
-
-		)
-
-		# Calculate the distance between the tips, which for an ultrametric tree 
-		# is twice the age of their most recent common ancestor
-		D %<>% mutate( distance = 2 * node_age )
-
-		return( D )
-
-	}
+	source( "functions.R" )
 
@@
 	# for node names
-	gene_trees = mclapply(
-		tree_lines,
-		function( x ){
-			tree_tc = textConnection( x )
-			tree = treeio::read.nhx( tree_tc )
-			close( tree_tc )
-			tree@data$label = c( tree@phylo$tip.label, tree@phylo$node.label )
-			return( tree )
-		}, mc.cores=cores
-	)
+	gene_trees = mclapply( tree_lines, parse_gene_trees, mc.cores=cores )
 
@@
 			tree@data %<>% left_join( expression, by = c( "G" = "Ensembl.Gene.ID" ) )
-
-			tree@data$Event = NA
-			tree@data$Event[ tree@data$D == "N" ] = "Speciation"
-			tree@data$Event[ tree@data$D == "Y" ] = "Duplication"
-			tree@data$Event = factor( tree@data$Event, levels=c( "Speciation", "Duplication" ) )
 

### 2017-02-12 functions.R

@@ -1,0 +1,516 @@
+#' Parses nhx text to a treeio::treedata object
+#' 
+#' @param tree_text Character string repressenting nhx tree
+#' @return A treeio::treedata object
+#' @export
+parse_gene_trees = function( tree_text ){
+	tree_tc = textConnection( tree_text )
+	tree = treeio::read.nhx( tree_tc )
+	close( tree_tc )
+	
+	# Parse clade labels from phylo object into the data frame
+	tree@data$label = c( tree@phylo$tip.label, tree@phylo$node.label )
+	
+	# Compara trees sometimes have speciation nodes whose descendants have the 
+	# same clade name. This isn't biologically possible, so change such nodes 
+	# from speciation events to NA so they don't interfere with tree calibration 
+	# and are not used for calculating speciation contrasts
+	
+	n_nodes = nrow( tree@data )
+	n_tips = length( tree@phylo$tip.label )
+	internal_nodes = ( n_tips + 1 ):n_nodes
+	is_speciation = tree@data$D == "N"
+	is_speciation[ is.na( is_speciation ) ] = FALSE
+	internal_speciation_nodes = tree@data$node[ ( tree@data$node > n_tips ) & is_speciation ]
+	
+	# Create a vector of clade names of speciation events, with NA for all other nodes
+	speciation_names = rep( NA, n_nodes )
+	speciation_names[ internal_speciation_nodes ] = tree@data$label[ internal_speciation_nodes ]
+	
+	# Loop over the internal speciation nodes
+	for( i in internal_speciation_nodes ){
+		# get descendent internal nodes
+		descendants = hutan::descendants( tree@phylo, i )
+		descendants = descendants[ descendants %in% internal_nodes ]
+		
+		descendant_names = speciation_names[ descendants ]
+		
+		if( speciation_names[i] %in% descendant_names ){
+			tree@data$D[i] = NA
+		}
+	}
+	
+	
+	# Create a human readable Event column
+	tree@data$Event = NA
+	tree@data$Event[ tree@data$D == "N" ] = "Speciation"
+	tree@data$Event[ tree@data$D == "Y" ] = "Duplication"
+	tree@data$Event = factor( tree@data$Event, levels=c( "Speciation", "Duplication" ) )
+	
+	return( tree )
+}
+
+
+#' Adds some random noise from a normal distribution to calibration
+#' times
+#' 
[…]

### 2017-02-12 manuscript.rmd
cleaned up plot

@@
 
-	g_age =
-		nodes_contrast %>%
-		ggplot( aes( x=node_age, y=pic, col=Event ) ) + 
-			geom_point( alpha=0.2 ) + 
-			xlab( "Node age (millions of years)" ) + 
-			ylab( "Phylogenetic independent \ncontrast" ) +
-			ylim( 0, 0.3 ) +
-			xlim( 0, 1000 ) +
-			ggtitle( "(a)" ) +
-			theme_classic()	+
-			theme( legend.justification=c( 1, 1 ), legend.position=c( 1, 1 ) )
-
 	max_speciation_age = 
@@
 
-	# Plot results of the ascertainment bias analyses
-
-	label_age = str_c( "Wilcoxon p=", wilcox_filtered_age )
+	g_age =
+		nodes_contrast %>%
+		ggplot( aes( x=node_age, y=pic, col=Event ) ) + 
+			geom_point( alpha=0.2, size = 0.5 ) + 
+			xlab( "Node age (millions of years)" ) + 
+			ylab( "Phylogenetic independent \ncontrast" ) +
+			ylim( 0, 0.3 ) +
+			xlim( 0, 1000 ) +
+			ggtitle( "(a)" ) +
+			theme_classic()	+
+			guides( colour = guide_legend( override.aes = list( alpha = 1, size = 2 ) ) ) +
+			theme( legend.justification=c( 1, 1 ), legend.position=c( 1, 1 ) )
+
+	label_age = str_c( "Wilcoxon p=", signif( wilcox_filtered_age, 2 ) )
 
@@
 		ggplot( aes( x=var_exp, y=pic, col=Event ) ) + 
-		geom_point( alpha=0.2 ) + 
+		geom_point( alpha=0.2, size = 0.5 ) + 
 		xlab( "Expected variance, \n depends mostly on branch length" ) + 
@@
 
-	label_var = str_c( "Wilcoxon p=", wilcox_filtered_var )
+	label_var = str_c( "Wilcoxon p=", signif ( wilcox_filtered_var, 2 ) )
 

### 2017-02-12 tests/testthat/test_general.R
added testthat framework

@@ -1,0 +1,51 @@
+library( treeio )
+library( tidyverse )
+library( magrittr )
+library( stringr )
+library( ape )
+library( hutan ) # From https://github.com/caseywdunn/hutan
+library( phangorn )
+
+source( "../../functions.R" )
+
+tree_4_text = "(((((((((((((((((ENSTNIP00000010478:0.061424[&&NHX:D=N:G=ENSTNIG00000007663:T=99883],ENSTRUP00000001252:0.083792[&&NHX:D=N:G=ENSTRUG00000000528:T=31033])Tetraodontidae:0.11473[&&NHX:D=N:B=99:T=31031],ENSGACP00000027655:0.102211[&&NHX:D=N:G=ENSGACG00000020915:T=69293])Percomorpha:0.032791[&&NHX:D=N:B=41:T=32485],((ENSXMAP00000006351:0.150553[&&NHX:D=N:G=ENSXMAG00000006333:T=8083],ENSORLP00000016021:0.151884[&&NHX:D=N:G=ENSORLG00000012795:T=8090])Atherinomorpha:0.021245[&&NHX:D=N:B=20:T=32456],ENSONIP00000022433:0.127146[&&NHX:D=N:G=ENSONIG00000017795:T=8128])Percomorpha:0.038431[&&NHX:D=N:B=4:T=32485])Percomorpha:0.076366[&&NHX:DD=Y:B=10:T=32485],ENSGMOP00000017370:0.255986[&&NHX:D=N:G=ENSGMOG00000016187:T=8049])Holacanthopterygii:0.151102[&&NHX:D=N:B=94:T=123370],ENSDARP00000105714:0.237813[&&NHX:D=N:G=ENSDARG00000044567:T=7955])Clupeocephala:0.048578[&&NHX:D=N:B=3:T=186625],ENSLOCP00000011549:0.200397[&&NHX:D=N:G=ENSLOCG00000009457:T=7918])Neopterygii:0[&&NHX:D=N:B=2:T=41665],(((((((ENSGALP00000002175:0.006783[&&NHX:D=N:G=ENSGALG00000001429:T=9031],ENSMGAP00000002621:0.012098[&&NHX:D=N:G=ENSMGAG00000002950:T=9103])Phasianidae:0.021242[&&NHX:D=
[…]

### 2017-02-12 manuscript.rmd
figure scaling

@@
 		species_table,
-		caption = "Number of tips for each species in trees that were used to calculate the independent contrasts."
+		caption = "Number of tips N for each species in trees that were used to calculate the independent contrasts."
 	)
@@
 					geom_tiplab( aes( label=species ), hjust=-0.1 ) + 
-					xlim( NA, 1400 ) + 
+					xlim( 0, 400 ) + 
 					geom_tippoint( aes( size=Tau ) ) +

### 2017-02-23 manuscript.rmd
added shortened url

@@
 	# Plot the contrasts for duplication events and speciation events
+	g_pic_empirical = 
 		nodes_contrast %>%
@@
 
-```
-
-We mapped the Tau values calculated by KMRR [@KryuchkovaMostacci:2016iw] for the Brawand *et al.* 2011 [@Brawand:2011du] datset onto `r n_gene_trees` gene trees parsed from ENSMBL Compara [@Herrero:2016jj].
+	print( g_pic_empirical )
+
+```
+
+We mapped the Tau values calculated by KMRR [@KryuchkovaMostacci:2016iw] for the Brawand *et al.* 2011 [@Brawand:2011du] datset onto `r n_gene_trees` gene trees parsed from ENSEMBL Compara [@Herrero:2016jj].
 These are the same pre-computed trees that the orthology/ paralogy annotations KMRR used are based on.
@@
 
+	g_pairwise_empirical = 
 		pairwise_summary %>% 
@@
 
+	print( g_pairwise_empirical )
+
 ```
@@
 In order to better understand why our phylogenetic analysis supports a different conclusion (*i.e.*, no support for the ortholog conjecture) than the published analysis of KMRR [@KryuchkovaMostacci:2016iw] (*i.e.*, strong support for the ortholog conjecture), we first checked to make sure we could reproduce their result based on pairwise analyses.
-This is important since we are only looking at at a subset of the data they considered, the Brawand *et al.* 2011 [@Brawand:2011du] dataset for gene trees that could be successfully time calibrated.
+This is important since we are only looking at a subset of the data they considered, the Brawand *et al.* 2011 [@Brawand:2011du] dataset for gene trees that could be successfully time calibrated.
 In their Figure 1, they present a higher Tau correlation coefficient between ortholog pairs than paralog ortholog pairs.

### 2017-02-27 manuscript.rmd
consolidated Figures 3-5

@@
 
-```{r show_contrast_result, echo=FALSE, cache=TRUE, message=F, warning=F, fig.height=2, fig.width=4, fig.cap="Phylogenetic reanalysis of the Tau values calculated by KMRR [@KryuchkovaMostacci:2016iw].
-Density plot of the magnitude of phylogenetic independent contrasts for Tau following duplication and speciation events.
-As the contrasts for duplication events are not larger than those of speciation events, the results are not consistent with the predictions of the ortholog conjecture."}
-
-	label_p = str_c( "Wilcoxon p=", signif( wilcox_test_result, 2 ) )
-
-	# Plot the contrasts for duplication events and speciation events
-	g_pic_empirical = 
-		nodes_contrast %>%
-			ggplot( aes( x=pic, y=..density.., col=Event ) ) + 
-			geom_freqpoly( binwidth=.001, position="identity" ) +
-			xlab( "Phylogenetic independent contrasts" ) +
-			ylab( "Density" ) +
-			xlim( 0, 0.075 ) + 
-			theme_classic() + 
-			annotate( "text", x = Inf, y = 50, hjust="right", vjust="top", label=label_p ) +
-			theme( legend.justification=c( 1, 1 ), legend.position=c( 1, 1 ) )
-
-	print( g_pic_empirical )
-
-```
-
 We mapped the Tau values calculated by KMRR [@KryuchkovaMostacci:2016iw] for the Brawand *et al.* 2011 [@Brawand:2011du] datset onto `r n_gene_trees` gene trees parsed from ENSEMBL Compara [@Herrero:2016jj].
@@
 These calibrated trees were used to calculate phylogenetic independent contrasts for `r nodes_contrast %>% filter(D=="Y") %>% nrow()` duplication nodes and `r nodes_contrast %>% filter(D=="N") %>% nrow()` speciation nodes.
-One of these trees is shown in Supplementary Figure 7 to demonstrate the analysis.
+One of these trees is shown in Supplementary Figure 5 to demonstrate the analysis.
 
@@
 
-We did not find increased evolutionary change in expression following duplication events (Figure 3).
+We did not find increased evolutionary change in expression following duplication events (Figure 3b).
 The Wilcoxon rank test does not reject the null hypothesis that the rate of evolution following duplications is the same as or less than the rate following speciation (p value = `r wilcox_test_result`).
@@
 We focused on two possible sources of bias, node depth and branch length.
-We found no evidence that either affected our results (Supplementary Information, Supplementary Figure 8).
+We found no evidence that either affected our results (Supplementary Information, Supplementary Figure 6).
 We also examined the sensitivity of the results to the calibration times applied to speciation events on the gene trees.
@@
 
-```{r pairwise, echo=FALSE, cache=TRUE, message=F, warning=F, fig.height=2, fig.width=5, fig.cap="The pairwise analyses presented in Figure 2a of KMRR [@KryuchkovaMostacci:2016iw] can be reproduced with the subset of data we consider here.
-Each circle indicates the Tau Pearson correlation coefficient for a set of pairwise comparisons annotated with a specific node name of a given age and event type, *i.e.*, whether the divergence at that nodes was due to speciation (giving rise to orthologs) or duplication (giving rise to paralogs).
-Following KMRR Figure 2a, this figure shows only the comparisons that include a human sequence."}
+```{r pairwise, echo=FALSE, cache=TRUE, message=F, warning=F}
 
@@
 
-	g_pairwise_empirical = 
+```
+
+In order to better understand why our phylogenetic analysis supports a different conclusion (*i.e.*, no support for the ortholog conjecture) than the published analysis of KMRR [@KryuchkovaMostacci:2016iw] (*i.e.*, strong support for the ortholog conjecture), we first checked to make sure we could reproduce their result based on pairwise analyses.
+This is important since we are only looking at a subset of the data they considered, the Brawand *et al.* 2011 [@Brawand:2011du] dataset for gene trees that could be successfully time calibrated.
+In their Figure 1, they present a higher Tau correlation coefficient between ortholog pairs than paralog ortholog pairs.
+We find the same here, with correlation coefficients of `r signif( ortholog_r, 2)` for orthologs and `r signif( paralog_r, 2)` for paralogs.
+
+Why is it that pairwise methods and phylogenetic methods lead to opposite conclusions?
+One reason is that multiple pairwise comparisons repeatedly sample the same evolutionary changes and in so doing violate statistical assumptions of independence, whereas phylogenetic comparative methods make multiple independent comparisons across non-overlapping branches of the tree.
+The other reason for the different conclusions is that pairwise comparisons and phylogenetic comparative methods describe different things.
+Pairwise comparisons describe contemporary patterns, while phylogenetic methods infer historical processes [@OMeara:2012gq].
+There need not be a different process of evolution following speciation and duplication for paralogs to be more different than orthologs.
+Any difference could be due to the structure of the gene phylogenies alone.
+If paralogs tend to be more distantly related to each other than orthologs, then there would be more time for differences to accumulate even if the rate of change is the same between the two.
+This is, in fact, the case for these data.
+While the mean distance (*i.e.*, total branch length) between orthologs is `r round( ortholog_distance_mean, 1 )` million years, the mean distance between paralogs is `r round( paralog_distance_mean, 1 )` million years.
+This is because the oldest speciation event is by definition the most recent common ancestor of the species included in the study, but many gene families underwent duplication before this time.
+
+To test the hypothesis that ancient duplications that precede the oldest speciation event (Supplementary Figure 6a) impact the lower correlation of Tau between paralogs than between orthologs, we removed them.
+When we consider only the duplication events the same age or younger than the oldest speciation event (Supplementary Figure 6b), the paralog correlation coefficient increases from `r signif( paralog_r, 2)` to `r signif( paralog_subset_r, 2)`.
+This is much closer to the ortholog correlation of `r signif( ortholog_r, 2) `.
+
+```{r estimate_tau_model, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
+	# Estimate model parameters and add them to the tree objects
+	# Subcalls to fitContinuous() are parallel, so don't use mclapply here
+	# This takes a while, so do it once here and reuse the values as needed
+	gene_trees_calibrated = lapply( gene_trees_calibrated, add_model_parameters )
+```
+
+```{r simulations_null, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
+
+	# Replace the actual Tau values with simulated Tau values, using the parameters 
+	# estimated for each tree
+	gene_trees_sim_null = mclapply( gene_trees_calibrated, sim_tau, mc.cores=cores )
+
+	# Perform contrasts 
+	gene_trees_sim_null %<>% add_pics_to_trees()
+	nodes_sim_null_contrast = gene_trees_sim_null %>% summarize_contrasts()
+
+	# Calculate pairwise comparisons
+	pairwise_sim_null_summary = 
+		mclapply( gene_trees_sim_null, get_pairwise_summary, mc.cores=cores ) %>% 
+		bind_rows()
+
+	# Calculate some summary statistics 
+	ortholog_r_sim_null = 
+		pairwise_sim_null_summary %>% 
+		filter( D=="N" ) %>% 
+		select( tau_a, tau_b ) %>% 
+		cor( ) %>% 
+		.[ 1, 2 ]
+
+	paralog_r_sim_null = 
+		pairwise_sim_null_summary %>% 
+		filter( D=="Y" ) %>% 
+		select( tau_a, tau_b ) %>% 
+		cor( ) %>% 
+		.[ 1, 2 ]
+
+```
+
+```{r simulations_oc, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
+
+	# Replace the actual Tau values with simulated Tau values
+	gene_trees_sim_oc = lapply( gene_trees_calibrated, sim_tau, dup_adjust=dup_adjust )
+
+	# Calculate contrasts 
+	gene_trees_sim_oc %<>% add_pics_to_trees()
+	nodes_sim_oc_contrast = gene_trees_sim_oc %>% summarize_contrasts()
+
+	# Calculate pairwise comparisons 
+	pairwise_sim_oc_summary = 
+		mclapply( gene_trees_sim_oc, get_pairwise_summary, mc.cores=cores ) %>% 
+		bind_rows()
+
+	ortholog_r_sim_oc = 
+		pairwise_sim_oc_summary %>% 
+		filter( D=="N" ) %>% 
+		select( tau_a, tau_b ) %>% 
+		cor( ) %>% 
+		.[ 1, 2 ]
+
+	paralog_r_sim_oc = 
+		pairwise_sim_oc_summary %>% 
+		filter( D=="Y" ) %>% 
+		select( tau_a, tau_b ) %>% 
+		cor( ) %>% 
+		.[ 1, 2 ]
+
+```
+
+```{r pairwise_and_phylo_plots, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.width=6.5, fig.height=6.5, fig.cap="Pairwise (left column - a, c, e) and phylogenetic (right column - b, d, f) phylogenetic analyses of the original data (first row - a, b), data simulated under the null hypothesis (second row - c, d), and data simulated under the ortholog conjecture (third row- e, f).
+In the simulation under the null hypothesis, the rate of evolution following duplication and speciation is the same.
+In the simulation under the ortholog conjecture, the rate of evolution following duplication is greater than following speciation.
+In the pairwise plots, each circle indicates the correlation coefficient of Tau for a set of pairwise comparisons annotated with a specific node name and event type, *i.e.*, whether the divergence at that nodes was due to speciation (giving rise to orthologs) or duplication (giving rise to paralogs).
+The top left pane (a) reproduces the pattern presented in Figure 2a of KMRR [@KryuchkovaMostacci:2016iw], which they took as evidence of the ortholog conjecture.
+The recovery of this same pairwise pattern under both simulations (c, e) indicates that it does not make distinct predictions that can test hyptheses about the evolution of expression.
+The phylogenetic analysis of the original data does not reject the null (b), providing no support for the ortholog conjecture.
+The bottom right panes validate the phylogenetic approach by showing that it does not reject the null when data are simulated under the null (d), but does reject the null when data are simulated under the ortholog conjecture (f)."}
+
+	g_pair_empirical = 
 		pairwise_summary %>% 
-			filter( species_a == "homo_sapiens" | species_b == "homo_sapiens" ) %>% 
 			filter( ! is.na( D ) ) %>%
@@
 				xlab( "Node Age (million years)" ) +
-				ylab( "Correlation Coefficient" )
-
-	print( g_pairwise_empirical )
-
-```
-
-In order to better understand why our phylogenetic analysis supports a different conclusion (*i.e.*, no support for the ortholog conjecture) than the published analysis of KMRR [@KryuchkovaMostacci:2016iw] (*i.e.*, strong support for the ortholog conjecture), we first checked to make sure we could reproduce their result based on pairwise analyses.
-This is important since we are only looking at a subset of the data they considered, the Brawand *et al.* 2011 [@Brawand:2011du] dataset for gene trees that could be successfully time calibrated.
-In their Figure 1, they present a higher Tau correlation coefficient between ortholog pairs than paralog ortholog pairs.
-We find the same here, with correlation coefficients of `r signif( ortholog_r, 2)` for orthologs and `r signif( paralog_r, 2)` for paralogs.
-
-Why is it that pairwise methods and phylogenetic methods lead to opposite conclusions?
-One reason is that multiple pairwise comparisons repeatedly sample the same evolutionary changes and in so doing violate statistical assumptions of independence, whereas phylogenetic comparative methods make multiple independent comparisons across non-overlapping branches of the tree.
-The other reason for the different conclusions is that pairwise comparisons and phylogenetic comparative methods describe different things.
-Pairwise comparisons describe contemporary patterns, while phylogenetic methods infer historical processes [@OMeara:2012gq].
-There need not be a different process of evolution following speciation and duplication for paralogs to be more different than orthologs.
-Any difference could be due to the structure of the gene phylogenies alone.
-If paralogs tend to be more distantly related to each other than orthologs, then there would be more time for differences to accumulate even if the rate of change is the same between the two.
-This is, in fact, the case for these data.
-While the mean distance (*i.e.*, total branch length) between orthologs is `r round( ortholog_distance_mean, 1 )` million years, the mean distance between paralogs is `r round( paralog_distance_mean, 1 )` million years.
-This is because the oldest speciation event is by definition the most recent common ancestor of the species included in the study, but many gene families underwent duplication before this time.
-
-To test the hypothesis that ancient duplications that precede the oldest speciation event (Supplementary Figure 8a) impact the lower correlation of Tau between paralogs than between orthologs, we removed them.
-When we consider only the duplication events the same age or younger than the oldest speciation event (Supplementary Figure 8b), the paralog correlation coefficient increases from `r signif( paralog_r, 2)` to `r signif( paralog_subset_r, 2)`.
-This is much closer to the ortholog correlation of `r signif( ortholog_r, 2) `.
-
-```{r estimate_tau_model, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
-	# Estimate model parameters and add them to the tree objects
-	# Subcalls to fitContinuous() are parallel, so don't use mclapply here
-	# This takes a while, so do it once here and reuse the values as needed
-	gene_trees_calibrated = lapply( gene_trees_calibrated, add_model_parameters )
-```
-
-```{r simulations_null, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
-
-	# Replace the actual Tau values with simulated Tau values, using the parameters 
-	# estimated for each tree
-	gene_trees_sim_null = mclapply( gene_trees_calibrated, sim_tau, mc.cores=cores )
-
-	# Perform contrasts 
-	gene_trees_sim_null %<>% add_pics_to_trees()
-	nodes_sim_null_contrast = gene_trees_sim_null %>% summarize_contrasts()
-
-	# Calculate pairwise comparisons
-	pairwise_sim_null_summary = 
-		mclapply( gene_trees_sim_null, get_pairwise_summary, mc.cores=cores ) %>% 
-		bind_rows()
-
-	# Calculate some summary statistics 
-	ortholog_r_sim_null = 
-		pairwise_sim_null_summary %>% 
-		filter( D=="N" ) %>% 
-		select( tau_a, tau_b ) %>% 
-		cor( ) %>% 
-		.[ 1, 2 ]
-
-	paralog_r_sim_null = 
-		pairwise_sim_null_summary %>% 
-		filter( D=="Y" ) %>% 
-		select( tau_a, tau_b ) %>% 
-		cor( ) %>% 
-		.[ 1, 2 ]
-
-```
-
-```{r simulations_oc, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
-
-	# Replace the actual Tau values with simulated Tau values
-	gene_trees_sim_oc = lapply( gene_trees_calibrated, sim_tau, dup_adjust=dup_adjust )
-
-	# Calculate contrasts 
-	gene_trees_sim_oc %<>% add_pics_to_trees()
-	nodes_sim_oc_contrast = gene_trees_sim_oc %>% summarize_contrasts()
-
-	# Calculate pairwise comparisons 
-	pairwise_sim_oc_summary = 
-		mclapply( gene_trees_sim_oc, get_pairwise_summary, mc.cores=cores ) %>% 
-		bind_rows()
-
-	ortholog_r_sim_oc = 
-		pairwise_sim_oc_summary %>% 
-		filter( D=="N" ) %>% 
-		select( tau_a, tau_b ) %>% 
-		cor( ) %>% 
-		.[ 1, 2 ]
-
-	paralog_r_sim_oc = 
-		pairwise_sim_oc_summary %>% 
-		filter( D=="Y" ) %>% 
-		select( tau_a, tau_b ) %>% 
-		cor( ) %>% 
-		.[ 1, 2 ]
-
-```
-
-```{r simulations_plot, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Analyses of simulated data reveal that pairwise methods produce similar results under both the null model and the ortholog conjecture (OC), while the phylogenetic method provides different results under each model.
-(a) Pairwise analysis of dataset simulated under the null model that there is no difference in the rate of expression evolution following duplication and speciation.
-(b) Phylogenetic analysis of data simulated under the null model.
-(c) Pairwise analysis of dataset simulated under the ortholog conjecture, in which there is a higher rate of expression evolution following duplication than speciation.
-This plot is very similar to the corresponding plot for the null model.
-(d) Phylogenetic analysis of data simulated under the ortholog conjecture.
-This plot is distinct from that for the null model in that phylogenetic independent contrasts for duplication events are significantly larger than after speciation events."}
-
-	label_p_sim_null = str_c( "Wilcoxon p=", signif( wilcox_oc( nodes_sim_null_contrast ), 2 ) )
-
-	g_phylo_null = 
-		nodes_sim_null_contrast %>% 
+				ylab( "Correlation Coefficient" ) +
+				ylim( 0, 1 ) +
+				ggtitle( "(a) Empirical, pairwise" ) +
+				theme(legend.position="none")
+
+	label_p = str_c( "Wilcoxon p=", signif( wilcox_test_result, 2 ) )
+	g_phylo_empirical = 
+		nodes_contrast %>%
 			ggplot( aes( x=pic, y=..density.., col=Event ) ) + 
-			geom_freqpoly( binwidth=.0025, position="identity" ) +
+			geom_freqpoly( binwidth=.001, position="identity" ) +
 			xlab( "Phylogenetic Independent Contrast" ) +
@@
 			ylim( 0, 75 ) +
-			annotate( "text", x = Inf, y = 20, hjust="right", vjust="top", label=label_p_sim_null ) +
 			theme_classic() + 
-			ggtitle( "(b) Null, phylogenetic" ) +
+			ggtitle( "(b) Empirical, phylogenetic" ) +
+			annotate( "text", x = Inf, y = 20, hjust="right", vjust="top", label=label_p ) +
 			theme( legend.justification=c( 1, 1 ), legend.position=c( 1, 1 ) )
@@
 			ylim( 0, 1 ) +
-			ggtitle( "(a) Null, pairwise" ) +
-			theme(legend.position="none")
-
-	label_p_sim_oc = str_c( "Wilcoxon p=", signif( wilcox_oc( nodes_sim_oc_contrast ), 2 ) )
-
-	g_phylo_oc = 
-	nodes_sim_oc_contrast %>% 
+			ggtitle( "(c) Null simulation, pairwise" ) +
+			theme( legend.position="none", axis.title.x=element_blank(), axis.title.y=element_blank() )
+
+	label_p_sim_null = str_c( "Wilcoxon p=", signif( wilcox_oc( nodes_sim_null_contrast ), 2 ) )
+	g_phylo_null = 
+		nodes_sim_null_contrast %>% 
 		ggplot( aes( x=pic, y=..density.., col=Event ) ) + 
-		geom_freqpoly( binwidth=.0025, position="identity" ) +
+			geom_freqpoly( binwidth=.001, position="identity" ) +
 			xlab( "Phylogenetic Independent Contrast" ) +
@@
 			ylim( 0, 75 ) +
-		annotate( "text", x = Inf, y = 30, hjust="right", vjust="top", label=label_p_sim_oc ) +
+			annotate( "text", x = Inf, y = 20, hjust="right", vjust="top", label=label_p_sim_null ) +
 			theme_classic() + 
-		ggtitle( "(d) OC, phylogenetic" ) +
-		theme(legend.position="none")
+			ggtitle( "(d) Null simulation, phylogenetic" ) +
+			theme( legend.position="none", axis.title.x=element_blank(), axis.title.y=element_blank() )
 
@@
 			ylim( 0, 1 ) +
-			ggtitle( "(c) OC, pairwise" ) +
-			theme(legend.position="none")
+			ggtitle( "(e) OC simulation, pairwise" ) +
+			theme( legend.position="none", axis.title.x=element_blank(), axis.title.y=element_blank() )
+
+	label_p_sim_oc = str_c( "Wilcoxon p=", signif( wilcox_oc( nodes_sim_oc_contrast ), 2 ) )
+	g_phylo_oc = 
+	nodes_sim_oc_contrast %>% 
+	ggplot( aes( x=pic, y=..density.., col=Event ) ) + 
+		geom_freqpoly( binwidth=.001, position="identity" ) +
+		xlab( "Phylogenetic Independent Contrast" ) +
+		ylab( "Density" ) +
+		xlim( 0, 0.05 ) + 
+		ylim( 0, 75 ) +
+		annotate( "text", x = Inf, y = 30, hjust="right", vjust="top", label=label_p_sim_oc ) +
+		theme_classic() + 
+		ggtitle( "(f) OC simulation, phylogenetic" ) +
+		theme( legend.position="none", axis.title.x=element_blank(), axis.title.y=element_blank() )
 
 	grid.arrange(
+		g_pair_empirical,
+		g_phylo_empirical,
 		g_pair_null,
@@
 They found that across the investigated range of node ages, ortholog pairs have higher Tau correlation than paralogs.
-We confirmed that we can replicate this result (Figure 4).
+We confirmed that we can replicate this result (Figure 3a).
 There are several difficulties with interpreting this plot, though.
@@
 
-To better understand this plot (Figure 4), we performed a couple simulations of Tau on the calibrated gene trees and then regenerated the figures.
+To better understand this plot (Figure 3b), we performed a couple simulations of Tau on the calibrated gene trees and then regenerated the figures.
 We did not modify the gene tree topologies or their inferred histories of duplication and loss.
 First, we simulated the evolution of Tau under the null model that it evolves at the same rate following duplication and speciation events.
-Under the null model, the plot of correlation coefficient to node age is very similar as for the observed data (Figure 5a).
+Under the null model (Figure 3b), the plot of correlation coefficient to node age is very similar as for the observed data (Figure 3a).
 As in the original study, there is higher correlation coefficient across orthologs (`r signif( ortholog_r_sim_null, 2 )`) than paralogs (`r signif( paralog_r_sim_null, 2 )`) when not considering node age.
-Phylogenetic analysis of the data simulated under the null hypothesis (Figure 5b) do not reject the null hypothesis (Wilcoxon p = $`r wilcox_oc( nodes_sim_null_contrast )`$), as expected.
+Phylogenetic analysis of the data simulated under the null hypothesis (Figure 3d) do not reject the null hypothesis (Wilcoxon p = $`r wilcox_oc( nodes_sim_null_contrast )`$), as expected.
 
 We next simulated the evolution of Tau under the ortholog conjecture, where the rate of evolution of Tau following duplication was `r dup_adjust` fold the rate following speciation.
-The pairwise results of this heterogeneous model are nearly indistinguishable from the null model (Figure 5c), and also have a higher correlation coefficient for orthologs (`r signif( ortholog_r_sim_oc, 2 )`) than paralogs (`r signif( paralog_r_sim_oc, 2 )`).
-The phylogenetic analysis of the ortholog conjecture simulation (Figure 5d) does reject the null hypothesis (Wilcoxon p = $`r signif( wilcox_oc( nodes_sim_oc_contrast ), 2 )`$).
+The pairwise results of this heterogeneous model (Figure 3e) are nearly indistinguishable from the results under the null model (Figure 3c), and also have a higher correlation coefficient for orthologs (`r signif( ortholog_r_sim_oc, 2 )`) than paralogs (`r signif( paralog_r_sim_oc, 2 )`).
+The phylogenetic analysis of the ortholog conjecture simulation (Figure 3f) does reject the null hypothesis (Wilcoxon p = $`r signif( wilcox_oc( nodes_sim_oc_contrast ), 2 )`$).
 
@@
 From this they conclude that following duplication there is a common trajectory of evolutionary change, where one paralog evolves to have lower expression and also become more tissue specific.
-There is, however, a negative relationship between Tau and maximum expression across all genes (Supplementary Figure 9).
+There is, however, a negative relationship between Tau and maximum expression across all genes (Supplementary Figure 7).
 A simpler explanation for this plot is that it reflects the global tendency for Tau to decrease with maximum expression.
@@
 They then calculated a similarity score for each temporal phase for each species pair based on the fraction of genes that exhibited the same patterns in each species.
-The distributions of similarity scores are plotted in their [Figure 4d](http://www.nature.com/nature/journal/v531/n7596/fig_tab/nature16994_F4.html), and their Kolmogorov–Smirnov (KS) tests indicated that the early distribution and late distribution were each significantly different from mid distribution (P < $10^{-6}$ and P < $10^{-12}$, respectively).
+The distributions of similarity scores are plotted in their Figure 4d, and their Kolmogorov–Smirnov (KS) tests indicated that the early distribution and late distribution were each significantly different from mid distribution (P < $10^{-6}$ and P < $10^{-12}$, respectively).
 This is the support they presented for the inverse hourglass model.
@@
 Once we addressed the issues above with the implementation of pairwise comparisons, we were able to explore more general issues that can be a problem when making multiple pairwise comparisons between species.
-We found that all five of the lowest values in the mid phase distribution (Figure 6a) are for pairwise comparisons that include the ctenophore (comb jelly).
-When the nine pairwise comparisons that include the ctenophore are removed, there is no significant difference between the early phase and mid phase distributions (p = 0.14 for the early-mid comparison and p < $10^{-5}$ for the late-mid comparison) and no support for the inverse hourglass (Figure 6b).
+We found that all five of the lowest values in the mid phase distribution (Figure 4a) are for pairwise comparisons that include the ctenophore (comb jelly).
+When the nine pairwise comparisons that include the ctenophore are removed, there is no significant difference between the early phase and mid phase distributions (p = 0.14 for the early-mid comparison and p < $10^{-5}$ for the late-mid comparison) and no support for the inverse hourglass (Figure 4b).
 This highlights a well understood property of pairwise comparisons across species [@Felsenstein:1985ua; @Ackerly:1999iu]: evolutionary changes along a given branch, like those along the ctenophore branch, impact each of the multiple pairwise comparisons that includes that branch.
@@
 Levin *et al.* helpfully provided data and clarification on methods.
-We obtained the matrix of pairwise scores that underlies their [Figure 4d](http://www.nature.com/nature/journal/v531/n7596/fig_tab/nature16994_F4.html) and confirmed we could reproduce their published results.
+We obtained the matrix of pairwise scores that underlies their Figure 4d and confirmed we could reproduce their published results.
 We then removed duplicate rows, applied the Wilcoxon test in place of the Kolmogorov-Smirnov test, and identified ctenophores as overrepresented among the low outliers in the mid-developmental transition column.
@@
 
-While the age of speciation nodes is constrained, duplication nodes can be much older and therefore have a wider range of ages (Supplementary Figure 8a).
+While the age of speciation nodes is constrained, duplication nodes can be much older and therefore have a wider range of ages (Supplementary Figure 6a).
 This is because many gene duplication events are older than the most recent common ancestor of the species in the study.
@@
 To test for such an effect, we remove duplication nodes that are older than the oldest speciation node.
-Our results are unchanged and this reduced dataset does not reject the null hypothesis that the rate of evolution following duplications is the same as or less than the rate following speciation (Supplementary Figure 8b).
+Our results are unchanged and this reduced dataset does not reject the null hypothesis that the rate of evolution following duplications is the same as or less than the rate following speciation (Supplementary Figure 6b).
 
@@
 This is because the same difference across a shorter total branch length indicates a greater rate of evolutionary change.
-The expected variance of contrasts for speciation nodes is constrained by the branch lengths on the species tree, but the expected variance of contrasts for duplications has a much wider range (Figure 8c).
+The expected variance of contrasts for speciation nodes is constrained by the branch lengths on the species tree, but the expected variance of contrasts for duplications has a much wider range (Figure 6c).
 This could lead to biases if the lengths of branches that descend from duplication nodes tend to be overestimated.
 We therefore examined only the contrasts that fell within in the range of expected variance seen for speciation contrasts, excluding duplication contrasts that fall outside of this range.
-This reanalysis does not reject the null hypothesis either (Figure 8d), indicating that branch length bias is not responsible for the result.
+This reanalysis does not reject the null hypothesis either (Figure 6d), indicating that branch length bias is not responsible for the result.
 
@@
 
-There is a negative relationship between miximum expression and Tau for each gene (Supplementary Figure 9).
-
 ```{r Tau_v_max, cache=TRUE, echo=FALSE, warning=FALSE, dev="png", dpi=300, message=FALSE, fig.cap="Genes with higher maximum observed expression tend to have lower Tau.
@@
 	nodes_tips$max_expression = max_expression
+
+	tau_v_max_lm = lm(nodes_tips$Tau ~ nodes_tips$max_expression) %>% summary()
 
@@
 
+There is a negative relationship between miximum expression and Tau for each gene (Supplementary Figure 7, $r^2=`r signif( tau_v_max_lm$r.squared, 3 )`$, $slope=`r signif( tau_v_max_lm$coefficients[2,1], 3 )`$, $p=`r signif( tau_v_max_lm$coefficients[2,4], 3 )`$).
+
 \pagebreak

### 2017-03-24 manuscript.rmd
started ti implement density difference plots

@@
 ```{r preliminaries, echo=FALSE, message=F, warning=F}
-	library( treeio )
+	library( treeio ) # https://bioconductor.org/packages/devel/bioc/html/treeio.html
+	library( ggtree ) # https://bioconductor.org/packages/release/bioc/html/ggtree.html
+	library( hutan ) # https://github.com/caseywdunn/hutan
+
+	# The rest can be installed from CRAN
 	library( tidyverse )
@@
 	library( parallel )
-	library( ggtree )
 	library( ape )
 	library( digest )
-	library( hutan ) # From https://github.com/caseywdunn/hutan
 	library( gridExtra )

### 2017-03-24 functions.R

@@ -514,3 +514,58 @@ get_pairwise_summary = function ( nhx ) {
 	return( D )
 	
 }
+
+#' Create a ggplot summary of independent contrast values
+#' 
+#' @param contrasts The contrasts to be displayed 
+#' @param mode The type of plot to produce 
+#' 	
+#' @return A ggplot object
+#' @export
+make_contrast_plot = function( nodes_contrast, mode="diff" ){
+	
+	wilcox_test_result = wilcox_oc( nodes_contrast )
+	label_p = str_c( "Wilcoxon p=", signif( wilcox_test_result, 2 ) )
+	
+	if( mode == "freqpoly" ){
+		ggcontrasts = nodes_contrast %>%
+			ggplot( aes( x=pic, y=..density.., col=Event ) ) + 
+				geom_freqpoly( binwidth=.001, position="identity" ) +
+				xlab( "Tau Phylogenetic Contrast" ) +
+				ylab( "Density" ) +
+				xlim( 0, 0.05 ) + 
+				ylim( 0, 75 ) +
+				theme_classic() + 
+				annotate( "text", x = Inf, y = 20, hjust="right", vjust="top", label=label_p ) +
+				theme( legend.position="none" )
+	} else if( mode == "boxplot" ) {
+		ggcontrasts = nodes_contrast %>%
+			filter( pic < 0.05 ) %>%
+			ggplot( aes(factor(Event), pic ) ) +
+			geom_boxplot( notch = TRUE ) +
+			coord_flip()
+	} else if( mode == "diff" ){
+		binwidth = 0.001
+		
+		ggcontrasts = 
+			nodes_contrast %>% 
+			group_by(Event) %>% 
+			# calculate densities for each group over same range; store in list column
+			summarise(d = list(density(pic, from = 0, to = 0.05))) %>% 
+			# make a new data.frame from two density objects
+			do(data.frame(x = .$d[[1]]$x,    # grab one set of x values (which are the same)
+										y = .$d[[1]]$y - .$d[[2]]$y)) %>%    # and subtract the y values
+			ggplot(aes(x, y)) +    # now plot
+				geom_line() +
+				xlab( "Tau Phylogenetic Contrast" ) +
+				ylab( "Density Difference" ) +
+				xlim( 0, 0.05 ) + 
+				ylim( -30, 30 ) +
+				geom_hline( yintercept = 0 ) +
+				theme_classic()
+		
+	}
+	
+	ggcontrasts
+	
+}

### 2017-03-28 manuscript.rmd
changed density plots to density difference plots, updated all dependencies to most recent versions

@@
 ```{r preliminaries, echo=FALSE, message=F, warning=F}
-	library( treeio ) # https://bioconductor.org/packages/devel/bioc/html/treeio.html
-	library( ggtree ) # https://bioconductor.org/packages/release/bioc/html/ggtree.html
-	library( hutan ) # https://github.com/caseywdunn/hutan
+
+	# The following should be installed from github with the specified 
+	# devtools command. You will need to install devtools first.
+	library( treeio ) # devtools::install_github("GuangchuangYu/treeio")
+	library( ggtree ) # devtools::install_github("GuangchuangYu/ggtree")
+	library( hutan ) # devtools::install_github("caseywdunn/hutan")
 
@@
 
-```{r expectation_plots, echo=FALSE, cache=TRUE, message=F, warning=F, fig.height=2, fig.cap="(a) Under the null hypothesis that there is no difference in the rate of evolution after duplication or speciation events, the distribution of phylogenetic independent contrasts would be drawn from the same distribution across both types of nodes.
-(b) Under the ortholog conjecture, contrasts across duplication nodes would tend to be larger than contrasts across speciation nodes."}
-
-	g_expectation_null =
-		sim_expectation() %>%
-		ggplot( aes( x=pic, y=..density.., col=Event ) ) + 
-			geom_freqpoly( binwidth=0.2, position="identity" ) +
-			xlab( "Phylogenetic independent contrasts" ) +
-			ylab( "Density" ) +
-			xlim( 0, 6 ) + 
-			theme_classic() +
-			theme( 
-				axis.text.x=element_blank(), 
-				axis.ticks.x=element_blank(), 
-				axis.text.y=element_blank(), 
-				axis.ticks.y=element_blank(),
-				legend.justification=c( 1, 1 ), 
-				legend.position=c( 1, 1 )
-			) +
-			ggtitle( "(a) Null expectation" )
-
-	g_expectation_oc =
-		sim_expectation( scale=2 ) %>%
-		ggplot( aes( x=pic, y=..density.., col=Event ) ) + 
-			geom_freqpoly( binwidth=0.2, position="identity" ) +
-			xlab( "Phylogenetic independent contrasts" ) +
-			ylab( "Density" ) +
-			xlim( 0, 6 ) + 
-			theme_classic() +
-			theme( 
-				axis.text.x=element_blank(), 
-				axis.ticks.x=element_blank(), 
-				axis.text.y=element_blank(), 
-				axis.ticks.y=element_blank(),
-				legend.position="none"
-			) +
-			ggtitle( "(b) Ortholog Conjecture \nexpectation" )
-
-	grid.arrange(
-			g_expectation_null,
-			g_expectation_oc,
-			ncol = 2
-		)
-
-```
-
 #### Original pairwise test of the ortholog conjecture
@@
 These calibrated trees were used to calculate phylogenetic independent contrasts for `r nodes_contrast %>% filter(D=="Y") %>% nrow()` duplication nodes and `r nodes_contrast %>% filter(D=="N") %>% nrow()` speciation nodes.
-One of these trees is shown in Supplementary Figure 5 to demonstrate the analysis.
+One of these trees is shown in Supplementary Figure 4 to demonstrate the analysis.
 
@@
 A suitable null hypothesis in this case is that there is no difference in the evolution of expression following speciation or duplication events [@Studer:2009bm].
-Under this hypothesis, we would predict that contrasts across speciation nodes and duplication nodes are drawn from the same distribution (Figure 2a).
-Under the alternative hypothesis specified by the ortholog conjecture, that there is a higher rate of change following duplication events than speciation events, we would expect to see the distribution of duplication contrasts shifted to higher values relative to the speciation contrasts (Figure 2b).
-
-We did not find increased evolutionary change in expression following duplication events (Figure 3b).
+Under this hypothesis, we would predict that contrasts across speciation nodes and duplication nodes are drawn from the same distribution.
+Under the alternative hypothesis specified by the ortholog conjecture, that there is a higher rate of change following duplication events than speciation events, we would expect to see the distribution of duplication contrasts shifted to higher values relative to the speciation contrasts.
+
+We did not find increased evolutionary change in expression following duplication events (Figure 2b).
 The Wilcoxon rank test does not reject the null hypothesis that the rate of evolution following duplications is the same as or less than the rate following speciation (p value = `r wilcox_test_result`).
@@
 We focused on two possible sources of bias, node depth and branch length.
-We found no evidence that either affected our results (Supplementary Information, Supplementary Figure 6).
+We found no evidence that either affected our results (Supplementary Information, Supplementary Figure 4).
 We also examined the sensitivity of the results to the calibration times applied to speciation events on the gene trees.
@@
 
-To test the hypothesis that ancient duplications that precede the oldest speciation event (Supplementary Figure 6a) impact the lower correlation of Tau between paralogs than between orthologs, we removed them.
-When we consider only the duplication events the same age or younger than the oldest speciation event (Supplementary Figure 6b), the paralog correlation coefficient increases from `r signif( paralog_r, 2)` to `r signif( paralog_subset_r, 2)`.
+To test the hypothesis that ancient duplications that precede the oldest speciation event (Supplementary Figure 5a) impact the lower correlation of Tau between paralogs than between orthologs, we removed them.
+When we consider only the duplication events the same age or younger than the oldest speciation event (Supplementary Figure 5b), the paralog correlation coefficient increases from `r signif( paralog_r, 2)` to `r signif( paralog_subset_r, 2)`.
 This is much closer to the ortholog correlation of `r signif( ortholog_r, 2) `.
@@
 
-```{r pairwise_and_phylo_plots, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.width=6.5, fig.height=6.5, fig.cap="Pairwise (left column - a, c, e) and phylogenetic (right column - b, d, f) phylogenetic analyses of the original data (first row - a, b), data simulated under the null hypothesis (second row - c, d), and data simulated under the ortholog conjecture (third row- e, f).
-In the simulation under the null hypothesis, the rate of evolution following duplication and speciation is the same.
-In the simulation under the ortholog conjecture, the rate of evolution following duplication is greater than following speciation.
-In the pairwise plots, each circle indicates the correlation coefficient of Tau for a set of pairwise comparisons annotated with a specific node name and event type, *i.e.*, whether the divergence at that nodes was due to speciation (giving rise to orthologs) or duplication (giving rise to paralogs).
+```{r pairwise_and_phylo_plots, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.width=6.5, fig.height=6.5, fig.cap="Pairwise (left column - a, c, e) and phylogenetic (right column - b, d, f) analyses of the original data (first row - a, b), data simulated under the null hypothesis (second row - c, d), and data simulated under the ortholog conjecture (third row- e, f).
+In the pairwise plots, each circle indicates the correlation coefficient of Tau for a set of pairwise comparisons annotated with a specific node name (e.g, Primates) and event type (speciation or duplication, giving rise respectively to orthologs and paralogs).
+The phylogenetic plots show the difference between the density distributions for Tau phylogenetic contrasts for speciation and duplication events, where a value above 0 indicates an excess of speciation contrasts in the indicated interval.
+A horizontal line at 0 would indicate that the density distributions are identical.
 The top left pane (a) reproduces the pattern presented in Figure 2a of KMRR [@KryuchkovaMostacci:2016iw], which they took as evidence of the ortholog conjecture.
 The recovery of this same pairwise pattern under both simulations (c, e) indicates that it does not make distinct predictions that can test hyptheses about the evolution of expression.
-The phylogenetic analysis of the original data does not reject the null (b), providing no support for the ortholog conjecture.
+The phylogenetic analysis of the original data (b) does not show an excess of larger contrasts for duplication events and does not reject the null hypothesis, providing no support for the ortholog conjecture.
 The bottom right panes validate the phylogenetic approach by showing that it does not reject the null when data are simulated under the null (d), but does reject the null when data are simulated under the ortholog conjecture (f)."}
@@
 				ggtitle( "(a) Empirical, pairwise" ) +
-				theme(legend.position="none")
-
-	label_p = str_c( "Wilcoxon p=", signif( wilcox_test_result, 2 ) )
-	g_phylo_empirical = 
-		nodes_contrast %>%
-			ggplot( aes( x=pic, y=..density.., col=Event ) ) + 
-			geom_freqpoly( binwidth=.001, position="identity" ) +
-			xlab( "Phylogenetic Independent Contrast" ) +
-			ylab( "Density" ) +
-			xlim( 0, 0.05 ) + 
-			ylim( 0, 75 ) +
-			theme_classic() + 
-			ggtitle( "(b) Empirical, phylogenetic" ) +
-			annotate( "text", x = Inf, y = 20, hjust="right", vjust="top", label=label_p ) +
-			theme( legend.justification=c( 1, 1 ), legend.position=c( 1, 1 ) )
+				guides( colour = guide_legend( override.aes = list( shape = 15 ) ) ) +
+				theme( 
+				legend.justification=c( 1, 1 ), 
+				legend.position=c( 1, 1 ), 
+				legend.title=element_blank() )
 
@@
 			ggtitle( "(c) Null simulation, pairwise" ) +
-			theme( legend.position="none", axis.title.x=element_blank(), axis.title.y=element_blank() )
-
-	label_p_sim_null = str_c( "Wilcoxon p=", signif( wilcox_oc( nodes_sim_null_contrast ), 2 ) )
-	g_phylo_null = 
-		nodes_sim_null_contrast %>% 
-		ggplot( aes( x=pic, y=..density.., col=Event ) ) + 
-			geom_freqpoly( binwidth=.001, position="identity" ) +
-			xlab( "Phylogenetic Independent Contrast" ) +
-			ylab( "Density" ) +
-			xlim( 0, 0.05 ) + 
-			ylim( 0, 75 ) +
-			annotate( "text", x = Inf, y = 20, hjust="right", vjust="top", label=label_p_sim_null ) +
-			theme_classic() + 
-			ggtitle( "(d) Null simulation, phylogenetic" ) +
-			theme( legend.position="none", axis.title.x=element_blank(), axis.title.y=element_blank() )
+			theme( legend.position="none" )
 
@@
 			ggtitle( "(e) OC simulation, pairwise" ) +
-			theme( legend.position="none", axis.title.x=element_blank(), axis.title.y=element_blank() )
-
-	label_p_sim_oc = str_c( "Wilcoxon p=", signif( wilcox_oc( nodes_sim_oc_contrast ), 2 ) )
-	g_phylo_oc = 
-	nodes_sim_oc_contrast %>% 
-	ggplot( aes( x=pic, y=..density.., col=Event ) ) + 
-		geom_freqpoly( binwidth=.001, position="identity" ) +
-		xlab( "Phylogenetic Independent Contrast" ) +
-		ylab( "Density" ) +
-		xlim( 0, 0.05 ) + 
-		ylim( 0, 75 ) +
-		annotate( "text", x = Inf, y = 30, hjust="right", vjust="top", label=label_p_sim_oc ) +
-		theme_classic() + 
-		ggtitle( "(f) OC simulation, phylogenetic" ) +
-		theme( legend.position="none", axis.title.x=element_blank(), axis.title.y=element_blank() )
+			theme( legend.position="none" )
 
@@
 		g_pair_empirical,
-		g_phylo_empirical,
-		g_pair_null,
-		g_phylo_null,
-		g_pair_oc,
-		g_phylo_oc,
+		make_contrast_plot(nodes_contrast),
+		# For the rest, put spaces in labels so you can't see them but alignment is 
+		# same as those with labels
+		g_pair_null + 
+			xlab( "  " ) + 
+			ylab( "  " ), 
+		make_contrast_plot(nodes_sim_null_contrast) + 
+			xlab( "  " ) + 
+			ylab( "  " ) + 
+			ggtitle( "(d) Null simulation, phylogenetic" ),
+		g_pair_oc + 
+			xlab( "  " ) + 
+			ylab( "  " ),
+		make_contrast_plot(nodes_sim_oc_contrast) + 
+			xlab( "  " ) + 
+			ylab( "  " ) + 
+			ggtitle( "(f) OC simulation, phylogenetic" ),
 		ncol = 2
@@
 They found that across the investigated range of node ages, ortholog pairs have higher Tau correlation than paralogs.
-We confirmed that we can replicate this result (Figure 3a).
+We confirmed that we can replicate this result (Figure 2a).
 There are several difficulties with interpreting this plot, though.
@@
 
-To better understand this plot (Figure 3b), we performed a couple simulations of Tau on the calibrated gene trees and then regenerated the figures.
+To better understand this plot (Figure 2a), we performed a couple simulations of Tau on the calibrated gene trees and then regenerated the figures.
 We did not modify the gene tree topologies or their inferred histories of duplication and loss.
 First, we simulated the evolution of Tau under the null model that it evolves at the same rate following duplication and speciation events.
-Under the null model (Figure 3b), the plot of correlation coefficient to node age is very similar as for the observed data (Figure 3a).
+Under the null model, the plot of correlation coefficient to node age is very similar as for the observed data (Figure 2c).
 As in the original study, there is higher correlation coefficient across orthologs (`r signif( ortholog_r_sim_null, 2 )`) than paralogs (`r signif( paralog_r_sim_null, 2 )`) when not considering node age.
-Phylogenetic analysis of the data simulated under the null hypothesis (Figure 3d) do not reject the null hypothesis (Wilcoxon p = $`r wilcox_oc( nodes_sim_null_contrast )`$), as expected.
+Phylogenetic analysis of the data simulated under the null hypothesis (Figure 2d) do not reject the null hypothesis (Wilcoxon p = $`r wilcox_oc( nodes_sim_null_contrast )`$), as expected.
 
 We next simulated the evolution of Tau under the ortholog conjecture, where the rate of evolution of Tau following duplication was `r dup_adjust` fold the rate following speciation.
-The pairwise results of this heterogeneous model (Figure 3e) are nearly indistinguishable from the results under the null model (Figure 3c), and also have a higher correlation coefficient for orthologs (`r signif( ortholog_r_sim_oc, 2 )`) than paralogs (`r signif( paralog_r_sim_oc, 2 )`).
-The phylogenetic analysis of the ortholog conjecture simulation (Figure 3f) does reject the null hypothesis (Wilcoxon p = $`r signif( wilcox_oc( nodes_sim_oc_contrast ), 2 )`$).
+The pairwise results of this heterogeneous model (Figure 2e) are nearly indistinguishable from the results under the null model (Figure 2c), and also have a higher correlation coefficient for orthologs (`r signif( ortholog_r_sim_oc, 2 )`) than paralogs (`r signif( paralog_r_sim_oc, 2 )`).
+The phylogenetic analysis of the ortholog conjecture simulation (Figure 2f) does reject the null hypothesis (Wilcoxon p = $`r signif( wilcox_oc( nodes_sim_oc_contrast ), 2 )`$).
 
@@
 From this they conclude that following duplication there is a common trajectory of evolutionary change, where one paralog evolves to have lower expression and also become more tissue specific.
-There is, however, a negative relationship between Tau and maximum expression across all genes (Supplementary Figure 7).
+There is, however, a negative relationship between Tau and maximum expression across all genes (Supplementary Figure 6).
 A simpler explanation for this plot is that it reflects the global tendency for Tau to decrease with maximum expression.
@@
 Once we addressed the issues above with the implementation of pairwise comparisons, we were able to explore more general issues that can be a problem when making multiple pairwise comparisons between species.
-We found that all five of the lowest values in the mid phase distribution (Figure 4a) are for pairwise comparisons that include the ctenophore (comb jelly).
-When the nine pairwise comparisons that include the ctenophore are removed, there is no significant difference between the early phase and mid phase distributions (p = 0.14 for the early-mid comparison and p < $10^{-5}$ for the late-mid comparison) and no support for the inverse hourglass (Figure 4b).
+We found that all five of the lowest values in the mid phase distribution (Figure 3a) are for pairwise comparisons that include the ctenophore (comb jelly).
+When the nine pairwise comparisons that include the ctenophore are removed, there is no significant difference between the early phase and mid phase distributions (p = 0.14 for the early-mid comparison and p < $10^{-5}$ for the late-mid comparison) and no support for the inverse hourglass (Figure 3b).
 This highlights a well understood property of pairwise comparisons across species [@Felsenstein:1985ua; @Ackerly:1999iu]: evolutionary changes along a given branch, like those along the ctenophore branch, impact each of the multiple pairwise comparisons that includes that branch.
@@
 Thanks to Steve Haddock, Alex Damian Serrano, August Guang, Bruno Vellutini, and Zack Lewis for feedback on the manuscript.
+Marc Robinson-Rechavi provided very helpful information about the KMRR analyses and valuable suggestions for our manuscript.
 CWD's contribution was supported by the National Science Foundation (DEB-1256695 and the Waterman Award).
@@
 
-While the age of speciation nodes is constrained, duplication nodes can be much older and therefore have a wider range of ages (Supplementary Figure 6a).
+While the age of speciation nodes is constrained, duplication nodes can be much older and therefore have a wider range of ages (Supplementary Figure 5a).
 This is because many gene duplication events are older than the most recent common ancestor of the species in the study.
@@
 To test for such an effect, we remove duplication nodes that are older than the oldest speciation node.
-Our results are unchanged and this reduced dataset does not reject the null hypothesis that the rate of evolution following duplications is the same as or less than the rate following speciation (Supplementary Figure 6b).
+Our results are unchanged and this reduced dataset does not reject the null hypothesis that the rate of evolution following duplications is the same as or less than the rate following speciation (Supplementary Figure 5b).
 
@@
 This is because the same difference across a shorter total branch length indicates a greater rate of evolutionary change.
-The expected variance of contrasts for speciation nodes is constrained by the branch lengths on the species tree, but the expected variance of contrasts for duplications has a much wider range (Figure 6c).
+The expected variance of contrasts for speciation nodes is constrained by the branch lengths on the species tree, but the expected variance of contrasts for duplications has a much wider range (Figure 5c).
 This could lead to biases if the lengths of branches that descend from duplication nodes tend to be overestimated.
 We therefore examined only the contrasts that fell within in the range of expected variance seen for speciation contrasts, excluding duplication contrasts that fall outside of this range.
-This reanalysis does not reject the null hypothesis either (Figure 6d), indicating that branch length bias is not responsible for the result.
+This reanalysis does not reject the null hypothesis either (Figure 5d), indicating that branch length bias is not responsible for the result.
 
@@
 Speciation nodes are calibrated to particular times, whereas duplication nodes have a wider range.
-(b) Density plot of contrasts for only the nodes that have an age less than or equal to the maximum age of speciation nodes.
+(b) Difference between density distributions of contrasts for only the nodes that have an age less than or equal to the maximum age of speciation nodes.
 (c) Magnitude of independent contrasts plotted against expected variance, which is largely determined by branch lengths.
 Contrasts for speciation nodes have a narrower range of expected variance than do contrasts for duplication nodes.
-(d) Density plot of contrasts for only the nodes that have expected variance within the range of contrasts across speciation nodes."}
+(d) Difference between density distributions of contrasts for only the nodes that have expected variance within the range of contrasts across speciation nodes."}
 
@@
 		g_age,
-		g_density_filtered_age,
+		make_contrast_plot( nodes_contrast_filtered_age ),
 		g_var_exp,
-		g_density_filtered_var,
+		make_contrast_plot( nodes_contrast_filtered_var ),
 		ncol = 2
@@
 
-There is a negative relationship between miximum expression and Tau for each gene (Supplementary Figure 7, $r^2=`r signif( tau_v_max_lm$r.squared, 3 )`$, $slope=`r signif( tau_v_max_lm$coefficients[2,1], 3 )`$, $p=`r signif( tau_v_max_lm$coefficients[2,4], 3 )`$).
+There is a negative relationship between miximum expression and Tau for each gene (Supplementary Figure 6, $r^2=`r signif( tau_v_max_lm$r.squared, 3 )`$, $slope=`r signif( tau_v_max_lm$coefficients[2,1], 3 )`$, $p=`r signif( tau_v_max_lm$coefficients[2,4], 3 )`$).
 

### 2017-03-28 functions.R

@@ -547,22 +547,31 @@ make_contrast_plot = function( nodes_contrast, mode="diff" ){
 	} else if( mode == "diff" ){
 		binwidth = 0.001
 		
-		ggcontrasts = 
+		D = 
 			nodes_contrast %>% 
 			group_by(Event) %>% 
 			# calculate densities for each group over same range; store in list column
 			summarise(d = list(density(pic, from = 0, to = 0.05))) %>% 
 			# make a new data.frame from two density objects
 			do(data.frame(x = .$d[[1]]$x,    # grab one set of x values (which are the same)
-										y = .$d[[1]]$y - .$d[[2]]$y)) %>%    # and subtract the y values
+										y = .$d[[1]]$y - .$d[[2]]$y))    # and subtract the y values
+		
+		D$Event[ D$y >= 0 ] = "Speciation"
+		D$Event[ D$y <= 0 ] = "Duplication"
+		D$Event = factor( D$Event, levels=c( "Speciation", "Duplication" ) )
+			
+		ggcontrasts = D %>%
 			ggplot(aes(x, y)) +    # now plot
 				geom_line() +
 				xlab( "Tau Phylogenetic Contrast" ) +
 				ylab( "Density Difference" ) +
 				xlim( 0, 0.05 ) + 
 				ylim( -30, 30 ) +
+				geom_area( aes( fill=Event ) ) +
 				geom_hline( yintercept = 0 ) +
-				theme_classic()
+				annotate( "text", x = Inf, y = -20, hjust="right", vjust="top", label=label_p ) +
+				theme_classic() +
+				theme(legend.position="none")
 		
 	}
 	

### 2017-04-01 manuscript.rmd
dup_adjust reduced from 5 to 2
added trees to repo

@@
 
-	# Name of the compara gene trees file. This is from 
+	# Name of the compara gene trees file. This was downloaded from 
 	# ftp://ftp.ensembl.org/pub/release-75/emf/ensembl-compara/homologies/
-	# but not included here due to its size. Put it in the kmrr/ directory
-	gene_tree_name = "Compara.75.protein.nhx.emf"
+	# and compressed with gz
+
+	gene_tree_name = "Compara.75.protein.nhx.emf.gz"
 
@@
 	# In simulation, the fold change in rate following duplication relative to speciation
-	dup_adjust = 5
+	dup_adjust = 2
 

### 2017-04-01 install_dependencies.R
added install_dependencies

@@ -1,0 +1,26 @@
+# Installs manuscript dependencies
+
+# From CRAN
+install.packages( 
+	c(
+		"devtools",
+		"tidyverse",
+		"magrittr",
+		"stringr",
+		"parallel",
+		"ape",
+		"digest",
+		"gridExtra",
+		"geiger",
+		"phytools"
+	) 
+)
+
+# From Bioconductor, needed for ggtree and treeio
+source("https://bioconductor.org/biocLite.R")
+biocLite()
+
+# From github
+devtools::install_github( "GuangchuangYu/treeio" )
+devtools::install_github( "GuangchuangYu/ggtree" )
+devtools::install_github( "caseywdunn/hutan" )

### 2017-04-01 install_dependencies.R
use all available cores

@@ -1,5 +1,10 @@
 # Installs manuscript dependencies
-
+# 
+# This installs all dependencies needed to run the 
+# manuscript, as verified by running it in a 
+# container based on the docker image 
+# https://hub.docker.com/r/rocker/tidyverse/
+#
 # From CRAN
 install.packages( 
 	c(
@@ -17,10 +22,10 @@ install.packages(
 )
 
 # From Bioconductor, needed for ggtree and treeio
-source("https://bioconductor.org/biocLite.R")
+source( "https://bioconductor.org/biocLite.R" )
 biocLite()
 
 # From github
-devtools::install_github( "GuangchuangYu/treeio" )
-devtools::install_github( "GuangchuangYu/ggtree" )
-devtools::install_github( "caseywdunn/hutan" )
+devtools::install_github( "GuangchuangYu/treeio", quiet=TRUE )
+devtools::install_github( "GuangchuangYu/ggtree", quiet=TRUE )
+devtools::install_github( "caseywdunn/hutan", quiet=TRUE )

### 2017-04-04 manuscript.rmd
use all available cores
applied suggestions from Marc Robinson-Rechavi, Matt Hahn, and others

@@
 	# Set system computational parameters
-	cores = detectCores() - 1
+	cores = detectCores()
 	if ( cores < 1 ) {
@@
 Phylogenetic comparative methods address this dependence.
-They have largely been applied to morphological and ecological traits, but are just as relevant to functional genomics.
+They have largely been applied to morphological and ecological traits, but are just as relevant to functional genomics [@Barker:2005be].
 Even so, most comparative functional genomics studies have abstained from phylogenetic approaches and instead rely on multiple pairwise comparisons across species (Figure 1a).
@@
 The KMRR analyses are based on pairwise comparisons (Figure 1a) between Tau within each gene family.
+Rather than make every pairwise comparison within each gene tree, they considered only a subset of pairwise comparisons in each particular analysis.
+They first selected a focal species, which varied from analysis to analysis.
+Ortholog comparisons were limitted to pairs that include this species, and the only paralogs considered were those with the highest expressin in this species.
+Note that this subset of pairwise comparisons still sample the same changes multiple times.
+
 They found the correlation coefficient of Tau for orthologs to be significantly greater than the correlation coefficient of Tau for paralogs, *i.e.* that orthologs tend to be more similar to each other than paralogs.
@@
 This is important since we are only looking at a subset of the data they considered, the Brawand *et al.* 2011 [@Brawand:2011du] dataset for gene trees that could be successfully time calibrated.
-In their Figure 1, they present a higher Tau correlation coefficient between ortholog pairs than paralog ortholog pairs.
+In their Figure 1, they present a higher Tau correlation coefficient between ortholog pairs than paralog pairs.
 We find the same here, with correlation coefficients of `r signif( ortholog_r, 2)` for orthologs and `r signif( paralog_r, 2)` for paralogs.
@@
 
-To better understand this plot (Figure 2a), we performed a couple simulations of Tau on the calibrated gene trees and then regenerated the figures.
+To better understand this plot (Figure 2a), we performed simulations of Tau on the calibrated gene trees and then regenerated the figures.
 We did not modify the gene tree topologies or their inferred histories of duplication and loss.
@@
 Our results suggest, at a minimum, that the ortholog conjecture is not a dominant pattern that is central to explaining the evolution of phenotypic diversity in gene families.
+This suggests that an alternative "neutral conjecture", *i.e.* the conjecture that the evolution of gene traits tends to be the same following gene duplication and speciation events, may better explain the process and patterns of most gene evolution.
+Under this neutral conjecture, phylogenetic distance is a better predictor of the similarity of gene function than is the history of gene duplication and speciation.
 The ortholog conjecture does not have to be an all or nothing question, though.
 It may be the case that the rates of phenotypic evolution following duplication may be greater than that following duplication in some organisms, gene families, and evolutionary processes [@Gabaldon:2013fi].
-We just don't find evidence for it when summarizing gene expression across tissues with Tau in these organisms.
+We just do not find evidence for it when summarizing gene expression across tissues with Tau in these organisms.
 This calls into question the general predictive power of the ortholog conjecture with respect to gene expression, and until these processes are better understood it will be necessary to test for it in each situation.
@@
 This in turn scales the phylogenetic independent contrasts (which take branch length into account) so they can be compared appropriately.
-We found a problem with the clade names on the ENSEMBL Compara trees, which others have also identified [@Daub:2016gd], that first had to be fixed.
-Hominini is the name for the clade that includes humans and chimps, while Homininae is the clade that includes humans, chimps, and gorillas.
-Both clades are labeled as Homininae in the Compara trees, though, which would interfere with calibration of these nodes.
-To fix the problem, we identified all clades labeled Homininae that have no gorilla sequence and renamed them Hominini.
+Before calibrating the trees, we had to slightly modify some of them.
+The node names in the ENSEMBL Compara [@Herrero:2016jj] gene phylogenies are parsed from the NCBI Taxonomy database, which has many polytomies, rather than a bifurcating species phylogeny.
+One implication of this is that node names can be resolved in such a way that a speciation node can have the same name as one of its speciation node ancestors, as others have noted [@Daub:2016gd].
+If left unaddressed, this would force all intervening branches to have length zero and interfere with calibration.
+In particular, Hominini is the name for the clade that includes humans and chimps, while Homininae is the clade that includes humans, chimps, and gorillas.
+Because of the structure of the NCBI Taxonomy, both clades are labeled as Homininae in the Compara trees.
+To remedy this, we identified all clades labeled Homininae that have no gorilla sequence and renamed them Hominini.
 We then calibrated the trees by fixing the speciation nodes to the dates specified in the KMRR code, with the exception of Hominini and Homininae.
@@
 
-Thanks to Steve Haddock, Alex Damian Serrano, August Guang, Bruno Vellutini, and Zack Lewis for feedback on the manuscript.
+Thanks to Steve Haddock, Alex Damian Serrano, August Guang, Bruno Vellutini, Zack Lewis, and Matthew Hahn for feedback on the manuscript.
 Marc Robinson-Rechavi provided very helpful information about the KMRR analyses and valuable suggestions for our manuscript.
+The Ensembl Comparative Genomics team, including Matthieu Muffato, provided excellent support for helping us better understand and use their Compara trees.
 CWD's contribution was supported by the National Science Foundation (DEB-1256695 and the Waterman Award).
@@
 There are also technical factors that can lead to an excess of duplication events deeper in the tree.
-Gene tree estimation errors errors, for example, tend to lead to the overestimation of deep duplications [@Hahn:2007cj].
+Gene tree estimation errors, for example, tend to lead to the overestimation of deep duplications [@Hahn:2007cj].
 If independent contrast values also tended to to be lower at greater node depth, it could interact with the preponderance of duplications at greater depth to create a pattern of lower contrasts associated with duplication events.

### 2017-04-18 manuscript.rmd
added explanation of lower duplication pairwise correlation

@@
 	library( phytools )
+	library( ggrepel )
 
@@
 ```{r pairwise_and_phylo_plots, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.width=6.5, fig.height=6.5, fig.cap="Pairwise (left column - a, c, e) and phylogenetic (right column - b, d, f) analyses of the original data (first row - a, b), data simulated under the null hypothesis (second row - c, d), and data simulated under the ortholog conjecture (third row- e, f).
-In the pairwise plots, each circle indicates the correlation coefficient of Tau for a set of pairwise comparisons annotated with a specific node name (e.g, Primates) and event type (speciation or duplication, giving rise respectively to orthologs and paralogs).
+In the pairwise plots, each point indicates the correlation coefficient of Tau for a set of pairwise comparisons annotated with a specific node name (e.g, Primates) and event type (speciation or duplication, giving rise respectively to orthologs and paralogs).
 The phylogenetic plots show the difference between the density distributions for Tau phylogenetic contrasts for speciation and duplication events, where a value above 0 indicates an excess of speciation contrasts in the indicated interval.
 A horizontal line at 0 would indicate that the density distributions are identical.
-The top left pane (a) reproduces the pattern presented in Figure 2a of KMRR [@KryuchkovaMostacci:2016iw], which they took as evidence of the ortholog conjecture.
-The recovery of this same pairwise pattern under both simulations (c, e) indicates that it does not make distinct predictions that can test hyptheses about the evolution of expression.
+The top left pane (a) reproduces the pattern presented in Figure 2a of KMRR [@KryuchkovaMostacci:2016iw] of higher correlation across speciation events than duplication events, which they took as evidence of the ortholog conjecture.
+The recovery of a similar  pattern under both simulations (c, e) indicates that it this pairwise approach does not make distinct testable predictions.
 The phylogenetic analysis of the original data (b) does not show an excess of larger contrasts for duplication events and does not reject the null hypothesis, providing no support for the ortholog conjecture.
@@
 				theme_classic() +
-				xlab( "Node Age (million years)" ) +
+				xlab( "Clade Age (million years)" ) +
 				ylab( "Correlation Coefficient" ) +
@@
 		pairwise_sim_null_summary %>% 
-		filter( species_a == "homo_sapiens" | species_b == "homo_sapiens" ) %>% 
 		filter( ! is.na( D ) ) %>%
@@
 			theme_classic() +
-			xlab( "Node Age (million years)" ) +
+			xlab( "Clade Age (million years)" ) +
 			ylab( "Correlation Coefficient" ) +
@@
 		pairwise_sim_oc_summary %>% 
-		filter( species_a == "homo_sapiens" | species_b == "homo_sapiens" ) %>% 
 		filter( ! is.na( D ) ) %>%
@@
 		g_pair_empirical,
-		make_contrast_plot(nodes_contrast),
+		make_contrast_plot(nodes_contrast) + 
+			ggtitle( "(b) Empirical, phylogenetic" ),
 		# For the rest, put spaces in labels so you can't see them but alignment is 
@@
 
-Their Figure 3 differs from their other analyses in that it presents independent changes in expression between triplets of genes.
-Each triplet has two paralogs that arose from a duplication event and one unduplicated homolog.
-They find that there is a tendency for the paralog with the lowest expression to have the highest Tau.
-From this they conclude that following duplication there is a common trajectory of evolutionary change, where one paralog evolves to have lower expression and also become more tissue specific.
-There is, however, a negative relationship between Tau and maximum expression across all genes (Supplementary Figure 6).
-A simpler explanation for this plot is that it reflects the global tendency for Tau to decrease with maximum expression.
-The null expectation for any two genes sampled at random is for one to have higher maximum expression and lower Tau, and the other to have lower maximum expression and higher Tau.
-This global pattern could be due to both biological factors and the technical details of how Tau is calculated.
+```{r cor_structure, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
+
+	# The threshold of allowed fractional devience of node age from clade age
+	age_window = 0.1
+
+	g_clade_age = pairwise_sim_null_summary %>% 
+		rename( clade = label ) %>% 
+		#filter( clade %in% speciation_clades ) %>%
+		left_join( calibration_times, clade=clade ) %>% 
+		filter( ! is.na( D ) ) %>%
+		ggplot( aes( x = age, y = node_age, col = Event ) ) +
+			geom_point( alpha = 0.1 ) +
+			xlab( "Clade Age (million years)" ) +
+			ylab( "Node Age (million years)" ) +
+			xlim( 0, 1000) +
+			ylim( 0, 1500 ) +
+			ggtitle( "A" ) +
+			theme_classic() +
+			theme( legend.position="none" )
+
+	pairwise_sim_null_summary_sub = 
+		pairwise_sim_null_summary %>% 
+		filter( ! is.na( D ) ) %>% rename( clade = label ) %>% 
+		left_join( calibration_times, clade=clade ) %>% 
+		filter( node_age > ( ( 1 - age_window ) * age ) ) %>%
+		filter( node_age < ( ( 1 + age_window ) * age ) )
+
+	g_sub = 
+		pairwise_sim_null_summary_sub %>%
+		group_by( Event, clade ) %>% 
+		summarise( cor.test( tau_a, tau_b )$estimate, n=n() ) %>% 
+		rename( rho=`cor.test(tau_a, tau_b)$estimate` ) %>%
+		left_join( calibration_times, clade=clade ) %>%
+		ggplot() + 
+			geom_point( aes( x=age, y=rho, col=Event ), size=3 ) +
+			theme_classic() +
+			xlab( "Clade Age (million years)" ) +
+			ylab( "Correlation Coefficient" ) +
+			xlim( 0, 1000) +
+			ylim( 0, 1 ) +
+			guides( colour = guide_legend( override.aes = list( shape = 15 ) ) ) +
+			theme( 
+				legend.justification=c( 1, 1 ), 
+				legend.position=c( 1, 1 ), 
+				legend.title=element_blank() ) +
+			ggtitle( "B" )
+
+	n_dup_original = 
+		pairwise_sim_null_summary %>%
+		filter( Event=="Duplication" ) %>%
+		nrow()
+
+	n_dup_remaining = 
+		pairwise_sim_null_summary_sub %>%
+		filter( Event=="Duplication" ) %>%
+		nrow()
+
+```
+
+Since greater rates of expression evolution following duplication do not explain the lower correlation of Tau values in the pairwise comparison plots (Figure 2 A, C, E), this pattern must reflect some other property that is shared across these analyses.
+We found that the lower correlatations for duplication comparisons can in largely be explained by the greater variance in the age of duplication nodes.
+Each point in the pairwise correlation plot (Figure 2 A, C, E) summarizes multiple pairwise comparisons across nodes of a given event type (speciation or duplication, as indicated by color) and clade in the species tree (Theria, Mammalia, Amniota, etc..., as indicated by the age of the clade along the x axis).
+These clade names in turn are derived from the node annotations in the ENSEMBL Compara trees, which apply clade names to both speciation and duplication nodes.
+While speciation nodes in the gene trees have a clear correspondence to clades in the species trees, duplucation nodes do not since duplication can occur at any point along any branch in the species tree.
+Neighboring clade ages are therefore a very rough approximation of the age of duplication events.
+This is apparent when node ages derived from calibrated trees are plotted against the age of the clade annotations for each node (Supplementary Figure 6A).
+Because duplication events that are all annotated with the same clade name can occur at very different times, pairwise comparisons across these nodes capture evolutionary changes in Tau across very different different branch lengths.
+They therefore have lower correlation than pairwise comparisons made across speciation nodes.
+Of the `r n_dup_original` duplication nodes considered here, `r n_dup_remaining` have a node age (as determined by the time calibrated trees) within `r age_window * 100`% of their clade annotation.
+If only these duplication events closest to their annotated clades are retained, the pattern of lower correlation of Tau evolution for duplicaiton events disappears for the dataset simulated under the null model (Supplementary Figure 6B).
+This correction, though, comes at the high cost of discarding `r round( (n_dup_original - n_dup_remaining )/ n_dup_original * 100, 1 )`% of duplication nodes.
 
@@
 
+#### Origin of pairwise correlation plot structure
+
+```{r cor_structure_plot, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, dev="png", dpi=600, fig.width=6.5, fig.height=6.5, fig.cap="Origin of pairwise correlation plot structure.
+(*A*) Each speciation and duplication node in the COMPARA trees is annotated with a clade name, and the age of that clade (shown on the x axis) is used as the age of the pairwise comparisons (Figure 2 A, C, E, KMRR[@KryuchkovaMostacci:2016iw] Figure 2).
+The clade ages have a clear correspondence to the speciation node ages (red).
+Duplication (blue) ages for nodes assigned to the same clade, however, have large variation in calibrated node age (y axis).
+This means that summaries of expression differences across duplication events represent evolutionary changes that occurred over a wide range of time scales.
+(*B*) When duplication nodes whose calibrated node age deviates more than 10% from their annotated clade age are removed from the data simulated under the null distribution, duplication comparisons no longer have a uniformly lower expression correlation than speciation comparisons."}
+
+	grid.arrange(
+		g_clade_age,
+		g_sub,
+		ncol = 1
+		)
+
+```
+
 #### Relationship between Tau and maximum expression
@@
 
-There is a negative relationship between miximum expression and Tau for each gene (Supplementary Figure 6, $r^2=`r signif( tau_v_max_lm$r.squared, 3 )`$, $slope=`r signif( tau_v_max_lm$coefficients[2,1], 3 )`$, $p=`r signif( tau_v_max_lm$coefficients[2,4], 3 )`$).
+Figure 3 if KMRR differs from their other analyses in that it presents independent changes in expression between triplets of genes.
+Each triplet has two paralogs that arose from a duplication event and one unduplicated homolog.
+They find that there is a tendency for the paralog with the lowest expression to have the highest Tau.
+From this they conclude that following duplication there is a common trajectory of evolutionary change, where one paralog evolves to have lower expression and also become more tissue specific.
+There is, however, a negative relationship between Tau and maximum expression across all genes (Supplementary Figure 7, $r^2=`r signif( tau_v_max_lm$r.squared, 3 )`$, $slope=`r signif( tau_v_max_lm$coefficients[2,1], 3 )`$, $p=`r signif( tau_v_max_lm$coefficients[2,4], 3 )`$).
+A simpler explanation for this plot is that it reflects the global tendency for Tau to decrease with maximum expression.
+The null expectation for any two genes sampled at random is for one to have higher maximum expression and lower Tau, and the other to have lower maximum expression and higher Tau.
+This global pattern could be due to both biological factors and the technical details of how Tau is calculated.
 

### 2017-04-18 manuscript.rmd
reformatted figure lettering

@@
 They have largely been applied to morphological and ecological traits, but are just as relevant to functional genomics [@Barker:2005be].
-Even so, most comparative functional genomics studies have abstained from phylogenetic approaches and instead rely on multiple pairwise comparisons across species (Figure 1a).
+Even so, most comparative functional genomics studies have abstained from phylogenetic approaches and instead rely on multiple pairwise comparisons across species (Figure 1*A*).
 This leaves comparative functional genomics studies susceptible to statistical problems and is a missed opportunity to ask questions that are only accessible in an explicit phylogenetic context.
 
-Phylogenetic comparative methods account for evolutionary history and explicitly model trait change along the branches of evolutionary trees (*e.g.*, Figure 1b).
+Phylogenetic comparative methods account for evolutionary history and explicitly model trait change along the branches of evolutionary trees (*e.g.*, Figure 1*B*).
 The value of these methods relative to to pairwise comparisons has been repeatedly shown in analyses of other types of character data [@Ricklefs:1996wb; @Chamberlain:2012vx; @OMeara:2012gq].
@@
 The first study, Kryuchkova-Mostacci and Robinson-Rechavi (KMRR) 2016 [@KryuchkovaMostacci:2016iw], analyzed multiple vertebrate expression datasets to test the ortholog conjecture - the hypothesis that orthologs tend to have more conserved attributes (specificity of expression across organs in this case) than do paralogs [@Nehrt:2011bm, @Koonin:2005fp].
-Using pairwise comparisons (Figure 1a), they found lower expression correlation between paralogs than between orthologs and interpreted this as strong support for the ortholog conjecture.
+Using pairwise comparisons (Figure 1*A*), they found lower expression correlation between paralogs than between orthologs and interpreted this as strong support for the ortholog conjecture.
 
@@
 The internal nodes of the tree are speciation and gene duplication events.
-(a) Many comparative functional genomic studies rely on pairwise comparisons, where traits of each gene are compared to traits of other genes across species.
+(*A*) Many comparative functional genomic studies rely on pairwise comparisons, where traits of each gene are compared to traits of other genes across species.
 This leads to many more comparisons than unique observations, making each comparison dependent on others.
-(b) Comparative phylogenetic methods, including phylogenetic independent contrasts [@Felsenstein:1985ua], make a smaller number of independent comparisons, where each contrast measures independent changes along different branches.
+(*B*) Comparative phylogenetic methods, including phylogenetic independent contrasts [@Felsenstein:1985ua], make a smaller number of independent comparisons, where each contrast measures independent changes along different branches.
 Phylogenetic approaches are rarely used for functional genomic studies.](Figure_overview.pdf)
@@
 
-The KMRR analyses are based on pairwise comparisons (Figure 1a) between Tau within each gene family.
+The KMRR analyses are based on pairwise comparisons (Figure 1*A*) between Tau within each gene family.
 Rather than make every pairwise comparison within each gene tree, they considered only a subset of pairwise comparisons in each particular analysis.
@@
 For each internal node in each gene tree, we calculated the phylogenetic independent contrast (PIC) of Tau.
-This is the difference in values of Tau for descendant nodes scaled by the expected variance, which is largely determined by the lengths of the two branches that connect the node to its two descendants (Figure 1b).
+This is the difference in values of Tau for descendant nodes scaled by the expected variance, which is largely determined by the lengths of the two branches that connect the node to its two descendants (Figure 1*B*).
 These contrasts were then annotated by whether each is made across a speciation or duplication event.
@@
 These calibrated trees were used to calculate phylogenetic independent contrasts for `r nodes_contrast %>% filter(D=="Y") %>% nrow()` duplication nodes and `r nodes_contrast %>% filter(D=="N") %>% nrow()` speciation nodes.
-One of these trees is shown in Supplementary Figure 4 to demonstrate the analysis.
+One of these trees is presented in *SI Appendix* Figure 4 to demonstrate the analysis.
 
@@
 
-We did not find increased evolutionary change in expression following duplication events (Figure 2b).
+We did not find increased evolutionary change in expression following duplication events (Figure 2*B*).
 The Wilcoxon rank test does not reject the null hypothesis that the rate of evolution following duplications is the same as or less than the rate following speciation (p value = `r wilcox_test_result`).
@@
 We focused on two possible sources of bias, node depth and branch length.
-We found no evidence that either affected our results (Supplementary Information, Supplementary Figure 4).
+We found no evidence that either affected our results (*SI Appendix*, Figure 5).
 We also examined the sensitivity of the results to the calibration times applied to speciation events on the gene trees.
@@
 There is also uncertainty associated with the timing of these speciation events.
-We added random noise to the calibration times in replicate analyses, and all still failed to reject the null hypothesis (Supplementary Information).
+We added random noise to the calibration times in replicate analyses, and all still failed to reject the null hypothesis (*SI Appendix*).
 
@@
 
-To test the hypothesis that ancient duplications that precede the oldest speciation event (Supplementary Figure 5a) impact the lower correlation of Tau between paralogs than between orthologs, we removed them.
-When we consider only the duplication events the same age or younger than the oldest speciation event (Supplementary Figure 5b), the paralog correlation coefficient increases from `r signif( paralog_r, 2)` to `r signif( paralog_subset_r, 2)`.
+To test the hypothesis that ancient duplications that precede the oldest speciation event (*SI Appendix*, Figure 5*A*) impact the lower correlation of Tau between paralogs than between orthologs, we removed them.
+When we consider only the duplication events the same age or younger than the oldest speciation event (*SI Appendix*, Figure 5*B*), the paralog correlation coefficient increases from `r signif( paralog_r, 2)` to `r signif( paralog_subset_r, 2)`.
 This is much closer to the ortholog correlation of `r signif( ortholog_r, 2) `.
@@
 
-```{r pairwise_and_phylo_plots, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.width=6.5, fig.height=6.5, fig.cap="Pairwise (left column - a, c, e) and phylogenetic (right column - b, d, f) analyses of the original data (first row - a, b), data simulated under the null hypothesis (second row - c, d), and data simulated under the ortholog conjecture (third row- e, f).
+```{r pairwise_and_phylo_plots, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.width=6.5, fig.height=6.5, fig.cap="Pairwise (left column - *A, C, E*) and phylogenetic (right column - *B, D, F*) analyses of the original data (first row - *A, B*), data simulated under the null hypothesis (second row - *C, D*), and data simulated under the ortholog conjecture (third row- *E, F*).
 In the pairwise plots, each point indicates the correlation coefficient of Tau for a set of pairwise comparisons annotated with a specific node name (e.g, Primates) and event type (speciation or duplication, giving rise respectively to orthologs and paralogs).
@@
 A horizontal line at 0 would indicate that the density distributions are identical.
-The top left pane (a) reproduces the pattern presented in Figure 2a of KMRR [@KryuchkovaMostacci:2016iw] of higher correlation across speciation events than duplication events, which they took as evidence of the ortholog conjecture.
-The recovery of a similar  pattern under both simulations (c, e) indicates that it this pairwise approach does not make distinct testable predictions.
-The phylogenetic analysis of the original data (b) does not show an excess of larger contrasts for duplication events and does not reject the null hypothesis, providing no support for the ortholog conjecture.
-The bottom right panes validate the phylogenetic approach by showing that it does not reject the null when data are simulated under the null (d), but does reject the null when data are simulated under the ortholog conjecture (f)."}
+The top left pane (*A*) reproduces the pattern presented in Figure 2A of KMRR [@KryuchkovaMostacci:2016iw] of higher correlation across speciation events than duplication events, which they took as evidence of the ortholog conjecture.
+The recovery of a similar  pattern under both simulations (*C, E*) indicates that it this pairwise approach does not make distinct testable predictions.
+The phylogenetic analysis of the original data (*B*) does not show an excess of larger contrasts for duplication events and does not reject the null hypothesis, providing no support for the ortholog conjecture.
+The bottom right panes validate the phylogenetic approach by showing that it does not reject the null when data are simulated under the null (*D*), but does reject the null when data are simulated under the ortholog conjecture (*F*)."}
 
@@
 				ylim( 0, 1 ) +
-				ggtitle( "(a) Empirical, pairwise" ) +
+				ggtitle( "A Empirical, pairwise" ) +
 				guides( colour = guide_legend( override.aes = list( shape = 15 ) ) ) +
@@
 			ylim( 0, 1 ) +
-			ggtitle( "(c) Null simulation, pairwise" ) +
+			ggtitle( "C Null simulation, pairwise" ) +
 			theme( legend.position="none" )
@@
 			ylim( 0, 1 ) +
-			ggtitle( "(e) OC simulation, pairwise" ) +
+			ggtitle( "E OC simulation, pairwise" ) +
 			theme( legend.position="none" )
@@
 		make_contrast_plot(nodes_contrast) + 
-			ggtitle( "(b) Empirical, phylogenetic" ),
+			ggtitle( "B Empirical, phylogenetic" ),
 		# For the rest, put spaces in labels so you can't see them but alignment is 
@@
 			ylab( "  " ) + 
-			ggtitle( "(d) Null simulation, phylogenetic" ),
+			ggtitle( "D Null simulation, phylogenetic" ),
 		g_pair_oc + 
@@
 			ylab( "  " ) + 
-			ggtitle( "(f) OC simulation, phylogenetic" ),
+			ggtitle( "F OC simulation, phylogenetic" ),
 		ncol = 2
@@
 They found that across the investigated range of node ages, ortholog pairs have higher Tau correlation than paralogs.
-We confirmed that we can replicate this result (Figure 2a).
+We confirmed that we can replicate this result (Figure 2*A*).
 There are several difficulties with interpreting this plot, though.
@@
 
-To better understand this plot (Figure 2a), we performed simulations of Tau on the calibrated gene trees and then regenerated the figures.
+To better understand this plot (Figure 2*A*), we performed simulations of Tau on the calibrated gene trees and then regenerated the figures.
 We did not modify the gene tree topologies or their inferred histories of duplication and loss.
 First, we simulated the evolution of Tau under the null model that it evolves at the same rate following duplication and speciation events.
-Under the null model, the plot of correlation coefficient to node age is very similar as for the observed data (Figure 2c).
+Under the null model, the plot of correlation coefficient to node age (Figure 2*C*) is very similar as for the observed data (Figure 2*A*).
 As in the original study, there is higher correlation coefficient across orthologs (`r signif( ortholog_r_sim_null, 2 )`) than paralogs (`r signif( paralog_r_sim_null, 2 )`) when not considering node age.
-Phylogenetic analysis of the data simulated under the null hypothesis (Figure 2d) do not reject the null hypothesis (Wilcoxon p = $`r wilcox_oc( nodes_sim_null_contrast )`$), as expected.
+Phylogenetic analysis of the data simulated under the null hypothesis (Figure 2*D*) do not reject the null hypothesis (Wilcoxon p = $`r wilcox_oc( nodes_sim_null_contrast )`$), as expected.
 
 We next simulated the evolution of Tau under the ortholog conjecture, where the rate of evolution of Tau following duplication was `r dup_adjust` fold the rate following speciation.
-The pairwise results of this heterogeneous model (Figure 2e) are nearly indistinguishable from the results under the null model (Figure 2c), and also have a higher correlation coefficient for orthologs (`r signif( ortholog_r_sim_oc, 2 )`) than paralogs (`r signif( paralog_r_sim_oc, 2 )`).
-The phylogenetic analysis of the ortholog conjecture simulation (Figure 2f) does reject the null hypothesis (Wilcoxon p = $`r signif( wilcox_oc( nodes_sim_oc_contrast ), 2 )`$).
+The pairwise results of this heterogeneous model (Figure 2*E*) are nearly indistinguishable from the results under the null model (Figure 2*C*), and also have a higher correlation coefficient for orthologs (`r signif( ortholog_r_sim_oc, 2 )`) than paralogs (`r signif( paralog_r_sim_oc, 2 )`).
+The phylogenetic analysis of the ortholog conjecture simulation (Figure 2*F*) does reject the null hypothesis (Wilcoxon p = $`r signif( wilcox_oc( nodes_sim_oc_contrast ), 2 )`$).
 
@@
 
-Since greater rates of expression evolution following duplication do not explain the lower correlation of Tau values in the pairwise comparison plots (Figure 2 A, C, E), this pattern must reflect some other property that is shared across these analyses.
-We found that the lower correlatations for duplication comparisons can in largely be explained by the greater variance in the age of duplication nodes.
-Each point in the pairwise correlation plot (Figure 2 A, C, E) summarizes multiple pairwise comparisons across nodes of a given event type (speciation or duplication, as indicated by color) and clade in the species tree (Theria, Mammalia, Amniota, etc..., as indicated by the age of the clade along the x axis).
+Since greater rates of expression evolution following duplication do not explain the lower correlation of Tau values in the pairwise comparison plots (Figure 2 *A, C, E*), this pattern must reflect some other property that is shared across these analyses.
+We found that the lower correlations for duplication comparisons can in largely be explained by the greater variance in the age of duplication nodes.
+Each point in the pairwise correlation plot (Figure 2 *A, C, E*) summarizes multiple pairwise comparisons across nodes of a given event type (speciation or duplication, as indicated by color) and clade in the species tree (Theria, Mammalia, Amniota, etc..., as indicated by the age of the clade along the x axis).
 These clade names in turn are derived from the node annotations in the ENSEMBL Compara trees, which apply clade names to both speciation and duplication nodes.
-While speciation nodes in the gene trees have a clear correspondence to clades in the species trees, duplucation nodes do not since duplication can occur at any point along any branch in the species tree.
+While speciation nodes in the gene trees have a clear correspondence to clades in the species trees, duplication nodes do not since duplication can occur at any point along any branch in the species tree.
 Neighboring clade ages are therefore a very rough approximation of the age of duplication events.
-This is apparent when node ages derived from calibrated trees are plotted against the age of the clade annotations for each node (Supplementary Figure 6A).
+This is apparent when node ages derived from calibrated trees are plotted against the age of the clade annotations for each node (*SI Appendix*, Figure 6*A*).
 Because duplication events that are all annotated with the same clade name can occur at very different times, pairwise comparisons across these nodes capture evolutionary changes in Tau across very different different branch lengths.
@@
 Of the `r n_dup_original` duplication nodes considered here, `r n_dup_remaining` have a node age (as determined by the time calibrated trees) within `r age_window * 100`% of their clade annotation.
-If only these duplication events closest to their annotated clades are retained, the pattern of lower correlation of Tau evolution for duplicaiton events disappears for the dataset simulated under the null model (Supplementary Figure 6B).
+If only these duplication events closest to their annotated clades are retained, the pattern of lower correlation of Tau evolution for duplication events disappears for the dataset simulated under the null model (*SI Appendix*, Figure 5*B*).
 This correction, though, comes at the high cost of discarding `r round( (n_dup_original - n_dup_remaining )/ n_dup_original * 100, 1 )`% of duplication nodes.
@@
 Model of variance, which is inversely related to similarity, is on the left.
-(a) The distributions as published by Levin *et al.* [@Levin:2016bp].
+(*A*) The distributions as published by Levin *et al.* [@Levin:2016bp].
 Low similarity (*i.e.*, high variance) in the mid phase of development was interpreted as support for an inverse hourglass model for the evolution of gene expression.
@@
 Haddock from phylopic.org.
-(b) The distributions after the exclusion of the ctenophore.
+(*B*) The distributions after the exclusion of the ctenophore.
 The early and mid phase distributions are not statistically distinct.](levin_etal/Figure_levin.pdf)
@@
 Once we addressed the issues above with the implementation of pairwise comparisons, we were able to explore more general issues that can be a problem when making multiple pairwise comparisons between species.
-We found that all five of the lowest values in the mid phase distribution (Figure 3a) are for pairwise comparisons that include the ctenophore (comb jelly).
+We found that all five of the lowest values in the mid phase distribution (Figure 3*A*) are for pairwise comparisons that include the ctenophore (comb jelly).
 When the nine pairwise comparisons that include the ctenophore are removed, there is no significant difference between the early phase and mid phase distributions (p = 0.14 for the early-mid comparison and p < $10^{-5}$ for the late-mid comparison) and no support for the inverse hourglass (Figure 3b).
 This highlights a well understood property of pairwise comparisons across species [@Felsenstein:1985ua; @Ackerly:1999iu]: evolutionary changes along a given branch, like those along the ctenophore branch, impact each of the multiple pairwise comparisons that includes that branch.
-The pairwise comparisons are therefore not independent - different pairwise comparisons are impacted by changes along some of the same branches (Figure 1a).
+The pairwise comparisons are therefore not independent - different pairwise comparisons are impacted by changes along some of the same branches (Figure 1*A*).
 This can give the impression of a general pattern across the tree that is instead specific to changes along one part of the tree.
@@
 Two paralogs that diverged from a duplication event deep in the tree may have many subsequent duplication and speciation events, and changes along all these branches will impact the final pairwise comparison.
-Phylogenetic methods have the advantage of isolating the changes under consideration (Figure 1b).
+Phylogenetic methods have the advantage of isolating the changes under consideration (Figure 1*B*).
 The phylogenetic methods avoid diluting the change that occurs along the branches that follow the node in question with changes along all subsequent branches.
@@
 We used the `chronos()` function from the R package ape [@Paradis:2004dv] for this calibration, with the `correlated` model.
-See Supplementary Information for additional sensitivity analyses to time calibration.
+See the *SI Appendix* for additional sensitivity analyses to time calibration.
 Some trees could not be calibrated with these hard node constraints, and were discarded.
@@
 
-While the age of speciation nodes is constrained, duplication nodes can be much older and therefore have a wider range of ages (Supplementary Figure 5a).
+While the age of speciation nodes is constrained, duplication nodes can be much older and therefore have a wider range of ages (*SI Appendix*, Figure 5*A*).
 This is because many gene duplication events are older than the most recent common ancestor of the species in the study.
@@
 To test for such an effect, we remove duplication nodes that are older than the oldest speciation node.
-Our results are unchanged and this reduced dataset does not reject the null hypothesis that the rate of evolution following duplications is the same as or less than the rate following speciation (Supplementary Figure 5b).
+Our results are unchanged and this reduced dataset does not reject the null hypothesis that the rate of evolution following duplications is the same as or less than the rate following speciation (*SI Appendix*, Figure 5*B*).
 
@@
 This is because the same difference across a shorter total branch length indicates a greater rate of evolutionary change.
-The expected variance of contrasts for speciation nodes is constrained by the branch lengths on the species tree, but the expected variance of contrasts for duplications has a much wider range (Figure 5c).
+The expected variance of contrasts for speciation nodes is constrained by the branch lengths on the species tree, but the expected variance of contrasts for duplications has a much wider range (*SI Appendix*, Figure 5*C*).
 This could lead to biases if the lengths of branches that descend from duplication nodes tend to be overestimated.
 We therefore examined only the contrasts that fell within in the range of expected variance seen for speciation contrasts, excluding duplication contrasts that fall outside of this range.
-This reanalysis does not reject the null hypothesis either (Figure 5d), indicating that branch length bias is not responsible for the result.
+This reanalysis does not reject the null hypothesis either (*SI Appendix*, Figure 5*D*), indicating that branch length bias is not responsible for the result.
 
 ```{r ascertainment_plot, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, dev="png", dpi=600, fig.cap="Investigation of possible ascertainment biases.
-(a) Magnitude of independent contrasts plotted against node age.
+(*A*) Magnitude of independent contrasts plotted against node age.
 Speciation nodes are calibrated to particular times, whereas duplication nodes have a wider range.
-(b) Difference between density distributions of contrasts for only the nodes that have an age less than or equal to the maximum age of speciation nodes.
-(c) Magnitude of independent contrasts plotted against expected variance, which is largely determined by branch lengths.
+(*B*) Difference between density distributions of contrasts for only the nodes that have an age less than or equal to the maximum age of speciation nodes.
+(*C*) Magnitude of independent contrasts plotted against expected variance, which is largely determined by branch lengths.
 Contrasts for speciation nodes have a narrower range of expected variance than do contrasts for duplication nodes.
-(d) Difference between density distributions of contrasts for only the nodes that have expected variance within the range of contrasts across speciation nodes."}
+(*D*) Difference between density distributions of contrasts for only the nodes that have expected variance within the range of contrasts across speciation nodes."}
 
@@
 			xlim( 0, 1000 ) +
-			ggtitle( "(a)" ) +
+			ggtitle( "A" ) +
 			theme_classic()	+
@@
 		xlim( 0, 0.075 ) + 
-		ggtitle( "(b)" ) +
+		ggtitle( "B" ) +
 		annotate( "text", x = Inf, y = Inf, hjust="right", vjust="top", label=label_age ) +
@@
 		xlim( 0, 2000 ) +
-		ggtitle( "(c)" ) +
+		ggtitle( "C" ) +
 		theme_classic()	+
@@
 		geom_freqpoly( binwidth=.001, position="identity" ) + 
-		ggtitle( "(d)" ) +
+		ggtitle( "D" ) +
 		theme( legend.justification=c( 1, 1 ), legend.position=c( 1, 1 ) ) +
@@
 ```{r cor_structure_plot, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, dev="png", dpi=600, fig.width=6.5, fig.height=6.5, fig.cap="Origin of pairwise correlation plot structure.
-(*A*) Each speciation and duplication node in the COMPARA trees is annotated with a clade name, and the age of that clade (shown on the x axis) is used as the age of the pairwise comparisons (Figure 2 A, C, E, KMRR[@KryuchkovaMostacci:2016iw] Figure 2).
+(*A*) Each speciation and duplication node in the COMPARA trees is annotated with a clade name, and the age of that clade (shown on the x axis) is used as the age of the pairwise comparisons (Figure 2 *A, C, E*, KMRR[@KryuchkovaMostacci:2016iw] Figure 2).
 The clade ages have a clear correspondence to the speciation node ages (red).
@@
 
-Figure 3 if KMRR differs from their other analyses in that it presents independent changes in expression between triplets of genes.
+KMRR[@KryuchkovaMostacci:2016iw] Figure 3 differs from their other analyses in that it presents independent changes in expression between triplets of genes.
 Each triplet has two paralogs that arose from a duplication event and one unduplicated homolog.
@@
 From this they conclude that following duplication there is a common trajectory of evolutionary change, where one paralog evolves to have lower expression and also become more tissue specific.
-There is, however, a negative relationship between Tau and maximum expression across all genes (Supplementary Figure 7, $r^2=`r signif( tau_v_max_lm$r.squared, 3 )`$, $slope=`r signif( tau_v_max_lm$coefficients[2,1], 3 )`$, $p=`r signif( tau_v_max_lm$coefficients[2,4], 3 )`$).
+There is, however, a negative relationship between Tau and maximum expression across all genes (*SI Appendix* Figure 7, $r^2=`r signif( tau_v_max_lm$r.squared, 3 )`$, $slope=`r signif( tau_v_max_lm$coefficients[2,1], 3 )`$, $p=`r signif( tau_v_max_lm$coefficients[2,4], 3 )`$).
 A simpler explanation for this plot is that it reflects the global tendency for Tau to decrease with maximum expression.

### 2017-05-01 manuscript.rmd
typos

@@
 If phylogenetic relationships are not considered, analyses indicate a negative correlation of leaf lifespan with leaf size.
-With the application of phylogenetic comparative methods, however, this correlation disappears because most of the observed variance is due to differences between conifers and flowering plants.
+With the application of phylogenetic comparative methods, however, this correlation disappears because most of the observed variance is due to differences between just two groups, conifers and flowering plants.
 Phylogenetic comparative methods capture this shift as change along a single branch deep in the tree, revealing that there is not a tendency of correlated change between these traits across the phylogeny.
@@
 
-The first study, Kryuchkova-Mostacci and Robinson-Rechavi (KMRR) 2016 [@KryuchkovaMostacci:2016iw], analyzed multiple vertebrate expression datasets to test the ortholog conjecture - the hypothesis that orthologs tend to have more conserved attributes (specificity of expression across organs in this case) than do paralogs [@Nehrt:2011bm, @Koonin:2005fp].
+The first study, Kryuchkova-Mostacci and Robinson-Rechavi (KMRR) [@KryuchkovaMostacci:2016iw], analyzed multiple vertebrate expression datasets to test the ortholog conjecture - the hypothesis that orthologs tend to have more conserved attributes (specificity of expression across organs in this case) than do paralogs [@Nehrt:2011bm, @Koonin:2005fp].
 Using pairwise comparisons (Figure 1*A*), they found lower expression correlation between paralogs than between orthologs and interpreted this as strong support for the ortholog conjecture.
@@
 
-The KMRR study [@KryuchkovaMostacci:2016iw] sought to test the ortholog conjecture.
+KMRR [@KryuchkovaMostacci:2016iw] sought to test the ortholog conjecture.
 The ortholog conjecture [@Nehrt:2011bm] is the proposition that orthologs (genes that diverged from each other due to a speciation event) have more similar attributes than do paralogs (genes that diverged from each other due to a gene duplication event).
@@
 
-The KMRR study considered several publicly available datasets of gene expression across tissues and species.
+KMRR considered several publicly available datasets of gene expression across tissues and species.
 Their expression summary statistic is Tau [@Yanai:2005dh, @KryuchkovaMostacci:2016in], an indicator of tissue specificity of gene expression.
@@
 They first selected a focal species, which varied from analysis to analysis.
-Ortholog comparisons were limitted to pairs that include this species, and the only paralogs considered were those with the highest expressin in this species.
+Ortholog comparisons were limited to pairs that include this species, and the only paralogs considered were those with the highest expression in this species.
 Note that this subset of pairwise comparisons still sample the same changes multiple times.
@@
 
-We mapped the Tau values calculated by KMRR [@KryuchkovaMostacci:2016iw] for the Brawand *et al.* 2011 [@Brawand:2011du] datset onto `r n_gene_trees` gene trees parsed from ENSEMBL Compara [@Herrero:2016jj].
+We mapped the Tau values calculated by KMRR [@KryuchkovaMostacci:2016iw] for the Brawand *et al.* 2011 [@Brawand:2011du] dataset onto `r n_gene_trees` gene trees parsed from ENSEMBL Compara [@Herrero:2016jj].
 These are the same pre-computed trees that the orthology/ paralogy annotations KMRR used are based on.
@@
 The Wilcoxon rank test does not reject the null hypothesis that the rate of evolution following duplications is the same as or less than the rate following speciation (p value = `r wilcox_test_result`).
-Our phylogenetic comparative analysis, unlike the previously published pairwise comparative analysis [@KryuchkovaMostacci:2016iw], therefore finds no evidence for the ortholog conjecture in this system.
+Our phylogenetic comparative analysis, unlike the previously published pairwise comparative analysis [@KryuchkovaMostacci:2016iw], therefore finds no support for the ortholog conjecture in this system.
 
@@
 Our phylogenetic analysis of independent contrasts does not reject the null hypothesis when data are simulated under the null model, and does reject the null hypothesis when the data are simulated under the ortholog conjecture.
-In contrast to pairwise methods, the phylogenetic analyses make testable predictions based on explicit hypotheses about evolutionary process.
+In contrast to pairwise methods, the phylogenetic analyses can test explicit predictions based on hypotheses about evolutionary process.
 
@@
 Since greater rates of expression evolution following duplication do not explain the lower correlation of Tau values in the pairwise comparison plots (Figure 2 *A, C, E*), this pattern must reflect some other property that is shared across these analyses.
-We found that the lower correlations for duplication comparisons can in largely be explained by the greater variance in the age of duplication nodes.
+We found that the lower correlations for duplication comparisons can largely be explained by the greater variance in the age of duplication nodes.
 Each point in the pairwise correlation plot (Figure 2 *A, C, E*) summarizes multiple pairwise comparisons across nodes of a given event type (speciation or duplication, as indicated by color) and clade in the species tree (Theria, Mammalia, Amniota, etc..., as indicated by the age of the clade along the x axis).
@@
 This creates statistical challenges as the resulting covariance matrices are singular and, if not treated appropriately, imply many false correlations that are artifacts of project design.
-We outlined these challenges and potential solutions in the context of gene expression [@Dunn:2013cpa].
+We outlined these challenges and potential solutions in the context of gene expression elsewhere [@Dunn:2013cpa].
 In the same manuscript we also considered another issue of relevance here -- the read counts generated by RNA-seq expression studies cannot be directly compared across species.
@@
 
-The fact that the first two comparative functional genomics studies we reanalyzed show serious problems with pairwise comparisons indicates that there likely to be similar problems in other studies that use these methods, and that future comparative studies will be compromised if they continue to use pairwise methods.
+The fact that the first two comparative functional genomics studies we reanalyzed show serious problems with pairwise comparisons indicates that there are likely to be similar problems in other studies that use these methods, and that future comparative studies will be compromised if they continue to use pairwise methods.
 Studies of evolutionary functional genomics should not be focused on the tips of the tree using pairwise comparisons.
 They should explicitly delve into the tree with phylogenetic comparative methods.
-Here we show a concrete case where the predictions of a null hypotheses and an alternatives hypothesis do not differ when considering pairwise comparisons, but make distinct predictions with phylogenetic comparative methods that can be used to test them.
+Here we show a concrete case where the predictions of a null hypotheses and an alternatives hypothesis do not differ when considering pairwise comparisons, but do differ with phylogenetic comparative methods.
 
@@
 
-	label_age = str_c( "Wilcoxon p=", signif( wilcox_filtered_age, 2 ) )
-
-	g_density_filtered_age =
-		nodes_contrast_filtered_age %>%
-		ggplot( aes( x=pic, y=..density.., col=Event ) ) + 
-		geom_freqpoly( binwidth=.001, position="identity" ) +
-		xlab( "Phylogenetic independent contrast" ) +
-		ylab( "Density" ) +
-		xlim( 0, 0.075 ) + 
-		ggtitle( "B" ) +
-		annotate( "text", x = Inf, y = Inf, hjust="right", vjust="top", label=label_age ) +
-		theme_classic() +
-		theme(legend.position="none")
-
 	# Examine by expected variance (branch length)
@@
 
-	label_var = str_c( "Wilcoxon p=", signif ( wilcox_filtered_var, 2 ) )
-
-	g_density_filtered_var =
-		nodes_contrast_filtered_var %>% 
-		ggplot( aes( x=pic, y=..density.., col=Event ) ) + 
-		geom_freqpoly( binwidth=.001, position="identity" ) + 
-		ggtitle( "D" ) +
-		theme( legend.justification=c( 1, 1 ), legend.position=c( 1, 1 ) ) +
-		xlim( 0, 0.075 ) + 
-		theme_classic() +
-		annotate( "text", x = Inf, y = Inf, hjust="right", vjust="top", label=label_var ) +
-		xlab( "Phylogenetic independent contrast" ) +
-		ylab( "Density" ) +
-		theme(legend.position="none")
-
 	grid.arrange(
 		g_age,
-		make_contrast_plot( nodes_contrast_filtered_age ),
+		make_contrast_plot( nodes_contrast_filtered_age ) + ggtitle( "B" ),
 		g_var_exp,
-		make_contrast_plot( nodes_contrast_filtered_var ),
+		make_contrast_plot( nodes_contrast_filtered_var ) + ggtitle( "D" ),
 		ncol = 2

### 2017-05-04 manuscript.rmd
minor edits throughout, shortened abstract

@@
 
-# Pairwise comparisons are problematic when analyzing functional genomic data across species
+# Pairwise comparisons across species are problematic when analyzing functional genomic data
 
@@
 
-There is now considerable interest in comparing functional genomics data across species.
-One goal of this work is to provide an integrated understanding of genome and phenotype evolution.
-Most studies of this type have relied on multiple pairwise comparisons of functional genomic data between species, an approach that does not incorporate information about the evolutionary relationships among species.
+There is considerable interest in comparing functional genomic data across species.
+One goal of such work is to provide an integrated understanding of genome and phenotype evolution.
+Most comparative functional genomic studies have relied on multiple pairwise comparisons between species, an approach that does not incorporate information about the evolutionary relationships among species.
 The statistical problems that arise from not considering these relationships can lead pairwise approaches to the wrong conclusions, and are a missed opportunity to learn about biology that can only be understood in an explicit phylogenetic context.
-Here we examine two recently published studies that compare gene expression across species with pairwise methods.
-We find problems with both that call their conclusions into question.
-One study interpreted higher expression correlation between orthologs than paralogs as evidence for the ortholog conjecture, *i.e.* the hypothesis that gene function evolution is more rapid after duplication than speciation.
-Instead, we find that this pattern is due to the structure of the gene phylogenies, not different rates of expression evolution, and is the expected pattern when rates are the same following duplication and speciation.
-The second study interpreted pairwise comparisons of embryonic gene expression across distantly related animals as evidence for a distinct evolutionary process that gave rise to animal phyla.
-Instead, we find that the pattern they identified is due to unique features of a single species that impact multiple pairwise comparisons that include that species.
-In each study, distinct patterns of pairwise similarity among species were interpreted as evidence of particular evolutionary processes, but instead reflect the structure of phylogenies.
-These reanalyses concretely demonstrate the inadequacy of pairwise comparisons for analyzing functional genomic data across species, and indicate that it will be critical to adopt phylogenetic comparative methods in future functional genomics work.
+Here we examine two recently published studies that compare gene expression across species with pairwise methods, and find reason to question the original conclusions of both.
+One study interpreted pairwise comparisons of gene expression as support for the ortholog conjecture, the hypothesis that orthologs tend to be more similar than paralogs.
+The other study interpreted pairwise comparisons of embryonic gene expression across distantly related animals as evidence for a distinct evolutionary process that gave rise to phyla.
+In each study, distinct patterns of pairwise similarity among species were originally interpreted as evidence of particular evolutionary processes, but instead we find they reflect species relationships.
+These reanalyses concretely demonstrate the inadequacy of pairwise comparisons for analyzing functional genomic data across species.
+It will be critical to adopt phylogenetic comparative methods in future functional genomic work.
 Fortunately, phylogenetic comparative biology is also a rapidly advancing field with many methods that can be directly applied to functional genomic data.
@@
 
-The focus of genomics research has quickly shifted from describing genome sequences to functional genomics, the study of how genomes "work" using tools that measure functional attributes such as expression, chromatin state, and transcription initiation.
+The focus of genomic research has quickly shifted from describing genome sequences to functional genomics, the study of how genomes "work" using tools that measure functional attributes such as expression, chromatin state, and transcription initiation.
 Functional genomics, in turn, is now becoming more comparative-- there is great interest in understanding how functional genomic variation across species gives rise to a diversity of development, morphology, physiology, and other phenotypes [@Wray:2013gh].
@@
 
-A rich theoretical and statistical methodology of phylogenetic comparative methods has been developed over the last three decades to address the challenges and opportunities of trait comparisons across species [@Felsenstein:1985ua; @Grafen:1989um; @Pagel:1999gc; Revell:2012jd; @FitzJohn:2012ey; @Uyeda:2014jl; @Garamszegi:2014ew].
+Over the last three decades, a rich set of phylogenetic comparative methods has been developed to address the challenges and opportunities of trait comparisons across species [@Felsenstein:1985ua; @Grafen:1989um; @Pagel:1999gc; Revell:2012jd; @FitzJohn:2012ey; @Uyeda:2014jl; @Garamszegi:2014ew].
 A central challenge is the dependence of observations across species due to the evolutionary history of species -- more closely related species share many traits that evolved once in a common ancestor.
@@
 They have largely been applied to morphological and ecological traits, but are just as relevant to functional genomics [@Barker:2005be].
-Even so, most comparative functional genomics studies have abstained from phylogenetic approaches and instead rely on multiple pairwise comparisons across species (Figure 1*A*).
-This leaves comparative functional genomics studies susceptible to statistical problems and is a missed opportunity to ask questions that are only accessible in an explicit phylogenetic context.
+Even so, most comparative functional genomic studies have abstained from phylogenetic approaches and instead rely on multiple pairwise comparisons across species (Figure 1*A*).
+This leaves comparative functional genomic studies susceptible to statistical problems and is a missed opportunity to ask questions that are only accessible in an explicit phylogenetic context.
 
@@
 Phylogenetic comparative methods capture this shift as change along a single branch deep in the tree, revealing that there is not a tendency of correlated change between these traits across the phylogeny.
-One reason that comparative functional genomic sudies have not embraced phylogenetic approaches is that it has not yet been concretely demonstrated that pairwise and phylogenetic comparative methods can lead to different results when considering functional genomics data.
+One reason that comparative functional genomic sudies have not embraced phylogenetic approaches is that there has not yet been a concrete demonstration that pairwise and phylogenetic comparative methods can lead to different results when considering functional genomic data.
 Here we examine this issue by re-evaluating the pairwise comparisons in two recent studies that compared gene expression across species.
@@
 
-The second comparative functional genomics study we evaluate here is Levin *et al.* [@Levin:2016bp].
+The second comparative functional genomic study we evaluate here is Levin *et al.* [@Levin:2016bp].
 This study analyzed gene expression through the course of embryonic development for ten animal species, each from a different phylum.
@@
 It is also used to relate findings from well-studied genes to related genes that have not been investigated in detail.
-It can be applied to any trait of genes, from gene sequence to biochemical properties to expression.
+It has be applied to many trait of genes, from gene sequence to biochemical properties to expression.
 While the ortholog conjecture describes a specific pattern of functional diversity across genes, it is also articulated as a hypothesis about the process of evolution-- that there is greater evolutionary change in gene attributes following a duplication event than a speciation event.
@@
 Their expression summary statistic is Tau [@Yanai:2005dh, @KryuchkovaMostacci:2016in], an indicator of tissue specificity of gene expression.
-It can range from a value of 0, which indicates no specificity (*i.e.*, uniform expression across tissues), to a value of 1, which indicates high specificity (*i.e.*, expression in only one tissue).
+Tau can range from a value of 0, which indicates no specificity (*i.e.*, uniform expression across tissues), to a value of 1, which indicates high specificity (*i.e.*, expression in only one tissue).
 Tau is convenient in that it is a single number of defined range for each gene, though of course since the original expression is multidimensional this means much information is discarded.
@@
 
-They found the correlation coefficient of Tau for orthologs to be significantly greater than the correlation coefficient of Tau for paralogs, *i.e.* that orthologs tend to be more similar to each other than paralogs.
+They found the correlation coefficient of Tau for orthologs to be significantly greater than the correlation coefficient of Tau for paralogs, *i.e.* orthologs tend to have more similar expression than do paralogs.
 From this they concluded that their analyses support the ortholog conjecture.
@@
 
-For each internal node in each gene tree, we calculated the phylogenetic independent contrast (PIC) of Tau.
+For each internal node in each gene tree, we calculated the phylogenetic independent contrast[@Felsenstein:1985ua] (PIC) of Tau.
 This is the difference in values of Tau for descendant nodes scaled by the expected variance, which is largely determined by the lengths of the two branches that connect the node to its two descendants (Figure 1*B*).
@@
 This is important since we are only looking at a subset of the data they considered, the Brawand *et al.* 2011 [@Brawand:2011du] dataset for gene trees that could be successfully time calibrated.
-In their Figure 1, they present a higher Tau correlation coefficient between ortholog pairs than paralog pairs.
+In their Figure 1[@KryuchkovaMostacci:2016iw], they present a higher Tau correlation coefficient between ortholog pairs than paralog pairs.
 We find the same here, with correlation coefficients of `r signif( ortholog_r, 2)` for orthologs and `r signif( paralog_r, 2)` for paralogs.
@@
 One reason is that multiple pairwise comparisons repeatedly sample the same evolutionary changes and in so doing violate statistical assumptions of independence, whereas phylogenetic comparative methods make multiple independent comparisons across non-overlapping branches of the tree.
-The other reason for the different conclusions is that pairwise comparisons and phylogenetic comparative methods describe different things.
+The other reason is that pairwise comparisons and phylogenetic comparative methods describe different things.
 Pairwise comparisons describe contemporary patterns, while phylogenetic methods infer historical processes [@OMeara:2012gq].
@@
 The KMRR study did investigate the impact of node age on correlation, but in a different way.
-In their Figure 2, they grouped orthologs and paralogs according to the ENSEMBL node name of their most recent common ancestor, and plotted the correlation of Tau for each of these groups by the node age.
+In their Figure 2[@KryuchkovaMostacci:2016iw], they grouped orthologs and paralogs according to the ENSEMBL node name of their most recent common ancestor, and plotted the correlation of Tau for each of these groups by the node age.
 They found that across the investigated range of node ages, ortholog pairs have higher Tau correlation than paralogs.
@@
 As in the original study, there is higher correlation coefficient across orthologs (`r signif( ortholog_r_sim_null, 2 )`) than paralogs (`r signif( paralog_r_sim_null, 2 )`) when not considering node age.
-Phylogenetic analysis of the data simulated under the null hypothesis (Figure 2*D*) do not reject the null hypothesis (Wilcoxon p = $`r wilcox_oc( nodes_sim_null_contrast )`$), as expected.
+Phylogenetic analysis of the data simulated under the null hypothesis (Figure 2*D*) do not reject the null hypothesis (Wilcoxon p = $`r wilcox_oc( nodes_sim_null_contrast ) %>% round( 3 )`$), as expected.
 
@@
 These simulations have several implications.
-The pairwise comparisons used by KMRR cannot distinguish between the null hypothesis and ortholog conjecture even when the evolution of expression is `r dup_adjust` fold higher following duplication than speciation.
-The pairwise results are strikingly similar under both hypotheses.
-These simulations also serve as a validation of phylogenetic methods as applied to this problem.
-Our phylogenetic analysis of independent contrasts does not reject the null hypothesis when data are simulated under the null model, and does reject the null hypothesis when the data are simulated under the ortholog conjecture.
+The pairwise comparisons used by KMRR cannot distinguish between the null hypothesis and ortholog conjecture.
+The pairwise results are strikingly similar under both hypotheses (Figure 2*C*, 2*E*).
+These simulations also serve to validate the phylogenetic methods applied to this problem.
+As expected, our phylogenetic analysis of independent contrasts does not reject the null hypothesis when data are simulated under the null model, and does reject the null hypothesis when the data are simulated under the ortholog conjecture.
 In contrast to pairwise methods, the phylogenetic analyses can test explicit predictions based on hypotheses about evolutionary process.
@@
 Of the `r n_dup_original` duplication nodes considered here, `r n_dup_remaining` have a node age (as determined by the time calibrated trees) within `r age_window * 100`% of their clade annotation.
-If only these duplication events closest to their annotated clades are retained, the pattern of lower correlation of Tau evolution for duplication events disappears for the dataset simulated under the null model (*SI Appendix*, Figure 5*B*).
+If only these duplication events closest to their annotated clade nodes are retained, the pattern of lower correlation of Tau evolution for duplication events disappears for the dataset simulated under the null model (*SI Appendix*, Figure 5*B*).
 This correction, though, comes at the high cost of discarding `r round( (n_dup_original - n_dup_remaining )/ n_dup_original * 100, 1 )`% of duplication nodes.
@@
 It indicates that the mechanism of gene divergence (speciation versus duplication) may not have as strong an impact on phenotypic divergence as sometimes proposed.
-It also has technical implications.
+It also has important technical implications.
 Having information on whether two genes are orthologs or paralogs provides little added information about expression beyond knowing how distantly related the two genes are.
@@
 They then calculated a similarity score for each temporal phase for each species pair based on the fraction of genes that exhibited the same patterns in each species.
-The distributions of similarity scores are plotted in their Figure 4d, and their Kolmogorov–Smirnov (KS) tests indicated that the early distribution and late distribution were each significantly different from mid distribution (P < $10^{-6}$ and P < $10^{-12}$, respectively).
+The distributions of similarity scores are plotted in their Figure 4d[@Levin:2016bp], and their Kolmogorov–Smirnov (KS) tests indicated that the early distribution and late distribution were each significantly different from mid distribution (P < $10^{-6}$ and P < $10^{-12}$, respectively).
 This is the support they presented for the inverse hourglass model.
@@
 
-We examined the matrix of pairwise comparisons used as the base for the KS tests and Figure 4d in *Levin et al.* [@Levin:2016bp], and thus as evidence to support the "inverse hourglass" model.
+We examined the matrix of pairwise comparisons used as the base for the KS tests and Figure 4d in *Levin et al.* [@Levin:2016bp], and thus as support for the "inverse hourglass" model.
 We found several problems resulting from the use of multiple pairwise comparisons.
@@
 Our results highlight the importance of explicitly incorporating information about phylogenetic relationships when comparing functional genomic traits across species.
-Some of the most widely used phylogenetic comparative methods [@Felsenstein:1985ua; @Grafen:1989um] are already directly applicable to comparative functional genomics studies.
-There are also interesting new challenges at this interdisciplinary interface that will need to be addressed to fully realize the potential of phylogenetic comparative functional genomics studies.
+Some of the most widely used phylogenetic comparative methods [@Felsenstein:1985ua; @Grafen:1989um] are already directly applicable to comparative functional genomic studies.
+There are also interesting new challenges at this interdisciplinary interface that will need to be addressed to fully realize the potential of phylogenetic comparative functional genomic studies.
 One such challenge is that most phylogenetic comparative analyses of covariance between traits have been developed to address problems with many more species (e.g., dozens or more) relative to the number of traits being examined.
-In comparative functional genomics analyses, there are often far fewer species because adding taxa is still expensive, but tens of thousands of traits.
+In comparative functional genomic analyses, there are often far fewer species because adding each species is still expensive, but high-throughput tools generate data for tens of thousands of traits per species.
 This creates statistical challenges as the resulting covariance matrices are singular and, if not treated appropriately, imply many false correlations that are artifacts of project design.
@@
 Most early phylogenetic comparative methods attempted to account for evolutionary signal to correct statistical tests for correlations between traits, while more recent methods tend to focus on testing hypotheses of evolutionary processes [@Pennell:2013cw].
-The application of this newer focus to functional genomics provides an exciting opportunity to address long standing questions of broad interest, including the order of changes in functional genomics traits and shifts in rates of evolution of one functional genomics trait following changes in another trait.
+The application of this newer focus to functional genomics provides an exciting opportunity to address long standing questions of broad interest, including the order of changes in functional genomic traits and shifts in rates of evolution of one functional genomic trait following changes in another trait.
 
 We are not the first to apply phylogenetic comparative methods to functional genomic data.
-While the vast majority of comparative functional genomics studies have used standard pairwise similarity methods, a small number of comparative functional genomics studies have employed phylogenetic comparative approaches [@Oakley:2005ky; @Eng:2009gp; @Chang:2014hq].
+While the vast majority of comparative functional genomic studies have used standard pairwise similarity methods, a small number of comparative functional genomic studies have employed phylogenetic comparative approaches [@Oakley:2005ky; @Eng:2009gp; @Chang:2014hq].
 For instance, a phylogenetic ANOVA [@Rohlfs:2015bd] of the evolution of gene expression improves statistical power and drastically reduces the rate of false positives relative to pairwise approaches.
 
-Addressing the statistical dependence of pairwise comparisons is not the only advantage of using phylogenetic comparative methods for functional genomics analyses.
+Addressing the statistical dependence of pairwise comparisons is not the only advantage of using phylogenetic comparative methods for functional genomic analyses.
 Another problem with the pairwise comparisons is that, except at the tips, they summarize changes along many branches in the phylogeny.
@@
 Phylogenetic methods have the advantage of isolating the changes under consideration (Figure 1*B*).
-The phylogenetic methods avoid diluting the change that occurs along the branches that follow the node in question with changes along all subsequent branches.
+Phylogenetic methods therefore avoid diluting the change that occurs along the branches that follow the node in question with changes along all subsequent branches.
 There may still be missing speciation events, due to extinction and incomplete taxon sampling, and missing duplication events, due to gene loss, but these omissions affect both methods.
@@
 
-The fact that the first two comparative functional genomics studies we reanalyzed show serious problems with pairwise comparisons indicates that there are likely to be similar problems in other studies that use these methods, and that future comparative studies will be compromised if they continue to use pairwise methods.
+The fact that the first two comparative functional genomic studies we reanalyzed show serious problems with pairwise comparisons indicates that there are likely to be similar problems in other studies that use these methods.
+Future studies that compare functional genomic data across species will be compromised if they continue to use pairwise methods.
 Studies of evolutionary functional genomics should not be focused on the tips of the tree using pairwise comparisons.
 They should explicitly delve into the tree with phylogenetic comparative methods.
-Here we show a concrete case where the predictions of a null hypotheses and an alternatives hypothesis do not differ when considering pairwise comparisons, but do differ with phylogenetic comparative methods.
 
@@
 To use patterns across living species to test hypotheses about evolutionary processes it is also necessary to incorporate information about evolutionary relationships, *i.e.* phylogenies.
-There have been decades of work on building comparative phylogenetic methods that do exactly that, and they are just as relevant to comparing functional genomics traits across species as they are to comparing morphology or any of the other traits they are already routinely applied to.
+There have been decades of work on building comparative phylogenetic methods that do exactly that, and they are just as relevant to comparing functional genomic traits across species as they are to comparing morphology or any of the other traits they are already routinely applied to.
 
@@
 The most recent commit at the time of the analysis presented here was `r system("git log | head -n 1", intern=TRUE) %>% str_replace("commit ", "")`.
-See the `readme.md` file in this repository for more information on the contents of the source file and how to re-execute them.
 
@@
 This modified script and their data files are available in the github repository for this paper, as are the intermediate files that were generated by their analysis script that we used in our own analyses.
-We also obtained the `Compara.75.protein.nh.emf` gene trees [@Herrero:2016jj] from ftp://ftp.ensembl.org/pub/release-75/emf/ensembl-compara/homologies/ and place them in the same directory as this file.
-These gene trees include branch lengths and annotate each internal node as being a duplication or speciation event.
+We\ obtained the `Compara.75.protein.nh.emf` gene trees [@Herrero:2016jj] from ftp://ftp.ensembl.org/pub/release-75/emf/ensembl-compara/homologies/ and include them in our github repository.
+These gene trees include branch lengths, annotate each internal node as being a duplication or speciation event, and provide a clade label for each internal node.
 
@@
 We left in sex chromosome genes and testes expression data, which KMRR removed in some of their sensitivity analyses.
-This corresponded to the analyses that provided the strongest support for the ortholog conjecture and therefore the most conservative reconsideration of it.
+This corresponded to the KMRR analyses that provided the strongest support for the ortholog conjecture and therefore the most conservative reconsideration of it.
 
@@
 The gene trees were then time calibrated.
-The goal is not necessarily to have precise dates for each node, but to scale branch lengths so that they are equivalent across the trees.
+The goal is not necessarily to have precise dates for each node, but to scale branch lengths so that they are equivalent across gene trees.
 This in turn scales the phylogenetic independent contrasts (which take branch length into account) so they can be compared appropriately.
@@
 To test for such an effect, we remove duplication nodes that are older than the oldest speciation node.
-Our results are unchanged and this reduced dataset does not reject the null hypothesis that the rate of evolution following duplications is the same as or less than the rate following speciation (*SI Appendix*, Figure 5*B*).
+The general results are unchanged and this reduced dataset does not reject the null hypothesis that the rate of evolution following duplications is the same as or less than the rate following speciation (*SI Appendix*, Figure 5*B*).
 
@@
 There is, however, a negative relationship between Tau and maximum expression across all genes (*SI Appendix* Figure 7, $r^2=`r signif( tau_v_max_lm$r.squared, 3 )`$, $slope=`r signif( tau_v_max_lm$coefficients[2,1], 3 )`$, $p=`r signif( tau_v_max_lm$coefficients[2,4], 3 )`$).
-A simpler explanation for this plot is that it reflects the global tendency for Tau to decrease with maximum expression.
+A simpler explanation for their plot is that it reflects the global tendency for Tau to decrease with maximum expression.
 The null expectation for any two genes sampled at random is for one to have higher maximum expression and lower Tau, and the other to have lower maximum expression and higher Tau.

### 2017-05-04 manuscript.rmd
added significance statement

@@
 bibliography: manuscript.bib
-csl: nature.csl
+csl: pnas.csl
 output:
@@
 
+## Significance
+
+Comparisons of genome function between species are providing important insight into the evolutionary origins of diversity.
+Here we demonstrate that comparative functional genomics studies can come to the wrong conclusions if they do not take the relationships of species into account and instead rely on pairwise comparisons between species, as is common practice.
+We re-examined two previously published studies and found problems with pairwise comparisons that draw both their original conclusions into question.
+One study had found support for the ortholog conjecture and the other had concluded that the evolution of gene expression was different between animal phyla than within them.
+Our results demonstrate that to answer evolutionary questions about genome function, it is critical to consider evolutionary relationships.
+
+\pagebreak
+
 ## Introduction
@@
 These analyses are also critical to transferring functional insight across species, and will grow in importance in coming years.
+
+![Pairwise and phylogenetic comparative approaches, illustrated on an example gene tree with multiple genes per species.
+The internal nodes of the tree are speciation and gene duplication events.
+(*A*) Many comparative functional genomic studies rely on pairwise comparisons, where traits of each gene are compared to traits of other genes across species.
+This leads to many more comparisons than unique observations, making each comparison dependent on others.
+(*B*) Comparative phylogenetic methods, including phylogenetic independent contrasts [@Felsenstein:1985ua], make a smaller number of independent comparisons, where each contrast measures independent changes along different branches.
+Phylogenetic approaches are rarely used for functional genomic studies.](Figure_overview.pdf)
 
@@
 Here we address the analyses themselves by examining the structure of the pairwise comparisons.
-
-![Pairwise and phylogenetic comparative approaches, illustrated on an example gene tree with multiple genes per species.
-The internal nodes of the tree are speciation and gene duplication events.
-(*A*) Many comparative functional genomic studies rely on pairwise comparisons, where traits of each gene are compared to traits of other genes across species.
-This leads to many more comparisons than unique observations, making each comparison dependent on others.
-(*B*) Comparative phylogenetic methods, including phylogenetic independent contrasts [@Felsenstein:1985ua], make a smaller number of independent comparisons, where each contrast measures independent changes along different branches.
-Phylogenetic approaches are rarely used for functional genomic studies.](Figure_overview.pdf)
 

### 2017-05-05 manuscript.rmd
removed hyphen in title
added space before citations

@@
 
+\hyphenation{analyzing}
+
 # Pairwise comparisons across species are problematic when analyzing functional genomic data
@@
 
-For each internal node in each gene tree, we calculated the phylogenetic independent contrast[@Felsenstein:1985ua] (PIC) of Tau.
+For each internal node in each gene tree, we calculated the phylogenetic independent contrast [@Felsenstein:1985ua] (PIC) of Tau.
 This is the difference in values of Tau for descendant nodes scaled by the expected variance, which is largely determined by the lengths of the two branches that connect the node to its two descendants (Figure 1*B*).
@@
 This is important since we are only looking at a subset of the data they considered, the Brawand *et al.* 2011 [@Brawand:2011du] dataset for gene trees that could be successfully time calibrated.
-In their Figure 1[@KryuchkovaMostacci:2016iw], they present a higher Tau correlation coefficient between ortholog pairs than paralog pairs.
+In their Figure 1 [@KryuchkovaMostacci:2016iw], they present a higher Tau correlation coefficient between ortholog pairs than paralog pairs.
 We find the same here, with correlation coefficients of `r signif( ortholog_r, 2)` for orthologs and `r signif( paralog_r, 2)` for paralogs.
@@
 The KMRR study did investigate the impact of node age on correlation, but in a different way.
-In their Figure 2[@KryuchkovaMostacci:2016iw], they grouped orthologs and paralogs according to the ENSEMBL node name of their most recent common ancestor, and plotted the correlation of Tau for each of these groups by the node age.
+In their Figure 2 [@KryuchkovaMostacci:2016iw], they grouped orthologs and paralogs according to the ENSEMBL node name of their most recent common ancestor, and plotted the correlation of Tau for each of these groups by the node age.
 They found that across the investigated range of node ages, ortholog pairs have higher Tau correlation than paralogs.
@@
 They then calculated a similarity score for each temporal phase for each species pair based on the fraction of genes that exhibited the same patterns in each species.
-The distributions of similarity scores are plotted in their Figure 4d[@Levin:2016bp], and their Kolmogorov–Smirnov (KS) tests indicated that the early distribution and late distribution were each significantly different from mid distribution (P < $10^{-6}$ and P < $10^{-12}$, respectively).
+The distributions of similarity scores are plotted in their Figure 4d [@Levin:2016bp], and their Kolmogorov–Smirnov (KS) tests indicated that the early distribution and late distribution were each significantly different from mid distribution (P < $10^{-6}$ and P < $10^{-12}$, respectively).
 This is the support they presented for the inverse hourglass model.
@@
 ```{r cor_structure_plot, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, dev="png", dpi=600, fig.width=6.5, fig.height=6.5, fig.cap="Origin of pairwise correlation plot structure.
-(*A*) Each speciation and duplication node in the COMPARA trees is annotated with a clade name, and the age of that clade (shown on the x axis) is used as the age of the pairwise comparisons (Figure 2 *A, C, E*, KMRR[@KryuchkovaMostacci:2016iw] Figure 2).
+(*A*) Each speciation and duplication node in the COMPARA trees is annotated with a clade name, and the age of that clade (shown on the x axis) is used as the age of the pairwise comparisons (Figure 2 *A, C, E*, KMRR [@KryuchkovaMostacci:2016iw] Figure 2).
 The clade ages have a clear correspondence to the speciation node ages (red).
@@
 
-KMRR[@KryuchkovaMostacci:2016iw] Figure 3 differs from their other analyses in that it presents independent changes in expression between triplets of genes.
+KMRR [@KryuchkovaMostacci:2016iw] Figure 3 differs from their other analyses in that it presents independent changes in expression between triplets of genes.
 Each triplet has two paralogs that arose from a duplication event and one unduplicated homolog.

### 2017-06-09 manuscript.rmd
updated author addresses

@@
 
-Casey W. Dunn^1^*, Felipe Zapata^1,2^, Catriona Munro^1^, Stefan Siebert^1,3^, Andreas Hejnol^4^
+Casey W. Dunn^1^*, Felipe Zapata^2^, Catriona Munro^1^, Stefan Siebert^3^, Andreas Hejnol^4^
 
@@
 
-^2^ Current address: Department of Ecology and Evolutionary Biology, University of California Los Angeles, Los Angeles, CA, USA
-
-^3^ Current address: Department of Molecular & Cellular Biology, University of California at Davis, Davis, CA, USA
+^2^ Department of Ecology and Evolutionary Biology, University of California Los Angeles, Los Angeles, CA, USA
+
+^3^ Department of Molecular & Cellular Biology, University of California at Davis, Davis, CA, USA
 

### 2017-06-10 manuscript.rmd
updated my address

@@
 
-Casey W. Dunn^1^*, Felipe Zapata^2^, Catriona Munro^1^, Stefan Siebert^3^, Andreas Hejnol^4^
+Casey W. Dunn^1,2^*, Felipe Zapata^3^, Catriona Munro^1^, Stefan Siebert^4^, Andreas Hejnol^5^
 
@@
 
-^2^ Department of Ecology and Evolutionary Biology, University of California Los Angeles, Los Angeles, CA, USA
-
-^3^ Department of Molecular & Cellular Biology, University of California at Davis, Davis, CA, USA
-
-^4^ Sars International Centre for Marine Molecular Biology, University of Bergen, Bergen, Norway
-
-\* Corresponding author, casey_dunn@brown.edu
+^2^ Current address: Department of Ecology and Evolutionary Biology, Yale University, New Haven, CT, USA
+
+^3^ Department of Ecology and Evolutionary Biology, University of California Los Angeles, Los Angeles, CA, USA
+
+^4^ Department of Molecular & Cellular Biology, University of California at Davis, Davis, CA, USA
+
+^5^ Sars International Centre for Marine Molecular Biology, University of Bergen, Bergen, Norway
+
+\* Corresponding author, casey.dunn@yale.edu
 

### 2017-07-16 manuscript.rmd
updated R and all libraries to most recent versions

@@
     fig_caption: yes
+    fig_height: 4
     fig_width: 6
-    fig_height: 4
-
+  word_document: default
 ---
@@
 
+#### Impact and importance
+
 Our results highlight the importance of explicitly incorporating information about phylogenetic relationships when comparing functional genomic traits across species.
-Some of the most widely used phylogenetic comparative methods [@Felsenstein:1985ua; @Grafen:1989um] are already directly applicable to comparative functional genomic studies.
-There are also interesting new challenges at this interdisciplinary interface that will need to be addressed to fully realize the potential of phylogenetic comparative functional genomic studies.
-One such challenge is that most phylogenetic comparative analyses of covariance between traits have been developed to address problems with many more species (e.g., dozens or more) relative to the number of traits being examined.
-In comparative functional genomic analyses, there are often far fewer species because adding each species is still expensive, but high-throughput tools generate data for tens of thousands of traits per species.
-This creates statistical challenges as the resulting covariance matrices are singular and, if not treated appropriately, imply many false correlations that are artifacts of project design.
-We outlined these challenges and potential solutions in the context of gene expression elsewhere [@Dunn:2013cpa].
-In the same manuscript we also considered another issue of relevance here -- the read counts generated by RNA-seq expression studies cannot be directly compared across species.
-This is because there are various species-specific technical factors that can be mistaken for differences in expression across species.
-These can be canceled out within species before making comparisons across species [@Dunn:2013cpa].
-
-Recent advances in phylogenetic comparative methods are particularly well suited to addressing questions about the evolution of functional genomic traits.
-Most early phylogenetic comparative methods attempted to account for evolutionary signal to correct statistical tests for correlations between traits, while more recent methods tend to focus on testing hypotheses of evolutionary processes [@Pennell:2013cw].
-The application of this newer focus to functional genomics provides an exciting opportunity to address long standing questions of broad interest, including the order of changes in functional genomic traits and shifts in rates of evolution of one functional genomic trait following changes in another trait.
-
-We are not the first to apply phylogenetic comparative methods to functional genomic data.
-While the vast majority of comparative functional genomic studies have used standard pairwise similarity methods, a small number of comparative functional genomic studies have employed phylogenetic comparative approaches [@Oakley:2005ky; @Eng:2009gp; @Chang:2014hq].
-For instance, a phylogenetic ANOVA [@Rohlfs:2015bd] of the evolution of gene expression improves statistical power and drastically reduces the rate of false positives relative to pairwise approaches.
 
@@
 There may still be missing speciation events, due to extinction and incomplete taxon sampling, and missing duplication events, due to gene loss, but these omissions affect both methods.
+
+#### Challenges and opportunities
+
+There are interesting new challenges at the interdisciplinary interface between phylogenetic comparative methods and functional genomics that will need to be addressed to fully realize the potential of integrating these approaches.
+One such challenge is that most phylogenetic comparative analyses of covariance between traits have been developed to address problems with many more species (e.g., dozens or more) relative to the number of traits being examined.
+In comparative functional genomic analyses, there are often far fewer species because adding each species is still expensive, but high-throughput tools generate data for tens of thousands of traits per species.
+This creates statistical challenges as the resulting covariance matrices are singular and, if not treated appropriately, imply many false correlations that are artifacts of project design.
+We outlined these challenges and potential solutions in the context of gene expression elsewhere [@Dunn:2013cpa].
+
+Another challenge is parameterizing differential expression in such a way that they can be analyzed in a phylogenetic context.
+Read counts generated by RNA-seq expression studies cannot be directly compared across species [@Dunn:2013cpa].
+This is because there are various species-specific technical factors that can be mistaken for differences in expression across species.
+These can be canceled out within species before making comparisons across species [@Dunn:2013cpa].
+
+Branch lengths are fundamental to phylogenetic comparative methods, since the null expectation is that more changes are expected along longer branches.
+In this study we applied three separate modifications of branch lengths.
+First, we time calibrated the gene trees so that nodes corresponding to the same speciation events had the same age across all trees.
+This is critical to making the comparisons of funcitonal genomic traits equivalent across genes.
+Second, phyogenetic independent contrasts adjust branch lengths so th
+
+#### Additional approaches and future directions
+
+We are not the first to apply phylogenetic comparative methods to functional genomic data.
+While the vast majority of comparative functional genomic studies have used standard pairwise similarity methods, a small number of comparative functional genomic studies have employed phylogenetic comparative approaches [@Oakley:2005ky; @Eng:2009gp; @Chang:2014hq].
+For instance, a phylogenetic ANOVA [@Rohlfs:2015bd] of the evolution of gene expression improves statistical power and drastically reduces the rate of false positives relative to pairwise approaches.
+
+Some of the most widely used phylogenetic comparative methods [@Felsenstein:1985ua; @Grafen:1989um] are already directly applicable to comparative functional genomic studies.
+
+Recent advances in phylogenetic comparative methods are particularly well suited to addressing questions about the evolution of functional genomic traits.
+Most early phylogenetic comparative methods attempted to account for evolutionary signal to correct statistical tests for correlations between traits, while more recent methods tend to focus on testing hypotheses of evolutionary processes [@Pennell:2013cw].
+The application of this newer focus to functional genomics provides an exciting opportunity to address long standing questions of broad interest, including the order of changes in functional genomic traits and shifts in rates of evolution of one functional genomic trait following changes in another trait.
 

### 2017-07-17 manuscript.rmd
added within species variance analyses

@@
 We added random noise to the calibration times in replicate analyses, and all still failed to reject the null hypothesis (*SI Appendix*).
+
+An additional concern is that within-species expression variation is not considered when making comparisons of tau across homologous genes.
+Neglecting within species variation is expected to mislead phylogenetic comparative analyses in some cases [@Ives:2007cw; Felsenstein:2008cl; Rohlfs:2015bd].
+Since branch length describes expected trait variation [@Felsenstein:1985ua], within species variation can be modeled by extending the terminal branches of each gene phylogeny [@Ives:2007cw].
+We modeled within-species variation by extending each terminal branch by 7 million years.
+This value was selected because it corresponds to the age of the clade Hominini (defined by the most recent common ancestor of humans and chimpanzees), the shallowest node in the phylogeny of examined species.
+This provides an opportunity to assess the impact that extreme within-species variation (*i.e.*, as large as the variance between humans and chimpanzees and their most recent common ancestor) would have on our findings.
+These analyses still do not reject the null hypothesis, and do not provide support for the ortholog conjecture in the presence of high within-species variance (*SI Appendix*, Figure 6).
 
@@
 Neighboring clade ages are therefore a very rough approximation of the age of duplication events.
-This is apparent when node ages derived from calibrated trees are plotted against the age of the clade annotations for each node (*SI Appendix*, Figure 6*A*).
+This is apparent when node ages derived from calibrated trees are plotted against the age of the clade annotations for each node (*SI Appendix*, Figure 7*A*).
 Because duplication events that are all annotated with the same clade name can occur at very different times, pairwise comparisons across these nodes capture evolutionary changes in Tau across very different different branch lengths.
@@
 Of the `r n_dup_original` duplication nodes considered here, `r n_dup_remaining` have a node age (as determined by the time calibrated trees) within `r age_window * 100`% of their clade annotation.
-If only these duplication events closest to their annotated clade nodes are retained, the pattern of lower correlation of Tau evolution for duplication events disappears for the dataset simulated under the null model (*SI Appendix*, Figure 5*B*).
+If only these duplication events closest to their annotated clade nodes are retained, the pattern of lower correlation of Tau evolution for duplication events disappears for the dataset simulated under the null model (*SI Appendix*, Figure 7*B*).
 This correction, though, comes at the high cost of discarding `r round( (n_dup_original - n_dup_remaining )/ n_dup_original * 100, 1 )`% of duplication nodes.
@@
 
+```{r within_species_variation, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, dev="png", dpi=600, fig.width=6.5, fig.height=3, fig.cap="Investigation of within-species variation.
+These plots correspond to Fig.
+2*A* and 2*B*, but with increased expected within-species variation modeled by increasing the terminal gene phylogeny branch lengths.
+The results are unchanged.
+(*A*) Pairwise comparisons still recover higher correlation across speciation events than duplication events.
+(*B*) There is not an excess of larger contrasts for duplication events and the analysis does not reject the null hypothesis.
+Modeling a large amount of within-species variation does not lead to support for the ortholog conjecture."}
+
+	x = calibration_times$age[calibration_times$clade == "Hominini"]
+	gene_trees_extended = mclapply( gene_trees_calibrated, extend_nhx, x=x, mc.cores=cores )
+	gene_trees_extended %<>% add_pics_to_trees()
+	nodes_extended_contrast = gene_trees_extended %>% summarize_contrasts()
+
+	pairwise_extended_summary = 
+			mclapply( gene_trees_extended, get_pairwise_summary, mc.cores=cores ) %>% 
+			bind_rows()
+
+	g_pair_extended = 
+		pairwise_extended_summary %>% 
+			filter( ! is.na( D ) ) %>%
+			group_by( Event, label ) %>% 
+			summarise( cor.test( tau_a, tau_b )$estimate, n=n() ) %>% 
+			rename( rho=`cor.test(tau_a, tau_b)$estimate` ) %>%
+			rename( clade = label ) %>% 
+			left_join( calibration_times, clade=clade ) %>%
+			ggplot() + 
+				geom_point( aes( x=age, y=rho, col=Event ), size=3 ) +
+				theme_classic() +
+				xlab( "Clade Age (million years)" ) +
+				ylab( "Correlation Coefficient" ) +
+				ylim( 0, 1 ) +
+				ggtitle( "A Pairwise" ) +
+				guides( colour = guide_legend( override.aes = list( shape = 15 ) ) ) +
+				theme( 
+				legend.justification=c( 1, 1 ), 
+				legend.position=c( 1, 1 ), 
+				legend.title=element_blank() )
+
+	grid.arrange(
+		g_pair_extended,
+		make_contrast_plot( nodes_extended_contrast ) + 
+			ggtitle( "B Phylogenetic" ),
+		ncol = 2
+	)
+```
+
 #### Investigation of sensitivity to calibration times
@@
 From this they conclude that following duplication there is a common trajectory of evolutionary change, where one paralog evolves to have lower expression and also become more tissue specific.
-There is, however, a negative relationship between Tau and maximum expression across all genes (*SI Appendix* Figure 7, $r^2=`r signif( tau_v_max_lm$r.squared, 3 )`$, $slope=`r signif( tau_v_max_lm$coefficients[2,1], 3 )`$, $p=`r signif( tau_v_max_lm$coefficients[2,4], 3 )`$).
+There is, however, a negative relationship between Tau and maximum expression across all genes (*SI Appendix* Figure 8, $r^2=`r signif( tau_v_max_lm$r.squared, 3 )`$, $slope=`r signif( tau_v_max_lm$coefficients[2,1], 3 )`$, $p=`r signif( tau_v_max_lm$coefficients[2,4], 3 )`$).
 A simpler explanation for their plot is that it reflects the global tendency for Tau to decrease with maximum expression.

### 2017-07-17 functions.R

@@ -590,3 +590,14 @@ make_contrast_plot = function( nodes_contrast, mode="diff" ){
 	ggcontrasts
 	
 }
+
+#' Extends each terminal branch by specified length
+#' 
+#' @param nhx A phylogenetic tree as a treeio::treedata object
+#' @param x Amount to extend each branch by
+#' @return A phylogenetic tree as a treeio::treedata object
+#' @export
+extend_nhx = function( nhx, x ){
+  nhx@phylo = hutan::extend_terminal_branches( nhx@phylo, x )
+  return( nhx )
+}

### 2017-07-18 manuscript.rmd
light edits throughout

@@
 
-Over the last three decades, a rich set of phylogenetic comparative methods has been developed to address the challenges and opportunities of trait comparisons across species [@Felsenstein:1985ua; @Grafen:1989um; @Pagel:1999gc; Revell:2012jd; @FitzJohn:2012ey; @Uyeda:2014jl; @Garamszegi:2014ew].
-A central challenge is the dependence of observations across species due to the evolutionary history of species -- more closely related species share many traits that evolved once in a common ancestor.
+Over the last three decades, a rich set of phylogenetic comparative methods has been developed to address the challenges and opportunities of trait comparisons across species [@Felsenstein:1985ua; @Grafen:1989um; @Pagel:1999gc; @Revell:2012jd; @FitzJohn:2012ey; @Uyeda:2014jl; @Garamszegi:2014ew; @Jhwueng:2013fi].
+A central challenge is the dependence of observations across species due to the evolutionary history of species -- more closely related species share more traits because they have a more recent common ancestor.
 This violates the fundamental assumption of observation independence in standard statistical methods.
@@
 Phylogenetic comparative methods account for evolutionary history and explicitly model trait change along the branches of evolutionary trees (*e.g.*, Figure 1*B*).
-The value of these methods relative to to pairwise comparisons has been repeatedly shown in analyses of other types of character data [@Ricklefs:1996wb; @Chamberlain:2012vx; @OMeara:2012gq].
+The value of these methods relative to pairwise comparisons has been repeatedly shown in analyses of other types of character data [@Garland:2005ek; @Ricklefs:1996wb; @Chamberlain:2012vx; @OMeara:2012gq].
 The application of these methods is illustrated by a study of leaf functional attributes for about 100 plant species [@Ackerly:1999iu].
@@
 Phylogenetic comparative methods capture this shift as change along a single branch deep in the tree, revealing that there is not a tendency of correlated change between these traits across the phylogeny.
-One reason that comparative functional genomic sudies have not embraced phylogenetic approaches is that there has not yet been a concrete demonstration that pairwise and phylogenetic comparative methods can lead to different results when considering functional genomic data.
+One reason that comparative functional genomic studies have not embraced phylogenetic approaches is that there has not yet been a concrete demonstration that pairwise and phylogenetic comparative methods can lead to different results when considering functional genomic data.
 Here we examine this issue by re-evaluating the pairwise comparisons in two recent studies that compared gene expression across species.
@@
 It shapes our understanding of the functional diversity of gene families.
-It is also used to relate findings from well-studied genes to related genes that have not been investigated in detail.
+It is also used to relate findings from well-studied genes to homologous genes that have not been investigated in detail.
 It has be applied to many trait of genes, from gene sequence to biochemical properties to expression.
@@
 Our use of contrasts is a bit different-- we look for differences in evolutionary changes of one trait (differential expression) between two categories of nodes (speciation and duplication).
+This is similar to previous applications of contrasts to examine shifts in evolutionary rates between clades [@Garland:1992jl].
+Rather than compare the distributions of contrasts for nodes in different clades, as this previous work did, we compare the distributions of contrasts for nodes with different annotations (speciation or duplication).
 
@@
 
-An additional concern is that within-species expression variation is not considered when making comparisons of tau across homologous genes.
+An additional concern is that within-species expression variation is not considered when making comparisons of Tau across homologous genes.
 Neglecting within species variation is expected to mislead phylogenetic comparative analyses in some cases [@Ives:2007cw; Felsenstein:2008cl; Rohlfs:2015bd].
@@
 This provides an opportunity to assess the impact that extreme within-species variation (*i.e.*, as large as the variance between humans and chimpanzees and their most recent common ancestor) would have on our findings.
-These analyses still do not reject the null hypothesis, and do not provide support for the ortholog conjecture in the presence of high within-species variance (*SI Appendix*, Figure 6).
+These analyses still do not reject the null hypothesis, and do not provide support for the ortholog conjecture (*SI Appendix*, Figure 6).
 
@@
 The top left pane (*A*) reproduces the pattern presented in Figure 2A of KMRR [@KryuchkovaMostacci:2016iw] of higher correlation across speciation events than duplication events, which they took as evidence of the ortholog conjecture.
-The recovery of a similar  pattern under both simulations (*C, E*) indicates that it this pairwise approach does not make distinct testable predictions.
+The recovery of a similar  pattern under both simulations (*C, E*) indicates that this pairwise approach does not make distinct testable predictions.
 The phylogenetic analysis of the original data (*B*) does not show an excess of larger contrasts for duplication events and does not reject the null hypothesis, providing no support for the ortholog conjecture.
@@
 Second, the correlation for each group is based on multiple non-independent pairwise comparisons.
-
-To better understand this plot (Figure 2*A*), we performed simulations of Tau on the calibrated gene trees and then regenerated the figures.
+Third, instead of using the actual age of duplication events, the age of an adjacent named node in the species tree is used.
+
+To better understand this plot (Figure 2*A*), we performed simulations of Tau on the calibrated gene trees.
 We did not modify the gene tree topologies or their inferred histories of duplication and loss.
@@
 Under this neutral conjecture, phylogenetic distance is a better predictor of the similarity of gene function than is the history of gene duplication and speciation.
-The ortholog conjecture does not have to be an all or nothing question, though.
+The ortholog conjecture does not have to be an all or nothing process, though.
 It may be the case that the rates of phenotypic evolution following duplication may be greater than that following duplication in some organisms, gene families, and evolutionary processes [@Gabaldon:2013fi].
@@
 The five least-similar mid phase scores were all from the ctenophore.
-Published KS p-values, based on duplicated data, are in parentheses.
 The inset ctenophore image is by S.
@@
 
-Branch lengths are fundamental to phylogenetic comparative methods, since the null expectation is that more changes are expected along longer branches.
-In this study we applied three separate modifications of branch lengths.
+Branch lengths are fundamental to phylogenetic comparative methods, since the null expectation is that there are more changes along longer branches.
+In this study we applied four separate modifications of branch lengths.
 First, we time calibrated the gene trees so that nodes corresponding to the same speciation events had the same age across all trees.
-This is critical to making the comparisons of funcitonal genomic traits equivalent across genes.
-Second, phyogenetic independent contrasts adjust branch lengths so th
+This is critical to making the comparisons of functional genomic traits equivalent across genes.
+Second, phylogenetic independent contrasts (including the implementation we use here) adjust internal branch lengths to accommodate the variation associated with estimating character states at internal nodes [@Felsenstein:1985ua].
+Third, we randomized the calibration times to determine the sensitivity of the analyses to calibration times and the branch length estimations they impact (*SI Appendix*).
+Fourth, we extended terminal branch lengths to understand the potential impact of within-species variation (*SI Appendix*, Figure 6).
+The first two of these modifications are technical requirements for the application of phylogenetic comparative methods.
+The second two explore the sensitivity of the analyses to branch length and the information they convey, and find that the results are robust to variation in branch length.
+This is encouraging for the future of phylogenetic comparative functional genomics, given that one concern of applying these methods is the challenge of estimating branch lengths.
 
@@
 We are not the first to apply phylogenetic comparative methods to functional genomic data.
-While the vast majority of comparative functional genomic studies have used standard pairwise similarity methods, a small number of comparative functional genomic studies have employed phylogenetic comparative approaches [@Oakley:2005ky; @Eng:2009gp; @Chang:2014hq].
-For instance, a phylogenetic ANOVA [@Rohlfs:2015bd] of the evolution of gene expression improves statistical power and drastically reduces the rate of false positives relative to pairwise approaches.
+While the vast majority of comparative functional genomic studies have used standard pairwise similarity methods, a small number of comparative functional genomic studies have employed phylogenetic comparative approaches [@Oakley:2005ky; @Eng:2009gp; @Chang:2014hq; @Whitney:2011fw].
+For instance, a phylogenetic ANOVA [@Rohlfs:2015bd] of the evolution of gene expression in strict orthologs improves statistical power and drastically reduces the rate of false positives relative to pairwise approaches.
 
 Some of the most widely used phylogenetic comparative methods [@Felsenstein:1985ua; @Grafen:1989um] are already directly applicable to comparative functional genomic studies.
-
 Recent advances in phylogenetic comparative methods are particularly well suited to addressing questions about the evolution of functional genomic traits.
@@
 
-These analyses illustrate how important it is to not conflate evolutionary patterns with the processes that generated them.
+These analyses also illustrate how important it is to not conflate evolutionary patterns with the processes that generated them.
 Finding a pattern wherein paralogs tend to be more different than orthologs is not evidence that there are different processes by which orthologs and paralogs evolve.
@@
 
-```{r within_species_variation, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, dev="png", dpi=600, fig.width=6.5, fig.height=3, fig.cap="Investigation of within-species variation.
-These plots correspond to Fig.
-2*A* and 2*B*, but with increased expected within-species variation modeled by increasing the terminal gene phylogeny branch lengths.
-The results are unchanged.
+```{r within_species_variation, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, dev="png", dpi=600, fig.width=6.5, fig.height=3, fig.cap="Investigation of the potential impact of within-species varianace.
+These plots correspond to Figures 2*A* and 2*B*, but with increased expected within-species variation modeled by increasing the terminal gene phylogeny branch lengths.
 (*A*) Pairwise comparisons still recover higher correlation across speciation events than duplication events.

### 2017-07-18 manuscript.rmd
added context for pic

@@
 It has gene expression data for six organs across ten species (nine mammals and one bird), eight of which were analyzed by KMRR and further considered here.
+
+Many phylogenetic comparative methods are now available for addressing a wide range of questions relevant to comparative functional genomics.
+Some recent methods have been developed specifically for analyses of expression.
+For example, the Expression Variance and Evolution (EVE) model compares the ratio of within-species expression variance to between-species evolutionary expression variance in an explicit phylogenetic framework [@Rohlfs:2015bd].
+It can test a variety of hypotheses, including lineage-specific expression level shifts.
+This method considers strict orthologs that all have the same gene tree, which is the same as the phylogeny of the species under consideration.
+We are addressing different questions that consider a broad diversity of gene trees that include both speciation and duplication.
+Phylogenetic independent contrasts [@Felsenstein:1985ua], the most widely used phylogenetic comparative method, is particularly well suited to this challenge.
 

### 2017-07-20 manuscript.rmd
cat edits

@@
 This method considers strict orthologs that all have the same gene tree, which is the same as the phylogeny of the species under consideration.
-We are addressing different questions that consider a broad diversity of gene trees that include both speciation and duplication.
+However, for the re-analysis presented here we address different questions that consider a broad diversity of gene trees that include both speciation and different patterns of duplication.
 Phylogenetic independent contrasts [@Felsenstein:1985ua], the most widely used phylogenetic comparative method, is particularly well suited to this challenge.
@@
 
-The independent contrast across a node is the amount of change observed between the daughter notes, scaled by the expected variance [@Felsenstein:1985ua].
+The independent contrast across a node is the amount of change observed between the daughter nodes, scaled by the expected variance [@Felsenstein:1985ua].
 The expected variance is principally determined by the lengths of the branches leading from the node to these daughters.

### 2017-07-21 manuscript.rmd
incorporated Stefan's comments

@@
 
-Over the last three decades, a rich set of phylogenetic comparative methods has been developed to address the challenges and opportunities of trait comparisons across species [@Felsenstein:1985ua; @Grafen:1989um; @Pagel:1999gc; @Revell:2012jd; @FitzJohn:2012ey; @Uyeda:2014jl; @Garamszegi:2014ew; @Jhwueng:2013fi].
+Over the last three decades, a rich set of phylogenetic comparative methods has been developed to address the challenges and opportunities of trait comparisons across species [@Felsenstein:1985ua; @Grafen:1989um; @Pagel:1999gc; @Revell:2012jd; @FitzJohn:2012ey; @Uyeda:2014jl; @Garamszegi:2014ew; @Jhwueng:2013fi; @Rohlfs:2015bd].
 A central challenge is the dependence of observations across species due to the evolutionary history of species -- more closely related species share more traits because they have a more recent common ancestor.
@@
 It is also used to relate findings from well-studied genes to homologous genes that have not been investigated in detail.
-It has be applied to many trait of genes, from gene sequence to biochemical properties to expression.
+It has been applied to many traits of genes, from gene sequence to biochemical properties to expression.
 While the ortholog conjecture describes a specific pattern of functional diversity across genes, it is also articulated as a hypothesis about the process of evolution-- that there is greater evolutionary change in gene attributes following a duplication event than a speciation event.
@@
 Analyses of domain structure were consistent with the ortholog conjecture [@Forslund:2011cw].
-There have been few tests of the ortholog conjecture with regards to gene expression [@Chen:2012jz], and KMRR is the most thorough such expression study to date.
+There have been few tests of the ortholog conjecture with regards to gene expression [@Chen:2012jz], and KMRR is one of the most detailed to date.
 
@@
 
-We mapped the Tau values calculated by KMRR [@KryuchkovaMostacci:2016iw] for the Brawand *et al.* 2011 [@Brawand:2011du] dataset onto `r n_gene_trees` gene trees parsed from ENSEMBL Compara [@Herrero:2016jj].
+We mapped the Tau values calculated by KMRR [@KryuchkovaMostacci:2016iw] for the Brawand *et al.* 2011 [@Brawand:2011du] dataset onto `r n_gene_trees %>% format(big.mark=",")` gene trees parsed from ENSEMBL Compara [@Herrero:2016jj].
 These are the same pre-computed trees that the orthology/ paralogy annotations KMRR used are based on.
-`r length(gene_trees_pruned)` gene trees passed taxon sampling criteria (`r min_genes_with_expression` genes) after removing tips without Tau values and had at least one speciation event.
-Of these, `r length(gene_trees_calibrated)` were successfully time calibrated.
-These calibrated trees were used to calculate phylogenetic independent contrasts for `r nodes_contrast %>% filter(D=="Y") %>% nrow()` duplication nodes and `r nodes_contrast %>% filter(D=="N") %>% nrow()` speciation nodes.
+`r length(gene_trees_pruned) %>% format(big.mark=",")` gene trees passed taxon sampling criteria (`r min_genes_with_expression %>% format(big.mark=",")` genes) after removing tips without Tau values and had at least one speciation event.
+Of these, `r length(gene_trees_calibrated)  %>% format(big.mark=",")` were successfully time calibrated.
+These calibrated trees were used to calculate phylogenetic independent contrasts for `r nodes_contrast %>% filter(D=="Y") %>% nrow() %>% format(big.mark=",")` duplication nodes and `r nodes_contrast %>% filter(D=="N") %>% nrow() %>% format(big.mark=",")` speciation nodes.
 One of these trees is presented in *SI Appendix* Figure 4 to demonstrate the analysis.
@@
 An additional concern is that within-species expression variation is not considered when making comparisons of Tau across homologous genes.
-Neglecting within species variation is expected to mislead phylogenetic comparative analyses in some cases [@Ives:2007cw; Felsenstein:2008cl; Rohlfs:2015bd].
+Neglecting within species variation is expected to mislead phylogenetic comparative analyses in some cases [@Ives:2007cw; @Felsenstein:2008cl; @Rohlfs:2015bd].
 Since branch length describes expected trait variation [@Felsenstein:1985ua], within species variation can be modeled by extending the terminal branches of each gene phylogeny [@Ives:2007cw].
@@
 They therefore have lower correlation than pairwise comparisons made across speciation nodes.
-Of the `r n_dup_original` duplication nodes considered here, `r n_dup_remaining` have a node age (as determined by the time calibrated trees) within `r age_window * 100`% of their clade annotation.
+Of the `r n_dup_original %>% format(big.mark=",")` duplication nodes considered here, `r n_dup_remaining %>% format(big.mark=",")` have a node age (as determined by the time calibrated trees) within `r age_window * 100`% of their clade annotation.
 If only these duplication events closest to their annotated clade nodes are retained, the pattern of lower correlation of Tau evolution for duplication events disappears for the dataset simulated under the null model (*SI Appendix*, Figure 7*B*).
@@
 We found that all five of the lowest values in the mid phase distribution (Figure 3*A*) are for pairwise comparisons that include the ctenophore (comb jelly).
-When the nine pairwise comparisons that include the ctenophore are removed, there is no significant difference between the early phase and mid phase distributions (p = 0.14 for the early-mid comparison and p < $10^{-5}$ for the late-mid comparison) and no support for the inverse hourglass (Figure 3b).
+When the nine pairwise comparisons that include the ctenophore are removed, there is no significant difference between the early phase and mid phase distributions (p = 0.14 for the early-mid comparison and p < $10^{-5}$ for the late-mid comparison) and no support for the inverse hourglass (Figure 3*B*).
 This highlights a well understood property of pairwise comparisons across species [@Felsenstein:1985ua; @Ackerly:1999iu]: evolutionary changes along a given branch, like those along the ctenophore branch, impact each of the multiple pairwise comparisons that includes that branch.
@@
 After parsing the trees from the Compara file with treeio, which was recently split from ggtree [@Yu:2016fo], we added Tau estimates generated by the KMRR `Rscript.R` to the tree data objects.
-We then pruned away tips without expression data, retaining only the trees with `r min_genes_with_expression` or more tips.
+We then pruned away tips without expression data, retaining only the trees with `r min_genes_with_expression %>% format(big.mark=",")` or more tips.
 We also only retained trees with one or more speciation events, as speciation events are required for calibration steps.
@@
 
-```{r within_species_variation, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, dev="png", dpi=600, fig.width=6.5, fig.height=3, fig.cap="Investigation of the potential impact of within-species varianace.
+```{r within_species_variation, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, dev="png", dpi=600, fig.width=6.5, fig.height=3, fig.cap="Investigation of the potential impact of within-species variance.
 These plots correspond to Figures 2*A* and 2*B*, but with increased expected within-species variation modeled by increasing the terminal gene phylogeny branch lengths.

### 2017-07-22 manuscript.rmd
converted figure formats to dev=pdf

@@
 output:
+  word_document: default
   pdf_document:
@@
     fig_width: 6
-  word_document: default
 ---
@@
 
-```{r pairwise_and_phylo_plots, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.width=6.5, fig.height=6.5, fig.cap="Pairwise (left column - *A, C, E*) and phylogenetic (right column - *B, D, F*) analyses of the original data (first row - *A, B*), data simulated under the null hypothesis (second row - *C, D*), and data simulated under the ortholog conjecture (third row- *E, F*).
+```{r pairwise_and_phylo_plots, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.width=6.5, fig.height=6.5, dev="pdf", fig.cap="Pairwise (left column - *A, C, E*) and phylogenetic (right column - *B, D, F*) analyses of the original data (first row - *A, B*), data simulated under the null hypothesis (second row - *C, D*), and data simulated under the ortholog conjecture (third row- *E, F*).
 In the pairwise plots, each point indicates the correlation coefficient of Tau for a set of pairwise comparisons annotated with a specific node name (e.g, Primates) and event type (speciation or duplication, giving rise respectively to orthologs and paralogs).
@@
 
-```{r example_trees, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, eval=TRUE, fig.cap="One of the gene trees from the phylogenetic analysis.
+```{r example_trees, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, eval=TRUE, dev="pdf", fig.cap="One of the gene trees from the phylogenetic analysis.
 The value of Tau (expression specificity) is indicated by the sizes of the circles at the tips of the tree.
@@
 
-```{r ascertainment_plot, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, dev="png", dpi=600, fig.cap="Investigation of possible ascertainment biases.
+```{r ascertainment_plot, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, dev="pdf", dpi=600, fig.cap="Investigation of possible ascertainment biases.
 (*A*) Magnitude of independent contrasts plotted against node age.
@@
 
-```{r within_species_variation, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, dev="png", dpi=600, fig.width=6.5, fig.height=3, fig.cap="Investigation of the potential impact of within-species variance.
+```{r within_species_variation, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, dev="pdf", dpi=600, fig.width=6.5, fig.height=3, fig.cap="Investigation of the potential impact of within-species variance.
 These plots correspond to Figures 2*A* and 2*B*, but with increased expected within-species variation modeled by increasing the terminal gene phylogeny branch lengths.
@@
 
-```{r cor_structure_plot, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, dev="png", dpi=600, fig.width=6.5, fig.height=6.5, fig.cap="Origin of pairwise correlation plot structure.
+```{r cor_structure_plot, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, dev="pdf", dpi=600, fig.width=6.5, fig.height=6.5, fig.cap="Origin of pairwise correlation plot structure.
 (*A*) Each speciation and duplication node in the COMPARA trees is annotated with a clade name, and the age of that clade (shown on the x axis) is used as the age of the pairwise comparisons (Figure 2 *A, C, E*, KMRR [@KryuchkovaMostacci:2016iw] Figure 2).
@@
 
-```{r Tau_v_max, cache=TRUE, echo=FALSE, warning=FALSE, dev="png", dpi=300, message=FALSE, fig.cap="Genes with higher maximum observed expression tend to have lower Tau.
+```{r Tau_v_max, cache=TRUE, echo=FALSE, warning=FALSE, dev="pdf", dpi=300, message=FALSE, fig.cap="Genes with higher maximum observed expression tend to have lower Tau.
 The blue line is the linear model for the relationship between the two."}

### 2017-10-04 manuscript.rmd
light edits throughout

--- 2017-07-21 manuscript.rmd
@@
     fig_caption: yes
-    fig_height: 4
-    fig_width: 6
   word_document: default
@@
 
-	knitr::opts_chunk$set( echo = TRUE )
+	knitr::opts_chunk$set( 
+	  echo = TRUE,
+    dev="tiff", 
+    dpi=1200
+  )
 
@@
 
-^1^ Department of Ecology and Evolutionary Biology, Brown University, Providence, RI, USA
-
-^2^ Current address: Department of Ecology and Evolutionary Biology, Yale University, New Haven, CT, USA
-
-^3^ Department of Ecology and Evolutionary Biology, University of California Los Angeles, Los Angeles, CA, USA
-
-^4^ Department of Molecular & Cellular Biology, University of California at Davis, Davis, CA, USA
-
-^5^ Sars International Centre for Marine Molecular Biology, University of Bergen, Bergen, Norway
+^1^ Department of Ecology and Evolutionary Biology, Brown University, Providence, RI, 02912, USA
+
+^2^ Current address: Department of Ecology and Evolutionary Biology, Yale University, New Haven, CT, 06511, USA
+
+^3^ Department of Ecology and Evolutionary Biology, University of California Los Angeles, Los Angeles, CA, 20095, USA
+
+^4^ Department of Molecular & Cellular Biology, University of California at Davis, Davis, CA, 95616, USA
+
+^5^ Sars International Centre for Marine Molecular Biology, University of Bergen, Bergen, 5006, Norway
 
@@
 Here we examine two recently published studies that compare gene expression across species with pairwise methods, and find reason to question the original conclusions of both.
-One study interpreted pairwise comparisons of gene expression as support for the ortholog conjecture, the hypothesis that orthologs tend to be more similar than paralogs.
+One study interpreted pairwise comparisons of gene expression as support for the ortholog conjecture, the hypothesis that orthologs tend to have more similar attributes (expression in this case) than do paralogs.
 The other study interpreted pairwise comparisons of embryonic gene expression across distantly related animals as evidence for a distinct evolutionary process that gave rise to phyla.
@@
 They have largely been applied to morphological and ecological traits, but are just as relevant to functional genomics [@Barker:2005be].
-Even so, most comparative functional genomic studies have abstained from phylogenetic approaches and instead rely on multiple pairwise comparisons across species (Figure 1*A*).
+Even so, most comparative functional genomic studies have abstained from phylogenetic approaches and instead rely on multiple pairwise comparisons across species (Fig.
+1*A*).
 This leaves comparative functional genomic studies susceptible to statistical problems and is a missed opportunity to ask questions that are only accessible in an explicit phylogenetic context.
 
-Phylogenetic comparative methods account for evolutionary history and explicitly model trait change along the branches of evolutionary trees (*e.g.*, Figure 1*B*).
+Phylogenetic comparative methods account for evolutionary history and explicitly model trait change along the branches of evolutionary trees (*e.g.*, Fig.
+1*B*).
 The value of these methods relative to pairwise comparisons has been repeatedly shown in analyses of other types of character data [@Garland:2005ek; @Ricklefs:1996wb; @Chamberlain:2012vx; @OMeara:2012gq].
@@
 The first study, Kryuchkova-Mostacci and Robinson-Rechavi (KMRR) [@KryuchkovaMostacci:2016iw], analyzed multiple vertebrate expression datasets to test the ortholog conjecture - the hypothesis that orthologs tend to have more conserved attributes (specificity of expression across organs in this case) than do paralogs [@Nehrt:2011bm, @Koonin:2005fp].
-Using pairwise comparisons (Figure 1*A*), they found lower expression correlation between paralogs than between orthologs and interpreted this as strong support for the ortholog conjecture.
+Using pairwise comparisons (Fig.
+1*A*), they found lower expression correlation between paralogs than between orthologs and interpreted this as strong support for the ortholog conjecture.
 
@@
 The ortholog conjecture [@Nehrt:2011bm] is the proposition that orthologs (genes that diverged from each other due to a speciation event) have more similar attributes than do paralogs (genes that diverged from each other due to a gene duplication event).
+The ortholog can and has been applied to diverse attributes, including molecular sequence, biochemical function, and, as in the study considered here, expression.
 The ortholog conjecture has important biological and technical implications.
@@
 
-The KMRR analyses are based on pairwise comparisons (Figure 1*A*) between Tau within each gene family.
+The KMRR analyses are based on pairwise comparisons (Fig.
+1*A*) between Tau within each gene family.
 Rather than make every pairwise comparison within each gene tree, they considered only a subset of pairwise comparisons in each particular analysis.
@@
 For each internal node in each gene tree, we calculated the phylogenetic independent contrast [@Felsenstein:1985ua] (PIC) of Tau.
-This is the difference in values of Tau for descendant nodes scaled by the expected variance, which is largely determined by the lengths of the two branches that connect the node to its two descendants (Figure 1*B*).
+This is the difference in values of Tau for descendant nodes scaled by the expected variance, which is largely determined by the lengths of the two branches that connect the node to its two descendants (Fig.
+1*B*).
 These contrasts were then annotated by whether each is made across a speciation or duplication event.
@@
 Of these, `r length(gene_trees_calibrated)  %>% format(big.mark=",")` were successfully time calibrated.
-These calibrated trees were used to calculate phylogenetic independent contrasts for `r nodes_contrast %>% filter(D=="Y") %>% nrow() %>% format(big.mark=",")` duplication nodes and `r nodes_contrast %>% filter(D=="N") %>% nrow() %>% format(big.mark=",")` speciation nodes.
-One of these trees is presented in *SI Appendix* Figure 4 to demonstrate the analysis.
+These calibrated trees were used to calculate phylogenetic independent contrasts for `r nodes_contrast %>% filter(D=="Y") %>% nrow() %>% format(big.mark=",")` duplication nodes and `r nodes_contrast %>% filter(D=="N") %>% nrow() %>% format(big.mark=",")` speciation nodes (Tables S1, S2).
+One of these trees is presented in *SI Appendix* Fig.
+4 to demonstrate the analysis.
 
@@
 
-We did not find increased evolutionary change in expression following duplication events (Figure 2*B*).
+We did not find increased evolutionary change in expression following duplication events as compared to speciation events (Fig.
+2*B*).
 The Wilcoxon rank test does not reject the null hypothesis that the rate of evolution following duplications is the same as or less than the rate following speciation (p value = `r wilcox_test_result`).
@@
 We focused on two possible sources of bias, node depth and branch length.
-We found no evidence that either affected our results (*SI Appendix*, Figure 5).
+We found no evidence that either affected our results (*SI Appendix*, Fig.
+5).
 We also examined the sensitivity of the results to the calibration times applied to speciation events on the gene trees.
@@
 This provides an opportunity to assess the impact that extreme within-species variation (*i.e.*, as large as the variance between humans and chimpanzees and their most recent common ancestor) would have on our findings.
-These analyses still do not reject the null hypothesis, and do not provide support for the ortholog conjecture (*SI Appendix*, Figure 6).
+These analyses still do not reject the null hypothesis, and do not provide support for the ortholog conjecture (*SI Appendix*, Fig.
+6).
 
@@
 This is important since we are only looking at a subset of the data they considered, the Brawand *et al.* 2011 [@Brawand:2011du] dataset for gene trees that could be successfully time calibrated.
-In their Figure 1 [@KryuchkovaMostacci:2016iw], they present a higher Tau correlation coefficient between ortholog pairs than paralog pairs.
+In their Fig.
+1 [@KryuchkovaMostacci:2016iw], they present a higher Tau correlation coefficient between ortholog pairs than paralog pairs.
 We find the same here, with correlation coefficients of `r signif( ortholog_r, 2)` for orthologs and `r signif( paralog_r, 2)` for paralogs.
@@
 
-To test the hypothesis that ancient duplications that precede the oldest speciation event (*SI Appendix*, Figure 5*A*) impact the lower correlation of Tau between paralogs than between orthologs, we removed them.
-When we consider only the duplication events the same age or younger than the oldest speciation event (*SI Appendix*, Figure 5*B*), the paralog correlation coefficient increases from `r signif( paralog_r, 2)` to `r signif( paralog_subset_r, 2)`.
+To test the hypothesis that ancient duplications that precede the oldest speciation event (*SI Appendix*, Fig.
+5*A*) impact the lower correlation of Tau between paralogs than between orthologs, we removed them.
+When we consider only the duplication events the same age or younger than the oldest speciation event (*SI Appendix*, Fig.
+5*B*), the paralog correlation coefficient increases from `r signif( paralog_r, 2)` to `r signif( paralog_subset_r, 2)`.
 This is much closer to the ortholog correlation of `r signif( ortholog_r, 2) `.
@@
 
-```{r pairwise_and_phylo_plots, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.width=6.5, fig.height=6.5, fig.cap="Pairwise (left column - *A, C, E*) and phylogenetic (right column - *B, D, F*) analyses of the original data (first row - *A, B*), data simulated under the null hypothesis (second row - *C, D*), and data simulated under the ortholog conjecture (third row- *E, F*).
+```{r Fig_2, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.width=7, fig.height=6.5, dev="quartz_pdf", fig.cap="Pairwise (left column - *A, C, E*) and phylogenetic (right column - *B, D, F*) analyses of the original data (first row - *A, B*), data simulated under the null hypothesis (second row - *C, D*), and data simulated under the ortholog conjecture (third row- *E, F*).
 In the pairwise plots, each point indicates the correlation coefficient of Tau for a set of pairwise comparisons annotated with a specific node name (e.g, Primates) and event type (speciation or duplication, giving rise respectively to orthologs and paralogs).
@@
 A horizontal line at 0 would indicate that the density distributions are identical.
-The top left pane (*A*) reproduces the pattern presented in Figure 2A of KMRR [@KryuchkovaMostacci:2016iw] of higher correlation across speciation events than duplication events, which they took as evidence of the ortholog conjecture.
+The top left pane (*A*) reproduces the pattern presented in Fig.
+2*A* of KMRR [@KryuchkovaMostacci:2016iw] of higher correlation across speciation events than duplication events, which they took as evidence of the ortholog conjecture.
 The recovery of a similar  pattern under both simulations (*C, E*) indicates that this pairwise approach does not make distinct testable predictions.
@@
 The KMRR study did investigate the impact of node age on correlation, but in a different way.
-In their Figure 2 [@KryuchkovaMostacci:2016iw], they grouped orthologs and paralogs according to the ENSEMBL node name of their most recent common ancestor, and plotted the correlation of Tau for each of these groups by the node age.
+In their Fig.
+2 [@KryuchkovaMostacci:2016iw], they grouped orthologs and paralogs according to the ENSEMBL node name of their most recent common ancestor, and plotted the correlation of Tau for each of these groups by the node age.
 They found that across the investigated range of node ages, ortholog pairs have higher Tau correlation than paralogs.
-We confirmed that we can replicate this result (Figure 2*A*).
+We confirmed that we can replicate this result (Fig.
+2*A*).
 There are several difficulties with interpreting this plot, though.
@@
 
-To better understand this plot (Figure 2*A*), we performed simulations of Tau on the calibrated gene trees.
+To better understand this plot (Fig.
+2*A*), we performed simulations of Tau on the calibrated gene trees.
 We did not modify the gene tree topologies or their inferred histories of duplication and loss.
 First, we simulated the evolution of Tau under the null model that it evolves at the same rate following duplication and speciation events.
-Under the null model, the plot of correlation coefficient to node age (Figure 2*C*) is very similar as for the observed data (Figure 2*A*).
+Under the null model, the plot of correlation coefficient to node age (Fig.
+2*C*) is very similar as for the observed data (Fig.
+2*A*).
 As in the original study, there is higher correlation coefficient across orthologs (`r signif( ortholog_r_sim_null, 2 )`) than paralogs (`r signif( paralog_r_sim_null, 2 )`) when not considering node age.
-Phylogenetic analysis of the data simulated under the null hypothesis (Figure 2*D*) do not reject the null hypothesis (Wilcoxon p = $`r wilcox_oc( nodes_sim_null_contrast ) %>% round( 3 )`$), as expected.
+Phylogenetic analysis of the data simulated under the null hypothesis (Fig.
+2*D*) do not reject the null hypothesis (Wilcoxon p = $`r wilcox_oc( nodes_sim_null_contrast ) %>% round( 3 )`$), as expected.
 
 We next simulated the evolution of Tau under the ortholog conjecture, where the rate of evolution of Tau following duplication was `r dup_adjust` fold the rate following speciation.
-The pairwise results of this heterogeneous model (Figure 2*E*) are nearly indistinguishable from the results under the null model (Figure 2*C*), and also have a higher correlation coefficient for orthologs (`r signif( ortholog_r_sim_oc, 2 )`) than paralogs (`r signif( paralog_r_sim_oc, 2 )`).
-The phylogenetic analysis of the ortholog conjecture simulation (Figure 2*F*) does reject the null hypothesis (Wilcoxon p = $`r signif( wilcox_oc( nodes_sim_oc_contrast ), 2 )`$).
+The pairwise results of this heterogeneous model (Fig.
+2*E*) are nearly indistinguishable from the results under the null model (Fig.
+2*C*), and also have a higher correlation coefficient for orthologs (`r signif( ortholog_r_sim_oc, 2 )`) than paralogs (`r signif( paralog_r_sim_oc, 2 )`).
+The phylogenetic analysis of the ortholog conjecture simulation (Fig.
+2*F*) does reject the null hypothesis (Wilcoxon p = $`r signif( wilcox_oc( nodes_sim_oc_contrast ), 2 )`$).
 
@@
 The pairwise comparisons used by KMRR cannot distinguish between the null hypothesis and ortholog conjecture.
-The pairwise results are strikingly similar under both hypotheses (Figure 2*C*, 2*E*).
+The pairwise results are strikingly similar under both hypotheses (Fig.
+2*C*, 2*E*).
 These simulations also serve to validate the phylogenetic methods applied to this problem.
@@
 
-Since greater rates of expression evolution following duplication do not explain the lower correlation of Tau values in the pairwise comparison plots (Figure 2 *A, C, E*), this pattern must reflect some other property that is shared across these analyses.
+Since greater rates of expression evolution following duplication do not explain the lower correlation of Tau values in the pairwise comparison plots (Fig.
+2 *A, C, E*), this pattern must reflect some other property that is shared across these analyses.
 We found that the lower correlations for duplication comparisons can largely be explained by the greater variance in the age of duplication nodes.
-Each point in the pairwise correlation plot (Figure 2 *A, C, E*) summarizes multiple pairwise comparisons across nodes of a given event type (speciation or duplication, as indicated by color) and clade in the species tree (Theria, Mammalia, Amniota, etc..., as indicated by the age of the clade along the x axis).
+Each point in the pairwise correlation plot (Fig.
+2 *A, C, E*) summarizes multiple pairwise comparisons across nodes of a given event type (speciation or duplication, as indicated by color) and clade in the species tree (Theria, Mammalia, Amniota, etc..., as indicated by the age of the clade along the x axis).
 These clade names in turn are derived from the node annotations in the ENSEMBL Compara trees, which apply clade names to both speciation and duplication nodes.
@@
 Neighboring clade ages are therefore a very rough approximation of the age of duplication events.
-This is apparent when node ages derived from calibrated trees are plotted against the age of the clade annotations for each node (*SI Appendix*, Figure 7*A*).
+This is apparent when node ages derived from calibrated trees are plotted against the age of the clade annotations for each node (*SI Appendix*, Fig.
+7*A*).
 Because duplication events that are all annotated with the same clade name can occur at very different times, pairwise comparisons across these nodes capture evolutionary changes in Tau across very different different branch lengths.
@@
 Of the `r n_dup_original %>% format(big.mark=",")` duplication nodes considered here, `r n_dup_remaining %>% format(big.mark=",")` have a node age (as determined by the time calibrated trees) within `r age_window * 100`% of their clade annotation.
-If only these duplication events closest to their annotated clade nodes are retained, the pattern of lower correlation of Tau evolution for duplication events disappears for the dataset simulated under the null model (*SI Appendix*, Figure 7*B*).
+If only these duplication events closest to their annotated clade nodes are retained, the pattern of lower correlation of Tau evolution for duplication events disappears for the dataset simulated under the null model (*SI Appendix*, Fig.
+7*B*).
 This correction, though, comes at the high cost of discarding `r round( (n_dup_original - n_dup_remaining )/ n_dup_original * 100, 1 )`% of duplication nodes.
@@
 While some studies have presented support for the ortholog conjecture, our results are consistent with multiple studies that have not [@Studer:2009bm; @Nehrt:2011bm; @Yanai:2004hk].
+Since the ortholog conjecture can be applied to many different gene attributes at both the genotype level (*e.g.*, molecular sequence) and phenotype level (*e.g.*, activity, biochemical function, binding partners, and expression), it is naturally a heterogeneous and potentially idiosyncratic topic [@Soria:2014bv].
+For example, a recent integrated analysis of multiple gene traits after duplication found extensive divergence in protein sequence but less divergence in expression than expected [@Soria:2014bv].
 
@@
 They then calculated a similarity score for each temporal phase for each species pair based on the fraction of genes that exhibited the same patterns in each species.
-The distributions of similarity scores are plotted in their Figure 4d [@Levin:2016bp], and their Kolmogorov–Smirnov (KS) tests indicated that the early distribution and late distribution were each significantly different from mid distribution (P < $10^{-6}$ and P < $10^{-12}$, respectively).
+The distributions of similarity scores are plotted in their Fig.
+4d [@Levin:2016bp], and their Kolmogorov–Smirnov (KS) tests indicated that the early distribution and late distribution were each significantly different from mid distribution (P < $10^{-6}$ and P < $10^{-12}$, respectively).
 This is the support they presented for the inverse hourglass model.
@@
 
-We examined the matrix of pairwise comparisons used as the base for the KS tests and Figure 4d in *Levin et al.* [@Levin:2016bp], and thus as support for the "inverse hourglass" model.
+We examined the matrix of pairwise comparisons used as the base for the KS tests and Fig.
+4d in *Levin et al.* [@Levin:2016bp], and thus as support for the "inverse hourglass" model.
 We found several problems resulting from the use of multiple pairwise comparisons.
@@
 Once we addressed the issues above with the implementation of pairwise comparisons, we were able to explore more general issues that can be a problem when making multiple pairwise comparisons between species.
-We found that all five of the lowest values in the mid phase distribution (Figure 3*A*) are for pairwise comparisons that include the ctenophore (comb jelly).
-When the nine pairwise comparisons that include the ctenophore are removed, there is no significant difference between the early phase and mid phase distributions (p = 0.14 for the early-mid comparison and p < $10^{-5}$ for the late-mid comparison) and no support for the inverse hourglass (Figure 3*B*).
+We found that all five of the lowest values in the mid phase distribution (Fig.
+3*A*) are for pairwise comparisons that include the ctenophore (comb jelly).
+When the nine pairwise comparisons that include the ctenophore are removed, there is no significant difference between the early phase and mid phase distributions (p = 0.14 for the early-mid comparison and p < $10^{-5}$ for the late-mid comparison) and no support for the inverse hourglass (Fig.
+3*B*).
 This highlights a well understood property of pairwise comparisons across species [@Felsenstein:1985ua; @Ackerly:1999iu]: evolutionary changes along a given branch, like those along the ctenophore branch, impact each of the multiple pairwise comparisons that includes that branch.
-The pairwise comparisons are therefore not independent - different pairwise comparisons are impacted by changes along some of the same branches (Figure 1*A*).
+The pairwise comparisons are therefore not independent - different pairwise comparisons are impacted by changes along some of the same branches (Fig.
+1*A*).
 This can give the impression of a general pattern across the tree that is instead specific to changes along one part of the tree.
@@
 While we demonstrate problems with pairwise comparisons that impacts the Levin *et al.* analysis, we did not perform a phylogenetic reanalysis of this study, as we did for the KMRR study.
-This is because the similarity metric computed in the pairwise comparisons of Levin *et al.* is not suitable for phylogenetic analysis.
-One issue with the similarity metric is that it is based on different genes for different species pairs, and therefore is not a trait whose evolution can be modeled across the phylogeny.
+This is because the similarity metric computed in the pairwise comparisons of Levin *et al.* is not suitable for phylogenetic analysis as it is based on different genes for different species pairs, and therefore cannot be modeled across the phylogeny.
 A full phylogenetic reanalysis would be possible using upstream analysis products to re-derive new expression summary statistics.
@@
 ### Phylogenetic comparative methods in functional genomics
-
-#### Impact and importance
-
-Our results highlight the importance of explicitly incorporating information about phylogenetic relationships when comparing functional genomic traits across species.
-
-Addressing the statistical dependence of pairwise comparisons is not the only advantage of using phylogenetic comparative methods for functional genomic analyses.
-Another problem with the pairwise comparisons is that, except at the tips, they summarize changes along many branches in the phylogeny.
-Two paralogs that diverged from a duplication event deep in the tree may have many subsequent duplication and speciation events, and changes along all these branches will impact the final pairwise comparison.
-Phylogenetic methods have the advantage of isolating the changes under consideration (Figure 1*B*).
-Phylogenetic methods therefore avoid diluting the change that occurs along the branches that follow the node in question with changes along all subsequent branches.
-There may still be missing speciation events, due to extinction and incomplete taxon sampling, and missing duplication events, due to gene loss, but these omissions affect both methods.
 
@@
 Third, we randomized the calibration times to determine the sensitivity of the analyses to calibration times and the branch length estimations they impact (*SI Appendix*).
-Fourth, we extended terminal branch lengths to understand the potential impact of within-species variation (*SI Appendix*, Figure 6).
+Fourth, we extended terminal branch lengths to understand the potential impact of within-species variation (*SI Appendix*, Fig.
+6).
 The first two of these modifications are technical requirements for the application of phylogenetic comparative methods.
@@
 
-The fact that the first two comparative functional genomic studies we reanalyzed show serious problems with pairwise comparisons indicates that there are likely to be similar problems in other studies that use these methods.
+The problems we identify with pairwise comparisons in two recent functional genomics studies indicate that there are likely to be similar problems in other studies that use these methods.
 Future studies that compare functional genomic data across species will be compromised if they continue to use pairwise methods.
@@
 Levin *et al.* helpfully provided data and clarification on methods.
-We obtained the matrix of pairwise scores that underlies their Figure 4d and confirmed we could reproduce their published results.
+We obtained the matrix of pairwise scores that underlies their Fig.
+4d and confirmed we could reproduce their published results.
 We then removed duplicate rows, applied the Wilcoxon test in place of the Kolmogorov-Smirnov test, and identified ctenophores as overrepresented among the low outliers in the mid-developmental transition column.
@@
 
-```{r example_trees, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, eval=TRUE, fig.cap="One of the gene trees from the phylogenetic analysis.
+```{r Fig_S1, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, eval=TRUE, dev="quartz_pdf", fig.width=7, fig.height=6.5, fig.cap="One of the gene trees from the phylogenetic analysis.
 The value of Tau (expression specificity) is indicated by the sizes of the circles at the tips of the tree.
@@
 
-While the age of speciation nodes is constrained, duplication nodes can be much older and therefore have a wider range of ages (*SI Appendix*, Figure 5*A*).
+While the age of speciation nodes is constrained, duplication nodes can be much older and therefore have a wider range of ages (*SI Appendix*, Fig.
+5*A*).
 This is because many gene duplication events are older than the most recent common ancestor of the species in the study.
@@
 Gene tree estimation errors, for example, tend to lead to the overestimation of deep duplications [@Hahn:2007cj].
-If independent contrast values also tended to to be lower at greater node depth, it could interact with the preponderance of duplications at greater depth to create a pattern of lower contrasts associated with duplication events.
+If independent contrast values also tended to be lower at greater node depth, it could interact with the preponderance of duplications at greater depth to create a pattern of lower contrasts associated with duplication events.
 To test for such an effect, we remove duplication nodes that are older than the oldest speciation node.
-The general results are unchanged and this reduced dataset does not reject the null hypothesis that the rate of evolution following duplications is the same as or less than the rate following speciation (*SI Appendix*, Figure 5*B*).
+The general results are unchanged and this reduced dataset does not reject the null hypothesis that the rate of evolution following duplications is the same as or less than the rate following speciation (*SI Appendix*, Fig.
+5*B*).
 
@@
 This is because the same difference across a shorter total branch length indicates a greater rate of evolutionary change.
-The expected variance of contrasts for speciation nodes is constrained by the branch lengths on the species tree, but the expected variance of contrasts for duplications has a much wider range (*SI Appendix*, Figure 5*C*).
+The expected variance of contrasts for speciation nodes is constrained by the branch lengths on the species tree, but the expected variance of contrasts for duplications has a much wider range (*SI Appendix*, Fig.
+5*C*).
 This could lead to biases if the lengths of branches that descend from duplication nodes tend to be overestimated.
 We therefore examined only the contrasts that fell within in the range of expected variance seen for speciation contrasts, excluding duplication contrasts that fall outside of this range.
-This reanalysis does not reject the null hypothesis either (*SI Appendix*, Figure 5*D*), indicating that branch length bias is not responsible for the result.
-
-```{r ascertainment_plot, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, dev="png", dpi=600, fig.cap="Investigation of possible ascertainment biases.
+This reanalysis does not reject the null hypothesis either (*SI Appendix*, Fig.
+5*D*), indicating that branch length bias is not responsible for the result.
+
+```{r Fig_S2, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.width=7, fig.height=6.5, dev="tiff", dev.args=list( compression="lzw" ), fig.cap="Investigation of possible ascertainment biases.
 (*A*) Magnitude of independent contrasts plotted against node age.
@@
 
-```{r within_species_variation, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, dev="png", dpi=600, fig.width=6.5, fig.height=3, fig.cap="Investigation of the potential impact of within-species variance.
-These plots correspond to Figures 2*A* and 2*B*, but with increased expected within-species variation modeled by increasing the terminal gene phylogeny branch lengths.
+```{r Fig_S3, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, dev="quartz_pdf", fig.width=7, fig.height=3, fig.cap="Investigation of the potential impact of within-species variance.
+These plots correspond to Fig.
+2*A* and 2*B*, but with increased expected within-species variation modeled by increasing the terminal gene phylogeny branch lengths.
 (*A*) Pairwise comparisons still recover higher correlation across speciation events than duplication events.
@@
 
-```{r cor_structure_plot, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, dev="png", dpi=600, fig.width=6.5, fig.height=6.5, fig.cap="Origin of pairwise correlation plot structure.
-(*A*) Each speciation and duplication node in the COMPARA trees is annotated with a clade name, and the age of that clade (shown on the x axis) is used as the age of the pairwise comparisons (Figure 2 *A, C, E*, KMRR [@KryuchkovaMostacci:2016iw] Figure 2).
+```{r Fig_S4, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.width=7, fig.height=3, dev.args=list( compression="lzw" ), fig.cap="Origin of pairwise correlation plot structure.
+(*A*) Each speciation and duplication node in the COMPARA trees is annotated with a clade name, and the age of that clade (shown on the x axis) is used as the age of the pairwise comparisons (Fig.
+2 *A, C, E*, KMRR [@KryuchkovaMostacci:2016iw] Fig.
+2).
 The clade ages have a clear correspondence to the speciation node ages (red).
@@
 		g_sub,
-		ncol = 1
+		ncol = 2
 		)
@@
 
-```{r Tau_v_max, cache=TRUE, echo=FALSE, warning=FALSE, dev="png", dpi=300, message=FALSE, fig.cap="Genes with higher maximum observed expression tend to have lower Tau.
+```{r Tau_v_max, cache=TRUE, echo=FALSE, warning=FALSE, fig.width=3.42, fig.height=3.42, dev.args=list( compression="lzw" ), message=FALSE, fig.cap="Genes with higher maximum observed expression tend to have lower Tau.
 The blue line is the linear model for the relationship between the two."}
@@
 
-KMRR [@KryuchkovaMostacci:2016iw] Figure 3 differs from their other analyses in that it presents independent changes in expression between triplets of genes.
+KMRR [@KryuchkovaMostacci:2016iw] Fig.
+3 differs from their other analyses in that it presents independent changes in expression between triplets of genes.
 Each triplet has two paralogs that arose from a duplication event and one unduplicated homolog.
@@
 From this they conclude that following duplication there is a common trajectory of evolutionary change, where one paralog evolves to have lower expression and also become more tissue specific.
-There is, however, a negative relationship between Tau and maximum expression across all genes (*SI Appendix* Figure 8, $r^2=`r signif( tau_v_max_lm$r.squared, 3 )`$, $slope=`r signif( tau_v_max_lm$coefficients[2,1], 3 )`$, $p=`r signif( tau_v_max_lm$coefficients[2,4], 3 )`$).
+There is, however, a negative relationship between Tau and maximum expression across all genes (*SI Appendix* Fig.
+8, $r^2=`r signif( tau_v_max_lm$r.squared, 3 )`$, $slope=`r signif( tau_v_max_lm$coefficients[2,1], 3 )`$, $p=`r signif( tau_v_max_lm$coefficients[2,4], 3 )`$).
 A simpler explanation for their plot is that it reflects the global tendency for Tau to decrease with maximum expression.

### 2017-10-04 manuscript.rmd
Dockerfile - added texshop

@@
 	# The rest can be installed from CRAN
-	library( tidyverse )
-	library( magrittr )
-	library( stringr )
-	library( parallel )
 	library( ape )
   library( digest )
+  library( geiger )
+  library( ggrepel )
   library( gridExtra )
-	library( geiger )
+	library( magrittr )
+  library( parallel )
   library( phytools )
-	library( ggrepel )
+	library( stringr )
+  library( tidyverse )
 

### 2017-10-06 manuscript.rmd
cache all chunks, save workplace image
cleaned up knitr options

@@
 	echo = TRUE,
-    dev="tiff", 
-    dpi=1200
+    dev="png", 
+    # dev.args=list( compression="lzw" ), # for tiff
+    dpi=600,
+    cache=TRUE
   )
@@
 
-```{r Fig_2, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.width=7, fig.height=6.5, dev="quartz_pdf", fig.cap="Pairwise (left column - *A, C, E*) and phylogenetic (right column - *B, D, F*) analyses of the original data (first row - *A, B*), data simulated under the null hypothesis (second row - *C, D*), and data simulated under the ortholog conjecture (third row- *E, F*).
+```{r Fig_2, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.width=7, fig.height=6.5, fig.cap="Pairwise (left column - *A, C, E*) and phylogenetic (right column - *B, D, F*) analyses of the original data (first row - *A, B*), data simulated under the null hypothesis (second row - *C, D*), and data simulated under the ortholog conjecture (third row- *E, F*).
 In the pairwise plots, each point indicates the correlation coefficient of Tau for a set of pairwise comparisons annotated with a specific node name (e.g, Primates) and event type (speciation or duplication, giving rise respectively to orthologs and paralogs).
@@
 
-```{r Fig_S1, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, eval=TRUE, dev="quartz_pdf", fig.width=7, fig.height=6.5, fig.cap="One of the gene trees from the phylogenetic analysis.
+```{r Fig_S1, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, eval=TRUE, fig.width=7, fig.height=6.5, fig.cap="One of the gene trees from the phylogenetic analysis.
 The value of Tau (expression specificity) is indicated by the sizes of the circles at the tips of the tree.
@@
 
-```{r Fig_S2, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.width=7, fig.height=6.5, dev="tiff", dev.args=list( compression="lzw" ), fig.cap="Investigation of possible ascertainment biases.
+```{r Fig_S2, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.width=7, fig.height=6.5, fig.cap="Investigation of possible ascertainment biases.
 (*A*) Magnitude of independent contrasts plotted against node age.
@@
 
-```{r Fig_S3, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, dev="quartz_pdf", fig.width=7, fig.height=3, fig.cap="Investigation of the potential impact of within-species variance.
+```{r Fig_S3, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.width=7, fig.height=3, fig.cap="Investigation of the potential impact of within-species variance.
 These plots correspond to Fig.
@@
 
-```{r Fig_S4, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.width=7, fig.height=3, dev.args=list( compression="lzw" ), fig.cap="Origin of pairwise correlation plot structure.
+```{r Fig_S4, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.width=7, fig.height=3, fig.cap="Origin of pairwise correlation plot structure.
 (*A*) Each speciation and duplication node in the COMPARA trees is annotated with a clade name, and the age of that clade (shown on the x axis) is used as the age of the pairwise comparisons (Fig.
@@
 
-```{r Tau_v_max, cache=TRUE, echo=FALSE, warning=FALSE, fig.width=3.42, fig.height=3.42, dev.args=list( compression="lzw" ), message=FALSE, fig.cap="Genes with higher maximum observed expression tend to have lower Tau.
+```{r Tau_v_max, cache=TRUE, echo=FALSE, warning=FALSE, fig.width=3.42, fig.height=3.42, message=FALSE, fig.cap="Genes with higher maximum observed expression tend to have lower Tau.
 The blue line is the linear model for the relationship between the two."}
@@
 
+```{r wrap_up, echo=FALSE}
+	save.image()
+```
+
 \pagebreak

### 2017-10-08 manuscript.rmd
added p values to correlation coefficients

@@
 		pairwise_summary %>% 
-		filter( D=="N" ) %>% 
-		select( tau_a, tau_b ) %>% 
-		cor( ) %>% 
-		.[ 1, 2 ]
+		filter( D=="N" ) %$% 
+		cor.test( tau_a, tau_b )
 
@@
 		pairwise_summary %>% 
-		filter( D=="Y" ) %>% 
-		select( tau_a, tau_b ) %>% 
-		cor( ) %>% 
-		.[ 1, 2 ]
+		filter( D=="Y" ) %$% 
+		cor.test( tau_a, tau_b )
 
@@
 		filter( D=="Y" ) %>% 
-		filter( node_age <= max_speciation_age ) %>% 
-		select( tau_a, tau_b ) %>% 
-		cor( ) %>% 
-		.[ 1, 2 ]
+		filter( node_age <= max_speciation_age ) %$% 
+		cor.test( tau_a, tau_b )
 
@@
 1 [@KryuchkovaMostacci:2016iw], they present a higher Tau correlation coefficient between ortholog pairs than paralog pairs.
-We find the same here, with correlation coefficients of `r signif( ortholog_r, 2)` for orthologs and `r signif( paralog_r, 2)` for paralogs.
+We find the same here, with correlation coefficients of `r signif( ortholog_r$estimate, 2)` (p `r format_p(ortholog_r)`) for orthologs and `r signif( paralog_r$estimate, 2)` (p `r format_p( paralog_r )`) for paralogs.
 
@@
 When we consider only the duplication events the same age or younger than the oldest speciation event (*SI Appendix*, Fig.
-5*B*), the paralog correlation coefficient increases from `r signif( paralog_r, 2)` to `r signif( paralog_subset_r, 2)`.
-This is much closer to the ortholog correlation of `r signif( ortholog_r, 2) `.
+5*B*), the paralog correlation coefficient increases from `r signif( paralog_r$estimate, 2)` to `r signif( paralog_subset_r$estimate, 2)` (p `r format_p(paralog_subset_r)`).
+This is much closer to the ortholog correlation of `r signif( ortholog_r$estimate, 2) `.
 
@@
 		pairwise_sim_null_summary %>% 
-		filter( D=="N" ) %>% 
-		select( tau_a, tau_b ) %>% 
-		cor( ) %>% 
-		.[ 1, 2 ]
+		filter( D=="N" ) %$% 
+		cor.test( tau_a, tau_b )
 
@@
 		pairwise_sim_null_summary %>% 
-		filter( D=="Y" ) %>% 
-		select( tau_a, tau_b ) %>% 
-		cor( ) %>% 
-		.[ 1, 2 ]
+		filter( D=="Y" ) %$% 
+		cor.test( tau_a, tau_b )
 
@@
 		pairwise_sim_oc_summary %>% 
-		filter( D=="N" ) %>% 
-		select( tau_a, tau_b ) %>% 
-		cor( ) %>% 
-		.[ 1, 2 ]
+		filter( D=="N" ) %$% 
+		cor.test( tau_a, tau_b )
 
@@
 		pairwise_sim_oc_summary %>% 
-		filter( D=="Y" ) %>% 
-		select( tau_a, tau_b ) %>% 
-		cor( ) %>% 
-		.[ 1, 2 ]
+		filter( D=="Y" ) %$% 
+		cor.test( tau_a, tau_b )
 
@@
 2*A*).
-As in the original study, there is higher correlation coefficient across orthologs (`r signif( ortholog_r_sim_null, 2 )`) than paralogs (`r signif( paralog_r_sim_null, 2 )`) when not considering node age.
+As in the original study, there is higher correlation coefficient across orthologs (`r signif( ortholog_r_sim_null$estimate, 2 )`, p `r format_p(ortholog_r_sim_null)`) than paralogs (`r signif( paralog_r_sim_null$estimate, 2 )`, p `r format_p(paralog_r_sim_null)`) when not considering node age.
 Phylogenetic analysis of the data simulated under the null hypothesis (Fig.
@@
 2*E*) are nearly indistinguishable from the results under the null model (Fig.
-2*C*), and also have a higher correlation coefficient for orthologs (`r signif( ortholog_r_sim_oc, 2 )`) than paralogs (`r signif( paralog_r_sim_oc, 2 )`).
+2*C*), and also have a higher correlation coefficient for orthologs (`r signif( ortholog_r_sim_oc$estimate, 2 )`, p `r format_p( ortholog_r_sim_oc )`) than paralogs (`r signif( paralog_r_sim_oc$estimate, 2 )`, p `r format_p( paralog_r_sim_oc )`).
 The phylogenetic analysis of the ortholog conjecture simulation (Fig.

### 2017-10-08 functions.R

@@ -1,3 +1,16 @@
+
+#' Create a string from a p.value
+#' 
+#' @param x htest object, like that returned from cor.test()
+#' @param digits the number of significant digits
+#' @return A character object
+#' @export
+format_p = function ( x, digits=3 ){
+		fp <- format.pval( x$p.value, digits = max(1L, digits - 3L) )
+	out = if(substr(fp, 1L, 1L) == "<") fp else paste("=",fp)
+		return( out )
+}
+
 #' Parses nhx text to a treeio::treedata object
 #' 
 #' @param tree_text Character string repressenting nhx tree

### 2017-10-09 manuscript.rmd
fixed parsing errors
revised code to mend lines with extra spaces, resolves #1. PDF not yet reexecuted

@@
 		lines[ grepl( "^SEQ", lines, perl = TRUE ) ] %>%	# Get lines that start with SEQ
+	  str_replace( " \\(\\d+ of \\d+\\)", "" ) %>%      # Remove instances of '(1 of 2)' and '(2 of 2)'
+	  str_replace( "(([^ ]+ ){8})([^ ]+) ([^ ]+)", "\\1\\3_\\4" ) %>%      # Get rid of spaces in last field of some entries
 		str_c( collapse="\n" ) %>%												# Combine them into one string

### 2017-10-24 manuscript.rmd
added tree summary code - untested

--- 2017-10-06 manuscript.rmd
@@
     fig_caption: yes
-  word_document: default
+    fig_width: 6
+    fig_height: 4
+
 ---
@@
 		echo = TRUE,
-    dev="png", 
-    # dev.args=list( compression="lzw" ), # for tiff
-    dpi=600,
-    cache=TRUE
+		cache = TRUE
 	)
@@
 	# The rest can be installed from CRAN
+	library( tidyverse )
+	library( magrittr )
+	library( stringr )
+	library( parallel )
 	library( ape )
 	library( digest )
+	library( gridExtra )
 	library( geiger )
+	library( phytools )
 	library( ggrepel )
-	library( gridExtra )
-	library( magrittr )
-	library( parallel )
-	library( phytools )
-	library( stringr )
-	library( tidyverse )
 
@@
 
-Casey W. Dunn^1,2^*, Felipe Zapata^3^, Catriona Munro^1^, Stefan Siebert^4^, Andreas Hejnol^5^
-
-^1^ Department of Ecology and Evolutionary Biology, Brown University, Providence, RI, 02912, USA
-
-^2^ Current address: Department of Ecology and Evolutionary Biology, Yale University, New Haven, CT, 06511, USA
-
-^3^ Department of Ecology and Evolutionary Biology, University of California Los Angeles, Los Angeles, CA, 20095, USA
-
-^4^ Department of Molecular & Cellular Biology, University of California at Davis, Davis, CA, 95616, USA
-
-^5^ Sars International Centre for Marine Molecular Biology, University of Bergen, Bergen, 5006, Norway
-
-\* Corresponding author, casey.dunn@yale.edu
+Casey W. Dunn^1^*, Felipe Zapata^1,2^, Catriona Munro^1^, Stefan Siebert^1,3^, Andreas Hejnol^4^
+
+^1^ Department of Ecology and Evolutionary Biology, Brown University, Providence, RI, USA
+
+^2^ Current address: Department of Ecology and Evolutionary Biology, University of California Los Angeles, Los Angeles, CA, USA
+
+^3^ Current address: Department of Molecular & Cellular Biology, University of California at Davis, Davis, CA, USA
+
+^4^ Sars International Centre for Marine Molecular Biology, University of Bergen, Bergen, Norway
+
+\* Corresponding author, casey_dunn@brown.edu
 
@@
 Here we examine two recently published studies that compare gene expression across species with pairwise methods, and find reason to question the original conclusions of both.
-One study interpreted pairwise comparisons of gene expression as support for the ortholog conjecture, the hypothesis that orthologs tend to have more similar attributes (expression in this case) than do paralogs.
+One study interpreted pairwise comparisons of gene expression as support for the ortholog conjecture, the hypothesis that orthologs tend to be more similar than paralogs.
 The other study interpreted pairwise comparisons of embryonic gene expression across distantly related animals as evidence for a distinct evolutionary process that gave rise to phyla.
@@
 
-Over the last three decades, a rich set of phylogenetic comparative methods has been developed to address the challenges and opportunities of trait comparisons across species [@Felsenstein:1985ua; @Grafen:1989um; @Pagel:1999gc; @Revell:2012jd; @FitzJohn:2012ey; @Uyeda:2014jl; @Garamszegi:2014ew; @Jhwueng:2013fi; @Rohlfs:2015bd].
-A central challenge is the dependence of observations across species due to the evolutionary history of species -- more closely related species share more traits because they have a more recent common ancestor.
+Over the last three decades, a rich set of phylogenetic comparative methods has been developed to address the challenges and opportunities of trait comparisons across species [@Felsenstein:1985ua; @Grafen:1989um; @Pagel:1999gc; Revell:2012jd; @FitzJohn:2012ey; @Uyeda:2014jl; @Garamszegi:2014ew].
+A central challenge is the dependence of observations across species due to the evolutionary history of species -- more closely related species share many traits that evolved once in a common ancestor.
 This violates the fundamental assumption of observation independence in standard statistical methods.
@@
 They have largely been applied to morphological and ecological traits, but are just as relevant to functional genomics [@Barker:2005be].
-Even so, most comparative functional genomic studies have abstained from phylogenetic approaches and instead rely on multiple pairwise comparisons across species (Fig.
-1*A*).
+Even so, most comparative functional genomic studies have abstained from phylogenetic approaches and instead rely on multiple pairwise comparisons across species (Figure 1*A*).
 This leaves comparative functional genomic studies susceptible to statistical problems and is a missed opportunity to ask questions that are only accessible in an explicit phylogenetic context.
 
-Phylogenetic comparative methods account for evolutionary history and explicitly model trait change along the branches of evolutionary trees (*e.g.*, Fig.
-1*B*).
-The value of these methods relative to pairwise comparisons has been repeatedly shown in analyses of other types of character data [@Garland:2005ek; @Ricklefs:1996wb; @Chamberlain:2012vx; @OMeara:2012gq].
+Phylogenetic comparative methods account for evolutionary history and explicitly model trait change along the branches of evolutionary trees (*e.g.*, Figure 1*B*).
+The value of these methods relative to to pairwise comparisons has been repeatedly shown in analyses of other types of character data [@Ricklefs:1996wb; @Chamberlain:2012vx; @OMeara:2012gq].
 The application of these methods is illustrated by a study of leaf functional attributes for about 100 plant species [@Ackerly:1999iu].
@@
 Phylogenetic comparative methods capture this shift as change along a single branch deep in the tree, revealing that there is not a tendency of correlated change between these traits across the phylogeny.
-One reason that comparative functional genomic studies have not embraced phylogenetic approaches is that there has not yet been a concrete demonstration that pairwise and phylogenetic comparative methods can lead to different results when considering functional genomic data.
+One reason that comparative functional genomic sudies have not embraced phylogenetic approaches is that there has not yet been a concrete demonstration that pairwise and phylogenetic comparative methods can lead to different results when considering functional genomic data.
 Here we examine this issue by re-evaluating the pairwise comparisons in two recent studies that compared gene expression across species.
@@
 The first study, Kryuchkova-Mostacci and Robinson-Rechavi (KMRR) [@KryuchkovaMostacci:2016iw], analyzed multiple vertebrate expression datasets to test the ortholog conjecture - the hypothesis that orthologs tend to have more conserved attributes (specificity of expression across organs in this case) than do paralogs [@Nehrt:2011bm, @Koonin:2005fp].
-Using pairwise comparisons (Fig.
-1*A*), they found lower expression correlation between paralogs than between orthologs and interpreted this as strong support for the ortholog conjecture.
+Using pairwise comparisons (Figure 1*A*), they found lower expression correlation between paralogs than between orthologs and interpreted this as strong support for the ortholog conjecture.
 
@@
 The ortholog conjecture [@Nehrt:2011bm] is the proposition that orthologs (genes that diverged from each other due to a speciation event) have more similar attributes than do paralogs (genes that diverged from each other due to a gene duplication event).
-The ortholog can and has been applied to diverse attributes, including molecular sequence, biochemical function, and, as in the study considered here, expression.
 The ortholog conjecture has important biological and technical implications.
 It shapes our understanding of the functional diversity of gene families.
-It is also used to relate findings from well-studied genes to homologous genes that have not been investigated in detail.
-It has been applied to many traits of genes, from gene sequence to biochemical properties to expression.
+It is also used to relate findings from well-studied genes to related genes that have not been investigated in detail.
+It has be applied to many trait of genes, from gene sequence to biochemical properties to expression.
 While the ortholog conjecture describes a specific pattern of functional diversity across genes, it is also articulated as a hypothesis about the process of evolution-- that there is greater evolutionary change in gene attributes following a duplication event than a speciation event.
@@
 Analyses of domain structure were consistent with the ortholog conjecture [@Forslund:2011cw].
-There have been few tests of the ortholog conjecture with regards to gene expression [@Chen:2012jz], and KMRR is one of the most detailed to date.
+There have been few tests of the ortholog conjecture with regards to gene expression [@Chen:2012jz], and KMRR is the most thorough such expression study to date.
 
@@
 
-The KMRR analyses are based on pairwise comparisons (Fig.
-1*A*) between Tau within each gene family.
+The KMRR analyses are based on pairwise comparisons (Figure 1*A*) between Tau within each gene family.
 Rather than make every pairwise comparison within each gene tree, they considered only a subset of pairwise comparisons in each particular analysis.
@@
 
-Many phylogenetic comparative methods are now available for addressing a wide range of questions relevant to comparative functional genomics.
-Some recent methods have been developed specifically for analyses of expression.
-For example, the Expression Variance and Evolution (EVE) model compares the ratio of within-species expression variance to between-species evolutionary expression variance in an explicit phylogenetic framework [@Rohlfs:2015bd].
-It can test a variety of hypotheses, including lineage-specific expression level shifts.
-This method considers strict orthologs that all have the same gene tree, which is the same as the phylogeny of the species under consideration.
-However, for the re-analysis presented here we address different questions that consider a broad diversity of gene trees that include both speciation and different patterns of duplication.
-Phylogenetic independent contrasts [@Felsenstein:1985ua], the most widely used phylogenetic comparative method, is particularly well suited to this challenge.
-
 For each internal node in each gene tree, we calculated the phylogenetic independent contrast [@Felsenstein:1985ua] (PIC) of Tau.
-This is the difference in values of Tau for descendant nodes scaled by the expected variance, which is largely determined by the lengths of the two branches that connect the node to its two descendants (Fig.
-1*B*).
+This is the difference in values of Tau for descendant nodes scaled by the expected variance, which is largely determined by the lengths of the two branches that connect the node to its two descendants (Figure 1*B*).
 These contrasts were then annotated by whether each is made across a speciation or duplication event.
@@
 Our use of contrasts is a bit different-- we look for differences in evolutionary changes of one trait (differential expression) between two categories of nodes (speciation and duplication).
-This is similar to previous applications of contrasts to examine shifts in evolutionary rates between clades [@Garland:1992jl].
-Rather than compare the distributions of contrasts for nodes in different clades, as this previous work did, we compare the distributions of contrasts for nodes with different annotations (speciation or duplication).
 
@@
 	gene_trees = mclapply( tree_lines, parse_gene_trees, mc.cores=cores )
+
+	# Uncomment for development to look at a subset of gene trees
+	gene_trees = gene_trees[1:100]
 
@@
 
+```{r estimate_tau_model, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
+	# Estimate model parameters and add them to the tree objects
+	# Subcalls to fitContinuous() are parallel, so don't use mclapply here
+	# This takes a while, so do it once here and reuse the values as needed
+	gene_trees_calibrated = lapply( gene_trees_calibrated, add_model_parameters )
+```
+
 ```{r trees_to_contrasts, cache=TRUE, echo=FALSE, message=F, warning=F}
@@
 
+	# Collect all the tree statistics in a single tibble
+	tree_summary = summarize_trees( gene_trees_pic )
+
 	# Test if the mean rank of duplication pics is greater than speciation pics
@@
 
-We mapped the Tau values calculated by KMRR [@KryuchkovaMostacci:2016iw] for the Brawand *et al.* 2011 [@Brawand:2011du] dataset onto `r n_gene_trees %>% format(big.mark=",")` gene trees parsed from ENSEMBL Compara [@Herrero:2016jj].
+We mapped the Tau values calculated by KMRR [@KryuchkovaMostacci:2016iw] for the Brawand *et al.* 2011 [@Brawand:2011du] dataset onto `r n_gene_trees` gene trees parsed from ENSEMBL Compara [@Herrero:2016jj].
 These are the same pre-computed trees that the orthology/ paralogy annotations KMRR used are based on.
-`r length(gene_trees_pruned) %>% format(big.mark=",")` gene trees passed taxon sampling criteria (`r min_genes_with_expression %>% format(big.mark=",")` genes) after removing tips without Tau values and had at least one speciation event.
-Of these, `r length(gene_trees_calibrated)  %>% format(big.mark=",")` were successfully time calibrated.
-These calibrated trees were used to calculate phylogenetic independent contrasts for `r nodes_contrast %>% filter(D=="Y") %>% nrow() %>% format(big.mark=",")` duplication nodes and `r nodes_contrast %>% filter(D=="N") %>% nrow() %>% format(big.mark=",")` speciation nodes (Tables S1, S2).
-One of these trees is presented in *SI Appendix* Fig.
-4 to demonstrate the analysis.
+`r length(gene_trees_pruned)` gene trees passed taxon sampling criteria (`r min_genes_with_expression` genes) after removing tips without Tau values and had at least one speciation event.
+Of these, `r length(gene_trees_calibrated)` were successfully time calibrated.
+These calibrated trees were used to calculate phylogenetic independent contrasts for `r nodes_contrast %>% filter(D=="Y") %>% nrow()` duplication nodes and `r nodes_contrast %>% filter(D=="N") %>% nrow()` speciation nodes.
+One of these trees is presented in *SI Appendix* Figure 4 to demonstrate the analysis.
 
@@
 
-We did not find increased evolutionary change in expression following duplication events as compared to speciation events (Fig.
-2*B*).
+We did not find increased evolutionary change in expression following duplication events (Figure 2*B*).
 The Wilcoxon rank test does not reject the null hypothesis that the rate of evolution following duplications is the same as or less than the rate following speciation (p value = `r wilcox_test_result`).
@@
 We focused on two possible sources of bias, node depth and branch length.
-We found no evidence that either affected our results (*SI Appendix*, Fig.
-5).
+We found no evidence that either affected our results (*SI Appendix*, Figure 5).
 We also examined the sensitivity of the results to the calibration times applied to speciation events on the gene trees.
@@
 We added random noise to the calibration times in replicate analyses, and all still failed to reject the null hypothesis (*SI Appendix*).
-
-An additional concern is that within-species expression variation is not considered when making comparisons of Tau across homologous genes.
-Neglecting within species variation is expected to mislead phylogenetic comparative analyses in some cases [@Ives:2007cw; @Felsenstein:2008cl; @Rohlfs:2015bd].
-Since branch length describes expected trait variation [@Felsenstein:1985ua], within species variation can be modeled by extending the terminal branches of each gene phylogeny [@Ives:2007cw].
-We modeled within-species variation by extending each terminal branch by 7 million years.
-This value was selected because it corresponds to the age of the clade Hominini (defined by the most recent common ancestor of humans and chimpanzees), the shallowest node in the phylogeny of examined species.
-This provides an opportunity to assess the impact that extreme within-species variation (*i.e.*, as large as the variance between humans and chimpanzees and their most recent common ancestor) would have on our findings.
-These analyses still do not reject the null hypothesis, and do not provide support for the ortholog conjecture (*SI Appendix*, Fig.
-6).
 
@@
 This is important since we are only looking at a subset of the data they considered, the Brawand *et al.* 2011 [@Brawand:2011du] dataset for gene trees that could be successfully time calibrated.
-In their Fig.
-1 [@KryuchkovaMostacci:2016iw], they present a higher Tau correlation coefficient between ortholog pairs than paralog pairs.
+In their Figure 1 [@KryuchkovaMostacci:2016iw], they present a higher Tau correlation coefficient between ortholog pairs than paralog pairs.
 We find the same here, with correlation coefficients of `r signif( ortholog_r, 2)` for orthologs and `r signif( paralog_r, 2)` for paralogs.
@@
 
-To test the hypothesis that ancient duplications that precede the oldest speciation event (*SI Appendix*, Fig.
-5*A*) impact the lower correlation of Tau between paralogs than between orthologs, we removed them.
-When we consider only the duplication events the same age or younger than the oldest speciation event (*SI Appendix*, Fig.
-5*B*), the paralog correlation coefficient increases from `r signif( paralog_r, 2)` to `r signif( paralog_subset_r, 2)`.
+To test the hypothesis that ancient duplications that precede the oldest speciation event (*SI Appendix*, Figure 5*A*) impact the lower correlation of Tau between paralogs than between orthologs, we removed them.
+When we consider only the duplication events the same age or younger than the oldest speciation event (*SI Appendix*, Figure 5*B*), the paralog correlation coefficient increases from `r signif( paralog_r, 2)` to `r signif( paralog_subset_r, 2)`.
 This is much closer to the ortholog correlation of `r signif( ortholog_r, 2) `.
-
-```{r estimate_tau_model, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
-	# Estimate model parameters and add them to the tree objects
-	# Subcalls to fitContinuous() are parallel, so don't use mclapply here
-	# This takes a while, so do it once here and reuse the values as needed
-	gene_trees_calibrated = lapply( gene_trees_calibrated, add_model_parameters )
-```
 
@@
 
-```{r Fig_2, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.width=7, fig.height=6.5, fig.cap="Pairwise (left column - *A, C, E*) and phylogenetic (right column - *B, D, F*) analyses of the original data (first row - *A, B*), data simulated under the null hypothesis (second row - *C, D*), and data simulated under the ortholog conjecture (third row- *E, F*).
+```{r pairwise_and_phylo_plots, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.width=6.5, fig.height=6.5, fig.cap="Pairwise (left column - *A, C, E*) and phylogenetic (right column - *B, D, F*) analyses of the original data (first row - *A, B*), data simulated under the null hypothesis (second row - *C, D*), and data simulated under the ortholog conjecture (third row- *E, F*).
 In the pairwise plots, each point indicates the correlation coefficient of Tau for a set of pairwise comparisons annotated with a specific node name (e.g, Primates) and event type (speciation or duplication, giving rise respectively to orthologs and paralogs).
@@
 A horizontal line at 0 would indicate that the density distributions are identical.
-The top left pane (*A*) reproduces the pattern presented in Fig.
-2*A* of KMRR [@KryuchkovaMostacci:2016iw] of higher correlation across speciation events than duplication events, which they took as evidence of the ortholog conjecture.
-The recovery of a similar  pattern under both simulations (*C, E*) indicates that this pairwise approach does not make distinct testable predictions.
+The top left pane (*A*) reproduces the pattern presented in Figure 2A of KMRR [@KryuchkovaMostacci:2016iw] of higher correlation across speciation events than duplication events, which they took as evidence of the ortholog conjecture.
+The recovery of a similar  pattern under both simulations (*C, E*) indicates that it this pairwise approach does not make distinct testable predictions.
 The phylogenetic analysis of the original data (*B*) does not show an excess of larger contrasts for duplication events and does not reject the null hypothesis, providing no support for the ortholog conjecture.
@@
 The KMRR study did investigate the impact of node age on correlation, but in a different way.
-In their Fig.
-2 [@KryuchkovaMostacci:2016iw], they grouped orthologs and paralogs according to the ENSEMBL node name of their most recent common ancestor, and plotted the correlation of Tau for each of these groups by the node age.
+In their Figure 2 [@KryuchkovaMostacci:2016iw], they grouped orthologs and paralogs according to the ENSEMBL node name of their most recent common ancestor, and plotted the correlation of Tau for each of these groups by the node age.
 They found that across the investigated range of node ages, ortholog pairs have higher Tau correlation than paralogs.
-We confirmed that we can replicate this result (Fig.
-2*A*).
+We confirmed that we can replicate this result (Figure 2*A*).
 There are several difficulties with interpreting this plot, though.
@@
 Second, the correlation for each group is based on multiple non-independent pairwise comparisons.
-Third, instead of using the actual age of duplication events, the age of an adjacent named node in the species tree is used.
-
-To better understand this plot (Fig.
-2*A*), we performed simulations of Tau on the calibrated gene trees.
+
+To better understand this plot (Figure 2*A*), we performed simulations of Tau on the calibrated gene trees and then regenerated the figures.
 We did not modify the gene tree topologies or their inferred histories of duplication and loss.
 First, we simulated the evolution of Tau under the null model that it evolves at the same rate following duplication and speciation events.
-Under the null model, the plot of correlation coefficient to node age (Fig.
-2*C*) is very similar as for the observed data (Fig.
-2*A*).
+Under the null model, the plot of correlation coefficient to node age (Figure 2*C*) is very similar as for the observed data (Figure 2*A*).
 As in the original study, there is higher correlation coefficient across orthologs (`r signif( ortholog_r_sim_null, 2 )`) than paralogs (`r signif( paralog_r_sim_null, 2 )`) when not considering node age.
-Phylogenetic analysis of the data simulated under the null hypothesis (Fig.
-2*D*) do not reject the null hypothesis (Wilcoxon p = $`r wilcox_oc( nodes_sim_null_contrast ) %>% round( 3 )`$), as expected.
+Phylogenetic analysis of the data simulated under the null hypothesis (Figure 2*D*) do not reject the null hypothesis (Wilcoxon p = $`r wilcox_oc( nodes_sim_null_contrast ) %>% round( 3 )`$), as expected.
 
 We next simulated the evolution of Tau under the ortholog conjecture, where the rate of evolution of Tau following duplication was `r dup_adjust` fold the rate following speciation.
-The pairwise results of this heterogeneous model (Fig.
-2*E*) are nearly indistinguishable from the results under the null model (Fig.
-2*C*), and also have a higher correlation coefficient for orthologs (`r signif( ortholog_r_sim_oc, 2 )`) than paralogs (`r signif( paralog_r_sim_oc, 2 )`).
-The phylogenetic analysis of the ortholog conjecture simulation (Fig.
-2*F*) does reject the null hypothesis (Wilcoxon p = $`r signif( wilcox_oc( nodes_sim_oc_contrast ), 2 )`$).
+The pairwise results of this heterogeneous model (Figure 2*E*) are nearly indistinguishable from the results under the null model (Figure 2*C*), and also have a higher correlation coefficient for orthologs (`r signif( ortholog_r_sim_oc, 2 )`) than paralogs (`r signif( paralog_r_sim_oc, 2 )`).
+The phylogenetic analysis of the ortholog conjecture simulation (Figure 2*F*) does reject the null hypothesis (Wilcoxon p = $`r signif( wilcox_oc( nodes_sim_oc_contrast ), 2 )`$).
 
@@
 The pairwise comparisons used by KMRR cannot distinguish between the null hypothesis and ortholog conjecture.
-The pairwise results are strikingly similar under both hypotheses (Fig.
-2*C*, 2*E*).
+The pairwise results are strikingly similar under both hypotheses (Figure 2*C*, 2*E*).
 These simulations also serve to validate the phylogenetic methods applied to this problem.
@@
 
-Since greater rates of expression evolution following duplication do not explain the lower correlation of Tau values in the pairwise comparison plots (Fig.
-2 *A, C, E*), this pattern must reflect some other property that is shared across these analyses.
+Since greater rates of expression evolution following duplication do not explain the lower correlation of Tau values in the pairwise comparison plots (Figure 2 *A, C, E*), this pattern must reflect some other property that is shared across these analyses.
 We found that the lower correlations for duplication comparisons can largely be explained by the greater variance in the age of duplication nodes.
-Each point in the pairwise correlation plot (Fig.
-2 *A, C, E*) summarizes multiple pairwise comparisons across nodes of a given event type (speciation or duplication, as indicated by color) and clade in the species tree (Theria, Mammalia, Amniota, etc..., as indicated by the age of the clade along the x axis).
+Each point in the pairwise correlation plot (Figure 2 *A, C, E*) summarizes multiple pairwise comparisons across nodes of a given event type (speciation or duplication, as indicated by color) and clade in the species tree (Theria, Mammalia, Amniota, etc..., as indicated by the age of the clade along the x axis).
 These clade names in turn are derived from the node annotations in the ENSEMBL Compara trees, which apply clade names to both speciation and duplication nodes.
@@
 Neighboring clade ages are therefore a very rough approximation of the age of duplication events.
-This is apparent when node ages derived from calibrated trees are plotted against the age of the clade annotations for each node (*SI Appendix*, Fig.
-7*A*).
+This is apparent when node ages derived from calibrated trees are plotted against the age of the clade annotations for each node (*SI Appendix*, Figure 6*A*).
 Because duplication events that are all annotated with the same clade name can occur at very different times, pairwise comparisons across these nodes capture evolutionary changes in Tau across very different different branch lengths.
 They therefore have lower correlation than pairwise comparisons made across speciation nodes.
-Of the `r n_dup_original %>% format(big.mark=",")` duplication nodes considered here, `r n_dup_remaining %>% format(big.mark=",")` have a node age (as determined by the time calibrated trees) within `r age_window * 100`% of their clade annotation.
-If only these duplication events closest to their annotated clade nodes are retained, the pattern of lower correlation of Tau evolution for duplication events disappears for the dataset simulated under the null model (*SI Appendix*, Fig.
-7*B*).
+Of the `r n_dup_original` duplication nodes considered here, `r n_dup_remaining` have a node age (as determined by the time calibrated trees) within `r age_window * 100`% of their clade annotation.
+If only these duplication events closest to their annotated clade nodes are retained, the pattern of lower correlation of Tau evolution for duplication events disappears for the dataset simulated under the null model (*SI Appendix*, Figure 5*B*).
 This correction, though, comes at the high cost of discarding `r round( (n_dup_original - n_dup_remaining )/ n_dup_original * 100, 1 )`% of duplication nodes.
@@
 While some studies have presented support for the ortholog conjecture, our results are consistent with multiple studies that have not [@Studer:2009bm; @Nehrt:2011bm; @Yanai:2004hk].
-Since the ortholog conjecture can be applied to many different gene attributes at both the genotype level (*e.g.*, molecular sequence) and phenotype level (*e.g.*, activity, biochemical function, binding partners, and expression), it is naturally a heterogeneous and potentially idiosyncratic topic [@Soria:2014bv].
-For example, a recent integrated analysis of multiple gene traits after duplication found extensive divergence in protein sequence but less divergence in expression than expected [@Soria:2014bv].
 
@@
 Under this neutral conjecture, phylogenetic distance is a better predictor of the similarity of gene function than is the history of gene duplication and speciation.
-The ortholog conjecture does not have to be an all or nothing process, though.
+The ortholog conjecture does not have to be an all or nothing question, though.
 It may be the case that the rates of phenotypic evolution following duplication may be greater than that following duplication in some organisms, gene families, and evolutionary processes [@Gabaldon:2013fi].
@@
 The five least-similar mid phase scores were all from the ctenophore.
+Published KS p-values, based on duplicated data, are in parentheses.
 The inset ctenophore image is by S.
@@
 They then calculated a similarity score for each temporal phase for each species pair based on the fraction of genes that exhibited the same patterns in each species.
-The distributions of similarity scores are plotted in their Fig.
-4d [@Levin:2016bp], and their Kolmogorov–Smirnov (KS) tests indicated that the early distribution and late distribution were each significantly different from mid distribution (P < $10^{-6}$ and P < $10^{-12}$, respectively).
+The distributions of similarity scores are plotted in their Figure 4d [@Levin:2016bp], and their Kolmogorov–Smirnov (KS) tests indicated that the early distribution and late distribution were each significantly different from mid distribution (P < $10^{-6}$ and P < $10^{-12}$, respectively).
 This is the support they presented for the inverse hourglass model.
@@
 
-We examined the matrix of pairwise comparisons used as the base for the KS tests and Fig.
-4d in *Levin et al.* [@Levin:2016bp], and thus as support for the "inverse hourglass" model.
+We examined the matrix of pairwise comparisons used as the base for the KS tests and Figure 4d in *Levin et al.* [@Levin:2016bp], and thus as support for the "inverse hourglass" model.
 We found several problems resulting from the use of multiple pairwise comparisons.
@@
 Once we addressed the issues above with the implementation of pairwise comparisons, we were able to explore more general issues that can be a problem when making multiple pairwise comparisons between species.
-We found that all five of the lowest values in the mid phase distribution (Fig.
-3*A*) are for pairwise comparisons that include the ctenophore (comb jelly).
-When the nine pairwise comparisons that include the ctenophore are removed, there is no significant difference between the early phase and mid phase distributions (p = 0.14 for the early-mid comparison and p < $10^{-5}$ for the late-mid comparison) and no support for the inverse hourglass (Fig.
-3*B*).
+We found that all five of the lowest values in the mid phase distribution (Figure 3*A*) are for pairwise comparisons that include the ctenophore (comb jelly).
+When the nine pairwise comparisons that include the ctenophore are removed, there is no significant difference between the early phase and mid phase distributions (p = 0.14 for the early-mid comparison and p < $10^{-5}$ for the late-mid comparison) and no support for the inverse hourglass (Figure 3b).
 This highlights a well understood property of pairwise comparisons across species [@Felsenstein:1985ua; @Ackerly:1999iu]: evolutionary changes along a given branch, like those along the ctenophore branch, impact each of the multiple pairwise comparisons that includes that branch.
-The pairwise comparisons are therefore not independent - different pairwise comparisons are impacted by changes along some of the same branches (Fig.
-1*A*).
+The pairwise comparisons are therefore not independent - different pairwise comparisons are impacted by changes along some of the same branches (Figure 1*A*).
 This can give the impression of a general pattern across the tree that is instead specific to changes along one part of the tree.
@@
 While we demonstrate problems with pairwise comparisons that impacts the Levin *et al.* analysis, we did not perform a phylogenetic reanalysis of this study, as we did for the KMRR study.
-This is because the similarity metric computed in the pairwise comparisons of Levin *et al.* is not suitable for phylogenetic analysis as it is based on different genes for different species pairs, and therefore cannot be modeled across the phylogeny.
+This is because the similarity metric computed in the pairwise comparisons of Levin *et al.* is not suitable for phylogenetic analysis.
+One issue with the similarity metric is that it is based on different genes for different species pairs, and therefore is not a trait whose evolution can be modeled across the phylogeny.
 A full phylogenetic reanalysis would be possible using upstream analysis products to re-derive new expression summary statistics.
@@
 
-#### Challenges and opportunities
-
-There are interesting new challenges at the interdisciplinary interface between phylogenetic comparative methods and functional genomics that will need to be addressed to fully realize the potential of integrating these approaches.
+Our results highlight the importance of explicitly incorporating information about phylogenetic relationships when comparing functional genomic traits across species.
+Some of the most widely used phylogenetic comparative methods [@Felsenstein:1985ua; @Grafen:1989um] are already directly applicable to comparative functional genomic studies.
+There are also interesting new challenges at this interdisciplinary interface that will need to be addressed to fully realize the potential of phylogenetic comparative functional genomic studies.
 One such challenge is that most phylogenetic comparative analyses of covariance between traits have been developed to address problems with many more species (e.g., dozens or more) relative to the number of traits being examined.
@@
 We outlined these challenges and potential solutions in the context of gene expression elsewhere [@Dunn:2013cpa].
-
-Another challenge is parameterizing differential expression in such a way that they can be analyzed in a phylogenetic context.
-Read counts generated by RNA-seq expression studies cannot be directly compared across species [@Dunn:2013cpa].
+In the same manuscript we also considered another issue of relevance here -- the read counts generated by RNA-seq expression studies cannot be directly compared across species.
 This is because there are various species-specific technical factors that can be mistaken for differences in expression across species.
@@
 
-Branch lengths are fundamental to phylogenetic comparative methods, since the null expectation is that there are more changes along longer branches.
-In this study we applied four separate modifications of branch lengths.
-First, we time calibrated the gene trees so that nodes corresponding to the same speciation events had the same age across all trees.
-This is critical to making the comparisons of functional genomic traits equivalent across genes.
-Second, phylogenetic independent contrasts (including the implementation we use here) adjust internal branch lengths to accommodate the variation associated with estimating character states at internal nodes [@Felsenstein:1985ua].
-Third, we randomized the calibration times to determine the sensitivity of the analyses to calibration times and the branch length estimations they impact (*SI Appendix*).
-Fourth, we extended terminal branch lengths to understand the potential impact of within-species variation (*SI Appendix*, Fig.
-6).
-The first two of these modifications are technical requirements for the application of phylogenetic comparative methods.
-The second two explore the sensitivity of the analyses to branch length and the information they convey, and find that the results are robust to variation in branch length.
-This is encouraging for the future of phylogenetic comparative functional genomics, given that one concern of applying these methods is the challenge of estimating branch lengths.
-
-#### Additional approaches and future directions
-
-We are not the first to apply phylogenetic comparative methods to functional genomic data.
-While the vast majority of comparative functional genomic studies have used standard pairwise similarity methods, a small number of comparative functional genomic studies have employed phylogenetic comparative approaches [@Oakley:2005ky; @Eng:2009gp; @Chang:2014hq; @Whitney:2011fw].
-For instance, a phylogenetic ANOVA [@Rohlfs:2015bd] of the evolution of gene expression in strict orthologs improves statistical power and drastically reduces the rate of false positives relative to pairwise approaches.
-
-Some of the most widely used phylogenetic comparative methods [@Felsenstein:1985ua; @Grafen:1989um] are already directly applicable to comparative functional genomic studies.
 Recent advances in phylogenetic comparative methods are particularly well suited to addressing questions about the evolution of functional genomic traits.
@@
 
+We are not the first to apply phylogenetic comparative methods to functional genomic data.
+While the vast majority of comparative functional genomic studies have used standard pairwise similarity methods, a small number of comparative functional genomic studies have employed phylogenetic comparative approaches [@Oakley:2005ky; @Eng:2009gp; @Chang:2014hq].
+For instance, a phylogenetic ANOVA [@Rohlfs:2015bd] of the evolution of gene expression improves statistical power and drastically reduces the rate of false positives relative to pairwise approaches.
+
+Addressing the statistical dependence of pairwise comparisons is not the only advantage of using phylogenetic comparative methods for functional genomic analyses.
+Another problem with the pairwise comparisons is that, except at the tips, they summarize changes along many branches in the phylogeny.
+Two paralogs that diverged from a duplication event deep in the tree may have many subsequent duplication and speciation events, and changes along all these branches will impact the final pairwise comparison.
+Phylogenetic methods have the advantage of isolating the changes under consideration (Figure 1*B*).
+Phylogenetic methods therefore avoid diluting the change that occurs along the branches that follow the node in question with changes along all subsequent branches.
+There may still be missing speciation events, due to extinction and incomplete taxon sampling, and missing duplication events, due to gene loss, but these omissions affect both methods.
+
 ## Conclusions
 
-The problems we identify with pairwise comparisons in two recent functional genomics studies indicate that there are likely to be similar problems in other studies that use these methods.
+The fact that the first two comparative functional genomic studies we reanalyzed show serious problems with pairwise comparisons indicates that there are likely to be similar problems in other studies that use these methods.
 Future studies that compare functional genomic data across species will be compromised if they continue to use pairwise methods.
@@
 
-These analyses also illustrate how important it is to not conflate evolutionary patterns with the processes that generated them.
+These analyses illustrate how important it is to not conflate evolutionary patterns with the processes that generated them.
 Finding a pattern wherein paralogs tend to be more different than orthologs is not evidence that there are different processes by which orthologs and paralogs evolve.
@@
 After parsing the trees from the Compara file with treeio, which was recently split from ggtree [@Yu:2016fo], we added Tau estimates generated by the KMRR `Rscript.R` to the tree data objects.
-We then pruned away tips without expression data, retaining only the trees with `r min_genes_with_expression %>% format(big.mark=",")` or more tips.
+We then pruned away tips without expression data, retaining only the trees with `r min_genes_with_expression` or more tips.
 We also only retained trees with one or more speciation events, as speciation events are required for calibration steps.
@@
 Levin *et al.* helpfully provided data and clarification on methods.
-We obtained the matrix of pairwise scores that underlies their Fig.
-4d and confirmed we could reproduce their published results.
+We obtained the matrix of pairwise scores that underlies their Figure 4d and confirmed we could reproduce their published results.
 We then removed duplicate rows, applied the Wilcoxon test in place of the Kolmogorov-Smirnov test, and identified ctenophores as overrepresented among the low outliers in the mid-developmental transition column.
@@
 
-```{r Fig_S1, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, eval=TRUE, fig.width=7, fig.height=6.5, fig.cap="One of the gene trees from the phylogenetic analysis.
+```{r example_trees, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, eval=TRUE, fig.cap="One of the gene trees from the phylogenetic analysis.
 The value of Tau (expression specificity) is indicated by the sizes of the circles at the tips of the tree.
@@
 
-While the age of speciation nodes is constrained, duplication nodes can be much older and therefore have a wider range of ages (*SI Appendix*, Fig.
-5*A*).
+While the age of speciation nodes is constrained, duplication nodes can be much older and therefore have a wider range of ages (*SI Appendix*, Figure 5*A*).
 This is because many gene duplication events are older than the most recent common ancestor of the species in the study.
@@
 Gene tree estimation errors, for example, tend to lead to the overestimation of deep duplications [@Hahn:2007cj].
-If independent contrast values also tended to be lower at greater node depth, it could interact with the preponderance of duplications at greater depth to create a pattern of lower contrasts associated with duplication events.
+If independent contrast values also tended to to be lower at greater node depth, it could interact with the preponderance of duplications at greater depth to create a pattern of lower contrasts associated with duplication events.
 To test for such an effect, we remove duplication nodes that are older than the oldest speciation node.
-The general results are unchanged and this reduced dataset does not reject the null hypothesis that the rate of evolution following duplications is the same as or less than the rate following speciation (*SI Appendix*, Fig.
-5*B*).
-
-The independent contrast across a node is the amount of change observed between the daughter nodes, scaled by the expected variance [@Felsenstein:1985ua].
+The general results are unchanged and this reduced dataset does not reject the null hypothesis that the rate of evolution following duplications is the same as or less than the rate following speciation (*SI Appendix*, Figure 5*B*).
+
+The independent contrast across a node is the amount of change observed between the daughter notes, scaled by the expected variance [@Felsenstein:1985ua].
 The expected variance is principally determined by the lengths of the branches leading from the node to these daughters.
@@
 This is because the same difference across a shorter total branch length indicates a greater rate of evolutionary change.
-The expected variance of contrasts for speciation nodes is constrained by the branch lengths on the species tree, but the expected variance of contrasts for duplications has a much wider range (*SI Appendix*, Fig.
-5*C*).
+The expected variance of contrasts for speciation nodes is constrained by the branch lengths on the species tree, but the expected variance of contrasts for duplications has a much wider range (*SI Appendix*, Figure 5*C*).
 This could lead to biases if the lengths of branches that descend from duplication nodes tend to be overestimated.
 We therefore examined only the contrasts that fell within in the range of expected variance seen for speciation contrasts, excluding duplication contrasts that fall outside of this range.
-This reanalysis does not reject the null hypothesis either (*SI Appendix*, Fig.
-5*D*), indicating that branch length bias is not responsible for the result.
-
-```{r Fig_S2, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.width=7, fig.height=6.5, fig.cap="Investigation of possible ascertainment biases.
+This reanalysis does not reject the null hypothesis either (*SI Appendix*, Figure 5*D*), indicating that branch length bias is not responsible for the result.
+
+```{r ascertainment_plot, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, dev="png", dpi=600, fig.cap="Investigation of possible ascertainment biases.
 (*A*) Magnitude of independent contrasts plotted against node age.
@@
 
-```{r Fig_S3, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.width=7, fig.height=3, fig.cap="Investigation of the potential impact of within-species variance.
-These plots correspond to Fig.
-2*A* and 2*B*, but with increased expected within-species variation modeled by increasing the terminal gene phylogeny branch lengths.
-(*A*) Pairwise comparisons still recover higher correlation across speciation events than duplication events.
-(*B*) There is not an excess of larger contrasts for duplication events and the analysis does not reject the null hypothesis.
-Modeling a large amount of within-species variation does not lead to support for the ortholog conjecture."}
-
-	x = calibration_times$age[calibration_times$clade == "Hominini"]
-	gene_trees_extended = mclapply( gene_trees_calibrated, extend_nhx, x=x, mc.cores=cores )
-	gene_trees_extended %<>% add_pics_to_trees()
-	nodes_extended_contrast = gene_trees_extended %>% summarize_contrasts()
-
-	pairwise_extended_summary = 
-			mclapply( gene_trees_extended, get_pairwise_summary, mc.cores=cores ) %>% 
-			bind_rows()
-
-	g_pair_extended = 
-		pairwise_extended_summary %>% 
-			filter( ! is.na( D ) ) %>%
-			group_by( Event, label ) %>% 
-			summarise( cor.test( tau_a, tau_b )$estimate, n=n() ) %>% 
-			rename( rho=`cor.test(tau_a, tau_b)$estimate` ) %>%
-			rename( clade = label ) %>% 
-			left_join( calibration_times, clade=clade ) %>%
-			ggplot() + 
-				geom_point( aes( x=age, y=rho, col=Event ), size=3 ) +
-				theme_classic() +
-				xlab( "Clade Age (million years)" ) +
-				ylab( "Correlation Coefficient" ) +
-				ylim( 0, 1 ) +
-				ggtitle( "A Pairwise" ) +
-				guides( colour = guide_legend( override.aes = list( shape = 15 ) ) ) +
-				theme( 
-				legend.justification=c( 1, 1 ), 
-				legend.position=c( 1, 1 ), 
-				legend.title=element_blank() )
-
-	grid.arrange(
-		g_pair_extended,
-		make_contrast_plot( nodes_extended_contrast ) + 
-			ggtitle( "B Phylogenetic" ),
-		ncol = 2
-	)
-```
-
 #### Investigation of sensitivity to calibration times
@@
 
-```{r Fig_S4, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.width=7, fig.height=3, fig.cap="Origin of pairwise correlation plot structure.
-(*A*) Each speciation and duplication node in the COMPARA trees is annotated with a clade name, and the age of that clade (shown on the x axis) is used as the age of the pairwise comparisons (Fig.
-2 *A, C, E*, KMRR [@KryuchkovaMostacci:2016iw] Fig.
-2).
+```{r cor_structure_plot, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, dev="png", dpi=600, fig.width=6.5, fig.height=6.5, fig.cap="Origin of pairwise correlation plot structure.
+(*A*) Each speciation and duplication node in the COMPARA trees is annotated with a clade name, and the age of that clade (shown on the x axis) is used as the age of the pairwise comparisons (Figure 2 *A, C, E*, KMRR [@KryuchkovaMostacci:2016iw] Figure 2).
 The clade ages have a clear correspondence to the speciation node ages (red).
@@
 		g_sub,
-		ncol = 2
+		ncol = 1
 		)
@@
 
-```{r Tau_v_max, cache=TRUE, echo=FALSE, warning=FALSE, fig.width=3.42, fig.height=3.42, message=FALSE, fig.cap="Genes with higher maximum observed expression tend to have lower Tau.
+```{r Tau_v_max, cache=TRUE, echo=FALSE, warning=FALSE, dev="png", dpi=300, message=FALSE, fig.cap="Genes with higher maximum observed expression tend to have lower Tau.
 The blue line is the linear model for the relationship between the two."}
@@
 
-KMRR [@KryuchkovaMostacci:2016iw] Fig.
-3 differs from their other analyses in that it presents independent changes in expression between triplets of genes.
+KMRR [@KryuchkovaMostacci:2016iw] Figure 3 differs from their other analyses in that it presents independent changes in expression between triplets of genes.
 Each triplet has two paralogs that arose from a duplication event and one unduplicated homolog.
@@
 From this they conclude that following duplication there is a common trajectory of evolutionary change, where one paralog evolves to have lower expression and also become more tissue specific.
-There is, however, a negative relationship between Tau and maximum expression across all genes (*SI Appendix* Fig.
-8, $r^2=`r signif( tau_v_max_lm$r.squared, 3 )`$, $slope=`r signif( tau_v_max_lm$coefficients[2,1], 3 )`$, $p=`r signif( tau_v_max_lm$coefficients[2,4], 3 )`$).
+There is, however, a negative relationship between Tau and maximum expression across all genes (*SI Appendix* Figure 7, $r^2=`r signif( tau_v_max_lm$r.squared, 3 )`$, $slope=`r signif( tau_v_max_lm$coefficients[2,1], 3 )`$, $p=`r signif( tau_v_max_lm$coefficients[2,4], 3 )`$).
 A simpler explanation for their plot is that it reflects the global tendency for Tau to decrease with maximum expression.
@@
 
-```{r wrap_up, echo=FALSE}
-	save.image()
-```
-
 \pagebreak

### 2017-10-24 functions.R

@@ -309,6 +309,47 @@ summarize_contrasts = function( gene_trees_pic ) {
 }
 
 
+#' Prepare summary statistics for each tree
+#' 
+#' @param gene_trees A list of phylogenetic trees with pics
+#' as treeio::treedata objects 
+#' @return A tibble with combined data
+#' @export
+summarize_trees = function( gene_trees_pic ) {
+	
+	tree_summary = 
+		lapply( 
+			gene_trees_pic, 
+			function( nhx ){
+				tags = nhx@data
+				phy = nhx@phylo
+				x = nhx@data$Tau[1:length(phy$tip.label)]
+				names( x ) = phy$tip.label
+				
+				tibble(
+					gene = digest( nhx ),
+					n_tips = length( x ),
+					tau_mean = mean( x ),
+					tau_var = var( x ),
+					K = phylosig( phy, x, method="K" ),
+					z0_bm = nhx@phylo$model_bm$opt$z0,
+					z0_ou = nhx@phylo$model_ou$opt$z0,
+					sigsq_bm = nhx@phylo$model_bm$opt$sigsq,
+					sigsq_ou = nhx@phylo$model_ou$opt$sigsq,
+					alpha_ou = nhx@phylo$model_ou$opt$alpha,
+					aic_bm = nhx@phylo$model_bm$opt$aic,
+					aic_ou = nhx@phylo$model_ou$opt$aic,
+					lnL_bm = nhx@phylo$model_bm$opt$lnL,
+					lnL_ou = nhx@phylo$model_ou$opt$lnL
+				)
+			}
+		) %>%
+		bind_rows()
+	
+	return( tree_summary )
+}
+
+
 #' Calibrate a list of gene trees given a set of speciation node dates
 #' 
 #' @param gene_trees_pruned A list of phylogenetic trees as treeio::treedata objects 
@@ -381,10 +422,12 @@ add_model_parameters = function( nhx, ... ) {
 	phy = nhx@phylo
 	tau_original = nhx@data$Tau [ 1:length( phy$tip.label ) ]
 	names( tau_original ) = phy$tip.label
-	brownian_model = fitContinuous( phy, tau_original, ... )
-	
-	nhx@phylo$model$opt$z0 = brownian_model$opt$z0
-	nhx@phylo$model$opt$sigsq = brownian_model$opt$sigsq
+
+	brownian_model = fitContinuous( phy, tau_original, model="BM", ... )
[…]

### 2017-10-24 functions.R
look at all trees

@@ -344,7 +344,7 @@ summarize_trees = function( gene_trees_pic ) {
 					n_tips = length( x ),
 					tau_mean = mean( x ),
 					tau_var = var( x ),
-					K = phylosig( phy, x, method="K" ),
+					K = phytools::phylosig( phy, x, method="K" ),
 					z0_bm = nhx@phylo$model_bm$opt$z0,
 					z0_ou = nhx@phylo$model_ou$opt$z0,
 					sigsq_bm = nhx@phylo$model_bm$opt$sigsq,

### 2017-11-02 manuscript.rmd
look at all trees
added phylogenetic signal and AIC code

--- 2017-10-09 manuscript.rmd
@@
 
+	# Uncomment for development to look at a subset of gene trees
+	# gene_trees = gene_trees[1:100]
+
 	n_gene_trees = length( gene_trees )
@@
 	nodes_contrast = summarize_contrasts( gene_trees_pic )
+
+	# Collect all the tree statistics in a single tibble
+	tree_summary = summarize_trees( gene_trees_pic )
 
@@
 This is much closer to the ortholog correlation of `r signif( ortholog_r$estimate, 2) `.
-
-```{r estimate_tau_model, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
-	# Estimate model parameters and add them to the tree objects
-	# Subcalls to fitContinuous() are parallel, so don't use mclapply here
-	# This takes a while, so do it once here and reuse the values as needed
-	gene_trees_calibrated = lapply( gene_trees_calibrated, add_model_parameters )
-```
 
@@
 
+#### Model selection and sensitivity
+
+```{r phylogenetic_signal, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
+
+	k_thresh = quantile( tree_summary$K, na.rm=TRUE )[3] # third quantile is the median
+	k_percentile = names(k_thresh)
+	genes_pass_k = tree_summary$gene[ tree_summary$K > k_thresh ]
+
+	nodes_contrast_filtered_k = 
+		nodes_contrast %>% 
+			filter( gene %in% genes_pass_k )
+
+	wilcox_test_result_k = wilcox_oc( nodes_contrast_filtered_k )
+
+```
+
+We removed the `r k_percentile` of genes with the lowest phylogenetic signal, corresponding to a K cutoff for retained genes of greater than `r signif( k_thresh, 3)`.
+Repeating the phylogenetic analyses on this subset of genes with strongest phylogenetic signal for Tau still do not reject the null hypothesis (Wilcoxon p = $`r  wilcox_test_result_k %>% round( 3 )`$).
+
+```{r model_selection, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
+
+	dAIC_thresh = 2
+	AIC_min = pmin( tree_summary$aic_bm, tree_summary$aic_ou )
+	tree_summary$dAIC = tree_summary$aic_ou - tree_summary$aic_bm
+	tree_summary$dAIC_bm = tree_summary$aic_bm - AIC_min
+	tree_summary$dAIC_ou = tree_summary$aic_ou - AIC_min
+	genes_pass_bm = tree_summary$gene[ tree_summary$dAIC_bm < dAIC_thresh ]
+	genes_pass_ou = tree_summary$gene[ tree_summary$dAIC_ou < dAIC_thresh ]
+
+	nodes_contrast_filtered_bm = 
+		nodes_contrast %>% 
+			filter( gene %in% genes_pass_bm )
+
+	wilcox_test_result_bm = wilcox_oc( nodes_contrast_filtered_bm )
+
+```
+
+Of the `r nrow(tree_summary)` genes considered, and a $\Delta$AIC threshold of `r dAIC_thresh`, `r sum(tree_summary$dAIC_bm < dAIC_thresh & tree_summary$dAIC_ou >= dAIC_thresh)` have $\Delta$AIC less than the threshold only for Brownian Motion (BM), `r sum(tree_summary$dAIC_bm >= dAIC_thresh & tree_summary$dAIC_ou < dAIC_thresh)` have $\Delta$AIC less than the threshold only for Ornstein Uhlenbeck (OU), and `r sum(tree_summary$dAIC_bm < dAIC_thresh & tree_summary$dAIC_ou < dAIC_thresh)` have $\Delta$AIC less than the threshold for both models.
+That gives a total of `r sum( tree_summary$dAIC_bm < dAIC_thresh )` genes that have $\Delta$AIC less than the threshold for BM and `r sum( tree_summary$dAIC_ou < dAIC_thresh )` genes that have $\Delta$AIC less than the threshold for OU.
+
+Repeating the BM phylogenetic analyses on the subset of `r sum( tree_summary$dAIC_bm < dAIC_thresh )` genes that have $\Delta$AIC less than the threshold for BM still does not reject the null hypothesis (Wilcoxon p = $`r  wilcox_test_result_bm %>% round( 3 )`$).
+
+```{r model_ou, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
+
+	# Get the trees that pass AIC criteria for OU
+	gene_trees_pic_digests = lapply( gene_trees_pic, digest )
+	gene_trees_pic_ou = gene_trees_pic[ gene_trees_pic_digests %in% genes_pass_ou ]
+
+	# Replace the BM pics with OU pics
+	gene_trees_pic_ou %<>% add_pics_to_trees( model_method="OU" )
+	nodes_ou_contrast = gene_trees_pic_ou %>% summarize_contrasts()
+	wilcox_ou = wilcox_oc( nodes_ou_contrast )
+
+```
+
 #### Investigation of potential ascertainment bias

### 2017-11-02 functions.R

@@ -248,9 +248,10 @@ calibrate_tree = function ( nhx, calibration_times, ... ) {
 #' 
 #' @param nhx A phylogenetic tree as a treeio::treedata object, with a 
 #' column @data$Tau containing expression data
+#' @param model_method The model of trait evolution. Can be one of c("BM", "OU")
 #' @return A treeio::treedata object, with new @data columns pic and var_exp
 #' @export
-pic.nhx = function( nhx ) {
+pic.nhx = function( nhx, model_method="BM" ) {
 	
 	tau = nhx@data$Tau[ is.tip.nhx( nhx ) ]
 	
@@ -259,8 +260,29 @@ pic.nhx = function( nhx ) {
 	}
 	
 	# Calculate the contrasts
-	p = pic( tau, nhx@phylo, var.contrasts=TRUE )
+	if ( model_method=="BM" ){
+		p = ape::pic( 
+			tau, 
+			nhx@phylo, 
+			var.contrasts=TRUE 
+		)
+	} else if ( model_method=="OU" ){
+		p = hutan::picx( 
+			tau, 
+			nhx@phylo, 
+			var.contrasts=TRUE , 
+			model_method="OU", 
+			model_parameters=nhx@phylo$model_ou, 
+			n_replicates=25 
+		)
+	} else {
+		stop("ERROR: Invalid model_method")
+	}
 	
+	# Remove previous values if they exist
+	nhx@data$pic = NULL
+	nhx@data$var_exp = NULL
+
 	# Add the results back to the @data slot, padding the rows that correspond to 
 	# tips with NA
 	nhx@data$pic = c( rep( NA, length( nhx@phylo$tip.label ) ), p[ ,1 ] )
@@ -274,14 +296,20 @@ pic.nhx = function( nhx ) {
 #' 
 #' @param gene_trees_calibrated A list of time calibrated phylogenetic trees 
 #' as treeio::treedata objects 
+#' @param model_method The model of trait evolution. Can be one of c("BM", "OU")
 #' @return A list of time phylogenetic trees with pics 
 #' as treeio::treedata objects
 #' @export
-add_pics_to_trees = function( gene_trees_calibrated ) {
+add_pics_to_trees = function( gene_trees_calibrated, model_method="BM" ) {
 	
 	# Calculate independent contrasts for Tau on each tree, storing the results 
 	# back into the @data slot of the tree objects
[…]

### 2017-11-02 functions.R
added fitContinuous bounds

@@ -464,10 +464,27 @@ add_model_parameters = function( nhx, ... ) {
 	tau_original = nhx@data$Tau [ 1:length( phy$tip.label ) ]
 	names( tau_original ) = phy$tip.label
 
-	brownian_model = fitContinuous( phy, tau_original, model="BM", ... )
+	# Set bounds on fitContinuous per manual, as unbond searches
+	# can get stuck. These bounds were selected according to 
+	# observed estimates on runs when no fitContinuous calls
+	# got stuck. 
+
+	brownian_model = fitContinuous( 
+		phy, 
+		tau_original, 
+		model="BM", 
+		bounds=list(sigsq=c(0.0, 1.0)),
+		... 
+	)
 	nhx@phylo$model_bm$opt = brownian_model$opt
 
-	ou_model = fitContinuous( phy, tau_original, model="OU", ... )
+	ou_model = fitContinuous( 
+		phy, 
+		tau_original, 
+		model="OU", 
+		bounds=list(sigsq=c(0.0, 1.0), alpha=c(0.0, 3.0)),
+		... 
+	)
 	nhx@phylo$model_ou$opt = ou_model$opt	
 	
 	return( nhx )

### 2017-11-02 manuscript.rmd
added mclapply

@@
 	# Estimate model parameters and add them to the tree objects
-	# Subcalls to fitContinuous() are parallel, so don't use mclapply here
 	# This takes a while, so do it once here and reuse the values as needed
-	gene_trees_calibrated = lapply( gene_trees_calibrated, add_model_parameters )
+	gene_trees_calibrated = mclapply( 
+		gene_trees_calibrated, 
+		add_model_parameters, 
+		mc.cores=cores 
+	)
 ```

### 2017-11-09 manuscript.rmd
created manuscript_kernel.R

@@
 
+```
+
+```{r load_data, echo=FALSE, message=F, warning=F}
+
+	source( "functions.R" )
+
+	# Load data and analysis results generated by 
+	# manuscript_kernel.R
+	load("manuscript.RData")
+
 	# Set system computational parameters
@@
 	set.seed( 23456 )
-
-	# Name of the compara gene trees file. This was downloaded from 
-	# ftp://ftp.ensembl.org/pub/release-75/emf/ensembl-compara/homologies/
-	# and compressed with gz
-
-	gene_tree_name = "Compara.75.protein.nhx.emf.gz"
-
-	# Files produced by KMRR Rscript.R with Tau values and other expression statistics
-	expression_file_names = c(
-		"ChickenBrawandTScomparisonTable_9_6chPC.txt", 
-		"ChimpBrawandTScomparisonTable_9_6cmPC.txt", 
-		"GorillaBrawandTScomparisonTable_9_6gPC.txt", 
-		"HumBrawandTScomparisonTable_9_6hPC.txt", 
-		"MacacaBrawandTScomparisonTable_9_6mcPC.txt", 
-		"MusBrawandTScomparisonTable_9_6mPC.txt", 
-		"OpossumBrawandTScomparisonTable_9_6oPC.txt", 
-		"PlatypusBrawandTScomparisonTable_9_6pPC.txt"
-	)
-
-	kmrr_directory = "kmrr/"
-	expression_file_names %<>% paste( kmrr_directory, ., sep="" )
-	gene_tree_name %<>% paste( kmrr_directory, ., sep="" )
-
-	# The minimum number of genes with expression data that a gene needs to have to be considered
-	min_genes_with_expression = 4
-
-	# These clade dates are from the KMRR code, they got them from http://timetree.org/
-	calibration_times = data.frame(
-		age =
-			c( 20, 92, 167, 9, 42, 74, 296, 535, 104, 937, 29, 722, 441, 65, 15, 1215, 
-				414, 371, 162, 25, 74, 77, 86, 7 ), 
-		clade = 
-			c("Hominoidea", "Euarchontoglires", "Mammalia", "Homininae", 
-				"Simiiformes", "Primates", "Amniota", "Vertebrata", "Eutheria", 
-				"Bilateria", "Catarrhini", "Chordata", "Euteleostomi", 
-				"Haplorrhini", "Hominidae", "Opisthokonta", "Sarcopterygii", 
-				"Tetrapoda", "Theria", "Murinae", "Sciurognathi", "Rodentia", "Glires", "Hominini"
-		),
-		stringsAsFactors=FALSE
-	)
-
-	# The calibration times used for speciation nodes
-	focal_calibrations_clades = 
-		c( "Hominini", "Homininae", "Catarrhini", "Euarchontoglires", "Theria", "Mammalia", "Amniota" )
-
-	# In simulation, the fold change in rate following duplication relative to speciation
-	dup_adjust = 2
-
-```
-
-```{r define_functions, echo=FALSE, message=F, warning=F}
-
-	source( "functions.R" )
 
@@
 
-```{r load_trees, echo=FALSE, cache=TRUE, message=F, warning=F}
-
-	# Read in the tree file as a vector of strings, one per line
-	lines = readLines( gene_tree_name )
-
-	# Isolate strings that are trees
-	# These start with (
-	tree_lines = lines[ grepl( "^\\(", lines, perl = TRUE ) ]
-
-	# Parse each string to a treeio::treedata object, adding a label column 
-	# for node names
-	gene_trees = mclapply( tree_lines, parse_gene_trees, mc.cores=cores )
-
-	# Uncomment for development to look at a subset of gene trees
-	# gene_trees = gene_trees[1:100]
-
-	n_gene_trees = length( gene_trees )
-
-	# Parse the tree annotations, these are lines that start with SEQ
-	# These have the species names, which are not in the trees themselves
-	tip_annotations = 
-		lines[ grepl( "^SEQ", lines, perl = TRUE ) ] %>%	# Get lines that start with SEQ
-	  str_replace( " \\(\\d+ of \\d+\\)", "" ) %>%      # Remove instances of '(1 of 2)' and '(2 of 2)'
-	  str_replace( "(([^ ]+ ){8})([^ ]+) ([^ ]+)", "\\1\\3_\\4" ) %>%      # Get rid of spaces in last field of some entries
-		str_c( collapse="\n" ) %>%												# Combine them into one string
-		read_delim( delim=" ", col_names = FALSE ) %>% 		# Read string as tibble
-		select( 2, 8 ) %>%																# Select the columns with species and gene ID
-		rename( Ensembl.Gene.ID = X8 ) %>%
-		rename( species = X2 )
-
-	# Remove the large string objects to free up memory
-	rm( lines )
-	rm( tree_lines )
-
-```
-
-```{r load_expression, echo=FALSE, message=F, warning=F}
-
-	# Read the expression data for each species and combine into a single tibble,
-	# adding the species names from tip_annotations
-	expression = 
-		lapply(
-			expression_file_names, 
-			read.table, 
-			stringsAsFactors=FALSE, 
-			header=TRUE
-		) %>%
-		bind_rows() %>%
-		left_join( tip_annotations, by="Ensembl.Gene.ID" )
-
-	# Free up memory
-	rm( tip_annotations )
-
-```
-
-```{r add_expression_to_trees, echo=FALSE, cache=TRUE, message=F, warning=F}
-
-	# Annotate each tree by joining the corresponding expression data to the 
-	# @data object. This maps the expression data to the tree tips
-	gene_trees_annotated = mclapply(
-		gene_trees,
-		function( tree ){
-			tree@data %<>% left_join( expression, by = c( "G" = "Ensembl.Gene.ID" ) )
-
-			return( tree )
-		}, 
-		mc.cores=cores
-	)
-
-	# Free up memory
-	rm( gene_trees )
-
-	# Remove tips from the trees that do not have expression data
-	gene_trees_pruned = mclapply( 
-		gene_trees_annotated, 
-		drop_empty_tips, 
-		min_genes_with_expression=min_genes_with_expression,
-		mc.cores=cores 
-	)
-
-	# Free up memory
-	rm( gene_trees_annotated )
-
-	# Remove trees with less than min_genes_with_expression tips with expression data
-	gene_trees_pruned = gene_trees_pruned[ ! is.na( gene_trees_pruned ) ]
-
-	n_removed_for_no_speciation = length( gene_trees_pruned )
-
-	# Remove trees with no speciation events
-	gene_trees_pruned = gene_trees_pruned[ unlist( lapply( gene_trees_pruned, get_n_speciation ) ) > 0 ]
-
-	n_removed_for_no_speciation = n_removed_for_no_speciation - length( gene_trees_pruned )
-
-	# fix clade names
-	gene_trees_pruned = lapply( gene_trees_pruned, fix_hominini )
-
-```
-
-```{r calibrate_trees, echo=FALSE, cache=TRUE, message=F, warning=F}
-
-	# Time calibrate all speciation nodes to the same years
-	gene_trees_calibrated = calibrate_trees( gene_trees_pruned, calibration_times, model="correlated" )
-
-```
-
-```{r estimate_tau_model, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
-	# Estimate model parameters and add them to the tree objects
-	# This takes a while, so do it once here and reuse the values as needed
-	gene_trees_calibrated = mclapply( 
-		gene_trees_calibrated, 
-		add_model_parameters, 
-		mc.cores=cores 
-	)
-```
-
-```{r trees_to_contrasts, cache=TRUE, echo=FALSE, message=F, warning=F}
-
-	# Calculate the contrasts from the trees and associated expression data
-	gene_trees_pic = add_pics_to_trees( gene_trees_calibrated )
-
-	# Collect all the contrasts in a single tibble
-	nodes_contrast = summarize_contrasts( gene_trees_pic )
-
-	# Collect all the tree statistics in a single tibble
-	tree_summary = summarize_trees( gene_trees_pic )
-
-	# Test if the mean rank of duplication pics is greater than speciation pics
-	wilcox_test_result = wilcox_oc( nodes_contrast )
-
-```
-
 We mapped the Tau values calculated by KMRR [@KryuchkovaMostacci:2016iw] for the Brawand *et al.* 2011 [@Brawand:2011du] dataset onto `r n_gene_trees %>% format(big.mark=",")` gene trees parsed from ENSEMBL Compara [@Herrero:2016jj].
@@
 Our phylogenetic comparative analysis, unlike the previously published pairwise comparative analysis [@KryuchkovaMostacci:2016iw], therefore finds no support for the ortholog conjecture in this system.
-
-```{r ascertainment, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
-
-	# Examine by node age
-
-	max_speciation_age = 
-		nodes_contrast %>% 
-		filter( D=="N" ) %>%
-		select( node_age ) %>%
-		max( na.rm=TRUE )
-
-	nodes_contrast_filtered_age = 
-		nodes_contrast %>% 
-		filter( node_age <= max_speciation_age )
-
-	wilcox_filtered_age = wilcox_oc( nodes_contrast_filtered_age )
-
-	# Examine by expected variance, ie branch length
-	var_exp_max = 
-		nodes_contrast %>% 
-		filter( D=="N" ) %>%
-		select( var_exp ) %>%
-		max( na.rm=TRUE )
-
-	var_exp_min = 
-		nodes_contrast %>% 
-		filter( D=="N" ) %>%
-		select( var_exp ) %>%
-		min( na.rm=TRUE )
-
-	nodes_contrast_filtered_var = 
-		nodes_contrast %>% 
-		filter ( ( var_exp >= var_exp_min ) & ( var_exp <= var_exp_max ) )
-
-	wilcox_filtered_var = wilcox_oc( nodes_contrast_filtered_var )
-
-```
 
@@
 #### Understanding the incongruence between pairwise and phylogenetic methods
-
-```{r pairwise, echo=FALSE, cache=TRUE, message=F, warning=F}
-
-	# Build a tibble of all pairwise comparisons between tips
-	pairwise_summary = 
-		mclapply( gene_trees_pic, get_pairwise_summary, mc.cores=cores ) %>% 
-		bind_rows()
-
-	# Calculate some summary statistics 
-	ortholog_r = 
-		pairwise_summary %>% 
-		filter( D=="N" ) %$% 
-		cor.test( tau_a, tau_b )
-
-	paralog_r = 
-		pairwise_summary %>% 
-		filter( D=="Y" ) %$% 
-		cor.test( tau_a, tau_b )
-
-	paralog_subset_r = 
-		pairwise_summary %>% 
-		filter( D=="Y" ) %>% 
-		filter( node_age <= max_speciation_age ) %$% 
-		cor.test( tau_a, tau_b )
-
-	ortholog_distance_mean = 
-		pairwise_summary %>% 
-		filter( D=="N" ) %>% 
-		.[["distance"]] %>% 
-		mean()
-
-	paralog_distance_mean = 
-		pairwise_summary %>% 
-		filter( D=="Y" ) %>% 
-		.[["distance"]] %>% 
-		mean()
-
-```
 
@@
 This is much closer to the ortholog correlation of `r signif( ortholog_r$estimate, 2) `.
-
-```{r simulations_null, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
-
-	# Replace the actual Tau values with simulated Tau values, using the parameters 
-	# estimated for each tree
-	gene_trees_sim_null = mclapply( gene_trees_calibrated, sim_tau, mc.cores=cores )
-
-	# Perform contrasts 
-	gene_trees_sim_null %<>% add_pics_to_trees()
-	nodes_sim_null_contrast = gene_trees_sim_null %>% summarize_contrasts()
-
-	# Calculate pairwise comparisons
-	pairwise_sim_null_summary = 
-		mclapply( gene_trees_sim_null, get_pairwise_summary, mc.cores=cores ) %>% 
-		bind_rows()
-
-	# Calculate some summary statistics 
-	ortholog_r_sim_null = 
-		pairwise_sim_null_summary %>% 
-		filter( D=="N" ) %$% 
-		cor.test( tau_a, tau_b )
-
-	paralog_r_sim_null = 
-		pairwise_sim_null_summary %>% 
-		filter( D=="Y" ) %$% 
-		cor.test( tau_a, tau_b )
-
-```
-
-```{r simulations_oc, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
-
-	# Replace the actual Tau values with simulated Tau values
-	gene_trees_sim_oc = lapply( gene_trees_calibrated, sim_tau, dup_adjust=dup_adjust )
-
-	# Calculate contrasts 
-	gene_trees_sim_oc %<>% add_pics_to_trees()
-	nodes_sim_oc_contrast = gene_trees_sim_oc %>% summarize_contrasts()
-
-	# Calculate pairwise comparisons 
-	pairwise_sim_oc_summary = 
-		mclapply( gene_trees_sim_oc, get_pairwise_summary, mc.cores=cores ) %>% 
-		bind_rows()
-
-	ortholog_r_sim_oc = 
-		pairwise_sim_oc_summary %>% 
-		filter( D=="N" ) %$% 
-		cor.test( tau_a, tau_b )
-
-	paralog_r_sim_oc = 
-		pairwise_sim_oc_summary %>% 
-		filter( D=="Y" ) %$% 
-		cor.test( tau_a, tau_b )
-
-```
 
@@
 
+```{r file_status, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
+	commit_render = system("git log | head -n 1", intern=TRUE) %>% str_replace("commit ", "")
+
+```
+
 All files needed to re-execute the analyses presented in this document are available at https://github.com/caseywdunn/comparative_expression_2017.
-The most recent commit at the time of the analysis presented here was `r system("git log | head -n 1", intern=TRUE) %>% str_replace("commit ", "")`.
+The most recent commit at the time the analyses were executed was `r commit_kernel %>% strtrim(8)`.
+The most recent commit at the time of the manuscript was rendered was `r commit_render %>% strtrim(8)`.
 
@@
 
-```{r phylogenetic_signal, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
-
-	k_thresh = quantile( tree_summary$K, na.rm=TRUE )[3] # third quantile is the median
-	k_percentile = names(k_thresh)
-	genes_pass_k = tree_summary$gene[ tree_summary$K > k_thresh ]
-
-	nodes_contrast_filtered_k = 
-		nodes_contrast %>% 
-			filter( gene %in% genes_pass_k )
-
-	wilcox_test_result_k = wilcox_oc( nodes_contrast_filtered_k )
-
-```
-
 We removed the `r k_percentile` of genes with the lowest phylogenetic signal, corresponding to a K cutoff for retained genes of greater than `r signif( k_thresh, 3)`.
@@
 
-```{r model_selection, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
-
-	dAIC_thresh = 2
-	AIC_min = pmin( tree_summary$aic_bm, tree_summary$aic_ou )
-	tree_summary$dAIC = tree_summary$aic_ou - tree_summary$aic_bm
-	tree_summary$dAIC_bm = tree_summary$aic_bm - AIC_min
-	tree_summary$dAIC_ou = tree_summary$aic_ou - AIC_min
-	genes_pass_bm = tree_summary$gene[ tree_summary$dAIC_bm < dAIC_thresh ]
-	genes_pass_ou = tree_summary$gene[ tree_summary$dAIC_ou < dAIC_thresh ]
-
-	nodes_contrast_filtered_bm = 
-		nodes_contrast %>% 
-			filter( gene %in% genes_pass_bm )
-
-	wilcox_test_result_bm = wilcox_oc( nodes_contrast_filtered_bm )
-
-```
-
 Of the `r nrow(tree_summary)` genes considered, and a $\Delta$AIC threshold of `r dAIC_thresh`, `r sum(tree_summary$dAIC_bm < dAIC_thresh & tree_summary$dAIC_ou >= dAIC_thresh)` have $\Delta$AIC less than the threshold only for Brownian Motion (BM), `r sum(tree_summary$dAIC_bm >= dAIC_thresh & tree_summary$dAIC_ou < dAIC_thresh)` have $\Delta$AIC less than the threshold only for Ornstein Uhlenbeck (OU), and `r sum(tree_summary$dAIC_bm < dAIC_thresh & tree_summary$dAIC_ou < dAIC_thresh)` have $\Delta$AIC less than the threshold for both models.
@@
 Repeating the BM phylogenetic analyses on the subset of `r sum( tree_summary$dAIC_bm < dAIC_thresh )` genes that have $\Delta$AIC less than the threshold for BM still does not reject the null hypothesis (Wilcoxon p = $`r  wilcox_test_result_bm %>% round( 3 )`$).
-
-```{r model_ou, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
-
-	# Get the trees that pass AIC criteria for OU
-	gene_trees_pic_digests = lapply( gene_trees_pic, digest )
-	gene_trees_pic_ou = gene_trees_pic[ gene_trees_pic_digests %in% genes_pass_ou ]
-
-	# Replace the BM pics with OU pics
-	gene_trees_pic_ou %<>% add_pics_to_trees( model_method="OU" )
-	nodes_ou_contrast = gene_trees_pic_ou %>% summarize_contrasts()
-	wilcox_ou = wilcox_oc( nodes_ou_contrast )
-
-```
 
@@
 
-```{r calibration_sensitivity, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
-
-	noised_replicate_n = 10
-	sd_fraction = 0.2
-
-	calibration_times_focal = 
-		calibration_times %>%
-		filter( clade %in% focal_calibrations_clades ) %>%
-		arrange( age )
-
-	noised_dates = 
-		replicate( 
-			noised_replicate_n, 
-			add_noise_calibration_times( calibration_times_focal, sd_fraction ),
-			simplify = FALSE
-		)
-
-	nodes_contrast_noised_list = lapply(
-		noised_dates,
-		function( x ){
-			gene_trees_pruned %>% 
-				calibrate_trees( x, model="correlated" ) %>% 
-				add_pics_to_trees() %>% 
-				summarize_contrasts()
-		}
-	)
-
-	p_noised = lapply( nodes_contrast_noised_list, wilcox_oc )
-
-```
-
 We examined the sensitivity of our results to the specification of calibration dates for the speciation nodes.
@@
 
-This manuscript was computed on `r format( Sys.time(), "%a %b %d %X %Y" )` with the following R package versions.
+The analyses presented in this manuscript were computed on `r format( system_time_kernel, "%a %b %d %X %Y" )` with the following R package versions.
 
 ```{r session_summary, echo=FALSE, comment=NA}
-	sessionInfo()
-```
-
-```{r wrap_up, echo=FALSE}
-	save.image()
-```
+	session_info_kernel
+```
+
+This manuscript was rendered from these analyses on `r format( system_time_kernel, "%a %b %d %X %Y" )`.
 

### 2017-11-09 manuscript_kernel.R

@@ -1,0 +1,419 @@
+
+## ----preliminaries, echo=FALSE, message=F, warning=F---------------------
+
+	# The following should be installed from github with the specified 
+	# devtools command. You will need to install devtools first.
+	library( treeio ) # devtools::install_github("GuangchuangYu/treeio")
+	library( ggtree ) # devtools::install_github("GuangchuangYu/ggtree")
+	library( hutan ) # devtools::install_github("caseywdunn/hutan")
+
+	# The rest can be installed from CRAN
+	library( ape )
+	library( digest )
+	library( geiger )
+	library( ggrepel )
+	library( gridExtra )
+	library( magrittr )
+	library( parallel )
+	library( phytools )
+	library( stringr )
+	library( tidyverse )
+	
+
+	# Set system computational parameters
+	cores = detectCores()
+	if ( cores < 1 ) {
+		cores = 1
+	}
+	set.seed( 23456 )
+	
+	# Name of the compara gene trees file. This was downloaded from 
+	# ftp://ftp.ensembl.org/pub/release-75/emf/ensembl-compara/homologies/
+	# and compressed with gz
+
+	gene_tree_name = "Compara.75.protein.nhx.emf.gz"
+	
+	# Files produced by KMRR Rscript.R with Tau values and other expression statistics
+	expression_file_names = c(
+		"ChickenBrawandTScomparisonTable_9_6chPC.txt", 
+		"ChimpBrawandTScomparisonTable_9_6cmPC.txt", 
+		"GorillaBrawandTScomparisonTable_9_6gPC.txt", 
+		"HumBrawandTScomparisonTable_9_6hPC.txt", 
+		"MacacaBrawandTScomparisonTable_9_6mcPC.txt", 
+		"MusBrawandTScomparisonTable_9_6mPC.txt", 
+		"OpossumBrawandTScomparisonTable_9_6oPC.txt", 
+		"PlatypusBrawandTScomparisonTable_9_6pPC.txt"
+	)
+	
+	kmrr_directory = "kmrr/"
+	expression_file_names %<>% paste( kmrr_directory, ., sep="" )
+	gene_tree_name %<>% paste( kmrr_directory, ., sep="" )
+	
+	# The minimum number of genes with expression data that a gene needs to have to be considered
+	min_genes_with_expression = 4
+	
+
[…]

### 2017-11-09 manuscript_kernel.R
added checkpoint

@@ -182,6 +182,7 @@
 	# Time calibrate all speciation nodes to the same years
 	gene_trees_calibrated = calibrate_trees( gene_trees_pruned, calibration_times, model="correlated" )
 	
+	save.image("manuscript_checkpoint_calibrate_trees.RData")
 
 ## ----estimate_tau_model, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE----
 	# Estimate model parameters and add them to the tree objects

### 2017-11-09 manuscript.rmd
set ncores=1 for fitContinuous()

@@
 
+	source( "functions.R" )
+	set.seed( 23456 )
+
 ```
@@
 
-	source( "functions.R" )
-
 	# Load data and analysis results generated by 
@@
 	load("manuscript.RData")
-
-	# Set system computational parameters
-	cores = detectCores()
-	if ( cores < 1 ) {
-		cores = 1
-	}
-	set.seed( 23456 )
 
@@
 
-	x = calibration_times$age[calibration_times$clade == "Hominini"]
-	gene_trees_extended = mclapply( gene_trees_calibrated, extend_nhx, x=x, mc.cores=cores )
-	gene_trees_extended %<>% add_pics_to_trees()
-	nodes_extended_contrast = gene_trees_extended %>% summarize_contrasts()
-
-	pairwise_extended_summary = 
-			mclapply( gene_trees_extended, get_pairwise_summary, mc.cores=cores ) %>% 
-			bind_rows()
-
 	g_pair_extended = 

### 2017-11-09 functions.R

@@ -467,13 +467,15 @@ add_model_parameters = function( nhx, ... ) {
 	# Set bounds on fitContinuous per manual, as unbond searches
 	# can get stuck. These bounds were selected according to 
 	# observed estimates on runs when no fitContinuous calls
-	# got stuck. 
+	# got stuck. Set ncores=1 so that can wrap in mclapply 
+	# without problems.
 
 	brownian_model = fitContinuous( 
 		phy, 
 		tau_original, 
 		model="BM", 
 		bounds=list(sigsq=c(0.0, 1.0)),
+		ncores=1
 		... 
 	)
 	nhx@phylo$model_bm$opt = brownian_model$opt
@@ -483,6 +485,7 @@ add_model_parameters = function( nhx, ... ) {
 		tau_original, 
 		model="OU", 
 		bounds=list(sigsq=c(0.0, 1.0), alpha=c(0.0, 3.0)),
+		ncores=1
 		... 
 	)
 	nhx@phylo$model_ou$opt = ou_model$opt	

### 2017-11-09 manuscript_kernel.R

@@ -18,14 +18,16 @@
 	library( phytools )
 	library( stringr )
 	library( tidyverse )
-	
+
+	source( "functions.R" )	
+	set.seed( 23456 )
+
 
 	# Set system computational parameters
 	cores = detectCores()
 	if ( cores < 1 ) {
 		cores = 1
 	}
-	set.seed( 23456 )
 	
 	# Name of the compara gene trees file. This was downloaded from 
 	# ftp://ftp.ensembl.org/pub/release-75/emf/ensembl-compara/homologies/
@@ -75,11 +77,6 @@
 	
 	# In simulation, the fold change in rate following duplication relative to speciation
 	dup_adjust = 2
-	
-
-## ----define_functions, echo=FALSE, message=F, warning=F------------------
-	
-	source( "functions.R" )
 
 
 ## ----load_trees, echo=FALSE, cache=TRUE, message=F, warning=F------------
@@ -407,7 +404,16 @@
 	
 	p_noised = lapply( nodes_contrast_noised_list, wilcox_oc )
 
-	
+
+## ----Fig_S3             -------------------------------------------------
+	x = calibration_times$age[calibration_times$clade == "Hominini"]
+	gene_trees_extended = mclapply( gene_trees_calibrated, extend_nhx, x=x, mc.cores=cores )
+	gene_trees_extended %<>% add_pics_to_trees()
+	nodes_extended_contrast = gene_trees_extended %>% summarize_contrasts()
+
+	pairwise_extended_summary = 
+			mclapply( gene_trees_extended, get_pairwise_summary, mc.cores=cores ) %>% 
+			bind_rows()
 
 ## ----session_summary, echo=FALSE, comment=NA-----------------------------
 	session_info_kernel = sessionInfo()

### 2017-11-09 functions.R
typo

@@ -485,7 +485,7 @@ add_model_parameters = function( nhx, ... ) {
 		tau_original, 
 		model="OU", 
 		bounds=list(sigsq=c(0.0, 1.0), alpha=c(0.0, 3.0)),
-		ncores=1
+		ncores=1,
 		... 
 	)
 	nhx@phylo$model_ou$opt = ou_model$opt	

### 2017-11-09 functions.R
typo

@@ -475,7 +475,7 @@ add_model_parameters = function( nhx, ... ) {
 		tau_original, 
 		model="BM", 
 		bounds=list(sigsq=c(0.0, 1.0)),
-		ncores=1
+		ncores=1,
 		... 
 	)
 	nhx@phylo$model_bm$opt = brownian_model$opt

### 2017-11-09 manuscript_kernel.R
replaced a mclapply with doParallel foreach

@@ -15,6 +15,8 @@
 	library( gridExtra )
 	library( magrittr )
 	library( parallel )
+	library( foreach )
+	library( doParallel )
 	library( phytools )
 	library( stringr )
 	library( tidyverse )
@@ -28,6 +30,9 @@
 	if ( cores < 1 ) {
 		cores = 1
 	}
+
+	# Register parallel workers for %dopar%
+	registerDoParallel( cores-1 )
 	
 	# Name of the compara gene trees file. This was downloaded from 
 	# ftp://ftp.ensembl.org/pub/release-75/emf/ensembl-compara/homologies/
@@ -184,11 +189,9 @@
 ## ----estimate_tau_model, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE----
 	# Estimate model parameters and add them to the tree objects
 	# This takes a while, so do it once here and reuse the values as needed
-	gene_trees_calibrated = mclapply( 
-		gene_trees_calibrated, 
-		add_model_parameters, 
-		mc.cores=cores 
-	)
+	
+	gene_trees_calibrated = 
+		foreach( nhx=gene_trees_calibrated ) %dopar% add_model_parameters( nhx )
 
 ## ----trees_to_contrasts, cache=TRUE, echo=FALSE, message=F, warning=F----
 

### 2017-11-10 manuscript.rmd
replaced all mclapply with doParallel foreach

@@
 	library( magrittr )
-	library( parallel )
 	library( phytools )

### 2017-11-10 functions.R

@@ -304,12 +304,8 @@ add_pics_to_trees = function( gene_trees_calibrated, model_method="BM" ) {
 	
 	# Calculate independent contrasts for Tau on each tree, storing the results 
 	# back into the @data slot of the tree objects
-	gene_trees_pic = mclapply( 
-		gene_trees_calibrated, 
-		pic.nhx, 
-		model_method=model_method, 
-		mc.cores=cores 
-	)
+	gene_trees_pic = foreach( tree=gene_trees_calibrated ) %dopar% 
+		pic.nhx( tree, model_method=model_method )
 	
 	return( gene_trees_pic )
 }
@@ -403,27 +399,17 @@ summarize_trees = function( gene_trees_pic ) {
 #' @export
 calibrate_trees = function( gene_trees_pruned, calibration_times, ... ) {
 	# Make the trees ultrametric and calibrate the speciation nodes to 
-	# specified times
-	gene_trees_calibrated = 
-		mclapply( 
-			gene_trees_pruned,
-			calibrate_tree, 
-			calibration_times=calibration_times,
-			...,
-			mc.cores=cores
-		)
-	
+	# specified times	
+	gene_trees_calibrated = foreach( tree=gene_trees_pruned ) %dopar% 
+		calibrate_tree( tree, calibration_times=calibration_times, ... )
+
 	# Remove trees that could not be successfully calibrated
 	gene_trees_calibrated = gene_trees_calibrated[ ! is.na( gene_trees_calibrated ) ]
 	
 	# Parse the calibrated node ages from the internal @phylo object and store them with the 
 	# corresponding rows in the @data object
-	gene_trees_calibrated = 
-		mclapply( 
-			gene_trees_calibrated, 
-			store_node_age, 
-			mc.cores=cores 
-		)
+	gene_trees_calibrated = foreach( tree=gene_trees_calibrated ) %dopar% 
+		store_node_age( tree )
 	
 	return( gene_trees_calibrated )
 }
@@ -467,8 +453,8 @@ add_model_parameters = function( nhx, ... ) {
 	# Set bounds on fitContinuous per manual, as unbond searches
 	# can get stuck. These bounds were selected according to 
 	# observed estimates on runs when no fitContinuous calls
-	# got stuck. Set ncores=1 so that can wrap in mclapply 
-	# without problems.
+	# got stuck. Set ncores=1 so that can wrap in parallel 
+	# code without problems.
 
 	brownian_model = fitContinuous( 
 		phy, 

### 2017-11-10 manuscript_kernel.R

@@ -14,7 +14,6 @@
 	library( ggrepel )
 	library( gridExtra )
 	library( magrittr )
-	library( parallel )
 	library( foreach )
 	library( doParallel )
 	library( phytools )
@@ -26,13 +25,13 @@
 
 
 	# Set system computational parameters
-	cores = detectCores()
+	cores = detectCores() - 1
 	if ( cores < 1 ) {
 		cores = 1
 	}
 
 	# Register parallel workers for %dopar%
-	registerDoParallel( cores-1 )
+	registerDoParallel( cores )
 	
 	# Name of the compara gene trees file. This was downloaded from 
 	# ftp://ftp.ensembl.org/pub/release-75/emf/ensembl-compara/homologies/
@@ -70,14 +69,16 @@
 				"Simiiformes", "Primates", "Amniota", "Vertebrata", "Eutheria", 
 				"Bilateria", "Catarrhini", "Chordata", "Euteleostomi", 
 				"Haplorrhini", "Hominidae", "Opisthokonta", "Sarcopterygii", 
-				"Tetrapoda", "Theria", "Murinae", "Sciurognathi", "Rodentia", "Glires", "Hominini"
+				"Tetrapoda", "Theria", "Murinae", "Sciurognathi", "Rodentia", 
+				"Glires", "Hominini"
 		),
 		stringsAsFactors=FALSE
 	)
 	
 	# The calibration times used for speciation nodes
 	focal_calibrations_clades = 
-		c( "Hominini", "Homininae", "Catarrhini", "Euarchontoglires", "Theria", "Mammalia", "Amniota" )
+		c( "Hominini", "Homininae", "Catarrhini", "Euarchontoglires", "Theria", 
+			"Mammalia", "Amniota" )
 	
 	
 	# In simulation, the fold change in rate following duplication relative to speciation
@@ -95,7 +96,8 @@
 
 	# Parse each string to a treeio::treedata object, adding a label column 
 	# for node names
-	gene_trees = mclapply( tree_lines, parse_gene_trees, mc.cores=cores )
+	gene_trees = foreach( tree_line=tree_lines ) %dopar%
+		parse_gene_trees( tree_line )
 	
 	# Uncomment for development to look at a subset of gene trees
 	# gene_trees = gene_trees[1:100]
@@ -141,26 +143,23 @@
 
 	# Annotate each tree by joining the corresponding expression data to the 
[…]

### 2017-11-10 manuscript_kernel.R
added more checkpointing

@@ -1,6 +1,7 @@
 
 ## ----preliminaries, echo=FALSE, message=F, warning=F---------------------
-
+	
+	time_start = Sys.time()
 	# The following should be installed from github with the specified 
 	# devtools command. You will need to install devtools first.
 	library( treeio ) # devtools::install_github("GuangchuangYu/treeio")
@@ -205,7 +206,8 @@
 	
 	# Test if the mean rank of duplication pics is greater than speciation pics
 	wilcox_test_result = wilcox_oc( nodes_contrast )
-	
+
+	save.image("manuscript_checkpoint_contrasts.RData")
 
 ## ----ascertainment, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE----
 
@@ -281,6 +283,7 @@
 		.[["distance"]] %>% 
 		mean()
 	
+	save.image("manuscript_checkpoint_pairwise.RData")
 
 	
 
@@ -338,6 +341,7 @@
 		filter( D=="Y" ) %$% 
 		cor.test( tau_a, tau_b ) 
 
+	save.image("manuscript_checkpoint_calibrate_simulations.RData")
 
 ## ----phylogenetic_signal, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE----
 
@@ -379,6 +383,8 @@
 	gene_trees_pic_ou %<>% add_pics_to_trees( model_method="OU" )
 	nodes_ou_contrast = gene_trees_pic_ou %>% summarize_contrasts()
 	wilcox_ou = wilcox_oc( nodes_ou_contrast )
+
+	save.image("manuscript_checkpoint_models.RData")
 
 
 ## ----calibration_sensitivity, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE----
@@ -432,4 +438,7 @@
 
 ## ----wrap_up, echo=FALSE-------------------------------------------------
 	save.image("manuscript.RData")
+
+	time_stop = Sys.time()
+	time_run = time_stop - time_start
 

### 2017-11-10 functions.R
increased picx replicates now that using simulated internal states

@@ -272,8 +272,7 @@ pic.nhx = function( nhx, model_method="BM" ) {
 			nhx@phylo, 
 			var.contrasts=TRUE , 
 			model_method="OU", 
-			model_parameters=nhx@phylo$model_ou, 
-			n_replicates=25 
+			model_parameters=nhx@phylo$model_ou
 		)
 	} else {
 		stop("ERROR: Invalid model_method")

### 2017-11-10 manuscript_kernel.R

@@ -301,12 +301,14 @@
 	# Calculate pairwise comparisons
 	pairwise_sim_null_summary = foreach( tree=gene_trees_sim_null ) %dopar% 
 		get_pairwise_summary( tree )
+
+	pairwise_sim_null_summary %<>% bind_rows()
 	
 	# Calculate some summary statistics 
 	ortholog_r_sim_null = 
 		pairwise_sim_null_summary %>% 
 		filter( D=="N" ) %$% 
-		cor.test( tau_a, tau_b ) 
+		cor.test( tau_a, tau_b )
 	
 	paralog_r_sim_null = 
 		pairwise_sim_null_summary %>% 
@@ -341,7 +343,7 @@
 		filter( D=="Y" ) %$% 
 		cor.test( tau_a, tau_b ) 
 
-	save.image("manuscript_checkpoint_calibrate_simulations.RData")
+	save.image("manuscript_checkpoint_simulations.RData")
 
 ## ----phylogenetic_signal, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE----
 
@@ -372,6 +374,7 @@
 	
 	wilcox_test_result_bm = wilcox_oc( nodes_contrast_filtered_bm )
 
+	save.image("manuscript_checkpoint_model_selection.RData")
 
 ## ----model_ou, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE------
 
@@ -384,7 +387,7 @@
 	nodes_ou_contrast = gene_trees_pic_ou %>% summarize_contrasts()
 	wilcox_ou = wilcox_oc( nodes_ou_contrast )
 
-	save.image("manuscript_checkpoint_models.RData")
+	save.image("manuscript_checkpoint_model_ou.RData")
 
 
 ## ----calibration_sensitivity, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE----

### 2017-11-10 manuscript_kernel.R
save stop time

@@ -440,8 +440,8 @@
 	commit_kernel = system("git log | head -n 1", intern=TRUE) %>% str_replace("commit ", "")
 
 ## ----wrap_up, echo=FALSE-------------------------------------------------
-	save.image("manuscript.RData")
 
 	time_stop = Sys.time()
 	time_run = time_stop - time_start
-
+	
+	save.image("manuscript.RData")

### 2017-11-11 install_dependencies.R
updated readmes with new execution instructions

@@ -1,31 +1,0 @@
-# Installs manuscript dependencies
-# 
-# This installs all dependencies needed to run the 
-# manuscript, as verified by running it in a 
-# container based on the docker image 
-# https://hub.docker.com/r/rocker/tidyverse/
-#
-# From CRAN
-install.packages( 
-	c(
-		"devtools",
-		"tidyverse",
-		"magrittr",
-		"stringr",
-		"parallel",
-		"ape",
-		"digest",
-		"gridExtra",
-		"geiger",
-		"phytools"
-	) 
-)
-
-# From Bioconductor, needed for ggtree and treeio
-source( "https://bioconductor.org/biocLite.R" )
-biocLite()
-
-# From github
-devtools::install_github( "GuangchuangYu/treeio", quiet=TRUE )
-devtools::install_github( "GuangchuangYu/ggtree", quiet=TRUE )
-devtools::install_github( "caseywdunn/hutan", quiet=TRUE )

### 2017-11-11 manuscript.rmd
added picx OU result

@@
 
-```{r setup, cache=TRUE, message=F, echo=FALSE, warning=F}
+```{r setup, cache=TRUE, message=FALSE, echo=FALSE, warning=FALSE}
 
 	knitr::opts_chunk$set( 
-	echo = TRUE,
+	  echo = FALSE,
+	  message=FALSE, 
+	  warning=FALSE,
     dev="png", 
@@
 
-```{r preliminaries, echo=FALSE, message=F, warning=F}
+```{r preliminaries}
 
@@
 
-```{r load_data, echo=FALSE, message=F, warning=F}
+```{r load_data}
 
@@
 
-```{r Fig_2, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.width=7, fig.height=6.5, fig.cap="Pairwise (left column - *A, C, E*) and phylogenetic (right column - *B, D, F*) analyses of the original data (first row - *A, B*), data simulated under the null hypothesis (second row - *C, D*), and data simulated under the ortholog conjecture (third row- *E, F*).
+```{r Fig_2, fig.width=7, fig.height=6.5, fig.cap="Pairwise (left column - *A, C, E*) and phylogenetic (right column - *B, D, F*) analyses of the original data (first row - *A, B*), data simulated under the null hypothesis (second row - *C, D*), and data simulated under the ortholog conjecture (third row- *E, F*).
 In the pairwise plots, each point indicates the correlation coefficient of Tau for a set of pairwise comparisons annotated with a specific node name (e.g, Primates) and event type (speciation or duplication, giving rise respectively to orthologs and paralogs).
@@
 
-```{r cor_structure, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
+```{r cor_structure}
 
@@
 
-```{r file_status, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
+```{r file_status}
+
 	commit_render = system("git log | head -n 1", intern=TRUE) %>% str_replace("commit ", "")
@@
 
-```{r contrast_table, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
+```{r contrast_table}
 
@@
 
-```{r species_table, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
+```{r species_table}
 
@@
 
-```{r Fig_S1, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, eval=TRUE, fig.width=7, fig.height=6.5, fig.cap="One of the gene trees from the phylogenetic analysis.
+```{r Fig_S1, fig.width=7, fig.height=6.5, fig.cap="One of the gene trees from the phylogenetic analysis.
 The value of Tau (expression specificity) is indicated by the sizes of the circles at the tips of the tree.
@@
 Of the `r nrow(tree_summary)` genes considered, and a $\Delta$AIC threshold of `r dAIC_thresh`, `r sum(tree_summary$dAIC_bm < dAIC_thresh & tree_summary$dAIC_ou >= dAIC_thresh)` have $\Delta$AIC less than the threshold only for Brownian Motion (BM), `r sum(tree_summary$dAIC_bm >= dAIC_thresh & tree_summary$dAIC_ou < dAIC_thresh)` have $\Delta$AIC less than the threshold only for Ornstein Uhlenbeck (OU), and `r sum(tree_summary$dAIC_bm < dAIC_thresh & tree_summary$dAIC_ou < dAIC_thresh)` have $\Delta$AIC less than the threshold for both models.
-That gives a total of `r sum( tree_summary$dAIC_bm < dAIC_thresh )` genes that have $\Delta$AIC less than the threshold for BM and `r sum( tree_summary$dAIC_ou < dAIC_thresh )` genes that have $\Delta$AIC less than the threshold for OU.
+That gives a total of `r sum( tree_summary$dAIC_bm < dAIC_thresh )` genes that have $\Delta$AIC less than the threshold for BM and `r sum( tree_summary$dAIC_ou < dAIC_thresh )` genes that have $\Delta$AIC less than the threshold for OU (these sets are overlapping, so their total is more than the number of genes considered).
 
 Repeating the BM phylogenetic analyses on the subset of `r sum( tree_summary$dAIC_bm < dAIC_thresh )` genes that have $\Delta$AIC less than the threshold for BM still does not reject the null hypothesis (Wilcoxon p = $`r  wilcox_test_result_bm %>% round( 3 )`$).
+
+We next approximated the Phylogenetic Independent Contrasts for the the genes that have $\Delta$AIC less than the threshold for OU under an OU model.
+These analyses do not reject the null hypothesis (Wilcoxon p = $`r  wilcox_ou %>% round( 3 )`$).
+The challenge with such an analysis is that BM is intrinsic to the classic implementation of Phylogenetic Independent Contrasts, which use the fact that the expected variance is proportional to branch length to scale each contrast.
+Under OU, branch length no longer has a linear relationship to expected variance.
+We therefore took an alternative simulation-based approach to scaling the OU contrasts.
+We implemented this new method, which we call picx (for "pic extended models"), in our R package hutan (https://github.com/caseywdunn/hutan).
+For each tree, picx first estimate the model parameters from the observed data.
+It then simulates two hundred datasets under these parameters, recording both the internal and tip node states.
+It then scales the observed contrasts by the mean magnitude of the corresponding contrasts in the simulations.
+We validated the method by comparing picx BM results to the standard pic results calcualted by ape.
+These validations indicated a strong linear realationship between the two ($R^2$ > 0.95).
+This approach could be extended to approximate Phylogenetic Independent Contrasts under other continuous models of evolution as well.
 
@@
 
-```{r Fig_S2, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.width=7, fig.height=6.5, fig.cap="Investigation of possible ascertainment biases.
+```{r Fig_S2, fig.width=7, fig.height=6.5, fig.cap="Investigation of possible ascertainment biases.
 (*A*) Magnitude of independent contrasts plotted against node age.
@@
 
-```{r Fig_S3, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.width=7, fig.height=3, fig.cap="Investigation of the potential impact of within-species variance.
+```{r Fig_S3, fig.width=7, fig.height=3, fig.cap="Investigation of the potential impact of within-species variance.
 These plots correspond to Fig.
@@
 
-```{r Fig_S4, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.width=7, fig.height=3, fig.cap="Origin of pairwise correlation plot structure.
+```{r Fig_S4, fig.width=7, fig.height=3, fig.cap="Origin of pairwise correlation plot structure.
 (*A*) Each speciation and duplication node in the COMPARA trees is annotated with a clade name, and the age of that clade (shown on the x axis) is used as the age of the pairwise comparisons (Fig.
@@
 
-```{r Tau_v_max, cache=TRUE, echo=FALSE, warning=FALSE, fig.width=3.42, fig.height=3.42, message=FALSE, fig.cap="Genes with higher maximum observed expression tend to have lower Tau.
+```{r Tau_v_max, fig.width=3.42, fig.height=3.42, fig.cap="Genes with higher maximum observed expression tend to have lower Tau.
 The blue line is the linear model for the relationship between the two."}
@@
 
-The analyses presented in this manuscript were computed on `r format( system_time_kernel, "%a %b %d %X %Y" )` with the following R package versions.
+These analyses were computed on `r format( system_time_kernel, "%a %b %d %X %Y" )` with the following R package versions.
 
-```{r session_summary, echo=FALSE, comment=NA}
+```{r session_summary, comment=NA}
 	session_info_kernel

### 2017-11-12 manuscript.rmd
refined model discussion, did not update pdf

@@
 Our phylogenetic comparative analysis, unlike the previously published pairwise comparative analysis [@KryuchkovaMostacci:2016iw], therefore finds no support for the ortholog conjecture in this system.
+This result holds when we exclude the genes with the lowest phylogenetic signal according to Blomberg's $K$ [@Blomberg:2003jg] and is robust to model selection (*SI Appendix*).
 
@@
 The distributions of similarity scores are plotted in their Fig.
-4d [@Levin:2016bp], and their Kolmogorov–Smirnov (KS) tests indicated that the early distribution and late distribution were each significantly different from mid distribution (P < $10^{-6}$ and P < $10^{-12}$, respectively).
+4d [@Levin:2016bp], and their Kolmogorov-Smirnov (KS) tests indicated that the early distribution and late distribution were each significantly different from mid distribution (P < $10^{-6}$ and P < $10^{-12}$, respectively).
 This is the support they presented for the inverse hourglass model.
@@
 CWD's contribution was supported by the National Science Foundation (DEB-1256695 and the Waterman Award).
-AH was supported by the European Research Council Community’s Framework Program Horizon 2020 (2014–2020) ERC grant agreement 648861.
+AH was supported by the European Research Council Community's Framework Program Horizon 2020 (2014-2020) ERC grant agreement 648861.
 \pagebreak
@@
 
-#### Model selection and sensitivity
+#### Phylogenetic signal and models
 
-We removed the `r k_percentile` of genes with the lowest phylogenetic signal, corresponding to a K cutoff for retained genes of greater than `r signif( k_thresh, 3)`.
+```{r Fig_SXX, fig.width=5, fig.height=3, fig.cap="Density plot of Blomberg's K.
+The shaded portion of the plot indicates the gene trees that were analyzed after excluding the trees that fell below the K threshold."}
+
+	max_k = 2
+	dens = density(tree_summary$K[ !is.na(tree_summary$K) ], from = 0, to = max_k)
+	D = data.frame(x = dens$x, y = dens$y)
+	thresh = k_thresh
+	shade = rbind(c(max_k, 0), c(thresh,0), subset(D, x > thresh))
+
+	thresh_label = paste( "\nthreshold: > ", round(thresh,2), sep='' )
+
+	g_k_density = 
+		ggplot(D, aes(x,y)) + 
+			geom_line() + 
+			geom_polygon( data = shade, aes(x, y) ) +
+			xlab("K") +
+			ylab("density") +
+			theme_classic() +
+			geom_vline(xintercept = thresh, col="grey") +
+			geom_text(aes(x=thresh, label=thresh_label, y=mean(range(y))), colour="grey", angle=90) +
+			ggtitle( "A Density of K" )
+
+	g_k_density
+
+```
+
+The gene trees considered here vary widely in their phylogenetic signal for Tau according to Blomberg's $K$ [@Blomberg:2003jg] (*SI Appendix*, Fig.
+XX).
+$K$ describes the tendancy for closely related tips to resemble each other, where $K=0$ indicates that close relatives resemble each other no more than they do distant relatives and $K=1$ indicates that the data have strong phylogenetic signal.
+To rule out the possibility that gene trees with little phylogenetic signal were diluting support for the Ortholog Conjecture, we removed the `r k_percentile` of genes with the lowest phylogenetic signal, corresponding to a $K$ cutoff for retained genes of greater than `r signif( k_thresh, 3)`.
 Repeating the phylogenetic analyses on this subset of genes with strongest phylogenetic signal for Tau still do not reject the null hypothesis (Wilcoxon p = $`r  wilcox_test_result_k %>% round( 3 )`$).
+This indicates that our inability to reject the null hypothesis with phylogenetic methods is not due to a lack of phylogenetic signal.
+
+Phylogenetic independent contrasts (PICs) are based on a Brownian Motion (BM) model of character evolution [@Felsenstein:1985ua].
+PICs could fail to reject the null hypothesis if BM is not a good model for the evolution of Tau on the gene phylogenies.
+We therefore compared the fit of BM to the more complex Ornstein Uhlenbeck (OU) model of evolution.
 
@@
 This approach could be extended to approximate Phylogenetic Independent Contrasts under other continuous models of evolution as well.
+
+These analyses indicate that our inability to reject the null hypothesis is robust under different models of Tau evolution.
 

### 2017-11-12 manuscript_kernel.R

@@ -347,7 +347,7 @@
 
 ## ----phylogenetic_signal, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE----
 
-	k_thresh = quantile( tree_summary$K, na.rm=TRUE )[3] # third quantile is the median
+	k_thresh = quantile( tree_summary$K, na.rm=TRUE )[4] # third quantile is the median, fourth is 75% quartile
 	k_percentile = names(k_thresh)
 	genes_pass_k = tree_summary$gene[ tree_summary$K > k_thresh ]
 

### 2017-11-12 manuscript.rmd
updated pdf

@@
 	knitr::opts_chunk$set( 
-	  echo = FALSE,
+	  echo=FALSE,
 	  message=FALSE, 
@@
 This modified script and their data files are available in the github repository for this paper, as are the intermediate files that were generated by their analysis script that we used in our own analyses.
-We\ obtained the `Compara.75.protein.nh.emf` gene trees [@Herrero:2016jj] from ftp://ftp.ensembl.org/pub/release-75/emf/ensembl-compara/homologies/ and include them in our github repository.
+We obtained the `Compara.75.protein.nh.emf` gene trees [@Herrero:2016jj] from ftp://ftp.ensembl.org/pub/release-75/emf/ensembl-compara/homologies/ and include them in our github repository.
 These gene trees include branch lengths, annotate each internal node as being a duplication or speciation event, and provide a clade label for each internal node.
@@
 			geom_text(aes(x=thresh, label=thresh_label, y=mean(range(y))), colour="grey", angle=90) +
-			ggtitle( "A Density of K" )
+			ggtitle( "" )
 

### 2017-11-12 manuscript.rmd
Added AIC citations

@@
 
-```{r setup, cache=TRUE, message=FALSE, echo=FALSE, warning=FALSE}
+```{r setup, message=FALSE, echo=FALSE, warning=FALSE}
 
@@
 Our phylogenetic comparative analysis, unlike the previously published pairwise comparative analysis [@KryuchkovaMostacci:2016iw], therefore finds no support for the ortholog conjecture in this system.
-This result holds when we exclude the genes with the lowest phylogenetic signal according to Blomberg's $K$ [@Blomberg:2003jg] and is robust to model selection (*SI Appendix*).
+This result holds when we exclude the genes with the lowest phylogenetic signal according to Blomberg's $K$ [@Blomberg:2003jg] and is robust to model selection [@Akaike:1974ih; Posada:2004ix] (*SI Appendix*).
 
@@
 PICs could fail to reject the null hypothesis if BM is not a good model for the evolution of Tau on the gene phylogenies.
-We therefore compared the fit of BM to the more complex Ornstein Uhlenbeck (OU) model of evolution.
+We therefore compared the fit of BM to the more complex Ornstein Uhlenbeck (OU) model of evolution with the Akaike Information Criterion (AIC) [@Akaike:1974ih; Posada:2004ix].
 

### 2017-11-13 manuscript_kernel.R
added comments throughout

@@ -1,5 +1,15 @@
+# The computationally intensive analyses for the manuscript 
+# presented at https://github.com/caseywdunn/comparative_expression_2017 .
+#
+# Executing this code generates manuscript.rmd, which contains analysis 
+# results. That file is then read by manuscript.rmd for rendering and 
+# presentation of the results.
+# 
+# The code presented here is roughly in the order of the analyses presented 
+# in the manuscript, though there are exceptions
 
-## ----preliminaries, echo=FALSE, message=F, warning=F---------------------
+
+## Preliminaries
 	
 	time_start = Sys.time()
 	# The following should be installed from github with the specified 
@@ -24,7 +34,6 @@
 	source( "functions.R" )	
 	set.seed( 23456 )
 
-
 	# Set system computational parameters
 	cores = detectCores() - 1
 	if ( cores < 1 ) {
@@ -33,7 +42,14 @@
 
 	# Register parallel workers for %dopar%
 	registerDoParallel( cores )
-	
+
+## Set parameters
+
+	# In simulation, the fold change in rate following duplication relative to speciation
+	dup_adjust = 2
+
+## Load data for KMRR analyses
+
 	# Name of the compara gene trees file. This was downloaded from 
 	# ftp://ftp.ensembl.org/pub/release-75/emf/ensembl-compara/homologies/
 	# and compressed with gz
@@ -80,15 +96,8 @@
 	focal_calibrations_clades = 
 		c( "Hominini", "Homininae", "Catarrhini", "Euarchontoglires", "Theria", 
 			"Mammalia", "Amniota" )
-	
-	
-	# In simulation, the fold change in rate following duplication relative to speciation
-	dup_adjust = 2
 
-
-## ----load_trees, echo=FALSE, cache=TRUE, message=F, warning=F------------
-
-	# Read in the tree file as a vector of strings, one per line
+	# Read in the gene tree file as a vector of strings, one per line
 	lines = readLines( gene_tree_name )
 
 	# Isolate strings that are trees
@@ -120,9 +129,6 @@
 	# Remove the large string objects to free up memory
 	rm( lines )
 	rm( tree_lines )
-	
-
-## ----load_expression, echo=FALSE, message=F, warning=F-------------------
 
 	# Read the expression data for each species and combine into a single tibble,
[…]

### 2017-11-13 manuscript_kernel.R
white space cleanup

@@ -8,7 +8,6 @@
 # The code presented here is roughly in the order of the analyses presented 
 # in the manuscript, though there are exceptions
 
-
 ## Preliminaries
 	
 	time_start = Sys.time()
@@ -43,10 +42,13 @@
 	# Register parallel workers for %dopar%
 	registerDoParallel( cores )
 
+
 ## Set parameters
 
-	# In simulation, the fold change in rate following duplication relative to speciation
+	# In simulation, the fold change in rate following duplication 
+	# relative to speciation
 	dup_adjust = 2
+
 
 ## Load data for KMRR analyses
 
@@ -56,7 +58,8 @@
 
 	gene_tree_name = "Compara.75.protein.nhx.emf.gz"
 	
-	# Files produced by KMRR Rscript.R with Tau values and other expression statistics
+	# Files produced by KMRR Rscript.R with Tau values and other 
+	# expression statistics
 	expression_file_names = c(
 		"ChickenBrawandTScomparisonTable_9_6chPC.txt", 
 		"ChimpBrawandTScomparisonTable_9_6cmPC.txt", 
@@ -72,15 +75,16 @@
 	expression_file_names %<>% paste( kmrr_directory, ., sep="" )
 	gene_tree_name %<>% paste( kmrr_directory, ., sep="" )
 	
-	# The minimum number of genes with expression data that a gene needs to have to be considered
+	# The minimum number of genes with expression data that a gene 
+	# needs to have to be considered
 	min_genes_with_expression = 4
 	
-
-	# These clade dates are from the KMRR code, they got them from http://timetree.org/
+	# These clade dates are from the KMRR code, they got them 
+	# from http://timetree.org/
 	calibration_times = data.frame(
 		age =
-			c( 20, 92, 167, 9, 42, 74, 296, 535, 104, 937, 29, 722, 441, 65, 15, 1215, 
-				414, 371, 162, 25, 74, 77, 86, 7 ), 
+			c( 20, 92, 167, 9, 42, 74, 296, 535, 104, 937, 29, 722, 441, 65, 
+				15, 1215, 414, 371, 162, 25, 74, 77, 86, 7 ), 
 		clade = 
 			c("Hominoidea", "Euarchontoglires", "Mammalia", "Homininae", 
 				"Simiiformes", "Primates", "Amniota", "Vertebrata", "Eutheria", 
@@ -117,12 +121,12 @@
[…]

### 2017-11-13 manuscript.rmd
updated figure numbering

@@
 One of these trees is presented in *SI Appendix* Fig.
-4 to demonstrate the analysis.
+1 to demonstrate the analysis.
 
@@
 Our phylogenetic comparative analysis, unlike the previously published pairwise comparative analysis [@KryuchkovaMostacci:2016iw], therefore finds no support for the ortholog conjecture in this system.
-This result holds when we exclude the genes with the lowest phylogenetic signal according to Blomberg's $K$ [@Blomberg:2003jg] and is robust to model selection [@Akaike:1974ih; Posada:2004ix] (*SI Appendix*).
+This result holds when we exclude the genes with the lowest phylogenetic signal according to Blomberg's $K$ [@Blomberg:2003jg] and is robust to model selection [@Akaike:1974ih; Posada:2004ix] (*SI Appendix*, *SI Appendix* Fig.
+2).
 
@@
 We found no evidence that either affected our results (*SI Appendix*, Fig.
-5).
+3).
 We also examined the sensitivity of the results to the calibration times applied to speciation events on the gene trees.
@@
 These analyses still do not reject the null hypothesis, and do not provide support for the ortholog conjecture (*SI Appendix*, Fig.
-6).
+4).
 
@@
 This is apparent when node ages derived from calibrated trees are plotted against the age of the clade annotations for each node (*SI Appendix*, Fig.
-7*A*).
+S5*A*).
 Because duplication events that are all annotated with the same clade name can occur at very different times, pairwise comparisons across these nodes capture evolutionary changes in Tau across very different different branch lengths.
@@
 If only these duplication events closest to their annotated clade nodes are retained, the pattern of lower correlation of Tau evolution for duplication events disappears for the dataset simulated under the null model (*SI Appendix*, Fig.
-7*B*).
+S5*B*).
 This correction, though, comes at the high cost of discarding `r round( (n_dup_original - n_dup_remaining )/ n_dup_original * 100, 1 )`% of duplication nodes.
@@
 Fourth, we extended terminal branch lengths to understand the potential impact of within-species variation (*SI Appendix*, Fig.
-6).
+S4).
 The first two of these modifications are technical requirements for the application of phylogenetic comparative methods.
@@
 
-```{r Fig_SXX, fig.width=5, fig.height=3, fig.cap="Density plot of Blomberg's K.
+```{r Fig_S2, fig.width=5, fig.height=3, fig.cap="Density plot of Blomberg's K.
 The shaded portion of the plot indicates the gene trees that were analyzed after excluding the trees that fell below the K threshold."}
@@
 The gene trees considered here vary widely in their phylogenetic signal for Tau according to Blomberg's $K$ [@Blomberg:2003jg] (*SI Appendix*, Fig.
-XX).
+S2).
 $K$ describes the tendancy for closely related tips to resemble each other, where $K=0$ indicates that close relatives resemble each other no more than they do distant relatives and $K=1$ indicates that the data have strong phylogenetic signal.
@@
 While the age of speciation nodes is constrained, duplication nodes can be much older and therefore have a wider range of ages (*SI Appendix*, Fig.
-5*A*).
+S3*A*).
 This is because many gene duplication events are older than the most recent common ancestor of the species in the study.
@@
 The general results are unchanged and this reduced dataset does not reject the null hypothesis that the rate of evolution following duplications is the same as or less than the rate following speciation (*SI Appendix*, Fig.
-5*B*).
+S3*B*).
 
@@
 The expected variance of contrasts for speciation nodes is constrained by the branch lengths on the species tree, but the expected variance of contrasts for duplications has a much wider range (*SI Appendix*, Fig.
-5*C*).
+3*C*).
 This could lead to biases if the lengths of branches that descend from duplication nodes tend to be overestimated.
@@
 This reanalysis does not reject the null hypothesis either (*SI Appendix*, Fig.
-5*D*), indicating that branch length bias is not responsible for the result.
-
-```{r Fig_S2, fig.width=7, fig.height=6.5, fig.cap="Investigation of possible ascertainment biases.
+3*D*), indicating that branch length bias is not responsible for the result.
+
+```{r Fig_S3, fig.width=7, fig.height=6.5, fig.cap="Investigation of possible ascertainment biases.
 (*A*) Magnitude of independent contrasts plotted against node age.
@@
 
-```{r Fig_S3, fig.width=7, fig.height=3, fig.cap="Investigation of the potential impact of within-species variance.
+```{r Fig_S4, fig.width=7, fig.height=3, fig.cap="Investigation of the potential impact of within-species variance.
 These plots correspond to Fig.
@@
 
-```{r Fig_S4, fig.width=7, fig.height=3, fig.cap="Origin of pairwise correlation plot structure.
+```{r Fig_S5, fig.width=7, fig.height=3, fig.cap="Origin of pairwise correlation plot structure.
 (*A*) Each speciation and duplication node in the COMPARA trees is annotated with a clade name, and the age of that clade (shown on the x axis) is used as the age of the pairwise comparisons (Fig.
@@
 
-```{r Tau_v_max, fig.width=3.42, fig.height=3.42, fig.cap="Genes with higher maximum observed expression tend to have lower Tau.
-The blue line is the linear model for the relationship between the two."}
+```{r Fig_S6, fig.width=3.42, fig.height=3.42, fig.cap="Genes with higher maximum observed expression tend to have lower Tau. The blue line is the linear model for the relationship between the two."}
 
@@
 There is, however, a negative relationship between Tau and maximum expression across all genes (*SI Appendix* Fig.
-8, $r^2=`r signif( tau_v_max_lm$r.squared, 3 )`$, $slope=`r signif( tau_v_max_lm$coefficients[2,1], 3 )`$, $p=`r signif( tau_v_max_lm$coefficients[2,4], 3 )`$).
+S6, $r^2=`r signif( tau_v_max_lm$r.squared, 3 )`$, $slope=`r signif( tau_v_max_lm$coefficients[2,1], 3 )`$, $p=`r signif( tau_v_max_lm$coefficients[2,4], 3 )`$).
 A simpler explanation for their plot is that it reflects the global tendency for Tau to decrease with maximum expression.

### 2017-11-14 manuscript_kernel.R
added edge summary code

@@ -232,6 +232,9 @@
 	# Collect tree statistics in a single tibble, one row per tree
 	tree_summary = summarize_trees( gene_trees_pic )
 
+	# Collect edge statistics in a single tibble, one row per edge
+	edge_summary = summarize_edges( gene_trees_pic )
+
 	save.image("manuscript_checkpoint_contrasts.RData")
 
 

### 2017-11-14 manuscript.rmd
reset figure numbers in Supplement

@@
 
-```{r preliminaries}
+```{r preliminaries, cache=FALSE}
 
@@
 
-#### Reexamination of pairwise comparisons
+#### Pairwise comparisons oversample lineage-specific changes
 
@@
 
+\setcounter{figure}{0}
+\makeatletter 
+\renewcommand{\thefigure}{S\@arabic\c@figure}
+\renewcommand{\thetable}{S\@arabic\c@table}
+\makeatother
+
 ### KMRR Analyses
@@
 $K$ describes the tendancy for closely related tips to resemble each other, where $K=0$ indicates that close relatives resemble each other no more than they do distant relatives and $K=1$ indicates that the data have strong phylogenetic signal.
-To rule out the possibility that gene trees with little phylogenetic signal were diluting support for the Ortholog Conjecture, we removed the `r k_percentile` of genes with the lowest phylogenetic signal, corresponding to a $K$ cutoff for retained genes of greater than `r signif( k_thresh, 3)`.
+To examine the possibility that gene trees with little phylogenetic signal were diluting support for the Ortholog Conjecture, we removed the `r k_percentile` of genes with the lowest phylogenetic signal, corresponding to a $K$ cutoff for retained genes of greater than `r signif( k_thresh, 3)`.
 Repeating the phylogenetic analyses on this subset of genes with strongest phylogenetic signal for Tau still do not reject the null hypothesis (Wilcoxon p = $`r  wilcox_test_result_k %>% round( 3 )`$).
-This indicates that our inability to reject the null hypothesis with phylogenetic methods is not due to a lack of phylogenetic signal.
+This indicates that gene trees with little phylogenetic signal are not obscuring support for the Ortholog Conjecture in genes with greater phylogenetic signal.
 

### 2017-11-14 manuscript.rmd
more figure spacing
updated P value reporting to P throughout

@@
     fig_caption: yes
+    keep_tex: yes
   word_document: default
@@
 2*B*).
-The Wilcoxon rank test does not reject the null hypothesis that the rate of evolution following duplications is the same as or less than the rate following speciation (p value = `r wilcox_test_result`).
+The Wilcoxon rank test does not reject the null hypothesis that the rate of evolution following duplications is the same as or less than the rate following speciation ($P=`r wilcox_test_result`$).
 Our phylogenetic comparative analysis, unlike the previously published pairwise comparative analysis [@KryuchkovaMostacci:2016iw], therefore finds no support for the ortholog conjecture in this system.
@@
 1 [@KryuchkovaMostacci:2016iw], they present a higher Tau correlation coefficient between ortholog pairs than paralog pairs.
-We find the same here, with correlation coefficients of `r signif( ortholog_r$estimate, 2)` (p `r format_p(ortholog_r)`) for orthologs and `r signif( paralog_r$estimate, 2)` (p `r format_p( paralog_r )`) for paralogs.
+We find the same here, with correlation coefficients of `r signif( ortholog_r$estimate, 2)` ($P`r format_p(ortholog_r)`$) for orthologs and `r signif( paralog_r$estimate, 2)` ($P`r format_p( paralog_r )`$) for paralogs.
 
@@
 When we consider only the duplication events the same age or younger than the oldest speciation event (*SI Appendix*, Fig.
-5*B*), the paralog correlation coefficient increases from `r signif( paralog_r$estimate, 2)` to `r signif( paralog_subset_r$estimate, 2)` (p `r format_p(paralog_subset_r)`).
+5*B*), the paralog correlation coefficient increases from `r signif( paralog_r$estimate, 2)` to `r signif( paralog_subset_r$estimate, 2)` ($P`r format_p(paralog_subset_r)`$).
 This is much closer to the ortholog correlation of `r signif( ortholog_r$estimate, 2) `.
@@
 2*A*).
-As in the original study, there is higher correlation coefficient across orthologs (`r signif( ortholog_r_sim_null$estimate, 2 )`, p `r format_p(ortholog_r_sim_null)`) than paralogs (`r signif( paralog_r_sim_null$estimate, 2 )`, p `r format_p(paralog_r_sim_null)`) when not considering node age.
+As in the original study, there is higher correlation coefficient across orthologs (`r signif( ortholog_r_sim_null$estimate, 2 )`, $P`r format_p(ortholog_r_sim_null)`$) than paralogs (`r signif( paralog_r_sim_null$estimate, 2 )`, $P`r format_p(paralog_r_sim_null)`$) when not considering node age.
 Phylogenetic analysis of the data simulated under the null hypothesis (Fig.
-2*D*) do not reject the null hypothesis (Wilcoxon p = $`r wilcox_oc( nodes_sim_null_contrast ) %>% round( 3 )`$), as expected.
+2*D*) do not reject the null hypothesis (Wilcoxon $P=`r wilcox_oc( nodes_sim_null_contrast ) %>% round( 3 )`$), as expected.
 
@@
 2*E*) are nearly indistinguishable from the results under the null model (Fig.
-2*C*), and also have a higher correlation coefficient for orthologs (`r signif( ortholog_r_sim_oc$estimate, 2 )`, p `r format_p( ortholog_r_sim_oc )`) than paralogs (`r signif( paralog_r_sim_oc$estimate, 2 )`, p `r format_p( paralog_r_sim_oc )`).
+2*C*), and also have a higher correlation coefficient for orthologs (`r signif( ortholog_r_sim_oc$estimate, 2 )`, $P`r format_p( ortholog_r_sim_oc )`$) than paralogs (`r signif( paralog_r_sim_oc$estimate, 2 )`, $P`r format_p( paralog_r_sim_oc )`$).
 The phylogenetic analysis of the ortholog conjecture simulation (Fig.
-2*F*) does reject the null hypothesis (Wilcoxon p = $`r signif( wilcox_oc( nodes_sim_oc_contrast ), 2 )`$).
+2*F*) does reject the null hypothesis (Wilcoxon $P=`r signif( wilcox_oc( nodes_sim_oc_contrast ), 2 )`$).
 
@@
 
+```{r implications}
+
+	edge_p = wilcox.test( 
+		edge_summary %>% filter( event_parent=="Duplication" ) %>% .$scaled_change, 
+		edge_summary %>% filter( event_parent=="Speciation" ) %>% .$scaled_change, 
+		alternative="greater" 
+	)$p.value
+
+```
+
 There has been considerable recent interest in, and controversy about, the ortholog conjecture [@Koonin:2005fp; @Studer:2009bm; @Nehrt:2011bm; @Gabaldon:2013fi; @Altenhoff:2012ea].
@@
 Pairwise scores for the ctenophore are red.
-Wilcoxon test p-values for the significance of the differences between early-mid distributions and late-mid distributions are on the right.
+Wilcoxon test $P$ values for the significance of the differences between early-mid distributions and late-mid distributions are on the right.
 Model of variance, which is inversely related to similarity, is on the left.
@@
 The distributions of similarity scores are plotted in their Fig.
-4d [@Levin:2016bp], and their Kolmogorov-Smirnov (KS) tests indicated that the early distribution and late distribution were each significantly different from mid distribution (P < $10^{-6}$ and P < $10^{-12}$, respectively).
+4d [@Levin:2016bp], and their Kolmogorov-Smirnov (KS) tests indicated that the early distribution and late distribution were each significantly different from mid distribution ($P < 10^{-6}$ and $P < 10^{-12}$, respectively).
 This is the support they presented for the inverse hourglass model.
@@
 The Wilcoxon test is instead appropriate in this case.
-When applied to the de-duplicated data, the significance of this test is 0.02 for the early-mid comparison and on the order of $10^{-7}$ for early-late.
+When applied to the de-duplicated data, the $P$ of this test is 0.02 for the early-mid comparison and on the order of $10^{-7}$ for early-late.
 
@@
 3*A*) are for pairwise comparisons that include the ctenophore (comb jelly).
-When the nine pairwise comparisons that include the ctenophore are removed, there is no significant difference between the early phase and mid phase distributions (p = 0.14 for the early-mid comparison and p < $10^{-5}$ for the late-mid comparison) and no support for the inverse hourglass (Fig.
+When the nine pairwise comparisons that include the ctenophore are removed, there is no significant difference between the early phase and mid phase distributions ($P=0.14$ for the early-mid comparison and $P<10^{-5}$ for the late-mid comparison) and no support for the inverse hourglass (Fig.
 3*B*).
@@
 To examine the possibility that gene trees with little phylogenetic signal were diluting support for the Ortholog Conjecture, we removed the `r k_percentile` of genes with the lowest phylogenetic signal, corresponding to a $K$ cutoff for retained genes of greater than `r signif( k_thresh, 3)`.
-Repeating the phylogenetic analyses on this subset of genes with strongest phylogenetic signal for Tau still do not reject the null hypothesis (Wilcoxon p = $`r  wilcox_test_result_k %>% round( 3 )`$).
+Repeating the phylogenetic analyses on this subset of genes with strongest phylogenetic signal for Tau still do not reject the null hypothesis (Wilcoxon $P=`r  wilcox_test_result_k %>% round( 3 )`$).
 This indicates that gene trees with little phylogenetic signal are not obscuring support for the Ortholog Conjecture in genes with greater phylogenetic signal.
@@
 
-Repeating the BM phylogenetic analyses on the subset of `r sum( tree_summary$dAIC_bm < dAIC_thresh )` genes that have $\Delta$AIC less than the threshold for BM still does not reject the null hypothesis (Wilcoxon p = $`r  wilcox_test_result_bm %>% round( 3 )`$).
+Repeating the BM phylogenetic analyses on the subset of `r sum( tree_summary$dAIC_bm < dAIC_thresh )` genes that have $\Delta$AIC less than the threshold for BM still does not reject the null hypothesis (Wilcoxon $P=`r  wilcox_test_result_bm %>% round( 3 )`$).
 
 We next approximated the Phylogenetic Independent Contrasts for the the genes that have $\Delta$AIC less than the threshold for OU under an OU model.
-These analyses do not reject the null hypothesis (Wilcoxon p = $`r  wilcox_ou %>% round( 3 )`$).
+These analyses do not reject the null hypothesis (Wilcoxon $P=`r  wilcox_ou %>% round( 3 )`$).
 The challenge with such an analysis is that BM is intrinsic to the classic implementation of Phylogenetic Independent Contrasts, which use the fact that the expected variance is proportional to branch length to scale each contrast.
@@
 If any daughter nodes became older than their parent, we repeated the sampling until the dates were congruent with the topology.
-The minimum Wilcoxon p in these reanalyses was $`r min( as.numeric( unlist( p_noised ) ) )`$, *i.e.* none of them reject the null hypothesis that the rate of evolution of Tau is greater following duplication events than speciation events.
+The minimum Wilcoxon $P$ in these reanalyses was $`r min( as.numeric( unlist( p_noised ) ) )`$, *i.e.* none of them reject the null hypothesis that the rate of evolution of Tau is greater following duplication events than speciation events.
 This is consistent with the analysis that uses the calibration dates as specified, indicating that our results are robust to the selection of calibration times for speciation nodes.

### 2017-11-14 functions.R

@@ -678,7 +678,7 @@ get_pairwise_summary = function ( nhx ) {
 make_contrast_plot = function( nodes_contrast, mode="diff" ){
 	
 	wilcox_test_result = wilcox_oc( nodes_contrast )
-	label_p = str_c( "Wilcoxon p=", signif( wilcox_test_result, 2 ) )
+	label_p = str_c( "Wilcoxon P=", signif( wilcox_test_result, 2 ) )
 	
 	if( mode == "freqpoly" ){
 		ggcontrasts = nodes_contrast %>%

### 2017-11-16 manuscript.rmd
incorporated ssiebert comments, pdf not rebuilt

@@
 The ortholog conjecture [@Nehrt:2011bm] is the proposition that orthologs (genes that diverged from each other due to a speciation event) have more similar attributes than do paralogs (genes that diverged from each other due to a gene duplication event).
-The ortholog can and has been applied to diverse attributes, including molecular sequence, biochemical function, and, as in the study considered here, expression.
-The ortholog conjecture has important biological and technical implications.
+The ortholog conjecture can and has been applied to diverse attributes, including molecular sequence, biochemical function, and, as in the study considered here, expression.
+It has important biological and technical implications.
 It shapes our understanding of the functional diversity of gene families.
@@
 This dataset is the best sampled in their analyses.
-It has gene expression data for six organs across ten species (nine mammals and one bird), eight of which were analyzed by KMRR and further considered here.
+It has gene expression data for six organs across ten species (nine mammals and one bird), eight of which were analyzed by KMRR and are further considered here.
 
@@
 One of these trees is presented in *SI Appendix* Fig.
-1 to demonstrate the analysis.
+S1 to demonstrate the analysis.
 
@@
 Our phylogenetic comparative analysis, unlike the previously published pairwise comparative analysis [@KryuchkovaMostacci:2016iw], therefore finds no support for the ortholog conjecture in this system.
-This result holds when we exclude the genes with the lowest phylogenetic signal according to Blomberg's $K$ [@Blomberg:2003jg] and is robust to model selection [@Akaike:1974ih; Posada:2004ix] (*SI Appendix*, *SI Appendix* Fig.
-2).
+This result holds when we exclude the genes with the lowest phylogenetic signal according to Blomberg's $K$ [@Blomberg:2003jg] and is robust to model selection [@Akaike:1974ih; @Posada:2004ix] (*SI Appendix*, *SI Appendix* Fig.
+S2).
 
@@
 We found no evidence that either affected our results (*SI Appendix*, Fig.
-3).
+S3).
 We also examined the sensitivity of the results to the calibration times applied to speciation events on the gene trees.
@@
 These analyses still do not reject the null hypothesis, and do not provide support for the ortholog conjecture (*SI Appendix*, Fig.
-4).
+S4).
 
@@
 To test the hypothesis that ancient duplications that precede the oldest speciation event (*SI Appendix*, Fig.
-5*A*) impact the lower correlation of Tau between paralogs than between orthologs, we removed them.
+S5*A*) impact the lower correlation of Tau between paralogs than between orthologs, we removed them.
 When we consider only the duplication events the same age or younger than the oldest speciation event (*SI Appendix*, Fig.
@@
 For each node in the remaining calibrated trees, we calculated the phylogenetic independent contrast for Tau across its daughter branches with the `pic()` function in ape [@Paradis:2004dv].
-We then collected the contrasts from all trees into a single table, along with other annotations including whether the node is a speciation or duplication event.
-This table, `nodes_contrast`, was then analyzed as described in the main text for the presented plots and tests.
+Summary statistics on taxon and node sampling are presented in *SI Appendix* Tables S1-S2.
+We then collected the contrasts from all trees into a single data frame, along with other annotations including whether the node is a speciation or duplication event.
+This data frame, `nodes_contrast`, was then analyzed as described in the main text for the presented plots and tests.
 
@@
 S2).
-$K$ describes the tendancy for closely related tips to resemble each other, where $K=0$ indicates that close relatives resemble each other no more than they do distant relatives and $K=1$ indicates that the data have strong phylogenetic signal.
+$K$ describes the tendency for closely related tips to resemble each other, where $K=0$ indicates that close relatives resemble each other no more than they do distant relatives and $K=1$ indicates that the data have strong phylogenetic signal.
 To examine the possibility that gene trees with little phylogenetic signal were diluting support for the Ortholog Conjecture, we removed the `r k_percentile` of genes with the lowest phylogenetic signal, corresponding to a $K$ cutoff for retained genes of greater than `r signif( k_thresh, 3)`.
@@
 PICs could fail to reject the null hypothesis if BM is not a good model for the evolution of Tau on the gene phylogenies.
-We therefore compared the fit of BM to the more complex Ornstein Uhlenbeck (OU) model of evolution with the Akaike Information Criterion (AIC) [@Akaike:1974ih; Posada:2004ix].
+We therefore compared the fit of BM to the more complex Ornstein Uhlenbeck (OU) model of evolution with the Akaike Information Criterion (AIC) [@Akaike:1974ih; @Posada:2004ix].
 
@@
 The expected variance of contrasts for speciation nodes is constrained by the branch lengths on the species tree, but the expected variance of contrasts for duplications has a much wider range (*SI Appendix*, Fig.
-3*C*).
+S3*C*).
 This could lead to biases if the lengths of branches that descend from duplication nodes tend to be overestimated.
-We therefore examined only the contrasts that fell within in the range of expected variance seen for speciation contrasts, excluding duplication contrasts that fall outside of this range.
+We therefore examined only the contrasts that fell within the range of expected variance seen for speciation contrasts, excluding duplication contrasts that fall outside of this range.
 This reanalysis does not reject the null hypothesis either (*SI Appendix*, Fig.
-3*D*), indicating that branch length bias is not responsible for the result.
+S3*D*), indicating that branch length bias is not responsible for the result.
 

### 2017-11-16 manuscript.rmd
spelling

@@
 
-	# The threshold of allowed fractional devience of node age from clade age
+	# The threshold of allowed fractional deviance of node age from clade age
 	age_window = 0.1
@@
 It then scales the observed contrasts by the mean magnitude of the corresponding contrasts in the simulations.
-We validated the method by comparing picx BM results to the standard pic results calcualted by ape.
-These validations indicated a strong linear realationship between the two ($R^2$ > 0.95).
+We validated the method by comparing picx BM results to the standard pic results calculated by ape.
+These validations indicated a strong linear relationship between the two ($R^2$ > 0.95).
 This approach could be extended to approximate Phylogenetic Independent Contrasts under other continuous models of evolution as well.

### 2017-11-16 manuscript.rmd
made zapta edits, pdf not rebuilt

@@
 
-^3^ Department of Ecology and Evolutionary Biology, University of California Los Angeles, Los Angeles, CA, 20095, USA
+^3^ Department of Ecology and Evolutionary Biology, University of California Los Angeles, Los Angeles, CA, 90095, USA
 
@@
 It is also used to relate findings from well-studied genes to homologous genes that have not been investigated in detail.
-It has been applied to many traits of genes, from gene sequence to biochemical properties to expression.
 While the ortholog conjecture describes a specific pattern of functional diversity across genes, it is also articulated as a hypothesis about the process of evolution-- that there is greater evolutionary change in gene attributes following a duplication event than a speciation event.
@@
 The phylogenetic analysis of the ortholog conjecture simulation (Fig.
-2*F*) does reject the null hypothesis (Wilcoxon $P=`r signif( wilcox_oc( nodes_sim_oc_contrast ), 2 )`$).
+2*F*) does reject the null hypothesis (Wilcoxon $P=`r signif( wilcox_oc( nodes_sim_oc_contrast ), 2 )`$), as expected if the phylogenetic methods are working correctly.
 
@@
 This is because there are various species-specific technical factors that can be mistaken for differences in expression across species.
-These can be canceled out within species before making comparisons across species [@Dunn:2013cpa].
+These can be canceled out within species before making comparisons across species.
 
@@
 
-#### Phylogenetic signal and models
+#### Phylogenetic signal and models of evolution
 
@@
 It then scales the observed contrasts by the mean magnitude of the corresponding contrasts in the simulations.
-We validated the method by comparing picx BM results to the standard pic results calculated by ape.
-These validations indicated a strong linear relationship between the two ($R^2$ > 0.95).
+We validated the method by comparing picx BM results to the standard pic results calculated by ape [@Paradis:2004dv].
+These validations indicated a strong linear relationship between the two ($R^2=0.98$, $P=1.39\times 10^{−23}$).
 This approach could be extended to approximate Phylogenetic Independent Contrasts under other continuous models of evolution as well.
@@
 There is, however, a negative relationship between Tau and maximum expression across all genes (*SI Appendix* Fig.
-S6, $r^2=`r signif( tau_v_max_lm$r.squared, 3 )`$, $slope=`r signif( tau_v_max_lm$coefficients[2,1], 3 )`$, $p=`r signif( tau_v_max_lm$coefficients[2,4], 3 )`$).
+S6, $r^2=`r signif( tau_v_max_lm$r.squared, 3 )`$, $slope=`r signif( tau_v_max_lm$coefficients[2,1], 3 )`$, $P=`r signif( tau_v_max_lm$coefficients[2,4], 3 )`$).
 A simpler explanation for their plot is that it reflects the global tendency for Tau to decrease with maximum expression.

### 2017-11-16 manuscript.rmd
updated significance

@@
 We re-examined two previously published studies and found problems with pairwise comparisons that draw both their original conclusions into question.
-One study had found support for the ortholog conjecture and the other had concluded that the evolution of gene expression was different between animal phyla than within them.
+One study found support for the ortholog conjecture and the other concluded that the evolution of gene expression differed in pattern and process between animal phyla versus within animal phyla.
 Our results demonstrate that to answer evolutionary questions about genome function, it is critical to consider evolutionary relationships.

### 2017-11-17 manuscript.rmd
incorporated cmunro edits, shortened, edits throughout. pdf not rebuilt

@@
 Functional genomics, in turn, is now becoming more comparative-- there is great interest in understanding how functional genomic variation across species gives rise to a diversity of development, morphology, physiology, and other phenotypes [@Wray:2013gh].
-These analyses are also critical to transferring functional insight across species, and will grow in importance in coming years.
+These analyses are critical to transferring functional insight across species, and will grow in importance in coming years.
 
@@
 The value of these methods relative to pairwise comparisons has been repeatedly shown in analyses of other types of character data [@Garland:2005ek; @Ricklefs:1996wb; @Chamberlain:2012vx; @OMeara:2012gq].
-The application of these methods is illustrated by a study of leaf functional attributes for about 100 plant species [@Ackerly:1999iu].
-If phylogenetic relationships are not considered, analyses indicate a negative correlation of leaf lifespan with leaf size.
-With the application of phylogenetic comparative methods, however, this correlation disappears because most of the observed variance is due to differences between just two groups, conifers and flowering plants.
-Phylogenetic comparative methods capture this shift as change along a single branch deep in the tree, revealing that there is not a tendency of correlated change between these traits across the phylogeny.
 One reason that comparative functional genomic studies have not embraced phylogenetic approaches is that there has not yet been a concrete demonstration that pairwise and phylogenetic comparative methods can lead to different results when considering functional genomic data.
@@
 KMRR considered several publicly available datasets of gene expression across tissues and species.
-Their expression summary statistic is Tau [@Yanai:2005dh, @KryuchkovaMostacci:2016in], an indicator of tissue specificity of gene expression.
+Their expression summary statistic is tau [@Yanai:2005dh, @KryuchkovaMostacci:2016in], an indicator of tissue specificity of gene expression.
 Tau can range from a value of 0, which indicates no specificity (*i.e.*, uniform expression across tissues), to a value of 1, which indicates high specificity (*i.e.*, expression in only one tissue).
@@
 This includes information about which tissue expression is specific to.
-For example, if one gene has expression specific to the brain and another expression specific to the kidney, both would have a Tau of 1.
+For example, if one gene has expression specific to the brain and another expression specific to the kidney, both would have a tau of 1.
 
 The KMRR analyses are based on pairwise comparisons (Fig.
-1*A*) between Tau within each gene family.
+1*A*) between tau within each gene family.
 Rather than make every pairwise comparison within each gene tree, they considered only a subset of pairwise comparisons in each particular analysis.
@@
 
-They found the correlation coefficient of Tau for orthologs to be significantly greater than the correlation coefficient of Tau for paralogs, *i.e.* orthologs tend to have more similar expression than do paralogs.
+They found the correlation coefficient of tau for orthologs to be significantly greater than the correlation coefficient of tau for paralogs, *i.e.* orthologs tend to have more similar expression than do paralogs.
 From this they concluded that their analyses support the ortholog conjecture.
@@
 
-For each internal node in each gene tree, we calculated the phylogenetic independent contrast [@Felsenstein:1985ua] (PIC) of Tau.
-This is the difference in values of Tau for descendant nodes scaled by the expected variance, which is largely determined by the lengths of the two branches that connect the node to its two descendants (Fig.
+For each internal node in each gene tree, we calculated the phylogenetic independent contrast [@Felsenstein:1985ua] (PIC) of tau.
+This is the difference in values of tau for descendant nodes scaled by the expected variance, which is largely determined by the lengths of the two branches that connect the node to its two descendants (Fig.
 1*B*).
@@
 
-We mapped the Tau values calculated by KMRR [@KryuchkovaMostacci:2016iw] for the Brawand *et al.* 2011 [@Brawand:2011du] dataset onto `r n_gene_trees %>% format(big.mark=",")` gene trees parsed from ENSEMBL Compara [@Herrero:2016jj].
+We mapped the tau values calculated by KMRR [@KryuchkovaMostacci:2016iw] for the Brawand *et al.* 2011 [@Brawand:2011du] dataset onto `r n_gene_trees %>% format(big.mark=",")` gene trees parsed from ENSEMBL Compara [@Herrero:2016jj].
 These are the same pre-computed trees that the orthology/ paralogy annotations KMRR used are based on.
-`r length(gene_trees_pruned) %>% format(big.mark=",")` gene trees passed taxon sampling criteria (`r min_genes_with_expression %>% format(big.mark=",")` genes) after removing tips without Tau values and had at least one speciation event.
+`r length(gene_trees_pruned) %>% format(big.mark=",")` gene trees passed taxon sampling criteria (`r min_genes_with_expression %>% format(big.mark=",")` genes) after removing tips without tau values and had at least one speciation event.
 Of these, `r length(gene_trees_calibrated)  %>% format(big.mark=",")` were successfully time calibrated.
@@
 Our phylogenetic comparative analysis, unlike the previously published pairwise comparative analysis [@KryuchkovaMostacci:2016iw], therefore finds no support for the ortholog conjecture in this system.
-This result holds when we exclude the genes with the lowest phylogenetic signal according to Blomberg's $K$ [@Blomberg:2003jg] and is robust to model selection [@Akaike:1974ih; @Posada:2004ix] (*SI Appendix*, *SI Appendix* Fig.
+This result holds when we exclude the genes with the lowest phylogenetic signal according to Blomberg's $K$ [@Blomberg:2003jg] and is robust to model selection [@Akaike:1974ih; @Posada:2004ix; @Pennell:2014jf] (*SI Appendix*, *SI Appendix* Fig.
 S2).
 
-We next examined the possibility that ascertainment biases were differentially impacting the inference of expression evolution following duplication and speciation events.
+We examined the possibility that ascertainment biases were differentially impacting the inference of expression evolution following duplication and speciation events.
 Such a bias might obscure support for the ortholog conjecture.
@@
 
-An additional concern is that within-species expression variation is not considered when making comparisons of Tau across homologous genes.
+An additional concern is that within-species expression variation is not considered when making comparisons of tau across homologous genes.
 Neglecting within species variation is expected to mislead phylogenetic comparative analyses in some cases [@Ives:2007cw; @Felsenstein:2008cl; @Rohlfs:2015bd].
@@
 In their Fig.
-1 [@KryuchkovaMostacci:2016iw], they present a higher Tau correlation coefficient between ortholog pairs than paralog pairs.
+1 [@KryuchkovaMostacci:2016iw], they present a higher tau correlation coefficient between ortholog pairs than paralog pairs.
 We find the same here, with correlation coefficients of `r signif( ortholog_r$estimate, 2)` ($P`r format_p(ortholog_r)`$) for orthologs and `r signif( paralog_r$estimate, 2)` ($P`r format_p( paralog_r )`$) for paralogs.
@@
 To test the hypothesis that ancient duplications that precede the oldest speciation event (*SI Appendix*, Fig.
-S5*A*) impact the lower correlation of Tau between paralogs than between orthologs, we removed them.
+S5*A*) impact the lower correlation of tau between paralogs than between orthologs, we removed them.
 When we consider only the duplication events the same age or younger than the oldest speciation event (*SI Appendix*, Fig.
@@
 ```{r Fig_2, fig.width=7, fig.height=6.5, fig.cap="Pairwise (left column - *A, C, E*) and phylogenetic (right column - *B, D, F*) analyses of the original data (first row - *A, B*), data simulated under the null hypothesis (second row - *C, D*), and data simulated under the ortholog conjecture (third row- *E, F*).
-In the pairwise plots, each point indicates the correlation coefficient of Tau for a set of pairwise comparisons annotated with a specific node name (e.g, Primates) and event type (speciation or duplication, giving rise respectively to orthologs and paralogs).
-The phylogenetic plots show the difference between the density distributions for Tau phylogenetic contrasts for speciation and duplication events, where a value above 0 indicates an excess of speciation contrasts in the indicated interval.
+In the pairwise plots, each point indicates the correlation coefficient of tau for a set of pairwise comparisons annotated with a specific node name (e.g, Primates) and event type (speciation or duplication, giving rise respectively to orthologs and paralogs).
+The phylogenetic plots show the difference between the density distributions for tau phylogenetic contrasts for speciation and duplication events, where a value above 0 indicates an excess of speciation contrasts in the indicated interval.
 A horizontal line at 0 would indicate that the density distributions are identical.
@@
 In their Fig.
-2 [@KryuchkovaMostacci:2016iw], they grouped orthologs and paralogs according to the ENSEMBL node name of their most recent common ancestor, and plotted the correlation of Tau for each of these groups by the node age.
-They found that across the investigated range of node ages, ortholog pairs have higher Tau correlation than paralogs.
+2 [@KryuchkovaMostacci:2016iw], they grouped orthologs and paralogs according to the ENSEMBL node name of their most recent common ancestor, and plotted the correlation of tau for each of these groups by the node age.
+They found that across the investigated range of node ages, ortholog pairs have higher tau correlation than paralogs.
 We confirmed that we can replicate this result (Fig.
@@
 To better understand this plot (Fig.
-2*A*), we performed simulations of Tau on the calibrated gene trees.
+2*A*), we performed simulations of tau on the calibrated gene trees.
 We did not modify the gene tree topologies or their inferred histories of duplication and loss.
-First, we simulated the evolution of Tau under the null model that it evolves at the same rate following duplication and speciation events.
+First, we simulated the evolution of tau under the null model that it evolves at the same rate following duplication and speciation events.
 Under the null model, the plot of correlation coefficient to node age (Fig.
@@
 
-We next simulated the evolution of Tau under the ortholog conjecture, where the rate of evolution of Tau following duplication was `r dup_adjust` fold the rate following speciation.
+We next simulated the evolution of tau under the ortholog conjecture, where the rate of evolution of tau following duplication was `r dup_adjust` fold the rate following speciation.
 The pairwise results of this heterogeneous model (Fig.
@@
 
-Since greater rates of expression evolution following duplication do not explain the lower correlation of Tau values in the pairwise comparison plots (Fig.
+Since greater rates of expression evolution following duplication do not explain the lower correlation of tau values in the pairwise comparison plots (Fig.
 2 *A, C, E*), this pattern must reflect some other property that is shared across these analyses.
@@
 S5*A*).
-Because duplication events that are all annotated with the same clade name can occur at very different times, pairwise comparisons across these nodes capture evolutionary changes in Tau across very different different branch lengths.
+Because duplication events that are all annotated with the same clade name can occur at very different times, pairwise comparisons across these nodes capture evolutionary changes in tau across very different different branch lengths.
 They therefore have lower correlation than pairwise comparisons made across speciation nodes.
 Of the `r n_dup_original %>% format(big.mark=",")` duplication nodes considered here, `r n_dup_remaining %>% format(big.mark=",")` have a node age (as determined by the time calibrated trees) within `r age_window * 100`% of their clade annotation.
-If only these duplication events closest to their annotated clade nodes are retained, the pattern of lower correlation of Tau evolution for duplication events disappears for the dataset simulated under the null model (*SI Appendix*, Fig.
+If only these duplication events closest to their annotated clade nodes are retained, the pattern of lower correlation of tau evolution for duplication events disappears for the dataset simulated under the null model (*SI Appendix*, Fig.
 S5*B*).
@@
 It may be the case that the rates of phenotypic evolution following duplication may be greater than that following duplication in some organisms, gene families, and evolutionary processes [@Gabaldon:2013fi].
-We just do not find evidence for it when summarizing gene expression across tissues with Tau in these organisms.
+We just do not find evidence for it when summarizing gene expression across tissues with tau in these organisms.
 This calls into question the general predictive power of the ortholog conjecture with respect to gene expression, and until these processes are better understood it will be necessary to test for it in each situation.
@@
 We found that every data point was included twice because both reciprocal pairwise comparisons (which have the same values) were retained.
-For example, there is both a nematode to arthropod comparison and an arthropod to nematode comparison.
 As a consequence, there are 90 entries for the 45 pairwise comparisons, and by doubling the data the significance of the result appears stronger than it actually is.
@@
 The KS test does not just evaluate whether one distribution is greater than the other, it also tests whether the shape of the distributions are the same.
-In addition, the samples in this dataset are matched (*i.e.*, for each pairwise comparison there is a early, mid, and late expression value), which the KS test does not take into account.
+The samples in this dataset are matched (*i.e.*, for each pairwise comparison there is a early, mid, and late expression value), which the KS test does not take into account.
 The Wilcoxon test is instead appropriate in this case.
@@
 
-Once we addressed the issues above with the implementation of pairwise comparisons, we were able to explore more general issues that can be a problem when making multiple pairwise comparisons between species.
+Once we addressed the issues above, we were able to explore more general issues that can be a problem when making multiple pairwise comparisons between species.
 We found that all five of the lowest values in the mid phase distribution (Fig.
@@
 
-#### Challenges and opportunities
-
-There are interesting new challenges at the interdisciplinary interface between phylogenetic comparative methods and functional genomics that will need to be addressed to fully realize the potential of integrating these approaches.
-One such challenge is that most phylogenetic comparative analyses of covariance between traits have been developed to address problems with many more species (e.g., dozens or more) relative to the number of traits being examined.
-In comparative functional genomic analyses, there are often far fewer species because adding each species is still expensive, but high-throughput tools generate data for tens of thousands of traits per species.
-This creates statistical challenges as the resulting covariance matrices are singular and, if not treated appropriately, imply many false correlations that are artifacts of project design.
-We outlined these challenges and potential solutions in the context of gene expression elsewhere [@Dunn:2013cpa].
-
-Another challenge is parameterizing differential expression in such a way that they can be analyzed in a phylogenetic context.
-Read counts generated by RNA-seq expression studies cannot be directly compared across species [@Dunn:2013cpa].
-This is because there are various species-specific technical factors that can be mistaken for differences in expression across species.
-These can be canceled out within species before making comparisons across species.
+We are not the first to apply phylogenetic comparative methods to functional genomic data.
+While the vast majority of comparative functional genomic studies have used standard pairwise similarity methods, a small number of comparative functional genomic studies have employed phylogenetic comparative approaches [@Oakley:2005ky; @Eng:2009gp; @Chang:2014hq; @Whitney:2011fw].
+For instance, a phylogenetic ANOVA [@Rohlfs:2015bd] of the evolution of gene expression in strict orthologs improves statistical power and drastically reduces the rate of false positives relative to pairwise approaches.
+
+Some of the most widely used phylogenetic comparative methods [@Felsenstein:1985ua; @Grafen:1989um] are already directly applicable to comparative functional genomic studies.
+There are also interesting new challenges, such as parameterizing differential expression so that species-specific technical biases are not mistaken for evolutionary changes in expression [@Dunn:2013cpa].
 
@@
 
-#### Additional approaches and future directions
-
-We are not the first to apply phylogenetic comparative methods to functional genomic data.
-While the vast majority of comparative functional genomic studies have used standard pairwise similarity methods, a small number of comparative functional genomic studies have employed phylogenetic comparative approaches [@Oakley:2005ky; @Eng:2009gp; @Chang:2014hq; @Whitney:2011fw].
-For instance, a phylogenetic ANOVA [@Rohlfs:2015bd] of the evolution of gene expression in strict orthologs improves statistical power and drastically reduces the rate of false positives relative to pairwise approaches.
-
-Some of the most widely used phylogenetic comparative methods [@Felsenstein:1985ua; @Grafen:1989um] are already directly applicable to comparative functional genomic studies.
-Recent advances in phylogenetic comparative methods are particularly well suited to addressing questions about the evolution of functional genomic traits.
-Most early phylogenetic comparative methods attempted to account for evolutionary signal to correct statistical tests for correlations between traits, while more recent methods tend to focus on testing hypotheses of evolutionary processes [@Pennell:2013cw].
-The application of this newer focus to functional genomics provides an exciting opportunity to address long standing questions of broad interest, including the order of changes in functional genomic traits and shifts in rates of evolution of one functional genomic trait following changes in another trait.
-
 ## Conclusions
@@
 
-After parsing the trees from the Compara file with treeio, which was recently split from ggtree [@Yu:2016fo], we added Tau estimates generated by the KMRR `Rscript.R` to the tree data objects.
+After parsing the trees from the Compara file with treeio, which was recently split from ggtree [@Yu:2016fo], we added tau estimates generated by the KMRR `Rscript.R` to the tree data objects.
 We then pruned away tips without expression data, retaining only the trees with `r min_genes_with_expression %>% format(big.mark=",")` or more tips.
@@
 The node names in the ENSEMBL Compara [@Herrero:2016jj] gene phylogenies are parsed from the NCBI Taxonomy database, which has many polytomies, rather than a bifurcating species phylogeny.
-One implication of this is that node names can be resolved in such a way that a speciation node can have the same name as one of its speciation node ancestors, as others have noted [@Daub:2016gd].
+One implication of this is that node names can be resolved so that a speciation node can have the same name as one of its speciation node ancestors, as others have noted [@Daub:2016gd].
 If left unaddressed, this would force all intervening branches to have length zero and interfere with calibration.
@@
 
-For each node in the remaining calibrated trees, we calculated the phylogenetic independent contrast for Tau across its daughter branches with the `pic()` function in ape [@Paradis:2004dv].
+For each node in the remaining calibrated trees, we calculated the phylogenetic independent contrast for tau across its daughter branches with the `pic()` function in ape [@Paradis:2004dv].
 Summary statistics on taxon and node sampling are presented in *SI Appendix* Tables S1-S2.
@@
 ```{r Fig_S1, fig.width=7, fig.height=6.5, fig.cap="One of the gene trees from the phylogenetic analysis.
-The value of Tau (expression specificity) is indicated by the sizes of the circles at the tips of the tree.
+The value of tau (expression specificity) is indicated by the sizes of the circles at the tips of the tree.
 Whether an internal node is a speciation or duplication is indicated by color.
@@
 
-The gene trees considered here vary widely in their phylogenetic signal for Tau according to Blomberg's $K$ [@Blomberg:2003jg] (*SI Appendix*, Fig.
+The gene trees considered here vary widely in their phylogenetic signal for tau according to Blomberg's $K$ [@Blomberg:2003jg] (*SI Appendix*, Fig.
 S2).
@@
 To examine the possibility that gene trees with little phylogenetic signal were diluting support for the Ortholog Conjecture, we removed the `r k_percentile` of genes with the lowest phylogenetic signal, corresponding to a $K$ cutoff for retained genes of greater than `r signif( k_thresh, 3)`.
-Repeating the phylogenetic analyses on this subset of genes with strongest phylogenetic signal for Tau still do not reject the null hypothesis (Wilcoxon $P=`r  wilcox_test_result_k %>% round( 3 )`$).
+Repeating the phylogenetic analyses on this subset of genes with strongest phylogenetic signal for tau still do not reject the null hypothesis (Wilcoxon $P=`r  wilcox_test_result_k %>% round( 3 )`$).
 This indicates that gene trees with little phylogenetic signal are not obscuring support for the Ortholog Conjecture in genes with greater phylogenetic signal.
@@
 Phylogenetic independent contrasts (PICs) are based on a Brownian Motion (BM) model of character evolution [@Felsenstein:1985ua].
-PICs could fail to reject the null hypothesis if BM is not a good model for the evolution of Tau on the gene phylogenies.
+PICs could fail to reject the null hypothesis if BM is not a good model for the evolution of tau on the gene phylogenies.
 We therefore compared the fit of BM to the more complex Ornstein Uhlenbeck (OU) model of evolution with the Akaike Information Criterion (AIC) [@Akaike:1974ih; @Posada:2004ix].
 
-Of the `r nrow(tree_summary)` genes considered, and a $\Delta$AIC threshold of `r dAIC_thresh`, `r sum(tree_summary$dAIC_bm < dAIC_thresh & tree_summary$dAIC_ou >= dAIC_thresh)` have $\Delta$AIC less than the threshold only for Brownian Motion (BM), `r sum(tree_summary$dAIC_bm >= dAIC_thresh & tree_summary$dAIC_ou < dAIC_thresh)` have $\Delta$AIC less than the threshold only for Ornstein Uhlenbeck (OU), and `r sum(tree_summary$dAIC_bm < dAIC_thresh & tree_summary$dAIC_ou < dAIC_thresh)` have $\Delta$AIC less than the threshold for both models.
-That gives a total of `r sum( tree_summary$dAIC_bm < dAIC_thresh )` genes that have $\Delta$AIC less than the threshold for BM and `r sum( tree_summary$dAIC_ou < dAIC_thresh )` genes that have $\Delta$AIC less than the threshold for OU (these sets are overlapping, so their total is more than the number of genes considered).
-
-Repeating the BM phylogenetic analyses on the subset of `r sum( tree_summary$dAIC_bm < dAIC_thresh )` genes that have $\Delta$AIC less than the threshold for BM still does not reject the null hypothesis (Wilcoxon $P=`r  wilcox_test_result_bm %>% round( 3 )`$).
-
-We next approximated the Phylogenetic Independent Contrasts for the the genes that have $\Delta$AIC less than the threshold for OU under an OU model.
-These analyses do not reject the null hypothesis (Wilcoxon $P=`r  wilcox_ou %>% round( 3 )`$).
-The challenge with such an analysis is that BM is intrinsic to the classic implementation of Phylogenetic Independent Contrasts, which use the fact that the expected variance is proportional to branch length to scale each contrast.
+We fit BM and OU models for the evolution of tau on each gene tree with the R package geiger [@Pennell:2014jf].
+This produced parameter estimates and AIC scores, $AIC_{BM}$ and $AIC_{OU}$, for each tree for each model.
+For each tree, we then calculated the AIC differences $\Delta AIC$ for each model as $\Delta AIC_{BM}=AIC_{BM}-min(AIC_{BM}, AIC_{OU})$ and $\Delta AIC_{OU}=AIC_{OU}-min(AIC_{BM}, AIC_{OU})$.
+The lower the $\Delta AIC$ for a model, the more plausible it is.
+$\Delta AIC<2$ is generally regarded as substantial relative support for a model.
+If multiple models have a low $\Delta AIC$, then there is little difference in the plausibility of the models.
+Since we are only considering two models, for each gene tree either $\Delta AIC_{BM}$ or $\Delta AIC_{OU}$ (or both) will be $0$.
+
+Of the `r nrow(tree_summary)` genes considered, `r sum(tree_summary$dAIC_bm < dAIC_thresh & tree_summary$dAIC_ou >= dAIC_thresh)` have $\Delta AIC_{BM} < `r dAIC_thresh`$ and $\Delta AIC_{OU} \geq `r dAIC_thresh`$, `r sum(tree_summary$dAIC_bm >= dAIC_thresh & tree_summary$dAIC_ou < dAIC_thresh)` have $\Delta AIC_{OU} < `r dAIC_thresh`$ and $\Delta AIC_{BM} \geq `r dAIC_thresh`$, and `r sum(tree_summary$dAIC_bm < dAIC_thresh & tree_summary$dAIC_ou < dAIC_thresh)` have both $\Delta AIC_{OU} < `r dAIC_thresh`$ and $\Delta AIC_{BM} < `r dAIC_thresh`$.
+That gives a total of `r sum( tree_summary$dAIC_bm < dAIC_thresh )` genes that have $\Delta AIC_{BM} < `r dAIC_thresh`$ and `r sum( tree_summary$dAIC_ou < dAIC_thresh )` genes that have $\Delta AIC_{OU} < `r dAIC_thresh`$ (these sets of genes are overlapping, so their total is more than the number of genes considered).
+
+Repeating the BM phylogenetic analyses on the subset of `r sum( tree_summary$dAIC_bm < dAIC_thresh )` genes for which $\Delta AIC_{BM} < `r dAIC_thresh`$ still does not reject the null hypothesis (Wilcoxon $P=`r  wilcox_test_result_bm %>% round( 3 )`$) that the rate of evolution of tau is the same after duplication and speciation events.
+
+We next approximated the phylogenetic independent contrasts for the the `r sum( tree_summary$dAIC_ou < dAIC_thresh )` genes for which $\Delta AIC_{OU} < `r dAIC_thresh`$.
+These analyses do not reject the null hypothesis either (Wilcoxon $P=`r  wilcox_ou %>% round( 3 )`$).
+The challenge with such an analysis is that BM is intrinsic to the classic implementation of phylogenetic independent contrasts, which use the fact that the expected variance is proportional to branch length to scale each contrast.
 Under OU, branch length no longer has a linear relationship to expected variance.
@@
 We validated the method by comparing picx BM results to the standard pic results calculated by ape [@Paradis:2004dv].
-These validations indicated a strong linear relationship between the two ($R^2=0.98$, $P=1.39\times 10^{−23}$).
-This approach could be extended to approximate Phylogenetic Independent Contrasts under other continuous models of evolution as well.
-
-These analyses indicate that our inability to reject the null hypothesis is robust under different models of Tau evolution.
+These validations indicated a strong linear relationship between the two ($R^2=0.98$, $P=1.39\times 10^{-23}$).
+This approach could be extended to approximate phylogenetic independent contrasts under other continuous models of evolution as well.
+
+These analyses indicate that our inability to reject the null hypothesis is robust under different models of tau evolution.
 
@@
 If any daughter nodes became older than their parent, we repeated the sampling until the dates were congruent with the topology.
-The minimum Wilcoxon $P$ in these reanalyses was $`r min( as.numeric( unlist( p_noised ) ) )`$, *i.e.* none of them reject the null hypothesis that the rate of evolution of Tau is greater following duplication events than speciation events.
+The minimum Wilcoxon $P$ in these reanalyses was $`r min( as.numeric( unlist( p_noised ) ) )`$, *i.e.* none of them reject the null hypothesis that the rate of evolution of tau is greater following duplication events than speciation events.
 This is consistent with the analysis that uses the calibration dates as specified, indicating that our results are robust to the selection of calibration times for speciation nodes.
@@
 
-#### Relationship between Tau and maximum expression
-
-```{r Fig_S6, fig.width=3.42, fig.height=3.42, fig.cap="Genes with higher maximum observed expression tend to have lower Tau. The blue line is the linear model for the relationship between the two."}
+#### Relationship between tau and maximum expression
+
+```{r Fig_S6, fig.width=3.42, fig.height=3.42, fig.cap="Genes with higher maximum observed expression tend to have lower tau. The blue line is the linear model for the relationship between the two."}
 
@@
 Each triplet has two paralogs that arose from a duplication event and one unduplicated homolog.
-They find that there is a tendency for the paralog with the lowest expression to have the highest Tau.
+They find that there is a tendency for the paralog with the lowest expression to have the highest tau.
 From this they conclude that following duplication there is a common trajectory of evolutionary change, where one paralog evolves to have lower expression and also become more tissue specific.
-There is, however, a negative relationship between Tau and maximum expression across all genes (*SI Appendix* Fig.
+There is, however, a negative relationship between tau and maximum expression across all genes (*SI Appendix* Fig.
 S6, $r^2=`r signif( tau_v_max_lm$r.squared, 3 )`$, $slope=`r signif( tau_v_max_lm$coefficients[2,1], 3 )`$, $P=`r signif( tau_v_max_lm$coefficients[2,4], 3 )`$).
-A simpler explanation for their plot is that it reflects the global tendency for Tau to decrease with maximum expression.
-The null expectation for any two genes sampled at random is for one to have higher maximum expression and lower Tau, and the other to have lower maximum expression and higher Tau.
-This global pattern could be due to both biological factors and the technical details of how Tau is calculated.
+A simpler explanation for their plot is that it reflects the global tendency for tau to decrease with maximum expression.
+The null expectation for any two genes sampled at random is for one to have higher maximum expression and lower tau, and the other to have lower maximum expression and higher tau.
+This global pattern could be due to both biological factors and the technical details of how tau is calculated.
 

### 2017-11-17 functions.R

@@ -301,7 +301,7 @@ pic.nhx = function( nhx, model_method="BM" ) {
 #' @export
 add_pics_to_trees = function( gene_trees_calibrated, model_method="BM" ) {
 	
-	# Calculate independent contrasts for Tau on each tree, storing the results 
+	# Calculate independent contrasts for tau on each tree, storing the results 
 	# back into the @data slot of the tree objects
 	gene_trees_pic = foreach( tree=gene_trees_calibrated ) %dopar% 
 		pic.nhx( tree, model_method=model_method )
@@ -486,10 +486,10 @@ wilcox_oc = function( nodes_contrast ) {
 }
 
 
-#' Estimate Tau evolution parameters from a tree::treedata tree
+#' Estimate tau evolution parameters from a tree::treedata tree
 #' and add them back into the tree object.
 #' 
-#' @param nhx A phylogenetic tree and associated Tau values 
+#' @param nhx A phylogenetic tree and associated tau values 
 #' as a treeio::treedata object
 #' @param ... Additional arguments to pass to geiger::fitContinuous()
 #' @return A phylogenetic trees as a treeio::treedata object, 
@@ -531,17 +531,17 @@ add_model_parameters = function( nhx, ... ) {
 }
 
 
-#' Simulate Tau on a tree::treedata. Existing observed values are
+#' Simulate tau on a tree::treedata. Existing observed values are
 #' fit to a brownian model to estimate parameters. These parameters
 #' are then used to replace the original values with similated 
 #' values.
 #' 
-#' @param nhx A phylogenetic tree and associated Tau values 
+#' @param nhx A phylogenetic tree and associated tau values 
 #' as a treeio::treedata object
 #' @param dup_adjust A multiplier for adjusting the branch lengths 
 #' following duplication to effectively change the rate following duplication
-#' @param a ancestral state for Tau. If not specified, use estimated value 
-#' @return A phylogenetic trees and simulated Tau values as a 
+#' @param a ancestral state for tau. If not specified, use estimated value 
+#' @return A phylogenetic trees and simulated tau values as a 
 #' treeio::treedata object
 #' @export
 sim_tau = function( nhx, dup_adjust=1, a=NA ) {

### 2017-11-20 manuscript.rmd
removed floating section header, added run time to sup info
revised implications section

@@
 Pairwise comparisons describe contemporary patterns, while phylogenetic methods infer historical processes [@OMeara:2012gq].
-There need not be a different process of evolution following speciation and duplication for paralogs to be more different than orthologs.
+Paralogs can be more different than orthologs even when the processes of evolution following speciation and duplication are the same.
 Any difference could be due to the structure of the gene phylogenies alone.
@@
 
+	edge_summary_pass_k = edge_summary %>% filter( gene %in% genes_pass_k )
+
+	edge_pass_k_p = wilcox.test( 
+		edge_summary_pass_k %>% filter( event_parent=="Duplication" ) %>% .$scaled_change, 
+		edge_summary_pass_k %>% filter( event_parent=="Speciation" ) %>% .$scaled_change, 
+		alternative="greater" 
+	)$p.value
+
 ```
@@
 While some studies have presented support for the ortholog conjecture, our results are consistent with multiple studies that have not [@Studer:2009bm; @Nehrt:2011bm; @Yanai:2004hk].
-Since the ortholog conjecture can be applied to many different gene attributes at both the genotype level (*e.g.*, molecular sequence) and phenotype level (*e.g.*, activity, biochemical function, binding partners, and expression), it is naturally a heterogeneous and potentially idiosyncratic topic [@Soria:2014bv].
-For example, a recent integrated analysis of multiple gene traits after duplication found extensive divergence in protein sequence but less divergence in expression than expected [@Soria:2014bv].
-
-Our results suggest, at a minimum, that the ortholog conjecture is not a dominant pattern that is central to explaining the evolution of phenotypic diversity in gene families.
-This suggests that an alternative "neutral conjecture", *i.e.* the conjecture that the evolution of gene traits tends to be the same following gene duplication and speciation events, may better explain the process and patterns of most gene evolution.
-Under this neutral conjecture, phylogenetic distance is a better predictor of the similarity of gene function than is the history of gene duplication and speciation.
-The ortholog conjecture does not have to be an all or nothing process, though.
-It may be the case that the rates of phenotypic evolution following duplication may be greater than that following duplication in some organisms, gene families, and evolutionary processes [@Gabaldon:2013fi].
-We just do not find evidence for it when summarizing gene expression across tissues with tau in these organisms.
-This calls into question the general predictive power of the ortholog conjecture with respect to gene expression, and until these processes are better understood it will be necessary to test for it in each situation.
-These tests should be articulated in terms of clear alternative hypotheses [@Studer:2009bm] that make distinct phylogenetic comparative predictions.
-
-Lack of support for the ortholog conjecture has important biological implications.
-It indicates that the mechanism of gene divergence (speciation versus duplication) may not have as strong an impact on phenotypic divergence as sometimes proposed.
-It also has important technical implications.
-Having information on whether two genes are orthologs or paralogs provides little added information about expression beyond knowing how distantly related the two genes are.
-Rather than focus on whether genes are orthologs or paralogs when attempting to predict function, it may be more effective to simply focus on how closely related or distantly related they are.
-Closely related paralogs, for example, may tend to have more similar phenotypes than more distantly related orthologs [@Studer:2009bm].
-
-This is also an example of the limitations of the concepts of orthology and paralogy [@Dunn:2016er].
-These terms can have straightforward meaning in small gene trees with simple duplication/speciation histories, but the utility of the terms breaks down on larger more complex gene trees.
-Orthology and paralogy are annotations on the tips of the phylogeny that are derived from the structure of the tree and history of duplication and speciation at internal tree nodes.
-In this sense, orthology and paralogy are statements about the internals of the tree that are distilled into statements at the tips of the tree.
-Much is lost in the process, though.
-For most questions it is much more direct to focus on the structure of the tree and the inferred processes within the tree, such as which internal nodes are duplication or speciation events and how much change occurs along the branches.
+It should be noted that many areas of biology restrict the term "function" to describe the biochemical function of a gene, but the literature on gene duplication often uses "function" in a broader sense to include any gene attribute that could impact fitness, such as biochemical function, spatial expression, or binding.
+Since the ortholog conjecture can be applied to such diverse attributes, it is a heterogeneous and potentially idiosyncratic topic.
+For example, a recent integrated analysis of multiple gene traits after duplication found extensive divergence in protein sequence but less divergence in expression [@Soria:2014bv].
+In our analyses, we found that phylogenetic distance is a better predictor of the similarity of gene expression than is the history of gene duplication and speciation.
+The ortholog conjecture does not have to be an all or nothing proposition, even as it applies to expression.
+It may be that the rates of tissue-specific expression evolution following duplication are greater in some organisms, gene families, and evolutionary processes [@Gabaldon:2013fi].
+The lack of support we find for the ortholog conjecture in this system, as well as the mixed results of others, does suggests that investigators should test for it in each situation rather than assume *a priori* that it is a dominant pattern in the diversity of gene expression.
+These tests should be done in an explicit phylogenetic framework.
+
+While empirical support for the ortholog conjecture has been mixed, theoretical work has suggested several mechanisms that could lead to increased rates of evolutionary change in expression and other gene traits following gene duplication.
+This has been a primary motivation for the extensive discussion of the OC.
+In his seminal work, Ohno [@Ohno:1970wm] outlined three outcomes of gene duplication, now often referred to as neofunctionalization (one gene copy can take on new functions while the other retains the old functions), subfunctionalization (each gene copy has a subset of the functions present in their most recent common ancestor), and conservation (each gene copy retains the same function).
+Neofunctionalization was the primary focus of early work on gene duplication, though it was not found to be a widespread outcome of duplication [@Hughes:1994ec].
+The highly influential duplication-degeneration-complementation (DDC) model [@Force:1999tg; @Lynch:2000um] suggested that subfunctionalization could be a better explanation for the retention of both copies of a gene.
+It is built on the idea that gene copies that have each lost different functions would both be needed to fulfill all the functions present in their most recent common ancestor.
+Tests of the DDC model have had mixed results [@Huminiecki:2004dc].
+A growing body of work suggests that retention and differentiation of genes is due to a variety of different mechanisms in different contexts, and that we should not necessarily expect one process to dominate over others [@Hahn:2009by].
+Our lack of support for the ortholog conjecture here suggests that whatever mechanisms lead to the retention of duplicated genes in the organisms considered here does not lead to changes in the tissue-specificity of gene expression as summarized by tau.
+
+The DDC model is particularly interesting to further consider in this context since the summary statistic tau describes tissue specificity of expression.
+In the context of tissue-specific expression, the OC predicts a greater rate of evolutionary change in tau following duplication, without specifying the direction of the change.
+The DDC model predicts a specific type of change in expression following duplication -- subfunctionalization, which would increase tissue specificity, and therefore tau, following duplication.
+To test for this more specific predictions of DDC, we conducted additional analyses to see if there is a greater rate of increase in tau following duplication.
+For each branch in each gene tree, we inferred the change in tau and standardized this change by branch lengths.
+We found no evidence of increased tau following duplication relative to speciation (Wilcoxon $P=`r  edge_p %>% round( 3 )`$).
+This result also holds when we discarded the `r k_percentile` of genes with the lowest phylogenetic signal (Wilcoxon $P=`r  edge_pass_k_p %>% round( 3 )`$).
+These results are consistent with another recent study that found little support for neofunctionalization or subfunctionalization in expression of duplicated human genes, and instead proposed that dosage compensation drives most changes in expression following duplication in mammals [@Lan:2016ix].
+Like the OC, DDC is not an all-or-nothing hypothesis.
+We fail to find support for it in these tissues in these species with the tau expression summary statistic, but it may hold in other contexts.
+DDC could still play a role in the retention of these genes, but with respect to other attributes such as biochemical functions rather than tissue-specific expression.
 
@@
 
-#### Origin of pairwise correlation plot structure
-
 ```{r Fig_S5, fig.width=7, fig.height=3, fig.cap="Origin of pairwise correlation plot structure.
@@
 These analyses were computed on `r format( system_time_kernel, "%a %b %d %X %Y" )` with the following R package versions.
+The total run time of these analyses was `r time_run %>% as.numeric() %>% round( 2 )` hours on a `r cores+1` core machine.
 

### 2017-11-20 functions.R
corrected scaling

@@ -367,7 +367,7 @@ summarize_nhx_edges = function( nhx ){
 	events_on_edges = sapply( phy$edge, function(x){ event_type[x] })
 	dim( events_on_edges ) = dim( phy$edge )
 	
-	scaled_change = (tau_child - tau_parent) / sqrt( edge_length )
+	scaled_change = (tau_child - tau_parent) / edge_length
 	
 	tibble(
 		gene = digest( nhx ),

### 2017-11-20 manuscript.rmd
spelled out OC

@@
 While empirical support for the ortholog conjecture has been mixed, theoretical work has suggested several mechanisms that could lead to increased rates of evolutionary change in expression and other gene traits following gene duplication.
-This has been a primary motivation for the extensive discussion of the OC.
+This has been a primary motivation for the extensive discussion of the ortholog conjecture.
 In his seminal work, Ohno [@Ohno:1970wm] outlined three outcomes of gene duplication, now often referred to as neofunctionalization (one gene copy can take on new functions while the other retains the old functions), subfunctionalization (each gene copy has a subset of the functions present in their most recent common ancestor), and conservation (each gene copy retains the same function).
@@
 The DDC model is particularly interesting to further consider in this context since the summary statistic tau describes tissue specificity of expression.
-In the context of tissue-specific expression, the OC predicts a greater rate of evolutionary change in tau following duplication, without specifying the direction of the change.
+In the context of tissue-specific expression, the ortholog conjecture predicts a greater rate of evolutionary change in tau following duplication, without specifying the direction of the change.
 The DDC model predicts a specific type of change in expression following duplication -- subfunctionalization, which would increase tissue specificity, and therefore tau, following duplication.
@@
 These results are consistent with another recent study that found little support for neofunctionalization or subfunctionalization in expression of duplicated human genes, and instead proposed that dosage compensation drives most changes in expression following duplication in mammals [@Lan:2016ix].
-Like the OC, DDC is not an all-or-nothing hypothesis.
+Like the ortholog conjecture, DDC is not an all-or-nothing hypothesis.
 We fail to find support for it in these tissues in these species with the tau expression summary statistic, but it may hold in other contexts.

### 2017-11-21 manuscript.rmd

@@
 Since the ortholog conjecture can be applied to such diverse attributes, it is a heterogeneous and potentially idiosyncratic topic.
-For example, a recent integrated analysis of multiple gene traits after duplication found extensive divergence in protein sequence but less divergence in expression [@Soria:2014bv].
+For example, a recent integrated analysis of multiple gene attributes after duplication found extensive divergence in protein sequence but less divergence in expression [@Soria:2014bv].
 In our analyses, we found that phylogenetic distance is a better predictor of the similarity of gene expression than is the history of gene duplication and speciation.
@@
 It may be that the rates of tissue-specific expression evolution following duplication are greater in some organisms, gene families, and evolutionary processes [@Gabaldon:2013fi].
-The lack of support we find for the ortholog conjecture in this system, as well as the mixed results of others, does suggests that investigators should test for it in each situation rather than assume *a priori* that it is a dominant pattern in the diversity of gene expression.
+The lack of support we find for the ortholog conjecture in this system, as well as the mixed results of others, does suggest that investigators should test for it in each situation rather than assume *a priori* that it is a dominant pattern in the diversity of gene expression.
 These tests should be done in an explicit phylogenetic framework.
@@
 A growing body of work suggests that retention and differentiation of genes is due to a variety of different mechanisms in different contexts, and that we should not necessarily expect one process to dominate over others [@Hahn:2009by].
-Our lack of support for the ortholog conjecture here suggests that whatever mechanisms lead to the retention of duplicated genes in the organisms considered here does not lead to changes in the tissue-specificity of gene expression as summarized by tau.
+Our lack of support for the ortholog conjecture suggests that whatever mechanisms lead to the retention of duplicated genes in the organisms considered here does not lead to changes in the tissue-specificity of gene expression as summarized by tau.
 
@@
 The DDC model predicts a specific type of change in expression following duplication -- subfunctionalization, which would increase tissue specificity, and therefore tau, following duplication.
-To test for this more specific predictions of DDC, we conducted additional analyses to see if there is a greater rate of increase in tau following duplication.
+To test for this more specific prediction of DDC, we conducted additional analyses to see if there is a greater rate of increase in tau following duplication.
 For each branch in each gene tree, we inferred the change in tau and standardized this change by branch lengths.

### 2017-12-18 manuscript.rmd
fixed a citation in the rmd, did not rebuild pdf

@@
 
-The first study, Kryuchkova-Mostacci and Robinson-Rechavi (KMRR) [@KryuchkovaMostacci:2016iw], analyzed multiple vertebrate expression datasets to test the ortholog conjecture - the hypothesis that orthologs tend to have more conserved attributes (specificity of expression across organs in this case) than do paralogs [@Nehrt:2011bm, @Koonin:2005fp].
+The first study, Kryuchkova-Mostacci and Robinson-Rechavi (KMRR) [@KryuchkovaMostacci:2016iw], analyzed multiple vertebrate expression datasets to test the ortholog conjecture - the hypothesis that orthologs tend to have more conserved attributes (specificity of expression across organs in this case) than do paralogs [@Nehrt:2011bm; @Koonin:2005fp].
 Using pairwise comparisons (Fig.

### 2018-01-04 manuscript.rmd
fixed another citation in the rmd, did not rebuild pdf
updated with paper citation and docker image link

@@
 KMRR considered several publicly available datasets of gene expression across tissues and species.
-Their expression summary statistic is tau [@Yanai:2005dh, @KryuchkovaMostacci:2016in], an indicator of tissue specificity of gene expression.
+Their expression summary statistic is tau [@Yanai:2005dh; @KryuchkovaMostacci:2016in], an indicator of tissue specificity of gene expression.
 Tau can range from a value of 0, which indicates no specificity (*i.e.*, uniform expression across tissues), to a value of 1, which indicates high specificity (*i.e.*, expression in only one tissue).
@@
 To test the hypothesis that ancient duplications that precede the oldest speciation event (*SI Appendix*, Fig.
-S5*A*) impact the lower correlation of tau between paralogs than between orthologs, we removed them.
+S3*A*) impact the lower correlation of tau between paralogs than between orthologs, we removed them.
 When we consider only the duplication events the same age or younger than the oldest speciation event (*SI Appendix*, Fig.
-5*B*), the paralog correlation coefficient increases from `r signif( paralog_r$estimate, 2)` to `r signif( paralog_subset_r$estimate, 2)` ($P`r format_p(paralog_subset_r)`$).
+S3*B*), the paralog correlation coefficient increases from `r signif( paralog_r$estimate, 2)` to `r signif( paralog_subset_r$estimate, 2)` ($P`r format_p(paralog_subset_r)`$).
 This is much closer to the ortholog correlation of `r signif( ortholog_r$estimate, 2) `.
@@
 If multiple models have a low $\Delta AIC$, then there is little difference in the plausibility of the models.
-Since we are only considering two models, for each gene tree either $\Delta AIC_{BM}$ or $\Delta AIC_{OU}$ (or both) will be $0$.
+Since we are only considering two models, for each gene tree either $\Delta AIC_{BM}$ or $\Delta AIC_{OU}$ will be $0$.
 
